### 2021-12-21 paper.tex
First commit

\documentclass[]{article}
\usepackage[round]{natbib}

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}

% lorem ipsum dummy text
\usepackage{lipsum}

\lstset{language=Python}

% local definitions
\newcommand{\comment}[1]{{\textcolor{red}{Comment: #1}}}

% cross-reference with supplement
\usepackage{xr}
\externaldocument{supplement}

\begin{document}

\title{A weakly structured stem for human origins in Africa}
\author[1]{Aaron P. Ragsdale}
\author[2]{Timothy D. Weaver}
\author[3]{Elizabeth G. Atkinson}
\author[4]{Eileen Hoal}
\author[5]{Marlo M\"{o}ller}
\author[2,6,*]{Brenna M. Henn}
\author[7,**]{Simon Gravel}
\affil[1]{FIXME: UW Madison}
\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA 95616, USA}
\affil[3]{FIXME: DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
\affil[4]{FIXME}
\affil[5]{FIXME}
\affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA 95616, USA}
\affil[7]{FIXME: McGill}
\affil[*]{bmhenn@ucdavis.edu}
\affil[**]{simon.gravel@mcgill.ca}
\maketitle

\abstract{
A very simple template for an article class document.
}

\section*{Introduction}

\section*{Results}

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
\caption{\textbf{The first figure}.
    It has a caption.
}
\label{fig:1}
\end{center}
\end{figure}

\subsection*{A Late Middle Stone Age common ancestry for contemporary human populations}

\subsection*{Evidence for deep population structure, but not archaic admixture within Africa}

\subsection*{Reconciling multiple lines of genetic evidence}

\section*{Discussion}

\subsection*{The Middle Stone Age in Africa}

\subsection*{Contrasting archaic admixture and a weakly structured stem}

\section*{Methods}
\cite{Kelleher2016-lw}

\section*{Acknowledgements}

\break

\bibliographystyle{plainnat}
\bibliography{paper}
\end{document}

### 2021-12-21 supplement.tex
First commit

\documentclass[]{article}
\usepackage[round]{natbib}

\usepackage{fullpage}
\usepackage{listings}
\usepackage{url}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}

% lorem ipsum dummy text
\usepackage{lipsum}

\lstset{language=Python}

% cross-reference with main text
\usepackage{xr}
\externaldocument{paper}

% local definitions
\newcommand{\comment}[1]{{\textcolor{red}{Comment: #1}}}

\begin{document}
\title{Supporting Information for ``A weakly structured stem for human origins in Africa''}
\author{Aaron P. Ragsdale, Timothy D. Weaver, Brenna M. Henn, and Simon Gravel}
\date{\today}
\maketitle

\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{equation}{0}

\tableofcontents
\newpage

\section{Data and sequencing}

\subsection{Sequencing and variant calling}
\cite{Kelleher2016-lw}

\subsection{Nama sample collection and consent}

\subsection{Details about populations used in analyses}

\subsection{Filtering and subsetting data}

\section{Computing statistics}

\subsection{Computing LD and diversity statistics}

\subsection{Estimating two-locus statistics with small sample sizes}

\subsection{Computing conditional SFS}

\section{Model specification and fitting}

\subsection{General strategy for building models and introducing complexity}

\subsection{Optimization using moments}

\subsection{Confidence intervals using godambe methods}

\section{Gene genealogy reconstruction}

\section{Predictions from inferred demographic models}

\subsection{$F_{ST}$ between coexisting populations over time}

\subsection{$f_4$ statistics between pairs of contemporary and pairs of ancient populations}

\section{Validations using simulatinos from inferred demographic models}

\subsection{Simulation details}

\subsection{cSFS prediction under inferred models}

\subsection{Relate curves from inferred models}

\subsection{Distribution of deep branch affinities to Neanderthal sequence}

\subsection{Mutation versus recombination rates}

\break

\addcontentsline{toc}{section}{References}
\bibliographystyle{plainnat}
\bibliography{paper}

\break

\addcontentsline{toc}{section}{Supporting figures and tables}
\section*{Supporting figures and tables}

\end{document}

### 2021-12-22 supplement.tex
supp tables started

@@
 % local definitions
-\newcommand{\comment}[1]{{\textcolor{red}{Comment: #1}}}
+\newcommand{\sgcomment}[1]{{\textcolor{red}{SG: #1}}}
+\newcommand{\bmhcomment}[1]{{\textcolor{blue}{BMH: #1}}}
+\newcommand{\aprcomment}[1]{{\textcolor{magenta}{APR: #1}}}
+\newcommand{\tdwcomment}[1]{{\textcolor{cyan}{TDW: #1}}}
 
@@
 
-\addcontentsline{toc}{section}{Supporting figures and tables}
-\section*{Supporting figures and tables}
+\addcontentsline{toc}{section}{Supporting tables and figures}
+\section*{Supporting tables and figures}
+
+\begin{table}[ht]
+\caption{
+    \label{tab:single_origin}
+    \textbf{Best-fit parameters from the Single-Origin model.}
+    \aprcomment{fill in caption}
+}
+\centering
+\begin{tabular}[t]{lp{8cm}ll}
+    \toprule
+    Parameter & Description & Value & Std. err.\\
+    \midrule
+    $N_e$ & Ancestral effective population size & 10198 & 403 \\
+    $N_{MH}$ & Size of modern-human lineage between Neanderthal and Nama splits & 21111 & 529 \\
+    $N_{Nama_0}$ & Initial Nama size & 10224 & 370 \\
+    $N_{Nama_F}$ & Final Nama size & 222 & 9 \\
+    $N_{MSL_0}$ & Initial Mende size & 17211 & 769 \\
+    $N_{MSL_F}$ & Final Mende size & 16822 & 606 \\
+    $N_{EA}$ & Size of East African Branch & 7139 & 273 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3831 & 131 \\
+    $N_{EP}$ & East African agriculturist size & 13033 & 491 \\
+    $N_{GBR_0}$ & Initial British size & 846 & 33 \\
+    $N_{GBR_F}$ & Final British size & 12121 & 507 \\
+    $N_{Neand}$ & Neanderthal size & 1867 & 105 \\
+    $T_{Nama}$ & Nama split time (years) & 110400 & 2525 \\
+    $m_{Nama-MSL}$ & Nama-Mende symmetric migration rate & $2.82\times10^{-5}$ & $0.158\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama-East Africa symmetric migration rate & $4.94\times10^{-5}$ & $0.197\times10^{-5}$ \\
+    \bottomrule
+\end{tabular}
+\end{table}
 

### 2021-12-22 supplement.tex
fill in best-fit tables

--- 2021-12-21 supplement.tex
@@
 % local definitions
-\newcommand{\comment}[1]{{\textcolor{red}{Comment: #1}}}
+\newcommand{\sgcomment}[1]{{\textcolor{red}{SG: #1}}}
+\newcommand{\bmhcomment}[1]{{\textcolor{blue}{BMH: #1}}}
+\newcommand{\aprcomment}[1]{{\textcolor{magenta}{APR: #1}}}
+\newcommand{\tdwcomment}[1]{{\textcolor{cyan}{TDW: #1}}}
+
+\usepackage{array}
+\newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
 
@@
 
-\addcontentsline{toc}{section}{Supporting figures and tables}
-\section*{Supporting figures and tables}
+\addcontentsline{toc}{section}{Supporting tables}
+\section*{Supporting tables}
+
+\begin{table}[ht]
+\caption{
+    \label{tab:single_origin}
+    \textbf{Best-fit parameters from the Single-Origin model.}
+    \aprcomment{fill in caption - generation time of 29 years, other details}
+}
+\centering
+\begin{tabular}[t]{rP{8cm}cc}
+    \toprule
+    Parameter & Description & Value & Std. err.\\
+    \midrule
+    $N_e$ & Ancestral effective population size & 10198 & 403 \\
+    $N_{MH}$ & Size of modern-human lineage between Neanderthal and Nama splits & 21111 & 529 \\
+    $N_{Nama_0}$ & Initial Nama size & 10224 & 370 \\
+    $N_{Nama_F}$ & Final Nama size & 222 & 9 \\
+    $N_{MSL_0}$ & Initial Mende size & 17211 & 769 \\
+    $N_{MSL_F}$ & Final Mende size & 16822 & 606 \\
+    $N_{EA}$ & Size of East African branch & 7139 & 273 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3831 & 131 \\
+    $N_{EP}$ & East African agriculturist size & 13033 & 491 \\
+    $N_{GBR_0}$ & Initial British size & 846 & 33 \\
+    $N_{GBR_F}$ & Final British size & 12121 & 507 \\
+    $N_{Neand}$ & Neanderthal size & 1867 & 105 \\
+    $T_{Nama}$ & Nama split time (years) & 110400 & 2525 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $2.82\times10^{-5}$ & $0.158\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.94\times10^{-5}$ & $0.197\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $18.76\times10^{-5}$ & $0.764\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.42\times10^{-5}$ & $0.239\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $41.28\times10^{-5}$ & $1.33\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.658 & 0.0039 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.279 & 0.0039 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.150 & 0.0019 \\
+    \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[ht]
+\caption{
+    \label{tab:continuous_migration}
+    \textbf{Best-fit parameters from the Continuous-Migration model.}
+    \aprcomment{fill in caption - generation time of 29 years, other details}
+}
+\centering
+\begin{tabular}[t]{rP{8cm}cc}
+    \toprule
+    Parameter & Description & Value & Std. err.\\
+    \midrule
+    $N_e$ & Ancestral effective population size & 7270 & 1777 \\
+    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 8256 & 1612 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 13547 & 2488 \\
+    $N_{Nama_0}$ & Initial Nama size & 11939 & 2989 \\
+    $N_{Nama_F}$ & Final Nama size & 221 & 54 \\
+    $N_{MSL_0}$ & Initial Mende size & 9738 & 2479 \\
+    $N_{MSL_F}$ & Final Mende size & 28150 & 6628 \\
+    $N_{EA}$ & Size of East African branch & 7489 & 1841 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3728 & 915 \\
+    $N_{EP}$ & East African agriculturist size & 13072 & 3246 \\
+    $N_{GBR_0}$ & Initial British size & 959 & 231 \\
+    $N_{GBR_F}$ & Final British size & 11822 & 2889 \\
+    $N_{Neand}$ & Neanderthal size & 2670 & 591 \\
+    $T_{stems}$ & Stem split time (years) & 1163072 & 390803 \\
+    $T_{Nama}$ & Nama split time (years) & 134745 & 17775 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.98\times10^{-5}$ & $0.366\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.08\times10^{-5}$ & $1.02\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $21.4\times10^{-5}$ & $5.32\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.17\times10^{-5}$ & $1.02\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $33.6\times10^{-5}$ & $8.35\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.642 & 0.0037 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.255 & 0.0043 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.0021 \\
+    $m_{stems}$ & Stem 1--stem 2 migration rate & $6.43\times10^{-5}$ & $1.05\times10^{-5}$ \\
+    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $5.82\times10^{-5}$ & $1.60\times10^{-5}$ \\
+    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $16.4\times10^{-5}$ & $4.19\times10^{-5}$ \\
+    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $3.10\times10^{-5}$ & $0.901\times10^{-5}$ \\
+    \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[ht]
+\caption{
+    \label{tab:merger_without_stem_migration}
+    \textbf{Best-fit parameters from the Merger-Without-Stem-Migration model.}
+    \aprcomment{fill in caption - generation time of 29 years, other details}
+}
+\centering
+\begin{tabular}[t]{rP{8cm}cc}
+    \toprule
+    Parameter & Description & Value & Std. err.\\
+    \midrule
+    $N_e$ & Ancestral effective population size & 11258 & 326 \\
+    $N_{stem1}$ & Size of stem 1 lineage between stem 1--stem 2 split and stem 1E--stem 1S split & 113 & 76 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 23984 & 1149 \\
+    $N_{Nama_0}$ & Initial Nama and stem 1S size & 13134 & 384 \\
+    $N_{Nama_F}$ & Final Nama size & 225 & 7.3 \\
+    $N_{MSL_0}$ & Initial Mende size & 11856 & 322 \\
+    $N_{MSL_F}$ & Final Mende size & 25558 & 987 \\
+    $N_{EA}$ & Size of East African and stem 1E branch & 9136 & 246 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3385 & 102 \\
+    $N_{EP}$ & East African agriculturist size & 13650 & 408 \\
+    $N_{GBR_0}$ & Initial British size & 931 & 29 \\
+    $N_{GBR_F}$ & Final British size & 12064 & 334 \\
+    $N_{Neand}$ & Neanderthal size & 1935 & 91 \\
+    $T_{stems}$ & Stem split time (years) & 420881 & 27380 \\
+    $T_{stem1}$ & Stem 1 split time into stem 1E and stem 1S (years) & 367434 & 19952 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.361\times10^{-5}$ & $0.113\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.00\times10^{-5}$ & $0.130\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.5\times10^{-5}$ & $0.548\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.77\times10^{-5}$ & $0.152\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $37.1\times10^{-5}$ & $1.26\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.647 & 0.0037 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.257 & 0.0042 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.0021 \\
+    $T_{Nama}$ & Time of Nama merger event & 117392 & 8253 \\
+    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancsestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.707 & 0.0086 \\
+    $T_{EA}$ & Time of East Africa merger event & 94892 & 3648 \\
+    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.481 & 0.0074 \\
+    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 23922 & 570 \\
+    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.168 & 0.0036 \\
+    \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[ht]
+\caption{
+    \label{tab:merger_with_stem_migration}
+    \textbf{Best-fit parameters from the Merger-With-Stem-Migration model.}
+    \aprcomment{fill in caption - generation time of 29 years, other details}
+}
+\centering
+\begin{tabular}[t]{rP{8cm}cc}
+    \toprule
+    Parameter & Description & Value & Std. err.\\
+    \midrule
+    $N_e$ & Ancestral effective population size & 11479 & 1369 \\
+    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal split and stem 1E--stem 1S split & 117 & 838 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 24393 & 6668 \\
+    $N_{Nama_0}$ & Initial Nama size & 13211 & 1514 \\
+    $N_{Nama_F}$ & Final Nama size & 223 & 31 \\
+    $N_{MSL_0}$ & Initial Mende size & 11444 & 1165 \\
+    $N_{MSL_F}$ & Final Mende size & 27417 & 4332 \\
+    $N_{EA}$ & Size of East African Branch & 9077 & 1628 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3402 & 337 \\
+    $N_{EP}$ & East African agriculturist size & 13506 & 1684 \\
+    $N_{GBR_0}$ & Initial British size & 953 & 122 \\
+    $N_{GBR_F}$ & Final British size & 12406 & 1678 \\
+    $N_{Neand}$ & Neanderthal size & 2416 & 235 \\
+    $T_{stems}$ & Stem split time (years) & 1442022 & 426449 \\
+    $T_{stem1}$ & Stem 1S--stem 1E split time (years) & 479401 & 166339 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.712\times10^{-5}$ & $0.401\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.35\times10^{-5}$ & $0.912\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.8\times10^{-5}$ & $2.57\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.87\times10^{-5}$ & $0.550\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.9\times10^{-5}$ & $5.36\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.640 & 0.0075 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.257 & 0.0049 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.157 & 0.0031 \\
+    $m_{stems}$ & Stem 1--stem 2 migration rate & $11.6\times10^{-5}$ & $8.74\times10^{-5}$ \\
+    $T_{Nama}$ & Time of Nama merger event & 118547 & 28170 \\
+    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancsestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.714 & 0.067 \\
+    $T_{EA}$ & Time of East Africa merger event & 98083 & 8865 \\
+    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.495 & 0.059 \\
+    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 25119 & 641 \\
+    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.181 & 0.0085 \\
+    \bottomrule
+\end{tabular}
+\end{table}
+
+\break
+
+\addcontentsline{toc}{section}{Supporting figures}
+\section*{Supporting figures}
 

### 2021-12-31 supplement.tex
fill in a few things
paper edits from a few days ago

@@
 \begin{document}
-\title{Supporting Information for ``A weakly structured stem for human origins in Africa''}
+\title{Supporting Information for\\
+``A weakly structured stem for human origins in Africa''}
 \author{Aaron P. Ragsdale, Timothy D. Weaver, Brenna M. Henn, and Simon Gravel}
@@
 \subsection{Sequencing and variant calling}
-\cite{Kelleher2016-lw}
+
+Low coverage (4-8x) Illumina short read data were generated for the Nama,
+Gumuz, Amhara and Oromo populations as part of the African Diversity Reference
+Panel (Sanger / Wellcome Trust) \citep{Gurdasani2015-qy,Pagani2015-pz} and
+approved through a secondary data analysis agreement for this project. Briefly,
+raw reads were aligned to GRCH37 with BWAmem \aprcomment{cite}, duplicates
+marked with Picard MarkDuplicates, reads were realigned around indels with GATK
+RealignerTargetCreater / IndelRealigner followed by BQSR with dbSNP 137.
+Contamination checks were performed requiring that FREEMIX $<0.05$;
+contamination checks resulted in the elimination of 22 Nama samples. We note
+that the high heterozygosity in these genomes both due to inherent genetic
+diversity and admixture may have violated base assumptions for this
+heterozygosity check. Genomes were then variant called with GATK3.2 Unified
+Genotyper \aprcomment{cite} using joint calling across 2,478 individuals within
+the ADRP dataset with a minimum base quality of 17. Data were merged with 1000
+Genomes Phase 3 \citep{1000_Genomes_Project_Consortium2015-zq} using the union
+of sites identified with bcftools isec (-n+1) \aprcomment{cite} then refined
+with Variant Recalibrator with a truth sensitivity threshold of 99.5\%. HapMap
+III and dbSNP 138 served as known sites while 1000 Genomes Phase 1 Omni2.5 and
+Phase 1 genomic SNP served as the training set. After VR, no batch effects were
+observed along PC1 and PC2 for 1000 Genomes vs. ADRP. Phasing on the combined
+dataset was performed via SHAPEIT2 \aprcomment{cite} and utilized the duoHMM
+option for duos and trios. We highlight 82 Nama genomes which are newly
+available (unpublished) under accession number EGAD00001006198. Among these 82
+Nama samples, we down-sampled the individuals to minimize close relatives and
+admixture, such that 44 Nama genomes were retained. 2nd and 3rd degrees
+relatives were inferred from reconstructed pedigrees with Omni2.5 SNP array
+data. Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
+analysis, after partitioning ancestry into $k=6$ clusters with ADMIXTURE
+\aprcomment{cite} where alternative ancestries represent European, West
+African, Near Eastern, and East African gene flow.
 
@@
 
+DNA samples were collected in South Africa (Richtersveld) with IRB approval
+from Stanford University, Stellenbosch University (ZA) and SUNY Stony Brook.
+\aprcomment{Does this need to be expanded on at all?}
+
 \subsection{Details about populations used in analyses}
+
+\sgcomment{To be expanded. TODO: Brenna}
+\begin{itemize}
+    \item From the merged dataset of the African Diversity Reference Panel and
+        1000 Genomes phase 3 data, subsampled to population included in this study
+    \item West Africa: Mende from Sierra Leone (MSL); South Africa: newly sequenced 
+        Nama; East Africa: Gumuz (traditionally hunter-gatherer with low levels of 
+        Eurasian admixture), Oromo and Amhara (combined as traditionally 
+        agriculturalists with large proportion of back-to-Africa Eurasian ancestry); 
+        Eurasian: British (GBR)
+    \item Combined with the high coverage Vindija neanderthal genome
+    \item For running Relate, we used a larger set of populations. We kept all 
+        African 1000 Genomes populations, along with the Nama, Gumuz, Oromo, and
+        Amhara, as well as 4 Eurasian populations from 1000 Genomes: GBR, CEU, PJL,
+        and CHB
+\end{itemize}
 
@@
 
+All analyses presented in this work focus on biallelic single nucleotide
+polymorphisms within the 1000 Genomes Phase 3 strict mask.
+
+For the moments-LD analysis, we focused on intergenic locations because these
+appear less affected by natural selection compared to both synonymous and
+nonsynonymous variation \citep{Ragsdale2018-dd}. To enable comparison with
+Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
+had less than 100 contiguous base pairs.
+
 \section{Computing statistics}
+
+\subsection{LD and diversity statistics used in model fits}
+
+\aprcomment{$D^2$, $D_z$, $\pi_2$, and $H$. Definitions, notation,
+indexing\ldots}
 
@@
 
+We compared single- and two-locus statistics in the data to predictions based
+on detailed demographic models. Model predictions were obtained using
+recursions described in \citet{Ragsdale2019-nt} and implemented in the software
+\emph{moments.LD} (\url{https://bitbucket.org/simongravel/moments/src/main/}).
+The model computes expected patterns of single-nucleotide pairwise diversity
+and linkage disequilibrium as a function of recombination distance between
+variants within and across populations, under the assumption of neutrality.
+
+For numerical convenience, observed genetic variants were binned by
+recombination distances. We assessed the robustness of the statistics to errors
+in the recombination maps by considering two different recombination maps, the
+OMNI YRI and HapMapII
+\citep{1000_Genomes_Project_Consortium2015-zq,International_HapMap_Consortium2007-vn}.
+Statistics were largely unchanged by using a different map
+(Figure~\ref{fig:sup-map_comparison}). \aprcomment{These maps are both inferred
+using array data, which is sparse. Comment on this.}
+
+We removed bins of recombination distance less than a recombination distance of
+$r = 5\times10^{-6}$ (at a rough estimate of 1 cM/Mb, this corresponds to a minimum
+distance of 500 bp on average) to avoid previously reported biases at short
+distances due to processes like multinucleotide mutations
+\citep{Harris2014-zg,Ragsdale2019-nt}. To avoid uncertainty in phasing, we used
+unphased genotypes to compute LD statistics, as proposed in
+\citet{Ragsdale2020-nz}.
+
+Finally, we estimated uncertainty due to the finite amount of genetic material
+used in inference using bootstrap over 500 segments along the genome with
+roughly equal lengths of retained sequences within each segment. First, for
+each distance bin, we used these bootstrap samples to obtain a
+variance-covariances matrix across all statistics. This variance-covariance
+matrix was used to obtain a model likelihood for each recombination distance
+bin and single-locus nucleotide diversity, as a multivariate Gaussian
+likelihood. The full model likelihood was taken as the product of likelihoods
+over each bin. In other words, we optimized a composite likelihood where
+observations in different bins were taken to be independent. To account for
+correlations across bins in uncertainty estimates, we estimated parameter
+confidence intervals using the same bootstrap set using the Godambe information
+matrix \citep{Coffman2016-yq}. \sgcomment{is redundancy here with section
+"Optimization using moments". I think we can get rid of this and say:
+"optimization and uncertainty calculations are described in section
+"Optimization using moments""?, This may be a bit more complicated, since the
+Optimizatino section relies on this description. I think we should just punt
+the bootstrap description to that section. TODO: discuss Aaron + Simon}
+
 \subsection{Estimating two-locus statistics with small sample sizes}
 
+The approach from \citet{Ragsdale2020-nz} provides unbiased estimates of the LD
+statistics considered here, with smaller sample sizes causing greater
+uncertainty in the estimated statistics, but is accounted for by computing
+variances/covariances via bootstrap. \sgcomment{I don't think that this is true
+if our bootstrap is over genomic regions. In an infinite genome with a tiny
+sample size, we would estimate no uncertainty...  Or at least, it is only true
+if the Neandertal form a truly randomly mating population...  }
+
+Some statistics, such as $D^2$, require at least two diploid samples to
+compute. Since we used a single Neanderthal sample, these statistics for the
+Neandertal population were not used in the fit. By contrast, there are
+statistics that only require a single sample per population to estimate. These
+include cross-population heterozygosity, as well as some statistics involving
+more than one population. For example, statistics of the form
+$D_{human}(1-2p_{human})(1-2q_{neanderthal})$ require a single Neanderthal
+sample and are informative of the Neanderthal demography. These statistics were
+included in the fit, but statistics requiring more than one Neanderthal sample
+to estimate were removed.
+
 \subsection{Computing conditional SFS}
+
+The conditional site frequency spectrum (or cSFS) is the distribution of allele
+frequencies restricted to loci that satisfy a given condition. Specifically, we
+consider the distribution of allele frequencies in present-day populations
+conditioned on the Vindija Neanderthal carrying the derived allele relative to
+the inferred ancestral allele. Ancestral alleles alleles were determined from a
+6 primate alignment \aprcomment{cite}. This cSFS is expected to be close to
+uniform under neutrality and a simple split model (with no subsequent
+migration) between the ancestors of modern humans and Neanderthal
+\citep{Chen2007-iy}. By contrast, a U-shaped distribution has been taken as
+evidence for archaic introgression from a population whose split from modern
+humans is at least as old as that of Neanderthal’s
+\citet{Durvasula2020-td,Yang2012-ze}. Because we wanted to compare our
+inferences (based on intergenic sites) to inferences from previous work (based
+on whole-genome data), we computed the cSFS for both intergenic and all sites
+genome-wide. Sites with no calls in the Vindija Neanderthal were excluded from
+this analysis. \aprcomment{Figures XX.}
+
+Because we were concerned that cSFS analyses may be affected by incorrect
+inference of the ancestral allele \cite{Hernandez2007-mf}, we computed the cSFS
+for all mutations, and for transitions and transversions separately
+\aprcomment{(Figure X, cSFS)}. Comparisons of these observed cSFS with model
+predictions are discussed in the model prediction section below.
 
@@
 
-\subsection{Confidence intervals using godambe methods}
+\subsection{Confidence intervals using Godambe methods}
 
@@
 \addcontentsline{toc}{section}{References}
-\bibliographystyle{plainnat}
+\bibliographystyle{genetics}
 \bibliography{paper}
 
-\break
+\clearpage
 
@@
 
-\break
+\clearpage
 

### 2022-01-01 paper.tex
supp tables started
placeholder figure
fill in a few things
paper updates

@@
 % local definitions
-\newcommand{\comment}[1]{{\textcolor{red}{Comment: #1}}}
+\newcommand{\sgcomment}[1]{{\textcolor{red}{SG: #1}}}
+\newcommand{\bmhcomment}[1]{{\textcolor{blue}{BMH: #1}}}
+\newcommand{\aprcomment}[1]{{\textcolor{purple}{APR: #1}}}
+\newcommand{\tdwcomment}[1]{{\textcolor{orange}{TDW: #1}}}
 
@@
 
-\begin{figure}[ht]
-\begin{center}
-\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
-\caption{\textbf{The first figure}.
-    It has a caption.
-}
-\label{fig:1}
-\end{center}
-\end{figure}
+\subsection*{A Late Middle Stone Age common ancestry for contemporary humans}
 
-\subsection*{A Late Middle Stone Age common ancestry for contemporary human populations}
-
-\subsection*{Evidence for deep population structure, but not archaic admixture within Africa}
+\subsection*{Deep population structure but not archaic admixture within Africa}
 
@@
 
+\begin{figure}[ht]
+\begin{center}
+\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
+\caption{\textbf{The first figure}.
+    The geographic and genetic diversity of populations across Africa.
+}
+\label{fig:1}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\makebox[\textwidth][c]{} % \includegraphics{figures/fig2.png}}
+\caption{\textbf{The second main figure}.
+    A placeholder - best fit model(s).
+}
+\label{fig:2}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
+\caption{\textbf{The third main figure}.
+    A placeholder - validation (Relate and cSFS).
+}
+\label{fig:r3}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
+\caption{\textbf{The fourth main figure}.
+    A placeholder - predictions (FST and/or f4).
+}
+\label{fig:r4}
+\end{center}
+\end{figure}
+
 \section*{Acknowledgements}

### 2022-01-01 supplement.tex
paper updates

@@
 
-DNA samples were collected in South Africa (Richtersveld) with IRB approval
-from Stanford University, Stellenbosch University (ZA) and SUNY Stony Brook.
-\aprcomment{Does this need to be expanded on at all?}
+DNA samples were collected from three Nama communities in the Richtersveld
+region of South Africa, which borders southern Namibia, in 2012. A community
+guide was present during each interview and facilitated consent in Afrikaans or
+Nama. Written consent was recorded per our IRB protocol with human subjects
+approval from Stanford University (Protocol \#13829), Stellenbosch University
+(N11/07/210) and later maintained via SUNY Stony Brook (Protocol \#727494).
+Saliva samples were extracted from Oragene [OGR-500] kits at Stellenbosch
+University. Results stemming from genetic analyses have been communicated in
+2015, 2019, 2021 via community presentations, a radio interview and to
+representatives of the Richtersveld National Park.
 
@@
 
-\subsection{Computing LD and diversity statistics}
+\subsection{Computing LD and diversity statistics}\label{sec:computing-stats}
 
@@
 
-\subsection{Computing conditional SFS}
+\subsection{Computing conditional SFS}\label{sec:computing-csfs}
 
@@
 evidence for archaic introgression from a population whose split from modern
-humans is at least as old as that of Neanderthal’s
+humans is at least as old as that of the human-Neanderthal split
 \citet{Durvasula2020-td,Yang2012-ze}. Because we wanted to compare our
@@
 
+For the early history, we tested model parameterizations that cover many of the
+proposed scenarios of population structure, size changes, and/or archaic
+admixture. The simplest model, in terms of number of parameters, was a
+single-origin expansion of modern humans, with no structure in the stem and no
+archaic admixture aside from the Neanderthal admixture in Eurasian populations
+following the out-of-Africa migration. This model allowed for a population size
+change in the stem of modern humans between the ancestral split of the
+human-Neanderthal lineages and the more recent split of branches leading to
+Southern and West/Eastern African populations.
+
+To include population structure in early human history, we considered multiple
+parameterizations of models that allowed more than a single stem population. In
+general, stem populations were allowed to vary in their sizes, split times, and
+migration rates, with parameterizations flexible enough to encompass proposed
+scenarios of either archaic admixture or population structure, both connected
+by gene flow or with isolation between stems.
+
+In one parameterization of early structure, which we refer to as a “continuous
+migration” model, a secondary stem population (stem 2) split from the primary
+stem (stem 1) that leads to modern humans. Stem 1 contributed to present-day
+populations via a series of population splits similar to the single-origin
+model, while stem 2 contributed through continuous symmetric migrations with
+contemporaneous populations. The symmetrical migration rates could differ
+across population pairs and over different epochs. This continuous migration
+was allowed until stem 2 disappeared, which occurred as recently as 5kya. We
+tested models that both allowed or disallowed migration between stems, i.e.,
+before stem 1 split into S/E/W African populations.
+
+In another parameterization of early structure, a secondary stem population
+(stem 2) contributed ancestry to present-day populations via a series of
+instantaneous admixture events (i.e., pulse admixture or “merger” events) to
+lineages leading to sampled present-day African populations. Merger events were
+allowed to occur in one or more of the Nama, Mende, and Gumuz branches, as well
+as the branch of East and West Africans prior to their split. Those admixture
+events were allowed to occur at any time along those branches, and with any
+proportions, and stem 2 was allowed to split from the primary stem at any time
+before subsequent divergences (and either before or after the split of the
+Neanderthal branch). We tested models that allowed migration between the early
+stems, before subsequent splits and admixture events, as well as models that
+were restricted to isolation between stem branches. Depending on the specific
+parameters, such models encompass commonly-considered ghost archaic admixture
+scenarios (e.g., if a long-isolated lineage more recently contributes a
+minority of ancestry to one or more populations), as well as relatively simple
+fragmentation-coalescence scenarios.
+
+Based on the geographical locations of present-day populations, we tentatively
+labeled ancestral branches using a parsimony in migration, referring to South,
+East, and West African branches (S/E/W AFR, Figure X). We do not know the
+geographical location of these ancestral populations (nor even if they
+correspond to populations with a well-defined geographic range), and these
+labels should be considered as tentative. However, we found it useful to name
+branches in reference to where in Africa their descendants are found.
+\sgcomment{Please double-check this paragraph! TODO Brenna?}
+
 \subsection{General strategy for building models and introducing complexity}
@@
 
+moments-LD uses a composite likelihood approach to simultaneously fit relative
+pairwise diversity and LD statistics over a range of recombination distances.
+Likelihoods were computed independently for pairwise diversity and each
+recombination distance bin using a multivariate Guassian likelihood function as
+described in Section \ref{sec:computing-stats}. These were multiplied across
+bins and the single set of heterozygosity statistics following the approach
+detailed in \citet{Ragsdale2019-nt}. For each model tested, we ran multiple
+rounds of optimization, alternating between the \emph{optimize\_log\_fim},
+\emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
+parameter space and hone in on the best fit parameters. Initial guesses for
+parameters were chosen from demographically plausible starting points and then
+perturbed to explore space, using gradient descent (on the log of the
+parameters). The best fits from these initial rounds of optimization were then
+chosen as the starting points for optimization using the Powell and/or the
+L-BFGS-B methods (as implemented in the SciPy optimization package
+\cite{Virtanen2020-kr}). This process was repeated with alternating
+optimization methods until the best fit parameter set converged consistently.
+
 \subsection{Confidence intervals using Godambe methods}
+
+The bootstrap replicates that were used to compute the variance-covariance
+structure of the observed statistics within bins were also used to build 500
+bootstrap replicates of the data by resampling with replacement. For the best
+fit parameters, we computed confidence intervals using the Godambe Information
+approach, which corrects composite likelihood estimates of confidence intervals
+to account for nonindependence in the data, including linkages between loci and
+nonindependence of recombination bins \cite{Coffman2016-yq}.
 
@@
 
+\begin{figure}[ht]
+\begin{center}
+    \includegraphics[width=0.65\textwidth]{figures/supp-possible-models.png}
+    \caption{\textbf{supp}.}
+    \label{fig:supp-possible-models}
+\end{center}
+\end{figure}
+
 \end{document}

### 2022-01-02 doc/conf.py
set up docs

@@ -1,0 +1,62 @@
+# Configuration file for the Sphinx documentation builder.
+#
+# This file only contains a selection of the most common options. For a full
+# list see the documentation:
+# https://www.sphinx-doc.org/en/master/usage/configuration.html
+
+# -- Path setup --------------------------------------------------------------
+
+# If extensions (or modules to document with autodoc) are in another directory,
+# add these directories to sys.path here. If the directory is relative to the
+# documentation root, use os.path.abspath to make it absolute, like shown here.
+
+import os
+import sys
+sys.path.insert(0, os.path.abspath('.'))
+
+
+# -- Project information -----------------------------------------------------
+
+project = 'Afr paper'
+copyright = '2021, Aaron Ragsdale'
+author = 'Aaron Ragsdale'
+
+# The full version, including alpha/beta/rc tags
+release = '0.0.1'
+
+
+# -- General configuration ---------------------------------------------------
+
+# Add any Sphinx extension module names here, as strings. They can be
+# extensions coming with Sphinx (named 'sphinx.ext.*') or your custom
+# ones.
+extensions = [
+    'sphinx.ext.autodoc',
+    'sphinx.ext.todo',
+    'sphinx.ext.viewcode',
+    'jupyter_sphinx',
+]
+
+# Add any paths that contain templates here, relative to this directory.
+templates_path = ['_templates']
+
+# List of patterns, relative to source directory, that match files and
+# directories to ignore when looking for source files.
+# This pattern also affects html_static_path and html_extra_path.
+exclude_patterns = ['_build', 'Thumbs.db', '.DS_Store']
+
+# If true, `todo` and `todoList` produce output, else they produce nothing.
+todo_include_todos = True
+
+# -- Options for HTML output -------------------------------------------------
+
+# The theme to use for HTML and HTML Help pages.  See the documentation for
+# a list of builtin themes.
+#
+html_theme = 'classic'
+
+# Add any paths that contain custom static files (such as style sheets) here,
+# relative to this directory. They are copied after the builtin static files,
+# so a file named "default.css" will overwrite the builtin "default.css".
+html_static_path = ['_static']
+numfig = True

### 2022-01-03 paper.tex
trigger action
Building models and complexity

@@
 \usepackage{color}
-
-% lorem ipsum dummy text
-\usepackage{lipsum}
 
@@
 \author[7,**]{Simon Gravel}
-\affil[1]{FIXME: UW Madison}
-\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA 95616, USA}
+\affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
+\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
 \affil[3]{FIXME: DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
@@
 \affil[5]{FIXME}
-\affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA 95616, USA}
-\affil[7]{FIXME: McGill}
+\affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}
+\affil[7]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
 \affil[*]{bmhenn@ucdavis.edu}
@@
 \section*{Introduction}
+
+The intro..
 
@@
 \section*{Methods}
-\cite{Kelleher2016-lw}
+A placeholder citation \citep{Kelleher2016-lw}.
 

### 2022-01-03 supplement.tex
Building models and complexity

@@
 \usepackage{booktabs}
-
-% lorem ipsum dummy text
-\usepackage{lipsum}
 
@@
 
+With up to six sampled populations in final demographic models that we fit,
+there are many parameters to learn. Even in the simplest model involving all
+populations (such as the tree-like single-origin model), there are a few dozen
+parameters defining split times, migration rates, admixture timings and
+proportions, and population sizes and size changes. Thus, parameter space for a
+given model topology is large. In addition, the space of possible model
+topologies is itself large – as the number of populations increases, the number
+of possible topologies also increases, as there are more possibilities for the
+order of divergence and admixture events.
+
+In order to narrow the set of possible models to plausible scenarios and to
+avoid overfitting, we took an approach that combined the incremental addition
+of complexity, starting with fewer populations before combining all
+populations, as well as fixing parameters that have been previously estimated
+or that fit consistently across all model scenarios. By initially considering
+sets of two or three populations, we were able to narrow down the relative
+orders of divergences between African and Eurasian populations. Assuming simple
+isolation-with-migration models, the Nama appeared to be the earliest diverging
+population of those we considered, with West African (Mende) and East African
+(Gumuz) populations diverging more recently, followed by the split of the
+Eurasian branch from the East African branch.
+
+We performed an initial round of optimization including all six populations
+with a family of models including single- and multiple-stem scenarios as
+described in the previous section. We identified parameters that reached
+consistent values across all models. These included the timings of recent or
+non-central divergences and admixture events:
+\begin{itemize}
+    \item East/West African population split: 60ka
+    \item East African/European split: 50 ka
+    \item Neanderthal introgression to Eurasian branch: 45 ka
+    \item Neanderthal/human split: 550 ka
+    \item Eurasian back-to-African migration: 12 ka
+\end{itemize}
+When testing multiple variations of the more complex models, we kept these
+values fixed. This reduced the potential for overfitting the more complex
+models, while reducing the computational cost of optimization.
+
+Our models also included recent events to account for known migrations,
+admixtures, and growths and declines in effective population sizes. Many of
+these parameters were fixed based on previous historical, genetic, or
+anthropological research, namely
+\begin{itemize}
+    \item East African pastoralist to South African admixture: 2 ka \cite{}
+    \item European to South African admixture: 10 generations ago \cite{}
+    \item Mende population expansion and Gumuz population decline: 5 ka \cite{}
+    \item South African population decline following colonial admixture: 9 generations ago \cite{}
+\end{itemize}
+While the dates of these events were fixed, the sizes and proportions were
+allowed to vary in the fits. The total number of parameters that were
+ultimately inferred were 18-26 \aprcomment{check!}, depending on the complexity
+of the model.
+
 \subsection{Optimization using moments}

### 2022-01-03 supplement.tex
diversity stats and relate inference

@@
 \usepackage{booktabs}
+\usepackage{amsfonts}
 
@@
 \newcommand{\tdwcomment}[1]{{\textcolor{cyan}{TDW: #1}}}
+
+\newcommand{\E}{\mathbb{E}}
 
@@
 All analyses presented in this work focus on biallelic single nucleotide
-polymorphisms within the 1000 Genomes Phase 3 strict mask.
-
-For the moments-LD analysis, we focused on intergenic locations because these
-appear less affected by natural selection compared to both synonymous and
-nonsynonymous variation \citep{Ragsdale2018-dd}. To enable comparison with
-Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
-had less than 100 contiguous base pairs.
+polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the moments-LD
+analysis, we focused on intergenic locations because these appear less affected
+by natural selection compared to both synonymous and nonsynonymous variation
+\citep{Ragsdale2018-dd}. To enable comparison with Neanderthal DNA, we excluded
+regions for which the Vindija Neanderthal sample had less than 100 contiguous
+base pairs.
 
@@
 
-\aprcomment{$D^2$, $D_z$, $\pi_2$, and $H$. Definitions, notation,
-indexing\ldots}
+We used multi-population linkage disequilibrium (LD) and pairwise diverisity
+statistics to fit demographic models to data. These statistics, introduced and
+described in detail in \citet{Ragsdale2019-nt}, are the multi-population analog
+of the classic LD statistics first described and computed by
+\citet{Hill1968-vu,Ohta1971-yd}.
+
+Given two biallelic loci, with alleles $A$ and $a$ and the left locus and
+allels $B$ and $b$ at the right locus, the standard covariance measure of LD is
+$D=p_{AB}p_{ab}-p_{Ab}p_{aB}$, where $p_{AB}$ is the frequency of $AB$
+haplotypes in a population (and thus the probability of drawing an $AB$
+haplotype in a random sample of that population). \citet{Hill1968-vu} solve for
+the expectation of $D^2$ using a system of equations that includes
+$\E[Dz] = \E[D(1-2p_A)(1-2p_B)]$ and $\E[\pi_2] =
+\E[p_A(1-p_A)p_B(1-p_B)]$, where $p_A$ and $p_B$ are the frequencies of
+$A$ and $B$ at the left and right loci, respectively. This system also requires
+the expected pairwise diversity (or expected heterozygosity, assuming random
+mating), denoted $\E[H] = \E[2p_A(1-p_A)] =
+\E[2p_B(1-pB)]$, assuming equal mutation rates at the two loci.
+
+\citet{Ragsdale2019-nt} showed how to compute the analogous multi-population
+set of LD statistics, and we refer readers there for details on their
+definitions, computation, and interpretations. In short, we obtain expectations
+of $D^2$ in each population, the cross-population product $D_iD_j$ (where $i$
+and $j$ index populations), as well as those additional statistics $Dz$ and
+$\pi_2$ taken over different combinations of population indexing. That is
+$Dz_{i, j, k} = D_i(1-2p_{A,j})(1-2p_{B,k})$ and
+$\pi_{2;i,j,k,l}=p_{A,i}(1-p_{A,j})p_{B,k}(1-2p_{B,l})$. We consider statistics
+normalized by $\pi_2$ in a reference population (throughout, we use the Mende
+$\pi_2$), which removes any dependence on the mutation rate. Thus, statistics
+take the form $\sigma_{d;i,j}^2 = \frac{\E[D^2_{i,j}]}{\E[\pi_2]}$,
+$\sigma_{Dz;i,j,k}=\frac{\E[Dz_{i,j,k}]}{\E[\pi_2]}$, and so on.
+Multi-population pairwise diversity statistics $\E[H_i]=\E[2p_i(1-p_i)]$ and
+$\E[H_{i,j}] = \E[p_i(1-p_j) + p_j(1-p_i)]$ were also normalized by $H$ in the
+Mende, so that all pairwise diversity measures are relative to the reference
+population.
+
+$\sigma_{Dz}$ has been shown to be sensitive to deep population structure and
+archaic admixture \citep{Ragsdale2019-nt}, and this statistic is closely
+related to $S^*$ statistics used to scan for introgressed haplotypes
+\citep{Plagnol2006-lt}. Pairwise diversity statistics $H_{i,j}$ have also been
+widely used in demographic inference involving ancient DNA and many samples, as
+$f_2$, $f_3$ and $f_4$ statistics can be expressed as linear combinations of
+$H_{i,j}$. $f$-statistics form the basis of admixture graph analysis
+\aprcomment{cite}. Therefore, the set of statistics used here encompass
+multiple features of genetic data that have been used to infer models of
+archaic admixture and population structure involving many populations.
 
@@
 
+We used Relate version 1.0.16 \citep{Speidel2019-nj} to reconstruct genome-wide
+gene genealogies using a combined set of 1000 Genomes and African Diversity
+Reference Panel datasets, retaining all AFR-labeled populations and GBR, CEU,
+PJL, and CHB from the 1000 Genomes panel and the Nama, Gumuz, Oromo, and Amhara
+from ADRP. We used all autosomes and applied the 1000 Genomes Phase 3 strict
+mask, we used an ancestral sequence determined by a 6-primate alignment
+(human\_ancestor\_GRCh37\_e59), and we used the HapMap II combined
+recombination map, all in GRCh37 coordinates. We assumed a generation time of
+29 years, a mutation rate of 1.25e-8, and followed the standard pipeline
+described in the Relate documentation.
+
+From the reconstructed gene genealogies, we computed coalescence rates within
+and between populations using Relate’s function RelateCoalescentRate. This
+allows for an estimate of the instantaneous inverse coalescence rate (IIRC) for
+samples drawn within each population (the inverse of which is often interpreted
+as the effective population size history), and the relative cross coalescence
+rates between pairs of populations (which are commonly used to estimate
+divergence times).
+
+Following \citet{Speidel2019-nj} we also identified “deep branches” within gene
+trees. Such a “deep branch” is a branch within a marginal tree that has its
+upper end (or node) extending to more than 1 million years in age, and we
+partitioned deep branches based on their lower end ages into bins between 0 and
+1Ma. Such branches can be categorized by their association with Neanderthals by
+comparing mutations that fall upon such a branch to the allele found in a
+Neanderthal genome sequence. For this, we used the published high-coverage
+Vindija Neanderthal \citep{Prufer2017-kk}. Again following the analysis in
+\citet{Speidel2019-nj}, if one or more mutations on a deep branch are shared
+with the Neanderthal sequence, the branch is inferred to have passed through
+the Neanderthal lineage.
+
+A deep branch is assigned to a contemporary population if at least one sample
+from that population has ancestry that passes through that branch.
+\citet{Speidel2019-nj} show that deep branches with lower ends more recent than
+the Neanderthal introgression event are enriched for Neanderthal-matching
+alleles in Eurasian populations, while a large majority of deep branches in
+1000 Genomes West African populations do not match either the Neanderthal or
+Denisovan samples. This observation was taken as further evidence for deep
+population structure or archaic admixture in West African populations from an
+unidentified hominin unrelated to the Neanderthal/Denisovan complex.
+
 \section{Predictions from inferred demographic models}
@@
 
-\section{Validations using simulatinos from inferred demographic models}
+\section{Validations using simulations from inferred demographic models}
 
@@
 \subsection{cSFS prediction under inferred models}
+
+\section{Supplementary results}
+
+\subsection{Conditional SFS}
 

### 2022-01-04 supplement.tex
add fit plots

@@
 haplotype in a random sample of that population). \citet{Hill1968-vu} solve for
-the expectation of $D^2$ using a system of equations that includes
-$\E[Dz] = \E[D(1-2p_A)(1-2p_B)]$ and $\E[\pi_2] =
-\E[p_A(1-p_A)p_B(1-p_B)]$, where $p_A$ and $p_B$ are the frequencies of
-$A$ and $B$ at the left and right loci, respectively. This system also requires
-the expected pairwise diversity (or expected heterozygosity, assuming random
-mating), denoted $\E[H] = \E[2p_A(1-p_A)] =
-\E[2p_B(1-pB)]$, assuming equal mutation rates at the two loci.
+the expectation of $D^2$ using a system of equations that includes $\E[Dz] =
+\E[D(1-2p_A)(1-2p_B)]$ and $\E[\pi_2] = \E[p_A(1-p_A)p_B(1-p_B)]$, where $p_A$
+and $p_B$ are the frequencies of $A$ and $B$ at the left and right loci,
+respectively. This system also requires the expected pairwise diversity (or
+expected heterozygosity, assuming random mating), denoted $\E[H] =
+\E[2p_A(1-p_A)] = \E[2p_B(1-pB)]$, assuming equal mutation rates at the two
+loci.
 
@@
 Statistics were largely unchanged by using a different map
-(Figure~\ref{fig:sup-map_comparison}). \aprcomment{These maps are both inferred
-using array data, which is sparse. Comment on this.}
+(Figure~\ref{fig:supp-omni-hapmap-comparison}). \aprcomment{These maps are both
+inferred using array data, which is sparse. Comment on this.}
 
@@
 
+Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}
+give the best fit values for all parameters in the four models considered here,
+along with standard errors using computed using the Godambe method.
+Figures~\ref{fig:supp-single-origin-fits}--\ref{fig:supp-merger-with-stem-migration-fits}
+show the corresponding fits to many of the LD statistics used in the inference.
+The $\sigma_{Dz}$ statistics are fit poorly by the single-origin model, while
+the models with multiple stems all provide reasonable fits to the data. The
+merger-with-stem-migration model performed the best of these four models,
+although the fits were nearly visually indistiguishable from the fits from the
+merger-with-stem-migration model.
+
 \section{Gene genealogy reconstruction}
@@
 Following \citet{Speidel2019-nj} we also identified “deep branches” within gene
-trees. Such a “deep branch” is a branch within a marginal tree that has its
-upper end (or node) extending to more than 1 million years in age, and we
-partitioned deep branches based on their lower end ages into bins between 0 and
-1Ma. Such branches can be categorized by their association with Neanderthals by
+trees. Such a “deep branch” is a branch within a marginal tree (meaning, the
+tree at a given locus) that has its upper end (i.e., the node of its
+coalescence with another branch) extending to more than 1 million years in age,
+and we partitioned deep branches based on their lower node ages (where two or
+more branches coalesce into this single branch) into bins between 0 and 1Ma.
+Such branches can be categorized by their association with Neanderthals by
 comparing mutations that fall upon such a branch to the allele found in a
@@
 \caption{
-    \label{tab:single_origin}
+    \label{tab:supp-single-origin}
     \textbf{Best-fit parameters from the Single-Origin model.}
@@
 \caption{
-    \label{tab:continuous_migration}
+    \label{tab:supp-continuous-migration}
     \textbf{Best-fit parameters from the Continuous-Migration model.}
@@
 \caption{
-    \label{tab:merger_without_stem_migration}
+    \label{tab:supp-merger-without-stem-migration}
     \textbf{Best-fit parameters from the Merger-Without-Stem-Migration model.}
@@
 \caption{
-    \label{tab:merger_with_stem_migration}
+    \label{tab:supp-merger-with-stem-migration}
     \textbf{Best-fit parameters from the Merger-With-Stem-Migration model.}
@@
 \begin{figure}[ht]
-\begin{center}
+    \centering
     \includegraphics[width=0.65\textwidth]{figures/supp-possible-models.png}
-    \caption{\textbf{supp}.}
+    \caption{\textbf{Proposed models of early human history in Africa.}}
     \label{fig:supp-possible-models}
-\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-omni-hapmap-comparison.pdf}
+    \caption{
+        \textbf{Comparison of LD statistics parsed using two recombination maps}
+        The recombination map determines genetic distances between pairs of SNPs,
+        so using different maps could result in systematic biases in parsed
+        statistics if the maps differ significantly. Here, the solid lines are
+        statistics parsed using the HapMapII recombination map, and dashed lines
+        are using the OMNI-YRI map. Shading represents 95\% confidence intervals
+        from the OMNI map. These two maps results in LD statistics that are
+        indistinguishable from one another.
+    }
+    \label{fig:supp-omni-hapmap-comparison}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-single-origin-fits.pdf}
+    \caption{\textbf{Single-origin model fit to LD statistics.}}
+    \label{fig:supp-single-origin-fits}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-continuous-migration-fits.pdf}
+    \caption{\textbf{Continuous-migration model fit to LD statistics.}}
+    \label{fig:supp-continuous-migration-fits}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-merger-without-stem-migration-fits.pdf}
+    \caption{\textbf{Merger-without-stem-migration model fit to LD statistics.}}
+    \label{fig:supp-merger-without-stem-migration-fits}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-merger-with-stem-migration-fits.pdf}
+    \caption{\textbf{Merger-with-stem-migration model fit to LD statistics.}}
+    \label{fig:supp-merger-with-stem-migration-fits}
 \end{figure}

### 2022-01-04 supplement.tex
add f4 and csfs figures

@@
 \usepackage{amsfonts}
+\usepackage{bm}
 
@@
 \subsection{LD and diversity statistics used in model fits}
+\label{sec:supp-ld-stats}
 
@@
         from the OMNI map. These two maps results in LD statistics that are
-        indistinguishable from one another.
+        indistinguishable from one another. Notations for each statistic are
+        described in section~\ref{sec:supp-ld-stats} and indexed by populations
+        (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
     }
@@
     \includegraphics{figures/supp-single-origin-fits.pdf}
-    \caption{\textbf{Single-origin model fit to LD statistics.}}
+    \caption{
+        \textbf{Single-origin model fit to LD statistics.}
+        Predicted vs observed two-locus statistics as a function of genetic
+        distance for the single-origin model. Each panel represents a different
+        set of two-locus statistics. Solid lines represent estimates using the
+        single-origin model. Dashed lines represent observed statistics.
+        Notation and indexing of statistics are described in the
+        Fig.~\ref{fig:supp-omni-hapmap-comparison} caption.
+    }
     \label{fig:supp-single-origin-fits}
@@
     \includegraphics{figures/supp-continuous-migration-fits.pdf}
-    \caption{\textbf{Continuous-migration model fit to LD statistics.}}
+    \caption{
+        \textbf{Continuous-migration model fit to LD statistics.}
+        Predicted vs observed two-locus statistics as a function of genetic
+        distance for the continuous-migration model. Each panel represents a different
+        set of two-locus statistics. Solid lines represent estimates using the
+        continuous-migration model. Dashed lines represent observed statistics.
+        Notation and indexing of statistics are described in the
+        Fig.~\ref{fig:supp-omni-hapmap-comparison} caption.
+    }
     \label{fig:supp-continuous-migration-fits}
@@
     \includegraphics{figures/supp-merger-without-stem-migration-fits.pdf}
-    \caption{\textbf{Merger-without-stem-migration model fit to LD statistics.}}
+    \caption{
+        \textbf{Merger-without-stem-migration model fit to LD statistics.}
+        Predicted vs observed two-locus statistics as a function of genetic
+        distance for the merger-without-stem-migration model. Each panel
+        represents a different set of two-locus statistics.
+        Solid lines represent estimates using the
+        merger-without-stem-migration model.
+        Dashed lines represent observed statistics.
+        Notation and indexing of statistics are described in the
+        Fig.~\ref{fig:supp-omni-hapmap-comparison} caption.
+    }
     \label{fig:supp-merger-without-stem-migration-fits}
@@
     \includegraphics{figures/supp-merger-with-stem-migration-fits.pdf}
-    \caption{\textbf{Merger-with-stem-migration model fit to LD statistics.}}
+    \caption{
+        \textbf{Merger-with-stem-migration model fit to LD statistics.}
+        Predicted vs observed two-locus statistics as a function of genetic
+        distance for the merger-with-stem-migration model. Each panel
+        represents a different set of two-locus statistics.
+        Solid lines represent estimates using the
+        merger-with-stem-migration model.
+        Dashed lines represent observed statistics.
+        Notation and indexing of statistics are described in the
+        Fig.~\ref{fig:supp-omni-hapmap-comparison} caption.
+    }
     \label{fig:supp-merger-with-stem-migration-fits}
@@
 
+\begin{figure}[ht]
+    \centering
+    %\includegraphics{figures/supp-FST-panels}
+    \caption{
+        \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
+    }
+    \label{fig:supp-FST-panels}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-f4s-single-origin.pdf}
+    \caption{
+        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
+        and ancient branches from the single-origin model.}
+    }
+    \label{fig:supp-f4s-single-origin}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-f4s-continuous-migration.pdf}
+    \caption{
+        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
+        and ancient branches from the continuous-migration model.}
+    }
+    \label{fig:supp-f4s-continuous-migration}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-f4s-merger-without-stem-migration.pdf}
+    \caption{
+        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
+        and ancient branches from the merger-without-stem-migration model.}
+    }
+    \label{fig:supp-f4s-merger-without-stem-migration}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-f4s-merger-with-stem-migration.pdf}
+    \caption{
+        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
+        and ancient branches from the merger-with-stem-migration model.}
+    }
+    \label{fig:supp-f4s-merger-with-stem-migration}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-csfs-single-origin.pdf}
+    \caption{caption}
+    \label{fig:supp-csfs-single-origin}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-csfs-continuous-migration.pdf}
+    \caption{caption}
+    \label{fig:supp-csfs-continuous-migration}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-csfs-merger-without-stem-migration.pdf}
+    \caption{caption}
+    \label{fig:supp-csfs-merger-without-stem-migration}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-csfs-merger-with-stem-migration.pdf}
+    \caption{caption}
+    \label{fig:supp-csfs-merger-with-stem-migration}
+\end{figure}
+
 \end{document}

### 2022-01-05 supplement.tex
add FST panel plot

@@
 The approach from \citet{Ragsdale2020-nz} provides unbiased estimates of the LD
-statistics considered here, with smaller sample sizes causing greater
-uncertainty in the estimated statistics, but is accounted for by computing
-variances/covariances via bootstrap. \sgcomment{I don't think that this is true
-if our bootstrap is over genomic regions. In an infinite genome with a tiny
-sample size, we would estimate no uncertainty...  Or at least, it is only true
-if the Neandertal form a truly randomly mating population...  }
+statistics considered here, which can be directly compared to expectations for
+those same statistics computed from the multi-population Hill-Robertson system.
+Estimated statistics for populations with smaller sample sizes have greater
+uncertainty, which is largely accounted for by computing variances and
+covariances via bootstrap, assuming our samples are unrelated and the
+population is randomly mating.
 
-Some statistics, such as $D^2$, require at least two diploid samples to
-compute. Since we used a single Neanderthal sample, these statistics for the
-Neandertal population were not used in the fit. By contrast, there are
-statistics that only require a single sample per population to estimate. These
-include cross-population heterozygosity, as well as some statistics involving
-more than one population. For example, statistics of the form
-$D_{human}(1-2p_{human})(1-2q_{neanderthal})$ require a single Neanderthal
-sample and are informative of the Neanderthal demography. These statistics were
-included in the fit, but statistics requiring more than one Neanderthal sample
-to estimate were removed.
+\sgcomment{I don't think that this is true if our bootstrap is
+over genomic regions. In an infinite genome with a tiny sample size, we would
+estimate no uncertainty...  Or at least, it is only true if the Neandertal form
+a truly randomly mating population...  }
+
+Some statistics, such as $\E[D_i D_j]$, require at least two diploid samples
+from a population to compute. Since we used a single Neanderthal sample, such
+statistics for the Neandertal population were not used in the fit. By contrast,
+there are statistics that only require a single sample in a given population to
+estimate. These include cross-population nucleotide diversity measures, as well
+as some LD statistics involving more than one population. For example,
+statistics of the form $D_{human}(1-2p_{human})(1-2q_{neanderthal})$ require a
+single Neanderthal sample and are informative of the Neanderthal demography.
+These statistics were included in the fit, but statistics requiring more than
+one Neanderthal sample to estimate were removed.
 
@@
 \subsection{$F_{ST}$ between coexisting populations over time}
+
+$F_{ST}$ can be interpreted as a measure of population structure, given by the
+proportion of genetic variance in a multi-population sample that can be
+attributed to drift between populations in that sample.
 
@@
     \centering
-    %\includegraphics{figures/supp-FST-panels}
+    \includegraphics[width=\textwidth]{figures/supp-fsts-all-models.pdf}
     \caption{
         \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
+        \aprcomment{Plot ordering, line colors and styles, etc need to be cleaned up}
     }
@@
     \includegraphics{figures/supp-csfs-single-origin.pdf}
-    \caption{caption}
+    \caption{
+        \textbf{Conditional SFS compared to single-origin model.}
+    }
     \label{fig:supp-csfs-single-origin}
@@
     \includegraphics{figures/supp-csfs-continuous-migration.pdf}
-    \caption{caption}
+    \caption{
+        \textbf{Conditional SFS compared to continuous-migration model.}
+    }
     \label{fig:supp-csfs-continuous-migration}
@@
     \includegraphics{figures/supp-csfs-merger-without-stem-migration.pdf}
-    \caption{caption}
+    \caption{
+        \textbf{Conditional SFS compared to merger-without-stem-migration model.}
+    }
     \label{fig:supp-csfs-merger-without-stem-migration}
@@
     \includegraphics{figures/supp-csfs-merger-with-stem-migration.pdf}
-    \caption{caption}
+    \caption{
+        \textbf{Conditional SFS compared to merger-with-stem-migration model.}
+    }
     \label{fig:supp-csfs-merger-with-stem-migration}

### 2022-01-05 supplement.tex
add f4 section

@@
 \usepackage{amsfonts}
+\usepackage{amsmath}
 \usepackage{bm}
@@
 proportion of genetic variance in a multi-population sample that can be
-attributed to drift between populations in that sample.
+attributed to drift between populations in that sample \cite{Bhatia2013-sq}.
+$F_{ST}$ depends on population divergence times as well as subsequent gene flow
+and population sizes: migration between populations reduces $F_{ST}$ while
+small effective population sizes can lead to large values. For each of the four
+inferred demographic models presented here, we computed predicted $F_{ST}$ for
+pairs of contemporaneous populations over time. For sampled populations,
+predicted $F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}). We
+computed $F_{ST}$ as a ratio of expected (or average) variances, using observed
+pairwise diversity within and between populations as
+\begin{equation}
+    F_{ST}(i, j) = \frac{\E[2H_{i,j} - H_i - H_j]}{\E[2H_{i,j} + H_i + H_j]}.
+    \label{eq:FST}
+\end{equation}
+
+In all inferred models, $F_{ST}$s between Neanderthals and the ancestors of
+modern humans were much larger than between any pair of modern human
+populations or their ancestors (stems). $F_{ST}$ between modern human
+populations over the last 100 thousand years remained relatively low, at
+$\sim0.1$ or less. In the continuous-migration model, the two stem populations
+also were predicted to have low $F_{ST}$ within this range, as migration rates
+were inferred to be high enough to maintain genetic similarity. In both the
+merger-without-stem-migration and the merger-with-stem-migration, one of the
+stem populations was inferred to have a short, severe bottleneck, which led to
+increased $F_{ST}$ between stem populations during that time. However, this
+period of relatively large $F_{ST}$ did not result in large $F_{ST}$ between
+more recent pairs of populations due to their roughly similar levels of
+contributions to the ancestors of modern humans. \aprcomment{We caution that
+the severe bottleneck may be an artifact of the model parameterization?}
 
 \subsection{$f_4$ statistics between pairs of contemporary and pairs of ancient populations}
+
+To evaluate the degree to which divergence among modern populations is
+inherited from divergence between early stem population, we computed f4
+statistics of the form
+\begin{equation}
+    f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = 
+    \E[(f_{\text{stem}_1} - f_{\text{stem}_2})(f_{\text{modern}_1}-f_{\text{modern}_2})]
+    \label{eq:f4}
+\end{equation}
+This statistic measures co-linearity of genetic drift between modern
+populations $(f_{\text{modern}_1}-f_{\text{modern}_2})$ and stem populations
+$(f_{\text{stem}_1}-f_{\text{stem}_2})$ (see
+Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}
+and \citet{Lipson2020-cq}).
+
+To contrast these statistics to divergence across modern populations,
+\begin{equation}
+    f_2(\text{modern}_1, \text{modern}_2) =
+    \E[(f_{\text{modern}_1}-f_{\text{modern}_2})^2]),
+\end{equation}
+we interpret the $f$-statistics as scalar products between vectors in 
+allele frequency space. In particular, the vector 
+$v=(f_{\text{modern}_1}-f_{\text{modern}_2})$, 
+representing differences in allele frequencies across modern populations,
+can be written as $v=v_\perp + v_\parallel$, a component perpendicular and
+a component parallel to the divergence $(f_{\text{stem}_1}-f_{\text{stem}_2})$
+across stems. We can then decompose 
+$f_2(\text{modern}_1,\text{modern}_2) = v_\perp^2+v_\parallel^2$, with  
+\begin{equation}
+    v_\perp^2 = f_4^2(\text{stem}_1, \text{stem}_2; 
+    \text{modern}_1, \text{modern}_2)/f_2(\text{stem}_1, \text{stem}_2)
+\end{equation}
+
+Finally, the proportion of modern-day structure
+$f_2(\text{modern}_1, \text{modern}_2)$ that is accounted for by the
+component parallel to ancient population structure is
+\begin{equation}
+    \alpha^2 = \frac{v_\perp^2}{f_2(\text{modern}_1, \text{modern}_2)} =
+    \frac{f_4^2(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)}
+    {f_2(\text{stem}_1,\text{stem}_2)f_2(\text{modern}_1, \text{modern}_2)}.
+\end{equation}
+$\alpha^2$ can be interpreted as the proportion of the $f_2$
+divergence across modern populations that is aligned with the differences in
+allele frequencies across ancient populations. However, drift occurring after
+the time of sampling of ancient populations is not correlated to drift
+occurring before that time. For this reason, we also interpret $\alpha^2$ as the
+proportion of the $f_2$ divergence across modern populations that is explained by
+the differences in allele frequencies across ancient populations. We report
+this predicted statistic for each of the four inferred demographic models in
+Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.75\textwidth]{figures/supp-f4-scenarios}
+    \caption{
+        \textbf{Present day structure attributable to ancient population structure.}
+        The statistic 
+        $f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)$
+        captures how much of present-day population structure can be attributed to 
+        population structure among the stems. The blue line represents genetic drift 
+        along ancestral branches between modern populations, while the red line 
+        represents drift between stem populations. $f_4$ measures the overlap 
+        between the blue and red line, which corresponds to shared genetic drift 
+        between the two pairs. 
+        In scenario A, modern populations are descended from distinct stems, and 
+        $f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)$
+        equals the drift between stems, $f_2(\text{stem}_1, \text{stem}_2)$.
+        In scenario B, despite deep population structure, there is no shared drift
+        and $f_4=0$. That is, present-day population structure is unrelated to
+        structure among stems. In more realistic scenario C, where $\text{modern}_1$
+        receives a proportion $m$ of its ancestry from the dotted arrow, the overlap
+        is only found along the corresponding path, and $f_4$ depends on both
+        ancestry proportion $m$ and drift between stems,
+        $f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)
+        = mf_2(\text{stem}_1, \text{stem}_2)$
+        \aprcomment{Get better resolution image from Simon}
+    }
+    \label{fig:supp-f4-scenarios}
+\end{figure}
 
@@
 \subsection{Mutation versus recombination rates}
+
+\subsection{Re-inference of an IM model from simulated data}
 

### 2022-01-05 paper.tex
add abstract, intro, and some results

@@
 \documentclass[]{article}
-\usepackage[round]{natbib}
+\usepackage[super,numbers]{natbib}
 
@@
 \newcommand{\bmhcomment}[1]{{\textcolor{blue}{BMH: #1}}}
-\newcommand{\aprcomment}[1]{{\textcolor{purple}{APR: #1}}}
-\newcommand{\tdwcomment}[1]{{\textcolor{orange}{TDW: #1}}}
+\newcommand{\aprcomment}[1]{{\textcolor{magenta}{APR: #1}}}
+\newcommand{\tdwcomment}[1]{{\textcolor{cyan}{TDW: #1}}}
 
@@
 \abstract{
-A very simple template for an article class document.
+While a modern human origin within Africa is now broadly accepted, considerable
+uncertainty surrounds specific models of divergence and migration across the
+continent. Progress is hampered by a paucity of fossil and genomic data, as
+well as variability in dating. Here we use linkage disequilibrium and
+diversity-based statistics, optimized for rapid, complex demographic inference
+to discriminate among such models. We infer detailed demographic models for
+populations across Africa, including representatives from eastern and western
+groups, as well as 44 newly whole-genome sequenced individuals from the Nama
+(Khoe-San). Despite the complexity of African population history, present-day
+population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
+population divergence among contemporary populations occurs 120-135ka, between
+the Khoe-San and other groups. Prior to the divergence of contemporary African
+groups, we infer long-lasting structure between two or more weakly
+differentiated ancestral Homo spp. populations connected by gene flow over
+hundreds of thousands of years (i.e. a weakly structured stem). We find that
+weakly structured stem models provide more likely explanations of polymorphism
+that had previously been attributed to contributions from archaic hominins in
+Africa.
+\sgcomment{IF we were not afraid of the racists, we could write something like:
+"Despite the genetic similarity between these populations, 4 to 10\% of random
+genetic differences among [modern human populations] can be attributed to this
+early genetic drift.}
+In contrast to models with archaic introgression, we predict that fossil
+remains from coexisting ancestral populations should be morphologically
+similar. We show that model misspecification explains variation in previous
+divergence time estimates and argue that studying a suite of models is key to
+robust inferences about deep history.
 }
@@
 
-The intro..
+Archaeological sites from the Middle Stone Age (approx. 300ka-40ka) are widely
+distributed across Africa, and are particularly well represented in the
+northern, eastern and southern parts of the continent. Similarly, fossil crania
+such as those from the sites of Jebel Irhoud \citep{Hublin2017-cq}, Herto
+\citep{White2003-bk} and Klasies River \citep{Deacon1995-rx} demonstrate that
+anatomically derived \emph{Homo sapiens} features were also present across the
+continent during this period. It has been difficult to reconcile these lines of
+evidence with evidence from genomics, which have suggested a predominantly
+tree-like model of recent population divergence from a single ancestral
+population. It is unclear whether fossil specimens and archaeological sites
+represent populations which contributed to our ancestors as population
+precedents, or were local “dead-ends” from which contemporary \emph{Homo
+sapiens} do not descend. Recently, synthetic attempts to reconcile genetic and
+paleoanthropological data include proposals for a Pan-African origin of
+\emph{Homo sapiens} by which populations in many regions of the continent
+contributed to the formation of Homo sapiens beginning at least 300ka
+\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
+
+Genetic models have been hampered in their contribution to this discussion
+because they primarily assume (or, at least, have been tested under) a
+tree-like model of isolation-with-migration. Alternative theoretical scenarios
+have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
+population coalescence and fragmentation \citep{Scerri2019-xg}. These
+approaches are more challenging to interpret and fit to data. However, new
+population genetic inference tools now allow for inference involving tens to
+hundreds of genomes from multiple populations and greater complexity
+\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
+Neanderthal admixture with modern humans in Eurasia, several recent articles
+have shown that introducing an archaic ghost population contributing to African
+populations in the period surrounding the Out-of-Africa migration event
+substantially improves the description of genetic data relative to
+single-origin models
+\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
+This has driven speculation about the geographic range of this ghost
+population, possible links to specific archaic remains, and the possibility of
+finding ancient DNA evidence (e.g., \citet{Hsieh2016-gk}). However, these prior
+articles share two weaknesses. First, they only contrast a single-origin model
+with an archaic admixture model, leaving out other plausible models
+(Figure~\ref{fig:supp-possible-models} and \citet{Henn2018-rf}). Second, they
+focus on a small subset of African diversity, either because of small sample
+sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
+recruited populations of recent West African ancestry. While ancient DNA from
+Eurasia has helped us understand early human history outside of Africa, there
+is no comparably ancient DNA to elucidate early history in Africa.
+
+Here, we therefore aim to discriminate among a broader set of demographic
+models by interrogating the genomes of contemporary populations. We take as our
+starting point 4 classes of models (single population expansion, single
+    population expansion with regional persistence, archaic admixture, and
+multi-regional evolution, Figure~\ref{fig:supp-possible-models}), using 290
+genomes from southern, eastern, and western Africa as well as Eurasia. By
+including geographically and genetically diverse populations across Africa, we
+infer demographic models that explain more aspects of genetic diversity in more
+populations than previously reported. These analyses confirm the inadequacy of
+tree-like models and provide an opportunity to directly evaluate a wide range
+of alternative models.
 
@@
 
+We inferred detailed demographic histories using 4x-8x whole-genome sequencing
+data for four diverse African populations, comprising the Nama (Khoe-San from
+South Africa, newly presented here), Mende (MSL from Sierre Leone
+\citep{1000_Genomes_Project_Consortium2015-zq}), Gumuz (recent descendents of a
+hunter-gatherer group from Ethiopia \citep{Gurdasani2015-qy,Gopalan2019-wd}),
+and Eastern African agriculturalists (Amhara and Oromo from Ethiopia
+\citep{Gurdasani2015-qy}). The Amhara and Oromo populations, despite speaking
+distinct Afro-Asiatic languages, are highly genetically similar
+\citep{Pagani2015-pz,Gopalan2019-wd} and thus the two groups were combined for
+a larger sample size (Figure~\ref{fig:1}). We also included the British
+(GBR) from the 1000 Genomes Project in our demographic models as a
+representative source of back-to-Africa gene flow and recent colonial admixture
+in South Africa. Finally, we added a high-coverage ancient Neanderthal genome
+from Vindija Cave, Croatia \citep{Prufer2017-kk} to account for archaic gene
+flow from Neanderthals in non-Africans and gauge the relative time depth of
+divergence, assuming Neanderthals diverged 550ka from a common stem. For each
+population, we computed low-order allele frequency and linkage disequilibrium
+(LD) statistics that are well suited for both low- and high-coverage genomes
+\citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a maximum-likelihood inference
+framework, we then fit to these statistics a family of parameterized
+demographic models that involve population splits, size changes, continuous and
+variable migration rates, and punctuated admixture events, to learn about the
+nature of population structure over the past million years.
+
+\begin{figure}[ht]
+\begin{center}
+\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
+\caption{\textbf{Geography and genetic diversity within Africa}.
+    \aprcomment{In the models presented below, we focus on the Nama, 
+    Mende (MSL), Gumuz, and Oromo and Amhara.}
+}
+\label{fig:1}
+\end{center}
+\end{figure}
+
 \subsection*{A Late Middle Stone Age common ancestry for contemporary humans}
+
+We began with a model of geographic expansion from a single ancestral source
+followed by migration between populations
+(Figure~\ref{fig:supp-possible-models}A), without allowing for contribution
+from an archaic African lineage. As expected \citep{Ragsdale2019-nt}, this
+first model was a poor fit to the data qualitatively
+(Figure~\ref{fig:supp-single-origin-fits}) and quantitatively (log-likelihood
+($LL$) $\approx -189,000$, Table~\ref{tab:supp-single-origin}). We next explored a
+suite of models in which population structure predates the differentiation of
+contemporary groups, including models allowing for ancestral reticulation
+(Figure~\ref{fig:supp-possible-models}B), archaic admixture
+(Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
+(Figure~\ref{fig:supp-possible-models}D).
+
+Regardless of the model choice for early epochs, inference of recent human
+demographic history from two thousand \aprcomment{why two thousand, instead of
+just ``up to 150ka''?} to 150ka was remarkably robust. The earliest divergence
+among contemporary sampled human populations differentiates the southern
+African Nama from other African groups between 110--135ka, with low to moderate
+levels of subsequent gene flow \aprcomment{(Table 1)}. In none of the
+high-likelihood models which we explored did the divergence between Nama and
+other populations exceed $\sim$150ka. We conclude that, regardless of model
+specification, geographic patterns of \emph{Homo sapiens} population structure
+date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
+Although we find evidence for earlier population structure in Africa (see
+below), present-day populations cannot be easily mapped onto these more ancient
+stem groups as only a small proportion of drift between modern populations can
+be attributed to drift between stems \aprcomment{(Figure 3)}.
+
+Given this consistency in inferred recent history and the numerical challenge
+of optimizing a large number of parameters, we fixed several parameters related
+to recent population history so as to focus on more ancient events (Supp.
+Methods). The time of divergence of Western and Eastern African populations was
+set to 60ka, just prior to the split of Eurasians and East Africans 50ka. Both
+models predict relatively high subsequent gene flow between the two regions
+($m=2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
+following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
+Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
+affected the ancestors of the Ethiopian agricultural populations, comprising
+over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
+years also saw major demographic changes, including strong population growth
+for Western Africans as they specialized in yam and oil palm agriculture
+(estimated 3-fold growth). We observe significant gene flow from the Amhara and
+Oromo into the Nama, a signal which is likely a proxy for the movement of
+Eastern African caprid and cattle pastoralists
+\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
+contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
+was estimated at 15\%, similar to proportions inferred by ADMIXTURE
+\aprcomment{cite} (Figure~\ref{fig:1}).
 
@@
 A placeholder citation \citep{Kelleher2016-lw}.
-
-\begin{figure}[ht]
-\begin{center}
-\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
-\caption{\textbf{The first figure}.
-    The geographic and genetic diversity of populations across Africa.
-}
-\label{fig:1}
-\end{center}
-\end{figure}
 
@@
 }
-\label{fig:r3}
+\label{fig:3}
 \end{center}
@@
 }
-\label{fig:r4}
+\label{fig:4}
 \end{center}
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{unsrtnat}
 \bibliography{paper}

### 2022-01-05 supplement.tex
add abstract, intro, and some results

@@
 \usepackage{bm}
+\usepackage{xspace}
 
@@
 \newcommand{\E}{\mathbb{E}}
+\newcommand{\moments}{\texttt{moments}\xspace}
+\newcommand{\Relate}{\texttt{Relate}\xspace}
+\newcommand{\msprime}{\texttt{msprime}\xspace}
+\newcommand{\demes}{\texttt{demes}\xspace}
 
@@
     \item Combined with the high coverage Vindija neanderthal genome
-    \item For running Relate, we used a larger set of populations. We kept all 
+    \item For running \Relate, we used a larger set of populations. We kept all 
         African 1000 Genomes populations, along with the Nama, Gumuz, Oromo, and
@@
 All analyses presented in this work focus on biallelic single nucleotide
-polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the moments-LD
-analysis, we focused on intergenic locations because these appear less affected
-by natural selection compared to both synonymous and nonsynonymous variation
-\citep{Ragsdale2018-dd}. To enable comparison with Neanderthal DNA, we excluded
-regions for which the Vindija Neanderthal sample had less than 100 contiguous
-base pairs.
+polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the
+\moments-LD analysis, we focused on intergenic regions because these
+appear less affected by natural selection compared to both synonymous and
+nonsynonymous variation \citep{Ragsdale2018-dd}. To enable comparison with
+Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
+had less than 100 contiguous base pairs.
 
@@
 recursions described in \citet{Ragsdale2019-nt} and implemented in the software
-\emph{moments.LD} (\url{https://bitbucket.org/simongravel/moments/src/main/}).
+\moments (\url{https://bitbucket.org/simongravel/moments/src/main/}).
 The model computes expected patterns of single-nucleotide pairwise diversity
@@
 Estimated statistics for populations with smaller sample sizes have greater
-uncertainty, which is largely accounted for by computing variances and
-covariances via bootstrap, assuming our samples are unrelated and the
-population is randomly mating.
-
-\sgcomment{I don't think that this is true if our bootstrap is
-over genomic regions. In an infinite genome with a tiny sample size, we would
-estimate no uncertainty...  Or at least, it is only true if the Neandertal form
-a truly randomly mating population...  }
+uncertainty, which is accounted for by computing variances and covariances via
+bootstrap over genomic regions, assuming that there is no correlation between
+distant genomic regions (i.e., no long-range LD).
 
@@
 
-moments-LD uses a composite likelihood approach to simultaneously fit relative
-pairwise diversity and LD statistics over a range of recombination distances.
-Likelihoods were computed independently for pairwise diversity and each
-recombination distance bin using a multivariate Guassian likelihood function as
-described in Section \ref{sec:computing-stats}. These were multiplied across
-bins and the single set of heterozygosity statistics following the approach
-detailed in \citet{Ragsdale2019-nt}. For each model tested, we ran multiple
-rounds of optimization, alternating between the \emph{optimize\_log\_fim},
-\emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
-parameter space and hone in on the best fit parameters. Initial guesses for
-parameters were chosen from demographically plausible starting points and then
-perturbed to explore space, using gradient descent (on the log of the
-parameters). The best fits from these initial rounds of optimization were then
-chosen as the starting points for optimization using the Powell and/or the
-L-BFGS-B methods (as implemented in the SciPy optimization package
-\cite{Virtanen2020-kr}). This process was repeated with alternating
-optimization methods until the best fit parameter set converged consistently.
+\moments-LD uses a composite likelihood approach to simultaneously fit
+relative pairwise diversity and LD statistics over a range of recombination
+distances.  Likelihoods were computed independently for pairwise diversity and
+each recombination distance bin using a multivariate Guassian likelihood
+function as described in Section \ref{sec:computing-stats}. These were
+multiplied across bins and the single set of heterozygosity statistics
+following the approach detailed in \citet{Ragsdale2019-nt}. For each model
+tested, we ran multiple rounds of optimization, alternating between the
+\emph{optimize\_log\_fim}, \emph{optimize\_log\_powell}, and
+\emph{optimize\_lbfgsb} methods to explore parameter space and hone in on the
+best fit parameters. Initial guesses for parameters were chosen from
+demographically plausible starting points and then perturbed to explore space,
+using gradient descent (on the log of the parameters). The best fits from these
+initial rounds of optimization were then chosen as the starting points for
+optimization using the Powell and/or the L-BFGS-B methods (as implemented in
+the SciPy optimization package \cite{Virtanen2020-kr}). This process was
+repeated with alternating optimization methods until the best fit parameter set
+converged consistently.
 
@@
 
-We used Relate version 1.0.16 \citep{Speidel2019-nj} to reconstruct genome-wide
+We used \Relate version 1.0.16 \citep{Speidel2019-nj} to reconstruct genome-wide
 gene genealogies using a combined set of 1000 Genomes and African Diversity
@@
 29 years, a mutation rate of 1.25e-8, and followed the standard pipeline
-described in the Relate documentation.
+described in the \Relate documentation.
 
 From the reconstructed gene genealogies, we computed coalescence rates within
-and between populations using Relate’s function RelateCoalescentRate. This
-allows for an estimate of the instantaneous inverse coalescence rate (IIRC) for
-samples drawn within each population (the inverse of which is often interpreted
-as the effective population size history), and the relative cross coalescence
-rates between pairs of populations (which are commonly used to estimate
-divergence times).
+and between populations using \Relate’s function \texttt{RelateCoalescentRate}.
+This allows for an estimate of the instantaneous inverse coalescence rate
+(IIRC) for samples drawn within each population (the inverse of which is often
+interpreted as the effective population size history), and the relative cross
+coalescence rates between pairs of populations (which are commonly used to
+estimate divergence times).
 
@@
 
+Here we focus on four parameterizations of early human history: (1) a “single
+origin” with recent expansion, (2) primary and second stem populations with
+“continuous migration”, (3) two stem populations that do not exchange migrants
+before merging to form regional African populations, referred to as a “merger
+without stem migration”, and (4) two stem populations that are allowed to
+exchange migrants before merging to form regional African populations, referred
+to as a “merger with stem migration”. Given the best fit parameters for each
+model
+(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
+we used \msprime version 1.0.4 \citep{Kelleher2016-lw,Baumdicker2021-mj} to
+simulate sampled individual genomes to match the data used in inference (see
+simulation details below). Using these simulations, we compared additional
+statistics and inferences between simulated and real data that were not used in
+the original optimization. These comparisons serve as independent validations
+of our inferred demographic models by assessing how well those models fit
+aspects of the data that have been used in other studies as evidence of
+population structure or archaic admixture from an unidentified hominin species
+within Africa.
+
 \subsection{Simulation details}
 
+The four inferred demographic models (single origin, continuous migration,
+merger without stem migration, and merger with stem migration) were converted
+to \demes format (Gower, Ragsdale, \emph{et al.}, in prep) and used as
+the input demography for \msprime simulations. We sampled the same
+number of individuals from each population as used in the inference (Nama: 44,
+Mende: 85, Gumuz: 23, combined Oromo/Amhara: 46, British: 91, and Vindija
+Neanderthal: 1). We assumed a generation time of 29 years, a per-base mutation
+rate of $1.25\times10^{-8}$, a per-base recombination rate of $1\times10^{-8}$
+(corresponding to 1cM/Mb), and we simulated 100 replicates with sequence
+lengths of 50Mb per replicate. We computed the conditional site frequency
+spectrum (cSFS) conditioned on observing the derived variant in the Neanderthal
+sample, and reconstructed gene genealogies using \Relate to obtain coalescence
+rate trajectories and distributions of deep branches passing through the
+Neanderthal branch. Each analysis followed the same procedure as described
+above in the analysis of the real data.
+
 \subsection{cSFS prediction under inferred models}
 
+We compared the observed cSFS to model predictions for the Nama, Mende, and
+Gumuz populations. The cSFS from data is sensitive to ancestral allele
+misidentification, but our simulations assume perfect knowledge of the
+ancestral allele. To investigate the role of misidentification, we compared
+predicted and observed cSFS for transitions and transversions separately as
+well as jointly.
+
+For each comparison, we fit a misidentification parameter $p_{misid}$ as well
+as a scaled mutation rate $\theta$. The misidentification parameter is the
+fraction $p_{misid}$ of loci that have the ancestral state misidentified.
+Accounting for $p_{misid}$ mixes the predicted joint population SFS with a
+small amount of the ancestral-allele-reversed SFS (i.e. for a two-population
+joint SFS, $SFS_{misid}(i, j)=(1-p_{misid})*SFS(i, j) + p_{misid}*SFS(n-i,
+m-j)$), where $n$ and $m$ are the sample sizes from the two populations. The
+cSFS for the first population (conditioning on the second) is found by
+projecting the sample size to $(n, 1)$ and taking $cSFS=SFS(:,1)$. $\theta$ and
+$p_{misid}$ were chosen to maximize a likelihood assuming that each observed
+entry in the cSFS is drawn from a Poisson distribution \citep{Sawyer1992-rt}.
+
 \section{Supplementary results}
 
-\subsection{Conditional SFS}
-
-\subsection{Relate curves from inferred models}
-
-\subsection{Distribution of deep branch affinities to Neanderthal sequence}
+\subsection{Alternative statistics for four models}
+
+\subsubsection{Conditional SFS}
+
+The single-origin model fit the cSFS relatively poorly across the three
+populations compared to the models that allow structure between stem
+populations (Fig.~\ref{fig:supp-csfs-single-origin}). The model that best fit
+the cSFS was the merger-with-stem-migration model, which also provided the best
+fit to pairwise diversity and LD statistics in the initial optimization. For
+the Mende and Gumuz, the merger-with-stem-migration model fits the data very
+well, while it underestimates rare variants in the Nama cSFS
+(Fig.~\ref{fig:supp-csfs-merger-with-stem-migration}). Models learned from the
+two-locus statistics provided a better qualitative description of the cSFS than
+models previously fit directly to the cSFS  (see, e.g., figure 1 in
+\citet{Durvasula2020-td}). We found that the inferred ancestral
+misidentification rates were concordant across populations. The inferred rates
+of ancestral misidentification were about twice as high for transitions
+relative to transversions as expected, qualitatively, from the higher
+transition mutation rate \citep{Hernandez2007-mf}.
+
+\subsubsection{Relate curves from inferred models}
+
+\aprcomment{TODO}
+
+\subsubsection{Distribution of deep branch affinities to Neanderthal sequence}
+
+\aprcomment{TODO}
 
@@
 
-\subsection{Re-inference of an IM model from simulated data}
+Prior studies primarily relied on the mutation clock in order to date events
+such as population divergence, and uncertainty about the mutation rate
+translates in uncertainty in timing estimates (e.g., \citet{Moorjani2016-qj}).
+Recent whole genome sequencing studies from non-pathological pedigrees have
+typically estimated mutation rates to be $1.2-1.3\times10^{-8}$ per site per
+generation \citep{Sasani2019-zp,Tian2019-zv}. A rate of $1.3\times10^{-8}$ also
+approximates the rate predicted by a mutational model from Gnomad
+\citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
+demographic inference in this study. However, \moments differs from other
+approaches by relying on a recombination clock, which has been argued to be
+more robust to estimation than a mutational clock \citep{Moorjani2016-ur}. To
+fit patterns of diversity which were inferred using the recombination map here,
+mutation rates on the order of $2\times10^{-8}$ would be needed.
+\aprcomment{This discrepancy between the mutation clock and recombination
+clock\dots}
+
+\subsection{Re-inference of IM models from simulated data}
+
+\aprcomment{TODO}
 
@@
 
-\addcontentsline{toc}{section}{Supporting tables}
-\section*{Supporting tables}
+\addcontentsline{toc}{section}{Supporting tables and figures}
+\section*{Supporting tables and figures}
 
@@
 
-\addcontentsline{toc}{section}{Supporting figures}
-\section*{Supporting figures}
-
 \begin{figure}[ht]

### 2022-01-05 paper.tex
fill in discussion

@@
 \usepackage{color}
+\usepackage{booktabs}
 
@@
 population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
-population divergence among contemporary populations occurs 120-135ka, between
+population divergence among contemporary populations occurs 120-135kya, between
 the Khoe-San and other groups. Prior to the divergence of contemporary African
@@
 
-Archaeological sites from the Middle Stone Age (approx. 300ka-40ka) are widely
+Archaeological sites from the Middle Stone Age (approx. 300kya-40kya) are widely
 distributed across Africa, and are particularly well represented in the
@@
 \emph{Homo sapiens} by which populations in many regions of the continent
-contributed to the formation of Homo sapiens beginning at least 300ka
+contributed to the formation of Homo sapiens beginning at least 300kya
 \citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
@@
 data for four diverse African populations, comprising the Nama (Khoe-San from
-South Africa, newly presented here), Mende (MSL from Sierre Leone
-\citep{1000_Genomes_Project_Consortium2015-zq}), Gumuz (recent descendents of a
-hunter-gatherer group from Ethiopia \citep{Gurdasani2015-qy,Gopalan2019-wd}),
-and Eastern African agriculturalists (Amhara and Oromo from Ethiopia
-\citep{Gurdasani2015-qy}). The Amhara and Oromo populations, despite speaking
-distinct Afro-Asiatic languages, are highly genetically similar
-\citep{Pagani2015-pz,Gopalan2019-wd} and thus the two groups were combined for
-a larger sample size (Figure~\ref{fig:1}). We also included the British
-(GBR) from the 1000 Genomes Project in our demographic models as a
-representative source of back-to-Africa gene flow and recent colonial admixture
-in South Africa. Finally, we added a high-coverage ancient Neanderthal genome
-from Vindija Cave, Croatia \citep{Prufer2017-kk} to account for archaic gene
-flow from Neanderthals in non-Africans and gauge the relative time depth of
-divergence, assuming Neanderthals diverged 550ka from a common stem. For each
-population, we computed low-order allele frequency and linkage disequilibrium
-(LD) statistics that are well suited for both low- and high-coverage genomes
-\citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a maximum-likelihood inference
-framework, we then fit to these statistics a family of parameterized
-demographic models that involve population splits, size changes, continuous and
-variable migration rates, and punctuated admixture events, to learn about the
-nature of population structure over the past million years.
+South Africa, newly presented here), Mende (MSL from Sierre Leone, Phase 3 of
+the 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}), Gumuz
+(recent descendents of a hunter-gatherer group from Ethiopia
+\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
+(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
+populations, despite speaking distinct Afro-Asiatic languages, are highly
+genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
+groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
+included the British (GBR) from the 1000 Genomes Project in our demographic
+models as a representative source of back-to-Africa gene flow and recent
+colonial admixture in South Africa. Finally, we added a high-coverage ancient
+Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
+for archaic gene flow from Neanderthals in non-Africans and gauge the relative
+time depth of divergence, assuming Neanderthals diverged 550kya from a common
+stem. For each population, we computed low-order allele frequency and linkage
+disequilibrium (LD) statistics that are well suited for both low- and
+high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
+maximum-likelihood inference framework, we then fit to these statistics a
+family of parameterized demographic models that involve population splits, size
+changes, continuous and variable migration rates, and punctuated admixture
+events, to learn about the nature of population structure over the past million
+years.
 
@@
 demographic history from two thousand \aprcomment{why two thousand, instead of
-just ``up to 150ka''?} to 150ka was remarkably robust. The earliest divergence
+just ``up to 150kya''?} to 150kya was remarkably robust. The earliest divergence
 among contemporary sampled human populations differentiates the southern
-African Nama from other African groups between 110--135ka, with low to moderate
+African Nama from other African groups between 110--135kya, with low to moderate
 levels of subsequent gene flow \aprcomment{(Table 1)}. In none of the
 high-likelihood models which we explored did the divergence between Nama and
-other populations exceed $\sim$150ka. We conclude that, regardless of model
+other populations exceed $\sim$150kya. We conclude that, regardless of model
 specification, geographic patterns of \emph{Homo sapiens} population structure
@@
 Methods). The time of divergence of Western and Eastern African populations was
-set to 60ka, just prior to the split of Eurasians and East Africans 50ka. Both
+set to 60kya, just prior to the split of Eurasians and East Africans 50kya. Both
 models predict relatively high subsequent gene flow between the two regions
@@
 
-\subsection*{Deep population structure but not archaic admixture within Africa}
-
-\subsection*{Reconciling multiple lines of genetic evidence}
-
-\section*{Discussion}
-
-\subsection*{The Middle Stone Age in Africa}
-
-\subsection*{Contrasting archaic admixture and a weakly structured stem}
-
-\section*{Methods}
-A placeholder citation \citep{Kelleher2016-lw}.
-
 \begin{figure}[ht]
@@
 \makebox[\textwidth][c]{} % \includegraphics{figures/fig2.png}}
-\caption{\textbf{The second main figure}.
+\caption{\textbf{}.
     A placeholder - best fit model(s).
@@
 \end{figure}
+
+\subsection*{Deep population structure but not archaic admixture within Africa}
+
+To account for population structure prior to 135kya, we allowed for two or more
+“stem” populations which could diverge either before or after the Neanderthal
+split. We considered models both with and without migration between these stem
+populations, and in both cases we tested two different types of gene exchange
+during the expansion phase. 1) One of the stem population expands (splits into
+contemporary populations), and the other stem population(s) has continuous
+symmetric migration with those populations; or 2) expansion of one or more of
+the stem populations, with an instantaneous pulse (or “merger”) from the other
+stem population. These models are meant to capture scenarios 2B and 2C (Figure
+2). \aprcomment{update this last sentence}
+
+Allowing for continuous migration between the stem populations substantially
+improved the fits relative to zero migration (e.g. $LL \approx -102,600$ vs.
+$-107,700$). With continuous migration, population structure extends back to
+1.1--1.4Mya \aprcomment{(Table 1)}. Migration between the stems in these models
+is low but non-negligible, with a fraction of migrant lineages each generation
+estimated as $m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is
+at least as large as migration inferred between the Ethiopian East Africans and
+British populations over the past 50kya \aprcomment{(Table 1)}. This ongoing
+(or, at least, periodic) gene flow qualitatively distinguishes these models
+from previously proposed archaic admixture models \aprcomment{(Figure S1C)} as
+the early branches remain closely related throughout (Figure 3).
+\aprcomment{this last phrase may need to be updated, with regard to Fig
+\ref{fig:supp-FST-panels}}
+
+Likelihood-based parameter confidence intervals are relatively narrow within
+each specified model
+(Tables~\ref{tab:supp-single-origin-fits}--\ref{tab:supp-merger-with-stem-migration-fits}),
+reflecting an informative statistical approach. As expected, model assumptions
+have a larger impact on parameter estimates. To convey model uncertainty, we
+highlight features of two models with the highest \aprcomment{not the
+highest\ldots merger-without-stem-migration at -107,650}log likelihoods. These
+differ primarily in whether early stems are assumed to give rise to modern
+populations in multiple discrete merger events or in continuous migrations
+\aprcomment{(Figure 2)}. Additional models are discussed in
+\aprcomment{Supplement XXX}.
+
+The “Multiple Merger” model has better log-likelihood than the “Continuous
+Migration” model (-102,600 vs. -116,000, respectively). The two models differ
+in the timing of divergence and the relative Ne of the two stems
+(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
+during the Middle Stone Age.
+
+Under the continuous-migration model, the two stems give rise to contemporary
+populations in western, southern and eastern Africa but with differing
+proportions of ancestry derived from “Stem 1” and “Stem 2” (Figure 2). For
+example, the Nama and Gumuz have lower levels of migration with Stem 2
+($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), relative to the
+Mende. The Mende continue to exchange migrants with Stem 2
+($m=1.6\times10^{-4}$) long after their separation 60kya; this is a similar
+level of gene flow as between the Mende and Eastern Africans (Table 1). 
+\aprcomment{paragraphs above need work, and condensing}
+
+\aprcomment{also below, too much information - we don't actually believe these
+models}
+
+In contrast, under the Multiple Merger model, we allowed for stem populations
+to have large instantaneous contributions, in different proportions, to form
+the different contemporary groups. However, the ghost populations do not neatly
+line up one-to-one with southern, western and eastern populations. We observe a
+deep bottleneck in Stem 1 down to $N_e=117$. This represents the lower bound
+allowed in our optimization (i.e. 1\% of the $N_{ancestral}$) and the size of
+this bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period
+of exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
+479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
+166$kya). These populations evolve independently until approximately 119kya
+when Stem 1S and Stem 2 merge to form the ancestors of the Nama, with
+proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 merge to
+form the ancestors of the Western Africans and Eastern Africans (and thus also
+all individuals who later disperse during the Out of Africa event). The
+ancestral Western / Eastern African group derives equal ancestry from the two
+stems (50\% from Stem 1E and Stem 2, Table 1). Finally, the Mende receive a
+large pulse of gene flow from Stem 2, replacing 18\% of their population 25kya.
+The later Stem 2 contribution to the Western African Mende resulted in better
+model fits (LL). This may indicate that an ancestral Stem 2 population
+occupied Western or Central Africa, broadly speaking. The differing proportions
+in the Nama and Eastern Africans may also indicate geographic separation of
+Stem 1S in southern Africa and Stem 1E in eastern Africa.
+
+\begin{table}
+    \centering
+    \begin{tabular}{l|l|l}
+        \toprule
+        Parameter & Value & Something else \\
+        \midrule
+        x & y & z \\
+        \bottomrule
+    \end{tabular}
+    \caption{caption}
+    \label{tab:migration-rates}
+\end{table}
+
+\subsection*{Reconciling multiple lines of genetic evidence}
+
+Previous studies have found support for archaic admixture in Africa using
+two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
+(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
+\citep{Speidel2019-nj}. \aprcomment{Will need to check how strongly the Relate
+paper claims that this supports archaic admixture vs simply deep structure. It
+wasn't a model fit, in any case.} However, none of these studies considered a
+weakly structured stem. We validated our Continuous Migration and Multiple
+Merger models with additional independent approaches. We find that the two
+models describe the cSFS accurately \aprcomment{(Figure SX)}, even though this
+statistic was not used in the fit.  Both models outperform archaic models fit
+directly to the cSFS (for example, compare with Figure~1 in
+\citet{Durvasula2020-td}).
+
+Similarly, we generated estimates of effective population size as inferred from
+coalescence rates with Relate. Changes in Ne over time from prior studies
+strongly support a bottleneck in Out of Africa populations, but African
+populations also dip in effective size around 200kya, although this decrease is
+more modest than the OOA event [Mallick 2016]. One possible explanation is that
+ancestral population structure during the Middle Stone Age inflated Ne and
+subsequent gene flow among ancestral Africans resulted in a lower Ne after
+100-200kya [Chikhi?]. To our knowledge, explicit comparisons of this hypothesis
+with SMC-style approaches have not been published. We simulated genomic data
+under 4 models (Figure SX). All models, including the Single Origin model,
+generate multi-modal curves indicating that the ancestral increase in Ne
+between 100kya-1Mya is a general feature of the coalescent rate in humans.
+Simulated Relate Ne curves under the CM and MM are distinguished by having 2
+and 3 modes, respectively. The Continuous Migration model had the better
+qualitative fit with the data, but the ancestral mode was shifted further back
+in time relative to the data. All simulated models underestimated the level of
+population growth during the Holocene, indicating additional free parameters
+are necessary to refine these growth rates. \aprcomment{Discussion on this
+paragraph moved to issues (issue \#1)}
+
+Relative cross-coalescence rates (rCCR) have recently been used to estimate
+divergence between a pair of populations, as measured by the rate of
+coalescence between two groups divided by the mean within population
+coalescence. Simulations of rCCR accuracy, however, focus on a ‘clean split’
+between populations whereby groups diverge without subsequent gene flow.
+Published estimates of the earliest human divergences with rCCR, which range
+from 150kya-100kya \citep{Bergstrom2021-iw}, may be significantly biased when
+compared to more complex models with gene flow as inferred here. We find that
+midpoint estimates of rCCR are poor estimates for population divergence, often
+underestimating divergence time by 50\% or greater (e.g. Mende vs. Gumuz ~15kya
+compared to a true divergence 60kya, \aprcomment{Supp Figure SX}). We suggest
+that rCCR analyses which do not fit multiple parameters including gene flow
+should be interpreted with caution.
 
@@
 
+\section*{Discussion}
+
+Any attempt at building detailed models of human history is subject to model
+misspecification. This is true of earlier studies, which often assumed that
+data inconsistent with a single origin model must be explained by archaic
+admixture. This is also true of this study. While it remains prohibitive to
+fully explore the space of plausible models of early human \bmhcomment{we use
+this phrase a lot. human or hominin? Tim?} population structure, we sought to
+capture model uncertainty by exploring multiple parameterizations of early
+human population structure. The best-fit models presented here include
+reticulation and low-level migration between early human populations rather
+than archaic admixture. We cannot rule out that more complex models involving
+additional stems, or hybrid models including both weak structure and archaic
+admixture may better explain the data. Because parameters related to the split
+time, migration rates, and relative sizes of the early stems were variable
+across models, reflecting a degree of confounding among these parameters, we
+refrained from introducing additional branches (and therefore additional
+parameters) during that period. \sgcomment{Some statement about hitting the
+boundary} Rather than interpreting the two stems as representing well-defined
+and stable populations over hundreds of thousands of years, we interpret the
+weakly structured stem as consistent, e.g., with a coalescence and
+fragmentation model \citep{Scerri2019-xg}.  Eventual ancient DNA samples from
+Africa could further distinguish the archaic admixture model from the weakly
+structured stem model. \sgcomment{Or models involving more diversity within
+Africa?}
+
+\subsection*{The Middle Stone Age in Africa}
+
+By contrast, our inferred models paint a more consistent picture of the late
+Middle Stone Age as a critical period of change, assuming that estimates from
+the recombination clock accurately relate to geological chronologies
+(Supplement). During the Middle Stone Age, the Multiple Merger indicates three
+major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
+eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
+isolation among the stems is variable across model fits, models with this
+period of divergence, isolation and then a merger event (i.e. a “reticulation”)
+out-performed models with bifurcating divergence and continuous gene flow.
+
+A population reticulation involves multiple stems contributing genetically to
+the formation of a group. One way in which this can happen is through the
+geographic expansion of one or both stems. For example, if during MIS 5, either
+Stem 1S from southern Africa moved northward thereby encountering the Stem 2,
+or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
+could observe disproportionate ancestry contributions from different stems in
+modern groups. We observed two merger events. The first, between Stem 1S and
+Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
+The second ~100kya between Stem 1E and Stem 2, results in the formation of the
+ancestors of East/West Africans as well as later “Out of Africans”.  The rapid
+rise in sea levels and increased precipitation during MIS 5e, following a
+glacial period of aridity across Africa \citep{Blome2012-lw}, might have
+triggered migration inland away from the coasts, as has been suggested for the
+Paleo-Agulhas plain \citep{Marean2014-pg}.
+
+Following these merger events, the stems subsequently fracture into
+subpopulations which then appear to persist over the past ~120kya. These
+subpopulations can be linked to modern-day groups despite subsequent gene flow
+across the continent; for example, a genetic lineage sampled in the Nama has an
+XX probability of being in the ancestral ‘southern’ subpopulation 50kya versus
+XX probability of being in the ‘eastern’ subpopulation (Figure XX). We also
+find that Stem 2 continued to contribute to western Africans during the Last
+Glacial Period, indicative that this gene flow likely occurred in
+western/central Africa (Table 1).
+
+\subsection*{Contrasting archaic admixture and a weakly structured stem}
+
+Evidence for archaic admixture in Eurasia has bolstered the plausibility of
+archaic admixture having also occurred in Africa. For this reason, previous
+work has focused on archaic admixture to explain patterns of polymorphism
+inconsistent with a single origin model. Here, we have shown that weakly
+structured stem models better capture these patterns. They also help explain an
+ecological riddle posed by the archaic admixture model. Neanderthal populations
+were separated from early \emph{Homo sapiens} by thousands of kilometers and
+continental geographic barriers. By contrast, an archaic hominin population in
+Africa would need to have stayed in relative reproductive isolation from the
+human lineage over hundreds of thousands of years despite closer geographic
+proximity and reproductive compatibility. The weakly structured stem model
+resolves this ecological riddle by accommodating continuous or recurrent
+contact between two or more groups present in Africa.
+
+There is evidence for both deleterious and adaptive archaic-derived alleles in
+modern genomes in the form of a depletion of Neanderthal ancestry in regulatory
+regions \citep{Petr2019-xo} or an increased frequency of archaic-related
+haplotypes such as at \emph{EPAS1} among Tibetans (e.g.,
+\citet{Zhang2021-xx}).\sgcomment{Probably not the best reference} Under an
+archaic African admixture model, the estimated 8--10\% introgression rate is
+much higher than Neanderthal gene flow, and would have plausibly been fertile
+ground for dramatic selection for or against archaic-derived haplotypes over a
+short period \citep{Wall2019-ao}. By contrast, adaptation under a weakly
+structured stem would have occurred continuously over much longer periods.
+Patterns of polymorphism that are inconsistent with the single-stem model
+predictions have been used to infer putative archaic admixed segments
+\citep{Plagnol2006-lt,Hsieh2016-gk,Wall2019-ao,Durvasula2020-td}, negative
+selection against such segments \citep{Wall2019-ao}, and pervasive positive
+selection \citep{Schrider2017-kl}. However, such approaches are subject to high
+false positives in the presence of population structure with migration
+\citep{Petr2019-xo}, and their interpretation should be re-examined in light of
+a weakly structured stem model in Africa.
+
+Multiple studies have shown a correspondence between phenotypic
+differentiation, usually assessed with measurements of the cranium, and genetic
+differentiation among human populations and between humans and Neanderthals
+\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. If we
+assume that this correspondence also holds for early human ancestral
+populations in Africa, the weakly structured stem model would predict that the
+fossils deriving from each of the two stems should be morphologically quite
+similar (supplementary text X)\aprcomment{Will we still include this?}, even if
+they are from different parts of Africa, with differences across stems closer
+to those across contemporary human populations than to human-Neanderthal
+differences. Differences within stems would be comparable to differences within
+recent human populations. Identifying the morphological signature of ancient
+structure in the fossil record will require careful analysis of many broadly
+similar remains. Our model further predicts that morphologically distinct
+crania such as XX and YY are unlikely representatives of branches that
+contributed appreciably to modern human ancestries.\sgcomment{Here I am being
+somewhat forceful. We could use conditional to emphasize that these conclusion
+are under the weakly structured stem model.}
+
+\section*{Methods}
+
+See Supplement, for now.
+
 \section*{Acknowledgements}
+
+We are grateful for the DNA contribution from each participant which enabled
+this study; in particular we wish to highlight the generous participation of
+the Richtersveld Nama community in South Africa and help from local research
+assistants Willem DeKlerk and Hendrik Kaimann. Additional assistance and
+community engagement was conducted by Justin Myrick, Chris Gignoux, Caitlen
+Uren and Cedric Werely. We thank the African Genome Diversity Project for data
+generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
+also thank Sriram Sankararam, \dots for helpful discussion. This research was
+supported by an NIH grant R35GM133531 (to BMH) and XXX. The content is solely
+the responsibility of the authors and does not necessarily represent the
+official views of the National Institutes of Health.
 

### 2022-01-06 paper.tex
add relate curves

@@
 populations also dip in effective size around 200kya, although this decrease is
-more modest than the OOA event [Mallick 2016]. One possible explanation is that
-ancestral population structure during the Middle Stone Age inflated Ne and
-subsequent gene flow among ancestral Africans resulted in a lower Ne after
-100-200kya [Chikhi?]. To our knowledge, explicit comparisons of this hypothesis
-with SMC-style approaches have not been published. We simulated genomic data
-under 4 models (Figure SX). All models, including the Single Origin model,
+more modest than the OOA event \citep{Mallick2016-lx}. One possible explanation
+is that ancestral population structure during the Middle Stone Age inflated Ne
+and subsequent gene flow among ancestral Africans resulted in a lower Ne after
+100-200kya \citep{Mazet2016-wn}\aprcomment{Is this a good ref here?}. To our
+knowledge, explicit comparisons of this hypothesis with SMC-style approaches
+have not been published. We simulated genomic data under 4 models
+\aprcomment{(Figure SX)}. All models, including the Single Origin model,
 generate multi-modal curves indicating that the ancestral increase in Ne

### 2022-01-06 supplement.tex
add relate curves

@@
 
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-relate-iicr-data}
+    \caption{
+        \textbf{Inverse instantaneous coalescence rates inferred from the data.}
+    }
+    \label{fig:supp-relate-sizes}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/supp-relate-iicr-simulation}
+    \caption{
+        \textbf{Inverse instantaneous coalescence rates reinferred from simulated data.}
+    }
+    \label{fig:supp-relate-sizes-sim}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/supp-relate-rccr-data}
+    \caption{
+        \textbf{Relative cross-coalescence rates inferred from the data.}
+    }
+    \label{fig:supp-relate-rccr}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/supp-relate-rccr-single-origin}
+    \caption{
+        \textbf{Relative cross-coalescence rates reinferred from simulated data
+        under the single-origin model.}
+    }
+    \label{fig:supp-relate-rccr-sim}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/supp-relate-rccr-continuous-migration}
+    \caption{
+        \textbf{Relative cross-coalescence rates reinferred from simulated data
+        under the continuous-migration model.}
+    }
+    \label{fig:supp-relate-rccr-sim}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/supp-relate-rccr-merger-without-stem-migration}
+    \caption{
+        \textbf{Relative cross-coalescence rates reinferred from simulated data
+        under the merger-without-stem-migration-model.}
+    }
+    \label{fig:supp-relate-rccr-sim}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{figures/supp-relate-rccr-merger-with-stem-migration}
+    \caption{
+        \textbf{Relative cross-coalescence rates reinferred from simulated data
+        under the merger-with-stem-migration model.}
+    }
+    \label{fig:supp-relate-rccr-sim}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    %\includegraphics{figures/supp-deep-brances-data}
+    \caption{
+        \textbf{Neanderthal-matching deep branch distribution from the data}
+    }
+    \label{fig:supp-deep-branches}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    %\includegraphics{figures/supp-deep-branches-sim}
+    \caption{
+        \textbf{Deep branch distribution reinferred from simulated data.}
+    }
+    \label{fig:supp-deep-branches-sim}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    %\includegraphics{figures/supp-IM-misspecification}
+    \caption{
+        \textbf{IM model split times reinferred from simulated data of full model.}
+    }
+    \label{fig:supp-IM-misspecification}
+\end{figure}
+
 \end{document}

### 2022-01-06 supplement.tex
deep branches and IM fit figures
relate coalescence rates

@@
 
+\aprcomment{We used the Omni YRI map (though see figure showing similarity to
+statistics under the HapMapII map). Assumed a generation time of 29 years for
+rescaling times in years to generations.}
+
 \subsection{Computing LD and diversity statistics}\label{sec:computing-stats}
@@
 
-\moments-LD uses a composite likelihood approach to simultaneously fit
-relative pairwise diversity and LD statistics over a range of recombination
-distances.  Likelihoods were computed independently for pairwise diversity and
-each recombination distance bin using a multivariate Guassian likelihood
-function as described in Section \ref{sec:computing-stats}. These were
-multiplied across bins and the single set of heterozygosity statistics
-following the approach detailed in \citet{Ragsdale2019-nt}. For each model
-tested, we ran multiple rounds of optimization, alternating between the
-\emph{optimize\_log\_fim}, \emph{optimize\_log\_powell}, and
-\emph{optimize\_lbfgsb} methods to explore parameter space and hone in on the
-best fit parameters. Initial guesses for parameters were chosen from
-demographically plausible starting points and then perturbed to explore space,
-using gradient descent (on the log of the parameters). The best fits from these
-initial rounds of optimization were then chosen as the starting points for
-optimization using the Powell and/or the L-BFGS-B methods (as implemented in
-the SciPy optimization package \cite{Virtanen2020-kr}). This process was
-repeated with alternating optimization methods until the best fit parameter set
-converged consistently.
+\moments-LD uses a composite likelihood approach to simultaneously fit relative
+pairwise diversity and LD statistics over a range of recombination distances.
+Likelihoods were computed independently for pairwise diversity and each
+recombination distance bin using a multivariate Guassian likelihood function as
+described in Section \ref{sec:computing-stats}. These were multiplied across
+bins and the single set of heterozygosity statistics following the approach
+detailed in \citet{Ragsdale2019-nt}. For each model tested, we ran multiple
+rounds of optimization, alternating between the \emph{optimize\_log\_fim},
+\emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
+parameter space and hone in on the best fit parameters. Initial guesses for
+parameters were chosen from demographically plausible starting points and then
+perturbed to explore space, using gradient descent (on the log of the
+parameters). The best fits from these initial rounds of optimization were then
+chosen as the starting points for optimization using the Powell and/or the
+L-BFGS-B methods (as implemented in the SciPy optimization package
+\cite{Virtanen2020-kr}). This process was repeated with alternating
+optimization methods until the best fit parameter set converged consistently.
 
@@
 This allows for an estimate of the instantaneous inverse coalescence rate
-(IIRC) for samples drawn within each population (the inverse of which is often
+(IICR) for samples drawn within each population (the inverse of which is often
 interpreted as the effective population size history), and the relative cross
@@
 
-\aprcomment{TODO}
+We used \Relate to reconstruct gene genealogies from the simulations under our
+four inferred demographic models. From each of these, following the same
+approach as for the analysis of the actual data, we computed coalescence rates
+within and between sampled populations. From the coalescence rates, we
+estimated the inverse instantaneous coalescence rate (IICR, often interpreted
+as a trajector of $N_e$) for each population, as well as relative cross
+coalescence rates between pairs of populations.
+
+We note that the IICR is a function of not only population sizes, but also of
+population structure and migration, and interpreting it as an actual population
+size history is tenuous.  However, it is still a statistic that can be computed
+from the data and compared to predictions under the model, and can therefore be
+informative about aspects of the data that our models fail to predict.
+\aprcomment{I think we want to say something along these lines above\dots} We
+also note that the coalescence rates inferred from data used a larger set of
+populations from the combined 1000 Genomes/AGRP dataset, while our model
+simulations only included a subset of 5 populations. This may lead to
+differences in resolution of the inferences.
+
+The data shows the characteristic ``hump'' of increased IICR ($N_e$) 100-300k
+for all populations, followed by a decrease in size among all populations that
+is sharpest in the Eurasian populations and also pronounced in the East African
+populations that have high proportions of Eurasian ancestry (Amhara and Oromo)
+(Fig.~\ref{fig:supp-iicr-data}). The model-inferred $N_e$ curves also show this
+general trend for the more recent reduction in $N_e$
+(Fig.~\ref{fig:supp-iicr-sim}). In the the more distant past, $N_e$ fluctuates
+in a manner that is broadly similar across models with observed ``humps'' of
+increased $N_e$. Despite the large differences in parameterizations and
+interpretations of the four inferred demographic models, each of them show
+reconstructed $N_e$ trajectories that are qualitatively similar, highlighting
+the difficulty in learning complex demography in the deep past from
+reconstructed coalescence rate trajectories.
+
+We also compared the relative cross-coalescence rates (RCCR) between pairs of
+populations in the four inferred demographic models
+(Figures~\ref{fig:supp-rccr-single-origin}--\ref{fig:supp-rccr-merger-with-stem-migration})
+to those in the data (Fig.~\ref{fig:supp-rccr-data}). In the recent past
+($<$10ka), each of the models show increased RCCR compared to the data.
+However, in the medium to more distant past, the RCCR provide a good match to
+the data in each of the four models, including both their timing and ordering
+of increased cross coalescences. The failure to match both the IIRC and RCCR in
+the recent past in all models suggests that additional parameters would be
+needed to account for recent increases in population sizes. As is the case with
+other methods that reconstruct size history trajectories from inferred coalesce
+rates (e.g. PSMC, MSMC), \Relate has difficulty inferring coalescences that
+occur in the recent past, \aprcomment{likely a combination of some differences
+between actual and simulated data, and inherent weaknesses of those methods.}
 
@@
     \label{tab:supp-single-origin}
-    \textbf{Best-fit parameters from the Single-Origin model.}
-    \aprcomment{fill in caption - generation time of 29 years, other details}
+    \textbf{Best-fit parameters from the single-origin model.}
+    Inferred values are scaled to physical units assuming a generation time of
+    29 years. This model gave a log-likelihood of -189,434. \aprcomment{See
+    figure XX with all parameters labeled.}
 }
@@
     \textbf{Best-fit parameters from the Continuous-Migration model.}
-    \aprcomment{fill in caption - generation time of 29 years, other details}
+    Inferred values are scaled to physical units assuming a generation time of
+    29 years. This model gave a log-likelihood of -116,000. \aprcomment{See
+    figure XX with all parameters labeled.} 
 }
@@
     \textbf{Best-fit parameters from the Merger-Without-Stem-Migration model.}
-    \aprcomment{fill in caption - generation time of 29 years, other details}
+    Inferred values are scaled to physical units assuming a generation time of
+    29 years. This model gave a log-likelihood of -107,652. \aprcomment{See
+    figure XX with all parameters labeled.} 
 }
@@
     \textbf{Best-fit parameters from the Merger-With-Stem-Migration model.}
-    \aprcomment{fill in caption - generation time of 29 years, other details}
+    Inferred values are scaled to physical units assuming a generation time of
+    29 years. This model gave a log-likelihood of -102,633. \aprcomment{See
+    figure XX with all parameters labeled.} 
 }
@@
     \includegraphics[width=0.65\textwidth]{figures/supp-possible-models.png}
-    \caption{\textbf{Proposed models of early human history in Africa.}}
+    \caption{
+        \textbf{Proposed models of early human history in Africa.}
+    }
     \label{fig:supp-possible-models}
@@
         \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
-        \aprcomment{Plot ordering, line colors and styles, etc need to be cleaned up}
+        \aprcomment{Plot ordering, line colors and styles, etc need to be cleaned up.
+        Add panel labels.}
+        All models (A: single origin, B: continuous migration, C: merger without stem
+        migration, D: merger with stem migration) match observed $F_{ST}$ between
+        sampled populations. The continuous-migration model predicts that human stem
+        populations remain genetically similar, while the merger with and without
+        stem migration models predict a period of increased $F_{ST}$ between ancestors
+        of modern humans. This is largely due to the very small inferred $N_e$ in one of
+        the stem branches, which leads to a rapid increase in differentiation due to
+        drift.
     }
@@
         and ancient branches from the single-origin model.}
+            \aprcomment{The y label should be $f_4^2/f_2f_2$ in these figures!}
     }
@@
     }
-    \label{fig:supp-relate-sizes}
+    \label{fig:supp-iicr-data}
 \end{figure}
@@
     }
-    \label{fig:supp-relate-sizes-sim}
+    \label{fig:supp-iicr-sim}
 \end{figure}
@@
     }
-    \label{fig:supp-relate-rccr}
+    \label{fig:supp-rccr-data}
 \end{figure}
@@
     }
-    \label{fig:supp-relate-rccr-sim}
+    \label{fig:supp-rccr-single-origin}
 \end{figure}
@@
     }
-    \label{fig:supp-relate-rccr-sim}
+    \label{fig:supp-rccr-continuous-migration}
 \end{figure}
@@
     }
-    \label{fig:supp-relate-rccr-sim}
+    \label{fig:supp-rccr-merger-without-stem-migration}
 \end{figure}
@@
     }
-    \label{fig:supp-relate-rccr-sim}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    %\includegraphics{figures/supp-deep-brances-data}
+    \label{fig:supp-rccr-merger-with-stem-migration}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-deep-brances-data}
     \caption{
@@
     \centering
-    %\includegraphics{figures/supp-deep-branches-sim}
+    \includegraphics[width=\textwidth]{figures/supp-deep-branches-sim}
     \caption{
@@
     \centering
-    %\includegraphics{figures/supp-IM-misspecification}
+    \includegraphics[width=\textwidth]{figures/supp-IM-misspecification}
     \caption{

### 2022-01-06 supplement.tex
fix up fig label

@@
 \section{Gene genealogy reconstruction}
+\label{sec:relate}
 
@@
 Vindija Neanderthal \citep{Prufer2017-kk}. Again following the analysis in
-\citet{Speidel2019-nj}, if one or more mutations on a deep branch are shared
-with the Neanderthal sequence, the branch is inferred to have passed through
-the Neanderthal lineage.
+\citet{Speidel2019-nj}, if one or more mutations on a deep branch (carrying at
+least two derived mutations) are shared with the Neanderthal sequence, the
+branch is inferred to have passed through the Neanderthal lineage.
 
@@
 
-We used \Relate to reconstruct gene genealogies from the simulations under our
-four inferred demographic models. From each of these, following the same
-approach as for the analysis of the actual data, we computed coalescence rates
-within and between sampled populations. From the coalescence rates, we
-estimated the inverse instantaneous coalescence rate (IICR, often interpreted
-as a trajector of $N_e$) for each population, as well as relative cross
-coalescence rates between pairs of populations.
+We used \Relate \citep{Speidel2019-nj} to reconstruct gene genealogies from the
+simulations under our four inferred demographic models. From each of these,
+following the same approach as for the analysis of the actual data, we computed
+coalescence rates within and between sampled populations. From the coalescence
+rates, we estimated the inverse instantaneous coalescence rate (IICR, often
+interpreted as a trajector of $N_e$) for each population, as well as relative
+cross coalescence rates between pairs of populations.
 
@@
 
-\aprcomment{TODO}
+Following \citet{Speidel2019-nj}, we also characterized Neanderthal-matching
+deep branches (see Section~\ref{sec:relate}) in our simulated data under the
+four inferred demographic models. Comparing deep-branch distributions from the
+data (Fig.~\ref{fig:supp-deep-branches-data}) to the models
+(Fig.~\ref{fig:supp-deep-branches-sim}), we find consistent trends between each
+of the four models and the data in the ordering of proportions of lineages from
+each population that are labeled as Neanderthal-matching deep branches. This
+proportion is slightly over-estimated in each of the African populations
+compared to the data, but is closest in the merger-with-stem-migration model
+(which also fit the LD statistics the best of the four models). Given the
+discrepancies between the data IICR and the model IICR, these differenes are
+not surprising \aprcomment{what more to say here\dots}
 
@@
 
-\aprcomment{TODO}
+Recent studies \aprcomment{have been quite interested in characterizing the
+``earliest divergence time among human populations''}, and such analyses
+typically fit simple isolation-with-migration (IM) or ``clean split'' models to
+observed allele frequencies or inferred coalescence times among and between
+populations \citep{Bergstrom2021-iw} \aprcomment{and others}. Simple IM models
+typically fit a constant $N_e$ within each diverged population, their split
+time, and a symmetric continuous migration rate (which could be fixed to zero,
+equivalent to the ``clean split'' model). We explored the effect of fitting
+such simplified models of history to data from more complex models by
+simulating the site frequency spectrum (SFS) under each of our four models for
+pairs of populations and reinferring an IM and clean split model for each pair.
+The simulations assumed 10 diploid individuals were sampled from each
+population.
+
+As expected, the clean split models consistently provide divergence times $T$
+that are more recent than the IM models. Because $T$ and subsequent migration
+can be inversely confounded in SFS inference, fixing migration rates to zero
+results in more recent inferred $T$ (Fig.~\ref{fig:supp-IM-misspecification}).
+Even though each of our four models inferred divergences of all human
+populations $\sim$120kya, ancient structure biases inferences and gives
+inflated estimates of $T$. A similar effect occurs when the ancestral
+population size increases prior to the population split
+\citep{Momigliano2021-th}, which is the case in our inferred single-origin
+model. \aprcomment{I want to highlight this more in the main text.} Inferred
+$T$ between the Nama and other populations were $>$200kya under the IM model,
+and $\sim$50--130kya under the clean split model.
+
+\aprcomment{There's plenty more we could say about this.}
 
@@
         and ancient branches from the single-origin model.}
-            \aprcomment{The y label should be $f_4^2/f_2f_2$ in these figures!}
     }

### 2022-01-06 supplement.tex
edits to relate discussion

@@
 rates, we estimated the inverse instantaneous coalescence rate (IICR, often
-interpreted as a trajector of $N_e$) for each population, as well as relative
+interpreted as an estimator of $N_e$) for each population, as well as relative
 cross coalescence rates between pairs of populations.
 
-We note that the IICR is a function of not only population sizes, but also of
-population structure and migration, and interpreting it as an actual population
-size history is tenuous.  However, it is still a statistic that can be computed
-from the data and compared to predictions under the model, and can therefore be
+The IICR is only a reliable estimator of $N_e$ in the absence of population structure.
+.
+However, since IIRC are readily estimated and visually interesting, it is commonly attempted even when the assumption is not met \cite{}.
+Even if the assumption is not met, the inferred IIRC can be interpreted as a summary statistic for which model predictions can be compared to data. and can therefore be
 informative about aspects of the data that our models fail to predict.
-\aprcomment{I think we want to say something along these lines above\dots} We
+\sgcomment{Insert inferences with same populations?}. We
 also note that the coalescence rates inferred from data used a larger set of

### 2022-01-06 supplement.tex
edits to fst discussion

@@
 
-\subsection{$F_{ST}$ between coexisting populations over time}
-
-$F_{ST}$ can be interpreted as a measure of population structure, given by the
+\subsection{Divergence between coexisting populations over time}
+
+We considered a few different measures of differentiation between populations  in four inferred models, starting with $F_{ST}.$ For each of the four
+inferred demographic models presented here, we computed predicted $F_{ST}$ for
+pairs of contemporaneous populations over time.
+
+$F_{ST}$ represents the
 proportion of genetic variance in a multi-population sample that can be
-attributed to drift between populations in that sample \cite{Bhatia2013-sq}.
-$F_{ST}$ depends on population divergence times as well as subsequent gene flow
-and population sizes: migration between populations reduces $F_{ST}$ while
-small effective population sizes can lead to large values. For each of the four
-inferred demographic models presented here, we computed predicted $F_{ST}$ for
-pairs of contemporaneous populations over time. For sampled populations,
-predicted $F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}). We
+attributed to drift between populations in that sample (e.g., \cite{Bhatia2013-sq}).
+$F_{ST}$ depends on population divergence times, gene flow,
+and population sizes. For present-day sampled populations,
+predicted $F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}), while past estimates represent model predictions about the differentiation between past populations. We
 computed $F_{ST}$ as a ratio of expected (or average) variances, using observed
-pairwise diversity within and between populations as
+pairwise diversity within and between populations as \sgcomment{WHich estimator is that? If this is the one recommended in Bhatia, cite?}
 \begin{equation}
@@
 \end{equation}
-
+\sgcomment{where $H_{ij}$ is the average number of differences between one haploid copy of the genome from population $i$ and one from population $j$.}
 In all inferred models, $F_{ST}$s between Neanderthals and the ancestors of
@@
 populations or their ancestors (stems). $F_{ST}$ between modern human
-populations over the last 100 thousand years remained relatively low, at
+populations over the last $100ka$ remained relatively low, at
 $\sim0.1$ or less. In the continuous-migration model, the two stem populations
@@
 stem populations was inferred to have a short, severe bottleneck, which led to
-increased $F_{ST}$ between stem populations during that time. However, this
-period of relatively large $F_{ST}$ did not result in large $F_{ST}$ between
-more recent pairs of populations due to their roughly similar levels of
-contributions to the ancestors of modern humans. \aprcomment{We caution that
-the severe bottleneck may be an artifact of the model parameterization?}
+increased $F_{ST}$ between stem populations during that time.
+In the merger with stem migration, $F_{ST}$ reached a maximum value of $0.3$ between stems, intermediate between human-Neanderthal divergence ($0.54$ \sgcomment{check}) and the maximum $F_{ST}$ observed between contemporary human populations ($0.07$.)
+
+Whereas the high $F_{ST}$ value between Neanderthal and humans reflects both elevated genetic differences between the genomes of modern humans and neanderthal and reduced diversity within the Neanderthal population (\sgcomment{Figure X}), the increased $F_{ST}$ between the stems is primarily due to an inferred bottleneck reducing stem diversity, while genetic differentiation between individuals from the two stems is within the range of genetic differentiation between (and even within) human populations.
+In other words, our best fit model describes two stems with one stems showing reduced diversity, but individuals across stems showing no more differentiation than two modern population.
+\sgcomment{Previous paragraph could use rereading.}
 

### 2022-01-06 supplement.tex
update figures

--- 2022-01-06 supplement.tex
@@
     \centering
+    \includegraphics[width=\textwidth]{figures/supp-relate-iicr-data-full-and-subset}
+    \caption{
+        \textbf{Inverse instantaneous coalescence rates inferred from the data,
+        subset to focal populations in the model inferences.}
+    }
+    \label{fig:supp-iicr-data-subset}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
     \includegraphics[width=\textwidth]{figures/supp-relate-iicr-simulation}
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/supp-relate-rccr-data}
+    \includegraphics[width=\textwidth]{figures/supp-relate-rccr-data-full-and-subset}
     \caption{

### 2022-01-07 paper.tex
add some figures

@@
 each specified model
-(Tables~\ref{tab:supp-single-origin-fits}--\ref{tab:supp-merger-with-stem-migration-fits}),
+(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
 reflecting an informative statistical approach. As expected, model assumptions

### 2022-01-07 supplement.tex
add some figures

@@
     \centering
+    \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
+    \caption{
+        \textbf{Predicted pairwise differentiation between contemporaneous populations.}
+        \aprcomment{Plot ordering, line colors and styles, etc need to be cleaned up.
+        Add panel labels.}
+        All models (A: single origin, B: continuous migration, C: merger without stem
+        migration, D: merger with stem migration)\dots
+    }
+    \label{fig:supp-h12-panels}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
     \includegraphics{figures/supp-f4s-single-origin.pdf}
@@
     \centering
+    \includegraphics[width=\textwidth]{figures/supp-exp-coal-rates}
+    \caption{
+        \textbf{Expected inverse instantaneous coalescence rates from models.}
+    }
+    \label{fig:supp-iicr-exp}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
     \includegraphics[width=\textwidth]{figures/supp-relate-rccr-data-full-and-subset}
@@
     \centering
+    \includegraphics[width=0.75\textwidth]{figures/supp-relate-rccr-model-exp.pdf}
+    \caption{
+        \textbf{Expected relative cross-coalescence rates from inferred models.}
+    }
+    \label{fig:supp-rccr-exp}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
     \includegraphics{figures/supp-deep-brances-data}

### 2022-01-07 supplement.tex
some captions

@@
 \subsection{$f_4$ statistics between pairs of contemporary and pairs of ancient populations}
+\label{sec:f4}
 
@@
         Add panel labels.}
-        All models (A: single origin, B: continuous migration, C: merger without stem
-        migration, D: merger with stem migration)\dots
+        $H_{i,j}$ is the predicted pairwise differentiaion between two populations,
+        which we normalize by the observed pairwise diversity in the Mende sample.
+        The single-origin model (A) shows larger deviations between observed and
+        expected differentiation than the models with early human structure
+        (B: continuousmigration, C: merger without stem migration, D: merger with 
+        stem migration). In the merger models (C and D), while $F_{ST}$ is large
+        between early stems due to the sharp bottleneck in one of those branches,
+        the pairwise differences are comparable to the continuous-migration model.
+        Stem 1E and Stem 1S have low differentiation, as the bottleneck in Stem 1
+        (which they both split from) sharply reduces diversity by the time of the
+        split. This has the effect of large $F_{ST}$ but small $H_{i,j}$.
     }
@@
         and ancient branches from the single-origin model.}
+        $f_4(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
+        amount of drift between sampled populations $A$ and $B$ that can be
+        attributed to drift between populations $X$ and $Y$, sampled in the past
+        (see Section~\ref{sec:f4}). \aprcomment{\dots}
     }

### 2022-01-07 paper.tex
abstract cleanup

@@
 Africa.
-\sgcomment{IF we were not afraid of the racists, we could write something like:
-"Despite the genetic similarity between these populations, 4 to 10\% of random
-genetic differences among [modern human populations] can be attributed to this
-early genetic drift.}
 In contrast to models with archaic introgression, we predict that fossil
 remains from coexisting ancestral populations should be morphologically
-similar. We show that model misspecification explains variation in previous
+similar. 
+Despite genetic similarity between these populations, an inferred \sgcomment{4 to 10\%} of random genetic differences among [modern human populations] can be attributed to genetic drift between stems.
+We show that model misspecification explains variation in previous
 divergence time estimates and argue that studying a suite of models is key to

### 2022-01-07 paper.tex
suggestion

@@
 
-Regardless of the model choice for early epochs, inference of recent human
-demographic history from two thousand \aprcomment{why two thousand, instead of
-just ``up to 150kya''?} to 150kya was remarkably robust. The earliest divergence
+Regardless of the model choice for early epochs, inference of human
+demographic history in the last 150kya was remarkably robust. The earliest divergence
 among contemporary sampled human populations differentiates the southern
@@
 
-Likelihood-based parameter confidence intervals are relatively narrow within
+Likelihood-based parameter confidence intervals  \sgcomment{or: Bootstrap-based confidence interval reflecting the finite length of the genome?}   are relatively narrow within
 each specified model

### 2022-01-07 paper.tex
added (a version of) table 1

@@
         \toprule
-        Parameter & Value & Something else \\
+        Parameter & Value & Value \\
         \midrule
@@
     \label{tab:migration-rates}
+\end{table}
+
+%Please add the following packages if necessary:
+%\usepackage{booktabs, multirow} % for borders and merged ranges
+%\usepackage{soul}% for underlines
+%\usepackage[table]{xcolor} % for cell colors
+%\usepackage{changepage,threeparttable} % for wide tables
+%If the table is too wide, replace \begin{table}[!htp]...\end{table} with
+%\begin{adjustwidth}{-2.5 cm}{-2.5 cm}\centering\begin{threeparttable}[!htb]...\end{threeparttable}\end{adjustwidth}
+\begin{table}[!htp]\centering
+
+\scriptsize
+\begin{tabular}{lrrrrrr}\toprule
+\textbf{Likelihood} &\textbf{Label} &\textbf{Population Pair} &\textbf{Divergence}&\textbf{Migration rate} &\textbf{Migration duration } \\
+& & & \textbf{Time (kya)} & \textbf{per generation} &\textbf{ (kya)}\\\midrule
+\textbf{Continuous Model} & & & & & \\
+LL= -116000 &missing &Stem 1 vs. Stem 2 &1,163 &6.43e-5 &1,028 \\
+&c, add copies &Nama vs. Other Africans* &135 &4.1E-5, 9.8E-6 &135, 60 \\
+& not shown &Mende vs. East Africans &60 &\textbf{2.14e-4} &55 \\
+& not shown &East Africans vs. British &50 &4.17e-5 &50 \\
+& labeled &Gumuz vs. Amhara/Oromo &12 &\textbf{3.36e-4} &12 \\
+& green &Stem 2 vs. Mende &NA &\textbf{1.64e-4} &55 \\
+& orange &Stem 2 vs. Nama &NA &5.82e-5 &130 \\
+\textbf{Merger Model} & & & & & \\
+LL= -102633 & &Stem 1 vs. Stem 2 &1,442 &\textbf{1.16e-4} &963 \\
+& &Stem 1S vs. Stem 1E &479 &0.00e+0 &NA \\
+& &Nama vs. Other Africans &119 &4.4E-5, 7.1E-6 &117 \\
+& &Stem 2 vs. Stem 1E &98 &\textbf{5.00e-1} &pulse \\
+& &Mende vs. East Africans &60 &\textbf{1.98e-4} &38 \\
+& &East Africans vs. British &50 &3.87e-5 &50 \\
+& &Gumuz vs. Amhara/Oromo &12 &\textbf{3.59e-4} &12 \\
+& &Stem 2 vs. Mende &NA &\textbf{0.18} &pulse \\
+\bottomrule
+\end{tabular}
+\caption{Table 1 copied from dropbox}
 \end{table}

### 2022-01-07 paper.tex
minor changes

--- 2022-01-07 paper.tex
@@
 Africa.
+\sgcomment{IF we were not afraid of the racists, we could write something like:
+"Despite the genetic similarity between these populations, 4 to 10\% of random
+genetic differences among [modern human populations] can be attributed to this
+early genetic drift.}
 In contrast to models with archaic introgression, we predict that fossil
 remains from coexisting ancestral populations should be morphologically
-similar. 
-Despite genetic similarity between these populations, an inferred \sgcomment{4 to 10\%} of random genetic differences among [modern human populations] can be attributed to genetic drift between stems.
-We show that model misspecification explains variation in previous
+similar. We show that model misspecification explains variation in previous
 divergence time estimates and argue that studying a suite of models is key to
@@
 The “Multiple Merger” model has better log-likelihood than the “Continuous
-Migration” model (-102,600 vs. -116,000, respectively). The two models differ
+Migration” model (-102,600 vs. -115,500, respectively). The two models differ
 in the timing of divergence and the relative Ne of the two stems

### 2022-01-07 supplement.tex
minor changes

@@
         $f_4(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
-        amount of drift between sampled populations $A$ and $B$ that can be
-        attributed to drift between populations $X$ and $Y$, sampled in the past
-        (see Section~\ref{sec:f4}). \aprcomment{\dots}
+        amount of drift between sampled populations $A$ and $B$ that aligns with
+        drift between populations $X$ and $Y$, sampled in the past
+        (see Section~\ref{sec:f4}). In the single-origin model, nearly all
+        differentiation between the Nama, Mende, and Gumuz populations arises
+        after the divergence events, so normalized $f_4$ decays to zero by the
+        time of those divergences.
     }
@@
         and ancient branches from the continuous-migration model.}
+        Despite population structure inferred to have extended up to 1Ma
+        into the past, differentiation between modern African populations
+        only traces back to differentiation between ancestral populations
+        $\sim$100-200ka, indicating that while stem populations were structured
+        (Fig.~\ref{fig:supp-FST-panels}), differentiation due to that
+        ancient structure does not contribute to differention between modern
+        populations.
     }
@@
         and ancient branches from the merger-without-stem-migration model.}
+        Unlike the continuous-migration model, 
     }

### 2022-01-07 supplement.tex
some clean up

@@
 
-We considered a few different measures of differentiation between populations  in four inferred models, starting with $F_{ST}.$ For each of the four
-inferred demographic models presented here, we computed predicted $F_{ST}$ for
-pairs of contemporaneous populations over time.
-
-$F_{ST}$ represents the
-proportion of genetic variance in a multi-population sample that can be
-attributed to drift between populations in that sample (e.g., \cite{Bhatia2013-sq}).
-$F_{ST}$ depends on population divergence times, gene flow,
-and population sizes. For present-day sampled populations,
-predicted $F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}), while past estimates represent model predictions about the differentiation between past populations. We
-computed $F_{ST}$ as a ratio of expected (or average) variances, using observed
-pairwise diversity within and between populations as \sgcomment{WHich estimator is that? If this is the one recommended in Bhatia, cite?}
+We considered a few different measures of differentiation between populations
+in four inferred models, starting with $F_{ST}.$ For each of the four inferred
+demographic models presented here, we computed predicted $F_{ST}$ for pairs of
+contemporaneous populations over time.
+
+$F_{ST}$ represents the proportion of genetic variance in a multi-population
+sample that can be attributed to drift between populations in that sample
+(e.g., \cite{Bhatia2013-sq}).  $F_{ST}$ depends on population divergence times,
+gene flow, and population sizes. For present-day sampled populations, predicted
+$F_{ST}$ can be compared to data (Fig.~\ref{fig:supp-FST-panels}), while past
+estimates represent model predictions about the differentiation between past
+populations. We computed $F_{ST}$ as a ratio of expected (or average)
+variances, using observed pairwise diversity within and between populations as
+in \citet{Peter2016-ys}:
 \begin{equation}
-    F_{ST}(i, j) = \frac{\E[2H_{i,j} - H_i - H_j]}{\E[2H_{i,j} + H_i + H_j]}.
+    F_{ST}(i, j) = \frac{\E[2H_{i,j} - H_i - H_j]}{\E[2H_{i,j} + H_i + H_j]},
     \label{eq:FST}
 \end{equation}
-\sgcomment{where $H_{ij}$ is the average number of differences between one haploid copy of the genome from population $i$ and one from population $j$.}
-In all inferred models, $F_{ST}$s between Neanderthals and the ancestors of
-modern humans were much larger than between any pair of modern human
-populations or their ancestors (stems). $F_{ST}$ between modern human
-populations over the last $100ka$ remained relatively low, at
-$\sim0.1$ or less. In the continuous-migration model, the two stem populations
-also were predicted to have low $F_{ST}$ within this range, as migration rates
-were inferred to be high enough to maintain genetic similarity. In both the
+
+where $H_{ij}$ is the average number of differences between one haploid copy of
+the genome from population $i$ and one from population $j$, and $H_i$ and $H_j$
+are the average number of differences for two haploid copies from the same
+population. In all inferred models, $F_{ST}$s between Neanderthals and the
+ancestors of modern humans were much larger than between any pair of modern
+human populations or their ancestors (stems). $F_{ST}$ between modern human
+populations over the last $100ka$ remained relatively low, at $\sim0.1$ or
+less. In the continuous-migration model, the two stem populations also were
+predicted to have low $F_{ST}$ within this range, as migration rates were
+inferred to be high enough to maintain genetic similarity. In both the
 merger-without-stem-migration and the merger-with-stem-migration, one of the
 stem populations was inferred to have a short, severe bottleneck, which led to
-increased $F_{ST}$ between stem populations during that time.
-In the merger with stem migration, $F_{ST}$ reached a maximum value of $0.3$ between stems, intermediate between human-Neanderthal divergence ($0.54$ \sgcomment{check}) and the maximum $F_{ST}$ observed between contemporary human populations ($0.07$.)
-
-Whereas the high $F_{ST}$ value between Neanderthal and humans reflects both elevated genetic differences between the genomes of modern humans and neanderthal and reduced diversity within the Neanderthal population (\sgcomment{Figure X}), the increased $F_{ST}$ between the stems is primarily due to an inferred bottleneck reducing stem diversity, while genetic differentiation between individuals from the two stems is within the range of genetic differentiation between (and even within) human populations.
-In other words, our best fit model describes two stems with one stems showing reduced diversity, but individuals across stems showing no more differentiation than two modern population.
-\sgcomment{Previous paragraph could use rereading.}
-
-\subsection{$f_4$ statistics between pairs of contemporary and pairs of ancient populations}
+increased $F_{ST}$ between stem populations during that time. In the merger
+with stem migration, $F_{ST}$ reached a maximum value of $0.3$ between stems,
+intermediate between human-Neanderthal divergence ($0.54$ \sgcomment{check})
+and the maximum $F_{ST}$ observed between contemporary human populations
+($0.07$.)
+
+Whereas the high $F_{ST}$ value between Neanderthal and humans reflects both
+elevated genetic differences between the genomes of modern humans and
+neanderthal and reduced diversity within the Neanderthal population
+(Fig.~\ref{fig:supp-FST-panels}), the increased $F_{ST}$ between the stems is
+primarily due to an inferred bottleneck reducing stem diversity, while genetic
+differentiation between individuals from the two stems is within the range of
+genetic differentiation between (and even within) human populations. In other
+words, our best fit model describes two stems with one stems showing reduced
+diversity, but individuals across stems showing no more differentiation than
+two modern population. \sgcomment{Previous paragraph could use rereading.}
+
+\subsection{\bm{$f_4$} statistics between pairs of contemporary
+and pairs of ancient populations}
 \label{sec:f4}
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -116,000. \aprcomment{See
+    29 years. This model gave a log-likelihood of -115,500. \aprcomment{See
     figure XX with all parameters labeled.} 

### 2022-01-09 paper.tex
add fig 2 and 3 drafts

--- 2022-01-07 paper.tex
@@
 
-\begin{figure}[ht]
+\begin{figure}[ht!]
 \begin{center}
@@
 
-\begin{figure}[ht]
-\begin{center}
-\makebox[\textwidth][c]{} % \includegraphics{figures/fig2.png}}
-\caption{\textbf{}.
-    A placeholder - best fit model(s).
+\begin{figure}[t!]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figures/fig2}
+    \caption{
+        \textbf{Two best fit models with a weakly structured stem.}
+        In both the the parameterizations of early population structure,
+        continuous migration (A) and fragmentation with mergers (B), models
+        that include ongoing migration between stem populations outperform
+        those in which stem populations are isolated. Most populations are
+        connected by continuous, reciprocal migration that are incidated by 
+        double-headed arrows (labels matched to migration rates and divergence
+        times in Table~\ref{tab:migration-rates}), and which lasts for the
+        duration of concurrence of contemporaneous populations. The
+        merger-with-stem-migration model (B) outperformed the
+        continuous-migration model (A), $LL=-102,600$ vs $-115,500$,
+        respectively, and both models provide a much better fit to diversity
+        and LD statistics than a single-origin model of recent expansion.
     }
     \label{fig:2}
-\end{center}
 \end{figure}
@@
 2). \aprcomment{update this last sentence}
+
+\begin{table}[t!]
+    \centering
+    \scriptsize
+    \begin{tabular}{lllrrrr}
+        \toprule
+        \textbf{Likelihood} & \textbf{Label} & \textbf{Population Pair} &
+            \textbf{Divergence} & \textbf{Migration rate} & \textbf{Migration} \\
+        & & & 
+            \textbf{Time (kya)} & \textbf{per generation} & \textbf{duration (ky)} \\
+        \midrule
+        \textbf{Continuous Model} & & & & & \\
+        LL= -115,500 & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
+        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
+        & c, d & Stem 2, Other Africans & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
+        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
+        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
+        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
+        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
+        \textbf{Merger Model} & & & & & \\
+        LL= -102,600 & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
+        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
+        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
+        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
+        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
+        & e, f & Nama, Other Africans & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
+        & g, & Mende, East Africans & 60 & \textbf{1.98e-4} & 60 \\
+        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
+        & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
+        \bottomrule
+    \end{tabular}
+    \caption{
+        \textbf{Migration and divergence parameters from best fit models.}\\
+        \aprcomment{Labels are keyed to Figure~\ref{fig:2}}
+    }
+    \label{tab:migration-rates}
+\end{table}
 
@@
 
-\begin{table}
-    \centering
-    \begin{tabular}{l|l|l}
-        \toprule
-        Parameter & Value & Something else \\
-        \midrule
-        x & y & z \\
-        \bottomrule
-    \end{tabular}
-    \caption{caption}
-    \label{tab:migration-rates}
-\end{table}
-
 \subsection*{Reconciling multiple lines of genetic evidence}
@@
 \citet{Durvasula2020-td}).
+
+\begin{figure}[t!]
+    \centering
+    \includegraphics{figures/fig3.pdf}
+    \caption{
+        \textbf{fig3}
+    }
+    \label{fig:3}
+\end{figure}
 
@@
 should be interpreted with caution.
-
-\begin{figure}[ht]
-\begin{center}
-    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
-\caption{\textbf{The third main figure}.
-    A placeholder - validation (Relate and cSFS).
-}
-\label{fig:3}
-\end{center}
-\end{figure}
 

### 2022-01-09 paper.tex
edits to figures and results

@@
 data for four diverse African populations, comprising the Nama (Khoe-San from
-South Africa, newly presented here), Mende (MSL from Sierre Leone, Phase 3 of
-the 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}), Gumuz
-(recent descendents of a hunter-gatherer group from Ethiopia
+South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
+Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
+Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
 \citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
@@
 models as a representative source of back-to-Africa gene flow and recent
-colonial admixture in South Africa. Finally, we added a high-coverage ancient
+colonial admixture in South Africa. Finally, we used a high-coverage ancient
 Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
@@
 (Figure~\ref{fig:supp-possible-models}A), without allowing for contribution
-from an archaic African lineage. As expected \citep{Ragsdale2019-nt}, this
-first model was a poor fit to the data qualitatively
-(Figure~\ref{fig:supp-single-origin-fits}) and quantitatively (log-likelihood
-($LL$) $\approx -189,000$, Table~\ref{tab:supp-single-origin}). We next explored a
-suite of models in which population structure predates the differentiation of
-contemporary groups, including models allowing for ancestral reticulation
+from an archaic African lineage or population structure prior to the expansion.
+As expected \citep{Ragsdale2019-nt}, this first model was a poor fit to the
+data qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and
+quantitatively (log-likelihood ($LL$) $\approx -189,400$,
+Table~\ref{tab:supp-single-origin}). We next explored a suite of models in
+which population structure predates the differentiation of contemporary groups,
+including models allowing for ancestral reticulation
 (Figure~\ref{fig:supp-possible-models}B), archaic admixture
@@
 African Nama from other African groups between 110--135kya, with low to moderate
-levels of subsequent gene flow \aprcomment{(Table 1)}. In none of the
+levels of subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
 high-likelihood models which we explored did the divergence between Nama and
-other populations exceed $\sim$150kya. We conclude that, regardless of model
+other populations exceed $\sim$140kya. We conclude that, regardless of model
 specification, geographic patterns of \emph{Homo sapiens} population structure
@@
 stem groups as only a small proportion of drift between modern populations can
-be attributed to drift between stems \aprcomment{(Figure 3)}.
+be attributed to drift between stems (Figures~\ref{fig:3} and
+\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
 
@@
 Methods). The time of divergence of Western and Eastern African populations was
-set to 60kya, just prior to the split of Eurasians and East Africans 50kya. Both
+set to 60kya, just prior to the split of Eurasians and East Africans 50kya. All
 models predict relatively high subsequent gene flow between the two regions
-($m=2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
+($m\approx2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
 following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
@@
 contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
-was estimated at 15\%, similar to proportions inferred by ADMIXTURE
-\aprcomment{cite} (Figure~\ref{fig:1}).
+was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE}
+\citep{Alexander2009-sw} (Figure~\ref{fig:1}).
 
@@
 contemporary populations), and the other stem population(s) has continuous
-symmetric migration with those populations; or 2) expansion of one or more of
-the stem populations, with an instantaneous pulse (or “merger”) from the other
-stem population. These models are meant to capture scenarios 2B and 2C (Figure
-2). \aprcomment{update this last sentence}
+symmetric migration with those populations; or 2) one or more of the stem
+populations expands, with instantaneous pulse (or “merger”) events from the
+other stem population, so that diversifying branches are formed through the
+coalescence of multiple ancestral populations. Depending on parameter values,
+this scenario encompasses archaic introgression and
+fragmentation-and-coalescence models.
 
@@
         & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
-        & c, d & Stem 2, Other Africans & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
+        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
         & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
@@
         & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
-        & e, f & Nama, Other Africans & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
-        & g, & Mende, East Africans & 60 & \textbf{1.98e-4} & 60 \\
+        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
+        & g, & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
         & h & East Africans, British & 50 & 3.87e-5 & 50 \\
@@
     \caption{
-        \textbf{Migration and divergence parameters from best fit models.}\\
-        \aprcomment{Labels are keyed to Figure~\ref{fig:2}}
+        \textbf{Migration and divergence parameters from best fit models.}
+        Labeled migration rates correspond to symmetric continuous migration
+        bands shown in Figure~\ref{fig:2}. Both the continuous-migration and
+        merger models \ldots
+        \aprcomment{also define *}
     }
@@
 improved the fits relative to zero migration (e.g. $LL \approx -102,600$ vs.
-$-107,700$). With continuous migration, population structure extends back to
-1.1--1.4Mya \aprcomment{(Table 1)}. Migration between the stems in these models
-is low but non-negligible, with a fraction of migrant lineages each generation
-estimated as $m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is
-at least as large as migration inferred between the Ethiopian East Africans and
-British populations over the past 50kya \aprcomment{(Table 1)}. This ongoing
-(or, at least, periodic) gene flow qualitatively distinguishes these models
-from previously proposed archaic admixture models \aprcomment{(Figure S1C)} as
-the early branches remain closely related throughout (Figure 3).
-\aprcomment{this last phrase may need to be updated, with regard to Fig
-\ref{fig:supp-FST-panels}}
+$-107,700$ in the merger scenario and $LL \approx -115,500$ vs. $-128,000$ with
+continuous migration). With continuous migration between stems, population
+structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
+Migration between the stems in these models is moderate, with a fraction of
+migrant lineages each generation estimated as
+$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this roughly compares
+to inferred migration rates between connected contemporary populations over the
+past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or, at least,
+periodic) gene flow qualitatively distinguishes these models from previously
+proposed archaic admixture models (Figure~\ref{fig:supp-proposed-models}C) as
+the early branches remain closely related (Figure 3). \aprcomment{this last
+phrase may need to be updated, with regard to Fig \ref{fig:supp-FST-panels}}
 
-Likelihood-based parameter confidence intervals  \sgcomment{or: Bootstrap-based confidence interval reflecting the finite length of the genome?}   are relatively narrow within
-each specified model
+Bootstrap-based confidence intervals are relatively narrow within each
+specified model
 (Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
@@
 have a larger impact on parameter estimates. To convey model uncertainty, we
-highlight features of two models with the highest \aprcomment{not the
-highest\ldots merger-without-stem-migration at -107,650}log likelihoods. These
-differ primarily in whether early stems are assumed to give rise to modern
-populations in multiple discrete merger events or in continuous migrations
-\aprcomment{(Figure 2)}. Additional models are discussed in
-\aprcomment{Supplement XXX}.
+highlight features of two models with high likelihoods. These differ primarily
+in whether early stems are assumed to give rise to modern populations in
+multiple discrete merger events or in continuous migrations
+(Figure~\ref{fig:2}). Additional models are discussed in the Supporting
+Material.
 
@@
 level of gene flow as between the Mende and Eastern Africans (Table 1).
-\aprcomment{paragraphs above need work, and condensing}
-
-\aprcomment{also below, too much information - we don't actually believe these
-models}
 
@@
     \centering
-    \includegraphics{figures/fig3.pdf}
+    %\includegraphics{figures/fig3.pdf}
     \caption{
@@
 
-\begin{figure}[ht]
-\begin{center}
-    \makebox[\textwidth][c]{} % \includegraphics{figures/fig3.pdf}}
-\caption{\textbf{The fourth main figure}.
-    A placeholder - predictions (FST and/or f4).
-}
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{figures/fig4}
+    \caption{fig4}
     \label{fig:4}
-\end{center}
 \end{figure}

### 2022-01-09 supplement.tex
edits to figures and results

@@
         \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
-        \aprcomment{Plot ordering, line colors and styles, etc need to be cleaned up.
-        Add panel labels.}
         All models (A: single origin, B: continuous migration, C: merger without stem
@@
         \textbf{Predicted pairwise differentiation between contemporaneous populations.}
-        \aprcomment{Plot ordering, line colors and styles, etc need to be cleaned up.
-        Add panel labels.}
         $H_{i,j}$ is the predicted pairwise differentiaion between two populations,

### 2022-01-09 paper.tex
edits to results

@@
 \begin{figure}[ht!]
-\begin{center}
-\makebox[\textwidth][c]{\includegraphics{figures/fig1.pdf}}
-\caption{\textbf{Geography and genetic diversity within Africa}.
-    \aprcomment{In the models presented below, we focus on the Nama, 
-    Mende (MSL), Gumuz, and Oromo and Amhara.}
+    \centering
+    \includegraphics{figures/fig1.pdf}
+    \caption{
+        \textbf{Geography and genetic diversity within Africa.}
+        The combined 1000 Genomes and AGRP datasets incude populations from
+        West, East, and South Africa. Both PCA (B, C) and \texttt{ADMIXTURE}
+        (D) illustrate the well-characterized high levels of genetic diversity
+        and signatures of recent migration and gene flow within Africa.
+        We built parameterized models using the newly-sequenced Nama, Mende
+        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
+        British and Vindija Neanderhtal individual (bold fonts).
+        \aprcomment{work on this caption}
     }
     \label{fig:1}
-\end{center}
 \end{figure}
@@
 
-We began with a model of geographic expansion from a single ancestral source
-followed by migration between populations
-(Figure~\ref{fig:supp-possible-models}A), without allowing for contribution
-from an archaic African lineage or population structure prior to the expansion.
-As expected \citep{Ragsdale2019-nt}, this first model was a poor fit to the
-data qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and
-quantitatively (log-likelihood ($LL$) $\approx -189,400$,
-Table~\ref{tab:supp-single-origin}). We next explored a suite of models in
-which population structure predates the differentiation of contemporary groups,
-including models allowing for ancestral reticulation
-(Figure~\ref{fig:supp-possible-models}B), archaic admixture
+We began with a model of geographic expansion from a single ancestral,
+unstructured source followed by migration between populations, without allowing
+for contribution from an archaic African lineage or population structure prior
+to the expansion (Figure~\ref{fig:supp-possible-models}A). As expected
+\citep{Ragsdale2019-nt}, this first model was a poor fit to the data
+qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
+(log-likelihood ($LL$) $\approx -189,400$, Table~\ref{tab:supp-single-origin}).
+We next explored a suite of models in which population structure predates the
+differentiation of contemporary groups, including models allowing for ancestral
+reticulation (Figure~\ref{fig:supp-possible-models}B), archaic admixture
 (Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
@@
 
-Regardless of the model choice for early epochs, inference of human
-demographic history in the last 150kya was remarkably robust. The earliest divergence
-among contemporary sampled human populations differentiates the southern
-African Nama from other African groups between 110--135kya, with low to moderate
-levels of subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
+Regardless of the model choice for early epochs, inference of human demographic
+history in the last 150kya was remarkably robust. The earliest divergence among
+contemporary human populations differentiates the southern African Nama from
+other African groups between 110--135kya, with low to moderate levels of
+subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
 high-likelihood models which we explored did the divergence between Nama and
@@
 
-\subsection*{Deep population structure but not archaic admixture within Africa}
-
-To account for population structure prior to 135kya, we allowed for two or more
-“stem” populations which could diverge either before or after the Neanderthal
-split. We considered models both with and without migration between these stem
-populations, and in both cases we tested two different types of gene exchange
-during the expansion phase. 1) One of the stem population expands (splits into
-contemporary populations), and the other stem population(s) has continuous
-symmetric migration with those populations; or 2) one or more of the stem
-populations expands, with instantaneous pulse (or “merger”) events from the
-other stem population, so that diversifying branches are formed through the
-coalescence of multiple ancestral populations. Depending on parameter values,
-this scenario encompasses archaic introgression and
-fragmentation-and-coalescence models.
-
 \begin{table}[t!]
@@
         bands shown in Figure~\ref{fig:2}. Both the continuous-migration and
-        merger models \ldots
-        \aprcomment{also define *}
+        merger models inferred a relatively deep split of human stem branches,
+        though they were connected by ongoing migration that maintained their
+        genetic similarity.
+        *: The ancestors of the Nama diverged from other African groups
+        $\sim$120--135kya, which later branch into West and East African
+        groups 60kya. Migration rates and durations are shown for 1) East Africans
+        and their ancestors, and 2) the Mende, respectively.
     }
@@
 
+\subsection*{Deep but connected population structure within Africa}
+
+To account for population structure prior to 135kya, we allowed for two or more
+“stem” populations which could diverge either before or after the Neanderthal
+split. We considered models both with and without migration between these stem
+populations, and in both cases we tested two different types of gene exchange
+during the expansion phase. 1) One of the stem population expands (splits into
+contemporary populations), and the other stem population(s) has continuous
+symmetric migration with those populations; or 2) one or more of the stem
+populations expands, with instantaneous pulse (or “merger”) events from the
+other stem population, so that diversifying branches are formed through the
+coalescence of multiple ancestral populations. Depending on parameter values,
+this scenario encompasses archaic introgression and
+fragmentation-and-coalescence models.
+
 Allowing for continuous migration between the stem populations substantially
-improved the fits relative to zero migration (e.g. $LL \approx -102,600$ vs.
+improved the fits relative to zero migration ($LL \approx -102,600$ vs.
 $-107,700$ in the merger scenario and $LL \approx -115,500$ vs. $-128,000$ with
@@
 to inferred migration rates between connected contemporary populations over the
-past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or, at least,
+past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
 periodic) gene flow qualitatively distinguishes these models from previously
-proposed archaic admixture models (Figure~\ref{fig:supp-proposed-models}C) as
-the early branches remain closely related (Figure 3). \aprcomment{this last
-phrase may need to be updated, with regard to Fig \ref{fig:supp-FST-panels}}
+proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
+the early branches remain closely related (Figure 3).
 
@@
 reflecting an informative statistical approach. As expected, model assumptions
-have a larger impact on parameter estimates. To convey model uncertainty, we
-highlight features of two models with high likelihoods. These differ primarily
-in whether early stems are assumed to give rise to modern populations in
-multiple discrete merger events or in continuous migrations
-(Figure~\ref{fig:2}). Additional models are discussed in the Supporting
-Material.
-
-The “Multiple Merger” model has better log-likelihood than the “Continuous
-Migration” model (-102,600 vs. -115,500, respectively). The two models differ
-in the timing of divergence and the relative Ne of the two stems
+have a larger impact on parameter estimates. To convey model uncertainty we
+highlight features of two inferred models with high likelihoods, the
+“multiple-merger” and the the “continuous-migration” model that both allow
+migration between stem branches. These differ primarily in the timing of the
+early divergence of stem populations and their relative $N_e$
 (Figure~\ref{fig:2}). The two models also differ in the mode of divergence
@@
 
-Under the continuous-migration model, the two stems give rise to contemporary
-populations in western, southern and eastern Africa but with differing
-proportions of ancestry derived from “Stem 1” and “Stem 2” (Figure 2). For
-example, the Nama and Gumuz have lower levels of migration with Stem 2
-($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), relative to the
-Mende. The Mende continue to exchange migrants with Stem 2
-($m=1.6\times10^{-4}$) long after their separation 60kya; this is a similar
-level of gene flow as between the Mende and Eastern Africans (Table 1).
+Under the continuous-migration model, one of the two stems diverges into
+lineages leading to contemporary populations in western, southern and eastern
+Africa, and the other (Stem 2) contributes variable ancestry to those
+populations. This migration from Stem 2 is highest with the Mende
+($m=1.6\times10^{-4}$) compared to the Nama and East African populations
+($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
+allowed to occur until the relatively recently demographic events 5kya. A
+sampled lineage from the Nama, Mende, and Gumuz have probabilities of being in
+Stem 2 at the time of Stem 1 expansion (135ka) of approximately $0.145$, $0.2$,
+and $0.13$, respectively, though these probabilities change over time,
+precluding the notion of a fixed admixture proportion.
 
-In contrast, under the Multiple Merger model, we allowed for stem populations
-to have large instantaneous contributions, in different proportions, to form
-the different contemporary groups. However, the ghost populations do not neatly
-line up one-to-one with southern, western and eastern populations. We observe a
-deep bottleneck in Stem 1 down to $N_e=117$. This represents the lower bound
-allowed in our optimization (i.e. 1\% of the $N_{ancestral}$) and the size of
-this bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period
-of exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
+In contrast, under the multiple-merger model, stem populations merge with
+varying proportions to form the different contemporary groups. 
+%However, the ghost populations do not neatly line up one-to-one with southern,
+%western and eastern populations.
+We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
+the Neanderthal branch. This represents the lower bound allowed in our
+optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
+bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
+exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
 479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
 166$kya). These populations evolve independently until approximately 119kya
-when Stem 1S and Stem 2 merge to form the ancestors of the Nama, with
-proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 merge to
-form the ancestors of the Western Africans and Eastern Africans (and thus also
-all individuals who later disperse during the Out of Africa event). The
-ancestral Western / Eastern African group derives equal ancestry from the two
-stems (50\% from Stem 1E and Stem 2, Table 1). Finally, the Mende receive a
-large pulse of gene flow from Stem 2, replacing 18\% of their population 25kya.
-The later Stem 2 contribution to the Western African Mende resulted in better
-model fits (LL). This may indicate that an ancestral Stem 2 population
-occupied Western or Central Africa, broadly speaking. The differing proportions
-in the Nama and Eastern Africans may also indicate geographic separation of
-Stem 1S in southern Africa and Stem 1E in eastern Africa.
-
-\subsection*{Reconciling multiple lines of genetic evidence}
-
-Previous studies have found support for archaic admixture in Africa using
-two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
-(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
-\citep{Speidel2019-nj}. \aprcomment{Will need to check how strongly the Relate
-paper claims that this supports archaic admixture vs simply deep structure. It
-wasn't a model fit, in any case.} However, none of these studies considered a
-weakly structured stem. We validated our Continuous Migration and Multiple
-Merger models with additional independent approaches. We find that the two
-models describe the cSFS accurately \aprcomment{(Figure SX)}, even though this
-statistic was not used in the fit.  Both models outperform archaic models fit
-directly to the cSFS (for example, compare with Figure~1 in
-\citet{Durvasula2020-td}).
+when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
+proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
+equal proportions (50\% each) to form the ancestors of the Western Africans and
+Eastern Africans (and thus also all individuals who later disperse during the
+Out of Africa event). Finally, the Mende receive a large additional pulse of
+gene flow from Stem 2, replacing 18\% of their population 25kya. The later Stem
+2 contribution to the Western African Mende resulted in better model fits (LL).
+This may indicate that an ancestral Stem 2 population occupied Western or
+Central Africa, broadly speaking. The differing proportions in the Nama and
+Eastern Africans may also indicate geographic separation of Stem 1S in southern
+Africa and Stem 1E in eastern Africa.
 
@@
 
-Similarly, we generated estimates of effective population size as inferred from
-coalescence rates with Relate. Changes in Ne over time from prior studies
-strongly support a bottleneck in Out of Africa populations, but African
-populations also dip in effective size around 200kya, although this decrease is
-more modest than the OOA event \citep{Mallick2016-lx}. One possible explanation
-is that ancestral population structure during the Middle Stone Age inflated Ne
-and subsequent gene flow among ancestral Africans resulted in a lower Ne after
-100-200kya \citep{Mazet2016-wn}\aprcomment{Is this a good ref here?}. To our
-knowledge, explicit comparisons of this hypothesis with SMC-style approaches
-have not been published. We simulated genomic data under 4 models
-\aprcomment{(Figure SX)}. All models, including the Single Origin model,
-generate multi-modal curves indicating that the ancestral increase in Ne
-between 100kya-1Mya is a general feature of the coalescent rate in humans.
-Simulated Relate Ne curves under the CM and MM are distinguished by having 2
-and 3 modes, respectively. The Continuous Migration model had the better
-qualitative fit with the data, but the ancestral mode was shifted further back
-in time relative to the data. All simulated models underestimated the level of
-population growth during the Holocene, indicating additional free parameters
-are necessary to refine these growth rates. \aprcomment{Discussion on this
-paragraph moved to issues (issue \#1)}
+\subsection*{Reconciling multiple lines of genetic evidence}
+
+Previous studies have found support for archaic admixture in Africa using
+two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
+(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
+\citep{Speidel2019-nj}. However, none of these studies considered a weakly
+structured stem. We validated our inferred models with additional independent
+approaches. We find that the observed cSFS (conditioned on the derived allele
+being carried in the Neanderthal sample) is very well-described by the merger
+model (Figures~\ref{fig:4}A-C and
+\ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
+even though this statistic was not used in the fit. Both models outperform
+archaic models fit directly to the cSFS (for example, compare with Figure~1 in
+\citet{Durvasula2020-td}).
+
+We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
+coalescence rates over time, which are commonly interpreted as effective
+population size histories. Trajectories of $N_e$ from prior studies strongly
+support a bottleneck in Out of Africa populations. African populations also dip
+in effective size around 100-200kya, although this decrease is more modest than
+the OOA event \citep{Mallick2016-lx}. One possible explanation is that
+ancestral population structure during the Middle Stone Age inflated $N_e$ and
+subsequent gene flow among ancestral Africans resulted in a more recent
+decreased $N_e$ \citep{Mazet2016-wn}.
+%To our knowledge, explicit comparisons of this hypothesis with SMC-style
+%approaches have not been published.
+We simulated genomic data under four inferred demographic models to compare
+with observed coalescence rate histories. All models, including the
+single-origin model, generate multi-modal curves indicating that the ancestral
+increase in $N_e$ between 100kya-1Mya is a general feature of the coalescent
+rate in humans, although none of the models perfectly match the exact timing of
+the early fluctuations in $N_e$ observed in the data.
+%Simulated Relate Ne curves under the CM and MM are distinguished by having 2
+%and 3 modes, respectively.  The Continuous Migration model had the better
+%qualitative fit with the data, but the ancestral mode was shifted further back
+%in time relative to the data.
+Each simulated model also underestimated the level of population growth during
+the Holocene, indicating additional free parameters are necessary to refine
+these growth rates.
 
 Relative cross-coalescence rates (rCCR) have recently been used to estimate
-divergence between a pair of populations, as measured by the rate of
+divergences between a pair of populations, as measured by the rate of
 coalescence between two groups divided by the mean within population
@@
 midpoint estimates of rCCR are poor estimates for population divergence, often
-underestimating divergence time by 50\% or greater (e.g. Mende vs. Gumuz ~15kya
-compared to a true divergence 60kya, \aprcomment{Supp Figure SX}). We suggest
+underestimating divergence time by 50\% or greater (e.g., Mende vs. Gumuz
+$\sim$15kya compared to a true divergence of 60kya), and recent migration can
+lead to the misordering of divergence events (Figure~\ref{fig:4}E). We suggest
 that rCCR analyses which do not fit multiple parameters including gene flow

### 2022-01-10 supplement.tex
add continuous migration with isolated stems

@@
     $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.150 & 0.0019 \\
+    \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[ht]
+\caption{
+    \label{tab:supp-continuous-migration-no-mig}
+    \textbf{Best-fit parameters from the Continuous-Migration model with isolation
+    between stems.}
+    Inferred values are scaled to physical units assuming a generation time of
+    29 years. This model gave a log-likelihood of -126,644. \aprcomment{See
+    figure XX with all parameters labeled.} 
+    \aprcomment{This table is a new fit, the analog of the merger-without-stem-migration
+    model. Here, it's the same as the continuous-migration model, just with migration
+    in the stem disallowed. Thus, there is one fewer parameter, and stems are forced
+    to be isolated. Note the tight constraint on $T_stems$, similar to what we see
+    in the merger with vs. without stem migration models.}
+}
+\centering
+\begin{tabular}[t]{rP{8cm}cc}
+    \toprule
+    Parameter & Description & Value & Std. err.\\
+    \midrule
+    $N_e$ & Ancestral effective population size & 10543 & 460 \\
+    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 15297 & 1227 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 9156 & 573 \\
+    $N_{Nama_0}$ & Initial Nama size & 11329 & 423 \\
+    $N_{Nama_F}$ & Final Nama size & 219 & 11 \\
+    $N_{MSL_0}$ & Initial Mende size & 11187 & 644 \\
+    $N_{MSL_F}$ & Final Mende size & 24056 & 945 \\
+    $N_{EA}$ & Size of East African branch & 7185 & 293 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3746 & 184 \\
+    $N_{EP}$ & East African agriculturist size & 13187 & 645 \\
+    $N_{GBR_0}$ & Initial British size & 937 & 41 \\
+    $N_{GBR_F}$ & Final British size & 11478 & 420 \\
+    $N_{Neand}$ & Neanderthal size & 2285 & 113 \\
+    $T_{stems}$ & Stem split time (years) & 459720 & 21704 \\
+    $T_{Nama}$ & Nama split time (years) & 139725 & 3250 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $1.93\times10^{-5}$ & $0.360\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.43\times10^{-5}$ & $0.309\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $20.8\times10^{-5}$ & $0.97\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.46\times10^{-5}$ & $0.171\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.4\times10^{-5}$ & $1.85\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.649 & 0.0038 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.268 & 0.0048 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.153 & 0.0022 \\
+    $m_{stems}$ & Stem 1--stem 2 migration rate & $0$ & fixed \\
+    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $4.33\times10^{-5}$ & $0.891\times10^{-5}$ \\
+    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $11.3\times10^{-5}$ & $1.83\times10^{-5}$ \\
+    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $2.77\times10^{-5}$ & $0.699\times10^{-5}$ \\
     \bottomrule
@@
     $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.0021 \\
+    $m_{stems}$ & Stem 1--stem 2 migration rate & $0$ & fixed \\
     $T_{Nama}$ & Time of Nama merger event & 117392 & 8253 \\

### 2022-01-11 paper.tex
add continuous migration with isolated stems
minor update to fig 4
add fig3 and captions

@@
-\documentclass[]{article}
+\documentclass{article}
 \usepackage[super,numbers]{natbib}
@@
 improved the fits relative to zero migration ($LL \approx -102,600$ vs.
-$-107,700$ in the merger scenario and $LL \approx -115,500$ vs. $-128,000$ with
+$-107,700$ in the merger scenario and $LL \approx -115,500$ vs. $-126,600$ with
 continuous migration). With continuous migration between stems, population
@@
     \centering
-    %\includegraphics{figures/fig3.pdf}
+    \includegraphics{figures/fig3.pdf}
     \caption{
-        \textbf{fig3}
+        \textbf{Predicted population structure over time.}
+        From the best fit models of our two parameterizations (top row:
+        continuous migration, bottom row: multiple merger), we computed
+        preditions of population structure between contemporaneous populations
+        over time ($F_{ST}$) and the proportion of drift between pairs of sampled
+        populations that aligns with drift between populations in the past
+        ($\alpha^2$).
     }
@@
 \citet{Durvasula2020-td}).
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{figures/fig4}
+    \caption{
+        \textbf{Model validation using independent statistics.} (A--C) Using
+        our best fit models to the LD and pairwise diversity statistics, we
+        simulated expected conditional site-frequency-spectra (cSFS) and
+        compared to the observed cSFS from the data. Our inferred models
+        provide a good fit to the data, even though these were not used in our
+        inference. Across the three populations, ancestral state
+        misidentification was consistently inferred to be 1.5--1.7\% for
+        intergenic loci (Supp. Methods). (D, E) We used Relate
+        \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
+        which we used to estimate coalescence rate trajectories. While
+        coalescence rate distributions are informative statistics about past
+        evolutionary processes, interpretation can be hindered by migration and
+        population structure. Translating relative cross-coalescence curves
+        into population divergences is especially prone to misinterpretation.
+    }
+    \label{fig:4}
+\end{figure}
 
@@
 should be interpreted with caution.
-
-\begin{figure}[ht!]
-    \centering
-    \includegraphics{figures/fig4}
-    \caption{fig4}
-    \label{fig:4}
-\end{figure}
 

### 2022-01-17 paper.tex

@@
         \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
-        which we used to estimate coalescence rate trajectories. While
+        which we used to estimate coalescence rate trajectories and
+        and cross-coalescence rates between pairs of populations. While
         coalescence rate distributions are informative statistics about past
         evolutionary processes, interpretation can be hindered by migration and
-        population structure. Translating relative cross-coalescence curves
-        into population divergences is especially prone to misinterpretation.
+        population structure, and translating relative cross-coalescence rate
+        curves (RCCR) into population divergence times is especially prone
+        to misinterpretation. For example, the Mende-Gumuz comparison shows
+        a more recent increased RCCR than either population with the British,
+        a pattern that is recapitulated under our best-fit model, even though
+        the Mende-Gumuz split occurs prior to the Gumuz-British split.
     }

### 2022-01-17 paper.tex
Made a change to the author affiliations (Eileen + Marlo)

@@
 \author[4]{Eileen Hoal}
-\author[5]{Marlo M\"{o}ller}
+\author[4]{Marlo M\"{o}ller}
 \author[2,6,*]{Brenna M. Henn}
@@
 \affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
-\affil[3]{FIXME: DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
-\affil[4]{FIXME}
-\affil[5]{FIXME}
+\affil[3]{FIXME: }
+\affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
 \affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}

### 2022-01-17 paper.tex
Fixed typo

@@
 Relative cross-coalescence rates (rCCR) have recently been used to estimate
-divergences between a pair of populations, as measured by the rate of
+divergence between a pair of populations, as measured by the rate of
 coalescence between two groups divided by the mean within population

### 2022-01-18 paper.tex
some minor edits to discussion

@@
 hundreds of thousands of years (i.e. a weakly structured stem). We find that
-weakly structured stem models provide more likely explanations of polymorphism
+weakly-structured-stem models provide more likely explanations of polymorphism
 that had previously been attributed to contributions from archaic hominins in
@@
 admixture. This is also true of this study. While it remains prohibitive to
-fully explore the space of plausible models of early human \bmhcomment{we use
-this phrase a lot. human or hominin? Tim?} population structure, we sought to
-capture model uncertainty by exploring multiple parameterizations of early
-human population structure. The best-fit models presented here include
-reticulation and low-level migration between early human populations rather
-than archaic admixture. We cannot rule out that more complex models involving
+fully explore the space of plausible models of early human population structure,
+we sought to capture model uncertainty by exploring multiple parameterizations of
+early history.
+\bmhcomment{we use this phrase a lot. human or hominin? Tim?}
+The best-fit models presented here include reticulation and migration between
+early human populations rather than archaic admixture from long-isolated branches.
+We cannot rule out that more complex models involving
 additional stems, or hybrid models including both weak structure and archaic
@@
 refrained from introducing additional branches (and therefore additional
-parameters) during that period. \sgcomment{Some statement about hitting the
-boundary} Rather than interpreting the two stems as representing well-defined
+parameters) during that period.
+\sgcomment{Some statement about hitting the boundary}
+Rather than interpreting the two stems as representing well-defined
 and stable populations over hundreds of thousands of years, we interpret the
 weakly structured stem as consistent, e.g., with a coalescence and
-fragmentation model \citep{Scerri2019-xg}.  Eventual ancient DNA samples from
-Africa could further distinguish the archaic admixture model from the weakly
-structured stem model. \sgcomment{Or models involving more diversity within
-Africa?}
+fragmentation model \citep{Scerri2019-xg}.
+Models including additional diversity within Africa,
+and eventual ancient DNA samples from Africa, could further distinguish
+the archaic admixture model from the weakly-structured-stem model.
 
@@
 the recombination clock accurately relate to geological chronologies
-(Supplement). During the Middle Stone Age, the Multiple Merger indicates three
+(Supporting Information).
+During the Middle Stone Age, the Multiple Merger indicates three
 major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
@@
 work has focused on archaic admixture to explain patterns of polymorphism
-inconsistent with a single origin model. Here, we have shown that weakly
-structured stem models better capture these patterns. They also help explain an
+inconsistent with a single origin model. Here, we have shown that
+weakly-structured-stem models better capture these patterns.
+They also help explain an
 ecological riddle posed by the archaic admixture model. Neanderthal populations
@@
 human lineage over hundreds of thousands of years despite closer geographic
-proximity and reproductive compatibility. The weakly structured stem model
+proximity and reproductive compatibility. The weakly-structured-stem model
 resolves this ecological riddle by accommodating continuous or recurrent
@@
 \citep{Petr2019-xo}, and their interpretation should be re-examined in light of
-a weakly structured stem model in Africa.
+a weakly-structured-stem model in Africa.
 
@@
 assume that this correspondence also holds for early human ancestral
-populations in Africa, the weakly structured stem model would predict that the
+populations in Africa, the weakly-structured-stem model would predict that the
 fossils deriving from each of the two stems should be morphologically quite
@@
 
-See Supplement, for now.
+Methods are detailed in the Supporting Information.
 

### 2022-01-21 paper.tex
multiple small edits and comments

@@
 similar. 
-Despite genetic similarity between these populations, an inferred \sgcomment{4 to 10\%} of random genetic differences among [modern human populations] can be attributed to genetic drift between stems.
+Despite genetic similarity between these populations, an inferred \sgcomment{4 to 10\%} of random genetic differences among [modern human populations] can be attributed to genetic drift between stem populations.
 We show that model misspecification explains variation in previous
@@
 groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
-included the British (GBR) from the 1000 Genomes Project in our demographic
+included the British (GBR) from the 1000 Genomes Project \sgcomment{phase 3} in our demographic
 models as a representative source of back-to-Africa gene flow and recent
@@
 other populations exceed $\sim$140kya. We conclude that, regardless of model
-specification, geographic patterns of \emph{Homo sapiens} population structure
+specification, geographic patterns of contemporary \emph{Homo sapiens} population structure
 date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
+\sgcomment{Alt formulation: Thus all models suggest that broad geographic patterns of contemporary population structure date back to MIS 5}
 Although we find evidence for earlier population structure in Africa (see
@@
 Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
+\sgcomment{Here it is not clear whether we are still talking about parameters that we fixed.
+I think we are also missing a clear description of the best-fit model.
+That *could* go after the description of the model uncertainty}
 affected the ancestors of the Ethiopian agricultural populations, comprising
@@
 
-To account for population structure prior to 135kya, we allowed for two or more
+To account for population structure prior to 135kya, we allowed \sgcomment {alt: three of our four models allowed} for two or more
 “stem” populations which could diverge either before or after the Neanderthal
@@
 other stem population, so that diversifying branches are formed through the
+\sgcomment{This sentence is not clear to me. Do we mean: In other words, the more recent populations are formed by mergers of multiple ancestral populations }
 coalescence of multiple ancestral populations. Depending on parameter values,
@@
 reflecting an informative statistical approach. As expected, model assumptions
-have a larger impact on parameter estimates. To convey model uncertainty we
+have a larger impact on parameter estimates (and thus, uncertainty). To convey model uncertainty, we
 highlight features of two inferred models with high likelihoods, the
@@
 
+\sgcomment{I don't love this new structure because it breaks the model description in two parts separated by the bootstrap.
+Maybe we have to, but I would have preferred a unified description of the models.
+For example, we could start this section by the bootstrap discussion ("In all models, likelihood differences are large and bootstrap confidence interval narrow within each specified model ...
+We therefore highlight two models".
+Then we describe model 1, describe model 2, and contrast the two. 
+ ) }
 Under the continuous-migration model, one of the two stems diverges into
@@
         preditions of population structure between contemporaneous populations
-        over time ($F_{ST}$) and the proportion of drift between pairs of sampled
+        over time ($F_{ST}$ \sgcomment{not sure I want Fst here, vs pi}) and the proportion of drift between pairs of sampled
         populations that aligns with drift between populations in the past
-        ($\alpha^2$).
+        ($\alpha^2$). \sgcomment{point to supplement, since nonstandard}
     }

### 2022-01-30 paper.tex
Addressing comments, Results

@@
 approaches are more challenging to interpret and fit to data. However, new
-population genetic inference tools now allow for inference involving tens to
+population genetic tools now allow for inference involving tens to
 hundreds of genomes from multiple populations and greater complexity
@@
 groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
-included the British (GBR) from the 1000 Genomes Project \sgcomment{phase 3} in our demographic
+included the British (GBR) from the 1000 Genomes Project in our demographic
 models as a representative source of back-to-Africa gene flow and recent
@@
 Regardless of the model choice for early epochs, inference of human demographic
-history in the last 150kya was remarkably robust. The earliest divergence among
+history for the last 150kya was remarkably robust. The earliest divergence among
 contemporary human populations differentiates the southern African Nama from
@@
 set to 60kya, just prior to the split of Eurasians and East Africans 50kya. All
-models predict relatively high subsequent gene flow between the two regions
+models infer relatively high subsequent gene flow between the two regions
 ($m\approx2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
@@
 
-To account for population structure prior to 135kya, we allowed \sgcomment {alt: three of our four models allowed} for two or more
+To account for population structure prior to 135kya, three of our four models allowed for two or more
 “stem” populations which could diverge either before or after the Neanderthal
@@
 populations expands, with instantaneous pulse (or “merger”) events from the
-other stem population, so that diversifying branches are formed through the
-\sgcomment{This sentence is not clear to me. Do we mean: In other words, the more recent populations are formed by mergers of multiple ancestral populations }
-coalescence of multiple ancestral populations. Depending on parameter values,
+other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
 this scenario encompasses archaic introgression and
-fragmentation-and-coalescence models.
+fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
+(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
+reflecting an informative statistical approach. Model assumptions
+have, however, a larger impact on parameter estimates (and thus, uncertainty). To convey model uncertainty, we
+highlight features of the two inferred models with high likelihoods. These are referred to as the
+“multiple-merger” and the the “continuous-migration” model. Both allow for
+migration between stem branches, but differ primarily in the timing of the
+early divergence of stem populations and their relative $N_e$
+(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
+during the Middle Stone Age.
 
@@
 improved the fits relative to zero migration ($LL \approx -102,600$ vs.
-$-107,700$ in the merger scenario and $LL \approx -115,500$ vs. $-126,600$ with
+$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
 continuous migration). With continuous migration between stems, population
@@
 migrant lineages each generation estimated as
-$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this roughly compares
+$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
 to inferred migration rates between connected contemporary populations over the
@@
 
-Bootstrap-based confidence intervals are relatively narrow within each
-specified model
-(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
-reflecting an informative statistical approach. As expected, model assumptions
-have a larger impact on parameter estimates (and thus, uncertainty). To convey model uncertainty, we
-highlight features of two inferred models with high likelihoods, the
-“multiple-merger” and the the “continuous-migration” model that both allow
-migration between stem branches. These differ primarily in the timing of the
-early divergence of stem populations and their relative $N_e$
-(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
-during the Middle Stone Age.
-
 \sgcomment{I don't love this new structure because it breaks the model description in two parts separated by the bootstrap.
@@
 Then we describe model 1, describe model 2, and contrast the two. 
- ) }
+ )} \bmh{what do you think of the revised text?}
+
 Under the continuous-migration model, one of the two stems diverges into

### 2022-02-04 supplement.tex
add a few supp figure captions

@@
         and ancient branches from the merger-without-stem-migration model.}
-        Unlike the continuous-migration model, 
+        Unlike the continuous-migration model, population structure in the stems
+        following the divergence of Neanderthal and human ancestors results
+        in differentiation among modern populations that aligns with drift
+        between the stems. In this model, migration between stem populations
+        is disallowed.
     }
@@
         and ancient branches from the merger-with-stem-migration model.}
+        Differentiation between modern present-day populations aligns with
+        up to $\approx 5\%$ of the drift between stems, although some
+        population pairs show even less shared drift.
     }
@@
         \textbf{Conditional SFS compared to single-origin model.}
+        The single-origin model provides a poor fit to the observed cSFS in
+        the Mende, Gumuz, and Nama. Ancestral state misidentification was
+        inferred to be between 1 and $2\%$, and misidentification rates were
+        consistent across model comparisons. Misidentification rates of
+        transition-type mutations were roughly double that of transversion-type
+        mutations, consistent with the higher mutation rates of transition
+        mutations vs. transversions.
     }
@@
         \textbf{Conditional SFS compared to merger-without-stem-migration model.}
+        The excess of intermediate-frequency variants in the Nama and Mende
+        cSFS results in a poorer fit to the cSFS compared to the continuous-migration
+        and merger-with-stem-migratino models. Along with the poorer likelihood this
+        model provides to the two-locus statistics, validation usin the cSFS
+        disfavors a model with strict isolation between stem populations.
     }
@@
         \textbf{Conditional SFS compared to merger-with-stem-migration model.}
+        The merger-with-stem-migration model, in addition to providing the best
+        fit to heterozygosity and LD statistics, provided the best fit to the cSFS
+        in the Nama, Gumuz, and Mende populations. Rates of ancestral state
+        misidentification were consistent across each comparison, with the
+        ancestral state of transversion mutations estimated to be $0.7-1.0\%$, and
+        that of transition mutations estimated to be $1.9-2.0\%$. These
+        misidentification rates are consistent with the known difference in mutation
+        rates between transitions and transversions, as well as other SFS-based
+        inferences in humans using the Thousand Genomes data and the 6-primate
+        alignment (see Methods).
     }
@@
         \textbf{Inverse instantaneous coalescence rates inferred from the data.}
+        Using combined data from the Thousand Genomes Project and the AGRP,
+        we reconstructed gene genealogies using Relate and calculated the
+        inverse instantaneous coalescence rate for samples within each population.
+        (A) Eurasian populations (CEU: European American, GBR: British,
+        CHB: Han Chinese, PJL: Punjabi), (B) Thousand Genomes African populations
+        (ACB: Afro-Caribbean, ASW: African American, ESN: Esan, GWD: Gambian,
+        LWK: Luhya, MSL: Mende, YRI: Yoruba), (C) AGRP populations.
     }

### 2022-02-04 supplement.tex
complete filling in supp fig captions

@@
 \subsection{Re-inference of IM models from simulated data}
+\label{sec:IM-reinfer}
 
@@
 The simulations assumed 10 diploid individuals were sampled from each
-population.
+population. \aprcomment{Simulation details - 500 Mb, r=1e-8, u=2e-8}
 
@@
         subset to focal populations in the model inferences.}
+        Fig.~\ref{fig:supp-iicr-data} shows IICR curves from reconstructed
+        genealogies inferred using a larger dataset (including nearly 1,200
+        individuals) than the subset of populations we primarily focused on
+        here. To assess the robustness of Relate-inferred genealogies and for a
+        direct comparison to our inferred demographic models, we ran Relate on
+        the 290 genomes from the Nama, Gumuz, Mende, Amhara/Oromo, and British.
+        While the reconstructed IICR curves are qualitatively similar to those
+        from the full dataset for these populations in the distant past, there
+        are noticeable discrepancies over the recent history in the last 10s of
+        thousands of years. Left: Relate run on full dataset, showing focal
+        populations. Right: Relate run directly on the subset dataset of focal
+        populations.
     }
@@
     \caption{
-        \textbf{Inverse instantaneous coalescence rates reinferred from simulated data.}
+        \textbf{Inverse instantaneous coalescence rates reinferred from
+        simulated data.} We simulated data under our four highlighted models,
+        sampling the same number of individuals per population as the original
+        dataset, and then ran Relate on each simulated dataset to reconstruct
+        IICR curves for each. Each model has qualitatively different histories,
+        particularly for the distant past with varying numbers and timings of
+        oscillations. However, due to the discrepancies in reconstructed
+        genealogies using even the same data as input
+        (Figs.~\ref{fig:supp-iicr-data-subset} and \ref{fig:supp-rccr-data}),
+        it is difficult to draw firm conclusions from qualitative comparisons
+        between reconstructed genealogies.
     }
@@
         \textbf{Expected inverse instantaneous coalescence rates from models.}
+        The inferred demographic models specify population sizes, split times,
+        and migration rates and events, making it possible to compute the exact
+        expected IICR curve for each population. Due to discrete events
+        (instantaneous size changes, for example), the models can produce
+        non-smooth IICR curves, although each predicts a period of increased
+        ``effective size'' in the past $\sim200-500$ka.
     }
@@
     \caption{
-        \textbf{Relative cross-coalescence rates inferred from the data.}
+        \textbf{Relative cross-coalescence rates inferred from the data.} Using
+        the same reconstructed genealogies from Relate, we computed RCCR curves
+        for pairs of populations using genealogies from the full dataset
+        ($\sim1,200$ individuals) and the subset of focal populations (290
+        individuals). Despite the overlap in samples, the two sets of inferred
+        genealogies provided inconsistent RCCR, with both the timing and
+        ordering of increased RCCR varying between the two datasets.
+        Genealogies reconstructed from the full dataset provide a better match
+        to each of inferred demographic models
+        (Figs.~\ref{fig:supp-rccr-single-origin}--\ref{fig:supp-rccr-merger-with-stem-migration}).
     }
@@
         \textbf{Relative cross-coalescence rates reinferred from simulated data
-        under the single-origin model.}
+        under the single-origin model.} RCCR curves from genealogies
+        reconstructed from simulated data using Relate match those inferred
+        from the full dataset (\label{fig:supp-rccr-data}, left) for distant
+        timepoints. There are large discrepancies in the very recent history,
+        where Relate and other genome-wide coalescence-based methods are known
+        to be underpowered.
     }
@@
         \textbf{Relative cross-coalescence rates reinferred from simulated data
-        under the continuous-migration model.}
+        under the continuous-migration model.} RCCR curves from genealogies
+        reconstructed from simulated data using Relate match those inferred
+        from the full dataset (\label{fig:supp-rccr-data}, left) for distant
+        timepoints. There are large discrepancies in the very recent history,
+        where Relate and other genome-wide coalescence-based methods are known
+        to be underpowered.
     }
@@
         \textbf{Relative cross-coalescence rates reinferred from simulated data
-        under the merger-without-stem-migration-model.}
+        under the merger-without-stem-migration-model.} RCCR curves from genealogies
+        reconstructed from simulated data using Relate match those inferred
+        from the full dataset (\label{fig:supp-rccr-data}, left) for distant
+        timepoints. There are large discrepancies in the very recent history,
+        where Relate and other genome-wide coalescence-based methods are known
+        to be underpowered.
     }
@@
         \textbf{Relative cross-coalescence rates reinferred from simulated data
-        under the merger-with-stem-migration model.}
+        under the merger-with-stem-migration model.} RCCR curves from genealogies
+        reconstructed from simulated data using Relate match those inferred
+        from the full dataset (\label{fig:supp-rccr-data}, left) for distant
+        timepoints. There are large discrepancies in the very recent history,
+        where Relate and other genome-wide coalescence-based methods are known
+        to be underpowered.
     }
@@
         \textbf{Expected relative cross-coalescence rates from inferred models.}
+        Our inferred demographic models allow for exact calculation of expected
+        RCCR curves, showing that while the parameterizations of the four models
+        are quite different, they each predict very similar RCCR profiles.
     }
@@
     \caption{
-        \textbf{Neanderthal-matching deep branch distribution from the data}
+        \textbf{Neanderthal-matching deep branch distribution from the data.}
+        Following the approach of \citet{Speidel2019-nj}, we categorized
+        Neanderthal-matching deep branches for each population from
+        Relate-inferred gene genealogies. The proportion of deep branches with
+        Neandethal affinity is highest for Eurasian populations, as expected,
+        and higher for African populations with recent Eurasian ancestry
+        (Oromo, Amhara, ACB, ASW, and Nama) than the Gumuz and West African
+        populations. 
     }
@@
     \caption{
-        \textbf{Deep branch distribution reinferred from simulated data.}
+        \textbf{Deep branch distribution reinferred from simulated data.} We
+        used Relate to reinfer gene genealogies from simulated data under the
+        four highlighted models and categorized Neanderthal-matching deep
+        branches in the simulated datasets. Each model provided a qualitative
+        match to the data (highest proportion of Neanderthal-matching deep
+        branches in the British, followed by Amhara/Oromo and Nama). Similar to
+        reinferred IIRC and RCCR, no model provided a perfect match to the
+        reconstructions using Relate with the data, but the similarities
+        between reconstructed genealogies in the simulated data suggest such
+        coalescence-rate based curves are underpowered to differentiate between
+        plausible models of history.
     }
@@
     \caption{
-        \textbf{IM model split times reinferred from simulated data of full model.}
+        \textbf{IM model split times reinferred from simulated data of full
+        model.} We simulated 10 diploid individuals under the four highlighted
+        models (Sec.~\ref{sec:IM-reinfer}). Using the joint-SFS, we reinferred
+        a simpler isolation-with-migration (IM) model for pairs of populations
+        including the Nama to explore the effect of model misspecification on
+        the earliest inferred divergence times between Khoe-San and other
+        populations. We simulated data with Nama individuals sampled at present
+        ($t=0$) as well as Nama individuals sampled 2,500 years ago, prior to
+        the recent gene flow from East Africans and Europeans. We fit two
+        models for each model and population pair: one that allowed for
+        migration between the populations (IM), and one that disallowed
+        migration (clean split). Despite all original models having recent
+        population divergences $\sim120$ka, the reinferred IM models all
+        settled on divergence times greater than 200ka and sometimes much
+        greater. If the true historical model includes reticulation or
+        long-lasting structure, simpler IM models may be severely biased in
+        favor of more ancient divergence times.
     }

### 2022-02-04 supplement.tex
missed on caption

@@
         \textbf{Conditional SFS compared to continuous-migration model.}
+        The continuous-migration model provides a better fit to the data than
+        either the single-origin model or the merger-without-stem-migration
+        model, although systematic deviations are still noticeable in each
+        population.
     }

### 2022-02-07 supplement.tex
update fig3 to use heterozygosity

@@
 
+\subsection{What does this mean for expectation of morphology btw branches?}
+
+\aprcomment{Tim to fill in, and edit section header}
+
 \section{Validations using simulations from inferred demographic models}
@@
 \subsection{Mutation versus recombination rates}
+\label{sec:supp-mutation}
 
@@
 \citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
-demographic inference in this study. However, \moments differs from other
-approaches by relying on a recombination clock, which has been argued to be
-more robust to estimation than a mutational clock \citep{Moorjani2016-ur}. To
-fit patterns of diversity which were inferred using the recombination map here,
-mutation rates on the order of $2\times10^{-8}$ would be needed.
-\aprcomment{This discrepancy between the mutation clock and recombination
-clock\dots}
+demographic inference in this study. \moments differs from other approaches by
+relying on a recombination clock, which has been argued to be more robust to
+estimation than a mutational clock \citep{Moorjani2016-ur}.
+
+From our recombination-rate-calibrated demographic models, we can estimate the
+mutation rates needed to fit patterns of pairwise diversity. With our best-fit
+model (merger-with-stem-migration,
+Table~\ref{tab:supp-merger-with-stem-migration}), a mutation rate of
+$1.33\times10^{-8}$ recovers the observed diversity within and between
+populations, closely matching the average mutation rate estimated from the
+Gnomad mutation model. The other inferred demographic models
+(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-without-stem-migration})
+would require mutation rates of $1.75-1.90\times10^{-8}$ to match pairwise
+diversity. \aprcomment{Because the merger-with-stem-migration model provides
+the best fit to the LD statistics, the cSFS, \emph{and} the Gnomad mutation
+rate, \dots}
+
+\aprcomment{The differences in recovered mutation rates between models could
+cut both ways: on one hand, it could help discriminate between models. One the
+other hand, it should mutation rate estimation in this manner is sensitive to
+the model parameterization, and we might expect it to be poorly constrained.
+Did we just get lucky that our best-fit model also matches the Gnomad mutation
+rate estimate for these same intergenic regions?}
 
@@
     \caption{
-        \textbf{Predicted pairwise differentiation between contemporaneous populations.}
-        $H_{i,j}$ is the predicted pairwise differentiaion between two populations,
-        which we normalize by the observed pairwise diversity in the Mende sample.
-        The single-origin model (A) shows larger deviations between observed and
-        expected differentiation than the models with early human structure
-        (B: continuousmigration, C: merger without stem migration, D: merger with 
-        stem migration). In the merger models (C and D), while $F_{ST}$ is large
-        between early stems due to the sharp bottleneck in one of those branches,
-        the pairwise differences are comparable to the continuous-migration model.
-        Stem 1E and Stem 1S have low differentiation, as the bottleneck in Stem 1
-        (which they both split from) sharply reduces diversity by the time of the
-        split. This has the effect of large $F_{ST}$ but small $H_{i,j}$.
+        \textbf{Predicted pairwise differentiation between contemporaneous
+        populations.} $H_{i,j}$ is the predicted pairwise differentiaion
+        between two populations per base-pair, using a mutation rate estimated
+        from matching predicted to observed pairwise diversity from the
+        intergenic data used in the fits (see Sec.~\ref{sec:supp-mutation}).
+        The single-origin model (A) shows larger deviations between observed
+        and expected differentiation than the models with early human structure
+        (B: continuous migration, C: merger without stem migration, D: merger
+        with stem migration). In the merger models (C and D), while $F_{ST}$ is
+        large between early stems due to the sharp bottleneck in one of those
+        branches, the pairwise differences are comparable to the
+        continuous-migration model. Stem 1E and Stem 1S have low
+        differentiation, as the bottleneck in Stem 1 (which they both split
+        from) sharply reduces diversity by the time of the split. This has the
+        effect of large $F_{ST}$ but small $H_{i,j}$.
     }

### 2022-02-08 paper.tex
fix bmh comment command
complete filling in supp fig captions
edits throughout

@@
-\documentclass{article}
+\documentclass[]{article}
 \usepackage[super,numbers]{natbib}
@@
 high-likelihood models which we explored did the divergence between Nama and
-other populations exceed $\sim$140kya. We conclude that, regardless of model
-specification, geographic patterns of contemporary \emph{Homo sapiens} population structure
+other populations exceed $\sim$140kya. We conclude that
+geographic patterns of contemporary \emph{Homo sapiens} population structure
 date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
-\sgcomment{Alt formulation: Thus all models suggest that broad geographic patterns of contemporary population structure date back to MIS 5}
+%\sgcomment{Alt formulation: Thus all models suggest that broad geographic patterns of contemporary population structure date back to MIS 5}
 Although we find evidence for earlier population structure in Africa (see
@@
 Then we describe model 1, describe model 2, and contrast the two. 
- )} \bmh{what do you think of the revised text?}
+ )} \bmhcomment{what do you think of the revised text?}
 
@@
         From the best fit models of our two parameterizations (top row:
-        continuous migration, bottom row: multiple merger), we computed
-        preditions of population structure between contemporaneous populations
-        over time ($F_{ST}$ \sgcomment{not sure I want Fst here, vs pi}) and the proportion of drift between pairs of sampled
+        continuous migration, bottom row: merger with stem migration), we computed
+        predicted differentiation between contemporaneous populations over time
+        as $H_{i,j}$, the expected pairwise diversity between two samples, one
+        each from populations $i$ and $j$.
+        We also computed the proportion of drift between pairs of sampled
         populations that aligns with drift between populations in the past
-        ($\alpha^2$). \sgcomment{point to supplement, since nonstandard}
+        ($\alpha^2$, see Section~\ref{sec:f4} in the Supplement).
+        While both models inferred population structure to extend deep into the
+        past, the ancient structure contributes relatively little to present-day
+        differentiation between human populations.
     }
@@
 
-\begin{figure}[ht!]
+\begin{figure}[t!]
     \centering

### 2022-02-08 supplement.tex
edits throughout

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{bm}
@@
 
+\small
 \tableofcontents
+\normalsize
 \newpage
@@
 approved through a secondary data analysis agreement for this project. Briefly,
-raw reads were aligned to GRCH37 with BWAmem \aprcomment{cite}, duplicates
+raw reads were aligned to GRCH37 with BWAmem, duplicates
 marked with Picard MarkDuplicates, reads were realigned around indels with GATK
-RealignerTargetCreater / IndelRealigner followed by BQSR with dbSNP 137.
+RealignerTargetCreater/IndelRealigner followed by BQSR with dbSNP 137.
 Contamination checks were performed requiring that FREEMIX $<0.05$;
@@
 heterozygosity check. Genomes were then variant called with GATK3.2 Unified
-Genotyper \aprcomment{cite} using joint calling across 2,478 individuals within
+Genotyper \citep{DePristo2011-up} using joint calling across 2,478 individuals within
 the ADRP dataset with a minimum base quality of 17. Data were merged with 1000
 Genomes Phase 3 \citep{1000_Genomes_Project_Consortium2015-zq} using the union
-of sites identified with bcftools isec (-n+1) \aprcomment{cite} then refined
+of sites identified with bcftools isec (-n+1) \citep{Danecek2021-kc} then refined
 with Variant Recalibrator with a truth sensitivity threshold of 99.5\%. HapMap
@@
 observed along PC1 and PC2 for 1000 Genomes vs. ADRP. Phasing on the combined
-dataset was performed via SHAPEIT2 \aprcomment{cite} and utilized the duoHMM
+dataset was performed via SHAPEIT2 \citep{Delaneau2013-aw} and utilized the duoHMM
 option for duos and trios. We highlight 82 Nama genomes which are newly
@@
 analysis, after partitioning ancestry into $k=6$ clusters with ADMIXTURE
-\aprcomment{cite} where alternative ancestries represent European, West
+\citep{Alexander2009-sw} where alternative ancestries represent European, West
 African, Near Eastern, and East African gene flow.
@@
 polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the
-\moments-LD analysis, we focused on intergenic regions because these
+\texttt{moments-LD} analysis, we focused on intergenic regions because these
 appear less affected by natural selection compared to both synonymous and
@@
 $H_{i,j}$. $f$-statistics form the basis of admixture graph analysis
-\aprcomment{cite}. Therefore, the set of statistics used here encompass
+\citep{Lipson2020-cq}. Therefore, the set of statistics used here encompass
 multiple features of genetic data that have been used to infer models of
 archaic admixture and population structure involving many populations.
-
-\aprcomment{We used the Omni YRI map (though see figure showing similarity to
-statistics under the HapMapII map). Assumed a generation time of 29 years for
-rescaling times in years to generations.}
 
@@
 Statistics were largely unchanged by using a different map
-(Figure~\ref{fig:supp-omni-hapmap-comparison}). \aprcomment{These maps are both
-inferred using array data, which is sparse. Comment on this.}
+(Figure~\ref{fig:supp-omni-hapmap-comparison}), although both maps were
+inferred using array data, which is sparser than whole-genome sequencing data.
+In subsequent analyses, we used data based on recombination rates fromthe Omni YRI map.
 
@@
 the inferred ancestral allele. Ancestral alleles alleles were determined from a
-6 primate alignment \aprcomment{cite}. This cSFS is expected to be close to
+6 primate alignment \citep{1000_Genomes_Project_Consortium2015-zq}.
+This cSFS is expected to be close to
 uniform under neutrality and a simple split model (with no subsequent
@@
 humans is at least as old as that of the human-Neanderthal split
-\citet{Durvasula2020-td,Yang2012-ze}. Because we wanted to compare our
-inferences (based on intergenic sites) to inferences from previous work (based
-on whole-genome data), we computed the cSFS for both intergenic and all sites
-genome-wide. Sites with no calls in the Vindija Neanderthal were excluded from
-this analysis. \aprcomment{Figures XX.}
+\citet{Durvasula2020-td,Yang2012-ze}.
+Sites with no calls in the Vindija Neanderthal were excluded from
+this analysis.
 
@@
 for all mutations, and for transitions and transversions separately
-\aprcomment{(Figure X, cSFS)}. Comparisons of these observed cSFS with model
+(Figures~\ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}).
+Comparisons of these observed cSFS with model
 predictions are discussed in the model prediction section below.
@@
 
+Parameters were converted to physical units
+(years, effective population sizes, etc)
+assuming a human generation time of 29 years.
+
 \subsection{General strategy for building models and introducing complexity}
@@
 allowed to vary in the fits. The total number of parameters that were
-ultimately inferred were 18-26 \aprcomment{check!}, depending on the complexity
+ultimately inferred were 21-31, depending on the complexity
 of the model.
 
-\subsection{Optimization using moments}
-
-\moments-LD uses a composite likelihood approach to simultaneously fit relative
+\subsection{Optimization using \moments}
+
+\texttt{moments-LD} uses a composite likelihood approach to simultaneously fit relative
 pairwise diversity and LD statistics over a range of recombination distances.
@@
 detailed in \citet{Ragsdale2019-nt}. For each model tested, we ran multiple
-rounds of optimization, alternating between the \emph{optimize\_log\_fim},
+rounds of optimization, alternating between the \emph{optimize\_log\_fmin},
 \emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
@@
 with stem migration, $F_{ST}$ reached a maximum value of $0.3$ between stems,
-intermediate between human-Neanderthal divergence ($0.54$ \sgcomment{check})
+intermediate between human-Neanderthal divergence ($0.45$)
 and the maximum $F_{ST}$ observed between contemporary human populations
-($0.07$.)
+($\lessapprox 0.1$.)
 
@@
 genetic differentiation between (and even within) human populations. In other
-words, our best fit model describes two stems with one stems showing reduced
-diversity, but individuals across stems showing no more differentiation than
-two modern population. \sgcomment{Previous paragraph could use rereading.}
+words, our best fit model describes two stems with one stem showing reduced
+diversity, but individuals across stems showing similar differentiation to
+that between two modern population.
 
@@
         = mf_2(\text{stem}_1, \text{stem}_2)$
-        \aprcomment{Get better resolution image from Simon}
     }
@@
 The IICR is only a reliable estimator of $N_e$ in the absence of population structure.
-.
-However, since IIRC are readily estimated and visually interesting, it is commonly attempted even when the assumption is not met \cite{}.
-Even if the assumption is not met, the inferred IIRC can be interpreted as a summary statistic for which model predictions can be compared to data. and can therefore be
+However, since IIRC are readily estimated and visually interesting,
+it is commonly attempted even when the assumption is not met
+\citep{}\aprcomment{what citation did you have in mind here, Simon?}.
+Even if the assumption is not met, the inferred IIRC can be interpreted as a
+summary statistic for which model predictions can be compared to data.
+It can therefore be
 informative about aspects of the data that our models fail to predict.
-\sgcomment{Insert inferences with same populations?}. We
-also note that the coalescence rates inferred from data used a larger set of
+We also note that the coalescence rates inferred from data used a larger set of
 populations from the combined 1000 Genomes/AGRP dataset, while our model
@@
 needed to account for recent increases in population sizes. As is the case with
-other methods that reconstruct size history trajectories from inferred coalesce
+other methods that reconstruct size history trajectories from inferred coalescence
 rates (e.g. PSMC, MSMC), \Relate has difficulty inferring coalescences that
-occur in the recent past, \aprcomment{likely a combination of some differences
-between actual and simulated data, and inherent weaknesses of those methods.}
+occur in the recent past.
 
@@
 four inferred demographic models. Comparing deep-branch distributions from the
-data (Fig.~\ref{fig:supp-deep-branches-data}) to the models
+data (Fig.~\ref{fig:supp-deep-branches}) to the models
 (Fig.~\ref{fig:supp-deep-branches-sim}), we find consistent trends between each
@@
 compared to the data, but is closest in the merger-with-stem-migration model
-(which also fit the LD statistics the best of the four models). Given the
-discrepancies between the data IICR and the model IICR, these differenes are
-not surprising \aprcomment{what more to say here\dots}
+(which also fit the LD statistics the best of the four highlighted models).
+Given the qualitative discrepancies between the data IICR and the model IICR,
+the differences in Neanderthal branch affinities is unsurprising.
 
@@
 \citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
-demographic inference in this study. \moments differs from other approaches by
+demographic inference in this study.
+\texttt{moments-LD} differs from other approaches by
 relying on a recombination clock, which has been argued to be more robust to
@@
 would require mutation rates of $1.75-1.90\times10^{-8}$ to match pairwise
-diversity. \aprcomment{Because the merger-with-stem-migration model provides
-the best fit to the LD statistics, the cSFS, \emph{and} the Gnomad mutation
-rate, \dots}
-
-\aprcomment{The differences in recovered mutation rates between models could
-cut both ways: on one hand, it could help discriminate between models. One the
-other hand, it should mutation rate estimation in this manner is sensitive to
-the model parameterization, and we might expect it to be poorly constrained.
-Did we just get lucky that our best-fit model also matches the Gnomad mutation
-rate estimate for these same intergenic regions?}
+diversity.
 
@@
 
-Recent studies \aprcomment{have been quite interested in characterizing the
-``earliest divergence time among human populations''}, and such analyses
+Recent studies have focused on characterizing the timing of the earliest
+divergences among human populations, and such analyses
 typically fit simple isolation-with-migration (IM) or ``clean split'' models to
 observed allele frequencies or inferred coalescence times among and between
-populations \citep{Bergstrom2021-iw} \aprcomment{and others}. Simple IM models
-typically fit a constant $N_e$ within each diverged population, their split
+populations (see \citet{Weaver2008-ho,Bergstrom2021-iw} for recent reviews).
+Simple IM models
+typically fit a constant $N_e$ within the two diverged population, the split
 time, and a symmetric continuous migration rate (which could be fixed to zero,
@@
 pairs of populations and reinferring an IM and clean split model for each pair.
-The simulations assumed 10 diploid individuals were sampled from each
-population. \aprcomment{Simulation details - 500 Mb, r=1e-8, u=2e-8}
+The simulations sampled 10 diploid individuals from each
+population, and we used \msprime \citep{Baumdicker2021-mj} to simulated
+500 Mb of total sequence length (the sum over500 replicates each 1 Mb in length)
+with constant recombination rate $r=10^{-8}$ and mutation rate $u=2\times10^{-8}.$
 
@@
 Even though each of our four models inferred divergences of all human
-populations $\sim$120kya, ancient structure biases inferences and gives
-inflated estimates of $T$. A similar effect occurs when the ancestral
+populations $\sim$120kya, ancient structure and size changes in the simulations
+biases re-inferences and gives inflated estimates of $T$.
+Such an effect has been shown to occur when the ancestral
 population size increases prior to the population split
-\citep{Momigliano2021-th}, which is the case in our inferred single-origin
-model. \aprcomment{I want to highlight this more in the main text.} Inferred
-$T$ between the Nama and other populations were $>$200kya under the IM model,
+\citep{Momigliano2021-th}, which is the case in our inferred single-origin model.
+Re-inferred $T$ between the Nama and other populations were $>$200kya under the IM model,
 and $\sim$50--130kya under the clean split model.
-
-\aprcomment{There's plenty more we could say about this.}
 
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -189,434. \aprcomment{See
-    figure XX with all parameters labeled.}
+    29 years. This model gave a log-likelihood of -189,434.
+    \aprcomment{include fixed parameters}
 }
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -126,644. \aprcomment{See
-    figure XX with all parameters labeled.} 
+    29 years. This model gave a log-likelihood of -126,644.
+    \aprcomment{include fixed parameters}
     \aprcomment{This table is a new fit, the analog of the merger-without-stem-migration
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -115,500. \aprcomment{See
-    figure XX with all parameters labeled.} 
+    29 years. This model gave a log-likelihood of -115,500.
+    \aprcomment{include fixed parameters}
 }
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -107,652. \aprcomment{See
-    figure XX with all parameters labeled.} 
+    29 years. This model gave a log-likelihood of -107,652.
+    \aprcomment{include fixed parameters} 
 }
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -102,633. \aprcomment{See
-    figure XX with all parameters labeled.} 
+    29 years. This model gave a log-likelihood of -102,633.
+    \aprcomment{include fixed parameters} 
 }
@@
         The single-origin model provides a poor fit to the observed cSFS in
-        the Mende, Gumuz, and Nama. Ancestral state misidentification was
+        the Mende, Gumuz, and Nama. Ancestral state misidentification
+        (labeled as $p_m$ in the figure legends) was
         inferred to be between 1 and $2\%$, and misidentification rates were
@@
         model, although systematic deviations are still noticeable in each
-        population.
+        population. Ancestral state misidentification ($p_m$), as estimated
+        with all models, is roughly doubled for transition-type mutations
+        compared to transversions.
     }
@@
         disfavors a model with strict isolation between stem populations.
+        Ancestral state misidentification ($p_m$), as estimated
+        with all models, is roughly doubled for transition-type mutations
+        compared to transversions.
     }
@@
         inferences in humans using the Thousand Genomes data and the 6-primate
-        alignment (see Methods).
+        alignment (see Methods). Ancestral state misidentification ($p_m$), as estimated
+        with all models, is roughly doubled for transition-type mutations
+        compared to transversions.
     }

### 2022-02-09 paper.tex
minor edits

@@
 similar.
-Despite genetic similarity between these populations, an inferred \sgcomment{4 to 10\%} of random genetic differences among [modern human populations] can be attributed to genetic drift between stem populations.
+Despite genetic similarity between these populations, an inferred 1--4\%
+of genetic differentiation among contemporary human populations can be
+attributed to genetic drift between stem populations.
 We show that model misspecification explains variation in previous
@@
         \textbf{Geography and genetic diversity within Africa.}
-        The combined 1000 Genomes and AGRP datasets incude populations from
+        (A) The combined 1000 Genomes and AGRP datasets include populations from
         West, East, and South Africa. Both PCA (B, C) and \texttt{ADMIXTURE}
@@
         from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
-        British and Vindija Neanderhtal individual (bold fonts).
+        British and Vindija Neanderthal individual (bold fonts).
         \aprcomment{work on this caption}
@@
 generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
-also thank Sriram Sankararam, \dots for helpful discussion. This research was
+thank Sriram Sankararaman for helpful discussion. This research was
 supported by an NIH grant R35GM133531 (to BMH) and XXX. The content is solely
@@
 
-\break
-
 \bibliographystyle{unsrtnat}

### 2022-02-13 paper.tex
Update paper.tex

@@
 \author[4]{Marlo M\"{o}ller}
-\author[2,6,*]{Brenna M. Henn}
-\author[7,**]{Simon Gravel}
+\author[2,6,^,*]{Brenna M. Henn}
+\author[7,^,**]{Simon Gravel}
 \affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
 \affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
-\affil[3]{FIXME: }
+\affil[3]{Department of Molecular and Human Genetics, Baylor College of Medicine, TX, USA}
 \affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
@@
 \affil[7]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
+\affil[^]{Co-Corresponding Authors}
 \affil[*]{bmhenn@ucdavis.edu}
@@
 groups, we infer long-lasting structure between two or more weakly
-differentiated ancestral Homo spp. populations connected by gene flow over
+differentiated ancestral \emph{Homo spp.} populations connected by gene flow over
 hundreds of thousands of years (i.e. a weakly structured stem). We find that
-weakly-structured-stem models provide more likely explanations of polymorphism
+weakly structured stem models provide more likely explanations of polymorphism
 that had previously been attributed to contributions from archaic hominins in
@@
         and signatures of recent migration and gene flow within Africa.
-        We built parameterized models using the newly-sequenced Nama, Mende
+        We built parameterized models using the newly-sequenced Nama from South Africa, Mende
         from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
@@
 other African groups between 110--135kya, with low to moderate levels of
-subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
+subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
 high-likelihood models which we explored did the divergence between Nama and
-other populations exceed $\sim$140kya. We conclude that
-geographic patterns of contemporary \emph{Homo sapiens} population structure
-date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
-%\sgcomment{Alt formulation: Thus all models suggest that broad geographic patterns of contemporary population structure date back to MIS 5}
+other populations exceed $\sim$140kya. 
+We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
+date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
 Although we find evidence for earlier population structure in Africa (see
@@
 of optimizing a large number of parameters, we fixed several parameters related
-to recent population history so as to focus on more ancient events (Supp.
-Methods). The time of divergence of Western and Eastern African populations was
-set to 60kya, just prior to the split of Eurasians and East Africans 50kya. All
-models infer relatively high subsequent gene flow between the two regions
-($m\approx2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
-following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
-Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
-\sgcomment{Here it is not clear whether we are still talking about parameters that we fixed.
-I think we are also missing a clear description of the best-fit model.
-That *could* go after the description of the model uncertainty}
+to recent population history so as to focus on more ancient events (see below). Fixed parameters included
+divergence time the time of divergence between Western and Eastern African populations,
+set to 60kya, just prior to the split of Eurasians and East Africans set to 50kya. 
+We also fixed the amount of admixture from Neanderthals to Europeans directly
+following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). 
+All models infer relatively high subsequent gene flow between the two regions
+($m\approx2\times10^{-4}$). Back-to-Africa gene flow at the beginning of the Holocene primarily
 affected the ancestors of the Ethiopian agricultural populations, comprising

### 2022-02-13 paper.tex
Update paper.tex
Typos, fixed parameters section, removed some comments

@@
 We also fixed the amount of admixture from Neanderthals to Europeans directly
-following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). 
-All models infer relatively high subsequent gene flow between the two regions
-($m\approx2\times10^{-4}$). Back-to-Africa gene flow at the beginning of the Holocene primarily
+following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). Such constraints, informed by prior research,
+allowed us to infer robust migration rates. For example all models infer relatively high gene flow between Eastern and Western Africa
+($m\approx2\times10^{-4}$). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
 affected the ancestors of the Ethiopian agricultural populations, comprising
@@
         In both the the parameterizations of early population structure,
-        continuous migration (A) and fragmentation with mergers (B), models
+        continuous migration (A) and multiple mergers (B), models
         that include ongoing migration between stem populations outperform
-        those in which stem populations are isolated. Most populations are
+        those in which stem populations are isolated. Most recent populations are
         connected by continuous, reciprocal migration that are incidated by 
@@
         continuous-migration model (A), $LL=-102,600$ vs $-115,500$,
-        respectively, and both models provide a much better fit to diversity
-        and LD statistics than a single-origin model of recent expansion.
+        respectively.
     }
@@
         & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
-        & g, & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
+        & g & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
         & h & East Africans, British & 50 & 3.87e-5 & 50 \\
@@
         Labeled migration rates correspond to symmetric continuous migration
-        bands shown in Figure~\ref{fig:2}. Both the continuous-migration and
+        bands shown in Figure~\ref{fig:2}. Both the continuous migration and
         merger models inferred a relatively deep split of human stem branches,
@@
 
-\sgcomment{I don't love this new structure because it breaks the model description in two parts separated by the bootstrap.
-Maybe we have to, but I would have preferred a unified description of the models.
-For example, we could start this section by the bootstrap discussion ("In all models, likelihood differences are large and bootstrap confidence interval narrow within each specified model ...
-We therefore highlight two models".
-Then we describe model 1, describe model 2, and contrast the two. 
- )} \bmhcomment{what do you think of the revised text?}
-
 Under the continuous-migration model, one of the two stems diverges into
@@
 ($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
-allowed to occur until the relatively recently demographic events 5kya. A
+allowed to occur until 5kya. A
 sampled lineage from the Nama, Mende, and Gumuz have probabilities of being in
@@
 varying proportions to form the different contemporary groups. 
-%However, the ghost populations do not neatly line up one-to-one with southern,
-%western and eastern populations.
 We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
@@
 across models, reflecting a degree of confounding among these parameters, we
-refrained from introducing additional branches (and therefore additional
-parameters) during that period.
+refrained from introducing additional branches associated with more
+parameters during that period.
 \sgcomment{Some statement about hitting the boundary}
@@
 and stable populations over hundreds of thousands of years, we interpret the
-weakly structured stem as consistent, e.g., with a coalescence and
+weakly structured stem as consistent with a population coalescence and
 fragmentation model \citep{Scerri2019-xg}.
 Models including additional diversity within Africa,
-and eventual ancient DNA samples from Africa, could further distinguish
+and ancient DNA samples from Africa, could further distinguish
 the archaic admixture model from the weakly-structured-stem model.
@@
 eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
-isolation among the stems is variable across model fits, models with this
-period of divergence, isolation and then a merger event (i.e. a “reticulation”)
+isolation among the stems is variable across model fits, models with a period of 
+divergence, isolation and then a merger event (i.e. a “reticulation”)
 out-performed models with bifurcating divergence and continuous gene flow.
@@
 Africa would need to have stayed in relative reproductive isolation from the
-human lineage over hundreds of thousands of years despite closer geographic
+ancestral human lineage over hundreds of thousands of years despite closer geographic
 proximity and reproductive compatibility. The weakly-structured-stem model
@@
 haplotypes such as at \emph{EPAS1} among Tibetans (e.g.,
-\citet{Zhang2021-xx}).\sgcomment{Probably not the best reference} Under an
-archaic African admixture model, the estimated 8--10\% introgression rate is
+\citet{Zhang2021-xx}). Under previous
+archaic African admixture models, the estimated 8--10\% introgression rate is
 much higher than Neanderthal gene flow, and would have plausibly been fertile
-ground for dramatic selection for or against archaic-derived haplotypes over a
-short period \citep{Wall2019-ao}. By contrast, adaptation under a weakly
+ground for dramatic selection for or against archaic-derived haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
 structured stem would have occurred continuously over much longer periods.
@@
 \citep{Petr2019-xo}, and their interpretation should be re-examined in light of
-a weakly-structured-stem model in Africa.
+a weakly-structured-stem model within Africa.
 
@@
 crania such as XX and YY are unlikely representatives of branches that
-contributed appreciably to modern human ancestries.\sgcomment{Here I am being
-somewhat forceful. We could use conditional to emphasize that these conclusion
-are under the weakly structured stem model.}
+contributed appreciably to modern human ancestries.
 

### 2022-02-18 paper.tex
fix up author block
FIgure 1 legend

@@
 \author[4]{Marlo M\"{o}ller}
-\author[2,6,^,*]{Brenna M. Henn}
-\author[7,^,**]{Simon Gravel}
+\author[2,6,$\dag$,*]{Brenna M. Henn}
+\author[7,$\dag$,**]{Simon Gravel}
 \affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
@@
 \affil[7]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
-\affil[^]{Co-Corresponding Authors}
+\affil[$\dag$]{Co-Corresponding Authors}
 \affil[*]{bmhenn@ucdavis.edu}
@@
     \caption{
-        \textbf{Geography and genetic diversity within Africa.}
-        (A) The combined 1000 Genomes and AGRP datasets include populations from
-        West, East, and South Africa. Both PCA (B, C) and \texttt{ADMIXTURE}
-        (D) illustrate the well-characterized high levels of genetic diversity
-        and signatures of recent migration and gene flow within Africa.
-        We built parameterized models using the newly-sequenced Nama from South Africa, Mende
+        \textbf{Representative genetic diversity within Africa.}
+        (A) Select populations from 1000 Genomes and ADRP illustrate diversity from
+        western, eastern and southern Africa. We chose representatives from each region
+        to build parameterized models, including the newly-sequenced Nama from South Africa, Mende
         from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
-        British and Vindija Neanderthal individual (bold fonts).
-        \aprcomment{work on this caption}
+        British and Vindija Neanderthal individual (bold fonts). (B, C) PCA highlights the range of genetic divergence anchored 
+        by western Africans, Nama, Gumuz and the British. (D) \texttt{ADMIXTURE}
+        illustrates signatures of recent gene flow in Africa which reflect colonial-period migration into the Nama,
+        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
     }
@@
 to recent population history so as to focus on more ancient events (see below). Fixed parameters included
-divergence time the time of divergence between Western and Eastern African populations,
+the time of divergence between Western and Eastern African populations,
 set to 60kya, just prior to the split of Eurasians and East Africans set to 50kya. 
@@
 following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). Such constraints, informed by prior research,
-allowed us to infer robust migration rates. For example all models infer relatively high gene flow between Eastern and Western Africa
+allowed us to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
 ($m\approx2\times10^{-4}$). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
@@
 contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
-was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE}
-\citep{Alexander2009-sw} (Figure~\ref{fig:1}).
+was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:1}).
 
@@
         those in which stem populations are isolated. Most recent populations are
-        connected by continuous, reciprocal migration that are incidated by 
+        connected by continuous, reciprocal migration that are indicated by 
         double-headed arrows (labels matched to migration rates and divergence

### 2022-02-18 paper.tex
minor edits

@@
 sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
-recruited populations of recent West African ancestry. While ancient DNA from
+recruited populations of recent West African ancestry (Figure \ref{fig:1}C). While ancient DNA from
 Eurasia has helped us understand early human history outside of Africa, there
@@
         though they were connected by ongoing migration that maintained their
-        genetic similarity.
+        genetic similarity. Migration rates above \sgcomment{XX} are highlighted in bold}
         *: The ancestors of the Nama diverged from other African groups

### 2022-02-18 paper.tex
fixed tex bug and added discussin of figure 3C

@@
         though they were connected by ongoing migration that maintained their
-        genetic similarity. Migration rates above \sgcomment{XX} are highlighted in bold}
+        genetic similarity. Bold text indicates migration rates above \sgcomment{XX} 
         *: The ancestors of the Nama diverged from other African groups
@@
 proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
-the early branches remain closely related (Figure 3).
+the early branches remain closely related (Figure 3). Because of this 
+relatedness, only $1\%$ to $4\%$ of genetic differentiation among modern populations can be traced back
+to this early population structure (Figure 3 and supplementary text)
 

### 2022-02-18 supplement.tex
minor edits
used p instead of f for frequencies, for consistency with rest of text

@@
 To evaluate the degree to which divergence among modern populations is
-inherited from divergence between early stem population, we computed f4
+inherited from divergence between early stem population, we computed $f_4$
 statistics of the form
@@
     f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = 
-    \E[(f_{\text{stem}_1} - f_{\text{stem}_2})(f_{\text{modern}_1}-f_{\text{modern}_2})]
+    \E[(p_{\text{stem}_1} - p_{\text{stem}_2} )(p_{\text{modern}_1} -p_{\text{modern}_2} )]
     \label{eq:f4}
@@
 This statistic measures co-linearity of genetic drift between modern
-populations $(f_{\text{modern}_1}-f_{\text{modern}_2})$ and stem populations
-$(f_{\text{stem}_1}-f_{\text{stem}_2})$ (see
+populations $(p_{\text{modern}_1} -p_{\text{modern}_2} )$ and stem populations
+$(p_{\text{stem}_1} -p_{\text{stem}_2} )$ (see
 Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}
@@
     f_2(\text{modern}_1, \text{modern}_2) =
-    \E[(f_{\text{modern}_1}-f_{\text{modern}_2})^2]),
+    \E[(p_{\text{modern}_1} -p_{\text{modern}_2} )^2]),
 \end{equation}
@@
 allele frequency space. In particular, the vector 
-$v=(f_{\text{modern}_1}-f_{\text{modern}_2})$, 
+$v=(p_{\text{modern}_1} -p_{\text{modern}_2} )$, 
 representing differences in allele frequencies across modern populations,
 can be written as $v=v_\perp + v_\parallel$, a component perpendicular and
-a component parallel to the divergence $(f_{\text{stem}_1}-f_{\text{stem}_2})$
+a component parallel to the divergence $(p_{\text{stem}_1} -p_{\text{stem}_2} )$
 across stems. We can then decompose 
@@
 and $\sim$50--130kya under the clean split model.
-
-\break
 

### 2022-02-19 supplement.tex
clean up alpha description

@@
 This statistic measures co-linearity of genetic drift between modern
-populations $(p_{\text{modern}_1} -p_{\text{modern}_2} )$ and stem populations
-$(p_{\text{stem}_1} -p_{\text{stem}_2} )$ (see
+populations $v= (p_{\text{modern}_1} -p_{\text{modern}_2} )$ and stem populations
+$w= (p_{\text{stem}_1} -p_{\text{stem}_2} )$ (see
 Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}
-and \citet{Lipson2020-cq}).
+and \citet{Lipson2020-cq} for an overview of $f_4$ statistics).
 
@@
     f_2(\text{modern}_1, \text{modern}_2) =
-    \E[(p_{\text{modern}_1} -p_{\text{modern}_2} )^2]),
+    \E[(p_{\text{modern}_1} -p_{\text{modern}_2} )^2],
 \end{equation}
-we interpret the $f$-statistics as scalar products between vectors in 
-allele frequency space. In particular, the vector 
-$v=(p_{\text{modern}_1} -p_{\text{modern}_2} )$, 
-representing differences in allele frequencies across modern populations,
-can be written as $v=v_\perp + v_\parallel$, a component perpendicular and
-a component parallel to the divergence $(p_{\text{stem}_1} -p_{\text{stem}_2} )$
-across stems. We can then decompose 
+we interpret the $f$-statistics as scalar products between vectors $v$ and $w$, i.e.,  
+$f_2(\text{modern}_1, \text{modern}_2) = v\cdot v$ and 
+$f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = v\cdot w.$
+We can then write $v=v_\perp + v_\parallel$ as the sum of a component $v_\perp$ perpendicular and
+a component $v_\parallel$ parallel to $w = (p_{\text{stem}_1} -p_{\text{stem}_2} ).$
+We can then decompose 
 $f_2(\text{modern}_1,\text{modern}_2) = v_\perp^2+v_\parallel^2$, with  
 \begin{equation}
-    v_\perp^2 = f_4^2(\text{stem}_1, \text{stem}_2; 
-    \text{modern}_1, \text{modern}_2)/f_2(\text{stem}_1, \text{stem}_2)
+    v_\parallel^2 = \frac{(v \cdot w)^2}{w\cdot w} = \frac{ f_4^2(\text{stem}_1, \text{stem}_2; 
+    \text{modern}_1, \text{modern}_2)}{f_2(\text{stem}_1, \text{stem}_2)}
 \end{equation}
@@
 Finally, the proportion of modern-day structure
-$f_2(\text{modern}_1, \text{modern}_2)$ that is accounted for by the
+$v^2 = f_2(\text{modern}_1, \text{modern}_2)$ that is accounted for by the
 component parallel to ancient population structure is
 \begin{equation}
-    \alpha^2 = \frac{v_\perp^2}{f_2(\text{modern}_1, \text{modern}_2)} =
+    \alpha^2 = \frac{v_\parallel^2}{v^2} =
     \frac{f_4^2(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)}
@@
 occurring before that time. For this reason, we also interpret $\alpha^2$ as the
-proportion of the $f_2$ divergence across modern populations that is explained by
+proportion of the $f_2$ divergence across modern populations that is \emph{explained by}
 the differences in allele frequencies across ancient populations. We report

### 2022-02-24 paper.tex
cleanup upon read-through

@@
 Although we find evidence for earlier population structure in Africa (see
-below), present-day populations cannot be easily mapped onto these more ancient
-stem groups as only a small proportion of drift between modern populations can
+below), present-day populations cannot be easily mapped onto the more ancient
+`stem' groups as only a small proportion of drift between modern populations can
 be attributed to drift between stems (Figures~\ref{fig:3} and
@@
 of optimizing a large number of parameters, we fixed several parameters related
-to recent population history so as to focus on more ancient events (see below). Fixed parameters included
+to recent population history so as to focus on more ancient events. 
+Fixed parameters included
 the time of divergence between Western and Eastern African populations,
@@
 We also fixed the amount of admixture from Neanderthals to Europeans directly
-following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). Such constraints, informed by prior research,
-allowed us to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
-($m\approx2\times10^{-4}$). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
+following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp.
+Methods).
+These constraints allowed us to integrate information from previous \sgcomment{genetic and archaelogical?} research,
+to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
+($m\approx2\times10^{-4},$ the proportion of migrant lineages per generation). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
 affected the ancestors of the Ethiopian agricultural populations, comprising
@@
 reflecting an informative statistical approach. Model assumptions
-have, however, a larger impact on parameter estimates (and thus, uncertainty). To convey model uncertainty, we
+have, however, a larger impact on parameter estimates (and thus, real uncertainty). To convey model uncertainty, we
 highlight features of the two inferred models with high likelihoods. These are referred to as the
-“multiple-merger” and the the “continuous-migration” model. Both allow for
+“multiple-merger” and the the “continuous-migration” models. Both allow for
 migration between stem branches, but differ primarily in the timing of the
@@
 proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
-the early branches remain closely related (Figure 3). Because of this 
+the early branches remain closely related, \sgcomment{and each branch contributes large amounts to all contemporary populations} (Figure 3). 
+Because of this 
 relatedness, only $1\%$ to $4\%$ of genetic differentiation among modern populations can be traced back
@@
 gene flow from Stem 2, replacing 18\% of their population 25kya. The later Stem
-2 contribution to the Western African Mende resulted in better model fits (LL).
+2 contribution to the Western African Mende resulted in better model fits (LL \sgcomment{Something is missing here}).
 This may indicate that an ancestral Stem 2 population occupied Western or
@@
 \ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
-even though this statistic was not used in the fit. Both models outperform
+even though this statistic was not used in the fit. Both models actually outperform
 archaic models fit directly to the cSFS (for example, compare with Figure~1 in
@@
 the Holocene, indicating additional free parameters are necessary to refine
-these growth rates.
+these growth rates. \sgcomment{I am not sure I am happy with these interpretations 
+-- let's discuss. I feel like this paragraph does not support our data well, and a likely culprit 
+in explaining the differences is the inaccuracy of TMRCA estimates from real data.}
 
@@
 parameters during that period.
-\sgcomment{Some statement about hitting the boundary}
 Rather than interpreting the two stems as representing well-defined
@@
 Models including additional diversity within Africa,
-and ancient DNA samples from Africa, could further distinguish
+and early ancient DNA samples from Africa, could further distinguish
 the archaic admixture model from the weakly-structured-stem model.
@@
 Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
-The second ~100kya between Stem 1E and Stem 2, results in the formation of the
+The second $\approx100$kya between Stem 1E and Stem 2, results in the formation of the
 ancestors of East/West Africans as well as later “Out of Africans”. The rapid
@@
 across the continent; for example, a genetic lineage sampled in the Nama has an
-XX probability of being in the ancestral ‘southern’ subpopulation 50kya versus
-XX probability of being in the ‘eastern’ subpopulation (Figure XX). We also
+\sgcomment{XX} probability of being inherited from the ancestral ‘southern’ subpopulation 50kya versus
+\sgcomment{XX} probability of being inherited from the ‘eastern’ subpopulation (Figure \sgcomment{XX}). We also
 find that Stem 2 continued to contribute to western Africans during the Last
@@
 fossils deriving from each of the two stems should be morphologically quite
-similar (supplementary text X)\aprcomment{Will we still include this?}, even if
+similar (supplementary text \sgcomment{X})\aprcomment{Will we still include this?} \sgcomment{yes!}, even if
 they are from different parts of Africa, with differences across stems closer
 to those across contemporary human populations than to human-Neanderthal
-differences. Differences within stems would be comparable to differences within
-recent human populations. Identifying the morphological signature of ancient
-structure in the fossil record will require careful analysis of many broadly
-similar remains. Our model further predicts that morphologically distinct
-crania such as XX and YY are unlikely representatives of branches that
+differences. Identifying the morphological signature of ancient
+structure in the fossil record will thus require careful analysis of many broadly
+similar remains. Morphologically distinct
+crania such as \sgcomment{XX} and \sgcomment{YY} are unlikely representatives of branches that
 contributed appreciably to modern human ancestries.

### 2022-02-24 paper.tex
edited relate conclusions

@@
 with observed coalescence rate histories. All models, including the
-single-origin model, generate multi-modal curves indicating that the ancestral
-increase in $N_e$ between 100kya-1Mya is a general feature of the coalescent
-rate in humans, although none of the models perfectly match the exact timing of
-the early fluctuations in $N_e$ observed in the data.
+single-origin model, generate multi-modal curves recapitulating an inferred ancestral
+increase in $N_e$ between 100kya-1Mya. Thus the reduction in inverse coalescence rate 
+is captured in our model, but is primarily explained by population structure in the best-fitting models \sgcomment{is that fair?}.     
+None of the models perfectly match the exact timing of
+the early fluctuations in $N_e$ observed in the data \ref{fig:supp-iicr-data-subset}. 
 %Simulated Relate Ne curves under the CM and MM are distinguished by having 2
@@
 Each simulated model also underestimated the level of population growth during
-the Holocene, indicating additional free parameters are necessary to refine
+the Holocene, indicating either artifacts of data inference in real data or that additional free parameters are necessary to refine
 these growth rates. \sgcomment{I am not sure I am happy with these interpretations 
 -- let's discuss. I feel like this paragraph does not support our data well, and a likely culprit 
-in explaining the differences is the inaccuracy of TMRCA estimates from real data.}
+in explaining the differences is the inaccuracy of TMRCA estimates from real data.
+Also, we would need a link to a supplementary figure!}
 

### 2022-02-24 paper.tex
cleaned up fig references and relate paragraph

@@
 single-origin model, generate multi-modal curves recapitulating an inferred ancestral
-increase in $N_e$ between 100kya-1Mya. Thus the reduction in inverse coalescence rate 
+increase in $N_e$ between 100kya-1Mya (Figure \ref{fig:supp-iicr-sim}). Thus the reduction in inverse coalescence rate 
 is captured in our model, but is primarily explained by population structure in the best-fitting models \sgcomment{is that fair?}.     
 None of the models perfectly match the exact timing of
-the early fluctuations in $N_e$ observed in the data \ref{fig:supp-iicr-data-subset}. 
+the early fluctuations in $N_e$ observed in the data (Figure \ref{fig:supp-iicr-data-subset}). 
+\sgcomment{Perhaps: These discrepancies are comparable to discrepancies observed in relate curves inferred 
+for the same populations using different subsets of the data (see Figures \ref{ig:supp-iicr-data-subset} and \ref{ig:supp-iicr-data-subset})}
 %Simulated Relate Ne curves under the CM and MM are distinguished by having 2
@@
 the Holocene, indicating either artifacts of data inference in real data or that additional free parameters are necessary to refine
-these growth rates. \sgcomment{I am not sure I am happy with these interpretations 
+these growth rates. \sgcomment{Perhaps: There are other discrepancies between odel }
+
+\sgcomment{I am not sure I am happy with these interpretations 
 -- let's discuss. I feel like this paragraph does not support our data well, and a likely culprit 

### 2022-02-25 paper.tex
IICR rewrite

--- 2022-02-24 paper.tex
@@
 We also fixed the amount of admixture from Neanderthals to Europeans directly
-following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp.
-Methods).
-These constraints allowed us to integrate information from previous \sgcomment{genetic and archaelogical?} research,
+following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). These constraints allowed us to integrate information from previous genetic and archaelogical research,
 to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
@@
         though they were connected by ongoing migration that maintained their
-        genetic similarity. Bold text indicates migration rates above \sgcomment{XX} 
+        genetic similarity. Bold text indicates migration rates above $1e-4$ 
         *: The ancestors of the Nama diverged from other African groups
@@
 proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
-the early branches remain closely related, \sgcomment{and each branch contributes large amounts to all contemporary populations} (Figure 3). 
+the early branches remain closely related, and each branch contributes large amounts to all contemporary populations (Figure 3). 
 Because of this 
@@
 We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
-coalescence rates over time, which are commonly interpreted as effective
-population size histories. Trajectories of $N_e$ from prior studies strongly
-support a bottleneck in Out of Africa populations. African populations also dip
-in effective size around 100-200kya, although this decrease is more modest than
-the OOA event \citep{Mallick2016-lx}. One possible explanation is that
-ancestral population structure during the Middle Stone Age inflated $N_e$ and
-subsequent gene flow among ancestral Africans resulted in a more recent
-decreased $N_e$ \citep{Mazet2016-wn}.
-%To our knowledge, explicit comparisons of this hypothesis with SMC-style
-%approaches have not been published.
-We simulated genomic data under four inferred demographic models to compare
-with observed coalescence rate histories. All models, including the
-single-origin model, generate multi-modal curves indicating that the ancestral
-increase in $N_e$ between 100kya-1Mya is a general feature of the coalescent
-rate in humans, although none of the models perfectly match the exact timing of
-the early fluctuations in $N_e$ observed in the data.
-%Simulated Relate Ne curves under the CM and MM are distinguished by having 2
-%and 3 modes, respectively.  The Continuous Migration model had the better
-%qualitative fit with the data, but the ancestral mode was shifted further back
-%in time relative to the data.
-Each simulated model also underestimated the level of population growth during
-the Holocene, indicating additional free parameters are necessary to refine
-these growth rates. \sgcomment{I am not sure I am happy with these interpretations 
--- let's discuss. I feel like this paragraph does not support our data well, and a likely culprit 
-in explaining the differences is the inaccuracy of TMRCA estimates from real data.}
+coalescence rates over time in real data and data simulated from our inferred models. Many 
+previous studies have found a reduction of coalescence rates 
+between $1000$kya and $100$kya in humans, and thus inferred an increase 
+in $N_e$ during the same period. \citep{Li, Durbin}. 
+This increase in $N_e$ could be due to either an increase in population or to 
+ancestral population structure during the Middle Stone Age \citep{Mazet2016-wn}.
+All models, including the
+single-origin model, recapitulate an inferred ancestral
+increase in $N_e$ between 100kya-1Mya (Figure \ref{fig:supp-iicr-sim} and supplementary section \sgcomment{X}). Whereas the single-origin model achieves this by
+an increase in $N_e$ during that period, the best-fit model recapitulate this pattern without corresponding population size changes.
 
@@
 fossils deriving from each of the two stems should be morphologically quite
-similar (supplementary text \sgcomment{X})\aprcomment{Will we still include this?} \sgcomment{yes!}, even if
+similar (supplementary text \sgcomment{X}), even if
 they are from different parts of Africa, with differences across stems closer

### 2022-02-28 paper.tex
comments on Figure 1

@@
     \caption{
-        \textbf{Representative genetic diversity within Africa.}
-        (A) Select populations from 1000 Genomes and ADRP illustrate diversity from
+        \textbf{Genetic diversity across Africa.}
+        (A) Select populations from 1000 Genomes and ADRP   \sgcomment{define} illustrate diversity from
         western, eastern and southern Africa. We chose representatives from each region
@@
         from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
-        British and Vindija Neanderthal individual (bold fonts). (B, C) PCA highlights the range of genetic divergence anchored 
-        by western Africans, Nama, Gumuz and the British. (D) \texttt{ADMIXTURE}
+        British and Vindija Neanderthal individual (bold fonts \sgcomment{I don't see bold fonts}). (B, C) PCA highlights the range of genetic divergence anchored 
+        by western Africans, Nama, Gumuz and the British. \sgcomment{I think we should indicate the origin on PCAS, also possible variance explained.} (D) \texttt{ADMIXTURE}
         illustrates signatures of recent gene flow in Africa which reflect colonial-period migration into the Nama,

### 2022-02-28 supplement.tex
add table with lineage probabilities

@@
 
+\newcommand{\specialcell}[2][c]{%
+  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+
+\begin{table}[ht]
+\caption{
+    \label{tab:supp-lineage-probabilitys}
+    \textbf{Probability that a lineage sampled 500ya is found in a given branch.}
+    Using our best-fit model (merger with stem migration), we calculated the probability
+    that a lineage sampled in a given population (Nama, Mende, or Gumuz) is found in
+    each branch at 10ka, 55ka, and 500ka. Lineages are sampled 500 years ago to
+    avoid the effects of very recent admixture, such as recent European admixture in
+    the Nama. Rows may not sum to one, as not every branch is listed here (such as
+    Neanderthal).
+}
+\centering
+\begin{tabular}[t]{lccccccc}
+    \toprule
+    & \specialcell{Nama\\branch} & \specialcell{Mende\\branch} &
+    \specialcell{Gumuz\\branch} & \specialcell{Oromo/Amhara\\branch} & 
+    \specialcell{British\\branch} & Stem 1 & Stem 2\\
+    \midrule
+    Found 10ka & & & & & & & \\
+    Nama & 0.724 & 0.015 & 0.032 & 0.226 & 0.003 & & \\
+    Mende & 0.003 & 0.881 & 0.058 & 0.058 & 0.001 & & \\
+    Gumuz & 0.013 & 0.058 & 0.819 & 0.098 & 0.012 & & \\
+    Oromo/Amhara & 0.013 & 0.058 & 0.098 & 0.819 & 0.012 & & \\
+    \midrule
+    Found 55ka & & & & & & & \\
+    Nama & 0.677 & 0.052 & 0.264 & & & & 0.005 \\
+    Mende & 0.021 & 0.562 & 0.272 & & & & 0.143 \\
+    Gumuz & 0.060 & 0.211 & 0.704 & & & & 0.023 \\
+    Oromo/Amhara & 0.041 & 0.142 & 0.793 & & & & 0.017 \\
+    \midrule
+    Found 500ka & & & & & & & \\
+    Nama & & & & & & 0.379 & 0.618 \\
+    Mende & & & & & & 0.430 & 0.569 \\
+    Gumuz  & & & & & & 0.473 & 0.525 \\
+    Oromo/Amhara & & & & & & 0.475 & 0.517 \\
+    \bottomrule
+\end{tabular}
+\end{table}
+
 \clearpage

### 2022-03-01 paper.tex
comments and edits, Caption to figure 2

@@
     \caption{
-        \textbf{Two best fit models with a weakly structured stem.}
-        In both the the parameterizations of early population structure,
+        \textbf{A weakly structured stem leads to best-fitting models}
+        In the two best-fitting parameterizations of early population structure,
         continuous migration (A) and multiple mergers (B), models
@@
         double-headed arrows (labels matched to migration rates and divergence
-        times in Table~\ref{tab:migration-rates}), and which lasts for the
-        duration of concurrence of contemporaneous populations. The
+        times in Table~\ref{tab:migration-rates}). These migrations last for the
+        duration of concurrence \sgcomment{concurrence or co-occurence?
+What does this mean in either case?} of contemporaneous populations \sgcomment{Is what we mean: Migration rates are constant over epochs delimited by population splits or mergers, see Supplementary table XX?
+}.
+The
         merger-with-stem-migration model (B) outperformed the

### 2022-03-01 paper.tex
more edits, Caption to figure 2

@@
     \caption{
-        \textbf{A weakly structured stem leads to best-fitting models}
+        \textbf{A weakly structured stem best describes two-locus statistics}
         In the two best-fitting parameterizations of early population structure,
@@
 The
-        merger-with-stem-migration model (B) outperformed the
-        continuous-migration model (A), $LL=-102,600$ vs $-115,500$,
-        respectively.
+        merger-with-stem-migration model (B, with  $LL=-102,600$) outperformed the
+        continuous-migration model (A, with $LL=-115,500$ ).
     }

### 2022-03-01 paper.tex
table 1 caption edits

@@
         genetic similarity. Bold text indicates migration rates above $1e-4$ 
-        *: The ancestors of the Nama diverged from other African groups
-        $\sim$120--135kya, which later branch into West and East African
-        groups 60kya. Migration rates and durations are shown for 1) East Africans
-        and their ancestors, and 2) the Mende, respectively.
+        In both models, the branch ancestral to the Nama shares a common ancestral population 
+        with the other African groups $\sim$120--135kya. Following this divergence, 
+        the population ancestral to other African groups branches into West and East African
+        groups 60kya. *Migration rates and durations are shown between branches ancestral to 1) Nama and East Africans
+        and their ancestors, and 2) Nama and Mende, respectively.
+        `Divergence times' correspond to the most recent common ancestral population, and does not account for continuous migration or earlier reticulations.
     }

### 2022-03-01 paper.tex
Figure 3 captions

@@
         From the best fit models of our two parameterizations (top row:
-        continuous migration, bottom row: merger with stem migration), we computed
-        predicted differentiation between contemporaneous populations over time
-        as $H_{i,j}$, the expected pairwise diversity between two samples, one
-        each from populations $i$ and $j$.
-        We also computed the proportion of drift between pairs of sampled
-        populations that aligns with drift between populations in the past
-        ($\alpha^2$, see Section~\ref{sec:f4} in the Supplement).
-        While both models inferred population structure to extend deep into the
-        past, the ancient structure contributes relatively little to present-day
-        differentiation between human populations.
-    }
+        continuous migration, bottom row: merger with stem migration), we predicted
+       pairwise differences $H_{i,j}$ between individuals sampled from populations $i$ and $j$ 
+       existing at time $t$. 
+        To understand how drift between stems explains present-day structure, 
+        we also computed the proportion $\alpha^2$ of drift between pairs of sampled present-day
+        populations that aligns with drift between past populations (Here Nama and Mende, see Section~S\ref{sec:f4} for details and additional comparisons \sgcomment{do we have extra comparisons there?}).
+        Both models infer deep population structure with modest contributions to present-day genetic differentiation. Most present differentiation dates back to the last 100kya.}
     \label{fig:3}

### 2022-03-03 paper.tex
address some comments and update figures

@@
         \textbf{Genetic diversity across Africa.}
-        (A) Select populations from 1000 Genomes and ADRP   \sgcomment{define} illustrate diversity from
-        western, eastern and southern Africa. We chose representatives from each region
-        to build parameterized models, including the newly-sequenced Nama from South Africa, Mende
+        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
+        illustrate diversity from western, eastern and southern Africa.
+        We chose representatives from each region to build parameterized models,
+        including the newly-sequenced Nama from South Africa, Mende
         from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
-        British and Vindija Neanderthal individual (bold fonts \sgcomment{I don't see bold fonts}). (B, C) PCA highlights the range of genetic divergence anchored 
-        by western Africans, Nama, Gumuz and the British. \sgcomment{I think we should indicate the origin on PCAS, also possible variance explained.} (D) \texttt{ADMIXTURE}
-        illustrates signatures of recent gene flow in Africa which reflect colonial-period migration into the Nama,
+        British and Vindija Neanderthal individual.
+        (B, C) PCA highlights the range of genetic divergence anchored 
+        by western Africans, Nama, Gumuz and the British.
+        Percentages show variance explained by each principal component.
+        (D) \texttt{ADMIXTURE} illustrates signatures of recent gene flow in Africa
+        which reflect colonial-period migration into the Nama,
         Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
@@
     \centering
-    \includegraphics[width=0.5\textwidth]{figures/fig2}
+    \includegraphics[width=0.6\textwidth]{figures/fig2-melisa}
     \caption{
@@
         times in Table~\ref{tab:migration-rates}). These migrations last for the
-        duration of concurrence \sgcomment{concurrence or co-occurence?
-What does this mean in either case?} of contemporaneous populations \sgcomment{Is what we mean: Migration rates are constant over epochs delimited by population splits or mergers, see Supplementary table XX?
-}.
-The
+        duration of co-existence of contemporaneous populations with constant
+        migration rates over those intervals. The
         merger-with-stem-migration model (B, with  $LL=-102,600$) outperformed the
@@
         \textbf{Continuous Model} & & & & & \\
-        LL= -115,500 & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
+        $LL= -115,500$ & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
         & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
@@
         \textbf{Merger Model} & & & & & \\
-        LL= -102,600 & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
+        $LL= -102,600$ & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
         & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
@@
 Out of Africa event). Finally, the Mende receive a large additional pulse of
-gene flow from Stem 2, replacing 18\% of their population 25kya. The later Stem
-2 contribution to the Western African Mende resulted in better model fits (LL \sgcomment{Something is missing here}).
+gene flow from Stem 2, replacing 18\% of their population 25kya.
+The later Stem 2 contribution to the Western African Mende resulted in
+better model fits ($\Delta LL \approx 60,000$).
 This may indicate that an ancestral Stem 2 population occupied Western or
@@
         \textbf{Predicted population structure over time.}
-        From the best fit models of our two parameterizations (top row:
-        continuous migration, bottom row: merger with stem migration), we predicted
-       pairwise differences $H_{i,j}$ between individuals sampled from populations $i$ and $j$ 
-       existing at time $t$. 
-        To understand how drift between stems explains present-day structure, 
-        we also computed the proportion $\alpha^2$ of drift between pairs of sampled present-day
-        populations that aligns with drift between past populations (Here Nama and Mende, see Section~S\ref{sec:f4} for details and additional comparisons \sgcomment{do we have extra comparisons there?}).
-        Both models infer deep population structure with modest contributions to present-day genetic differentiation. Most present differentiation dates back to the last 100kya.}
+        From the best fit models of our two parameterizations (A and B:
+        continuous migration, C and D: merger with stem migration),
+        we predicted pairwise differences $H_{i,j}$ between individuals
+        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
+        (B and D) To understand how drift between stems explains present-day structure, 
+        we also computed the proportion $\alpha^2$ of drift between pairs of
+        sampled present-day
+        populations that aligns with drift between past populations
+        (here Nama and Mende, see Section~S\ref{sec:f4} for details and additional
+        comparisons in Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
+        Both models infer deep population structure with modest contributions to
+        present-day genetic differentiation.
+        Most present differentiation dates back to the last 100kya.
+    }
     \label{fig:3}
@@
 between $1000$kya and $100$kya in humans, and thus inferred an increase 
-in $N_e$ during the same period. \citep{Li, Durbin}. 
+in $N_e$ during the same period \citep{Li2011-le}.
 This increase in $N_e$ could be due to either an increase in population or to 
@@
 Following these merger events, the stems subsequently fracture into
-subpopulations which then appear to persist over the past ~120kya. These
+subpopulations which then appear to persist over the past $\sim120$kya. These
 subpopulations can be linked to modern-day groups despite subsequent gene flow
-across the continent; for example, a genetic lineage sampled in the Nama has an
-\sgcomment{XX} probability of being inherited from the ancestral ‘southern’ subpopulation 50kya versus
-\sgcomment{XX} probability of being inherited from the ‘eastern’ subpopulation (Figure \sgcomment{XX}). We also
+across the continent; for example, a genetic lineage sampled in the Gumuz has a
+$0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
+(Stem 1E) 150kya versus $0.03$ probability of being inherited from the ‘southern’
+subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2
+(see Table~\ref{tab:supp-lineage-probabilitys} for additional comparisons). We also
 find that Stem 2 continued to contribute to western Africans during the Last
 Glacial Period, indicative that this gene flow likely occurred in
-western/central Africa (Table 1).
+western/central Africa (Table~\ref{tab:migration-rates}).
 

### 2022-03-03 supplement.tex
address some comments and update figures

@@
 However, since IIRC are readily estimated and visually interesting,
-it is commonly attempted even when the assumption is not met
-\citep{}\aprcomment{what citation did you have in mind here, Simon?}.
+it is commonly attempted even when the assumption is not met.
 Even if the assumption is not met, the inferred IIRC can be interpreted as a

### 2022-03-03 supplement.tex
add section on pca and admixture methods

@@
 had less than 100 contiguous base pairs.
+
+\subsection{PCA and \texttt{ADMIXTURE} analysis}
+
+We used principal component analysis (PCA) from \texttt{scikit-learn}
+\citep{Pedregosa2011-ke} and \texttt{ADMIXTURE} version 1.3.0
+\citep{Alexander2011-aq} to summarize genetic diversity among a broader set of
+1000 Genomes and ADRP populations (Figure~\ref{fig:1} in the main text). Data
+filtering was identical for both analyses. We retained biallelic SNPs with no
+missing data and minor allele frequency greater that 5\%. We then thinned the
+remaining sites to remove SNPs in high linkage disequilibrium with one another.
+We used the \texttt{locate\_unlinked} function from \texttt{scikit-allel}
+version 1.2.1 \citep{Miles2021-yq} to identify SNPs in low linkage
+disequilibrium, using a threshold of $r^2=0.1$, a window size of 100 SNPs, and
+a step size of 20 SNPs, removing the remaining SNPs. This process was iterated
+five times for each chromosome, resulting in 327,656 roughly unlinked SNPs used
+in the PCA and \texttt{ADMIXTURE} analyses.
 

### 2022-03-04 supplement.tex
minor edits and clarifications

@@
 labeled ancestral branches using a parsimony in migration, referring to South,
-East, and West African branches (S/E/W AFR, Figure X). We do not know the
+East, and West African branches (S/E/W AFR,
+Figures~\ref{fig:1},~\ref{fig:supp-possible-models}). We do not know the
 geographical location of these ancestral populations (nor even if they
@@
 branches in reference to where in Africa their descendants are found.
-\sgcomment{Please double-check this paragraph! TODO Brenna?}
-
 Parameters were converted to physical units
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -126,644.
+    29 years. This model gave a log-likelihood of -126,644. This model allows for
+    continuous migration between Stem 2 and modern populations, but disallows
+    migration between Stem 2 and Stem 1, which later splits into branches leading
+    to South, East, and West African populations.
     \aprcomment{include fixed parameters}
-    \aprcomment{This table is a new fit, the analog of the merger-without-stem-migration
-    model. Here, it's the same as the continuous-migration model, just with migration
-    in the stem disallowed. Thus, there is one fewer parameter, and stems are forced
-    to be isolated. Note the tight constraint on $T_stems$, similar to what we see
-    in the merger with vs. without stem migration models.}
 }

### 2022-03-04 supplement.tex
add table with fixed parameters

@@
 \subsection{General strategy for building models and introducing complexity}
+\label{sec:models}
 
@@
 When testing multiple variations of the more complex models, we kept these
-values fixed. This reduced the potential for overfitting the more complex
+values fixed (see Table~\ref{tab:fixed_params} for a complete list of all fixed
+parameters). This reduced the potential for overfitting the more complex
 models, while reducing the computational cost of optimization.
@@
     \caption{
+        \label{tab:fixed-params}
+        \textbf{Fixed parameters in inferred demographic models.}
+        Many of the fixed parameters are specific to the Neanderthal branch, are
+        constrained by known history, or were consistently fit across multiple
+        model parameterizations. We assumed a generation time of 29 years throughout.
+        See Section~\ref{sec:models} for details.
+    }
+    \centering
+    \begin{tabular}[t]{rP{8cm}c}
+        \toprule
+        Parameter & Description & Value \\
+        \midrule
+        $T_{MH-Neand}$ & The time of the split between branches leading to Neanderthals and modern humans & 550ka \\
+        $T_{NI-Vin}$ & The time of the split between branches leading to the Vindija sample and the Neanderthal population that mixed with expanding modern humans & 80ka \\
+        $T_{Vin}$ & The sample time of the Vindija Neanderthal individual & 50ka \\
+        $T_{NI\rightarrow EUR}$ & The timing of admixture from Neanderthals to modern humans & 45ka \\
+        $f_{NI\rightarrow EUR}$ & The proportion of admixture from Neanderthals to modern humans & 0.015 \\
+        $T_{E-W}$ & The split time of East and West African populations & 60ka \\
+        $T_{GBR}$ & The split time of Eurasian and East African populations & 50ka \\
+        $T_{GBR\rightarrow EA}$ & The time of the back-to-Africa admixture event, giving rise to the Amhara/Oromo branch & 12ka \\
+        $T_{expan}$ & The expansion time of the Mende and contraction time of the Gumuz & 5ka \\
+        $T_{EA\rightarrow Nama}$ & The admixture time of East African pastoralists to the Nama & 67 gen. \\
+        $T_{GBR\rightarrow Nama}$ & The admixture time of Europeans to the Nama & 10 gen. \\
+        $T_{bottle}$ & The bottleneck time of the Nama population & 9 gen. \\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+\begin{table}[ht]
+\caption{
     \label{tab:supp-single-origin}
     \textbf{Best-fit parameters from the single-origin model.}
+    This model includes a single stem population that splits into modern human
+    groups. Fixed parameters are specifed in Table~\ref{tab:fixed-params}.
     Inferred values are scaled to physical units assuming a generation time of
     29 years. This model gave a log-likelihood of -189,434.
-    \aprcomment{include fixed parameters}
 }
@@
     to South, East, and West African populations.
-    \aprcomment{include fixed parameters}
 }
@@
     29 years. This model gave a log-likelihood of -115,500.
-    \aprcomment{include fixed parameters}
 }
@@
     29 years. This model gave a log-likelihood of -107,652.
-    \aprcomment{include fixed parameters} 
 }
@@
     29 years. This model gave a log-likelihood of -102,633.
-    \aprcomment{include fixed parameters} 
 }

### 2022-03-04 supplement.tex
cleanup small comments

@@
 
-Finally, we estimated uncertainty due to the finite amount of genetic material
-used in inference using bootstrap over 500 segments along the genome with
-roughly equal lengths of retained sequences within each segment. First, for
-each distance bin, we used these bootstrap samples to obtain a
-variance-covariances matrix across all statistics. This variance-covariance
-matrix was used to obtain a model likelihood for each recombination distance
-bin and single-locus nucleotide diversity, as a multivariate Gaussian
-likelihood. The full model likelihood was taken as the product of likelihoods
-over each bin. In other words, we optimized a composite likelihood where
-observations in different bins were taken to be independent. To account for
-correlations across bins in uncertainty estimates, we estimated parameter
-confidence intervals using the same bootstrap set using the Godambe information
-matrix \citep{Coffman2016-yq}. \sgcomment{is redundancy here with section
-"Optimization using moments". I think we can get rid of this and say:
-"optimization and uncertainty calculations are described in section
-"Optimization using moments""?, This may be a bit more complicated, since the
-Optimizatino section relies on this description. I think we should just punt
-the bootstrap description to that section. TODO: discuss Aaron + Simon}
-
 \subsection{Estimating two-locus statistics with small sample sizes}
@@
 
-\subsection{Confidence intervals using Godambe methods}
-
-The bootstrap replicates that were used to compute the variance-covariance
-structure of the observed statistics within bins were also used to build 500
-bootstrap replicates of the data by resampling with replacement. For the best
-fit parameters, we computed confidence intervals using the Godambe Information
-approach, which corrects composite likelihood estimates of confidence intervals
-to account for nonindependence in the data, including linkages between loci and
-nonindependence of recombination bins \cite{Coffman2016-yq}.
+\subsection{Uncertainty and confidence intervals using Godambe methods}
+
+To compute likelihoods, we needed an estimate of the joint distribution of summary statistics. 
+We estimated uncertainty due to the finite amount of genetic material
+used in inference using bootstrap over 500 segments along the genome with
+roughly equal lengths of retained sequences within each segment. First, for
+each distance bin, we used these bootstrap samples to obtain a
+variance-covariances matrix across all statistics. This variance-covariance
+matrix was used to obtain a model likelihood for each recombination distance
+bin and single-locus nucleotide diversity, as a multivariate Gaussian
+likelihood. The full model likelihood was taken as the product of likelihoods
+over each bin. In other words, we optimized a composite likelihood where
+observations in different bins were taken to be independent.
+
+For the best fit parameters, we computed confidence intervals using the 
+same set of bootstraped samples and Godambe Information
+approach \citep{Coffman2016-yq}, which corrects composite likelihood estimates of confidence intervals
+to account for nonindependence in the data, including linkage between loci and
+nonindependence of recombination bins.
 

### 2022-03-04 paper.tex
update figure 1 with Luke's help
cleaned up small remaining bits

@@
 single-origin model, recapitulate an inferred ancestral
-increase in $N_e$ between 100kya-1Mya (Figure \ref{fig:supp-iicr-sim} and supplementary section \sgcomment{X}). Whereas the single-origin model achieves this by
+increase in $N_e$ between 100kya-1Mya (Figure \ref{fig:supp-iicr-sim}). Whereas the single-origin model achieves this by
 an increase in $N_e$ during that period, the best-fit model recapitulate this pattern without corresponding population size changes.
@@
 early history.
-\bmhcomment{we use this phrase a lot. human or hominin? Tim?}
 The best-fit models presented here include reticulation and migration between
@@
 generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
-thank Sriram Sankararaman for helpful discussion. This research was
+thank Sriram Sankararaman for helpful discussion and Luke Anderson-Trocm\'e for
+assistance in creating the map in Figure~\ref{fig:1}. This research was
 supported by an NIH grant R35GM133531 (to BMH) and XXX. The content is solely

### 2022-03-04 paper.tex
add Tim's section to conclusion

--- 2022-03-03 paper.tex
@@
 
-\usepackage{fullpage}
-\usepackage{listings}
+\usepackage[margin=1in]{geometry}
+%\usepackage{showframe}
 \usepackage{url}
@@
 \usepackage{booktabs}
-
-\lstset{language=Python}
 
@@
     \includegraphics{figures/fig1.pdf}
-    \caption{
+    \caption[width=\textwidth]{
         \textbf{Genetic diversity across Africa.}
@@
 differentiation among human populations and between humans and Neanderthals
-\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. If we
-assume that this correspondence also holds for early human ancestral
-populations in Africa, the weakly-structured-stem model would predict that the
-fossils deriving from each of the two stems should be morphologically quite
-similar (supplementary text \sgcomment{X}), even if
-they are from different parts of Africa, with differences across stems closer
-to those across contemporary human populations than to human-Neanderthal
-differences. Identifying the morphological signature of ancient
-structure in the fossil record will thus require careful analysis of many broadly
-similar remains. Morphologically distinct
-crania such as \sgcomment{XX} and \sgcomment{YY} are unlikely representatives of branches that
-contributed appreciably to modern human ancestries.
+\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. The fossil
+record of Africa is sparse during the time period of the stems, but of the
+available fossils, some are very similar to present-day humans (e.g., from Omo
+Kibish, Ethiopia), others are similar in some features but not others (e.g.,
+from Jebel Irhoud, Morocco), and others are very different (e.g., from
+Dinaledi, South Africa) in morphology. If, as our model predicts, the
+morphological differences between the stems were comparable to those among
+present-day human populations, the most divergent fossils are unlikely to
+represent the branches that contributed appreciably to present-day human
+ancestries.
+
+%If we
+%assume that this correspondence also holds for early human ancestral
+%populations in Africa, the weakly-structured-stem model would predict that the
+%fossils deriving from each of the two stems should be morphologically quite
+%similar (supplementary text \sgcomment{X}), even if
+%they are from different parts of Africa, with differences across stems closer
+%to those across contemporary human populations than to human-Neanderthal
+%differences. Identifying the morphological signature of ancient
+%structure in the fossil record will thus require careful analysis of many broadly
+%similar remains. Morphologically distinct
+%crania such as \sgcomment{XX} and \sgcomment{YY} are unlikely representatives of branches that
+%contributed appreciably to modern human ancestries.
 

### 2022-03-04 paper.tex
fix quotes

@@
 represent populations which contributed to our ancestors as population
-precedents, or were local “dead-ends” from which contemporary \emph{Homo
+precedents, or were local ``dead-ends'' from which contemporary \emph{Homo
 sapiens} do not descend. Recently, synthetic attempts to reconcile genetic and
@@
 To account for population structure prior to 135kya, three of our four models allowed for two or more
-“stem” populations which could diverge either before or after the Neanderthal
+``stem'' populations which could diverge either before or after the Neanderthal
 split. We considered models both with and without migration between these stem
@@
 symmetric migration with those populations; or 2) one or more of the stem
-populations expands, with instantaneous pulse (or “merger”) events from the
+populations expands, with instantaneous pulse (or ``merger'') events from the
 other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
@@
 highlight features of the two inferred models with high likelihoods. These are referred to as the
-“multiple-merger” and the the “continuous-migration” models. Both allow for
+``multiple-merger'' and the the ``continuous-migration'' models. Both allow for
 migration between stem branches, but differ primarily in the timing of the
@@
 bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
-exchange with Stem 2, Stem 1 then fractures into “Stem 1E” and “Stem 1S”
+exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
 479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
@@
 isolation among the stems is variable across model fits, models with a period of 
-divergence, isolation and then a merger event (i.e. a “reticulation”)
+divergence, isolation and then a merger event (i.e. a ``reticulation'')
 out-performed models with bifurcating divergence and continuous gene flow.
@@
 The second $\approx100$kya between Stem 1E and Stem 2, results in the formation of the
-ancestors of East/West Africans as well as later “Out of Africans”. The rapid
+ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
 rise in sea levels and increased precipitation during MIS 5e, following a

### 2022-03-04 paper.tex

@@
 
-\abstract{
+\begin{abstract}
 While a modern human origin within Africa is now broadly accepted, considerable
@@
 robust inferences about deep history.
-}
+\end{abstract}
 

### 2022-03-04 paper.tex
tiny cleanup discussion'

@@
 differentiation among human populations and between humans and Neanderthals
-\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. The fossil
+\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also Section \ref{morpho_diff}). The fossil
 record of Africa is sparse during the time period of the stems, but of the
@@
 present-day human populations, the most divergent fossils are unlikely to
-represent the branches that contributed appreciably to present-day human
+represent branches that contributed appreciably to present-day human
 ancestries.

### 2022-03-04 paper.tex
minor edits

--- 2022-03-04 paper.tex
@@
 \usepackage[margin=1in]{geometry}
-%\usepackage{showframe}
 \usepackage{url}
@@
 We also fixed the amount of admixture from Neanderthals to Europeans directly
-following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Methods). These constraints allowed us to integrate information from previous genetic and archaelogical research,
+following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Information). These constraints allowed us to integrate information from previous genetic and archaelogical research,
 to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
@@
     \centering
-    \includegraphics[width=0.6\textwidth]{figures/fig2-melisa}
+    \includegraphics[width=0.65\textwidth]{figures/fig2-melisa}
     \caption{
@@
         though they were connected by ongoing migration that maintained their
-        genetic similarity. Bold text indicates migration rates above $1e-4$ 
+        genetic similarity. Bold text indicates migration rates above $10^{-4}$. 
         In both models, the branch ancestral to the Nama shares a common ancestral population 
@@
 relatedness, only $1\%$ to $4\%$ of genetic differentiation among modern populations can be traced back
-to this early population structure (Figure 3 and supplementary text)
+to this early population structure (Figure 3 and Supp. Information)
 
@@
         misidentification was consistently inferred to be 1.5--1.7\% for
-        intergenic loci (Supp. Methods). (D, E) We used Relate
+        intergenic loci (Supp. Information). (D, E) We used Relate
         \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
@@
 the recombination clock accurately relate to geological chronologies
-(Supporting Information).
+(Supp. Information).
 During the Middle Stone Age, the Multiple Merger indicates three
@@
 
-%If we
-%assume that this correspondence also holds for early human ancestral
-%populations in Africa, the weakly-structured-stem model would predict that the
-%fossils deriving from each of the two stems should be morphologically quite
-%similar (supplementary text \sgcomment{X}), even if
-%they are from different parts of Africa, with differences across stems closer
-%to those across contemporary human populations than to human-Neanderthal
-%differences. Identifying the morphological signature of ancient
-%structure in the fossil record will thus require careful analysis of many broadly
-%similar remains. Morphologically distinct
-%crania such as \sgcomment{XX} and \sgcomment{YY} are unlikely representatives of branches that
-%contributed appreciably to modern human ancestries.
-
-\section*{Methods}
-
-Methods are detailed in the Supporting Information.
-
 \section*{Acknowledgements}
@@
 generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
-thank Sriram Sankararaman for helpful discussion and Luke Anderson-Trocm\'e for
-assistance in creating the map in Figure~\ref{fig:1}. This research was
-supported by an NIH grant R35GM133531 (to BMH) and XXX. The content is solely
-the responsibility of the authors and does not necessarily represent the
-official views of the National Institutes of Health.
+thank Luke Anderson-Trocm\'e for assistance in creating the map in
+Figure~\ref{fig:1}. This research was supported by an NIH grant R35GM133531 (to
+BMH). The content is solely the responsibility of the authors and does
+not necessarily represent the official views of the National Institutes of
+Health.
 

### 2022-03-07 paper.tex
Update paper.tex
Modified the last paragraph and added citations in comments.

@@
 differentiation among human populations and between humans and Neanderthals
-\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also Section \ref{morpho_diff}). The fossil
+\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also Section \ref{morpho_diff}).
+This correspondence allows predictions of our model to be related to the fossil record.
+The fossil
 record of Africa is sparse during the time period of the stems, but of the
 available fossils, some are very similar to present-day humans (e.g., from Omo
-Kibish, Ethiopia), others are similar in some features but not others (e.g.,
-from Jebel Irhoud, Morocco), and others are very different (e.g., from
-Dinaledi, South Africa) in morphology. If, as our model predicts, the
-morphological differences between the stems were comparable to those among
-present-day human populations, the most divergent fossils are unlikely to
+Kibish, Ethiopia\tdwcomment{add citations: Day (1969) doi.org/10.1038/2221135a0 and Vidal et al. (2022) doi.org/10.1038/s41586-021-04275-8}), others are similar in some features but not others (e.g.,
+from Jebel Irhoud, Morocco\tdwcomment{add citations: Hublin et al. (2017) doi.org/10.1038/nature22336 and Richter et al. (2017) doi.org/10.1038/nature22335}), and others are very different (e.g., from
+Dinaledi, South Africa\tdwcomment{add citations: Berger et al.
+(2015) doi.org/10.7554/eLife.09560 and Dirks et al.
+(2017) doi.org/10.7554/eLife.24231.001}) in morphology.
+If, as our model predicts, the
+genetic differences between the stems were comparable to those among
+present-day human populations, the most morphologically divergent fossils are unlikely to
 represent branches that contributed appreciably to present-day human

### 2022-03-07 paper.tex
Update paper.tex
I tried to make the language about hominin, modern human, present-day human, and contemporary human more precise, less ambiguous, and more consistent.

@@
 \begin{abstract}
-While a modern human origin within Africa is now broadly accepted, considerable
+While it is now broadly accepted that \emph{Homo sapiens} origined within Africa, considerable
 uncertainty surrounds specific models of divergence and migration across the
@@
 groups, as well as 44 newly whole-genome sequenced individuals from the Nama
-(Khoe-San). Despite the complexity of African population history, present-day
+(Khoe-San). Despite the complexity of African population history, contemporary
 population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
-population divergence among contemporary populations occurs 120-135kya, between
+population divergence among contemporary populations occurs 120-135kya\tdwcomment{The standard notation for anyone who studies the past (e.g., paleoanthropologists, geologists) for thousands of years ago is ka, and Ma for millions of years ago.
+Given this, I think we should use these throughout the paper.
+I can make the changes, but I didn't want to do so without discussing them}, between
 the Khoe-San and other groups. Prior to the divergence of contemporary African
@@
 \citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
-Neanderthal admixture with modern humans in Eurasia, several recent articles
-have shown that introducing an archaic ghost population contributing to African
+Neanderthal admixture with humans in Eurasia, several recent articles
+have shown that introducing an archaic hominin ghost population contributing to African
 populations in the period surrounding the Out-of-Africa migration event
@@
 This has driven speculation about the geographic range of this ghost
-population, possible links to specific archaic remains, and the possibility of
+population, possible links to specific fossils, and the possibility of
 finding ancient DNA evidence (e.g., \citet{Hsieh2016-gk}). However, these prior
 articles share two weaknesses. First, they only contrast a single-origin model
-with an archaic admixture model, leaving out other plausible models
+with an archaic hominin admixture model, leaving out other plausible models
 (Figure~\ref{fig:supp-possible-models} and \citet{Henn2018-rf}). Second, they
@@
 starting point 4 classes of models (single population expansion, single
-population expansion with regional persistence, archaic admixture, and
+population expansion with regional persistence, archaic hominin admixture, and
 multi-regional evolution, Figure~\ref{fig:supp-possible-models}), using 290
@@
 Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
-for archaic gene flow from Neanderthals in non-Africans and gauge the relative
+for gene flow from Neanderthals into non-Africans and gauge the relative
 time depth of divergence, assuming Neanderthals diverged 550kya from a common
@@
 unstructured source followed by migration between populations, without allowing
-for contribution from an archaic African lineage or population structure prior
+for contribution from an African archaic hominin lineage or population structure prior
 to the expansion (Figure~\ref{fig:supp-possible-models}A). As expected
@@
 differentiation of contemporary groups, including models allowing for ancestral
-reticulation (Figure~\ref{fig:supp-possible-models}B), archaic admixture
+reticulation (Figure~\ref{fig:supp-possible-models}B), archaic hominin admixture
 (Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
@@
 Although we find evidence for earlier population structure in Africa (see
-below), present-day populations cannot be easily mapped onto the more ancient
-`stem' groups as only a small proportion of drift between modern populations can
+below), contemporary populations cannot be easily mapped onto the more ancient
+`stem' groups as only a small proportion of drift between contemporary populations can
 be attributed to drift between stems (Figures~\ref{fig:3} and
@@
 other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
-this scenario encompasses archaic introgression and
+this scenario encompasses archaic hominin introgression and
 fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
@@
 periodic) gene flow qualitatively distinguishes these models from previously
-proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
+proposed archaic hominin admixture models (Figure~\ref{fig:supp-possible-models}C) as
 the early branches remain closely related, and each branch contributes large amounts to all contemporary populations (Figure 3). 
 Because of this 
-relatedness, only $1\%$ to $4\%$ of genetic differentiation among modern populations can be traced back
+relatedness, only $1\%$ to $4\%$ of genetic differentiation among contemporary populations can be traced back
 to this early population structure (Figure 3 and Supp. Information)
@@
         sampled from populations $i$ and $j$ existing at time $t$ (A and C).
-        (B and D) To understand how drift between stems explains present-day structure, 
+        (B and D) To understand how drift between stems explains contemporary structure, 
         we also computed the proportion $\alpha^2$ of drift between pairs of
-        sampled present-day
+        sampled contemporary
         populations that aligns with drift between past populations
@@
         Both models infer deep population structure with modest contributions to
-        present-day genetic differentiation.
+        contemporary genetic differentiation.
         Most present differentiation dates back to the last 100kya.
@@
 
-Previous studies have found support for archaic admixture in Africa using
+Previous studies have found support for archaic hominin admixture in Africa using
 two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
@@
 even though this statistic was not used in the fit. Both models actually outperform
-archaic models fit directly to the cSFS (for example, compare with Figure~1 in
+archaic hominin admixture models fit directly to the cSFS (for example, compare with Figure~1 in
 \citet{Durvasula2020-td}).
@@
 data inconsistent with a single origin model must be explained by archaic
-admixture. This is also true of this study. While it remains prohibitive to
+hominin admixture. This is also true of this study. While it remains prohibitive to
 fully explore the space of plausible models of early human population structure,
@@
 The best-fit models presented here include reticulation and migration between
-early human populations rather than archaic admixture from long-isolated branches.
+early human populations rather than archaic hominin admixture from long-isolated branches.
 We cannot rule out that more complex models involving
 additional stems, or hybrid models including both weak structure and archaic
-admixture may better explain the data. Because parameters related to the split
+hominin admixture may better explain the data. Because parameters related to the split
 time, migration rates, and relative sizes of the early stems were variable
@@
 and early ancient DNA samples from Africa, could further distinguish
-the archaic admixture model from the weakly-structured-stem model.
+the archaic hominin admixture model from the weakly-structured-stem model.
 
@@
 could observe disproportionate ancestry contributions from different stems in
-modern groups. We observed two merger events. The first, between Stem 1S and
+contemporary groups. We observed two merger events. The first, between Stem 1S and
 Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
@@
 subpopulations which then appear to persist over the past $\sim120$kya. These
-subpopulations can be linked to modern-day groups despite subsequent gene flow
+subpopulations can be linked to contemporary groups despite subsequent gene flow
 across the continent; for example, a genetic lineage sampled in the Gumuz has a
@@
 
-\subsection*{Contrasting archaic admixture and a weakly structured stem}
+\subsection*{Contrasting archaic hominin admixture and a weakly structured stem}
 
-Evidence for archaic admixture in Eurasia has bolstered the plausibility of
-archaic admixture having also occurred in Africa. For this reason, previous
-work has focused on archaic admixture to explain patterns of polymorphism
+Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of
+archaic hominin admixture having also occurred in Africa. For this reason, previous
+work has focused on archaic hominin admixture to explain patterns of polymorphism
 inconsistent with a single origin model. Here, we have shown that
@@
 They also help explain an
-ecological riddle posed by the archaic admixture model. Neanderthal populations
+ecological riddle posed by the archaic hominin admixture model. Neanderthal populations
 were separated from early \emph{Homo sapiens} by thousands of kilometers and
@@
 
-There is evidence for both deleterious and adaptive archaic-derived alleles in
-modern genomes in the form of a depletion of Neanderthal ancestry in regulatory
-regions \citep{Petr2019-xo} or an increased frequency of archaic-related
+There is evidence for both deleterious and adaptive archaic-hominin-derived alleles in
+contemporary genomes in the form of a depletion of Neanderthal ancestry in regulatory
+regions \citep{Petr2019-xo} or an increased frequency of archaic-hominin-related
 haplotypes such as at \emph{EPAS1} among Tibetans (e.g.,
 \citet{Zhang2021-xx}). Under previous
-archaic African admixture models, the estimated 8--10\% introgression rate is
+African archaic hominin admixture models, the estimated 8--10\% introgression rate is
 much higher than Neanderthal gene flow, and would have plausibly been fertile
-ground for dramatic selection for or against archaic-derived haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
+ground for dramatic selection for or against archaic-hominin-derived haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
 structured stem would have occurred continuously over much longer periods.
@@
 record of Africa is sparse during the time period of the stems, but of the
-available fossils, some are very similar to present-day humans (e.g., from Omo
+available fossils, some are very similar to contemporary humans (e.g., from Omo
 Kibish, Ethiopia\tdwcomment{add citations: Day (1969) doi.org/10.1038/2221135a0 and Vidal et al. (2022) doi.org/10.1038/s41586-021-04275-8}), others are similar in some features but not others (e.g.,
@@
 genetic differences between the stems were comparable to those among
-present-day human populations, the most morphologically divergent fossils are unlikely to
-represent branches that contributed appreciably to present-day human
+contemporary human populations, the most morphologically divergent fossils are unlikely to
+represent branches that contributed appreciably to contemporary human
 ancestries.

### 2022-03-07 paper.tex
add comments from coathors to paper

--- 2022-03-04 paper.tex
@@
 % local definitions
+\newcommand{\comment}[1]{{\textcolor{red}{Co-author: #1}}}
 \newcommand{\sgcomment}[1]{{\textcolor{red}{SG: #1}}}
@@
 \affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
-\affil[3]{Department of Molecular and Human Genetics, Baylor College of Medicine, TX, USA}
+\affil[3]{Department of Molecular and Human Genetics, Baylor College of Medicine, Houston, TX, USA}
 \affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
@@
 
+\comment{About title: "Contrasting archaic admixture and a weakly structured stem for human originas in Africa" Too long? Inconclusive?}
+
 \begin{abstract}
@@
 continent. Progress is hampered by a paucity of fossil and genomic data, as
-well as variability in dating. Here we use linkage disequilibrium and
+well as variability in dating \comment{prior dataing estimates}.
+Here we use linkage disequilibrium and
 diversity-based statistics, optimized for rapid, complex demographic inference
-to discriminate among such models. We infer detailed demographic models for
+to discriminate among such models \comment{\ldots to evaluate the fit of various models}.
+We infer detailed demographic models for
 populations across Africa, including representatives from eastern and western
@@
 groups, we infer long-lasting structure between two or more weakly
-differentiated ancestral \emph{Homo spp.} populations connected by gene flow over
+differentiated ancestral \emph{Homo spp.}
+\comment{maybe don't abbreviate; won't be understood outside of evolutionary biology as much}
+populations connected by gene flow over
 hundreds of thousands of years (i.e. a weakly structured stem). We find that
@@
 precedents, or were local ``dead-ends'' from which contemporary \emph{Homo
-sapiens} do not descend. Recently, synthetic attempts to reconcile genetic and
+sapiens} do not descend. 
+\comment{unpack more; explain why these are at odds}
+Recently, synthetic \comment{synthetic doesn't feel quite right here, can delete}
+attempts to reconcile genetic and
 paleoanthropological data include proposals for a Pan-African origin of
@@
 population coalescence and fragmentation \citep{Scerri2019-xg}. These
-approaches are more challenging to interpret and fit to data. However, new
+approaches are more challenging to interpret and fit to data.
+\comment{I think a schematic illustrating models being tested would be extremely useful in the main text.
+This could be Fig S2, or even a new figure with a further simplified illustration o fthe general idea of the models being tested to help orient the reader.}
+However, new
 population genetic tools now allow for inference involving tens to
@@
 sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
-recruited populations of recent West African ancestry (Figure \ref{fig:1}C). While ancient DNA from
+recruited populations of recent West African ancestry (Figure \ref{fig:1}C). 
+\comment{What about the LWK?}
+While ancient DNA from
 Eurasia has helped us understand early human history outside of Africa, there
 is no comparably ancient DNA to elucidate early history in Africa.
+\comment{Any point in citing this paper: https://www.nature.com/articles/s41586-022-04430-9 ?}
 
@@
 including geographically and genetically diverse populations across Africa, we
-infer demographic models that explain more aspects of genetic diversity in more
+infer demographic models that explain more aspects of genetic diversity
+\comment{this is a bit vague; make more detailed. e.g.: better fit patterns of LD and \ldots}
+in more
 populations than previously reported. These analyses confirm the inadequacy of
@@
 time depth of divergence, assuming Neanderthals diverged 550kya from a common
-stem. For each population, we computed low-order allele frequency and linkage
+stem. For each population, we computed low-order
+\comment{what do you mean by low-order? not a commonly used phrase}
+allele frequency and linkage
 disequilibrium (LD) statistics that are well suited for both low- and
@@
         Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
+        \comment{How did you land on the best fit k? Show scree plot of CV errors, or otherwise justify choice}
     }
@@
         continuous-migration model (A, with $LL=-115,500$ ).
+        \comment{If colors have a meaning, explain; also, it may be easier to change the color scheme of figure 1 to match figure 2}
     }
@@
 this scenario encompasses archaic introgression and
-fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
+fragmentation-and-coalescence models.
+\comment{another supp schematic figure may be helpful}
+For many parameters, confidence intervals based on bootstrapping are relatively narrow
 (Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
@@
 highlight features of the two inferred models with high likelihoods. These are referred to as the
-``multiple-merger'' and the the ``continuous-migration'' models. Both allow for
+``multiple-merger'' and the ``continuous-migration'' models. Both allow for
 migration between stem branches, but differ primarily in the timing of the
@@
         \textbf{Predicted population structure over time.}
+        \comment{This could have a mroe descriptive title saying the take-home point}
         From the best fit models of our two parameterizations (A and B:
@@
 an increase in $N_e$ during that period, the best-fit model recapitulate this pattern without corresponding population size changes.      
+\comment{These are pretty complicated ideas that are not explained in detail. Might be worth unpacking a bit more to make sure the reader understands}
 
@@
 isolation among the stems is variable across model fits, models with a period of 
-divergence, isolation and then a merger event (i.e. a ``reticulation'')
+divergence, isolation and then a merger event (i.e. a ``reticulation'' \comment{define at the first mention})
 out-performed models with bifurcating divergence and continuous gene flow.
@@
 geographic expansion of one or both stems. For example, if during MIS 5, either
-Stem 1S from southern Africa moved northward thereby encountering the Stem 2,
+Stem 1S \comment{where is this illustrated/labeled?}
+from southern Africa moved northward thereby encountering the Stem 2,
 or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
@@
 Paleo-Agulhas plain \citep{Marean2014-pg}.
+\comment{very specific... might be worth expanding a bit}
 
@@
 \citep{Petr2019-xo}, and their interpretation should be re-examined in light of
-a weakly-structured-stem model within Africa.
+a weakly-structured-stem model within Africa. \comment{Reviewers may suggest this demonstrated here}
 
@@
 represent branches that contributed appreciably to present-day human
-ancestries.
+ancestries. \comment{this needs more analyses/citations to back up if included}\aprcomment{but this comment was on the previous version}
 

### 2022-03-07 paper.tex

--- 2022-03-07 paper.tex
@@
 this scenario encompasses archaic hominin introgression and
-fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
+fragmentation-and-coalescence models.
+\comment{another supp schematic figure may be helpful}
+For many parameters, confidence intervals based on bootstrapping are relatively narrow
 (Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
@@
 ancestries.
+\comment{this needs more analyses/citations to back up if included}\aprcomment{but this comment was on the previous version}
 

### 2022-03-07 supplement.tex
add Tim's section to conclusion
label
fix abstract and margins

@@
 
-\usepackage{fullpage}
+\usepackage[margin=1in]{geometry}
 \usepackage{listings}
@@
 $\sigma_{Dz}$ has been shown to be sensitive to deep population structure and
-archaic admixture \citep{Ragsdale2019-nt}, and this statistic is closely
+archaic hominin admixture \citep{Ragsdale2019-nt}, and this statistic is closely
 related to $S^*$ statistics used to scan for introgressed haplotypes
@@
 multiple features of genetic data that have been used to infer models of
-archaic admixture and population structure involving many populations.
+archaic hominin admixture and population structure involving many populations.
 
@@
 uniform under neutrality and a simple split model (with no subsequent
-migration) between the ancestors of modern humans and Neanderthal
+migration) between the ancestors of humans and Neanderthals
 \citep{Chen2007-iy}. By contrast, a U-shaped distribution has been taken as
-evidence for archaic introgression from a population whose split from modern
-humans is at least as old as that of the human-Neanderthal split
+evidence for introgression from an archaic hominin population that split from humans at least as far back in time as the human-Neanderthal split
 \citet{Durvasula2020-td,Yang2012-ze}.
@@
 proposed scenarios of population structure, size changes, and/or archaic
-admixture. The simplest model, in terms of number of parameters, was a
+hominin admixture. The simplest model, in terms of number of parameters, was a
 single-origin expansion of modern humans, with no structure in the stem and no
-archaic admixture aside from the Neanderthal admixture in Eurasian populations
+archaic hominin admixture aside from the Neanderthal admixture in Eurasian populations
 following the out-of-Africa migration. This model allowed for a population size
@@
 migration rates, with parameterizations flexible enough to encompass proposed
-scenarios of either archaic admixture or population structure, both connected
+scenarios of either archaic hominin admixture or population structure, both connected
 by gene flow or with isolation between stems.
@@
 were restricted to isolation between stem branches. Depending on the specific
-parameters, such models encompass commonly-considered ghost archaic admixture
-scenarios (e.g., if a long-isolated lineage more recently contributes a
+parameters, such models encompass commonly-considered scenarios for admixture from a ghost archaic hominin population (e.g., if a long-isolated lineage more recently contributes a
 minority of ancestry to one or more populations), as well as relatively simple
@@
 Denisovan samples. This observation was taken as further evidence for deep
-population structure or archaic admixture in West African populations from an
+population structure or archaic hominin admixture in West African populations from an
 unidentified hominin unrelated to the Neanderthal/Denisovan complex.
@@
 \subsection{What does this mean for expectation of morphology btw branches?}
-
+\label{morpho_diff}
 \aprcomment{Tim to fill in, and edit section header}
@@
 aspects of the data that have been used in other studies as evidence of
-population structure or archaic admixture from an unidentified hominin species
+population structure or archaic hominin admixture from an unidentified hominin species
 within Africa.

### 2022-03-07 paper.tex
edits in response to comments

@@
 
-\comment{About title: "Contrasting archaic admixture and a weakly structured stem for human originas in Africa" Too long? Inconclusive?}
+\comment{About title: "Contrasting archaic admixture and a weakly structured stem for human originas in Africa" Too long? Inconclusive?} \sgcomment{Disagree.}
 
@@
 continent. Progress is hampered by a paucity of fossil and genomic data, as
-well as variability in dating \comment{prior dataing estimates}.
+well as variability in prior dating estimates.
 Here we use linkage disequilibrium and
@@
 is no comparably ancient DNA to elucidate early history in Africa.
-\comment{Any point in citing this paper: https://www.nature.com/articles/s41586-022-04430-9 ?}
+\comment{Any point in citing this paper: https://www.nature.com/articles/s41586-022-04430-9 ?} \sgcomment{Yes, we can literally just cite it here. }
 
@@
         continuous-migration model (A, with $LL=-115,500$ ).
-        \comment{If colors have a meaning, explain; also, it may be easier to change the color scheme of figure 1 to match figure 2}
-    }
+        Colors are used to distinguish overlapping branches and link to Figure \ref{fig:1}.}
     \label{fig:2}
@@
     \caption{
-        \textbf{Predicted population structure over time.}
-        \comment{This could have a mroe descriptive title saying the take-home point}
+        \textbf{Structure among stems is weak and present-day structure is mostly recent}
         From the best fit models of our two parameterizations (A and B:

### 2022-03-08 supplement.tex
comments to supplement

@@
 % local definitions
+\newcommand{\comment}[1]{{\textcolor{red}{Co-author: #1}}} 
 \newcommand{\sgcomment}[1]{{\textcolor{red}{SG: #1}}}
@@
 option for duos and trios. We highlight 82 Nama genomes which are newly
-available (unpublished) under accession number EGAD00001006198. Among these 82
+available (unpublished) under accession number EGAD00001006198.
+\comment{This makes is sound like these weren't generated for this project.}
+Among these 82
 Nama samples, we down-sampled the individuals to minimize close relatives and
@@
 data. Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
-analysis, after partitioning ancestry into $k=6$ clusters with ADMIXTURE
+analysis, after partitioning ancestry into $k=6$ clusters with \texttt{ADMIXTURE}
+\comment{Justification for k=6}
 \citep{Alexander2009-sw} where alternative ancestries represent European, West
@@
 Given two biallelic loci, with alleles $A$ and $a$ and the left locus and
-allels $B$ and $b$ at the right locus, the standard covariance measure of LD is
+alleles $B$ and $b$ at the right locus, the standard covariance measure of LD is
 $D=p_{AB}p_{ab}-p_{Ab}p_{aB}$, where $p_{AB}$ is the frequency of $AB$
@@
 normalized by $\pi_2$ in a reference population (throughout, we use the Mende
+\comment{Why use the Mende to normalize?}
 $\pi_2$), which removes any dependence on the mutation rate. Thus, statistics
@@
 labeled ancestral branches using a parsimony in migration, referring to South,
-East, and West African branches (S/E/W AFR,
-Figures~\ref{fig:1},~\ref{fig:supp-possible-models}). We do not know the
+East, and West African branches
+\comment{would also be interesting to organize by ethnolinguistic group...}
+(S/E/W AFR, Figures~\ref{fig:1},~\ref{fig:supp-possible-models}). We do not know the
 geographical location of these ancestral populations (nor even if they
@@
 \end{itemize}
+\comment{This could be organized with a bit more structure. E.g. by region, or by timepoint}
 While the dates of these events were fixed, the sizes and proportions were
@@
 statistics and inferences between simulated and real data that were not used in
-the original optimization. These comparisons serve as independent validations
+the original optimization.
+\comment{Describe in more detail}
+These comparisons serve as independent validations
 of our inferred demographic models by assessing how well those models fit
@@
 generation \citep{Sasani2019-zp,Tian2019-zv}. A rate of $1.3\times10^{-8}$ also
-approximates the rate predicted by a mutational model from Gnomad
+approximates the rate predicted by a mutational model from gnomAD
 \citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
@@
 populations, closely matching the average mutation rate estimated from the
-Gnomad mutation model. The other inferred demographic models
+gnomAD mutation model. The other inferred demographic models
 (Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-without-stem-migration})
@@
 \clearpage
+\comment{30 is a lot of supp figures... can these be condensed to be a bit less daunting to readers?}
 
@@
         \textbf{Proposed models of early human history in Africa.}
+        \comment{Lovely figure, may even want to move to main text}
     }

### 2022-03-08 supplement.tex
edits to figure captions and supp. cleanup

@@
 $\pi_{2;i,j,k,l}=p_{A,i}(1-p_{A,j})p_{B,k}(1-2p_{B,l})$. We consider statistics
-normalized by $\pi_2$ in a reference population (throughout, we use the Mende
-\comment{Why use the Mende to normalize?}
+normalized by $\pi_2$ in an arbitrary reference population (throughout, we use the Mende
 $\pi_2$), which removes any dependence on the mutation rate. Thus, statistics
@@
 simulate sampled individual genomes to match the data used in inference (see
-simulation details below). Using these simulations, we compared additional
-statistics and inferences between simulated and real data that were not used in
-the original optimization.
-\comment{Describe in more detail}
+simulation details below). We used these simulations to compare our model predictions to 
+predictions based on the conditional SFS (cSFS) and on relate curves (see below).
+
 These comparisons serve as independent validations
@@
     \label{tab:supp-continuous-migration-no-mig}
-    \textbf{Best-fit parameters from the Continuous-Migration model with isolation
-    between stems.}
+    \textbf{Best-fit parameters from the Continuous-Migration model without stem migration.}
     Inferred values are scaled to physical units assuming a generation time of
@@
 \clearpage
-\comment{30 is a lot of supp figures... can these be condensed to be a bit less daunting to readers?}
 
@@
     \caption{
-        \textbf{Proposed models of early human history in Africa.}
+        \textbf{Models of early human history in Africa} seeking to reconcile genetic and paleoanthropological evidence \cite{Henn2018-rf}.
          \comment{Lovely figure, may even want to move to main text}
@@
     \includegraphics{figures/supp-omni-hapmap-comparison.pdf}
-    \caption{
+    \caption{\sgcomment{$\pi$ in figure are not rendered properly}
         \textbf{Comparison of LD statistics parsed using two recombination maps}
@@
         Predicted vs observed two-locus statistics as a function of genetic
-        distance for the continuous-migration model. Each panel represents a different
+        distance for the continuous-migration model \sgcomment{with or without migration?}. Each panel represents a different
         set of two-locus statistics. Solid lines represent estimates using the
@@
     \includegraphics[width=\textwidth]{figures/supp-fsts-all-models.pdf}
-    \caption{
+    \caption{\sgcomment{Need ABCD in figure!}
         \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
@@
     \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
-    \caption{
+    \caption{\sgcomment{can we sync with text colors? Need ABCD in figure}
         \textbf{Predicted pairwise differentiation between contemporaneous
-        populations.} $H_{i,j}$ is the predicted pairwise differentiaion
+        populations.} $H_{i,j}$ is the predicted pairwise differentiation
         between two populations per base-pair, using a mutation rate estimated
@@
         and ancient branches from the single-origin model.}
-        $f_4(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
+        $f_4^2(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
         amount of drift between sampled populations $A$ and $B$ that aligns with
@@
         (Fig.~\ref{fig:supp-FST-panels}), differentiation due to that
-        ancient structure does not contribute to differention between modern
+        ancient structure contributes little to differentiation between modern
         populations.

### 2022-03-08 paper.tex
edits to figure captions and supp. cleanup

@@
 similar.
-Despite genetic similarity between these populations, an inferred 1--4\%
+Despite genetic similarity between these populations, an inferred 1--4\%\sgcomment{explain in the main text where this estimate is from!}
 of genetic differentiation among contemporary human populations can be

### 2022-03-08 paper.tex
address some comments

--- 2022-03-07 paper.tex
@@
 \begin{abstract}
-While it is now broadly accepted that \emph{Homo sapiens} origined within Africa, considerable
+While it is now broadly accepted that \emph{Homo sapiens}
+origined within Africa, considerable
 uncertainty surrounds specific models of divergence and migration across the
@@
 population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
-population divergence among contemporary populations occurs 120-135kya\tdwcomment{The standard notation for anyone who studies the past (e.g., paleoanthropologists, geologists) for thousands of years ago is ka, and Ma for millions of years ago.
-Given this, I think we should use these throughout the paper.
-I can make the changes, but I didn't want to do so without discussing them}, between
+population divergence among contemporary populations occurs 120-135ka, between
 the Khoe-San and other groups. Prior to the divergence of contemporary African
@@
 
-Archaeological sites from the Middle Stone Age (approx. 300kya-40kya) are widely
+Archaeological sites from the Middle Stone Age (approx. 300ka-40ka) are widely
 distributed across Africa, and are particularly well represented in the
@@
 \emph{Homo sapiens} by which populations in many regions of the continent
-contributed to the formation of Homo sapiens beginning at least 300kya
+contributed to the formation of Homo sapiens beginning at least 300ka
 \citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
@@
 population, possible links to specific fossils, and the possibility of
-finding ancient DNA evidence (e.g., \citet{Hsieh2016-gk}). However, these prior
+finding ancient DNA evidence \citep{Hsieh2016-gk}). However, these prior
 articles share two weaknesses. First, they only contrast a single-origin model
 with an archaic hominin admixture model, leaving out other plausible models
-(Figure~\ref{fig:supp-possible-models} and \citet{Henn2018-rf}). Second, they
+\citep{Henn2018-rf} (Figure~\ref{fig:supp-possible-models}).
+Second, they
 focus on a small subset of African diversity, either because of small sample
@@
 for gene flow from Neanderthals into non-Africans and gauge the relative
-time depth of divergence, assuming Neanderthals diverged 550kya from a common
+time depth of divergence, assuming Neanderthals diverged 550ka from a common
 stem. For each population, we computed low-order
@@
 Regardless of the model choice for early epochs, inference of human demographic
-history for the last 150kya was remarkably robust. The earliest divergence among
+history for the last 150ka was remarkably robust. The earliest divergence among
 contemporary human populations differentiates the southern African Nama from
-other African groups between 110--135kya, with low to moderate levels of
+other African groups between 110--135ka, with low to moderate levels of
 subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
 high-likelihood models which we explored did the divergence between Nama and
-other populations exceed $\sim$140kya. 
+other populations exceed $\sim$140ka. 
 We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
@@
 the time of divergence between Western and Eastern African populations,
-set to 60kya, just prior to the split of Eurasians and East Africans set to 50kya. 
+set to 60ka, just prior to the split of Eurasians and East Africans set to 50ka. 
 We also fixed the amount of admixture from Neanderthals to Europeans directly
-following the out-of-Africa migration which was set to 1.5\% at 45kya (Supp. Information). These constraints allowed us to integrate information from previous genetic and archaelogical research,
+following the out-of-Africa migration which was set to 1.5\% at 45ka (Supp. Information). These constraints allowed us to integrate information from previous genetic and archaelogical research,
 to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
@@
         & & & 
-            \textbf{Time (kya)} & \textbf{per generation} & \textbf{duration (ky)} \\
+            \textbf{Time (ka)} & \textbf{per generation} & \textbf{duration (ky)} \\
         \midrule
@@
         In both models, the branch ancestral to the Nama shares a common ancestral population 
-        with the other African groups $\sim$120--135kya. Following this divergence, 
+        with the other African groups $\sim$120--135ka. Following this divergence, 
         the population ancestral to other African groups branches into West and East African
-        groups 60kya. *Migration rates and durations are shown between branches ancestral to 1) Nama and East Africans
+        groups 60ka. *Migration rates and durations are shown between branches ancestral to 1) Nama and East Africans
         and their ancestors, and 2) Nama and Mende, respectively.
@@
 
-To account for population structure prior to 135kya, three of our four models allowed for two or more
+To account for population structure prior to 135ka, three of our four models allowed for two or more
 ``stem'' populations which could diverge either before or after the Neanderthal
@@
 continuous migration). With continuous migration between stems, population
-structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
+structure extends back to 1.1--1.4Ma (Table~\ref{tab:migration-rates}).
 Migration between the stems in these models is moderate, with a fraction of
@@
 to inferred migration rates between connected contemporary populations over the
-past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
+past 50ka (Table~\ref{tab:migration-rates}). This ongoing (or at least,
 periodic) gene flow qualitatively distinguishes these models from previously
@@
 ($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
-allowed to occur until 5kya. A
+allowed to occur until 5ka. A
 sampled lineage from the Nama, Mende, and Gumuz have probabilities of being in
@@
 exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
-479kya. The timing of this divergence was also poorly constrained ($\sigma_T=
-166$kya). These populations evolve independently until approximately 119kya
+479ka. The timing of this divergence was also poorly constrained ($\sigma_T=
+166$ka). These populations evolve independently until approximately 119ka
 when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
@@
 Out of Africa event). Finally, the Mende receive a large additional pulse of
-gene flow from Stem 2, replacing 18\% of their population 25kya.
+gene flow from Stem 2, replacing 18\% of their population 25ka.
 The later Stem 2 contribution to the Western African Mende resulted in
@@
         contemporary genetic differentiation.
-        Most present differentiation dates back to the last 100kya.
+        Most present differentiation dates back to the last 100ka.
     }
@@
 even though this statistic was not used in the fit. Both models actually outperform
-archaic hominin admixture models fit directly to the cSFS (for example, compare with Figure~1 in
-\citet{Durvasula2020-td}).
+archaic hominin admixture models fit directly to the cSFS
+(for example, compare with Figure~1 in
+Durvasula and Sankararaman (2020)\citep{Durvasula2020-td}).
 
@@
 previous studies have found a reduction of coalescence rates 
-between $1000$kya and $100$kya in humans, and thus inferred an increase 
+between $1000$ka and $100$ka in humans, and thus inferred an increase 
 in $N_e$ during the same period \citep{Li2011-le}.
@@
 single-origin model, recapitulate an inferred ancestral
-increase in $N_e$ between 100kya-1Mya (Figure \ref{fig:supp-iicr-sim}). Whereas the single-origin model achieves this by
+increase in $N_e$ between 100ka-1Ma (Figure \ref{fig:supp-iicr-sim}). Whereas the single-origin model achieves this by
 an increase in $N_e$ during that period, the best-fit model recapitulate this pattern without corresponding population size changes.      
@@
 Published estimates of the earliest human divergences with rCCR, which range
-from 150kya-100kya \citep{Bergstrom2021-iw}, may be significantly biased when
+from 150ka-100ka \citep{Bergstrom2021-iw}, may be significantly biased when
 compared to more complex models with gene flow as inferred here. We find that
@@
 underestimating divergence time by 50\% or greater (e.g., Mende vs. Gumuz
-$\sim$15kya compared to a true divergence of 60kya), and recent migration can
+$\sim$15ka compared to a true divergence of 60ka), and recent migration can
 lead to the misordering of divergence events (Figure~\ref{fig:4}E). We suggest
@@
 contemporary groups. We observed two merger events. The first, between Stem 1S and
-Stem 2, results in the formation of an ancestral Khoe-San population ~120kya.
-The second $\approx100$kya between Stem 1E and Stem 2, results in the formation of the
+Stem 2, results in the formation of an ancestral Khoe-San population ~120ka.
+The second $\approx100$ka between Stem 1E and Stem 2, results in the formation of the
 ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
@@
 Following these merger events, the stems subsequently fracture into
-subpopulations which then appear to persist over the past $\sim120$kya. These
+subpopulations which then appear to persist over the past $\sim120$ka. These
 subpopulations can be linked to contemporary groups despite subsequent gene flow
@@
 $0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
-(Stem 1E) 150kya versus $0.03$ probability of being inherited from the ‘southern’
+(Stem 1E) 150ka versus $0.03$ probability of being inherited from the ‘southern’
 subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2
@@
 regions \citep{Petr2019-xo} or an increased frequency of archaic-hominin-related
-haplotypes such as at \emph{EPAS1} among Tibetans (e.g.,
-\citet{Zhang2021-xx}). Under previous
+haplotypes such as at \emph{EPAS1} among Tibetans \citep{Zhang2021-xx}. Under previous
 African archaic hominin admixture models, the estimated 8--10\% introgression rate is
@@
 differentiation among human populations and between humans and Neanderthals
-\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also Section \ref{morpho_diff}).
+\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}
+(see also Section \ref{morpho_diff}).
 This correspondence allows predictions of our model to be related to the fossil record.
-The fossil
-record of Africa is sparse during the time period of the stems, but of the
-available fossils, some are very similar to contemporary humans (e.g., from Omo
-Kibish, Ethiopia\tdwcomment{add citations: Day (1969) doi.org/10.1038/2221135a0 and Vidal et al. (2022) doi.org/10.1038/s41586-021-04275-8}), others are similar in some features but not others (e.g.,
-from Jebel Irhoud, Morocco\tdwcomment{add citations: Hublin et al. (2017) doi.org/10.1038/nature22336 and Richter et al. (2017) doi.org/10.1038/nature22335}), and others are very different (e.g., from
-Dinaledi, South Africa\tdwcomment{add citations: Berger et al.
-(2015) doi.org/10.7554/eLife.09560 and Dirks et al.
-(2017) doi.org/10.7554/eLife.24231.001}) in morphology.
+The fossil record of Africa is sparse during the time period of the stems, but of the
+available fossils, some are very similar to contemporary humans
+(e.g., from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}),
+others are similar in some features but not others
+(e.g., from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}),
+and others are very different
+(e.g., from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}) in morphology.
 If, as our model predicts, the
@@
 ancestries.
-\comment{this needs more analyses/citations to back up if included}\aprcomment{but this comment was on the previous version}
 
@@
 
-\bibliographystyle{unsrtnat}
+\bibliographystyle{naturemag}
 \bibliography{paper}

### 2022-03-08 supplement.tex
address some comments

--- 2022-03-08 supplement.tex
@@
         reconstructed from simulated data using Relate match those inferred
-        from the full dataset (\label{fig:supp-rccr-data}, left) for distant
+        from the full dataset (\ref{fig:supp-rccr-data}, left) for distant
         timepoints. There are large discrepancies in the very recent history,

### 2022-03-08 paper.tex
address some comments

@@
 \comment{What about the LWK?}
-While ancient DNA from
-Eurasia has helped us understand early human history outside of Africa, there
-is no comparably ancient DNA to elucidate early history in Africa.
-\comment{Any point in citing this paper: https://www.nature.com/articles/s41586-022-04430-9 ?} \sgcomment{Yes, we can literally just cite it here. }
+While ancient DNA from Eurasia has helped us understand early human history
+outside of Africa, there is no comparably ancient DNA to elucidate early history
+in Africa \citep{Lipson2022-xf}.
 
@@
 \comment{what do you mean by low-order? not a commonly used phrase}
+\aprcomment{I think this is a fairly common term in statistics, right?}
 allele frequency and linkage
@@
 differentiation of contemporary groups, including models allowing for ancestral
-reticulation (Figure~\ref{fig:supp-possible-models}B), archaic hominin admixture
+reticulation (such as fragmentation-and-coalescence or meta-population models,
+\aprcomment{defining reticulation at first mention -- something better here?}
+Figure~\ref{fig:supp-possible-models}B), archaic hominin admixture
 (Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
@@
         the population ancestral to other African groups branches into West and East African
-        groups 60ka. *Migration rates and durations are shown between branches ancestral to 1) Nama and East Africans
-        and their ancestors, and 2) Nama and Mende, respectively.
-        `Divergence times' correspond to the most recent common ancestral population, and does not account for continuous migration or earlier reticulations.
+        groups 60ka.
+        *Migration rates and durations are shown between branches ancestral to 
+        1) Nama and East Africans and their ancestors, and
+        2) Nama and Mende, respectively.
+        ``Divergence times'' correspond to the most recent common ancestral population,
+        and does not account for continuous migration or earlier reticulations.
     }
@@
 single-origin model, recapitulate an inferred ancestral
-increase in $N_e$ between 100ka-1Ma (Figure \ref{fig:supp-iicr-sim}). Whereas the single-origin model achieves this by
-an increase in $N_e$ during that period, the best-fit model recapitulate this pattern without corresponding population size changes.      
-\comment{These are pretty complicated ideas that are not explained in detail. Might be worth unpacking a bit more to make sure the reader understands}
+increase in $N_e$ between 100ka-1Ma (Figure \ref{fig:supp-iicr-sim}).
+Whereas the single-origin model achieves this by
+an increase in $N_e$ during that period, the best-fit model recapitulate
+this pattern without corresponding population size changes.      
+%\comment{These are pretty complicated ideas that are not explained in detail.
+%Might be worth unpacking a bit more to make sure the reader understands}
 
@@
 isolation among the stems is variable across model fits, models with a period of 
-divergence, isolation and then a merger event (i.e. a ``reticulation'' \comment{define at the first mention})
+divergence, isolation and then a merger event (i.e. a ``reticulation''
 out-performed models with bifurcating divergence and continuous gene flow.
@@
 \citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}
-(see also Section \ref{morpho_diff}).
+(see also Section \ref{sec:orpho_diff}).
 This correspondence allows predictions of our model to be related to the fossil record.

### 2022-03-10 supplement.tex
address some comments
Update supplement.tex
Added section about relating model predictions to morphological variation

@@
 
-\subsection{What does this mean for expectation of morphology btw branches?}
-\label{morpho_diff}
-\aprcomment{Tim to fill in, and edit section header}
+\subsection{Relating model predictions to morphological variation} \label{sec:morpho_diff}
+
+Here we show how the predictions of our model can be related to morphological variation in the fossil record.
+Let $\Delta^2$ be the expected squared difference in a morphological measurement between fossil 1 and fossil 2.
+
+\begin{equation}
+	\Delta^2=\mathrm{E}\left\{\left(\bar{X}_1-\bar{X}_2\right)^2\right\}+\mathrm{E}\left\{V_1\right\}+\mathrm{E}\left\{V_2\right\}
+\end{equation}
+
+where $\bar{X}_1$ and $V_1$ and $\bar{X}_2$ and $V_2$ are measurement means and variances for the populations from which fossil 1 and fossil 2 are sampled.
+If we assume that the trait being measured is evolving neutrally, which seems reasonable as an approximation based on studies of morphological (cranial) variation in humans and Neanderthals \tdwcomment{cite here: main text references 36,37, and 38}, then
+
+\begin{equation}
+	\mathrm{E}\left\{\left(\bar{X}_1-\bar{X}_2\right)^2\right\}=4\sigma_m^2\tau_B-2\sigma_m^2\tau_{W_1}-2\sigma_m^2\tau_{W_2}
+\end{equation}
+
+\begin{equation}
+\mathrm{E}\left\{V_A\right\}=\frac{1}{h^2}\sigma_m^2\tau_{W_1}
+\end{equation}
+
+\begin{equation}
+\mathrm{E}\left\{V_B\right\}=\frac{1}{h^2}\sigma_m^2\tau_{W_2}
+\end{equation}
+
+where $\sigma^2_m$ is the additive genetic variance introduced by mutation (per zygote per generation), $h^2$ is the (narrow sense) heritability, $\tau_{W_1}$ and $\tau_{W_2}$ are the average coalescence time of pairs of alleles from the same population and $\tau_B$ is the average coalescence time of pairs of alleles from different populations \tdwcomment{cite here: Whitlock 1999 doi.org/10.1017/S0016672399004127 and Weaver 2016 doi.org/10.1002/ece3.2522}.
+Substituting and simplifying gives
+
+\begin{equation}
+	\Delta^2=\sigma_m^2 \left( 4\tau_B+\frac{1-2h^2}{h^2} \left( \tau_{W_1}+\tau_{W_2} \right) \right)
+\end{equation}
+
+which relates $\Delta^2$ to between and within population coalescence times. As a further approximation, if we assume that the heritability of the measurement is close to one half, then
+
+\begin{equation}
+	\Delta^2\approx4\sigma_m^2\tau_B
+\end{equation}
+
+which shows that under these assumptions $\Delta^2$ is proportional to $\tau_B$.
+
+Nucleotide diversity, $\pi$, for pairs of individuals sampled from different populations is also proportional to $\tau_B$.
+Therefore, ratios of $\Delta^2$ can be predicted from ratios of $\pi$ for the same comparisons.
+For example, a ratio for which $\pi$ for stem 1 and stem 2 is the numerator and $\pi$ for a pair of contemporary human populations is the denominator can be used to predict $\Delta^2$ for two fossils sampled from each of the two stems based on $\Delta^2$ for contemporary humans.
+In this way, the predictions of our model can be related to morphological variation in the fossil record.
 

### 2022-03-10 supplement.tex
Update supplement.tex

@@
 \end{equation}
-
 where $\bar{X}_1$ and $V_1$ and $\bar{X}_2$ and $V_2$ are measurement means and variances for the populations from which fossil 1 and fossil 2 are sampled.
@@
 \end{equation}
-
 where $\sigma^2_m$ is the additive genetic variance introduced by mutation (per zygote per generation), $h^2$ is the (narrow sense) heritability, $\tau_{W_1}$ and $\tau_{W_2}$ are the average coalescence time of pairs of alleles from the same population and $\tau_B$ is the average coalescence time of pairs of alleles from different populations \tdwcomment{cite here: Whitlock 1999 doi.org/10.1017/S0016672399004127 and Weaver 2016 doi.org/10.1002/ece3.2522}.
@@
 \end{equation}
-
 which relates $\Delta^2$ to between and within population coalescence times. As a further approximation, if we assume that the heritability of the measurement is close to one half, then
@@
 \end{equation}
-
 which shows that under these assumptions $\Delta^2$ is proportional to $\tau_B$.

### 2022-03-10 supplement.tex
Update supplement.tex
addressing loose ends

--- 2022-03-08 supplement.tex
@@
 option for duos and trios. We highlight 82 Nama genomes which are newly
-available (unpublished) under accession number EGAD00001006198.
+available (and previously unpublished) under accession number EGAD00001006198.
 \comment{This makes is sound like these weren't generated for this project.}
-Among these 82
-Nama samples, we down-sampled the individuals to minimize close relatives and
-admixture, such that 44 Nama genomes were retained. 2nd and 3rd degrees
-relatives were inferred from reconstructed pedigrees with Omni2.5 SNP array
-data. Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
+Among these 82 Nama samples, we down-sampled the individuals to minimize
+close relatives and admixture, such that 44 Nama genomes were retained.
+Second- and third-degrees relatives were inferred from reconstructed
+pedigrees with Omni2.5 SNP array data.
+Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
 analysis, after partitioning ancestry into $k=6$ clusters with \texttt{ADMIXTURE}
@@
 
-\sgcomment{To be expanded. TODO: Brenna}
-\begin{itemize}
-    \item From the merged dataset of the African Diversity Reference Panel and
-        1000 Genomes phase 3 data, subsampled to population included in this study
-    \item West Africa: Mende from Sierra Leone (MSL); South Africa: newly sequenced 
-        Nama; East Africa: Gumuz (traditionally hunter-gatherer with low levels of 
-        Eurasian admixture), Oromo and Amhara (combined as traditionally 
-        agriculturalists with large proportion of back-to-Africa Eurasian ancestry); 
-        Eurasian: British (GBR)
-    \item Combined with the high coverage Vindija neanderthal genome
-    \item For running \Relate, we used a larger set of populations. We kept all 
-        African 1000 Genomes populations, along with the Nama, Gumuz, Oromo, and
-        Amhara, as well as 4 Eurasian populations from 1000 Genomes: GBR, CEU, PJL,
-        and CHB
-\end{itemize}
+From the merged dataset of the African Diversity Reference Panel (ADRP,
+citet{}) and the \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data,
+we subsampled to the populations included in this study. From the 1000 Genomes
+data, these included all populations in the African super-population (ESN, GWD,
+LWK, MSL, and YRI) and four non-African populations (CEU, GBR, CHB, and PJL).
+From the ADRP, these included the Amhara, Bakiga, Gumuz, Nama, Oromo, Somali,
+Nama, and Zulu. This larger set of populations were used in the PCA and
+\texttt{Admixture} analyses. The \texttt{Relate} analysis used a slightly
+smaller set, discluding the Bakiga, Somali, and Zulu samples.
+
+For the demographic modeling, we further subset to the regional populations,
+from West Africa: Mende from Sierra Leone (MSL); South Africa: Nama from South
+Africa (newly sequenced); East Africa: Gumuz (traditionally hunter-gatherer
+with low levels of Eurasian admixture), Oromo and Amhara (combined as
+traditionally agriculturalists with large proportion of back-to-Africa Eurasian
+ancestry); and Europe: British (GBR). These were combined with the
+high-coverage Vindija Neanderhal genome \citep{Prufer2017-kk}, resulting in 290
+total samples used in demographic inference.
 
@@
 When testing multiple variations of the more complex models, we kept these
-values fixed (see Table~\ref{tab:fixed_params} for a complete list of all fixed
+values fixed (see Table~\ref{tab:fixed-params} for a complete list of all fixed
 parameters). This reduced the potential for overfitting the more complex
@@
 
-\subsection{What does this mean for expectation of morphology btw branches?}
-\label{morpho_diff}
-\aprcomment{Tim to fill in, and edit section header}
+\subsection{Relating model predictions to morphological variation}
+\label{sec:morpho_diff}
+
+Here we show how the predictions of our model can be related to morphological
+variation in the fossil record.
+Let $\Delta^2$ be the expected squared difference in a morphological measurement
+between fossil 1 and fossil 2,
+\begin{equation}
+    \Delta^2=\E\left[\left(\bar{X}_1-\bar{X}_2\right)^2\right]+\E\left[V_1\right]+\E\left[V_2\right],
+\end{equation}
+where $\bar{X}_1$ and $V_1$ and $\bar{X}_2$ and $V_2$ are measurement means and
+variances for the populations from which fossil 1 and fossil 2 are sampled.
+
+If we assume that the trait being measured is evolving neutrally, which seems
+reasonable as an approximation based on studies of morphological (cranial)
+variation in humans and Neanderthals
+\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}, then
+\begin{equation}
+    \E\left[\left(\bar{X}_1-\bar{X}_2\right)^2\right]=4\sigma_m^2\tau_B-2\sigma_m^2\tau_{W_1}-2\sigma_m^2\tau_{W_2},
+\end{equation}
+\begin{equation}
+    \E\left[V_A\right]=\frac{1}{h^2}\sigma_m^2\tau_{W_1},
+\end{equation}
+and
+\begin{equation}
+    \E\left[V_B\right]=\frac{1}{h^2}\sigma_m^2\tau_{W_2},
+\end{equation}
+\aprcomment{switch between 1/2 and A/B - is that intentional, or a typo?}
+where $\sigma^2_m$ is the additive genetic variance introduced by mutation
+(per zygote per generation), $h^2$ is the (narrow sense) heritability,
+$\tau_{W_1}$ and $\tau_{W_2}$ are the average coalescence time of pairs
+of alleles from the same population and $\tau_B$ is the average coalescence
+time of pairs of alleles from different populations
+\citep{Whitlock1999-un,Weaver2016-id}.
+Substituting and simplifying gives
+\begin{equation}
+    \Delta^2=\sigma_m^2 \left( 4\tau_B+\frac{1-2h^2}{h^2} \left( \tau_{W_1}+\tau_{W_2} \right) \right),
+\end{equation}
+which relates $\Delta^2$ to between and within population coalescence times.
+
+As a further approximation, if we assume that the heritability of the
+measurement is close to one half, then
+\begin{equation}
+    \Delta^2\approx4\sigma_m^2\tau_B,
+\end{equation}
+which shows that under these assumptions $\Delta^2$ is proportional to $\tau_B$.
+
+Nucleotide diversity, $\pi$, for pairs of individuals sampled from different
+populations is also proportional to $\tau_B$.
+Therefore, ratios of $\Delta^2$ can be predicted from ratios of $\pi$
+for the same comparisons.
+For example, a ratio for which $\pi$ for Stem 1 and Stem 2 is the numerator
+and $\pi$ for a pair of contemporary human populations is the denominator
+can be used to predict $\Delta^2$ for two fossils sampled from each of the
+two stems based on $\Delta^2$ for contemporary humans.
+In this way, the predictions of our model can be related to morphological
+variation in the fossil record.
 
@@
     \includegraphics{figures/supp-omni-hapmap-comparison.pdf}
-    \caption{\sgcomment{$\pi$ in figure are not rendered properly}
+    \caption{
         \textbf{Comparison of LD statistics parsed using two recombination maps}
@@
         Predicted vs observed two-locus statistics as a function of genetic
-        distance for the continuous-migration model \sgcomment{with or without migration?}. Each panel represents a different
+        distance for the continuous-migration model with migration between stems.
+        Each panel represents a different
         set of two-locus statistics. Solid lines represent estimates using the
@@
     \includegraphics[width=\textwidth]{figures/supp-fsts-all-models.pdf}
-    \caption{\sgcomment{Need ABCD in figure!}
+    \caption{
         \textbf{Predicted \bm{$F_{ST}$} between contemporaneous populations over time.}
@@
     \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
-    \caption{\sgcomment{can we sync with text colors? Need ABCD in figure}
+    \caption{\sgcomment{can we sync with text colors?}
         \textbf{Predicted pairwise differentiation between contemporaneous

### 2022-03-10 supplement.tex
Modified morphology section

@@
 \begin{equation}
-    \E\left[V_A\right]=\frac{1}{h^2}\sigma_m^2\tau_{W_1},
+    \E\left[V_1\right]=\frac{1}{h^2}\sigma_m^2\tau_{W_1},
 \end{equation}
@@
 \begin{equation}
-    \E\left[V_B\right]=\frac{1}{h^2}\sigma_m^2\tau_{W_2},
+    \E\left[V_2\right]=\frac{1}{h^2}\sigma_m^2\tau_{W_2},
 \end{equation}
@@
 $\tau_{W_1}$ and $\tau_{W_2}$ are the average coalescence time of pairs
-of alleles from the same population and $\tau_B$ is the average coalescence
+of alleles from the same population, and $\tau_B$ is the average coalescence
 time of pairs of alleles from different populations

### 2022-03-10 supplement.tex
Fixed the typo

@@
 \end{equation}
-\aprcomment{switch between 1/2 and A/B - is that intentional, or a typo?}
 where $\sigma^2_m$ is the additive genetic variance introduced by mutation

### 2022-03-10 supplement.tex
edits to captions up to S22

@@
 \subsubsection{Conditional SFS}
-
+\label{sec:cSFS}
 The single-origin model fit the cSFS relatively poorly across the three
@@
     \caption{
-        \textbf{Conditional SFS compared to single-origin model.}
-        The single-origin model provides a poor fit to the observed cSFS in
-        the Mende, Gumuz, and Nama. Ancestral state misidentification
+        \textbf{Conditional SFS compared to single-origin model predictions.}
+        The single-origin model provides a poor fit to the Neanderthal-conditional SFS in
+        the Mende, Gumuz, and Nama (see Section \ref{sec:cSFS}). Ancestral state misidentification
         (labeled as $p_m$ in the figure legends) was
@@
     \caption{
-        \textbf{Conditional SFS compared to continuous-migration model.}
-        The continuous-migration model provides a better fit to the data than
+        \textbf{Conditional SFS compared to continuous-migration model predictions.}
+        The continuous-migration model provides a better fit to the Neanderthal-conditional SFS  than
         either the single-origin model or the merger-without-stem-migration
@@
     \caption{
-        \textbf{Conditional SFS compared to merger-without-stem-migration model.}
+        \textbf{Conditional SFS compared to merger-without-stem-migration model predictions.}
         The excess of intermediate-frequency variants in the Nama and Mende
         cSFS results in a poorer fit to the cSFS compared to the continuous-migration
-        and merger-with-stem-migratino models. Along with the poorer likelihood this
-        model provides to the two-locus statistics, validation usin the cSFS
-        disfavors a model with strict isolation between stem populations.
-        Ancestral state misidentification ($p_m$), as estimated
-        with all models, is roughly doubled for transition-type mutations
-        compared to transversions.
+        and merger-with-stem-migratino models. Thus the model without stem migrations
+        provides a poorer fit 
+        to both two-locus statistics and the cSFS.
+        Ancestral state misidentification for transition-type mutations ($p_m$), as estimated
+        with all models, is roughly double that of transversions.
     }
@@
     \caption{
-        \textbf{Conditional SFS compared to merger-with-stem-migration model.}
+        \textbf{Conditional SFS compared to merger-with-stem-migration model predictions.}
         The merger-with-stem-migration model, in addition to providing the best
@@
         inferences in humans using the Thousand Genomes data and the 6-primate
-        alignment (see Methods). Ancestral state misidentification ($p_m$), as estimated
-        with all models, is roughly doubled for transition-type mutations
-        compared to transversions.
+        alignment (see Methods). 
     }
@@
         from the full dataset for these populations in the distant past, there
-        are noticeable discrepancies over the recent history in the last 10s of
+        are noticeable discrepancies over the recent history in the last tens of
         thousands of years. Left: Relate run on full dataset, showing focal
         populations. Right: Relate run directly on the subset dataset of focal
-        populations.
+        populations. \sgcomment{should the labels on right and left panels be identical?}
     }
@@
         (Figs.~\ref{fig:supp-iicr-data-subset} and \ref{fig:supp-rccr-data}),
-        it is difficult to draw firm conclusions from qualitative comparisons
+        it is difficult to draw firm conclusions from quantitative comparisons
         between reconstructed genealogies.
@@
         \textbf{Relative cross-coalescence rates inferred from the data.} Using
-        the same reconstructed genealogies from Relate, we computed RCCR curves
+        the same \sgcomment{same was what? Figure S19 presumably?} reconstructed genealogies from Relate, we computed RCCR curves
         for pairs of populations using genealogies from the full dataset

### 2022-03-10 supplement.tex
final comments and edits to supplementary captions

@@
         \textbf{Relative cross-coalescence rates inferred from the data.} Using
-        the same \sgcomment{same was what? Figure S19 presumably?} reconstructed genealogies from Relate, we computed RCCR curves
+        the same \sgcomment{same was what? Figure S19 presumably?} reconstructed genealogies from Relate, 
+        we computed relative cross-coalescence (RCCR) curves
         for pairs of populations using genealogies from the full dataset
-        ($\sim1,200$ individuals) and the subset of focal populations (290
-        individuals). Despite the overlap in samples, the two sets of inferred
+        ($\sim1,200$ individuals, \sgcomment{top?}) and the subset of focal populations (290
+        individuals, \sgcomment{bottom?}). Despite the overlap in samples, the two sets of inferred
         genealogies provided inconsistent RCCR, with both the timing and
@@
         reconstructed from simulated data using Relate match those inferred
-        from the full dataset (\ref{fig:supp-rccr-data}, left) for distant
+        from the full dataset (\ref{fig:supp-rccr-data}, left \sgcomment{top?}) for distant
         timepoints. There are large discrepancies in the very recent history,
@@
         \textbf{Expected relative cross-coalescence rates from inferred models.}
+        \sgcomment{I think we need model labels for the four rows }
         Our inferred demographic models allow for exact calculation of expected

### 2022-03-10 supplement.tex
update some figures
address comments on supp figures

--- 2022-03-10 supplement.tex
@@
     \includegraphics[width=\textwidth]{figures/supp-h12s-all-models.pdf}
-    \caption{\sgcomment{can we sync with text colors?}
+    \caption{
         \textbf{Predicted pairwise differentiation between contemporaneous
@@
         populations. Right: Relate run directly on the subset dataset of focal
-        populations. \sgcomment{should the labels on right and left panels be identical?}
+        populations.
     }
@@
         \textbf{Relative cross-coalescence rates inferred from the data.} Using
-        the same \sgcomment{same was what? Figure S19 presumably?} reconstructed genealogies from Relate, we computed RCCR curves
-        for pairs of populations using genealogies from the full dataset
-        ($\sim1,200$ individuals) and the subset of focal populations (290
-        individuals). Despite the overlap in samples, the two sets of inferred
+        the same reconstructed genealogies from Relate as displayed in
+        Fig.~\ref{fig:supp-iicr-data-subset}, 
+        we computed relative cross-coalescence (RCCR) curves
+        for pairs of populations using genealogies from the either the full dataset
+        ($\sim1,200$ individuals, top) and the subset of focal populations (290
+        individuals, bottom). Despite the overlap in samples, the two sets of inferred
         genealogies provided inconsistent RCCR, with both the timing and
@@
         reconstructed from simulated data using Relate match those inferred
-        from the full dataset (\ref{fig:supp-rccr-data}, left) for distant
+        from the full dataset (\ref{fig:supp-rccr-data}, top) for distant
         timepoints. There are large discrepancies in the very recent history,

### 2022-03-10 paper.tex
Update paper.tex
bold labels in figure 1

@@
         illustrate diversity from western, eastern and southern Africa.
-        We chose representatives from each region to build parameterized models,
+        We chose representatives from each region (bold labels)
+        to build parameterized models,
         including the newly-sequenced Nama from South Africa, Mende
@@
 \citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}
-(see also Section \ref{sec:orpho_diff}).
+(see also Section \ref{sec:morpho_diff}).
 This correspondence allows predictions of our model to be related to the fossil record.

### 2022-03-11 supplement.tex
Make terminology about human, hominin unambiguous

@@
 hominin admixture. The simplest model, in terms of number of parameters, was a
-single-origin expansion of modern humans, with no structure in the stem and no
+single-origin expansion of humans, with no structure in the stem and no
 archaic hominin admixture aside from the Neanderthal admixture in Eurasian populations
 following the out-of-Africa migration. This model allowed for a population size
-change in the stem of modern humans between the ancestral split of the
+change in the stem of humans between the ancestral split of the
 human-Neanderthal lineages and the more recent split of branches leading to
@@
 migration” model, a secondary stem population (stem 2) split from the primary
-stem (stem 1) that leads to modern humans. Stem 1 contributed to present-day
+stem (stem 1) that leads to contemporary humans. Stem 1 contributed to present-day
 populations via a series of population splits similar to the single-origin
@@
 population. In all inferred models, $F_{ST}$s between Neanderthals and the
-ancestors of modern humans were much larger than between any pair of modern
-human populations or their ancestors (stems). $F_{ST}$ between modern human
+ancestors of contemporary humans were much larger than between any pair of contemporary
+human populations or their ancestors (stems). $F_{ST}$ between human
 populations over the last $100ka$ remained relatively low, at $\sim0.1$ or
@@
 Whereas the high $F_{ST}$ value between Neanderthal and humans reflects both
-elevated genetic differences between the genomes of modern humans and
-neanderthal and reduced diversity within the Neanderthal population
+elevated genetic differences between the genomes of humans and
+Neanderthals and reduced diversity within the Neanderthal population
 (Fig.~\ref{fig:supp-FST-panels}), the increased $F_{ST}$ between the stems is
@@
 diversity, but individuals across stems showing similar differentiation to
-that between two modern population.
+that between two contemporary human population.
 
@@
 
-To evaluate the degree to which divergence among modern populations is
+\tdwcomment{I have tried to make the terminology about human, hominin, etc.
+consistent and unambiguous. To do this, I have avoided using "modern" because 
+"modern human" can mean "anatomically modern human" or \emph{Homo sapiens}, 
+but mostly in the paper we use it to mean "contemporary" or "present-day". I 
+think we should use "contemporary human" or "present-day human" when we mean 
+the latter and just "human" when we mean the former. I have made these changes 
+in the main text and supplement, preferentially using "contemporary" but 
+leaving "present-day" when it was already there. However, I think it would be 
+best if Simon made these changes in the equations in this section, so I don't mess 
+something up.}
+
+To evaluate the degree to which divergence among contemporary human populations is
 inherited from divergence between early stem population, we computed $f_4$
@@
 \end{equation}
-This statistic measures co-linearity of genetic drift between modern
+This statistic measures co-linearity of genetic drift between contemporary
 populations $v= (p_{\text{modern}_1} -p_{\text{modern}_2} )$ and stem populations
@@
 
-To contrast these statistics to divergence across modern populations,
+To contrast these statistics to divergence across contemporary populations,
 \begin{equation}
@@
 
-Finally, the proportion of modern-day structure
+Finally, the proportion of present-day structure
 $v^2 = f_2(\text{modern}_1, \text{modern}_2)$ that is accounted for by the
@@
 $\alpha^2$ can be interpreted as the proportion of the $f_2$
-divergence across modern populations that is aligned with the differences in
+divergence across contemporary populations that is aligned with the differences in
 allele frequencies across ancient populations. However, drift occurring after
@@
 occurring before that time. For this reason, we also interpret $\alpha^2$ as the
-proportion of the $f_2$ divergence across modern populations that is \emph{explained by}
+proportion of the $f_2$ divergence across contemporary populations that is \emph{explained by}
 the differences in allele frequencies across ancient populations. We report
@@
     \caption{
-        \textbf{Present day structure attributable to ancient population structure.}
+        \textbf{Present-day structure attributable to ancient population structure.}
         The statistic 
@@
         population structure among the stems. The blue line represents genetic drift 
-        along ancestral branches between modern populations, while the red line 
+        along ancestral branches between contemporary populations, while the red line 
         represents drift between stem populations. $f_4$ measures the overlap 
@@
         between the two pairs. 
-        In scenario A, modern populations are descended from distinct stems, and 
+        In scenario A, contemporary populations are descended from distinct stems, and 
         $f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)$
@@
         \midrule
-        $T_{MH-Neand}$ & The time of the split between branches leading to Neanderthals and modern humans & 550ka \\
-        $T_{NI-Vin}$ & The time of the split between branches leading to the Vindija sample and the Neanderthal population that mixed with expanding modern humans & 80ka \\
+        $T_{MH-Neand}$ & The time of the split between branches leading to Neanderthals and humans & 550ka \\
+        $T_{NI-Vin}$ & The time of the split between branches leading to the Vindija sample and the Neanderthal population that mixed with expanding humans & 80ka \\
         $T_{Vin}$ & The sample time of the Vindija Neanderthal individual & 50ka \\
-        $T_{NI\rightarrow EUR}$ & The timing of admixture from Neanderthals to modern humans & 45ka \\
-        $f_{NI\rightarrow EUR}$ & The proportion of admixture from Neanderthals to modern humans & 0.015 \\
+        $T_{NI\rightarrow EUR}$ & The timing of admixture from Neanderthals to humans & 45ka \\
+        $f_{NI\rightarrow EUR}$ & The proportion of admixture from Neanderthals to humans & 0.015 \\
         $T_{E-W}$ & The split time of East and West African populations & 60ka \\
@@
     \textbf{Best-fit parameters from the single-origin model.}
-    This model includes a single stem population that splits into modern human
+    This model includes a single stem population that splits into contemporary human
     groups. Fixed parameters are specifed in Table~\ref{tab:fixed-params}.
@@
     $N_e$ & Ancestral effective population size & 10198 & 403 \\
-    $N_{MH}$ & Size of modern-human lineage between Neanderthal and Nama splits & 21111 & 529 \\
+    $N_{MH}$ & Size of human lineage between Neanderthal and Nama splits & 21111 & 529 \\
     $N_{Nama_0}$ & Initial Nama size & 10224 & 370 \\
@@
     29 years. This model gave a log-likelihood of -126,644. This model allows for
-    continuous migration between Stem 2 and modern populations, but disallows
+    continuous migration between Stem 2 and contemporary populations, but disallows
     migration between Stem 2 and Stem 1, which later splits into branches leading
@@
         stem migration models predict a period of increased $F_{ST}$ between ancestors
-        of modern humans. This is largely due to the very small inferred $N_e$ in one of
+        of contemporary humans. This is largely due to the very small inferred $N_e$ in one of
         the stem branches, which leads to a rapid increase in differentiation due to
@@
     \caption{
-        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
+        \textbf{Predicted \bm{$f_4$} between pairs of contemporary human populations
         and ancient branches from the single-origin model.}
@@
     \caption{
-        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
+        \textbf{Predicted \bm{$f_4$} between pairs of contemporary human populations
         and ancient branches from the continuous-migration model.}
         Despite population structure inferred to have extended up to 1Ma
-        into the past, differentiation between modern African populations
+        into the past, differentiation between contemporary African populations
         only traces back to differentiation between ancestral populations
@@
         (Fig.~\ref{fig:supp-FST-panels}), differentiation due to that
-        ancient structure contributes little to differentiation between modern
+        ancient structure contributes little to differentiation between contemporary
         populations.
@@
     \caption{
-        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
+        \textbf{Predicted \bm{$f_4$} between pairs of contemporary human populations
         and ancient branches from the merger-without-stem-migration model.}
@@
         following the divergence of Neanderthal and human ancestors results
-        in differentiation among modern populations that aligns with drift
+        in differentiation among contemporary populations that aligns with drift
         between the stems. In this model, migration between stem populations
@@
     \caption{
-        \textbf{Predicted \bm{$f_4$} between pairs of modern populations
+        \textbf{Predicted \bm{$f_4$} between pairs of contemporary populations
         and ancient branches from the merger-with-stem-migration model.}
-        Differentiation between modern present-day populations aligns with
+        Differentiation between present-day populations aligns with
         up to $\approx 5\%$ of the drift between stems, although some

### 2022-03-11 supplement.tex
Changed to "ka"

@@
 across population pairs and over different epochs. This continuous migration
-was allowed until stem 2 disappeared, which occurred as recently as 5kya. We
+was allowed until stem 2 disappeared, which occurred as recently as 5ka. We
 tested models that both allowed or disallowed migration between stems, i.e.,
@@
 Even though each of our four models inferred divergences of all human
-populations $\sim$120kya, ancient structure and size changes in the simulations
+populations $\sim$120ka, ancient structure and size changes in the simulations
 biases re-inferences and gives inflated estimates of $T$.
@@
 \citep{Momigliano2021-th}, which is the case in our inferred single-origin model.
-Re-inferred $T$ between the Nama and other populations were $>$200kya under the IM model,
-and $\sim$50--130kya under the clean split model.
+Re-inferred $T$ between the Nama and other populations were $>$200ka under the IM model,
+and $\sim$50--130ka under the clean split model.
 

### 2022-03-11 paper.tex
Update paper.tex
some formatting

@@
 \usepackage{url}
-\usepackage{authblk}
 \usepackage{graphicx}
@@
 \usepackage{booktabs}
+
+\usepackage{authblk}
+\renewcommand{\Affilfont}{\normalsize}
+\renewcommand{\Authfont}{\normalsize}
 
@@
 \author[4]{Marlo M\"{o}ller}
-\author[2,6,$\dag$,*]{Brenna M. Henn}
-\author[7,$\dag$,**]{Simon Gravel}
+\author[2,5,$\dag$,*]{Brenna M. Henn}
+\author[6,$\dag$,**]{Simon Gravel}
 \affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
@@
 \affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
-\affil[6]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}
-\affil[7]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
+\affil[5]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}
+\affil[6]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
 \affil[$\dag$]{Co-Corresponding Authors}
@@
 \affil[**]{simon.gravel@mcgill.ca}
+\date{\normalsize \today}
 \maketitle
@@
 that had previously been attributed to contributions from archaic hominins in
-Africa.
+Africa. \aprcomment{.. to contributions from archaic hominin introgression in Africa?}
 In contrast to models with archaic introgression, we predict that fossil
@@
 similar.
-Despite genetic similarity between these populations, an inferred 1--4\%\sgcomment{explain in the main text where this estimate is from!}
+Despite genetic similarity between these populations, an inferred 1--4\%
+\sgcomment{explain in the main text where this estimate is from!}
 of genetic differentiation among contemporary human populations can be

### 2022-03-11 supplement.tex

@@
     \item East/West African population split: 60ka
-    \item East African/European split: 50 ka
-    \item Neanderthal introgression to Eurasian branch: 45 ka
-    \item Neanderthal/human split: 550 ka
-    \item Eurasian back-to-African migration: 12 ka
+    \item East African/European split: 50ka
+    \item Neanderthal introgression to Eurasian branch: 45ka
+    \item Neanderthal/human split: 550ka
+    \item Eurasian back-to-African migration: 12ka
 \end{itemize}
@@
 \begin{itemize}
-    \item East African pastoralist to South African admixture: 2 ka \cite{}
+    \item East African pastoralist to South African admixture: 2ka \cite{}
     \item European to South African admixture: 10 generations ago \cite{}
-    \item Mende population expansion and Gumuz population decline: 5 ka \cite{}
+    \item Mende population expansion and Gumuz population decline: 5ka \cite{}
     \item South African population decline following colonial admixture: 9 generations ago \cite{}
@@
     $m_{EA-EA}$ & Intra-East Africa migration rate & $41.28\times10^{-5}$ & $1.33\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.658 & 0.0039 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.279 & 0.0039 \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.658 & 0.0039 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.279 & 0.0039 \\
     $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.150 & 0.0019 \\
@@
     $m_{EA-EA}$ & Intra-East Africa migration rate & $35.4\times10^{-5}$ & $1.85\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.649 & 0.0038 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.268 & 0.0048 \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.649 & 0.0038 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.268 & 0.0048 \\
     $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.153 & 0.0022 \\
@@
     $m_{EA-EA}$ & Intra-East Africa migration rate & $33.6\times10^{-5}$ & $8.35\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.642 & 0.0037 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.255 & 0.0043 \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.642 & 0.0037 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.255 & 0.0043 \\
     $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.0021 \\
@@
     $m_{EA-EA}$ & Intra-East Africa migration rate & $37.1\times10^{-5}$ & $1.26\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.647 & 0.0037 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.257 & 0.0042 \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.647 & 0.0037 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.257 & 0.0042 \\
     $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.0021 \\
@@
     $m_{EA-EA}$ & Intra-East Africa migration rate & $35.9\times10^{-5}$ & $5.36\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12 ka ($1-f$ from Gumuz) & 0.640 & 0.0075 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2 ka & 0.257 & 0.0049 \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.640 & 0.0075 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.257 & 0.0049 \\
     $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.157 & 0.0031 \\

### 2022-03-11 paper.tex
cleaned up some comments

--- 2022-03-10 paper.tex
@@
 disequilibrium (LD) statistics that are well suited for both low- and
-high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
+high-coverage genomes \sgcomment{We computed one- and two-locus statistics whose expectation within and across populations can be computed efficiently and that are well suited for both low- and
+high-coverage genomes} \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
 maximum-likelihood inference framework, we then fit to these statistics a
@@
 reticulation (such as fragmentation-and-coalescence or meta-population models,
-\aprcomment{defining reticulation at first mention -- something better here?}
 Figure~\ref{fig:supp-possible-models}B), archaic hominin admixture
@@
         & & & 
-            \textbf{Time (ka)} & \textbf{per generation} & \textbf{duration (ky)} \\
+            \textbf{Time (ka)} & \textbf{per generation} & \textbf{duration (kyr)} \\
         \midrule

### 2022-03-11 paper.tex
real-time edits

@@
 
-\comment{About title: "Contrasting archaic admixture and a weakly structured stem for human originas in Africa" Too long? Inconclusive?} \sgcomment{Disagree.}
-
 \begin{abstract}
@@
 groups, we infer long-lasting structure between two or more weakly
-differentiated ancestral \emph{Homo spp.}
-\comment{maybe don't abbreviate; won't be understood outside of evolutionary biology as much}
+differentiated ancestral \emph{Homo}
 populations connected by gene flow over
@@
 that had previously been attributed to contributions from archaic hominins in
-Africa. \aprcomment{.. to contributions from archaic hominin introgression in Africa?}
+Africa.
 In contrast to models with archaic introgression, we predict that fossil
@@
 sapiens} do not descend. 
-\comment{unpack more; explain why these are at odds}
-Recently, synthetic \comment{synthetic doesn't feel quite right here, can delete}
+Recently,
 attempts to reconcile genetic and
@@
 sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
-recruited populations of recent West African ancestry (Figure \ref{fig:1}C). 
-\comment{What about the LWK?}
+recruited populations of recent West African or Bantu-speaking ancestry (Figure \ref{fig:1}C). 
 While ancient DNA from Eurasia has helped us understand early human history
@@
 including geographically and genetically diverse populations across Africa, we
-infer demographic models that explain more aspects of genetic diversity
-\comment{this is a bit vague; make more detailed. e.g.: better fit patterns of LD and \ldots}
+infer demographic models that explain more features of genetic diversity
 in more
@@
 time depth of divergence, assuming Neanderthals diverged 550ka from a common
-stem. For each population, we computed low-order
-\comment{what do you mean by low-order? not a commonly used phrase}
-\aprcomment{I think this is a fairly common term in statistics, right?}
-allele frequency and linkage
-disequilibrium (LD) statistics that are well suited for both low- and
-high-coverage genomes \sgcomment{We computed one- and two-locus statistics whose expectation within and across populations can be computed efficiently and that are well suited for both low- and
-high-coverage genomes} \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
+stem. We computed one- and two-locus statistics whose expectation within and across 
+populations can be computed efficiently and that are well suited for both low- and
+high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
 maximum-likelihood inference framework, we then fit to these statistics a
@@
 events, to learn about the nature of population structure over the past million
-years.
+years. \sgcomment{define LD}
 
@@
         \textbf{Genetic diversity across Africa.}
-        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
+        (A) \sgcomment{secure coordinates} Select populations from the 1000 Genomes and African Diversity Reference Panels
         illustrate diversity from western, eastern and southern Africa.
@@
         Percentages show variance explained by each principal component.
-        (D) \texttt{ADMIXTURE} illustrates signatures of recent gene flow in Africa
+        (D) \texttt{ADMIXTURE} with $K=4$ illustrates signatures of recent gene flow in Africa
         which reflect colonial-period migration into the Nama,
         Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
-        \comment{How did you land on the best fit k? Show scree plot of CV errors, or otherwise justify choice}
+        %\comment{How did you land on the best fit k? Show scree plot of CV errors, or otherwise justify choice}
     }
@@
 geographic expansion of one or both stems. For example, if during MIS 5, either
-Stem 1S \comment{where is this illustrated/labeled?}
+Stem 1S (Figure \ref{fig:2}B) 
 from southern Africa moved northward thereby encountering the Stem 2,
@@
 glacial period of aridity across Africa \citep{Blome2012-lw}, might have
-triggered migration inland away from the coasts, as has been suggested for the
+triggered migration inland away from the coasts, as has been suggested, e.g., for the
 Paleo-Agulhas plain \citep{Marean2014-pg}.
-\comment{very specific... might be worth expanding a bit}
 
@@
 \citep{Petr2019-xo}, and their interpretation should be re-examined in light of
-a weakly-structured-stem model within Africa. \comment{Reviewers may suggest this demonstrated here}
+a weakly-structured-stem model within Africa.
 

### 2022-03-11 supplement.tex
clean up comments

@@
 dataset was performed via SHAPEIT2 \citep{Delaneau2013-aw} and utilized the duoHMM
-option for duos and trios. We highlight 82 Nama genomes which are newly
-available (and previously unpublished) under accession number EGAD00001006198.
+option for duos and trios. We present new whole-genome data for 82 Nama individuals 
+available under accession number EGAD00001006198.
 \comment{This makes is sound like these weren't generated for this project.}
@@
 analysis, after partitioning ancestry into $k=6$ clusters with \texttt{ADMIXTURE}
-\comment{Justification for k=6}
 \citep{Alexander2009-sw} where alternative ancestries represent European, West
@@
 East, and West African branches
-\comment{would also be interesting to organize by ethnolinguistic group...}
 (S/E/W AFR, Figures~\ref{fig:1},~\ref{fig:supp-possible-models}). We do not know the
@@
 \end{itemize}
-\comment{This could be organized with a bit more structure. E.g. by region, or by timepoint}
+\comment{This could be organized with a bit more structure. E.g. by region, or by timepoint} \sgcomment{order chronologically}
 While the dates of these events were fixed, the sizes and proportions were

### 2022-03-11 paper.tex
some updates

--- 2022-03-11 paper.tex
@@
 Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
-\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
+\citep{Gurdasani2015-qy,Gopalan2022-pw}), and Eastern African agriculturalists
 (Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
 populations, despite speaking distinct Afro-Asiatic languages, are highly
-genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
+genetically similar \citep{Pagani2015-pz,Gopalan2022-pw} and thus the two
 groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
@@
 disequilibrium (LD) statistics that are well suited for both low- and
-high-coverage genomes \sgcomment{We computed one- and two-locus statistics whose expectation within and across populations can be computed efficiently and that are well suited for both low- and
-high-coverage genomes} \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
+high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
 maximum-likelihood inference framework, we then fit to these statistics a
@@
 reticulation (such as fragmentation-and-coalescence or meta-population models,
+\aprcomment{defining reticulation at first mention -- something better here?}
 Figure~\ref{fig:supp-possible-models}B), archaic hominin admixture
@@
 Stem 1S \comment{where is this illustrated/labeled?}
-from southern Africa moved northward thereby encountering the Stem 2,
+from southern Africa moved northward thereby encountering Stem 2,
 or Stem 2 moved from central/western Africa southward into Stem 1S -- then we

### 2022-03-11 paper.tex

--- 2022-03-11 paper.tex
@@
 diversity-based statistics, optimized for rapid, complex demographic inference
-to discriminate among such models \comment{\ldots to evaluate the fit of various models}.
+to discriminate among such models.
 We infer detailed demographic models for
@@
 Despite genetic similarity between these populations, an inferred 1--4\%
-\sgcomment{explain in the main text where this estimate is from!}
 of genetic differentiation among contemporary human populations can be
@@
 approaches are more challenging to interpret and fit to data.
-\comment{I think a schematic illustrating models being tested would be extremely useful in the main text.
-This could be Fig S2, or even a new figure with a further simplified illustration o fthe general idea of the models being tested to help orient the reader.}
-However, new
-population genetic tools now allow for inference involving tens to
+However, new population genetic tools now allow for inference involving tens to
 hundreds of genomes from multiple populations and greater complexity
@@
 with an archaic hominin admixture model, leaving out other plausible models
-\citep{Henn2018-rf} (Figure~\ref{fig:supp-possible-models}).
+\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
 Second, they
@@
 sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
-recruited populations of recent West African or Bantu-speaking ancestry (Figure \ref{fig:1}C). 
+recruited populations of recent West African or Bantu-speaking ancestry
+(Figure~\ref{fig:diversity}C). 
 While ancient DNA from Eurasia has helped us understand early human history
@@
 population expansion with regional persistence, archaic hominin admixture, and
-multi-regional evolution, Figure~\ref{fig:supp-possible-models}), using 290
+multi-regional evolution, Figure~\ref{fig:proposed-models}), using 290
 genomes from southern, eastern, and western Africa as well as Eurasia. By
@@
 
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.5\textwidth]{figures/proposed-models.png}
+    \caption{
+        \textbf{Proposed models of early human history in Africa.}
+        seeking to reconcile genetic and paleoanthropological evidence \citep{Henn2018-rf}.
+        \aprcomment{needs a caption}
+    }
+    \label{fig:proposed-models}
+\end{figure}
+
 \section*{Results}
@@
 Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
-\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
+\citep{Gurdasani2015-qy,Gopalan2022-pw}), and Eastern African agriculturalists
 (Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
 populations, despite speaking distinct Afro-Asiatic languages, are highly
-genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
-groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
+genetically similar \citep{Pagani2015-pz,Gopalan2022-pw} and thus the two
+groups were combined for a larger sample size (Figure~\ref{fig:diversity}). We also
 included the British (GBR) from the 1000 Genomes Project in our demographic
@@
 events, to learn about the nature of population structure over the past million
-years. \sgcomment{define LD}
+years.
 
@@
     \centering
-    \includegraphics{figures/fig1.pdf}
+    \includegraphics{figures/samples-diversity.pdf}
     \caption[width=\textwidth]{
         \textbf{Genetic diversity across Africa.}
-        (A) \sgcomment{secure coordinates} Select populations from the 1000 Genomes and African Diversity Reference Panels
+        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
         illustrate diversity from western, eastern and southern Africa.
@@
         Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
-        %\comment{How did you land on the best fit k? Show scree plot of CV errors, or otherwise justify choice}
     }
-    \label{fig:1}
+    \label{fig:diversity}
 \end{figure}
@@
 for contribution from an African archaic hominin lineage or population structure prior
-to the expansion (Figure~\ref{fig:supp-possible-models}A). As expected
+to the expansion (Figure~\ref{fig:proposed-models}A). As expected
 \citep{Ragsdale2019-nt}, this first model was a poor fit to the data
@@
 reticulation (such as fragmentation-and-coalescence or meta-population models,
-Figure~\ref{fig:supp-possible-models}B), archaic hominin admixture
-(Figure~\ref{fig:supp-possible-models}C), and African multi-regionalism
-(Figure~\ref{fig:supp-possible-models}D).
+Figure~\ref{fig:proposed-models}B), archaic hominin admixture
+(Figure~\ref{fig:proposed-models}C), and African multi-regionalism
+(Figure~\ref{fig:proposed-models}D).
 
@@
 `stem' groups as only a small proportion of drift between contemporary populations can
-be attributed to drift between stems (Figures~\ref{fig:3} and
+be attributed to drift between stems (Figures~\ref{fig:predictions} and
 \ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
@@
 contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
-was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:1}).
+was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:diversity}).
 
@@
     \centering
-    \includegraphics[width=0.65\textwidth]{figures/fig2-melisa}
+    \includegraphics[width=0.65\textwidth]{figures/best-fit-models}
     \caption{
-        \textbf{A weakly structured stem best describes two-locus statistics}
+        \textbf{A weakly structured stem best describes two-locus statistics.}
         In the two best-fitting parameterizations of early population structure,
@@
         continuous-migration model (A, with $LL=-115,500$ ).
-        Colors are used to distinguish overlapping branches and link to Figure \ref{fig:1}.}
-    \label{fig:2}
+        Colors are used to distinguish overlapping branches and link to
+        Figure~\ref{fig:diversity}.}
+    \label{fig:best-fit-models}
 \end{figure}
@@
         Labeled migration rates correspond to symmetric continuous migration
-        bands shown in Figure~\ref{fig:2}. Both the continuous migration and
+        bands shown in Figure~\ref{fig:best-fit-models}. Both the continuous migration and
         merger models inferred a relatively deep split of human stem branches,
@@
 this scenario encompasses archaic hominin introgression and
-fragmentation-and-coalescence models.
-\comment{another supp schematic figure may be helpful}
+fragmentation-and-coalescence models (such as Figure~\ref{fig:proposed-models}A and C).
 For many parameters, confidence intervals based on bootstrapping are relatively narrow
@@
 early divergence of stem populations and their relative $N_e$
-(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
+(Figure~\ref{fig:best-fit-models}). The two models also differ in the mode of divergence
 during the Middle Stone Age.
@@
 periodic) gene flow qualitatively distinguishes these models from previously
-proposed archaic hominin admixture models (Figure~\ref{fig:supp-possible-models}C) as
-the early branches remain closely related, and each branch contributes large amounts to all contemporary populations (Figure 3). 
-Because of this 
-relatedness, only $1\%$ to $4\%$ of genetic differentiation among contemporary populations can be traced back
-to this early population structure (Figure 3 and Supp. Information)
+proposed archaic hominin admixture models (Figure~\ref{fig:proposed-models}C) as
+the early branches remain closely related, and each branch contributes large
+amounts to all contemporary populations (Figure~\ref{fig:predictions}). 
+Because of this relatedness, only $1\%$ to $4\%$ of genetic differentiation
+among contemporary populations can be traced back
+to this early population structure (Supp. Information)
 
@@
     \centering
-    \includegraphics{figures/fig3.pdf}
+    \includegraphics{figures/predictions.pdf}
     \caption{
-        \textbf{Structure among stems is weak and present-day structure is mostly recent}
+        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
         From the best fit models of our two parameterizations (A and B:
@@
     }
-    \label{fig:3}
+    \label{fig:predictions}
 \end{figure}
@@
 being carried in the Neanderthal sample) is very well-described by the merger
-model (Figures~\ref{fig:4}A-C and
+model (Figures~\ref{fig:validation}A-C and
 \ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
@@
     \centering
-    \includegraphics{figures/fig4}
+    \includegraphics{figures/validation}
     \caption{
         \textbf{Model validation using independent statistics.} (A--C) Using
-        our best fit models to the LD and pairwise diversity statistics, we
-        simulated expected conditional site-frequency-spectra (cSFS) and
+        our best fit models to linkage disequilibrium and pairwise diversity statistics,
+        we simulated expected conditional site-frequency-spectra (cSFS) and
         compared to the observed cSFS from the data. Our inferred models
@@
     }
-    \label{fig:4}
+    \label{fig:validation}
 \end{figure}
@@
 This increase in $N_e$ could be due to either an increase in population or to 
-ancestral population structure during the Middle Stone Age \citep{Mazet2016-wn}.
+ancestral population structure during the Middle Stone Age \citep{Mazet2015-wn}.
 All models, including the
 single-origin model, recapitulate an inferred ancestral
-increase in $N_e$ between 100ka-1Ma (Figure \ref{fig:supp-iicr-sim}).
+increase in $N_e$ between 100ka-1Ma (Figure~\ref{fig:supp-iicr-sim}).
 Whereas the single-origin model achieves this by
@@
 $\sim$15ka compared to a true divergence of 60ka), and recent migration can
-lead to the misordering of divergence events (Figure~\ref{fig:4}E). We suggest
+lead to the misordering of divergence events (Figure~\ref{fig:validation}E). We suggest
 that rCCR analyses which do not fit multiple parameters including gene flow
@@
 geographic expansion of one or both stems. For example, if during MIS 5, either
-Stem 1S (Figure \ref{fig:2}B) 
-from southern Africa moved northward thereby encountering the Stem 2,
+Stem 1S (Figure~\ref{fig:best-fit-models}B) 
+from southern Africa moved northward thereby encountering Stem 2,
 or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
@@
 thank Luke Anderson-Trocm\'e for assistance in creating the map in
-Figure~\ref{fig:1}. This research was supported by an NIH grant R35GM133531 (to
+Figure~\ref{fig:diversity}. This research was supported by an NIH grant R35GM133531 (to
 BMH). The content is solely the responsibility of the authors and does

### 2022-03-11 supplement.tex

@@
 ``A weakly structured stem for human origins in Africa''}
-\author{Aaron P. Ragsdale, Timothy D. Weaver, Brenna M. Henn, and Simon Gravel}
+\author{Aaron P. Ragsdale, Timothy D. Weaver, Elizabeth G. Atkinson, Eileen Hoal, Marlo M\"{o}ller, Brenna M. Henn, and Simon Gravel}
 \date{\today}
@@
 available under accession number EGAD00001006198.
-\comment{This makes is sound like these weren't generated for this project.}
 Among these 82 Nama samples, we down-sampled the individuals to minimize
@@
 \citep{Alexander2011-aq} to summarize genetic diversity among a broader set of
-1000 Genomes and ADRP populations (Figure~\ref{fig:1} in the main text). Data
+1000 Genomes and ADRP populations (Figure~\ref{fig:diversity} in the main text). Data
 filtering was identical for both analyses. We retained biallelic SNPs with no
@@
 Because we were concerned that cSFS analyses may be affected by incorrect
-inference of the ancestral allele \cite{Hernandez2007-mf}, we computed the cSFS
+inference of the ancestral allele \citep{Hernandez2007-mf}, we computed the cSFS
 for all mutations, and for transitions and transversions separately
@@
 East, and West African branches
-(S/E/W AFR, Figures~\ref{fig:1},~\ref{fig:supp-possible-models}). We do not know the
+(S/E/W AFR, Figures~\ref{fig:proposed-models},~\ref{fig:diversity}). We do not know the
 geographical location of these ancestral populations (nor even if they
@@
 \begin{itemize}
+    \item Neanderthal/human split: 550ka
     \item East/West African population split: 60ka
@@
     \item Neanderthal introgression to Eurasian branch: 45ka
-    \item Neanderthal/human split: 550ka
     \item Eurasian back-to-African migration: 12ka
@@
 \begin{itemize}
-    \item East African pastoralist to South African admixture: 2ka \cite{}
-    \item European to South African admixture: 10 generations ago \cite{}
-    \item Mende population expansion and Gumuz population decline: 5ka \cite{}
-    \item South African population decline following colonial admixture: 9 generations ago \cite{}
+    \item Mende population expansion and Gumuz population decline: 5ka \citep{Gopalan2022-pw}
+    \item East African pastoralist to South African admixture: 2ka \citep{Uren2016-nn}
+    \item European to South African admixture: 10 generations ago
+    \item South African population decline following colonial admixture: 9 generations ago
 \end{itemize}
-\comment{This could be organized with a bit more structure. E.g. by region, or by timepoint} \sgcomment{order chronologically}
-While the dates of these events were fixed, the sizes and proportions were
 allowed to vary in the fits. The total number of parameters that were
@@
 L-BFGS-B methods (as implemented in the SciPy optimization package
-\cite{Virtanen2020-kr}). This process was repeated with alternating
+\citep{Virtanen2020-kr}). This process was repeated with alternating
 optimization methods until the best fit parameter set converged consistently.
@@
 
-\tdwcomment{I have tried to make the terminology about human, hominin, etc.
-consistent and unambiguous. To do this, I have avoided using "modern" because 
-"modern human" can mean "anatomically modern human" or \emph{Homo sapiens}, 
-but mostly in the paper we use it to mean "contemporary" or "present-day". I 
-think we should use "contemporary human" or "present-day human" when we mean 
-the latter and just "human" when we mean the former. I have made these changes 
-in the main text and supplement, preferentially using "contemporary" but 
-leaving "present-day" when it was already there. However, I think it would be 
-best if Simon made these changes in the equations in this section, so I don't mess 
-something up.}
+%\tdwcomment{I have tried to make the terminology about human, hominin, etc.
+%consistent and unambiguous. To do this, I have avoided using "modern" because 
+%"modern human" can mean "anatomically modern human" or \emph{Homo sapiens}, 
+%but mostly in the paper we use it to mean "contemporary" or "present-day". I 
+%think we should use "contemporary human" or "present-day human" when we mean 
+%the latter and just "human" when we mean the former. I have made these changes 
+%in the main text and supplement, preferentially using "contemporary" but 
+%leaving "present-day" when it was already there. However, I think it would be 
+%best if Simon made these changes in the equations in this section, so I don't mess 
+%something up.}
 
@@
     f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = 
-    \E[(p_{\text{stem}_1} - p_{\text{stem}_2} )(p_{\text{modern}_1} -p_{\text{modern}_2} )]
+    \E[(p_{\text{stem}_1} - p_{\text{stem}_2} )(p_{\text{present}_1} -p_{\text{present}_2} )]
     \label{eq:f4}
@@
 This statistic measures co-linearity of genetic drift between contemporary
-populations $v= (p_{\text{modern}_1} -p_{\text{modern}_2} )$ and stem populations
+populations $v= (p_{\text{present}_1} -p_{\text{present}_2} )$ and stem populations
 $w= (p_{\text{stem}_1} -p_{\text{stem}_2} )$ (see
@@
 \begin{equation}
-    f_2(\text{modern}_1, \text{modern}_2) =
-    \E[(p_{\text{modern}_1} -p_{\text{modern}_2} )^2],
+    f_2(\text{present}_1, \text{present}_2) =
+    \E[(p_{\text{present}_1} -p_{\text{present}_2} )^2],
 \end{equation}
 we interpret the $f$-statistics as scalar products between vectors $v$ and $w$, i.e.,  
-$f_2(\text{modern}_1, \text{modern}_2) = v\cdot v$ and 
-$f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = v\cdot w.$
-We can then write $v=v_\perp + v_\parallel$ as the sum of a component $v_\perp$ perpendicular and
-a component $v_\parallel$ parallel to $w = (p_{\text{stem}_1} -p_{\text{stem}_2} ).$
+$f_2(\text{present}_1, \text{present}_2) = v\cdot v$ and 
+$f_4(\text{stem}_1, \text{stem}_2; \text{present}_1, \text{present}_2) = v\cdot w.$
+We can then write $v=v_\perp + v_\parallel$ as the sum of
+a component $v_\perp$ perpendicular and
+a component $v_\parallel$ parallel to
+$w = (p_{\text{stem}_1} -p_{\text{stem}_2} ).$
 We can then decompose 
-$f_2(\text{modern}_1,\text{modern}_2) = v_\perp^2+v_\parallel^2$, with  
+$f_2(\text{present}_1,\text{present}_2) = v_\perp^2+v_\parallel^2$, with  
 \begin{equation}
     v_\parallel^2 = \frac{(v \cdot w)^2}{w\cdot w} = \frac{ f_4^2(\text{stem}_1, \text{stem}_2; 
-    \text{modern}_1, \text{modern}_2)}{f_2(\text{stem}_1, \text{stem}_2)}
+    \text{present}_1, \text{present}_2)}{f_2(\text{stem}_1, \text{stem}_2)}
 \end{equation}
@@
 Finally, the proportion of present-day structure
-$v^2 = f_2(\text{modern}_1, \text{modern}_2)$ that is accounted for by the
+$v^2 = f_2(\text{present}_1, \text{present}_2)$ that is accounted for by the
 component parallel to ancient population structure is
@@
     \alpha^2 = \frac{v_\parallel^2}{v^2} =
-    \frac{f_4^2(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)}
-    {f_2(\text{stem}_1,\text{stem}_2)f_2(\text{modern}_1, \text{modern}_2)}.
+    \frac{f_4^2(\text{stem}_1, \text{stem}_2; \text{present}_1, \text{present}_2)}
+    {f_2(\text{stem}_1,\text{stem}_2)f_2(\text{present}_1, \text{present}_2)}.
 \end{equation}
@@
 occurring before that time. For this reason, we also interpret $\alpha^2$ as the
-proportion of the $f_2$ divergence across contemporary populations that is \emph{explained by}
-the differences in allele frequencies across ancient populations. We report
-this predicted statistic for each of the four inferred demographic models in
+proportion of the $f_2$ divergence across contemporary populations that is
+\emph{explained by} the differences in allele frequencies across ancient populations.
+We report this predicted statistic for each of the four inferred demographic models in
 Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}.
+In our best fit models (continuous migration and merger with stem migration),
+differentiation between Stem 1 and Stem 2
+(before the split of Stem 1 into Stem 1E and Stem 1S in the merger model)
+contributes a maximum of $4\%$ to differentiation between the Nama and Gumuz, and closer
+to $1\%$ to differentiation between the Nama and Mende or the Mende and Gumuz.
 
@@
         The statistic 
-        $f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)$
+        $f_4(\text{stem}_1, \text{stem}_2; \text{present}_1, \text{present}_2)$
         captures how much of present-day population structure can be attributed to 
@@
         In scenario A, contemporary populations are descended from distinct stems, and 
-        $f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)$
+        $f_4(\text{stem}_1, \text{stem}_2; \text{present}_1, \text{present}_2)$
         equals the drift between stems, $f_2(\text{stem}_1, \text{stem}_2)$.
@@
         and $f_4=0$. That is, present-day population structure is unrelated to
-        structure among stems. In more realistic scenario C, where $\text{modern}_1$
+        structure among stems. In more realistic scenario C, where $\text{present}_1$
         receives a proportion $m$ of its ancestry from the dotted arrow, the overlap
@@
         ancestry proportion $m$ and drift between stems,
-        $f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2)
+        $f_4(\text{stem}_1, \text{stem}_2; \text{present}_1, \text{present}_2)
         = mf_2(\text{stem}_1, \text{stem}_2)$
@@
     \centering
-    \includegraphics[width=0.65\textwidth]{figures/supp-possible-models.png}
-    \caption{
-        \textbf{Models of early human history in Africa} seeking to reconcile genetic and paleoanthropological evidence \cite{Henn2018-rf}.
-         \comment{Lovely figure, may even want to move to main text}
-    }
-    \label{fig:supp-possible-models}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
     \includegraphics{figures/supp-omni-hapmap-comparison.pdf}

### 2022-03-11 supplement.tex
more modern to present changes
formatting

@@
 \usepackage{url}
+
 \usepackage{authblk}
+\renewcommand{\Affilfont}{\normalsize}
+\renewcommand{\Authfont}{\normalsize}
+
 \usepackage{graphicx}
@@
 \author{Aaron P. Ragsdale, Timothy D. Weaver, Elizabeth G. Atkinson, Eileen Hoal, Marlo M\"{o}ller, Brenna M. Henn, and Simon Gravel}
-\date{\today}
+\date{\normalsize \today}
 \maketitle
@@
 (Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
-we used \msprime version 1.0.4 \citep{Kelleher2016-lw,Baumdicker2021-mj} to
+we used \msprime version 1.0.4 \citep{Kelleher2016-lw,Baumdicker2022-mj} to
 simulate sampled individual genomes to match the data used in inference (see
@@
 The simulations sampled 10 diploid individuals from each
-population, and we used \msprime \citep{Baumdicker2021-mj} to simulated
+population, and we used \msprime \citep{Baumdicker2022-mj} to simulated
 500 Mb of total sequence length (the sum over500 replicates each 1 Mb in length)

### 2022-03-11 paper.tex
figure 1 caption

@@
     \caption{
-        \textbf{Proposed models of early human history in Africa.}
-        seeking to reconcile genetic and paleoanthropological evidence \citep{Henn2018-rf}.
-        \aprcomment{needs a caption}
+        \textbf{Proposed conceptual models of early human history in Africa.}
+        These models have been designed to translate models from the 
+        paleoanthropological literature into genetically testable demographicmodels \citep{Henn2018-rf}.
+       We used these conceptual models as starting points to build detailed parameterized demographic models 
+       (Supplementary section \ref{modelspec}) that were then fit to genetic data. 
     }

### 2022-03-11 supplement.tex
figure 1 caption

--- 2022-03-11 supplement.tex
@@
 \section{Model specification and fitting}
-
+\label{modelspec}
 For the early history, we tested model parameterizations that cover many of the
@@
 \begin{equation}
-    f_4(\text{stem}_1, \text{stem}_2; \text{modern}_1, \text{modern}_2) = 
+    f_4(\text{stem}_1, \text{stem}_2; \text{present}_1, \text{present}_2) = 
     \E[(p_{\text{stem}_1} - p_{\text{stem}_2} )(p_{\text{present}_1} -p_{\text{present}_2} )]

### 2022-03-11 paper.tex
clean up figure and caption

--- 2022-03-11 paper.tex
@@
     \caption{
-        \textbf{Proposed models of early human history in Africa.}
-        seeking to reconcile genetic and paleoanthropological evidence \citep{Henn2018-rf}.
-        \aprcomment{needs a caption}
+        \textbf{Proposed conceptual models of early human history in Africa.}
+        These models have been designed to translate models from the 
+        paleoanthropological literature into genetically testable demographic
+        models \citep{Henn2018-rf}.
+        We used these conceptual models as starting points to build detailed
+        parameterized demographic models (Supp. Information section~\ref{modelspec})
+        that were then fit to genetic data. 
     }

### 2022-03-11 paper.tex
Made a few small edits to the introduction

@@
 
-Archaeological sites from the Middle Stone Age (approx. 300ka-40ka) are widely
+Archaeological sites from the Middle Stone Age (approx. 300-40ka) are widely
 distributed across Africa, and are particularly well represented in the
-northern, eastern and southern parts of the continent. Similarly, fossil crania
-such as those from the sites of Jebel Irhoud \citep{Hublin2017-cq}, Herto
-\citep{White2003-bk} and Klasies River \citep{Deacon1995-rx} demonstrate that
+northern, eastern and southern parts of the continent. Similarly, fossils 
+such as those from the sites of Jebel Irhoud, Morocco \citep{Hublin2017-cq}, Herto, Ethiopia
+\citep{White2003-bk} and Klasies River, South Africa \citep{Deacon1995-rx} demonstrate that
 anatomically derived \emph{Homo sapiens} features were also present across the
@@
 \emph{Homo sapiens} by which populations in many regions of the continent
-contributed to the formation of Homo sapiens beginning at least 300ka
+contributed to the formation of /emph{Homo sapiens} beginning at least 300ka
 \citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
@@
 have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
-population coalescence and fragmentation \citep{Scerri2019-xg}. These
+population coalescence and fragmentation \citep{Scerri2019-xg}, but these
 approaches are more challenging to interpret and fit to data.
@@
 population, possible links to specific fossils, and the possibility of
-finding ancient DNA evidence \citep{Hsieh2016-gk}). However, these prior
+finding ancient DNA evidence \citep{Hsieh2016-gk}. However, these prior
 articles share two weaknesses. First, they only contrast a single-origin model

### 2022-03-11 paper.tex
A few small changes to the last paragraph

@@
 The fossil record of Africa is sparse during the time period of the stems, but of the
-available fossils, some are very similar to contemporary humans
+available fossils, some are very similar in morphology to contemporary humans
 (e.g., from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}),
-others are similar in some features but not others
+others are similar in some morphological features but not others
 (e.g., from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}),
-and others are very different
-(e.g., from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}) in morphology.
+and others are very different in morphology
+(e.g., from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}).
 If, as our model predicts, the

### 2022-03-12 paper.tex
added merger model figure

@@
 populations, and in both cases we tested two different types of gene exchange
-during the expansion phase. 1) One of the stem population expands (splits into
+during the expansion phase, as illustrated in Figure \ref{fig:mergers}: 1) One of the stem population expands (splits into
 contemporary populations), and the other stem population(s) has continuous

### 2022-03-12 supplement.tex
added merger model figure

@@
     \centering
+    \includegraphics{figures/mergers.pdf}
+    \caption{
+        \textbf{Two scenarios for population rearrangement.}
+        In our demographic model, we considered A) continuous migration models, in which one of the stem population expands (splits into
+contemporary populations), and the other stem population(s) has continuous
+symmetric migration with those populations; and B) one or more of the stem
+populations expands, with instantaneous pulse (or ``merger'') events from the
+other stem population, so that recent populations are formed by mergers of multiple ancestral populations.
+    }
+    \label{fig:mergers}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
     \includegraphics{figures/supp-omni-hapmap-comparison.pdf}

### 2022-03-12 paper.tex
minor edits to intro

--- 2022-03-11 paper.tex
@@
 \author[6,$\dag$,**]{Simon Gravel}
-\affil[1]{Department of Integrative Biology, University of Wisconsin, Madison, WI, USA}
+\affil[1]{Department of Integrative Biology, University of Wisconsin--Madison, WI, USA}
 \affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
@@
 
-Archaeological sites from the Middle Stone Age (approx. 300-40ka) are widely
+Archaeological sites from the Middle Stone Age (approx. 300ka-40ka) are widely
 distributed across Africa, and are particularly well represented in the
-northern, eastern and southern parts of the continent. Similarly, fossils 
-such as those from the sites of Jebel Irhoud, Morocco \citep{Hublin2017-cq}, Herto, Ethiopia
-\citep{White2003-bk} and Klasies River, South Africa \citep{Deacon1995-rx} demonstrate that
+northern, eastern and southern parts of the continent. Similarly, fossil crania
+such as those from the sites of Jebel Irhoud \citep{Hublin2017-cq}, Herto
+\citep{White2003-bk} and Klasies River \citep{Deacon1995-rx} demonstrate that
 anatomically derived \emph{Homo sapiens} features were also present across the
@@
 \emph{Homo sapiens} by which populations in many regions of the continent
-contributed to the formation of /emph{Homo sapiens} beginning at least 300ka
+contributed to the formation of Homo sapiens beginning at least 300ka
 \citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
@@
 have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
-population coalescence and fragmentation \citep{Scerri2019-xg}, but these
+population coalescence and fragmentation \citep{Scerri2019-xg}. These
 approaches are more challenging to interpret and fit to data.
@@
 single-origin models
-\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
+\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}.
 This has driven speculation about the geographic range of this ghost
 population, possible links to specific fossils, and the possibility of
-finding ancient DNA evidence \citep{Hsieh2016-gk}. However, these prior
+finding ancient DNA evidence (e.g.\citep{Hsieh2016-gk}). However, these prior
 articles share two weaknesses. First, they only contrast a single-origin model

### 2022-03-12 supplement.tex

--- 2022-03-11 supplement.tex
@@
     \centering
+    \includegraphics{figures/mergers.pdf}
+    \caption{
+        \textbf{Two scenarios for population rearrangement.}
+        In our demographic models, we considered
+        A) continuous migration models, in which one of the stem population expands 
+        (splits into contemporary populations), and the other stem population(s)
+        has continuous symmetric migration with those populations;
+        and B) one or more of the stem populations expands, with instantaneous pulse
+        (or ``merger'') events from the other stem population,
+        so that recent populations are formed by mergers
+        of multiple ancestral populations.
+    }
+    \label{fig:mergers}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
     \includegraphics{figures/supp-omni-hapmap-comparison.pdf}
     \caption{
-        \textbf{Comparison of LD statistics parsed using two recombination maps}
+        \textbf{Comparison of LD statistics parsed using two recombination maps.}
         The recombination map determines genetic distances between pairs of SNPs,
@@
         stem migration models predict a period of increased $F_{ST}$ between ancestors
-        of contemporary humans. This is largely due to the very small inferred $N_e$ in one of
+        of contemporary humans.
+        This is largely due to the very small inferred $N_e$ in one of
         the stem branches, which leads to a rapid increase in differentiation due to
@@
         \textbf{Conditional SFS compared to continuous-migration model predictions.}
-        The continuous-migration model provides a better fit to the Neanderthal-conditional SFS  than
+        The continuous-migration model provides a better fit to the
+        Neanderthal-conditional SFS than
         either the single-origin model or the merger-without-stem-migration

### 2022-03-12 paper.tex
a final read-through for small errors

@@
 \emph{Homo sapiens} by which populations in many regions of the continent
-contributed to the formation of /emph{Homo sapiens} beginning at least 300ka
+contributed to the formation of \emph{Homo sapiens} beginning at least 300ka
 \citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
@@
 
-Here, we therefore aim to discriminate among a broader set of demographic
-models by interrogating the genomes of contemporary populations. We take as our
+We therefore aim to discriminate among a broader set of demographic
+models by studying the genomes of contemporary populations. We take as our
 starting point 4 classes of models (single population expansion, single
@@
 Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
-\citep{Gurdasani2015-qy,Gopalan2022-pw}), and Eastern African agriculturalists
+\citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
 (Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
@@
 Fixed parameters included
-the time of divergence between Western and Eastern African populations,
+the time of divergence between western and eastern African populations,
 set to 60ka, just prior to the split of Eurasians and East Africans set to 50ka. 
 We also fixed the amount of admixture from Neanderthals to Europeans directly
-following the out-of-Africa migration which was set to 1.5\% at 45ka (Supp. Information). These constraints allowed us to integrate information from previous genetic and archaelogical research,
-to infer robust migration rates. For example, all models infer relatively high gene flow between Eastern and Western Africa
-($m\approx2\times10^{-4},$ the proportion of migrant lineages per generation). We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
+following the out-of-Africa migration which was set to 1.5\% at 45ka (Supp. Information).
+These constraints allowed us to integrate information from previous genetic
+and archaelogical research to infer robust migration rates.
+For example, all models infer relatively high gene flow between eastern and western Africa
+($m\approx2\times10^{-4},$ the proportion of migrant lineages per generation).
+We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
 affected the ancestors of the Ethiopian agricultural populations, comprising
@@
 years also saw major demographic changes, including strong population growth
-for Western Africans as they specialized in yam and oil palm agriculture
+for western Africans as they specialized in yam and oil palm agriculture
 (estimated 3-fold growth). We observe significant gene flow from the Amhara and
 Oromo into the Nama, a signal which is likely a proxy for the movement of
-Eastern African caprid and cattle pastoralists
+eastern African caprid and cattle pastoralists
 \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
 contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
-was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE} (Figure~\ref{fig:diversity}).
+was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
+(Figure~\ref{fig:diversity}).
 
@@
 populations, and in both cases we tested two different types of gene exchange
-during the expansion phase, as illustrated in Figure \ref{fig:mergers}: 1) One of the stem population expands (splits into
+during the expansion phase, as illustrated in Figure \ref{fig:mergers}:
+1) One of the stem population expands (splits into
 contemporary populations), and the other stem population(s) has continuous
@@
 populations expands, with instantaneous pulse (or ``merger'') events from the
-other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
+other stem population, so that recent populations are formed by mergers
+of multiple ancestral populations. Depending on parameter values,
 this scenario encompasses archaic hominin introgression and
-fragmentation-and-coalescence models (such as Figure~\ref{fig:proposed-models}A and C).
+fragmentation-and-coalescence models (such as Figure~\ref{fig:proposed-models}B and C).
 For many parameters, confidence intervals based on bootstrapping are relatively narrow
@@
 reflecting an informative statistical approach. Model assumptions
-have, however, a larger impact on parameter estimates (and thus, real uncertainty). To convey model uncertainty, we
-highlight features of the two inferred models with high likelihoods. These are referred to as the
+have, however, a larger impact on parameter estimates (and thus, real uncertainty).
+To convey model uncertainty, we highlight features of the two inferred models
+with high likelihoods. These are referred to as the
 ``multiple-merger'' and the ``continuous-migration'' models. Both allow for
@@
 Allowing for continuous migration between the stem populations substantially
-improved the fits relative to zero migration ($LL \approx -102,600$ vs.
-$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
-continuous migration). With continuous migration between stems, population
+improved the fits relative to zero migration between stems
+($LL \approx -102,600$ vs. $-107,700$ in the merger model and
+$LL \approx -115,500$ vs. $-126,600$ in the continuous migration model).
+With continuous migration between stems, population
 structure extends back to 1.1--1.4Ma (Table~\ref{tab:migration-rates}).
@@
 migrant lineages each generation estimated as
-$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
+$m=6.4\times10^{-5}-1.2\times10^{-4}$. For comparison, this is similar
 to inferred migration rates between connected contemporary populations over the
@@
 proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
-equal proportions (50\% each) to form the ancestors of the Western Africans and
-Eastern Africans (and thus also all individuals who later disperse during the
+equal proportions (50\% each) to form the ancestors of the western Africans and
+eastern Africans (and thus also all individuals who later disperse during the
 Out of Africa event). Finally, the Mende receive a large additional pulse of
 gene flow from Stem 2, replacing 18\% of their population 25ka.
-The later Stem 2 contribution to the Western African Mende resulted in
+The later Stem 2 contribution to the western African Mende resulted in
 better model fits ($\Delta LL \approx 60,000$).
-This may indicate that an ancestral Stem 2 population occupied Western or
+This may indicate that an ancestral Stem 2 population occupied western or
 Central Africa, broadly speaking. The differing proportions in the Nama and
-Eastern Africans may also indicate geographic separation of Stem 1S in southern
+eastern Africans may also indicate geographic separation of Stem 1S in southern
 Africa and Stem 1E in eastern Africa.
@@
 \ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
-even though this statistic was not used in the fit. Both models actually outperform
+even though this statistic was not used in the fit. Our best-fit models outperform
 archaic hominin admixture models fit directly to the cSFS
-(for example, compare with Figure~1 in
+(for example, compare to Figure~1 in
 Durvasula and Sankararaman (2020)\citep{Durvasula2020-td}).
@@
 in $N_e$ during the same period \citep{Li2011-le}.
-This increase in $N_e$ could be due to either an increase in population or to 
+This increase in inferred $N_e$ could be due to either an increase in population size or to 
 ancestral population structure during the Middle Stone Age \citep{Mazet2015-wn}.
@@
 (Supp. Information).
-During the Middle Stone Age, the Multiple Merger indicates three
+During the Middle Stone Age, the multiple merger model indicates three
 major stem lineages in Africa, tentatively assigned to southern (Stem 1S),

### 2022-03-12 supplement.tex
a final read-through for small errors

@@
 University. Results stemming from genetic analyses have been communicated in
-2015, 2019, 2021 via community presentations, a radio interview and to
+2015, 2019 and 2021 via community presentations, a radio interview and to
 representatives of the Richtersveld National Park.
@@
 
-From the merged dataset of the African Diversity Reference Panel (ADRP,
-citet{}) and the \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data,
+From the merged dataset of the African Diversity Reference Panel (ADRP)
+\citep{Gurdasani2015-qy,Pagani2015-pz}
+and the \citet{1000_Genomes_Project_Consortium2015-zq} phase 3 data,
 we subsampled to the populations included in this study. From the 1000 Genomes
@@
 appear less affected by natural selection compared to both synonymous and
-nonsynonymous variation \citep{Ragsdale2018-dd}. To enable comparison with
+nonsynonymous variation \citep{Ragsdale2018-dd}, removing regions that fell
+outside of the 1000 Genomes strict callability mask. To enable comparison with
 Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
@@
 
-We used principal component analysis (PCA) from \texttt{scikit-learn}
+We used principal component analysis (PCA) implemented in \texttt{scikit-learn}
 \citep{Pedregosa2011-ke} and \texttt{ADMIXTURE} version 1.3.0
@@
 expected heterozygosity, assuming random mating), denoted $\E[H] =
-\E[2p_A(1-p_A)] = \E[2p_B(1-pB)]$, assuming equal mutation rates at the two
+\E[2p_A(1-p_A)] = \E[2p_B(1-p_B)]$, assuming equal mutation rates at the two
 loci.
@@
 \citep{Plagnol2006-lt}. Pairwise diversity statistics $H_{i,j}$ have also been
-widely used in demographic inference involving ancient DNA and many samples, as
+widely used in demographic inference involving ancient DNA and many statistics, as
 $f_2$, $f_3$ and $f_4$ statistics can be expressed as linear combinations of
@@
 inferred using array data, which is sparser than whole-genome sequencing data.
-In subsequent analyses, we used data based on recombination rates fromthe Omni YRI map.
+In subsequent analyses, we used data based on recombination rates from the OMNI YRI map.
 
@@
 as some LD statistics involving more than one population. For example,
-statistics of the form $D_{human}(1-2p_{human})(1-2q_{neanderthal})$ require a
+statistics of the form
+$D_{\text{human}}(1-2p_{\text{human}})(1-2q_{\text{neanderthal}})$ require a
 single Neanderthal sample and are informative of the Neanderthal demography.
@@
 \citep{Chen2007-iy}. By contrast, a U-shaped distribution has been taken as
-evidence for introgression from an archaic hominin population that split from humans at least as far back in time as the human-Neanderthal split
-\citet{Durvasula2020-td,Yang2012-ze}.
+evidence for introgression from an archaic hominin population that split
+from humans at least as far back in time as the human-Neanderthal split
+\citep{Yang2012-ze,Durvasula2020-td}.
 Sites with no calls in the Vindija Neanderthal were excluded from
@@
 human-Neanderthal lineages and the more recent split of branches leading to
-Southern and West/Eastern African populations.
+southern and western/eastern African populations.
 
@@
 From the reconstructed gene genealogies, we computed coalescence rates within
-and between populations using \Relate’s function \texttt{RelateCoalescentRate}.
+and between populations using \texttt{Relate}’s function \texttt{RelateCoalescentRate}.
 This allows for an estimate of the instantaneous inverse coalescence rate
@@
 \end{equation}
-
 where $H_{ij}$ is the average number of differences between one haploid copy of
@@
 are the average number of differences for two haploid copies from the same
-population. In all inferred models, $F_{ST}$s between Neanderthals and the
+population.
+
+In all inferred models, $F_{ST}$ between Neanderthals and the
 ancestors of contemporary humans were much larger than between any pair of contemporary
@@
 stem populations was inferred to have a short, severe bottleneck, which led to
-increased $F_{ST}$ between stem populations during that time. In the merger
-with stem migration, $F_{ST}$ reached a maximum value of $0.3$ between stems,
+increased $F_{ST}$ between stem populations during that time. In the
+merger-with-stem-migration, $F_{ST}$ reached a maximum value of $0.3$ between stems,
 intermediate between human-Neanderthal divergence ($0.45$)
@@
 
-\subsection{\bm{$f_4$} statistics between pairs of contemporary
+\subsection{Shared differentiation between pairs of contemporary
 and pairs of ancient populations}
 \label{sec:f4}
-
-%\tdwcomment{I have tried to make the terminology about human, hominin, etc.
-%consistent and unambiguous. To do this, I have avoided using "modern" because 
-%"modern human" can mean "anatomically modern human" or \emph{Homo sapiens}, 
-%but mostly in the paper we use it to mean "contemporary" or "present-day". I 
-%think we should use "contemporary human" or "present-day human" when we mean 
-%the latter and just "human" when we mean the former. I have made these changes 
-%in the main text and supplement, preferentially using "contemporary" but 
-%leaving "present-day" when it was already there. However, I think it would be 
-%best if Simon made these changes in the equations in this section, so I don't mess 
-%something up.}
 
@@
 population, and we used \msprime \citep{Baumdicker2022-mj} to simulated
-500 Mb of total sequence length (the sum over500 replicates each 1 Mb in length)
+500 Mb of total sequence length (the sum over 500 replicates each 1 Mb in length)
 with constant recombination rate $r=10^{-8}$ and mutation rate $u=2\times10^{-8}.$

### 2022-03-12 paper.tex
spelling

@@
 models by studying the genomes of contemporary populations. We take as our
-starting point 4 classes of models (single population expansion, single
+starting point four classes of models (single population expansion, single
 population expansion with regional persistence, archaic hominin admixture, and
@@
 data for four diverse African populations, comprising the Nama (Khoe-San from
-South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
+South Africa, newly presented here), Mende (from Sierra Leone, MSL from the
 Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),

### 2022-03-14 paper.tex
origined -> originated

@@
 While it is now broadly accepted that \emph{Homo sapiens}
-origined within Africa, considerable
+originated within Africa, considerable
 uncertainty surrounds specific models of divergence and migration across the

### 2022-03-14 paper.tex
spelling in main text

@@
 Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
-Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
+Gumuz (recent descendants of a hunter-gatherer group from Ethiopia
 \citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
@@
 These constraints allowed us to integrate information from previous genetic
-and archaelogical research to infer robust migration rates.
+and archaeological research to infer robust migration rates.
 For example, all models infer relatively high gene flow between eastern and western Africa

### 2022-03-14 supplement.tex
spelling in supplement

@@
 \texttt{Admixture} analyses. The \texttt{Relate} analysis used a slightly
-smaller set, discluding the Bakiga, Somali, and Zulu samples.
+smaller set, excluding the Bakiga, Somali, and Zulu samples.
 
@@
 ancestry); and Europe: British (GBR). These were combined with the
-high-coverage Vindija Neanderhal genome \citep{Prufer2017-kk}, resulting in 290
+high-coverage Vindija Neanderthal genome \citep{Prufer2017-kk}, resulting in 290
 total samples used in demographic inference.
@@
 
-We used multi-population linkage disequilibrium (LD) and pairwise diverisity
+We used multi-population linkage disequilibrium (LD) and pairwise diversity
 statistics to fit demographic models to data. These statistics, introduced and
@@
 from a population to compute. Since we used a single Neanderthal sample, such
-statistics for the Neandertal population were not used in the fit. By contrast,
+statistics for the Neanderthal population were not used in the fit. By contrast,
 there are statistics that only require a single sample in a given population to
@@
 For the best fit parameters, we computed confidence intervals using the 
-same set of bootstraped samples and Godambe Information
+same set of bootstrapped samples and Godambe Information
 approach \citep{Coffman2016-yq}, which corrects composite likelihood estimates of confidence intervals
@@
         cSFS results in a poorer fit to the cSFS compared to the continuous-migration
-        and merger-with-stem-migratino models. Thus the model without stem migrations
+        and merger-with-stem-migration models. Thus the model without stem migrations
         provides a poorer fit 
@@
         from the full dataset (\ref{fig:supp-rccr-data}, top) for distant
-        timepoints. There are large discrepancies in the very recent history,
+        time points. There are large discrepancies in the very recent history,
         where Relate and other genome-wide coalescence-based methods are known
@@
         Relate-inferred gene genealogies. The proportion of deep branches with
-        Neandethal affinity is highest for Eurasian populations, as expected,
+        Neanderthal affinity is highest for Eurasian populations, as expected,
         and higher for African populations with recent Eurasian ancestry

### 2022-03-14 paper.tex
Update paper.tex

@@
 continent. Progress is hampered by a paucity of fossil and genomic data, as
-well as variability in prior dating estimates.
+well as variability in prior divergence time estimates.
 Here we use linkage disequilibrium and
@@
 \citep{White2003-bk} and Klasies River, South Africa \citep{Deacon1995-rx} demonstrate that 
-anatomically derived \emph{Homo sapiens} features were also present across the
+derived \emph{Homo sapiens} anatomical features were also present across the
 continent during this period. It has been difficult to reconcile these lines of

### 2022-06-03 supplement.tex
Add Neand column to lineage probs

@@
     avoid the effects of very recent admixture, such as recent European admixture in
-    the Nama. Rows may not sum to one, as not every branch is listed here (such as
-    Neanderthal).
-}
-\centering
-\begin{tabular}[t]{lccccccc}
+    the Nama.
+}
+\centering
+\begin{tabular}[t]{lcccccccc}
     \toprule
@@
     \specialcell{Gumuz\\branch} & \specialcell{Oromo/Amhara\\branch} & 
-    \specialcell{British\\branch} & Stem 1 & Stem 2\\
+    \specialcell{British\\branch} & Neand. & Stem 1 & Stem 2\\
     \midrule
     Found 10ka & & & & & & & \\
-    Nama & 0.724 & 0.015 & 0.032 & 0.226 & 0.003 & & \\
-    Mende & 0.003 & 0.881 & 0.058 & 0.058 & 0.001 & & \\
-    Gumuz & 0.013 & 0.058 & 0.819 & 0.098 & 0.012 & & \\
-    Oromo/Amhara & 0.013 & 0.058 & 0.098 & 0.819 & 0.012 & & \\
+    Nama & 0.724 & 0.015 & 0.032 & 0.226 & 0.003 & & & \\
+    Mende & 0.003 & 0.881 & 0.058 & 0.058 & 0.001 & & & \\
+    Gumuz & 0.013 & 0.058 & 0.819 & 0.098 & 0.012 & & & \\
+    Oromo/Amhara & 0.013 & 0.058 & 0.098 & 0.819 & 0.012 & & & \\
     \midrule
     Found 55ka & & & & & & & \\
-    Nama & 0.677 & 0.052 & 0.264 & & & & 0.005 \\
-    Mende & 0.021 & 0.562 & 0.272 & & & & 0.143 \\
-    Gumuz & 0.060 & 0.211 & 0.704 & & & & 0.023 \\
-    Oromo/Amhara & 0.041 & 0.142 & 0.793 & & & & 0.017 \\
+    Nama & 0.677 & 0.052 & 0.264 & & & 0.002 & & 0.005 \\
+    Mende & 0.021 & 0.562 & 0.272 & & & 0.001 & & 0.143 \\
+    Gumuz & 0.060 & 0.211 & 0.704 & & & 0.002 & & 0.023 \\
+    Oromo/Amhara & 0.041 & 0.142 & 0.793 & & & 0.008 & & 0.017 \\
     \midrule
-    Found 500ka & & & & & & & \\
-    Nama & & & & & & 0.379 & 0.618 \\
-    Mende & & & & & & 0.430 & 0.569 \\
-    Gumuz  & & & & & & 0.473 & 0.525 \\
-    Oromo/Amhara & & & & & & 0.475 & 0.517 \\
+    Found 500ka & & & & & & & & \\
+    Nama & & & & & & 0.002 & 0.379 & 0.618 \\
+    Mende & & & & & & 0.001 & 0.430 & 0.569 \\
+    Gumuz  & & & & & & 0.002 & 0.473 & 0.525 \\
+    Oromo/Amhara & & & & & & 0.008 & 0.475 & 0.517 \\
     \bottomrule

### 2022-06-09 supplement.tex
back-to-Africa migration

@@
 \end{figure}
+
+\subsection{Back-to-Africa Gene flow and Neanderthal ancestry in Africa}
+
+Back-to-Africa migrations likely introduced Neanderthal ancestry into Africa \cite{Chen2020}.
+This Neanderthal ancestry would cause reticulation in the ancestry of the African populations, and is therefore relevant to the modelling presented here.
+Our models do account for both Neanderthal admixture out of Africa and Back-to-Africa migrations, such that they predict the amount of Neanderthal ancestry that we would expect to find in African individuals.
+We find that the MSL are predicted to carry \sgcomment{TODO} $XX-YY \%$ Neanderthal ancestry, which is remarkably consistent with the results of \cite{Chen2020}.
+Since the IBD statistics from \cite{Chen2020} are not used in the present analysis, this serves as an independent validation of the present models.
+More importantly, since the Neanderthal admixture in Africa is properly accounted for in the model, we do not expect the presence of such ancestry to bias our estimate of reticulation in Africa.
 

### 2022-06-09 paper.tex
reticulation IM

@@
 Regardless of the model choice for early epochs, inference of human demographic
-history for the last 150ka was remarkably robust. The earliest divergence among
+history for the last 150ka was remarkably robust. In a reticulated model, we use
+"divergence" between populations to describe the time of their most recent shared ancestry.  
+The earliest divergence among
 contemporary human populations differentiates the southern African Nama from
@@
 
+Other studies have fit tree-like demographic models to African populations using 
+distributions of allele frequencies or related statistics, finding inconsistent divergence times \cite{Henn2018, Bergstrom2021}.  
+Section \ref{sec:IM-reinfer} explains this behaviour: 
+split times inferred using isolation with migration models and distributions of allele frequencies simulated under our reticulated 
+models are highly variable and much deeper than the more recent recent population divergence.
+
 \section*{Discussion}

### 2022-06-17 paper.tex
fill in neand ancestry details

--- 2022-03-14 paper.tex
@@
 Regardless of the model choice for early epochs, inference of human demographic
-history for the last 150ka was remarkably robust. The earliest divergence among
+history for the last 150ka was remarkably robust.
+In a reticulated model, we use ``divergence'' between populations to describe
+the time of their most recent shared ancestry.
+The earliest divergence among
 contemporary human populations differentiates the southern African Nama from
@@
 
+Other studies have fit tree-like demographic models to African populations using 
+distributions of allele frequencies or related statistics,
+finding inconsistent divergence times \cite{Henn2018, Bergstrom2021}.
+Section~\ref{sec:IM-reinfer} explains this behaviour:
+split times inferred using isolation with migration models and
+distributions of allele frequencies simulated under our reticulated
+models are highly variable and much deeper than the more recent
+estimated population divergence.
+
 \section*{Discussion}

### 2022-06-17 supplement.tex
fill in neand ancestry details

--- 2022-06-03 supplement.tex
@@
 \end{figure}
+
+\subsection{Back-to-Africa Gene flow and Neanderthal ancestry in Africa}
+
+Back-to-Africa migrations likely introduced Neanderthal ancestry into Africa
+\citep{Chen2020-wh}. This Neanderthal ancestry would cause reticulation in the
+ancestry of the African populations, and is therefore relevant to the modeling
+presented here. The models we inferred account for both Neanderthal admixture
+in Eurasia and Back-to-Africa migrations, such that they predict the amount of
+Neanderthal ancestry that we would expect to find in African individuals. We
+find that African populations without high proportions of recent Eurasian
+ancestry are predicted to carry $0.1-0.2\%$ Neanderthal
+ancestry (Table~\ref{tab:supp-lineage-probabilitys}),
+which is remarkably consistent with the results of
+\citet{Chen2020-wh}. Since the IBD statistics from \citet{Chen2020-wh} are not
+used in the present analysis, this serves as an independent validation of the
+present models. More importantly, since the Neanderthal admixture in Africa is
+properly accounted for in the model, we do not expect the presence of such
+ancestry to bias our estimate of reticulation in Africa.
 

### 2022-07-15 supplement.tex
start adding supplementary results

@@
 
+\aprcomment{TODO: expand this section to include further discussion about why
+we chose each population, and why we have omitted some (such as those suggested
+in the reviews).}
+
 \subsection{Filtering and subsetting data}
@@
 
-\subsection{PCA and \texttt{ADMIXTURE} analysis}
+\subsection{PCA and \texttt{ADMIXTURE} analysis}\label{sec:dimred}
 
@@
 recombination distances. We assessed the robustness of the statistics to errors
-in the recombination maps by considering two different recombination maps, the
-OMNI YRI and HapMapII
-\citep{1000_Genomes_Project_Consortium2015-zq,International_HapMap_Consortium2007-vn}.
+in the recombination maps by considering three different recombination maps,
+the OMNI YRI and HapMapII
+\citep{1000_Genomes_Project_Consortium2015-zq,International_HapMap_Consortium2007-vn},
+and a recombination map inferred from the Nama \citep{Van_Eeden2021-od}.
 Statistics were largely unchanged by using a different map
-(Figure~\ref{fig:supp-omni-hapmap-comparison}), although both maps were
-inferred using array data, which is sparser than whole-genome sequencing data.
-In subsequent analyses, we used data based on recombination rates from the OMNI YRI map.
+(Figures~\ref{fig:supp-omni-hapmap-comparison},~\ref{fig:supp-omni-nama-comparison}).
+The OMNI and HapMap recombination maps resulted in LD statistics that were
+nearly identical, with only minor deviations between the OMNI/HapMap and the
+Nama-inferred recombination maps. Both the OMNI and HapMap maps were inferred
+using array data, which is sparser than whole-genome sequencing data as was
+used in the Nama map inference. In subsequent analyses, we used data based on
+recombination rates from the OMNI YRI map, and in the Supplemental Results we
+show that the inferences are robust when using the Nama recombination map
+(Section~\ref{sec:supp-alt-data}).
 
@@
 were restricted to isolation between stem branches. Depending on the specific
-parameters, such models encompass commonly-considered scenarios for admixture from a ghost archaic hominin population (e.g., if a long-isolated lineage more recently contributes a
-minority of ancestry to one or more populations), as well as relatively simple
-fragmentation-coalescence scenarios.
+parameters, such models encompass commonly-considered scenarios for admixture
+from a ghost archaic hominin population (e.g., if a long-isolated lineage more
+recently contributes a minority of ancestry to one or more populations), as
+well as relatively simple fragmentation-coalescence scenarios.
 
@@
 
-\subsection{Uncertainty and confidence intervals using Godambe methods}
-
-To compute likelihoods, we needed an estimate of the joint distribution of summary statistics. 
+\subsection{Uncertainty and confidence intervals}
+
+To compute likelihoods, we needed an estimate of the joint
+distribution of summary statistics. 
 We estimated uncertainty due to the finite amount of genetic material
@@
 
-\section{Validations using simulations from inferred demographic models}
+\section{Model validation}
+
+We took two approaches to validate demographic model inferences. First, we
+explored the effects of using alternative populations, sample sets, and
+recombination maps, and the effect of relaxing fixed parameters when fitting
+the demographic models to diversity and LD statistics. We then use simulations
+to test our inferred model predictions against other informative summary
+statistics. Results are shown in the Supplementary Results
+(Section~\ref{sec:supp-results}).
+
+\subsection{Validations using alternative data sets and model specification}
+
+\subsubsection{Alternate Thousand Genomes population sets}
+
+In our primary model fits, we used the Mende in Sierra Leone (MSL) and the
+British from England and Scotland (GBR). We tested the same models against a
+different West African population (the Yoruba from Ibidan, Nigeria, YRI), and
+two different European populations (Western Europeans from Utah, CEU, and
+Toscani in Italy, TSI). The CEU are closely related to GBR and are thus likely
+to be an equivalently close proxy for European ancestry in African individuals
+with admixed European ancestry. The TSI are likely more distantly related to
+the source of this European ancestry. We refit each of our best-fit demographic
+models using the following alternate population sets: (Nama, YRI, Gumuz,
+Oromo/Amhara, and CEU), and (Nama, MSL, Gumuz, Oromo/Amhara, and TSI).
+
+\subsubsection{Variable thresholds of Nama-associated ancestry}
+
+The Nama have recent European ancestry, and Nama individuals in our dataset
+show varying levels of inferred ancestry relating to European populations
+(Figure~\ref{fig:diversity}D, in the main text). Each of our inferred models
+allows for gene flow from Eurasia to Africa after the out-of-Africa migration
+event, through early Holocene back-to-Africa migration in East Africa, recent
+colonial admixture in South Africa, and ongoing continuous migration between
+branches leading to East African and European populations. While our
+demographic models directly account for European ancestry in Nama individuals,
+inferred parameters relating to early divergences and population structure may
+be impacted by recent admixture.
+
+To assess the impact of recent European admixture in Nama individuals on
+inferred early divergence times and population structure, we subset the set of
+Nama individuals that we included based on inferred European-matching ancestry
+proportions. We used inferred ancestry proportions from \texttt{ADMIXTURE} with
+the number of clusters $K=4$ (Section~\ref{sec:dimred}) and kept individuals
+with either at least a) 90\% or b) 99\% proportion of inferred ancestry that
+clusters within the Nama group (colored green in Figure~\ref{fig:diversity}D).
+This left a) 18 Nama individuals with at least 90\% inferred South African
+ancestry, and b) 14 Nama individuals with at least 99\%. For each subset, we
+recomputed diversity and heterozygosity statistics and fit each of our
+demographic models to these statistics.
+
+\subsubsection{Varying fixed divergence times}
+
+In order to reduce parameter space to avoid overfitting, we fixed parameters
+that have previously been inferred, including the timings of the early
+back-to-Africa migration contributing to the ancestry of East African
+agricultural populations (12ka), the divergence of West and East Africans
+(60ka), and the out-of-Africa migration event (50ka). Due to variability in
+inference and dating methods, there is some uncertainty surrounding these
+dates. We explored how variable timings of these events effect our inferred
+models by 1) increasing the fixed times of the E-W African split (to 100ka) and
+the out-of-Africa event (to 60ka), and 2) allowing those two parameters to be
+jointly fit in model optimization.
+
+\subsubsection{Recombination map inferred from Nama}
+
+Throughout, we used inferred recombination maps based on the YRI OMNI array
+data \citep{1000_Genomes_Project_Consortium2015-zq}. Because recombination
+rates can vary between populations, we tested the robustness of our demographic
+inference to the recombination map used to compute LD. We recomputed diversity
+and LD statistics using recombination rates from the Nama recombination map
+\citep{Van_Eeden2021-od}, and refit our best-fit demographic models to these
+statistics.
+
+\subsubsection{\aprcomment{TODO:} High-coverage Nama samples}
+
+\subsection{Validations using simulations from inferred demographic models}
 
@@
 
-\subsection{Simulation details}
+\subsubsection{Simulation details}
 
@@
 
-\subsection{cSFS prediction under inferred models}
+\subsubsection{cSFS prediction under inferred models}
 
@@
 
-\section{Supplementary results}
+\section{Supplementary results}\label{sec:supp-results}
+
+\subsection{Model fit using alternative datasets}
+\label{sec:supp-alt-data}
+
+\begin{itemize}
+    \item Using YRI and CEU, inferred parameters for each of our highlighted models
+        did not change (all well within one SD of the inferred parameters of the
+        original fit models)
+    \item TODO: run fits with TSI
+    \item Fits using the Nama recombination map, even though there were slight
+        differences in some of the LD statistics (primarily a few noticeable
+        differences in the $\pi_2$ statistics), also resulted in model fits that
+        were very similar to the original fit models.
+    \item Fits with alternate thresholds of Nama-inferred ancestry: TODO:
+        fit continuous migration and the merger with stem migration model
+\end{itemize}
+
+\subsection{Relaxing fixed divergence times}
+
+\begin{itemize}
+    \item Fixed to older dates (100ka: E/W split, 60ka: OOA event). These fixed
+        values stretched the other inferred parameters to older times, and resulted
+        in a worse log-likelihood
+    \item Allowed to vary, the best-fit inferred E/W split time and OOA event
+        were more recent 35-50ka, instead of our fixed 50-60ka. \aprcomment{I
+        still don't find this entirely unreasonable, as E Africans and Eurasian
+        populations may be more closely related than the oldest evidence of H.
+        sapiens in Europe (up to 50ka), or those earliest H. sapiens are likely
+        not to be the direct ancestors of present-day Europeans. Still, it's
+        probably too recent of a split, and fixing to 50ka is defensible.}
+\end{itemize}
 

### 2022-07-15 supplement.tex
added label

--- 2022-06-09 supplement.tex
@@
 $\pi_2$), which removes any dependence on the mutation rate. Thus, statistics
-take the form $\sigma_{d;i,j}^2 = \frac{\E[D^2_{i,j}]}{\E[\pi_2]}$,
-$\sigma_{Dz;i,j,k}=\frac{\E[Dz_{i,j,k}]}{\E[\pi_2]}$, and so on.
+take the form 
+\begin{equation}
+\begin{split}
+\sigma_{D;i,j}^2 = \frac{\E[D^2_{i,j}]}{\E[\pi_2]}\\
+\sigma_{Dz;i,j,k}=\frac{\E[Dz_{i,j,k}]}{\E[\pi_2]},
+\end{split}
+\label{def_ld}
+\end{equation}
+and so on.
 Multi-population pairwise diversity statistics $\E[H_i]=\E[2p_i(1-p_i)]$ and
@@
 \subsubsection{Relate curves from inferred models}
-
+\label{inferred_relate}
 We used \Relate \citep{Speidel2019-nj} to reconstruct gene genealogies from the

### 2022-07-15 paper.tex
small edits to address reviewer comments

--- 2022-06-09 paper.tex
@@
 better model fits ($\Delta LL \approx 60,000$).
-This may indicate that an ancestral Stem 2 population occupied western or
+Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
 Central Africa, broadly speaking. The differing proportions in the Nama and
@@
 single-origin model, recapitulate an inferred ancestral
-increase in $N_e$ between 100ka-1Ma (Figure~\ref{fig:supp-iicr-sim}).
+increase in $N_e$ between 100ka-1Ma extending deeper in the past than the relate estimate (Figure~\ref{fig:supp-iicr-sim} and Section \ref{inferred_relate}).
 Whereas the single-origin model achieves this by
@@
 this pattern without corresponding population size changes.
-%\comment{These are pretty complicated ideas that are not explained in detail.
-%Might be worth unpacking a bit more to make sure the reader understands}
 
@@
 We cannot rule out that more complex models involving
-additional stems, or hybrid models including both weak structure and archaic
+additional stems, continuous structure, or hybrid models including both weak structure and archaic
 hominin admixture may better explain the data. Because parameters related to the split

### 2022-07-15 paper.tex

@@
 Regardless of the model choice for early epochs, inference of human demographic
-history for the last 150ka was remarkably robust. In a reticulated model, we use
-"divergence" between populations to describe the time of their most recent shared ancestry.  
+history for the last 150ka was remarkably robust.
+In a reticulated model, we use ``divergence'' between populations to describe
+the time of their most recent shared ancestry.
 The earliest divergence among
@@
 Other studies have fit tree-like demographic models to African populations using 
-distributions of allele frequencies or related statistics, finding inconsistent divergence times \cite{Henn2018, Bergstrom2021}.  
-Section \ref{sec:IM-reinfer} explains this behaviour: 
-split times inferred using isolation with migration models and distributions of allele frequencies simulated under our reticulated 
-models are highly variable and much deeper than the more recent recent population divergence.
+distributions of allele frequencies or related statistics,
+finding inconsistent divergence times \cite{Henn2018, Bergstrom2021}.
+Section~\ref{sec:IM-reinfer} explains this behaviour:
+split times inferred using isolation with migration models and
+distributions of allele frequencies simulated under our reticulated
+models are highly variable and much deeper than the more recent
+estimated population divergence.
 

### 2022-07-15 paper.tex
Addressing Reviewer #3 point #2

@@
 and archaeological research to infer robust migration rates.
-For example, all models infer relatively high gene flow between eastern and western Africa
-($m\approx2\times10^{-4},$ the proportion of migrant lineages per generation).
+For example, all models infer relatively high gene flow between Ethiopia and western Africa
+($m\approx2\times10^{-4},$ the constant proportion of migrant lineages per generation over 60ka).
 We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily

### 2022-07-15 paper.tex
Addressing Reviewer #3 point #3a

@@
         $LL= -115,500$ & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
-        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
+        (Table S4) & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
         & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
@@
         $LL= -102,600$ & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
-        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
+        (Table S6) & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
         & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
@@
 479ka. The timing of this divergence was also poorly constrained ($\sigma_T=
-166$ka). These populations evolve independently until approximately 119ka
-when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
+166$ka). These populations evolve independently until 119ka
+($\sigma_T=
+28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
 proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
@@
 Out of Africa event). Finally, the Mende receive a large additional pulse of
-gene flow from Stem 2, replacing 18\% of their population 25ka.
+gene flow from Stem 2, replacing 18\% of their population 25ka ($\sigma_T=
+0.6$ka).
 The later Stem 2 contribution to the western African Mende resulted in
@@
 contemporary groups. We observed two merger events. The first, between Stem 1S and
-Stem 2, results in the formation of an ancestral Khoe-San population ~120ka.
+Stem 2, results in the formation of an ancestral Khoe-San population $\approx120$ka.
 The second $\approx100$ka between Stem 1E and Stem 2, results in the formation of the

### 2022-07-15 supplement.tex
add some supplementary results

--- 2022-07-15 supplement.tex
@@
 with admixed European ancestry. The TSI are likely more distantly related to
-the source of this European ancestry. We refit each of our best-fit demographic
-models using the following alternate population sets: (Nama, YRI, Gumuz,
-Oromo/Amhara, and CEU), and (Nama, MSL, Gumuz, Oromo/Amhara, and TSI).
+the source of this European ancestry. We chose not to replace the GBR
+non-European Eurasian populations, as these would be poorer proxies for early
+Holocene and recent colonial migration from the Near East and Europe back to
+Africa, and it is unclear how such a model should be interpreted. We therefore
+refit each of our best-fit demographic models using the following alternate
+population sets: (Nama, YRI, Gumuz, Oromo/Amhara, and CEU), and (Nama, MSL,
+Gumuz, Oromo/Amhara, and TSI).
 
@@
 \end{itemize}
+
+\subsubsection{Replacing Thousand Genomes populations}
+
+We replaced the Mende (MSL) and British (GBR) population sets with Yoruba (YRI)
+and Western and Northern Europeans (CEU), respectively, and refit our inferred
+demographic models to diversity and LD statistics computed from those
+populations (along with Nama, Oromo/Amhara, and Gumuz). Because the MSL and YRI
+share recent West African ancestry, and the GBR and CEU share recent European
+ancestry, we expected most parameters in inferred models using these two
+datasets to be largely similar. Indeed, all parameters inferred using the YRI
+and CEU data were well within one standard error of the same parameters
+inferred using the MSL and GBR data, across all tested models. The one
+exception to this was the recent size of the YRI branch, inferred to be
+somewhat smaller than the final size of the analogous MSL branch ($\approx
+20,000$ vs. $\approx 30,000$, resp.). Other population sizes, migration rates,
+and divergence times were highly consistent.
+
+Because CEU and GBR are genetically closely related, we also tested the
+consistency of inferred parameters when replacing GBR with Toscani from Italy,
+a more distantly related European population set than CEU. \aprcomment{TODO}
+
+\subsubsection{Fit to LD statistics from Nama recombination map}
 
@@
     \caption{
-        \textbf{Comparison of LD statistics parsed using two recombination maps.}
+        \textbf{Comparison of LD statistics parsed using two array-based
+        recombination maps.}
         The recombination map determines genetic distances between pairs of SNPs,
@@
         (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
+    }
+    \label{fig:supp-omni-hapmap-comparison}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-omni-nama-comparison.pdf}
+    \caption{
+        \textbf{Comparison of LD statistics parsed using OMNI YRI and
+        Nama-specific recombination map.}
+        The OMNI YRI recombination map and the Nama-inferred map
+        \citep{Van_Eeden2021-od} differ in two primary ways: 1) the YRI and
+        Nama have relatively highly diverged ancestries, and 2) the YRI
+        map was inferred from low-density array data, while the Nama map
+        was inferred from whole-genome sequencing data.
+        Nonetheless, LD statistics computed under recombination rates from
+        the two maps remain largely concordant, and inferences are not
+        strongly affected by recombination map choice (Section~\ref{sec:supp-results}).
+        Shading represents 95\% confidence intervals from the Nama-inferred
+        map. Notations for each statistic are described in
+        section~\ref{sec:supp-ld-stats} and indexed by populations (0: Nama, 1:
+        Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
     }

### 2022-07-15 paper.tex

@@
 
-Archaeological sites from the Middle Stone Age (approx. 300-40ka) are widely
-distributed across Africa, and are particularly well represented in the
-northern, eastern and southern parts of the continent. Similarly, fossils 
-such as those from the sites of Jebel Irhoud, Morocco \citep{Hublin2017-cq}, Herto, Ethiopia
-\citep{White2003-bk} and Klasies River, South Africa \citep{Deacon1995-rx} demonstrate that 
-derived \emph{Homo sapiens} anatomical features were also present across the
-continent during this period. It has been difficult to reconcile these lines of
-evidence with evidence from genomics, which have suggested a predominantly
-tree-like model of recent population divergence from a single ancestral
-population. It is unclear whether fossil specimens and archaeological sites
-represent populations which contributed to our ancestors as population
-precedents, or were local ``dead-ends'' from which contemporary \emph{Homo
-sapiens} do not descend. 
-Recently,
-attempts to reconcile genetic and
-paleoanthropological data include proposals for a Pan-African origin of
-\emph{Homo sapiens} by which populations in many regions of the continent
-contributed to the formation of \emph{Homo sapiens} beginning at least 300ka
-\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
+Decades of study of human genome variation have suggested a predominantly tree-like model of recent population divergence from a single ancestral population in Africa.
+It has been difficult to reconcile this finding with the fossil and archaeological record of human occupation across the vast African continent.
+For example, fossils such as those from the sites of Jebel Irhoud, Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo sapiens} anatomical features were present across the continent 300-100ka.
+Archaeological sites from the Middle Stone Age, which has been associated with \emph{Homo sapiens} at some sites, are also widely distributed across Africa.
+It is unclear whether fossil specimens and archaeological sites represent populations which contributed to our ancestors as population precedents, or were local ``dead-ends'' from which contemporary \emph{Homo sapiens} do not descend.
+Recently, attempts to reconcile genetic and paleoanthropological data include proposals for a Pan-African origin of \emph{Homo sapiens} by which populations in many regions of the continent contributed to the formation of \emph{Homo sapiens} beginning at least 300ka \citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
 

### 2022-07-15 paper.tex
Addressing Reviewer #3 point 1f

@@
 It has been difficult to reconcile this finding with the fossil and archaeological record of human occupation across the vast African continent.
-For example, fossils such as those from the sites of Jebel Irhoud, Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo sapiens} anatomical features were present across the continent 300-100ka.
+For example, fossils such as those from the sites of Jebel Irhoud, Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo sapiens} anatomical features were intermittantly present across the continent 300-100ka.
 Archaeological sites from the Middle Stone Age, which has been associated with \emph{Homo sapiens} at some sites, are also widely distributed across Africa.
-It is unclear whether fossil specimens and archaeological sites represent populations which contributed to our ancestors as population precedents, or were local ``dead-ends'' from which contemporary \emph{Homo sapiens} do not descend.
+It is unclear whether these fossils and archaeological sites represent populations which contributed to contemporary \emph{Homo sapiens} as population precedents or were local ``dead-ends''.
 Recently, attempts to reconcile genetic and paleoanthropological data include proposals for a Pan-African origin of \emph{Homo sapiens} by which populations in many regions of the continent contributed to the formation of \emph{Homo sapiens} beginning at least 300ka \citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.

### 2022-07-18 supplement.tex
discussion of OOA parameter fix

@@
 
-\subsection{Estimating two-locus statistics with small sample sizes}
+\subsection{Estimating two-locus statistics with finite sample sizes}
 
@@
 with a family of models including single- and multiple-stem scenarios as
-described in the previous section. We identified parameters that reached
-consistent values across all models. These included the timings of recent or
-non-central divergences and admixture events:
+described in the previous section. We identified some parameters that reached
+consistent values across all models. :
 \begin{itemize}
     \item Neanderthal/human split: 550ka
-    \item East/West African population split: 60ka
-    \item East African/European split: 50ka
     \item Neanderthal introgression to Eurasian branch: 45ka
@@
 of the model.
+
+In our preliminary models, the unconstrained inferred split times between East African populations 
+and Europeans was inferred to have occurred around 40kya \sgcomment{check}, which is 
+more recent than we may have expected given 
+the well-documented presence of modern humans out of Africa at that time\sgcomment{cite}, as 
+well as previous genetics-based dating of the split \sgcomment{cite}. 
+The inferred split between East and West African populations occurred about $10ka$ earlier. 
+We see two non-exclusive explanations for this apparent discrepancy. 
+First, there is uncertainty in our model fit of this parameter.
+The $95\%$ confidence interval allows for a split as far back as \sgcomment{$XX$}. 
+Second, our confidence intervals do not capture all sources of uncertainty. 
+For example, model mis-specification can lead to a mismatch between an inferred split time and the first time populations actually split. 
+Our model supposes a single OOA event with constant symmetric migrations 
+between populations, which is certainly an oversimplification \cite{}. 
+Rather than adding the many parameters required to account for these effects, 
+we fixed the split time between Europeans and the East African ancestral population at $50ka,$ and
+the split between East/West African population at $60ka$. 
+We discuss the sensitivity of our model to these assumptions in Section \ref{sec:change_dates}.
 
@@
 \subsubsection{Varying fixed divergence times}
-
+\label{sec:change_dates}
 In order to reduce parameter space to avoid overfitting, we fixed parameters

### 2022-07-18 supplement.tex
supp results for nama map

--- 2022-07-15 supplement.tex
@@
 
-\begin{itemize}
-    \item Using YRI and CEU, inferred parameters for each of our highlighted models
-        did not change (all well within one SD of the inferred parameters of the
-        original fit models)
-    \item TODO: run fits with TSI
-    \item Fits using the Nama recombination map, even though there were slight
-        differences in some of the LD statistics (primarily a few noticeable
-        differences in the $\pi_2$ statistics), also resulted in model fits that
-        were very similar to the original fit models.
-    \item Fits with alternate thresholds of Nama-inferred ancestry: TODO:
-        fit continuous migration and the merger with stem migration model
-\end{itemize}
-
 \subsubsection{Replacing Thousand Genomes populations}
@@
 consistency of inferred parameters when replacing GBR with Toscani from Italy,
-a more distantly related European population set than CEU. \aprcomment{TODO}
+a more distantly related European population set than GBR. Most parameters in
+each of the tested models were consistent with parameters inferred using GBR.
+The only large difference was found in fitting the continuous migration model,
+Nama divergence was $\approx200$ka, rather than $135$ka. Migration rates during
+this early period of the model also shifted to larger values between the stem
+populations, demonstrating that divergence times and migration rates are
+confounded parameters in this model. Notably, the multiple merger model (which
+provides a better fit to the data) was inferred to have consistent divergence
+and merger times between the models fit using GBR and TSI, with ancestral
+populations ``reshuffling'' between 100 and 120ka.
 
 \subsubsection{Fit to LD statistics from Nama recombination map}
+
+The LD statistics computed using the OMNI YRI recombination map and the
+Nama-specific recombination map were very similar, with only a handful of
+statistics showing slight visible differences
+(Figure~\ref{fig:supp-omni-nama-comparison}). We therefore expected inferred
+demographic models to be highly consistent when fit to the two sets of
+statistics. Across each of our models, inferred parameters from fits using data
+from the Nama-specific map were unchanged, with best-fit values all within one
+estimated standard error from the inferred parameters from the OMNI YRI map.
+
+Among human populations, the Nama are relatively highly diverged from the
+sampled populations used in estimating the OMNI recombination maps (such as the
+YRI), and the two recombination maps were inferred using different
+computational methods from different types of sequencing data (arrays vs. whole
+genomes). Therefore, differences between these two maps are likely near the
+upper bound for differences in recombination rates between present-day human
+populations \citep{Van_Eeden2021-od}. Because our fits were highly consistent
+when using the two recombination maps, we conclude that inferences using the
+diversity and LD statistics here are robust to variation in recombination rates
+between present-day populations as well as to any differences in inferred
+recombination rates due to true recombination rate differences or typical
+statistical error in recombination rate estimation.
+
+\subsubsection{Varying Nama ancestry-proportion thresholds}
 
@@
         Nonetheless, LD statistics computed under recombination rates from
-        the two maps remain largely concordant, and inferences are not
-        strongly affected by recombination map choice (Section~\ref{sec:supp-results}).
+        the two maps remain largely concordant, and inferences are
+        unaffected by recombination map choice (Section~\ref{sec:supp-results}).
         Shading represents 95\% confidence intervals from the Nama-inferred
         map. Notations for each statistic are described in
-        section~\ref{sec:supp-ld-stats} and indexed by populations (0: Nama, 1:
-        Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
-    }
-    \label{fig:supp-omni-hapmap-comparison}
+        section~\ref{sec:supp-ld-stats}, and indexes represent populations
+        (0: Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
+    }
+    \label{fig:supp-omni-nama-comparison}
 \end{figure}

### 2022-07-18 paper.tex

@@
         $LL= -115,500$ & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
-        (Table S4) & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
+        (Table~\ref{tab:supp-continuous-migration}) & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
         & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
@@
         $LL= -102,600$ & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
-        (Table S6) & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
+        (Table~\ref{tab:supp-merger-with-stem-migration}) & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
         & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\

### 2022-07-18 supplement.tex
nama threshold section

@@
 merger without stem migration, and merger with stem migration) were converted
-to \demes format (Gower, Ragsdale, \emph{et al.}, in prep) and used as
+to \demes format \citep{Gower2022-yn} and used as
 the input demography for \msprime simulations. We sampled the same
@@
 \label{sec:supp-alt-data}
+
+All supplementary inference results are provided in Demes format
+\citep{Gower2022-yn} as a supplemental zipped archive.
 
@@
 \subsubsection{Varying Nama ancestry-proportion thresholds}
+
+In the demographic inference results presented in the main text, we did not
+impose a minimum threshold on inferred Nama ancestry (as estimated using
+AMDIXTURE). Because our proposed models allow for post-divergence gene flow as
+well as recent admixture events, we can jointly learn deeper history and recent
+admixture dynamics. This is in contrast to methods that do not account for
+ongoing migration or recent admixture events, for which un-admixed genomes (if
+they even exist) are needed for unbiased estimates of earlier history.
+Although the methods used in this paper (\texttt{moments-LD}) are able to infer
+early and recent history jointly, we tested the robustness of the inferred
+early history to varying ancestry proportions in the Nama population. To do
+this, we used ADMIXTURE to cluster inferred ancestry components, noting that
+ADMIXTURE-inferred ancestries are coarse and can be misleading about true
+underlying admixture proportions. Nonetheless, we included Nama individuals
+that exceeded either 90\% ro 99\% ancestry primarily shared among Nama
+individuals.
+
+For both thresholds, early history was robust in all tested models. Divergence
+times and migration rates did not vary significantly from the original fits
+that included all unrelated Nama individuals. The primary differences between
+these fits and the original fits that did not impose an ancestry threshold were
+the inferred admixture proportions of East African agriculturalists (2ka) and
+Europeans (10 generations ago) in the Nama population. The East African
+admixture proportion was reduced from $\approx 25\%$ to $20-22\%$, while the
+European admixture proportion was reduced from $\approx 15\%$ to either $5-6\%$
+(with a $90\%$ ancestry cutoff) or $3-5\%$ (with a $99\%$ ancestry cutoff). No
+other parameters were strongly affected by subsetting the Nama individuals by
+ADMIXTURE-inferred ancestry proportions. Because only the inferred proportions
+of recent admixture in our model fits changed, we conclude that our inferences
+of early history are robust to recent admixture, as long as that recent
+admixture is accounted for in the demographic models.
 

### 2022-07-18 paper.tex
minor wording change to intro para, formatting

@@
 
-Decades of study of human genome variation have suggested a predominantly tree-like model of recent population divergence from a single ancestral population in Africa.
-It has been difficult to reconcile this finding with the fossil and archaeological record of human occupation across the vast African continent.
-For example, fossils such as those from the sites of Jebel Irhoud, Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo sapiens} anatomical features were intermittantly present across the continent 300-100ka.
-Archaeological sites from the Middle Stone Age, which has been associated with \emph{Homo sapiens} at some sites, are also widely distributed across Africa.
-It is unclear whether these fossils and archaeological sites represent populations which contributed to contemporary \emph{Homo sapiens} as population precedents or were local ``dead-ends''.
-Recently, attempts to reconcile genetic and paleoanthropological data include proposals for a Pan-African origin of \emph{Homo sapiens} by which populations in many regions of the continent contributed to the formation of \emph{Homo sapiens} beginning at least 300ka \citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
+Decades of study of human genome variation have suggested a predominantly
+tree-like model of recent population divergence from a single ancestral
+population in Africa. It has been difficult to reconcile this finding with the
+fossil and archaeological record of human occupation across the vast African
+continent. For example, fossils such as those from the sites of Jebel Irhoud,
+Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies
+River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
+sapiens} anatomical features were intermittantly present across the continent
+300-100ka. Archaeological sites from the Middle Stone Age, of which some have
+been associated with \emph{Homo sapiens}, are also widely distributed across
+Africa. It is unclear whether these fossils and archaeological sites represent
+populations which contributed to contemporary \emph{Homo sapiens} as population
+precedents or were local ``dead-ends''. Recently, attempts to reconcile genetic
+and paleoanthropological data include proposals for a Pan-African origin of
+\emph{Homo sapiens} by which populations in many regions of the continent
+contributed to the formation of \emph{Homo sapiens} beginning at least 300ka
+\citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
 

### 2022-07-22 supplement.tex
more supp results

@@
 described in the previous section. We identified some parameters that reached
-consistent values across all models. :
+consistent values across all models:
 \begin{itemize}
     \item Neanderthal/human split: 550ka
-    \item Neanderthal introgression to Eurasian branch: 45ka
     \item Eurasian back-to-African migration: 12ka
@@
 \begin{itemize}
+    \item Neanderthal introgression to Eurasian branch: 45ka
     \item Mende population expansion and Gumuz population decline: 5ka \citep{Gopalan2022-pw}
@@
 \end{itemize}
-allowed to vary in the fits. The total number of parameters that were
+The total number of parameters that were
 ultimately inferred were 21-31, depending on the complexity
@@
 
-In our preliminary models, the unconstrained inferred split times between East African populations 
-and Europeans was inferred to have occurred around 40kya \sgcomment{check}, which is 
-more recent than we may have expected given 
-the well-documented presence of modern humans out of Africa at that time\sgcomment{cite}, as 
-well as previous genetics-based dating of the split \sgcomment{cite}. 
-The inferred split between East and West African populations occurred about $10ka$ earlier. 
-We see two non-exclusive explanations for this apparent discrepancy. 
-First, there is uncertainty in our model fit of this parameter.
-The $95\%$ confidence interval allows for a split as far back as \sgcomment{$XX$}. 
-Second, our confidence intervals do not capture all sources of uncertainty. 
-For example, model mis-specification can lead to a mismatch between an inferred split time and the first time populations actually split. 
-Our model supposes a single OOA event with constant symmetric migrations 
-between populations, which is certainly an oversimplification \cite{}. 
-Rather than adding the many parameters required to account for these effects, 
-we fixed the split time between Europeans and the East African ancestral population at $50ka,$ and
-the split between East/West African population at $60ka$. 
-We discuss the sensitivity of our model to these assumptions in Section \ref{sec:change_dates}.
+In our preliminary models, the unconstrained inferred split times between East
+African populations and Europeans was inferred to have occurred around $40$ka,
+which is more recent than we may have expected given the well-documented
+presence of modern humans outside of Africa at that time \citep{}, as well as
+previous genetics-based dating of splits between African and Eurasian
+populations \citep{}. The unconstrained inferred split between East and West
+African populations occurred about $10$ka earlier
+(Section~\ref{sec:supp-results}).
+%We see two non-exclusive explanations for this apparent discrepancy. First,
+%there is uncertainty in our model fit of this parameter. The $95\%$ confidence
+%interval allows for a split as far back as \sgcomment{$XX$}.
+Across all models, the $95\%$ confidence intervals for the OOA event was
+roughly $38 \pm 5$ka, and for the East/West African split was $45 \pm 10$ka.
+However, our confidence intervals do not capture all sources of uncertainty.
+For example, model mis-specification can lead to a mismatch between an inferred
+split time and the first time populations actually split. Our model supposes a
+single OOA event followed by constant symmetric migrations between populations,
+which is certainly an oversimplification \citep{}. Additionally, most previous
+estimates of the time of the OOA expansion compared Eurasian and West African
+populations. Our inferred time of the East/West African divergence aligns with
+previous estimates of the OOA event at $\approx 60$ka (themselves inferred from
+oversimplified demographic models). Rather than adding the many parameters
+required to account for these effects, we fixed the split time between
+Europeans and the East African ancestral population at $50$ka, and the split
+between East/West African population at $60$ka.
+%We discuss the sensitivity of our model to these assumptions in Section
+%\ref{sec:change-dates}.
 
@@
 
-\subsection{Relaxing fixed divergence times}
-
-\begin{itemize}
-    \item Fixed to older dates (100ka: E/W split, 60ka: OOA event). These fixed
-        values stretched the other inferred parameters to older times, and resulted
-        in a worse log-likelihood
-    \item Allowed to vary, the best-fit inferred E/W split time and OOA event
-        were more recent 35-50ka, instead of our fixed 50-60ka. \aprcomment{I
-        still don't find this entirely unreasonable, as E Africans and Eurasian
-        populations may be more closely related than the oldest evidence of H.
-        sapiens in Europe (up to 50ka), or those earliest H. sapiens are likely
-        not to be the direct ancestors of present-day Europeans. Still, it's
-        probably too recent of a split, and fixing to 50ka is defensible.}
-\end{itemize}
+\subsection{Relaxed inference of divergence times}
+\label{sec:change-dates}
+
+In the primary results presented in the main text
+(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
+we fixed a number of parameters that have been previously inferred or are well
+constrained by historical records (Section~\ref{sec:models}), including the
+Eurasian out-of-Africa expansion at 50ka and the divergence of Western and
+Eastern African groups at 60ka. To evaluate the effects of these choices on the
+inference of other parameters in the model, we first fixed those modeled
+divergence times to 60ka (for the out-of-Africa expansion) and 100ka (for the
+E/W African divergence). In the next section, we allow those parameters to be
+fit along with all other model parameters.
+
+By fixing those divergence times to older dates, the inferred models preferred
+older dates for the other free divergence time parameters. These dates were
+stretched by a similar proportion that our fixed parameters were shifted, so
+that the Nama divergence time was inferred to be $\approx150$ka, rather than
+$\approx130$ka. Notably, older divergence times resulted in models with worse
+likelihoods than our primary presented results. Overall, these inferred models
+have a similar interpretation to the primary inference results, and the deep
+population structure was unaffected by these choices.
+
+When we allowed the out-of-Africa expansion and Eastern/Western African
+divergence times to be jointly fit in model optimization, we found that the
+best-fit inferred E/W divergence time and the OOA expansion were more recent
+(35--40ka, 45--55ka, resp.). The remainder of the parameters were largely
+unaffected by these shifts to more recent divergence times, and our
+interpretation of the inferred early history remains unchanged. The $95\%$
+confidence intervals around these parameters were relatively narrow ($\approx
+5$ka for the OOA split, $\approx 10$ka for the East/West African split),
+although this uncertainty does not capture uncertainty due to model
+mis-specification (as discussed in Section~\ref{sec:models}).
 
@@
         a simpler isolation-with-migration (IM) model for pairs of populations
-        including the Nama to explore the effect of model misspecification on
+        including the Nama to explore the effect of model mis-specification on
         the earliest inferred divergence times between Khoe-San and other

### 2022-07-22 paper.tex
minor edits

@@
 
-Decades of study of human genome variation have suggested a predominantly
+Archaeological sites from the Middle Stone Age (approx. 300-40ka) are widely
+distributed across Africa, and are particularly well represented in the
+northern, eastern and southern parts of the continent. Fossils 
+such as those from the sites of Jebel Irhoud, Morocco \citep{Hublin2017-cq}, Herto, Ethiopia
+\citep{White2003-bk} and Klasies River, South Africa \citep{Deacon1995-rx} demonstrate that 
+derived \emph{Homo sapiens} anatomical features were also present across the
+continent during this period. It has been difficult to reconcile these lines of
+evidence with evidence from genomics, which have suggested a predominantly
 tree-like model of recent population divergence from a single ancestral
-population in Africa. It has been difficult to reconcile this finding with the
-fossil and archaeological record of human occupation across the vast African
-continent. For example, fossils such as those from the sites of Jebel Irhoud,
-Morocco \citep{Hublin2017-cq}, Herto, Ethiopia \citep{White2003-bk} and Klasies
-River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
-sapiens} anatomical features were intermittantly present across the continent
-300-100ka. Archaeological sites from the Middle Stone Age, of which some have
-been associated with \emph{Homo sapiens}, are also widely distributed across
-Africa. It is unclear whether these fossils and archaeological sites represent
-populations which contributed to contemporary \emph{Homo sapiens} as population
-precedents or were local ``dead-ends''. Recently, attempts to reconcile genetic
-and paleoanthropological data include proposals for a Pan-African origin of
+population. It is unclear whether fossil specimens and archaeological sites
+represent populations which contributed to our ancestors as population
+precedents, or were local ``dead-ends'' from which contemporary \emph{Homo
+sapiens} do not descend. 
+Recently,
+attempts to reconcile genetic and
+paleoanthropological data include proposals for a Pan-African origin of
 \emph{Homo sapiens} by which populations in many regions of the continent
@@
 Regardless of the model choice for early epochs, inference of human demographic
-history for the last 150ka was remarkably robust.
-In a reticulated model, we use ``divergence'' between populations to describe
-the time of their most recent shared ancestry.
+history for the last 150ka was remarkably robust. In a reticulated model, we use
+"divergence" between populations to describe the time of their most recent shared ancestry.  
 The earliest divergence among
@@
 `stem' groups as only a small proportion of drift between contemporary populations can
-be attributed to drift between stems (Figures~\ref{fig:predictions} and
+be attributed to drift between stems (Section \ref{sec:f4},  Figures~\ref{fig:predictions} and
 \ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
@@
 and archaeological research to infer robust migration rates.
-For example, all models infer relatively high gene flow between Ethiopia and western Africa
-($m\approx2\times10^{-4},$ the constant proportion of migrant lineages per generation over 60ka).
+For example, all models infer relatively high gene flow between eastern and western Africa
+($m\approx2\times10^{-4},$ the proportion of migrant lineages per generation).
 We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
@@
         $LL= -115,500$ & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
-        (Table~\ref{tab:supp-continuous-migration}) & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
+        & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
         & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
@@
         $LL= -102,600$ & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
-        (Table~\ref{tab:supp-merger-with-stem-migration}) & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
+        & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
         & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
@@
 479ka. The timing of this divergence was also poorly constrained ($\sigma_T=
-166$ka). These populations evolve independently until 119ka
-($\sigma_T=
-28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
+166$ka). These populations evolve independently until approximately 119ka
+when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
 proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
@@
 Out of Africa event). Finally, the Mende receive a large additional pulse of
-gene flow from Stem 2, replacing 18\% of their population 25ka ($\sigma_T=
-0.6$ka).
+gene flow from Stem 2, replacing 18\% of their population 25ka.
 The later Stem 2 contribution to the western African Mende resulted in
 better model fits ($\Delta LL \approx 60,000$).
-Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
+This may indicate that an ancestral Stem 2 population occupied western or
 Central Africa, broadly speaking. The differing proportions in the Nama and
@@
 single-origin model, recapitulate an inferred ancestral
-increase in $N_e$ between 100ka-1Ma extending deeper in the past than the relate estimate (Figure~\ref{fig:supp-iicr-sim} and Section \ref{inferred_relate}).
+increase in $N_e$ between 100ka-1Ma (Figure~\ref{fig:supp-iicr-sim}).
 Whereas the single-origin model achieves this by
@@
 this pattern without corresponding population size changes.      
+%\comment{These are pretty complicated ideas that are not explained in detail.
+%Might be worth unpacking a bit more to make sure the reader understands}
 
@@
 Other studies have fit tree-like demographic models to African populations using 
-distributions of allele frequencies or related statistics,
-finding inconsistent divergence times \cite{Henn2018, Bergstrom2021}.
-Section~\ref{sec:IM-reinfer} explains this behaviour:
-split times inferred using isolation with migration models and
-distributions of allele frequencies simulated under our reticulated
-models are highly variable and much deeper than the more recent
-estimated population divergence.
+distributions of allele frequencies or related statistics, finding inconsistent divergence times \cite{Henn2018, Bergstrom2021}.  
+Section \ref{sec:IM-reinfer} explains this behaviour: 
+split times inferred using isolation with migration models and distributions of allele frequencies simulated under our reticulated 
+models are highly variable and much deeper than the more recent recent population divergence.
 
@@
 We cannot rule out that more complex models involving
-additional stems, continuous structure, or hybrid models including both weak structure and archaic
+additional stems, or hybrid models including both weak structure and archaic
 hominin admixture may better explain the data. Because parameters related to the split
@@
 contemporary groups. We observed two merger events. The first, between Stem 1S and
-Stem 2, results in the formation of an ancestral Khoe-San population $\approx120$ka.
+Stem 2, results in the formation of an ancestral Khoe-San population ~120ka.
 The second $\approx100$ka between Stem 1E and Stem 2, results in the formation of the

### 2022-07-22 supplement.tex
fix OOA parameters

@@
 African populations occurred about $10$ka earlier
-(Section~\ref{sec:supp-results}).
-%We see two non-exclusive explanations for this apparent discrepancy. First,
-%there is uncertainty in our model fit of this parameter. The $95\%$ confidence
-%interval allows for a split as far back as \sgcomment{$XX$}.
-Across all models, the $95\%$ confidence intervals for the OOA event was
-roughly $38 \pm 5$ka, and for the East/West African split was $45 \pm 10$ka.
-However, our confidence intervals do not capture all sources of uncertainty.
-For example, model mis-specification can lead to a mismatch between an inferred
-split time and the first time populations actually split. Our model supposes a
-single OOA event followed by constant symmetric migrations between populations,
-which is certainly an oversimplification \citep{}. Additionally, most previous
+These recent split dates seems to be at odds with archaelogical evidence for earlier
+ Out-of-Africa (OOA) expansion, and with the estimated date for Neanderthal admixture
+ into OOA populations.  The parameter uncertainty due to finite genome
+ do not account for this difference:  in the high-likelihood models, the $95\%$ confidence 
+ intervals for the east African-European split was roughly $38 \pm 5$ka, 
+ and for the East/West African split was $45 \pm 10$ka, with the uncertainties 
+ reflecting 2 standard deviations according to bootstrap of genomic regions.  
+Thus we expect these discrepancies to result for model misspecification in the 
+details of the divergence between the ancestors of OOA and east African 
+populations. Our models supposes a
+single split prior to the OOA event followed by constant symmetric migrations 
+between eastern and OOA populations.
+which is certainly an oversimplification \citep{}.
+
+Since a detailed model for the Out of Africa and Back-to-Africa was not our main focus, 
+we chose to keep this simplified model, but fix the split between east African-European 
+split at $50ka.$ This allowed us to include Neanderthal admixture into the OOA 
+population at the previously-estimated $45ka$. Fixing this parameter pushed back the 
+estimated value for the split between the ancestors of east and West African populations at around 
+$60ka$, a value that we took as fixed in most subsequent analyses.
+
+We discuss the sensitivity of our model to these assumptions in Section
+\ref{sec:change-dates}.
+
+\sgcomment{I feel like we should improve the discussion of relationship with previous estimates below, or remove.}
+
+In addition to archaeological evidence, previous genetic dating suggested 
+earlier dates for a split between European and African populations. Most of 
+these results focused on populations of recent West African ancestry.
+
+Additionally, most previous
 estimates of the time of the OOA expansion compared Eurasian and West African
@@
 previous estimates of the OOA event at $\approx 60$ka (themselves inferred from
-oversimplified demographic models). Rather than adding the many parameters
-required to account for these effects, we fixed the split time between
-Europeans and the East African ancestral population at $50$ka, and the split
-between East/West African population at $60$ka.
+oversimplified demographic models).
+
 %We discuss the sensitivity of our model to these assumptions in Section

### 2022-07-22 paper.tex
Update paper.tex

@@
 population in Africa. It has been difficult to reconcile this finding with the
-fossil and archaeological record of human occupation across the vast African
+fossil and archaeological records of human occupation across the vast African
 continent. For example, fossils such as those from the sites of Jebel Irhoud,
@@
 substantially improves the description of genetic data relative to
-single-origin models
+single origin models
 \citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}.
@@
 finding ancient DNA evidence (e.g.\citep{Hsieh2016-gk}). However, these prior
-articles share two weaknesses. First, they only contrast a single-origin model
+articles share two weaknesses. First, they only contrast a single origin model
 with an archaic hominin admixture model, leaving out other plausible models
@@
 All models, including the
-single-origin model, recapitulate an inferred ancestral
+single origin model, recapitulate an inferred ancestral
 increase in $N_e$ between 100ka-1Ma extending deeper in the past than the relate estimate (Figure~\ref{fig:supp-iicr-sim} and Section \ref{inferred_relate}).
-Whereas the single-origin model achieves this by
+Whereas the single origin model achieves this by
 an increase in $N_e$ during that period, the best-fit model recapitulate

### 2022-07-22 paper.tex
Update paper.tex

--- 2022-07-22 paper.tex
@@
 misspecification. This is true of earlier studies, which often assumed that
-data inconsistent with a single origin model must be explained by archaic
+data inconsistent with a single-origin model must be explained by archaic
 hominin admixture. This is also true of this study. While it remains prohibitive to
@@
 work has focused on archaic hominin admixture to explain patterns of polymorphism
-inconsistent with a single origin model. Here, we have shown that
+inconsistent with a single-origin model. Here, we have shown that
 weakly-structured-stem models better capture these patterns.

### 2022-07-22 paper.tex
Revised "ecological riddle" paragraph

@@
 work has focused on archaic hominin admixture to explain patterns of polymorphism 
-inconsistent with a single-origin model. Here, we have shown that
-weakly-structured-stem models better capture these patterns.
-They also help explain an
-ecological riddle posed by the archaic hominin admixture model. Neanderthal populations
-were separated from early \emph{Homo sapiens} by thousands of kilometers and
-continental geographic barriers. By contrast, an archaic hominin population in
-Africa would need to have stayed in relative reproductive isolation from the
-ancestral human lineage over hundreds of thousands of years despite closer geographic
-proximity and reproductive compatibility. The weakly-structured-stem model
-resolves this ecological riddle by accommodating continuous or recurrent
-contact between two or more groups present in Africa.
+inconsistent with a single-origin model. Here, we have shown that weakly-structured-stem 
+models better capture these patterns. Preferring weakly-structured-stem models over 
+archaic-admixture models potentially has ecological implications. With a weakly 
+structured stem, there is no need to posit that an archaic hominin population stayed 
+reproductively isolated from the ancestral human lineage for thousands of years before 
+the initiation of gene flow. Instead, there would have been continuous or recurrent 
+contact between two or more groups present in Africa. It would be fruitful to 
+investigate the ecological reasons for continued contact between groups within Africa 
+but long periods of isolation between African and Eurasian groups.
 

### 2022-07-22 supplement.tex
Update supplement.tex

@@
  Out-of-Africa (OOA) expansion, and with the estimated date for Neanderthal admixture
- into OOA populations.  The parameter uncertainty due to finite genome
- do not account for this difference:  in the high-likelihood models, the $95\%$ confidence 
+ into OOA populations. The parameter uncertainty due to finite genomes
+ does not account for this difference:  in the high-likelihood models, the $95\%$ confidence 
  intervals for the east African-European split was roughly $38 \pm 5$ka, 
@@
         direct comparison to our inferred demographic models, we ran Relate on
-        the 290 genomes from the Nama, Gumuz, Mende, Amhara/Oromo, and British.
+        the 290 genomes from the Nama, Gumuz, Mende, Amhara/Oromo, and British individuals.
         While the reconstructed IICR curves are qualitatively similar to those

### 2022-07-22 paper.tex
Update paper.tex

@@
 multi-regional evolution, Figure~\ref{fig:proposed-models}), using 290
-genomes from southern, eastern, and western Africa as well as Eurasia. By
+genomes of individuals from southern, eastern, and western Africa as well as Eurasia. By
 including geographically and genetically diverse populations across Africa, we

### 2022-07-27 supplement.tex
clean up fixed parameters section

--- 2022-07-22 supplement.tex
@@
 In our preliminary models, the unconstrained inferred split times between East
-African populations and Europeans was inferred to have occurred around $40$ka,
-which is more recent than we may have expected given the well-documented
-presence of modern humans outside of Africa at that time \citep{}, as well as
-previous genetics-based dating of splits between African and Eurasian
-populations \citep{}. The unconstrained inferred split between East and West
-African populations occurred about $10$ka earlier
-(Section~\ref{sec:supp-results}).
-%We see two non-exclusive explanations for this apparent discrepancy. First,
-%there is uncertainty in our model fit of this parameter. The $95\%$ confidence
-%interval allows for a split as far back as \sgcomment{$XX$}.
-Across all models, the $95\%$ confidence intervals for the OOA event was
-roughly $38 \pm 5$ka, and for the East/West African split was $45 \pm 10$ka.
-However, our confidence intervals do not capture all sources of uncertainty.
-For example, model mis-specification can lead to a mismatch between an inferred
-split time and the first time populations actually split. Our model supposes a
-single OOA event followed by constant symmetric migrations between populations,
-which is certainly an oversimplification \citep{}. Additionally, most previous
-estimates of the time of the OOA expansion compared Eurasian and West African
-populations. Our inferred time of the East/West African divergence aligns with
-previous estimates of the OOA event at $\approx 60$ka (themselves inferred from
-oversimplified demographic models). Rather than adding the many parameters
-required to account for these effects, we fixed the split time between
-Europeans and the East African ancestral population at $50$ka, and the split
-between East/West African population at $60$ka.
-%We discuss the sensitivity of our model to these assumptions in Section
-%\ref{sec:change-dates}.
+African populations and Europeans was inferred to have occurred by around
+$40$ka, which is more recent than we may have expected given the
+well-documented presence of modern humans outside of Africa at that time
+\citep[e.g.]{Hublin2020-nk,Hajdinjak2021-jo}, as well as previous
+genetics-based dating of splits between African and Eurasian populations
+\citep[e.g.]{Jouganous2017-pq,Kamm2020-vn}. The unconstrained inferred split
+between East and West African populations occurred about $10$ka earlier. These
+recent split dates seems to be at odds with archaeological evidence for earlier
+Out-of-Africa (OOA) expansion, and with the estimated date for Neanderthal
+admixture into OOA populations. The parameter uncertainty due to finite genomes
+does not account for this difference: in the high-likelihood models, the $95\%$
+confidence intervals for the East African-European split was roughly $38 \pm
+5$ka, and for the East/West African split was $45 \pm 10$ka, with the
+uncertainties reflecting 2 standard deviations according to bootstrap of
+genomic regions.  Thus we expect these discrepancies to result from model
+misspecification in the details of the divergence between the ancestors of OOA
+and east African populations. Our models supposes a single split prior to the
+OOA event followed by constant symmetric migrations between eastern and OOA
+populations, which is certainly an oversimplification.
+
+Since a detailed model for the Out of Africa and Back-to-Africa was not our
+main focus, we chose to keep this simplified model, but fix the split between
+east African-European split at $50$ka. This allowed us to include Neanderthal
+admixture into the OOA population at the previously-estimated $45$ka. Fixing
+this parameter pushed back the estimated value for the split between the
+ancestors of East and West African populations at around $60$ka, a value that
+we took as fixed in most subsequent analyses. 
+We discuss the sensitivity of our model to these assumptions in Section
+\ref{sec:change-dates}.
+
+%\sgcomment{I feel like we should improve the discussion of relationship
+%with previous estimates below, or remove.}
+%
+%In addition to archaeological evidence, previous genetic dating suggested 
+%earlier dates for a split between European and African populations. Most of 
+%these results focused on populations of recent West African ancestry. 
+%
+%Additionally, most previous
+%estimates of the time of the OOA expansion compared Eurasian and West African
+%populations. Our inferred time of the East/West African divergence aligns with
+%previous estimates of the OOA event at $\approx 60$ka (themselves inferred from
+%oversimplified demographic models).
 
@@
 
-\subsubsection{\aprcomment{TODO:} High-coverage Nama samples}
-
 \subsection{Validations using simulations from inferred demographic models}
@@
 All supplementary inference results are provided in Demes format
-\citep{Gower2022-yn} as a supplemental zipped archive.
+\citep{Gower2022-yn} as a supplemental zipped archive. For each scenario,
+we refit the continuous migration and multiple merger models, with
+stem migration.
 

### 2022-07-27 supplement.tex
ADRP throughout

@@
 We also note that the coalescence rates inferred from data used a larger set of
-populations from the combined 1000 Genomes/AGRP dataset, while our model
+populations from the combined 1000 Genomes/ADRP dataset, while our model
 simulations only included a subset of 5 populations. This may lead to
@@
         \textbf{Inverse instantaneous coalescence rates inferred from the data.}
-        Using combined data from the Thousand Genomes Project and the AGRP,
+        Using combined data from the Thousand Genomes Project and the ADRP,
         we reconstructed gene genealogies using Relate and calculated the
@@
         (ACB: Afro-Caribbean, ASW: African American, ESN: Esan, GWD: Gambian,
-        LWK: Luhya, MSL: Mende, YRI: Yoruba), (C) AGRP populations.
+        LWK: Luhya, MSL: Mende, YRI: Yoruba), (C) ADRP populations.
     }

### 2022-07-27 paper.tex
fix up a few references

@@
 We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
-coalescence rates over time in real data and data simulated from our inferred models. Many 
-previous studies have found a reduction of coalescence rates 
-between $1000$ka and $100$ka in humans, and thus inferred an increase 
-in $N_e$ during the same period \citep{Li2011-le}.
-This increase in inferred $N_e$ could be due to either an increase in population size or to 
-ancestral population structure during the Middle Stone Age \citep{Mazet2015-wn}.
-All models, including the
-single-origin model, recapitulate an inferred ancestral
-increase in $N_e$ between 100ka-1Ma extending deeper in the past than the relate estimate (Figure~\ref{fig:supp-iicr-sim} and Section \ref{inferred_relate}).
-Whereas the single-origin model achieves this by
-an increase in $N_e$ during that period, the best-fit model recapitulate
-this pattern without corresponding population size changes.
+coalescence rates over time in real data and data simulated from our inferred
+models. Many previous studies have found a reduction of coalescence rates
+between $1000$ka and $100$ka in humans, and thus inferred an increase in $N_e$
+during the same period \citep{Li2011-le}.  This increase in inferred $N_e$
+could be due to either an increase in population size or to ancestral
+population structure during the Middle Stone Age \citep{Mazet2016-wn}.  All
+models, including the single-origin model, recapitulate an inferred ancestral
+increase in $N_e$ between 100ka-1Ma extending deeper in the past than the
+relate estimate (Figure~\ref{fig:supp-iicr-sim} and Section
+\ref{inferred_relate}). Whereas the single-origin model achieves this by an
+increase in $N_e$ during that period, the best-fit model recapitulate this
+pattern without corresponding population size changes.
 
@@
 $\sim$15ka compared to a true divergence of 60ka), and recent migration can
-lead to the misordering of divergence events (Figure~\ref{fig:validation}E). We suggest
-that rCCR analyses which do not fit multiple parameters including gene flow
-should be interpreted with caution.
+lead to the misordering of divergence events (Figure~\ref{fig:validation}E). We
+suggest that rCCR analyses which do not fit multiple parameters including gene
+flow should be interpreted with caution.
 
-Other studies have fit tree-like demographic models to African populations using 
-distributions of allele frequencies or related statistics,
-finding inconsistent divergence times \cite{Henn2018, Bergstrom2021}.
-Section~\ref{sec:IM-reinfer} explains this behaviour:
-split times inferred using isolation with migration models and
-distributions of allele frequencies simulated under our reticulated
-models are highly variable and much deeper than the more recent
-estimated population divergence.
+Other studies have fit tree-like demographic models to African populations
+using distributions of allele frequencies or related statistics, finding
+inconsistent divergence times \cite{Henn2018-rf,Bergstrom2021-iw}.
+Section~\ref{sec:IM-reinfer} explains this behaviour: split times inferred
+using isolation with migration models and distributions of allele frequencies
+simulated under our reticulated models are highly variable and much deeper than
+the more recent estimated population divergence.
 

### 2022-07-29 supplement.tex
expand details of populations used in supp

@@
 \normalsize
-\newpage
+%\newpage
 
@@
 
-\aprcomment{TODO: expand this section to include further discussion about why
-we chose each population, and why we have omitted some (such as those suggested
-in the reviews).}
+The number of publicly available whole-genome sequenced individuals from a
+diverse set of African groups has grown in recent years
+\citep[e.g.,][]{Mallick2016-lx,Bergstrom2020-cy,Schlebusch2020-zo}. Our choice
+of populations was motivated by both technical and study design considerations.
+First, while the diversity and linkage disequilibrium statistics used in our
+inferences are robust to low coverage sequencing data \citep{Ragsdale2019-nt},
+we wanted to have at least 20 individuals per population to increase accuracy
+in statistical measurements by reducing the variance due to small sample sizes
+\citep{Ragsdale2020-nz}. Second, merging of datasets post-variant calling can
+introduce biases in demographic inference due to batch effects. We therefore
+wanted to be able to perform joint calling in all of our samples. Together,
+these considerations make the jointly called Thousand Genomes and ADRP dataset
+best-suited for our demographic inference.
+
+From a study design perspective, our choice of populations aimed to improve our
+understanding of early human demography in Africa. The Nama derive ancestry
+from a relatively deeply diverged population branch, and thus their demographic
+history is of unique interest. Given previous archaeological and genetic
+evidence, we expected that a minimal model for the Nama must include
+pastoralist East African groups that contributed to the Nama
+\citep[e.g.,][]{Uren2016-nn}, as well as recent European admixture, each of
+which are represented in our jointly called dataset. Finally, because previous
+studies have suggested a role for ``ghost'' archaic admixture or deep structure
+in West African populations \citep[e.g.,][]{Speidel2019-nj,Durvasula2020-td},
+and because these same sample sets are included in our dataset, we chose to
+also include representatives of West African ancestry.
 

### 2022-07-29 paper.tex

@@
 archaic-admixture models potentially has ecological implications. With a weakly 
-structured stem, there is no need to posit that an archaic hominin population stayed 
-reproductively isolated from the ancestral human lineage for thousands of years before 
-the initiation of gene flow. Instead, there would have been continuous or recurrent 
-contact between two or more groups present in Africa. It would be fruitful to 
-investigate the ecological reasons for continued contact between groups within Africa 
-but long periods of isolation between African and Eurasian groups.
+structured stem, there is no need to posit that an archaic hominin population in Africa stayed 
+reproductively isolated from the ancestral human lineage for tens thousands of years before 
+the initiation of gene flow. Instead, there would simply have been continuous or recurrent 
+contact between two or more groups present in Africa.
 

### 2022-07-29 paper.tex
tried to clarify the distinction between conceptual and fitted models

@@
 for contribution from an African archaic hominin lineage or population structure prior
-to the expansion (Figure~\ref{fig:proposed-models}A). As expected
-\citep{Ragsdale2019-nt}, this first model was a poor fit to the data
+to the expansion (Figure~\ref{fig:proposed-models}A or Figure~\ref{fig:proposed-models}D). As expected
+\citep{Ragsdale2019-nt}, this first model preferred a recent expansion model but was also a poor fit to the data
 qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
 (log-likelihood ($LL$) $\approx -189,400$, Table~\ref{tab:supp-single-origin}).
-We next explored a suite of models in which population structure predates the
-differentiation of contemporary groups, including models allowing for ancestral
+We next explored a suite of parameterized models in which population structure predates the
+differentiation of contemporary groups (Supp. Information section~\ref{modelspec}). 
+Depending on the parameters, these encompassed models allowing for ancestral
 reticulation (such as fragmentation-and-coalescence or meta-population models,
@@
 
-Regardless of the model choice for early epochs, inference of human demographic
+Regardless of the model choice for early epochs, maximum-likelihood inference of human demographic
 history for the last 150ka was remarkably robust.

### 2022-07-29 supplement.tex

@@
 \citep{Ragsdale2020-nz}. Second, merging of datasets post-variant calling can
-introduce biases in demographic inference due to batch effects. We therefore
-wanted to be able to perform joint calling in all of our samples. Together,
-these considerations make the jointly called Thousand Genomes and ADRP dataset
-best-suited for our demographic inference.
+introduce biases in demographic inference due to batch effects (as demonstrated
+in Section~\ref{sec:merged-bias}). We therefore wanted to be able to perform
+joint calling in all of our samples. Together, these considerations make the
+jointly called Thousand Genomes and ADRP dataset best-suited for our
+demographic inference.
 
@@
 pastoralist East African groups that contributed to the Nama
-\citep[e.g.,][]{Uren2016-nn}, as well as recent European admixture, each of
-which are represented in our jointly called dataset. Finally, because previous
-studies have suggested a role for ``ghost'' archaic admixture or deep structure
-in West African populations \citep[e.g.,][]{Speidel2019-nj,Durvasula2020-td},
-and because these same sample sets are included in our dataset, we chose to
-also include representatives of West African ancestry.
+\citep[e.g.,][]{Uren2016-nn}, as well as European colonial admixture.  The
+inclusion of three Ethiopian populations here represent some uncertainty in the
+possible source population for the pastoralist contribution to the Nama, and
+the Amhara and Oromo also have substantive ``back-to-Africa'' ancestry. The
+Gumuz serve as a representative of southwest Ethipian ancestry that pre-dates
+the Holocene back-to-Africa migration. Finally, because previous studies have
+suggested a role for ``ghost'' archaic admixture or deep structure in West
+African populations \citep[e.g.,][]{Speidel2019-nj,Durvasula2020-td}, we chose
+to also include representatives of West African ancestry.
 
@@
 \label{modelspec}
+
 For the early history, we tested model parameterizations that cover many of the
@@
 hominin admixture. The simplest model, in terms of number of parameters, was a
-single-origin expansion of humans, with no structure in the stem and no
-archaic hominin admixture aside from the Neanderthal admixture in Eurasian populations
+single-origin expansion of humans, with no structure in the stem and no archaic
+hominin admixture aside from the Neanderthal admixture in Eurasian populations
 following the out-of-Africa migration. This model allowed for a population size
@@
 human-Neanderthal lineages and the more recent split of branches leading to
-southern and western/eastern African populations.
+southern and western/eastern African populations. The recent expansion and the
+African multi-regional models (Figure~\ref{fig:proposed-models}A and D) have
+the same topology, so interpretation of the model depends on the specified or
+inferred divergence times. Thus, because model optimization explores a wide
+range of parameter values, we account for both scenarios with this
+parameterization.
 
@@
 \subsection{Optimization using \moments}
+\label{sec:optimization}
 
@@
 populations $\sim$120ka, ancient structure and size changes in the simulations
-biases re-inferences and gives inflated estimates of $T$.
-Such an effect has been shown to occur when the ancestral
-population size increases prior to the population split
-\citep{Momigliano2021-th}, which is the case in our inferred single-origin model.
-Re-inferred $T$ between the Nama and other populations were $>$200ka under the IM model,
-and $\sim$50--130ka under the clean split model.
+biases re-inferences and gives inflated estimates of $T$. Such an effect has
+been shown to occur when the ancestral population size increases prior to the
+population split \citep{Momigliano2021-th}, which is the case in our inferred
+single-origin model. Re-inferred $T$ between the Nama and other populations
+were $>$200ka under the IM model, and $\sim$50--130ka under the clean split
+model.
+
+\subsection{Batch effects from merged data biases inference}\label{sec:merged-bias}
+
+Merging samples from multiple independently called datasets can cause
+systematic biases in computed diversity statistics (i.e., batch effects). In
+order to explore the severity of batch effects on the LD statistics used in our
+inferences, we took advantage of multiple independently sequenced and called
+datasets using the same individuals from the Thousand Genomes project. We used
+the phase 3 low coverage \citet{1000_Genomes_Project_Consortium2015-zq} data
+and the high coverage resequenced data of the same individuals
+\citep{Byrska-Bishop2021-jl} to 1) verify that computed diversity statistics
+and inferences are robust to low coverage data, and 2) demonstrate that merging
+data without joint variant calling can lead to substantial biases in inferred
+demographic parameters. We analyzed two populations, the Mende (MSL) and
+British (GBR), and considered a two-population isolation-with-migration model.
+
+We computed four sets of diversity and LD statistics, one set of statistics
+from the jointly called high coverage data, one from the jointly called low
+coverage data, and one each from merged high and low coverage (MSL high
+coverage and GBR low coverage; and GBR high coverage and MSL low coverage).
+Using the measurement variances from each two-locus statistic in the high
+coverage data, we estimated measurement error of the other three sets of
+statistics (low/low, low/high, and high/low) against the high coverage data.
+The jointly called low coverage data provided statistics that closely matched
+the high coverage statistics, with each statistic within 1--2 standard errors.
+On the other hand, both the merged datasets provided statistics that deviated
+from the jointly called data, particularly for the cross-population statistics
+(Figure~\ref{fig:supp-merged-data-SEs}.
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=5in]{figures/supp-SE-merged-data}
+    \caption{
+        \textbf{Average error of statistics from low coverage and merged datasets.}
+        Low coverage data provides consistent allelic diversity and linkage
+        disequilibrium measures to high coverage data. On the other hand,
+        merging separately called datasets can lead to large discrepancies
+        in computed LD statistics.
+    }
+    \label{fig:supp-merged-data-SEs}
+\end{figure}
+
+With the observation that the low and high coverage data provided broadly
+consistent statistics while the merged data provided inconsistent statistics,
+we next wanted to assess whether their differences affect model inference. To
+do this, we fit a two-population demographic model to the diversity and LD
+statistics as described in Section~\ref{sec:optimization}. The modest
+differences between the low and high coverage did not result in noticeable
+differences in the optimized model (with an MSL-GBR inferred divergence time of
+60ka, consistent with previous genomics studies). On the other hand, both
+merged datasets provided substantially inconsistent fits, both to the high and
+low coverage data and to each other. Notably, the inferred divergence time is
+significantly older, due to the overestimation of mutational differences
+between populations due to batch effects.
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics{figures/supp-merged-data-model-fits}
+    \caption{
+        \textbf{Merged datasets bias demographic inferences.}
+        While statistics from low and high coverage datasets provide consistent
+        best-fit demographic models, merged datasets result in biased LD
+        statistics (Figure~\ref{fig:supp-merged-data-SEs}). These biased LD
+        statistics lead to large downstream biases in inferred demographic
+        models.
+    }
+    \label{fig:supp-merged-data-fits}
+\end{figure}
+
+More careful post-merging quality control may alleviate some of these
+discrepancies. However, low coverage data provides a consistent model fit to
+the high coverage data. Therefore, it is preferable to favor jointly called low
+coverage data over a mix of high coverage but separately called data, at least
+when using the diversity and LD statistics that we have in our analyses.
 

### 2022-08-04 supplement.tex
typo fix

@@
 between East and West African populations occurred about $10$ka earlier. These
-recent split dates seems to be at odds with archaeological evidence for earlier
+recent split dates seem to be at odds with archaeological evidence for earlier
 Out-of-Africa (OOA) expansion, and with the estimated date for Neanderthal

### 2022-08-04 paper.tex
Update paper.tex

@@
 history for the last 150ka was remarkably robust.
-In a reticulated model, we use ``divergence'' between populations to describe
+In a reticulated model, we use ``divergence'' between populations to mean
 the time of their most recent shared ancestry.
@@
 We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
-date back to the late Middle Stone Age in Africa, likely arising during MIS 5.
+likely arise during MIS 5.
 Although we find evidence for earlier population structure in Africa (see
@@
 of optimizing a large number of parameters, we fixed several parameters related
-to recent population history so as to focus on more ancient events. 
-Fixed parameters included
-the time of divergence between western and eastern African populations,
+to recent population history so as to focus on more ancient events. These parameters were ones supported by multiple genetic and archaeological studies.
+Fixed parameters included the time of divergence between western and eastern African populations,
 set to 60ka, just prior to the split of Eurasians and East Africans set to 50ka. 
@@
 following the out-of-Africa migration which was set to 1.5\% at 45ka (Supp. Information).
-These constraints allowed us to integrate information from previous genetic
-and archaeological research to infer robust migration rates.
-For example, all models infer relatively high gene flow between Ethiopia and western Africa
+
+We quantify several migration rates among populations after their divergence ~120ka. 
+All models infer relatively high gene flow between Ethiopia and western Africa
 ($m\approx2\times10^{-4},$ the constant proportion of migrant lineages per generation over 60ka).
@@
 affected the ancestors of the Ethiopian agricultural populations, comprising
-over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
-years also saw major demographic changes, including strong population growth
-for western Africans as they specialized in yam and oil palm agriculture
-(estimated 3-fold growth). We observe significant gene flow from the Amhara and
-Oromo into the Nama, a signal which is likely a proxy for the movement of
-eastern African caprid and cattle pastoralists
-\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
-contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
+over half of their genetic ancestry, estimated to be 64--65\%. 
+We observe significant gene flow from the Amhara and Oromo into the Nama, a signal which is likely a proxy for the movement of
+eastern African caprid and cattle pastoralists \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
+contribution 2,000 ya.
+While this gene flow is not apparent from the ADMIXTURE plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestry Eastern African source.
+Colonial period admixture from Europeans into the Nama
 was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}

### 2022-08-04 paper.tex
Update paper.tex

@@
 
-We quantify several migration rates among populations after their divergence ~120ka. 
+We quantify migration rates among populations after their divergence ~120ka.
+Migration, prior to the Bantu expansion, between the ancestors of the Nama and other groups  
+is an order of magnitude weaker than that observed between between western and eastern Africans (Table \ref{tab:migration-rates}).
 All models infer relatively high gene flow between Ethiopia and western Africa

### 2022-08-05 paper.tex
fixed references to supplementary sections
clarified discrepancy in main text

@@
         We used these conceptual models as starting points to build detailed
-        parameterized demographic models (Supp. Information section~\ref{modelspec})
+        parameterized demographic models (Supp. Information Section~\ref{modelspec})
         that were then fit to genetic data. 
@@
 We also fixed the amount of admixture from Neanderthals to Europeans directly
-following the out-of-Africa migration which was set to 1.5\% at 45ka (Supp. Information).
+following the out-of-Africa migration which was set to 1.5\% at 45ka (Supp. Information \ref{sec:models}).
 
@@
 among contemporary populations can be traced back
-to this early population structure (Supp. Information)
+to this early population structure (Supp. Information \ref{sec:f4})
 
@@
         misidentification was consistently inferred to be 1.5--1.7\% for
-        intergenic loci (Supp. Information). (D, E) We used Relate
+        intergenic loci (Supp. Information \ref{sec:misspecification}). (D, E) We used Relate
         \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
@@
 using distributions of allele frequencies or related statistics, finding
-inconsistent divergence times \cite{Henn2018-rf,Bergstrom2021-iw}.
-Section~\ref{sec:IM-reinfer} explains this behaviour: split times inferred
-using isolation with migration models and distributions of allele frequencies
-simulated under our reticulated models are highly variable and much deeper than
-the more recent estimated population divergence.
+inconsistent divergence times, many of which older than the ones we find here
+\cite{Henn2018-rf,Bergstrom2021-iw}. In section~\ref{sec:IM-reinfer} , we show that this discrepancy
+can be explained by model misspecification: if divergence is estimated using an isolation with migration (IM) model
+with constant population sizes, but the correct model has ancient population growth or population structure,
+the inferred IM models have split times that are much deeper than in the correct model.
 
@@
 the recombination clock accurately relate to geological chronologies
-(Supp. Information).
+(Supp. Information \sgcomment{section}).
 During the Middle Stone Age, the multiple merger model indicates three

### 2022-08-05 paper.tex
minor word change

@@
 using distributions of allele frequencies or related statistics, finding
-inconsistent divergence times, many of which older than the ones we find here
-\cite{Henn2018-rf,Bergstrom2021-iw}. In section~\ref{sec:IM-reinfer} , we show that this discrepancy
+inconsistent divergence times, some of which older than the ones we find here
+\cite{Henn2018-rf,Bergstrom2021-iw}. In Section~\ref{sec:IM-reinfer} , we show that this discrepancy
 can be explained by model misspecification: if divergence is estimated using an isolation with migration (IM) model

### 2022-08-05 paper.tex
Typo fix and add cites

@@
 inconsistent divergence times, some of which older than the ones we find here
-\cite{Henn2018-rf,Bergstrom2021-iw}. In Section~\ref{sec:IM-reinfer} , we show that this discrepancy
+\cite{Henn2018-rf,Bergstrom2021-iw}. In Section~\ref{sec:IM-reinfer}, we show that this discrepancy
 can be explained by model misspecification: if divergence is estimated using an isolation with migration (IM) model
 with constant population sizes, but the correct model has ancient population growth or population structure,
-the inferred IM models have split times that are much deeper than in the correct model.
+the inferred IM models have split times that are much deeper than in the correct model \citep{Momigliano2021-th,Shchur2022-wa}.
 

### 2022-08-08 paper.tex
IM and coalescence times

--- 2022-08-05 paper.tex
@@
 \title{A weakly structured stem for human origins in Africa}
-\author[1]{Aaron P. Ragsdale}
-\author[2]{Timothy D. Weaver}
-\author[3]{Elizabeth G. Atkinson}
-\author[4]{Eileen Hoal}
-\author[4]{Marlo M\"{o}ller}
-\author[2,5,$\dag$,*]{Brenna M. Henn}
-\author[6,$\dag$,**]{Simon Gravel}
+\author[1]{Aaron~P.~Ragsdale}
+\author[2]{Timothy~D.~Weaver}
+\author[3]{Elizabeth~G.~Atkinson}
+\author[4]{Eileen~Hoal}
+\author[4]{Marlo~M\"{o}ller}
+\author[2,5,$\dag$,*]{Brenna~M.~Henn}
+\author[6,$\dag$,**]{Simon~Gravel}
 \affil[1]{Department of Integrative Biology, University of Wisconsin--Madison, WI, USA}
@@
 using distributions of allele frequencies or related statistics, finding
-inconsistent divergence times, some of which older than the ones we find here
-\cite{Henn2018-rf,Bergstrom2021-iw}. In Section~\ref{sec:IM-reinfer} , we show that this discrepancy
-can be explained by model misspecification: if divergence is estimated using an isolation with migration (IM) model
-with constant population sizes, but the correct model has ancient population growth or population structure,
-the inferred IM models have split times that are much deeper than in the correct model.
+inconsistent divergence times, some of which are older than those we find here
+\cite{Henn2018-rf,Bergstrom2021-iw}. In the Supp. Information
+(Section~\ref{sec:IM-reinfer}), we show that this discrepancy can be explained
+by model misspecification: if divergence is estimated using an isolation with
+migration (IM) model with constant population sizes, but the correct model has
+ancient population growth or population structure, the split time in the
+inferred IM model is much deeper than in the correct model. Intuitively, growth
+or structure in the ancestral population will each increase coalescence times
+relative to a randomly mating population of contant size, so a model that
+assumes constant population sizes would require an older divergence time to fit
+the observed distribution of coalescence times and related statistics
+\citep{Momigliano2021-th,Shchur2022-wa}.
 
@@
 data inconsistent with a single-origin model must be explained by archaic
-hominin admixture. This is also true of this study. While it remains prohibitive to
-fully explore the space of plausible models of early human population structure,
-we sought to capture model uncertainty by exploring multiple parameterizations of
-early history.
-The best-fit models presented here include reticulation and migration between
-early human populations rather than archaic hominin admixture from long-isolated branches.
-We cannot rule out that more complex models involving
-additional stems, continuous structure, or hybrid models including both weak structure and archaic
-hominin admixture may better explain the data. Because parameters related to the split
-time, migration rates, and relative sizes of the early stems were variable
-across models, reflecting a degree of confounding among these parameters, we
-refrained from introducing additional branches associated with more
-parameters during that period.
-Rather than interpreting the two stems as representing well-defined
-and stable populations over hundreds of thousands of years, we interpret the
-weakly structured stem as consistent with a population coalescence and
-fragmentation model \citep{Scerri2019-xg}.
-Models including additional diversity within Africa,
-and early ancient DNA samples from Africa, could further distinguish
-the archaic hominin admixture model from the weakly-structured-stem model.
+hominin admixture. This is also true of this study. While it remains
+prohibitive to fully explore the space of plausible models of early human
+population structure, we sought to capture model uncertainty by exploring
+multiple parameterizations of early history. The best-fit models presented
+here include reticulation and migration between early human populations rather
+than archaic hominin admixture from long-isolated branches. We cannot rule out
+that more complex models involving additional stems, continuous structure, or
+hybrid models including both weak structure and archaic hominin admixture may
+better explain the data. Because parameters related to the split time,
+migration rates, and relative sizes of the early stems were variable across
+models, reflecting a degree of confounding among these parameters, we refrained
+from introducing additional branches associated with more parameters during
+that period.  Rather than interpreting the two stems as representing
+well-defined and stable populations over hundreds of thousands of years, we
+interpret the weakly structured stem as consistent with a population
+coalescence and fragmentation model \citep{Scerri2019-xg}. Models including
+additional diversity within Africa, and early ancient DNA samples from Africa,
+could further distinguish the archaic hominin admixture model from the
+weakly-structured-stem model.
 
@@
 the recombination clock accurately relate to geological chronologies
-(Supp. Information \sgcomment{section}).
+(Supp. Information, Section~\ref{sec:supp-mutation}).
 During the Middle Stone Age, the multiple merger model indicates three
@@
 $0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
-(Stem 1E) 150ka versus $0.03$ probability of being inherited from the ‘southern’
+(Stem 1E) $150$ka versus $0.03$ probability of being inherited from the ‘southern’
 subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2

### 2022-08-11 supplement.tex
fixed references to supplementary sections
typo in supp

@@
 \subsubsection{cSFS prediction under inferred models}
-
+\label{sec:misspecification}
 We compared the observed cSFS to model predictions for the Nama, Mende, and
@@
 joint SFS, $SFS_{misid}(i, j)=(1-p_{misid})*SFS(i, j) + p_{misid}*SFS(n-i,
-m-j)$), where $n$ and $m$ are the sample sizes from the two populations. The
+m-j)$, where $n$ and $m$ are the sample sizes from the two populations). The
 cSFS for the first population (conditioning on the second) is found by

### 2022-08-12 paper.tex
change sigma to SE

@@
 optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
-bottleneck is poorly constrained ($\sigma_{N_e}=838$). After a long period of
+bottleneck is poorly constrained (standard error (SE) $=838$). After a long period of
 exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
-479ka. The timing of this divergence was also poorly constrained ($\sigma_T=
-166$ka). These populations evolve independently until 119ka
-($\sigma_T=
-28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
+479ka. The timing of this divergence was also poorly constrained (SE $=166$ka).
+These populations evolve independently until 119ka
+(SE $=28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
 proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
@@
 Out of Africa event). Finally, the Mende receive a large additional pulse of
-gene flow from Stem 2, replacing 18\% of their population 25ka ($\sigma_T=
-0.6$ka).
+gene flow from Stem 2, replacing 18\% of their population 25ka (SE $=0.6$ka).
 The later Stem 2 contribution to the western African Mende resulted in
@@
 Stem 2, results in the formation of an ancestral Khoe-San population $\approx120$ka.
-The second $\approx100$ka between Stem 1E and Stem 2, results in the formation of the
+The second event $\approx100$ka between Stem 1E and Stem 2,
+results in the formation of the
 ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
@@
 Following these merger events, the stems subsequently fracture into
-subpopulations which then appear to persist over the past $\sim120$ka. These
+subpopulations which then appear to persist over the past $\sim$120ka. These
 subpopulations can be linked to contemporary groups despite subsequent gene flow

### 2022-08-12 paper.tex
point to validations in supplement

@@
 high-likelihood models which we explored did the divergence between Nama and
-other populations exceed $\sim$140ka. 
-We conclude that geographic patterns of contemporary \emph{Homo sapiens} population structure
-likely arise during MIS 5.
-Although we find evidence for earlier population structure in Africa (see
-below), contemporary populations cannot be easily mapped onto the more ancient
+other populations exceed $\sim$140ka
+(Supp. Information~\ref{sec:validation-data-models} describes validation using
+alternative datasets and model specifications). 
+We conclude that geographic patterns of contemporary \emph{Homo sapiens}
+population structure likely arise during MIS 5.
+Although we find evidence for earlier population structure in Africa (next section),
+contemporary populations cannot be easily mapped onto the more ancient
 `stem' groups as only a small proportion of drift between contemporary populations can

### 2022-08-12 supplement.tex
point to validations in supplement
intuition

@@
 model.
+
+To understand the relationship between model specification and demographic inference, 
+consider the five toy models shown on Figure \ref{fig:toymodels}: A) clean split 100ka, 
+B) clean split 200ka, C) population growth at 200ka followed by split 100ka, 
+D) split 200ka followed by migration until 100ka, and E) split 200ka followed by 
+merger 105ka and a new split 100ka.
+
+Models C), D), and E) show qualitatively similar patterns of coalescence rate. 
+Given the same data, an IM model (such as D) will infer a potentially much 
+deeper population divergence than growth model C and reticulation model E.
 

### 2022-08-12 supplement.tex
paragraph on TSI

@@
 \subsection{Validations using alternative data sets and model specification}
+\label{sec:validation-data-models}
 
@@
 populations ``reshuffling'' between 100 and 120ka.
+
+Across all validations to alternative datasets described in this section, model
+fits were largely robust and matched those shown in the main text
+(Figure~\ref{fig:best-fit-models})---the single outlier in these validations
+was the continuous migration model fit to the dataset that included TSI instead
+of GBR. The continuous migration model did not provide the best log-likelihood
+among tested models, and our original model using the dataset with GBR
+underestimated both $\sigma_{Dz}$ and $\sigma_{\pi_2}$ compared to data
+(Figure~\ref{fig:supp-continuous-migration-fits}). In fitting this same model
+to the TSI, the model provided a better fit to the $\sigma_{Dz}$ statistics but
+a worse fit to the $\sigma_{\pi_2}$ statistics (and still provided a
+significantly worse fit to the data compared to the multiple merger model).
+These differences are likely driven by a combination of factors: the continuous
+migration model mis-specifies historical dynamics in the range of 100-200ka;
+the TSI are an inappropriate proxy for recent European admixture in East and
+South Africa; and the TSI may also carry recent North African ancestry, which
+is not captured in our model. Together, this highlights the sensitivity of
+complex demographic inference to the choices of populations to include, as well
+as to choices of model specification.
+\aprcomment{not sure about these last two sentences}
 

### 2022-08-12 supplement.tex
figure caption

--- 2022-08-12 supplement.tex
@@
 populations ``reshuffling'' between 100 and 120ka.
+
+\aprcomment{Short paragraph here explaining the discrepancy in the TSI fit
+continuous migration model}
 
@@
 merger 105ka and a new split 100ka.
+
+\begin{figure}[ht!]
+    \centering
+    \includegraphics[width=5in]{figures/supp-toy-models}
+    \caption{The impact of model specification on coalescence rate and inference.
+    We show five simplified demographic models that differ in demographic events, 
+    occurring between 200ka and 100ka, that mark the transition between a single 
+    ancestral population and two daughter populations. A) A clean split 100ka, 
+B) A clean split 200ka, C) population growth at 200ka followed by split 100ka, 
+D) split 200ka followed by migration until 100ka, and E) split 200ka followed by 
+merger 105ka and a new split 100ka. The left column shows model representation, 
+the middle column shows instantaneous coalescence rates for pairs of lineages from the
+same daughter population, the right column shows instantaneous coalescence rates for
+pairs of lineages coming from distinct populations.  
+        \label{fig:toymodels}
+    }
+    \end{figure}
 

### 2022-08-12 supplement.tex
added figures

--- 2022-08-12 supplement.tex
@@
 \usepackage{xspace}
+
+\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{graphicx}
 
@@
 the source of this European ancestry. We chose not to replace the GBR
-non-European Eurasian populations, as these would be poorer proxies for early
+by non-European Eurasian populations, as these would be poorer proxies for early
 Holocene and recent colonial migration from the Near East and Europe back to
@@
 The only large difference was found in fitting the continuous migration model,
-Nama divergence was $\approx200$ka, rather than $135$ka. Migration rates during
+Nama divergence was $\approx200$ka, rather than $135$ka. 
+Our original model using the dataset with GBR
+underestimated both $\sigma_{Dz}$ and $\sigma_{\pi_2}$ compared to data
+(Figure~\ref{fig:supp-continuous-migration-fits}), and as a result provided a poorer 
+fit to the data than the multiple merger model. 
+When fitting this same model to the TSI, the model provided a better fit to the $\sigma_{Dz}$ statistics but
+a worse fit to the $\sigma_{\pi_2}$ statistics (and still provided a
+significantly worse fit to the data compared to the multiple merger model).
+Migration rates during
 this early period of the model also shifted to larger values between the stem
 populations, demonstrating that divergence times and migration rates are
-confounded parameters in this model. Notably, the multiple merger model (which
+confounded parameters in this model. 
+These differences in inferred parameters are likely driven by a combination of factors: the continuous
+migration model mis-specifies historical dynamics in the range of 100-200ka;
+the TSI are an inappropriate proxy for recent European admixture in East and
+South Africa; the TSI may also carry recent North African ancestry, which
+is not captured in our model. Finally, there is an amount of confounding between divergence time
+and migration rates.
+
+By contrast, the multiple merger model (which
 provides a better fit to the data) was inferred to have consistent divergence
@@
 
-\aprcomment{Short paragraph here explaining the discrepancy in the TSI fit
-continuous migration model}
+Across all validations to alternative datasets described in this section, model
+fits were largely robust and matched those shown in the main text
+(Figure~\ref{fig:best-fit-models})---the single outlier in these validations
+was the continuous migration model fit to the dataset that included TSI instead
+of GBR. We conclude that while there remain important issues
+
+Together, this highlights the sensitivity of
+complex demographic inference to the choices of populations to include, as well
+as to choices of model specification.
+\aprcomment{not sure about these last two sentences}
 
@@
 
-\begin{figure}[ht!]
-    \centering
-    \includegraphics[width=5in]{figures/supp-toy-models}
+\begin{figure}
+
+    \includegraphics{figures/toy_models/toy_model_combo}
+
     \caption{The impact of model specification on coalescence rate and inference.
@@
     }
+
     \end{figure}

### 2022-08-12 supplement.tex
finalized discussion of TSI

@@
 
-Across all validations to alternative datasets described in this section, model
+In summary, across all validations to alternative datasets described in this section, model
 fits were largely robust and matched those shown in the main text
 (Figure~\ref{fig:best-fit-models})---the single outlier in these validations
-was the continuous migration model fit to the dataset that included TSI instead
-of GBR. We conclude that while there remain important issues
-
-Together, this highlights the sensitivity of
-complex demographic inference to the choices of populations to include, as well
-as to choices of model specification.
-\aprcomment{not sure about these last two sentences}
+was the estimated divergence time for the Nama in the continuous migration model
+fit to the dataset that included TSI instead of GBR.  
+Thus, while specific aspects of the inferred demographic inference can be sensitive
+to the choices of populations to include and the choices of model specification, 
+the results presented in the main text are broadly robust to choices of model, dataset, and 
+analysis choices. 
+Our conclusions in the main text took into account both this overall robustness 
+and the important sources of uncertainty.
 

### 2022-08-15 supplement.tex
small changes and formatting

@@
 All supplementary inference results are provided in Demes format
-\citep{Gower2022-yn} as a supplemental zipped archive. For each scenario,
-we refit the continuous migration and multiple merger models, with
-stem migration.
+\citep{Gower2022-yn} as a supplemental zipped archive. For each scenario, we
+refit the continuous migration and multiple merger models, with stem migration.
 
@@
 The only large difference was found in fitting the continuous migration model,
-Nama divergence was $\approx200$ka, rather than $135$ka. 
-Our original model using the dataset with GBR
-underestimated both $\sigma_{Dz}$ and $\sigma_{\pi_2}$ compared to data
-(Figure~\ref{fig:supp-continuous-migration-fits}), and as a result provided a poorer 
-fit to the data than the multiple merger model. 
-When fitting this same model to the TSI, the model provided a better fit to the $\sigma_{Dz}$ statistics but
-a worse fit to the $\sigma_{\pi_2}$ statistics (and still provided a
-significantly worse fit to the data compared to the multiple merger model).
-Migration rates during
-this early period of the model also shifted to larger values between the stem
-populations, demonstrating that divergence times and migration rates are
-confounded parameters in this model. 
-These differences in inferred parameters are likely driven by a combination of factors: the continuous
-migration model mis-specifies historical dynamics in the range of 100-200ka;
-the TSI are an inappropriate proxy for recent European admixture in East and
-South Africa; the TSI may also carry recent North African ancestry, which
-is not captured in our model. Finally, there is an amount of confounding between divergence time
-and migration rates.
-
-By contrast, the multiple merger model (which
-provides a better fit to the data) was inferred to have consistent divergence
-and merger times between the models fit using GBR and TSI, with ancestral
-populations ``reshuffling'' between 100 and 120ka.
-
-In summary, across all validations to alternative datasets described in this section, model
-fits were largely robust and matched those shown in the main text
-(Figure~\ref{fig:best-fit-models})---the single outlier in these validations
-was the estimated divergence time for the Nama in the continuous migration model
-fit to the dataset that included TSI instead of GBR.  
-Thus, while specific aspects of the inferred demographic inference can be sensitive
-to the choices of populations to include and the choices of model specification, 
-the results presented in the main text are broadly robust to choices of model, dataset, and 
-analysis choices. 
-Our conclusions in the main text took into account both this overall robustness 
-and the important sources of uncertainty.
+where the Nama divergence was $\approx200$ka rather than $135$ka. Our original
+model using the dataset with GBR underestimated both $\sigma_{Dz}$ and
+$\sigma_{\pi_2}$ compared to data
+(Figure~\ref{fig:supp-continuous-migration-fits}), and as a result provided a
+poorer fit to the data than the multiple merger model. When fitting this same
+model to the TSI, the model provided a better fit to the $\sigma_{Dz}$
+statistics but a worse fit to the $\sigma_{\pi_2}$ statistics (and still
+provided a significantly worse fit to the data compared to the multiple
+merger model). Migration rates during this early period of the model also
+shifted to larger values between the stem populations, demonstrating that
+divergence times and migration rates are confounded parameters in this model.
+These differences in inferred parameters are likely driven by a combination of
+factors: the continuous migration model mis-specifies historical dynamics in
+the range of 100-200ka; the TSI are an inappropriate proxy for recent European
+admixture in East and South Africa; and the TSI may also carry recent North
+African ancestry, which is not captured in our model. By contrast, the
+multiple merger model (which provides a better fit to the data) was inferred to
+have consistent divergence and merger times between the models fit using GBR
+and TSI, with ancestral populations ``reshuffling'' between 100 and 120ka.
+
+In summary, across all validations to alternative datasets described in this
+section, model fits were largely robust and matched those shown in the main
+text (Figure~\ref{fig:best-fit-models})---the single outlier in these
+validations was the estimated divergence time for the Nama in the continuous
+migration model fit to the dataset that included TSI instead of GBR. Thus,
+while specific aspects of the inferred demographic inference can be sensitive
+to the choices of populations to include and the choices of model
+specification, the results presented in the main text are broadly robust to
+choices of model, dataset, and analysis choices. Our conclusions in the main
+text took into account both this overall robustness and the important sources
+of uncertainty.
 

### 2022-08-15 supplement.tex
Section on statistics computed from whole genome

@@
 
-\subsection{Filtering and subsetting data}
+\subsection{Filtering and subsetting data}\label{sec:filtering}
 
@@
 \citep{1000_Genomes_Project_Consortium2015-zq,International_HapMap_Consortium2007-vn},
-and a recombination map inferred from the Nama \citep{Van_Eeden2021-od}.
+and a recombination map inferred from the Nama \citep{Van_Eeden2022-od}.
 Statistics were largely unchanged by using a different map
@@
 \citet{Ragsdale2020-nz}.
+
+Statistics were computed from intergenic loci, as described in
+Section~\ref{sec:filtering}, to avoid biases due to direct or indirect
+selection in protein-coding regions. To assess how strongly selection in coding
+regions affects LD and diversity statistics, we also computed statistics from
+all regions of the genome that pass the Thousand Genomes strict mask.
+Genome-wide statistics showed slight deviations from statistics inferred from
+intergenic regions alone (Figure~\ref{fig:supp-inter-all-comparison}), although
+the differences were smaller than the differences between our best-fit models
+and the data they were fit to. Nonetheless, we chose to use only intergenic
+data to avoid possible biases due to selection in and around protein-coding
+regions.
 
@@
 and LD statistics using recombination rates from the Nama recombination map
-\citep{Van_Eeden2021-od}, and refit our best-fit demographic models to these
+\citep{Van_Eeden2022-od}, and refit our best-fit demographic models to these
 statistics.
@@
 upper bound for differences in recombination rates between present-day human
-populations \citep{Van_Eeden2021-od}. Because our fits were highly consistent
+populations \citep{Van_Eeden2022-od}. Because our fits were highly consistent
 when using the two recombination maps, we conclude that inferences using the
@@
         The OMNI YRI recombination map and the Nama-inferred map
-        \citep{Van_Eeden2021-od} differ in two primary ways: 1) the YRI and
+        \citep{Van_Eeden2022-od} differ in two primary ways: 1) the YRI and
         Nama have relatively highly diverged ancestries, and 2) the YRI
@@
     \label{fig:supp-omni-nama-comparison}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-intergenic-all-snps-comparison.pdf}
+    \caption{
+        \textbf{Comparison of LD statistics parsed from intergenic loci and
+        from all loci across the genome.} Statistics computed from intergenic
+        regions (dashed lines, shown with computed standard errors as shaded
+        regions) and those computed from all regions genome wide (solid lines)
+        do not show large differences. We chose to perform model inference
+        using intergenic regions alone to reduce bias due to selection in and
+        around protein-coding regions (Section~\ref{sec:computing-stats}).
+        Notations for each statistic are described in
+        section~\ref{sec:supp-ld-stats}, and indexes represent populations (0:
+        Nama, 1: Mende, 2: Gumuz, 3: Amhara/Oromo, 4: British).
+    }
+    \label{fig:supp-inter-all-comparison}
 \end{figure}

### 2022-08-31 paper.tex
discussion of discordances

@@
 \end{figure}
+To assess robustness of the inferred models to analysis and reference population choices, 
+Sections \ref{sec:validation_strategy} and \ref{sec:validation_results} present re-analyses changing the European and West African populations, as well as the recombination maps, 
+filtering strategies and parameter optimization strategies.
+While we find some differences in the inferred parameters (see in particular \ref{sec:replacing} and \ref{sec:change-dates}), the best fit model across all reanalyses are quantitatively consistent.
 

### 2022-08-31 supplement.tex
discussion of discordances

@@
 
-\section{Model validation}
-
+\section{Model validation strategy}
+\label{sec:validation_strategy}
 We took two approaches to validate demographic model inferences. First, we
@@
 
-\section{Supplementary results}\label{sec:supp-results}
-
+\section{Model validation results}\label{sec:supp-results}
+\label{sec:validation_results}
 \subsection{Model fit using alternative datasets}
@@
 \subsubsection{Replacing Thousand Genomes populations}
-
+\label{sec:replacing}
 We replaced the Mende (MSL) and British (GBR) population sets with Yoruba (YRI)
@@
 underlying admixture proportions. Nonetheless, we included Nama individuals
-that exceeded either 90\% ro 99\% ancestry primarily shared among Nama
+that exceeded either $90\%$ or $99\%$ ancestry primarily shared among Nama
 individuals.
@@
 
-\subsection{Mutation versus recombination rates}
-\label{sec:supp-mutation}
-
-Prior studies primarily relied on the mutation clock in order to date events
-such as population divergence, and uncertainty about the mutation rate
-translates in uncertainty in timing estimates (e.g., \citet{Moorjani2016-qj}).
-Recent whole genome sequencing studies from non-pathological pedigrees have
-typically estimated mutation rates to be $1.2-1.3\times10^{-8}$ per site per
-generation \citep{Sasani2019-zp,Tian2019-zv}. A rate of $1.3\times10^{-8}$ also
-approximates the rate predicted by a mutational model from gnomAD
-\citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
-demographic inference in this study.
-\texttt{moments-LD} differs from other approaches by
-relying on a recombination clock, which has been argued to be more robust to
-estimation than a mutational clock \citep{Moorjani2016-ur}.
-
-From our recombination-rate-calibrated demographic models, we can estimate the
-mutation rates needed to fit patterns of pairwise diversity. With our best-fit
-model (merger-with-stem-migration,
-Table~\ref{tab:supp-merger-with-stem-migration}), a mutation rate of
-$1.33\times10^{-8}$ recovers the observed diversity within and between
-populations, closely matching the average mutation rate estimated from the
-gnomAD mutation model. The other inferred demographic models
-(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-without-stem-migration})
-would require mutation rates of $1.75-1.90\times10^{-8}$ to match pairwise
-diversity.
-
 \subsection{Re-inference of IM models from simulated data}
@@
 when using the diversity and LD statistics that we have in our analyses.
+
+\section{Mutation versus recombination rates}
+\label{sec:supp-mutation}
+
+Prior studies primarily relied on the mutation clock in order to date events
+such as population divergence, and uncertainty about the mutation rate
+translates in uncertainty in timing estimates (e.g., \citet{Moorjani2016-qj}).
+Recent whole genome sequencing studies from non-pathological pedigrees have
+typically estimated mutation rates to be $1.2-1.3\times10^{-8}$ per site per
+generation \citep{Sasani2019-zp,Tian2019-zv}. A rate of $1.3\times10^{-8}$ also
+approximates the rate predicted by a mutational model from gnomAD
+\citep{Karczewski2020-le} in the $\sim$650Mb of intergenic regions selected for
+demographic inference in this study.
+\texttt{moments-LD} differs from other approaches by
+relying on a recombination clock, which has been argued to be more robust to
+estimation than a mutational clock \citep{Moorjani2016-ur}.
+
+From our recombination-rate-calibrated demographic models, we can estimate the
+mutation rates needed to fit patterns of pairwise diversity. With our best-fit
+model (merger-with-stem-migration,
+Table~\ref{tab:supp-merger-with-stem-migration}), a mutation rate of
+$1.33\times10^{-8}$ recovers the observed diversity within and between
+populations, closely matching the average mutation rate estimated from the
+gnomAD mutation model. The other inferred demographic models
+(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-without-stem-migration})
+would require mutation rates of $1.75-1.90\times10^{-8}$ to match pairwise
+diversity.
 

### 2022-08-31 paper.tex
minor changes

--- 2022-08-12 paper.tex
@@
 River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
-sapiens} anatomical features were intermittantly present across the continent
+sapiens} anatomical features were intermittently present across the continent
 300-100ka. Archaeological sites from the Middle Stone Age, of which some have
@@
 `stem' groups as only a small proportion of drift between contemporary populations can
-be attributed to drift between stems (Section \ref{sec:f4},  Figures~\ref{fig:predictions} and
+be attributed to drift between stems (Supp. Information Section \ref{sec:f4},
+Figures~\ref{fig:predictions} and
 \ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
@@
         populations that aligns with drift between past populations
-        (here Nama and Mende, see Section~S\ref{sec:f4} for details and additional
-        comparisons in Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
+        (here Nama and Mende, see Supp. Information Section~S\ref{sec:f4}
+        for details and additional comparisons in
+        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
         Both models infer deep population structure with modest contributions to
@@
 increase in $N_e$ between 100ka-1Ma extending deeper in the past than the
-relate estimate (Figure~\ref{fig:supp-iicr-sim} and Section
+relate estimate (Figure~\ref{fig:supp-iicr-sim} and Supp. Information Section
 \ref{inferred_relate}). Whereas the single-origin model achieves this by an
@@
 or structure in the ancestral population will each increase coalescence times
-relative to a randomly mating population of contant size, so a model that
+relative to a randomly mating population of constant size, so a model that
 assumes constant population sizes would require an older divergence time to fit
@@
 the recombination clock accurately relate to geological chronologies
-(Supp. Information, Section~\ref{sec:supp-mutation}).
+(Supp. Information Section~\ref{sec:supp-mutation}).
 During the Middle Stone Age, the multiple merger model indicates three
@@
 \citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}
-(see also Section \ref{sec:morpho_diff}).
+(see also Supp. Information Section \ref{sec:morpho_diff}).
 This correspondence allows predictions of our model to be related to the fossil record.

### 2022-08-31 supplement.tex
minor changes

--- 2022-08-15 supplement.tex
@@
 Genotyper \citep{DePristo2011-up} using joint calling across 2,478 individuals within
-the ADRP dataset with a minimum base quality of 17. Data were merged with 1000
+the African Diversity Reference Panel (ADRP) dataset with a
+minimum base quality of 17. Data were merged with 1000
 Genomes Phase 3 \citep{1000_Genomes_Project_Consortium2015-zq} using the union
@@
         regions (dashed lines, shown with computed standard errors as shaded
-        regions) and those computed from all regions genome wide (solid lines)
+        regions) and those computed from all regions genome-wide (solid lines)
         do not show large differences. We chose to perform model inference

### 2022-08-31 paper.tex
spelling

--- 2022-08-31 paper.tex
@@
 \end{figure}
+To assess robustness of the inferred models to analysis and reference
+population choices, Supp. Information Sections \ref{sec:validation_strategy}
+and \ref{sec:validation_results} present re-analyses changing the European and
+West African populations, as well as the recombination maps, filtering
+strategies and parameter optimization strategies. While we find some
+differences in the inferred parameters (see in particular Sections
+\ref{sec:replacing} and \ref{sec:change-dates}), the best fit model across all
+reanalyses are quantitatively consistent.
 

### 2022-08-31 supplement.tex
spelling

@@
 the Amhara and Oromo also have substantive ``back-to-Africa'' ancestry. The
-Gumuz serve as a representative of southwest Ethipian ancestry that pre-dates
+Gumuz serve as a representative of southwest Ethiopian ancestry that pre-dates
 the Holocene back-to-Africa migration. Finally, because previous studies have
@@
 Likelihoods were computed independently for pairwise diversity and each
-recombination distance bin using a multivariate Guassian likelihood function as
+recombination distance bin using a multivariate Gaussian likelihood function as
 described in Section \ref{sec:computing-stats}. These were multiplied across
@@
 merger-with-stem-migration model performed the best of these four models,
-although the fits were nearly visually indistiguishable from the fits from the
+although the fits were nearly visually indistinguishable from the fits from the
 merger-with-stem-migration model.
@@
 impose a minimum threshold on inferred Nama ancestry (as estimated using
-AMDIXTURE). Because our proposed models allow for post-divergence gene flow as
+ADMIXTURE). Because our proposed models allow for post-divergence gene flow as
 well as recent admixture events, we can jointly learn deeper history and recent
@@
     This model includes a single stem population that splits into contemporary human
-    groups. Fixed parameters are specifed in Table~\ref{tab:fixed-params}.
+    groups. Fixed parameters are specified in Table~\ref{tab:fixed-params}.
     Inferred values are scaled to physical units assuming a generation time of
@@
     $T_{Nama}$ & Time of Nama merger event & 117392 & 8253 \\
-    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancsestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.707 & 0.0086 \\
+    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.707 & 0.0086 \\
     $T_{EA}$ & Time of East Africa merger event & 94892 & 3648 \\
@@
     $T_{Nama}$ & Time of Nama merger event & 118547 & 28170 \\
-    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancsestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.714 & 0.067 \\
+    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.714 & 0.067 \\
     $T_{EA}$ & Time of East Africa merger event & 98083 & 8865 \\

### 2022-08-31 paper.tex
SI formatting

--- 2022-08-31 paper.tex
@@
         We used these conceptual models as starting points to build detailed
-        parameterized demographic models (Supp. Information Section~\ref{modelspec})
+        parameterized demographic models (SI Section~\ref{modelspec})
         that were then fit to genetic data. 
@@
 We next explored a suite of parameterized models in which population structure predates the
-differentiation of contemporary groups (Supp. Information section~\ref{modelspec}). 
+differentiation of contemporary groups (Supporting Information (SI) Section~\ref{modelspec}). 
 Depending on the parameters, these encompassed models allowing for ancestral
@@
 other populations exceed $\sim$140ka
-(Supp. Information~\ref{sec:validation-data-models} describes validation using
+(SI Section~\ref{sec:validation-data-models} describes validation using
 alternative datasets and model specifications). 
@@
 `stem' groups as only a small proportion of drift between contemporary populations can
-be attributed to drift between stems (Supp. Information Section \ref{sec:f4},
+be attributed to drift between stems (SI Section \ref{sec:f4},
 Figures~\ref{fig:predictions} and
@@
 We also fixed the amount of admixture from Neanderthals to Europeans directly
-following the out-of-Africa migration which was set to 1.5\% at 45ka (Supp. Information \ref{sec:models}).
+following the out-of-Africa migration which was set to 1.5\% at 45ka (SI Section~\ref{sec:models}).
 
@@
 among contemporary populations can be traced back
-to this early population structure (Supp. Information \ref{sec:f4})
+to this early population structure (SI Section~\ref{sec:f4})
 
@@
         populations that aligns with drift between past populations
-        (here Nama and Mende, see Supp. Information Section~S\ref{sec:f4}
+        (here Nama and Mende, see SI Section~S\ref{sec:f4}
         for details and additional comparisons in
@@
 \end{figure}
+To assess robustness of the inferred models to analysis and reference
+population choices, SI Sections \ref{sec:validation_strategy}
+and \ref{sec:validation_results} present re-analyses changing the European and
+West African populations, as well as the recombination maps, filtering
+strategies and parameter optimization strategies. While we find some
+differences in the inferred parameters (see in particular Sections
+\ref{sec:replacing} and \ref{sec:change-dates}), the best fit model across all
+reanalyses are quantitatively consistent.
 
@@
         misidentification was consistently inferred to be 1.5--1.7\% for
-        intergenic loci (Supp. Information \ref{sec:misspecification}). (D, E) We used Relate
+        intergenic loci (SI \ref{sec:misspecification}). (D, E) We used Relate
         \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
@@
 increase in $N_e$ between 100ka-1Ma extending deeper in the past than the
-relate estimate (Figure~\ref{fig:supp-iicr-sim} and Supp. Information Section
-\ref{inferred_relate}). Whereas the single-origin model achieves this by an
+Relate estimate (Figure~\ref{fig:supp-iicr-sim} and SI
+Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
 increase in $N_e$ during that period, the best-fit model recapitulate this
@@
 inconsistent divergence times, some of which are older than those we find here
-\cite{Henn2018-rf,Bergstrom2021-iw}. In the Supp. Information
+\cite{Henn2018-rf,Bergstrom2021-iw}. In the SI
 (Section~\ref{sec:IM-reinfer}), we show that this discrepancy can be explained
@@
 the recombination clock accurately relate to geological chronologies
-(Supp. Information Section~\ref{sec:supp-mutation}).
+(SI Section~\ref{sec:supp-mutation}).
 During the Middle Stone Age, the multiple merger model indicates three
@@
 \citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}
-(see also Supp. Information Section \ref{sec:morpho_diff}).
+(see also SI Section~\ref{sec:morpho_diff}).
 This correspondence allows predictions of our model to be related to the fossil record.

### 2022-08-31 supplement.tex
lengths and number of SNPs

@@
 available under accession number EGAD00001006198.
+Second- and third-degrees relatives were inferred from reconstructed
+pedigrees with Omni2.5 SNP array data.
 Among these 82 Nama samples, we down-sampled the individuals to minimize
 close relatives and admixture, such that 44 Nama genomes were retained.
-Second- and third-degrees relatives were inferred from reconstructed
-pedigrees with Omni2.5 SNP array data.
 Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
@@
 All analyses presented in this work focus on biallelic single nucleotide
-polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the
-\texttt{moments-LD} analysis, we focused on intergenic regions because these
+polymorphisms within the 1000 Genomes Phase 3 strict mask ($L_{included}=XXXbp$).  
+To enable comparison with
+Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
+had less than 100 contiguous base pairs ($L_{included}=XXXbp$).
+For the
+\texttt{moments-LD} analysis, we further focused on intergenic regions because these
 appear less affected by natural selection compared to both synonymous and
-nonsynonymous variation \citep{Ragsdale2018-dd}, removing regions that fell
-outside of the 1000 Genomes strict callability mask. To enable comparison with
-Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
-had less than 100 contiguous base pairs.
+nonsynonymous variation \citep{Ragsdale2018-dd} ($L_{included}=XXX$).
+
+The 290 individuals used in the moments.LD inference 
+featured 18,461,915 polymorphic biallelic SNPs genome-wide in the callable regions with 
+Neanderthal alignment, of which 8,115,115 were in intergenic and thus putatively neutral regions.
 

### 2022-08-31 supplement.tex
Update filtering section

--- 2022-08-31 supplement.tex
@@
 All analyses presented in this work focus on biallelic single nucleotide
-polymorphisms within the 1000 Genomes Phase 3 strict mask.  For the
-\texttt{moments-LD} analysis, we focused on intergenic regions because these
-appear less affected by natural selection compared to both synonymous and
-nonsynonymous variation \citep{Ragsdale2018-dd}, removing regions that fell
-outside of the 1000 Genomes strict callability mask. To enable comparison with
-Neanderthal DNA, we excluded regions for which the Vindija Neanderthal sample
-had less than 100 contiguous base pairs.
+polymorphisms within the 1000 Genomes Phase 3 strict mask
+($L_{included}=2,064,554,803$bp). To enable comparison with Neanderthal DNA, we
+excluded regions for which the Vindija Neanderthal sample had less than 100
+contiguous base pairs with high quality variant calls
+($L_{included}=1,519,643,507$bp). For the \texttt{moments-LD} analysis, we
+further focused on intergenic regions because these appear less affected by
+natural selection compared to both synonymous and nonsynonymous variation
+\citep{Ragsdale2018-dd} ($L_{included}=637,639,065$bp).
+
+The 290 individuals used in the \texttt{moments-LD} inference featured
+18,461,915 polymorphic biallelic SNPs genome-wide in the callable regions with
+Neanderthal alignment, of which 8,115,115 were in intergenic and thus
+putatively neutral regions.
 

### 2022-09-01 paper.tex
minor edits

@@
 high-likelihood models which we explored did the divergence between Nama and
-other populations exceed $\sim$140ka
-(SI Section~\ref{sec:validation-data-models} describes validation using
-alternative datasets and model specifications). 
+other populations exceed $\sim$140ka. 
 We conclude that geographic patterns of contemporary \emph{Homo sapiens}
@@
 
-We quantify migration rates among populations after their divergence ~120ka.
+We quantify migration rates among populations after their divergence $\approx120ka$.
 Migration, prior to the Bantu expansion, between the ancestors of the Nama and other groups  
@@
 All models infer relatively high gene flow between Ethiopia and western Africa
-($m\approx2\times10^{-4},$ the constant proportion of migrant lineages per generation over 60ka).
+($m\approx2\times10^{-4},$ the constant proportion of immigrant lineages per generation over 60ka).
 We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
@@
 eastern African caprid and cattle pastoralists \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
-contribution 2,000 ya.
-While this gene flow is not apparent from the ADMIXTURE plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestry Eastern African source.
+contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE plot (Figure~\ref{fig:diversity}), 
+the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestral Eastern African source. 
 Colonial period admixture from Europeans into the Nama
@@
 Africa and Stem 1E in eastern Africa. 
+\sgcomment{Here my read is that the statement about stem 2 in Mende is shared across both models.
+The statement about the Nama is specific to the merger model. Make this explicit?
+}
 
@@
         \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
-        which we used to estimate coalescence rate trajectories and
+        which we used to estimate coalescence rate trajectories
         and cross-coalescence rates between pairs of populations. While

### 2022-09-01 paper.tex
more minor edits
added Georgette in acknowledgements

@@
 isolation among the stems is variable across model fits, models with a period of 
-divergence, isolation and then a merger event (i.e. a ``reticulation''
+divergence, isolation and then a merger event (i.e. a ``reticulation'')
 out-performed models with bifurcating divergence and continuous gene flow.
@@
 generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
-thank Luke Anderson-Trocm\'e for assistance in creating the map in
-Figure~\ref{fig:diversity}. This research was supported by an NIH grant R35GM133531 (to
+thank Luke Anderson-Trocm\'e and Georgette Femerling for assistance in creating the map in
+Figure~\ref{fig:diversity} and Figure \ref{fig:toymodels}, respectively. This research was supported by an NIH grant R35GM133531 (to
 BMH). The content is solely the responsibility of the authors and does

### 2022-09-01 supplement.tex
cleaned up captions

@@
         \textbf{Average error of statistics from low coverage and merged datasets.}
-        Low coverage data provides consistent allelic diversity and linkage
-        disequilibrium measures to high coverage data. On the other hand,
+        Low coverage data provides allelic diversity and linkage
+        disequilibrium measures consistent with those from high coverage data (left panel). On the other hand,
         merging separately called datasets can lead to large discrepancies
-        in computed LD statistics.
+        in computed LD statistics (middle and right panels).
     }
@@
         \textbf{Merged datasets bias demographic inferences.}
+        The panels represent best-fit models according to 
+        different combinations of high- and low- coverage data.
         While statistics from low and high coverage datasets provide consistent

### 2022-09-01 supplement.tex
cleaned up captions v2

@@
         and ancient branches from the continuous-migration model.} 
+        $f_4^2(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
+        amount of drift between sampled populations $A$ and $B$ that aligns with
+        drift between populations $X$ and $Y$, sampled in the past
+        (see Section~\ref{sec:f4}).
         Despite population structure inferred to have extended up to 1Ma
@@
         and ancient branches from the merger-without-stem-migration model.}
+        $f_4^2(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
+        amount of drift between sampled populations $A$ and $B$ that aligns with
+        drift between populations $X$ and $Y$, sampled in the past
+        (see Section~\ref{sec:f4}).
         Unlike the continuous-migration model, population structure in the stems
@@
         and ancient branches from the merger-with-stem-migration model.}
+        $f_4^2(A, B; X, Y)/f_2(A, B)f_2(X, Y)$ is interpreted as the proportion of the
+        amount of drift between sampled populations $A$ and $B$ that aligns with
+        drift between populations $X$ and $Y$, sampled in the past
+        (see Section~\ref{sec:f4}).
         Differentiation between present-day populations aligns with

### 2022-09-01 supplement.tex
started admixture discussion

@@
 Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
-analysis, after partitioning ancestry into $k=6$ clusters with \texttt{ADMIXTURE}
-\citep{Alexander2009-sw} where alternative ancestries represent European, West
+analysis, after partitioning ancestry into $k=6$ clusters with an initial \texttt{ADMIXTURE}
+\citep{Alexander2009-sw} analysis on \sgcomment{TODO: Brenna complete}, where alternative ancestries represent European, West
 African, Near Eastern, and East African gene flow. 
+Other \texttt{ADMIXTURE} analyses
+presented in this manuscript were restricted to a subset of 1157 individuals \sgcomment{TODO:  Aaron complete}.
 

### 2022-09-01 supplement.tex
admixture analysis

--- 2022-09-01 supplement.tex
@@
 Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
-analysis, after partitioning ancestry into $k=6$ clusters with \texttt{ADMIXTURE}
-\citep{Alexander2009-sw} where alternative ancestries represent European, West
+analysis, after partitioning ancestry into $K=6$ clusters with an initial
+\texttt{ADMIXTURE} \citep{Alexander2009-sw} analysis on the full ADRP
+dataset, where alternative ancestries represent European, West
 African, Near Eastern, and East African gene flow.
+Other \texttt{ADMIXTURE} analyses presented in this manuscript were
+restricted to a subset of 1157 individuals representing a subset
+of the full ADRP/Thousand Genomes dataset used in analyses presented here
+(see Section~\ref{sec:populations-used}).
 
@@
 \subsection{Details about populations used in analyses}
+\label{sec:populations-used}
 
@@
 \citep{Alexander2011-aq} to summarize genetic diversity among a broader set of
-1000 Genomes and ADRP populations (Figure~\ref{fig:diversity} in the main text). Data
+1000 Genomes and ADRP populations (Figure~\ref{fig:diversity} in the main text,
+and Figure~\ref{fig:supp-admixture}). Data
 filtering was identical for both analyses. We retained biallelic SNPs with no
@@
 
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=6in]{figures/supp-admixture-panels}
+    \caption{
+        \textbf{\texttt{ADMIXTURE} clustering using $K=2$ to $9$ clusters.}
+        \texttt{ADMIXTURE} \citep{Alexander2009-sw} was used to cluster all
+        individuals in the subset of populations used in analyses in this
+        paper. At $K=5$ clusters, West African populations begin to show
+        variable separation in assigned clusters, likely due to
+        isolation-by-distance demographic processes rather than recent
+        admixture events. Figure~\ref{fig:diversity}D in the main text shows
+        the \texttt{ADMIXTURE} results for $K=4$, which most clearly displays
+        known admixture events in the histories of East and South African
+        populations.
+    }
+    \label{fig:supp-admixture}
+\end{figure}
 \end{document}

### 2022-09-02 paper.tex
added comment on CSFS
removed comment

@@
 Africa and Stem 1E in eastern Africa.
-\sgcomment{Here my read is that the statement about stem 2 in Mende is shared across both models.
-The statement about the Nama is specific to the merger model. Make this explicit?
-}
 
@@
 Durvasula and Sankararaman (2020)\citep{Durvasula2020-td}). 
+Specifically, it is the addition of migration between stems that results in a qualitative improvement 
+of the agreement 
+(compare Figures \ref{fig:supp-csfs-merger-without-stem-migration}  and \ref{fig:supp-csfs-merger-with-stem-migration}).
 

### 2022-09-02 paper.tex
Molinaro reference

@@
 
-\subsection*{A Late Middle Stone Age common ancestry for contemporary humans}
+\subsection*{A Late Pleistocene common ancestry for contemporary humans}
 
@@
 We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
-affected the ancestors of the Ethiopian agricultural populations, comprising
-over half of their genetic ancestry, estimated to be 64--65\%. 
+affected the ancestors of the Ethiopian agricultural populations \citep{Molinaro2019-rf},
+comprising over half of their genetic ancestry, estimated to be 64--65\%.
 We observe significant gene flow from the Amhara and Oromo into the Nama, a signal which is likely a proxy for the movement of
@@
 contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE plot (Figure~\ref{fig:diversity}),
-the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestral Eastern African source. 
+the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestral eastern African source.
 Colonial period admixture from Europeans into the Nama
@@
 early divergence of stem populations and their relative $N_e$
-(Figure~\ref{fig:best-fit-models}). The two models also differ in the mode of divergence
-during the Middle Stone Age.
+(Figure~\ref{fig:best-fit-models}). The two models also differ in the mode of divergence, with the multiple-merger featuring a population reticulation during the Middle Pleistocene.
 
@@
 Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
-Central Africa, broadly speaking. The differing proportions in the Nama and
+central Africa, broadly speaking. The differing proportions in the Nama and
 eastern Africans may also indicate geographic separation of Stem 1S in southern
@@
 could be due to either an increase in population size or to ancestral
-population structure during the Middle Stone Age \citep{Mazet2016-wn}.  All
+population structure during the Middle Pleistocene \citep{Mazet2016-wn}.  All
 models, including the single-origin model, recapitulate an inferred ancestral
@@
 
-\subsection*{The Middle Stone Age in Africa}
+\subsection*{The Formation of Population Structure in Africa}
 
-By contrast, our inferred models paint a more consistent picture of the late
-Middle Stone Age as a critical period of change, assuming that estimates from
+By contrast, our inferred models paint a more consistent picture of the Middle to Late Pleistocene as a critical period of change, assuming that estimates from
 the recombination clock accurately relate to geological chronologies
 (SI Section~\ref{sec:supp-mutation}).
-During the Middle Stone Age, the multiple merger model indicates three
+During the late Middle Pleistocene, the multiple merger model indicates three
 major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
@@
 results in the formation of the
-ancestors of East/West Africans as well as later ``Out of Africans''. The rapid
-rise in sea levels and increased precipitation during MIS 5e, following a
+ancestors of East/West Africans including the ancestors of ``Out of Africans''. The rapid
+rise in sea levels and increased precipitation during MIS 5e interglacial, following a
 glacial period of aridity across Africa \citep{Blome2012-lw}, might have

### 2022-09-09 paper.tex
pushed Middle Stone Age changes

@@
 We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
-affected the ancestors of the Ethiopian agricultural populations \citep{Molinaro2019-rf},
-comprising over half of their genetic ancestry, estimated to be 64--65\%.
+affected the ancestors of the Ethiopian agricultural populations, comprising
+over half of their genetic ancestry, estimated to be 64--65\%. 
 We observe significant gene flow from the Amhara and Oromo into the Nama, a signal which is likely a proxy for the movement of
@@
 contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE plot (Figure~\ref{fig:diversity}), 
-the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestral eastern African source.
+the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestral Eastern African source. 
 Colonial period admixture from Europeans into the Nama
@@
 Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
-central Africa, broadly speaking. The differing proportions in the Nama and
+Central Africa, broadly speaking. The differing proportions in the Nama and
 eastern Africans may also indicate geographic separation of Stem 1S in southern

### 2022-09-09 supplement.tex
add Brenna's supp comments

@@
 close relatives and admixture, such that 44 Nama genomes were retained.
-Individuals with $>70\%$ estimated Khoe-San ancestry were retained for
-analysis, after partitioning ancestry into $K=6$ clusters with an initial
-\texttt{ADMIXTURE} \citep{Alexander2009-sw} analysis on the full ADRP
-dataset, where alternative ancestries represent European, West
-African, Near Eastern, and East African gene flow.
-Other \texttt{ADMIXTURE} analyses presented in this manuscript were
-restricted to a subset of 1157 individuals representing a subset
-of the full ADRP/Thousand Genomes dataset used in analyses presented here
-(see Section~\ref{sec:populations-used}).
+
+\subsubsection{ADMIXTURE} Nama individuals with $>70\%$ estimated Khoe-San
+ancestry were retained for analysis here, after partitioning ancestry into $K =
+6$ clusters with an \texttt{ADMIXTURE} \citep{Alexander2009-sw} analysis on the
+full ADRP dataset \citep{Van_Eeden2022-od}. The 27 ADRP populations represent a
+wide variety of West African, East African, North African and Bantu-speaking
+South Africans. Europeans (1000 Genomes: CEU) were included to account for
+colonial admixture. A Khoe-San specific ancestry emerged at $K = 3$,
+predominant in the Nama. This ancestry remained stable at higher $K$'s. At $K =
+6$, an ancestry representing the southern Bantu-speakers (Sotho and Zulu) also
+emerged, at low frequency in some Nama. Higher $K$'s highlighted
+population-specific ancestries or those with no representation in the Nama,
+therefore we focused on $K = 6$ for selection of high Khoe-San ancestry
+individuals. For illustration, \texttt{ADMIXTURE} analysis in
+Figures~\ref{fig:diversity}D~and~\ref{fig:supp-admixture} was restricted to
+1,157 individuals representing a subset of the full ADRP and several 1000
+Genomes populations used in analyses presented here (see
+Section~\ref{sec:populations-used}).
 
@@
 
+\subsubsection{Ethics and Initiation of DNA Collection}
+
 DNA samples were collected from three Nama communities in the Richtersveld
-region of South Africa, which borders southern Namibia, in 2012. A community
-guide was present during each interview and facilitated consent in Afrikaans or
-Nama. Written consent was recorded per our IRB protocol with human subjects
-approval from Stanford University (Protocol \#13829), Stellenbosch University
-(N11/07/210) and later maintained via SUNY Stony Brook (Protocol \#727494).
-Saliva samples were extracted from Oragene [OGR-500] kits at Stellenbosch
-University. Results stemming from genetic analyses have been communicated in
-2015, 2019 and 2021 via community presentations, a radio interview and to
-representatives of the Richtersveld National Park.
+region of South Africa, which borders southern Namibia. Members of the Nama
+community in the Richtersveld were initially approached regarding a genetic
+ancestry study in 2011. The members included church elders, representatives at
+the Richtersveld National Park including the appointee for Peoples and
+Conservation (from which we obtained permission to sample individuals who still
+live in the park), the Richtersveld Cultural World Heritage Site Coordinator,
+and the Kuboes tourist office. All of these individuals self-identified as
+Nama. There was strong support for research that examined the extent to which
+the Nama are related to other Khoe-San populations, the impact of colonial
+migration, the time depth of their presence in southern Africa and
+relationships to other African groups. One primary concern was to make sure the
+research results were reported back to the community. Based on the positive
+response we received, permission was granted from the Stellenbosch University
+HREC to continue the study. Written consent was recorded per our IRB protocol
+with human subjects approval from Stanford University (Protocol \#13829),
+Stellenbosch University (N11/07/210) and later maintained via SUNY Stony Brook
+(Protocol \#727494). DNA were collected for the express purpose of understanding
+Nama population history, their relationship to other Africans and human
+evolution. No commercial activities are allowed. The IRB has approved the data
+for public deposition and sharing provided that the research is related to
+population history or human evolution.
+
+\subsubsection{Collection Procedures}
+
+Demographic and DNA collection was initiated in 2012 with a joint team of South
+African and American researchers which included both geneticists and
+anthropologists. Research assistants were fluent in Afrikaans, Nama and
+English. Collection primarily occurred at home, by first approaching a family
+member, gauging interest in the study, entering the home or sitting outside by
+their invitation, oral and written consent occurring with family members
+present and then finally -- completing the DNA sampling. By sampling
+individuals at home, members of the family are able to voice concerns, whether
+or not they decide to participate, and thus the final decisions are made in a
+slow and ethical manner. This process sometimes involved introducing the study
+and then returning at a later day in order to give participants sufficient time
+to consider. Saliva samples were extracted from Oragene [OGR-500] kits at
+Stellenbosch University. A copy of the English consent form is available as
+a Supplementalal Document.
+
+\subsubsection{Return of Results}
+
+Results stemming from genetic analyses have been communicated in 2015, 2019 and
+2021 via community presentations, a radio interview and in person to
+representatives of the Richtersveld National Park. We created a poster
+illustrating Khoe-San ancestry proportions, discriminating ancestry on maternal
+vs. paternal lineages, the relationship between the Nama and Kalahari Khoe-San
+populations, and the evolution of light skin pigmentation in southern Africa.
+The poster was printed in Afrikaans and presented to the tourism office in
+Kuboes (a central location) for display. A display panel on genetic ancestry is
+currently under construction for the Richtersveld National Park entrance,
+created in consultation with the authors here. Future community meetings are
+anticipated in 2023. We remain committed to upholding this decade long
+relationship with the Nama, and value the community’s input in the research
+process.
+
+%DNA samples were collected from three Nama communities in the Richtersveld
+%region of South Africa, which borders southern Namibia, in 2012. A community
+%guide was present during each interview and facilitated consent in Afrikaans or
+%Nama. Written consent was recorded per our IRB protocol with human subjects
+%approval from Stanford University (Protocol \#13829), Stellenbosch University
+%(N11/07/210) and later maintained via SUNY Stony Brook (Protocol \#727494).
+%Saliva samples were extracted from Oragene [OGR-500] kits at Stellenbosch
+%University. Results stemming from genetic analyses have been communicated in
+%2015, 2019 and 2021 via community presentations, a radio interview and to
+%representatives of the Richtersveld National Park.
 
@@
 from West Africa: Mende from Sierra Leone (MSL); South Africa: Nama from South
-Africa (newly sequenced); East Africa: Gumuz (traditionally hunter-gatherer
-with low levels of Eurasian admixture), Oromo and Amhara (combined as
-traditionally agriculturalists with large proportion of back-to-Africa Eurasian
-ancestry); and Europe: British (GBR). These were combined with the
-high-coverage Vindija Neanderthal genome \citep{Prufer2017-kk}, resulting in 290
-total samples used in demographic inference.
+Africa (newly sequenced, Figure~\ref{fig:supp-nama}); East Africa: Gumuz
+(traditionally hunter-gatherer with low levels of Eurasian admixture), Oromo
+and Amhara (combined as traditionally agriculturalists with large proportion of
+back-to-Africa Eurasian ancestry); and Europe: British (GBR). These were
+combined with the high-coverage Vindija Neanderthal genome
+\citep{Prufer2017-kk}, resulting in 290 total samples used in demographic
+inference.
 
@@
     \centering
+    \includegraphics[width=\textwidth]{figures/SuppFig_HDV.60_CEU.Nama}
+    \caption{caption text}
+    \label{fig:supp-nama}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=5.75in]{figures/supp-admixture-panels}
+    \caption{
+        \textbf{\texttt{ADMIXTURE} clustering using $K=2$ to $9$ clusters.}
+        \texttt{ADMIXTURE} \citep{Alexander2009-sw} was used to cluster all
+        individuals in the subset of populations used in analyses in this
+        paper. At $K=5$ clusters, West African populations begin to show
+        variable separation in assigned clusters, likely due to
+        isolation-by-distance demographic processes rather than recent
+        admixture events. Figure~\ref{fig:diversity}D in the main text shows
+        the \texttt{ADMIXTURE} results for $K=4$, which most clearly displays
+        known admixture events in the histories of East and South African
+        populations.
+    }
+    \label{fig:supp-admixture}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
     \includegraphics{figures/mergers.pdf}
@@
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=6in]{figures/supp-admixture-panels}
-    \caption{
-        \textbf{\texttt{ADMIXTURE} clustering using $K=2$ to $9$ clusters.}
-        \texttt{ADMIXTURE} \citep{Alexander2009-sw} was used to cluster all
-        individuals in the subset of populations used in analyses in this
-        paper. At $K=5$ clusters, West African populations begin to show
-        variable separation in assigned clusters, likely due to
-        isolation-by-distance demographic processes rather than recent
-        admixture events. Figure~\ref{fig:diversity}D in the main text shows
-        the \texttt{ADMIXTURE} results for $K=4$, which most clearly displays
-        known admixture events in the histories of East and South African
-        populations.
-    }
-    \label{fig:supp-admixture}
-\end{figure}
 \end{document}

### 2022-09-09 supplement.tex
supp figure edit

@@
 from West Africa: Mende from Sierra Leone (MSL); South Africa: Nama from South
-Africa (newly sequenced, Figure~\ref{fig:supp-nama}); East Africa: Gumuz
-(traditionally hunter-gatherer with low levels of Eurasian admixture), Oromo
-and Amhara (combined as traditionally agriculturalists with large proportion of
-back-to-Africa Eurasian ancestry); and Europe: British (GBR). These were
-combined with the high-coverage Vindija Neanderthal genome
-\citep{Prufer2017-kk}, resulting in 290 total samples used in demographic
-inference.
+Africa (newly sequenced); East Africa: Gumuz (traditionally hunter-gatherer
+with low levels of Eurasian admixture), Oromo and Amhara (combined as
+traditionally agriculturalists with large proportion of back-to-Africa Eurasian
+ancestry); and Europe: British (GBR). These were combined with the
+high-coverage Vindija Neanderthal genome \citep{Prufer2017-kk}, resulting in
+290 total samples used in demographic inference.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/SuppFig_HDV.60_CEU.Nama}
-    \caption{caption text}
-    \label{fig:supp-nama}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=5.75in]{figures/supp-admixture-panels}
-    \caption{
-        \textbf{\texttt{ADMIXTURE} clustering using $K=2$ to $9$ clusters.}
-        \texttt{ADMIXTURE} \citep{Alexander2009-sw} was used to cluster all
-        individuals in the subset of populations used in analyses in this
-        paper. At $K=5$ clusters, West African populations begin to show
-        variable separation in assigned clusters, likely due to
-        isolation-by-distance demographic processes rather than recent
-        admixture events. Figure~\ref{fig:diversity}D in the main text shows
-        the \texttt{ADMIXTURE} results for $K=4$, which most clearly displays
-        known admixture events in the histories of East and South African
-        populations.
-    }
-    \label{fig:supp-admixture}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
     \includegraphics{figures/mergers.pdf}
@@
 
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=5.75in]{figures/supp-admixture-panels}
+    \caption{
+        \textbf{\texttt{ADMIXTURE} clustering using $K=2$ to $9$ clusters.}
+        \texttt{ADMIXTURE} \citep{Alexander2009-sw} was used to cluster all
+        individuals in the subset of populations used in analyses in this
+        paper. At $K=5$ clusters, West African populations begin to show
+        variable separation in assigned clusters, likely due to
+        isolation-by-distance demographic processes rather than recent
+        admixture events. Figure~\ref{fig:diversity}D in the main text shows
+        the \texttt{ADMIXTURE} results for $K=4$, which most clearly displays
+        known admixture events in the histories of East and South African
+        populations.
+    }
+    \label{fig:supp-admixture}
+\end{figure}
+
 \end{document}

### 2022-09-15 supplement.tex
fixed typo

@@
 misspecification in the details of the divergence between the ancestors of OOA
-and east African populations. Our models supposes a single split prior to the
+and east African populations. Our models suppose a single split prior to the
 OOA event followed by constant symmetric migrations between eastern and OOA

### 2022-11-16 paper.tex
added line numbers
add code availability, need data availability info

--- 2022-09-02 paper.tex
@@
 
+\section*{Data availability}
+
+All sequencing data are available from XXX,
+accession number: YYY.
+
+\section*{Code availability}
+
+Code for the software used in this paper is found at the following locations:\\
+moments-LD (\url{https://bitbucket.org/simongravel/moments}),\\
+Demes (\url{https://github.com/popsim-consortium/demes-python}),\\
+Relate (\url{https://myersgroup.github.io/relate/}),\\
+msprime (\url{https://github.com/tskit-dev/msprime}),\\
+tskit (\url{https://github.com/tskit-dev/tskit}),\\
+and scripts implementing analyses using each of these software are available at\\
+\url{https://github.com/apragsdale/african-structure-paper}.
+
 \section*{Acknowledgements}

### 2022-12-19 paper.tex
add Godambe table, unfinished

@@
 \usepackage{authblk}
+
+\usepackage{lineno}
+\linenumbers
 \renewcommand{\Affilfont}{\normalsize}
@@
 of optimizing a large number of parameters, we fixed several parameters related
-to recent population history so as to focus on more ancient events. These parameters were ones supported by multiple genetic and archaeological studies.
-Fixed parameters included the time of divergence between western and eastern African populations,
-set to 60ka, just prior to the split of Eurasians and East Africans set to 50ka.
-We also fixed the amount of admixture from Neanderthals to Europeans directly
-following the out-of-Africa migration which was set to 1.5\% at 45ka (SI Section~\ref{sec:models}).
+to recent population history so as to focus on more ancient events.  These
+parameters were ones supported by multiple genetic and archaeological studies.
+Fixed parameters included the time of divergence between western and eastern
+African populations, set to 60ka, just prior to the split of Eurasians and East
+Africans set to 50ka.  We also fixed the amount of admixture from Neanderthals
+to Europeans directly following the out-of-Africa migration which was set to
+1.5\% at 45ka (SI Section~\ref{sec:models}).
 
-We quantify migration rates among populations after their divergence $\approx120ka$.
-Migration, prior to the Bantu expansion, between the ancestors of the Nama and other groups
-is an order of magnitude weaker than that observed between between western and eastern Africans (Table \ref{tab:migration-rates}).
+We quantify migration rates among populations after their divergence
+$\approx120ka$. Prior to the Bantu expansion, migration between the ancestors
+of the Nama and other groups is an order of magnitude weaker than that observed
+between between western and eastern Africans (Table \ref{tab:migration-rates}).
 All models infer relatively high gene flow between Ethiopia and western Africa
-($m\approx2\times10^{-4},$ the constant proportion of immigrant lineages per generation over 60ka).
-We further find that Back-to-Africa gene flow at the beginning of the Holocene primarily
-affected the ancestors of the Ethiopian agricultural populations \citep{Molinaro2019-rf},
-comprising over half of their genetic ancestry, estimated to be 64--65\%.
-We observe significant gene flow from the Amhara and Oromo into the Nama, a signal which is likely a proxy for the movement of
-eastern African caprid and cattle pastoralists \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
-contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE plot (Figure~\ref{fig:diversity}),
-the ancestry is likely grouped into the Khoe-San component which has drifted appreciably from its ancestral eastern African source.
-Colonial period admixture from Europeans into the Nama
-was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
+($m\approx2\times10^{-4}$, the constant proportion of immigrant lineages per
+generation over 60ka). We further find that Back-to-Africa gene flow at the
+beginning of the Holocene primarily affected the ancestors of the Ethiopian
+agricultural populations \citep{Molinaro2019-rf}, comprising over half of their
+genetic ancestry, estimated to be 64--65\%.  We observe significant gene flow
+from the Amhara and Oromo into the Nama, a signal which is likely a proxy for
+the movement of eastern African caprid and cattle pastoralists
+\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
+contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE
+plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped into the
+Khoe-San component which has drifted appreciably from its ancestral eastern
+African source.  Colonial period admixture from Europeans into the Nama was
+estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
 (Figure~\ref{fig:diversity}).

### 2022-12-19 supplement.tex
add Godambe table, unfinished

@@
 \usepackage{graphicx}
+\usepackage{lineno}
+\linenumbers
 
@@
 
+\aprcomment{TODO: add (here?) our reasoning and validation of the choice of
+60/50ka for fixed divergence times.}
+
 %\sgcomment{I feel like we should improve the discussion of relationship
@@
 
-For the best fit parameters, we computed confidence intervals using the 
-same set of bootstrapped samples and Godambe Information
-approach \citep{Coffman2016-yq}, which corrects composite likelihood estimates of confidence intervals
-to account for nonindependence in the data, including linkage between loci and
-nonindependence of recombination bins.
+For the best fit parameters, we computed confidence intervals using the same
+set of bootstrapped samples and Godambe Information approach
+\citep{Coffman2016-yq}, which corrects composite likelihood estimates of
+confidence intervals to account for nonindependence in the data, including
+linkage between loci and nonindependence of recombination bins.
+
+\aprcomment{To ensure that our confidence intervals computed using the approach
+    from \citet{Coffman2016-yq} do not underestimate true CIs, we also
+    performed block bootstrap resampling of our data, 200 times, and refit each
+    model to the bootstrapped datasets. From these, we evaluated the
+distribution of inferred parameters, which should be similar to or narrower
+than the CIs estimated from the GIM approach. See
+Section~\ref{sec:bootstrap-CIs}.}
 
@@
 when using the diversity and LD statistics that we have in our analyses.
+
+\subsection{Evaluation of confidence intervals}
+\label{sec:bootstrap-CIs}
+
+\aprcomment{Here, point to tables for CIs, or SEs from Godambe vs reinferred
+models from bootstrapped data. See Table~\ref{tab:supp-confidence-intervals}}
+
+\subsection{Allowing for additional migrations}
+
+\aprcomment{Here, addressing major comment 3 from reviewer 4, migration between
+Nama and Mene and Nama and GBR.}
+
+\subsection{Replacing the reference population for statistics normalization}
+
+\aprcomment{Minor comment 3 from reviewer 4.}
 
@@
 
+\begin{table}[ht]
+    \caption{
+        \label{tab:supp-confidence-intervals}
+        \textbf{GIM-computed confidence intervals are not underestimated.}
+        CIs computed with the Godambe information matrix (GIM) are larger than
+        those estimated by creating block bootstrap-replicate datasets and
+        refitting each model (200 times) to these datasets. Standard deviations
+        for each inferred parameter were then computed from these 200 model fits.
+        Parameters are as labeled and described in the above tables.
+        Blank standard errors indicate that parameter was not relevant
+        to the given model.
+        \aprcomment{Fill in after perturbing starting conditions}
+    }
+    \centering
+    \begin{tabular}[t]{l|cc|cc|cc|cc}
+        \toprule
+        & \multicolumn{2}{c|}{Single origin} & \multicolumn{2}{c|}{Cont. mig.}
+        & \multicolumn{2}{c|}{Merger w/o mig.} & \multicolumn{2}{c}{Merger w/mig.} \\
+        Parameter & GIM & BS & GIM & BS & GIM & BS & GIM & BS \\
+        \midrule
+        $N_e$ & 403 & 127 & 1777 & 951 & 326 &  & 1369 &  \\
+        $N_{MH}$ & 529 & 158 & & & & & & \\
+        $N_{stem1}$ & & & 1612 & 406 & 76 &  & 838 &  \\
+        $N_{stem2}$ & & & 2488 & 1001 & 1149 &  & 6668 &  \\
+        $N_{Nama_0}$ & 370 & 79 & 2989 & 512 & 384 &  & 1514 &  \\
+        $N_{Nama_F}$ & 9 & 2 & 54 & 4 & 7 &  & 31 &  \\
+        $N_{MSL_0}$ & 769 & 190 & 2479 & 285 & 322 &  & 1165 &  \\
+        $N_{MSL_F}$ & 606 & 185 & 6628 & 1250 & 987 &  & 4332 &  \\
+        $N_{EA}$ & 273 & 49 & 1841 & 138 & 246 &  & 1628 &  \\
+        $N_{Gumuz_F}$ & 131 & 56 & 915 & 80 & 102 &  & 337 &  \\
+        $N_{EP}$ & 491 & 106 & 3246 & 168 & 408 &  & 1684 &  \\
+        $N_{GBR_0}$ & 33 & 7 & 231 & 17 & 29 &  & 122 &  \\
+        $N_{GBR_F}$ & 507 & 117 & 2889 & 295 & 334 &  & 1678 &  \\
+        $N_{Neand}$ & 105 & 88 & 591 & 97 & 91 &  & 235 &  \\
+        $T_{stems}$ & & & 391E3 & 67E3 & 27E3 &  & 426E3 &  \\
+        $T_{stem1}$ & & & & & 20E3 &  & 166E3 &  \\
+        $m_{Nama-MSL}$ & 1.6E-6 & 1.2E-6 & 3.7E-6 & 2.7E-6 & 1.1E-6 &  & 4.0E-6 &  \\
+        $m_{Nama-EA}$ & 2.0E-6 & 1.3E-6 & 1.0E-5 & 1.8E-6 & 1.3E-6 &  & 9.1E-6 &  \\
+        $m_{MSL-EA}$ & 7.6E-6 & 1.1E-6 & 5.3E-5 & 3.0E-6 & 5.5E-6 &  & 2.6E-5 &  \\
+        $m_{EA-GBR}$ & 2.4E-6 & 1.1E-6 & 1.0E-5 & 1.9E-6 & 1.5E-6 &  & 5.5E-6 &  \\
+        $m_{EA-EA}$ & 1.3E-5 & 7.5E-6 & 8.4E-5 & 1.4E-5 & 1.3E-5 &  & 5.4E-5 &  \\
+        $f_{GBR \rightarrow EP}$ & 0.0039 & 0.0036 & 0.0037 & 0.0071 & 0.0037 &  & 0.0075 &  \\
+        $f_{EP \rightarrow Nama}$ & 0.0039 & 0.0040 & 0.0043 & 0.0072 & 0.0042 &  & 0.0049 &  \\
+        $f_{GBR \rightarrow Nama}$ & 0.0019 & 0.0020 & 0.0021 & 0.0030 & 0.0021 & & 0.0031 &  \\
+        $m_{stems}$ & & & 1.1E-5 & 7.7E-6 & & & 8.7E-5 &  \\
+        $m_{stem2-Nama}$ & & & 1.6E-5 & 1.1E-5 & & & & \\
+        $m_{stem2-MSL}$ & & & 4.2E-5 & 1.7E-5 & & & & \\
+        $m_{stem2-EA}$ & & & 9.0E-6 & 5.4E-6 & & & & \\
+        $T_{Nama}$ & 2525 & 1188 & 18E3 & 6400 & 8253 &  & 28E3 &  \\
+        $f_{stem 2 \rightarrow Nama}$ & & & & & 0.0086 &  & 0.067 &  \\
+        $T_{EA}$ & & & & & 3648 &  & 8865 &  \\
+        $f_{stem 2 \rightarrow EA}$ & & & & & 0.0074 &  & 0.059 &  \\
+        $T_{MSL}$ & & & & & 570 &  & 641 &  \\
+        $f_{stem 2 \rightarrow MSL}$ & & & & & 0.0036 &  & 0.0085 &  \\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
 \newcommand{\specialcell}[2][c]{%

### 2022-12-19 supplement.tex
finish filling in table S7

--- 2022-09-15 supplement.tex
@@
 
+\begin{table}[ht]
+    \caption{
+        \label{tab:supp-confidence-intervals}
+        \textbf{GIM-computed confidence intervals are not underestimated.}
+        CIs computed with the Godambe information matrix (GIM) are largely
+        similar (i.e., within an order of magnitude) to those estimated by
+        creating 200 block bootstrap-replicate datasets (BS) and refitting each
+        modelto these datasets (see Section~\ref{sec:bootstrap-CIs}). Standard
+        deviations for each inferred parameter were computed from these 200
+        model fits. Parameters are as labeled and described in the above
+        tables. Blank standard errors indicate that parameter was not relevant
+        to the given model.
+    }
+    \centering
+    \begin{tabular}[t]{l|cc|cc|cc|cc}
+        \toprule
+        & \multicolumn{2}{c|}{Single origin} & \multicolumn{2}{c|}{Cont. mig.}
+        & \multicolumn{2}{c|}{Merger w/o mig.} & \multicolumn{2}{c}{Merger w/mig.} \\
+        Parameter & GIM & BS & GIM & BS & GIM & BS & GIM & BS \\
+        \midrule
+        $N_e$ & 403 & 127 & 1777 & 951 & 326 & 406 & 1369 & 1408 \\
+        $N_{MH}$ & 529 & 158 & & & & & & \\
+        $N_{stem1}$ & & & 1612 & 406 & 76 & 122 & 838 & 164 \\
+        $N_{stem2}$ & & & 2488 & 1001 & 1149 & 2344 & 6668 & 1154 \\
+        $N_{Nama_0}$ & 370 & 79 & 2989 & 512 & 384 & 496 & 1514 & 538 \\
+        $N_{Nama_F}$ & 9 & 2 & 54 & 4 & 7 & 4 & 31 & 5 \\
+        $N_{MSL_0}$ & 769 & 190 & 2479 & 285 & 322 & 384 & 1165 & 421 \\
+        $N_{MSL_F}$ & 606 & 185 & 6628 & 1250 & 987 & 1031 & 4332 & 1404 \\
+        $N_{EA}$ & 273 & 49 & 1841 & 138 & 246 & 154 & 1628 & 252 \\
+        $N_{Gumuz_F}$ & 131 & 56 & 915 & 80 & 102 & 72 & 337 & 83 \\
+        $N_{EP}$ & 491 & 106 & 3246 & 168 & 408 & 177 & 1684 & 185 \\
+        $N_{GBR_0}$ & 33 & 7 & 231 & 17 & 29 & 17 & 122 & 21 \\
+        $N_{GBR_F}$ & 507 & 117 & 2889 & 295 & 334 & 203 & 1678 & 218 \\
+        $N_{Neand}$ & 105 & 88 & 591 & 97 & 91 & 94 & 235 & 158 \\
+        $T_{stems}$ & & & 391E3 & 67E3 & 27E3 & 7.6E3 & 426E3 & 284E3 \\
+        $T_{stem1}$ & & & & & 20E3 & 45E3 & 166E3 & 58E3 \\
+        $m_{Nama-MSL}$ & 1.6E-6 & 1.2E-6 & 3.7E-6 & 2.7E-6 & 1.1E-6 & 1.3E-6 & 4.0E-6 & 2.6E-6 \\
+        $m_{Nama-EA}$ & 2.0E-6 & 1.3E-6 & 1.0E-5 & 1.8E-6 & 1.3E-6 & 1.8E-6 & 9.1E-6 & 2.8E-6 \\
+        $m_{MSL-EA}$ & 7.6E-6 & 1.1E-6 & 5.3E-5 & 3.0E-6 & 5.5E-6 & 3.3E-6 & 2.6E-5 & 4.4E-6 \\
+        $m_{EA-GBR}$ & 2.4E-6 & 1.1E-6 & 1.0E-5 & 1.9E-6 & 1.5E-6 & 2.2E-6 & 5.5E-6 & 2.6E-6 \\
+        $m_{EA-EA}$ & 1.3E-5 & 7.5E-6 & 8.4E-5 & 1.4E-5 & 1.3E-5 & 1.4E-5 & 5.4E-5 & 1.9E-5 \\
+        $f_{GBR \rightarrow EP}$ & 0.0039 & 0.0036 & 0.0037 & 0.0071 & 0.0037 & 0.0066 & 0.0075 & 0.0084 \\
+        $f_{EP \rightarrow Nama}$ & 0.0039 & 0.0040 & 0.0043 & 0.0072 & 0.0042 & 0.0068 & 0.0049 & 0.0085 \\
+        $f_{GBR \rightarrow Nama}$ & 0.0019 & 0.0020 & 0.0021 & 0.0030 & 0.0021 & 0.0030 & 0.0031 & 0.0035 \\
+        $m_{stems}$ & & & 1.1E-5 & 7.7E-6 & & & 8.7E-5 & 2.5E-5 \\
+        $m_{stem2-Nama}$ & & & 1.6E-5 & 1.1E-5 & & & & \\
+        $m_{stem2-MSL}$ & & & 4.2E-5 & 1.7E-5 & & & & \\
+        $m_{stem2-EA}$ & & & 9.0E-6 & 5.4E-6 & & & & \\
+        $T_{Nama}$ & 2525 & 1188 & 18E3 & 6400 & 8253 & 4763 & 28E3 & 6283 \\
+        $f_{stem 2 \rightarrow Nama}$ & & & & & 0.0086 & 0.017 & 0.067 & 0.018 \\
+        $T_{EA}$ & & & & & 3648 & 2047 & 8865 & 4114 \\
+        $f_{stem 2 \rightarrow EA}$ & & & & & 0.0074 & 0.015 & 0.059 & 0.021 \\
+        $T_{MSL}$ & & & & & 570 & 1373 & 641 & 1613 \\
+        $f_{stem 2 \rightarrow MSL}$ & & & & & 0.0036 & 0.0066 & 0.0085 & 0.0076 \\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
 \newcommand{\specialcell}[2][c]{%

### 2022-12-21 supplement.tex
added text about initialization

@@
 
-For the early history, we tested model parameterizations that cover many of the
+As is commonly done in demographic inference, 
+all models we considered are initialized from an ancestral panmictic population at 
+mutation-recombination-drift equilibrium. This allows for rapid initialization of the 
+summary statistics to reasonable values.
+The size $N_e$ of this ancestral population is a model-specific adjustable parameter 
+that also serves as a reference size to define subsequent population size changes.
+
+For the early history following this initialization, 
+we tested model parameterizations that cover many of the
 proposed scenarios of population structure, size changes, and/or archaic

### 2022-12-21 supplement.tex
composite likelihood and Gaussian

@@
 %oversimplified demographic models). 
-
-\subsection{Optimization using \moments}
-\label{sec:optimization}
-
-\texttt{moments-LD} uses a composite likelihood approach to simultaneously fit relative
-pairwise diversity and LD statistics over a range of recombination distances.
-Likelihoods were computed independently for pairwise diversity and each
-recombination distance bin using a multivariate Gaussian likelihood function as
-described in Section \ref{sec:computing-stats}. These were multiplied across
-bins and the single set of heterozygosity statistics following the approach
-detailed in \citet{Ragsdale2019-nt}. For each model tested, we ran multiple
-rounds of optimization, alternating between the \emph{optimize\_log\_fmin},
-\emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
-parameter space and hone in on the best fit parameters. Initial guesses for
-parameters were chosen from demographically plausible starting points and then
-perturbed to explore space, using gradient descent (on the log of the
-parameters). The best fits from these initial rounds of optimization were then
-chosen as the starting points for optimization using the Powell and/or the
-L-BFGS-B methods (as implemented in the SciPy optimization package
-\citep{Virtanen2020-kr}). This process was repeated with alternating
-optimization methods until the best fit parameter set converged consistently.
-
-\subsection{Uncertainty and confidence intervals}
-
-To compute likelihoods, we needed an estimate of the joint
+\subsection{Likelihoods, uncertainty, and confidence intervals}
+
+\texttt{moments-LD} uses a composite multivariate Gaussian likelihood approach 
+to simultaneously fit relative pairwise diversity and LD statistics over a range of 
+recombination distances.
+This composite likelihood and Gaussian assumption are described in more detail 
+ and validated in \citet{Ragsdale2019-nt}, but we outline them here.  
+To compute likelihoods, we need an estimate of the joint
 distribution of summary statistics. 
-We estimated uncertainty due to the finite amount of genetic material
+We estimate uncertainty due to the finite amount of genetic material
 used in inference using bootstrap over 500 segments along the genome with
 roughly equal lengths of retained sequences within each segment. First, for
-each distance bin, we used these bootstrap samples to obtain a
+each distance bin, we use these bootstrap samples to obtain a
 variance-covariances matrix across all statistics. This variance-covariance
-matrix was used to obtain a model likelihood for each recombination distance
+matrix is used to obtain a model likelihood for each recombination distance
 bin and single-locus nucleotide diversity, as a multivariate Gaussian
-likelihood. The full model likelihood was taken as the product of likelihoods
-over each bin. In other words, we optimized a composite likelihood where
+likelihood. 
+The full model likelihood is taken as the product of likelihoods
+over each bin. In other words, we followed the approach of  \citet{Ragsdale2019-nt} 
+in optimizing a composite likelihood where
 observations in different bins were taken to be independent.
@@
 merger-with-stem-migration model.
+
+\subsection{Optimization using \moments}
+\label{sec:optimization}
+
+For each model tested, we ran multiple rounds of likelihood optimization, 
+alternating between the \emph{optimize\_log\_fmin},
+\emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
+parameter space and hone in on the best fit parameters. Initial guesses for
+parameters were chosen from demographically plausible starting points and then
+perturbed to explore space, using gradient descent (on the log of the
+parameters). The best fits from these initial rounds of optimization were then
+chosen as the starting points for optimization using the Powell and/or the
+L-BFGS-B methods (as implemented in the SciPy optimization package
+\citep{Virtanen2020-kr}). This process was repeated with alternating
+optimization methods until the best fit parameter set converged consistently.
 

### 2022-12-23 paper.tex
Updated Discussion

@@
 We quantify migration rates among populations after their divergence
-$\approx120ka$. Prior to the Bantu expansion, migration between the ancestors
+$\approx120ka$. Prior to agripastoralist expansion 5ka, migration between the ancestors
 of the Nama and other groups is an order of magnitude weaker than that observed
@@
 population structure, we sought to capture model uncertainty by exploring
-multiple parameterizations of early history. The best-fit models presented
+multiple parameterizations of early history. The best fit models presented
 here include reticulation and migration between early human populations rather
-than archaic hominin admixture from long-isolated branches. We cannot rule out
-that more complex models involving additional stems, continuous structure, or
+than archaic hominin admixture from long-isolated branches (Figure~\ref{fig:proposed-models}C). 
+Elements of a recent single-origin or African multi-regionalism (Figure~\ref{fig:proposed-models}A,D)
+feature in our best fit models, as indicated in the recent time of contemporary population divergence
+or the gene flow between disparate stems, respectively.
+We caution that we cannot rule out that more complex models involving additional stems, continuous structure, or
 hybrid models including both weak structure and archaic hominin admixture may
@@
 interpret the weakly structured stem as consistent with a population
-coalescence and fragmentation model \citep{Scerri2019-xg}. Models including
-additional diversity within Africa, and early ancient DNA samples from Africa,
-could further distinguish the archaic hominin admixture model from the
-weakly-structured-stem model.
+coalescence and fragmentation model \citep{Scerri2019-xg}.
+
+Additional African populations such as those 
+from Central Africa, or pre-Holocene ancient DNA samples, could further test our .
 
@@
 
-By contrast, our inferred models paint a more consistent picture of the Middle to Late Pleistocene as a critical period of change, assuming that estimates from
-the recombination clock accurately relate to geological chronologies
+Our inferred models paint a more consistent picture of the Middle to Late Pleistocene as a 
+critical period of change, assuming that estimates from the recombination clock accurately relate to geological chronologies
 (SI Section~\ref{sec:supp-mutation}). 

### 2022-12-23 paper.tex
Update paper.tex

@@
 interpret the weakly structured stem as consistent with a population
-coalescence and fragmentation model \citep{Scerri2019-xg}.
-
-Additional African populations such as those 
-from Central Africa, or pre-Holocene ancient DNA samples, could further test our .
+coalescence and fragmentation model \citep{Scerri2019-xg}. Additional African populations such as those 
+from Central Africa, or pre-Holocene ancient DNA samples, could further test our proposed models.
+Given that our samples cover major ancestry components across Africa, we predict that any new admixture signal is likely to be localized.
 
@@
 
-Our inferred models paint a more consistent picture of the Middle to Late Pleistocene as a 
+Our inferred models paint a consistent picture of the Middle to Late Pleistocene as a 
 critical period of change, assuming that estimates from the recombination clock accurately relate to geological chronologies
@@
 results in the formation of the
-ancestors of East/West Africans including the ancestors of ``Out of Africans''. The rapid
+ancestors of Eastern/Western Africans including the ancestors of ``Out of Africans''. The rapid
 rise in sea levels and increased precipitation during MIS 5e interglacial, following a
@@
 Following these merger events, the stems subsequently fracture into
-subpopulations which then appear to persist over the past $\sim$120ka. These
+subpopulations which then persist over the past $\sim$120ka. These
 subpopulations can be linked to contemporary groups despite subsequent gene flow

### 2022-12-23 supplement.tex
work on fixing date at 60ka

--- 2022-12-19 supplement.tex
@@
 Since a detailed model for the Out of Africa and Back-to-Africa was not our
-main focus, we chose to keep this simplified model, but fix the split between
-east African-European split at $50$ka. This allowed us to include Neanderthal
-admixture into the OOA population at the previously-estimated $45$ka. Fixing
-this parameter pushed back the estimated value for the split between the
-ancestors of East and West African populations at around $60$ka, a value that
-we took as fixed in most subsequent analyses. 
-We discuss the sensitivity of our model to these assumptions in Section
-\ref{sec:change-dates}.
-
-\aprcomment{TODO: add (here?) our reasoning and validation of the choice of
-60/50ka for fixed divergence times.}
+main focus, we chose to keep this simplified model, but fix the east
+African-Eurasian split at $50$ka \aprcomment{references of archaeological
+evidence of humans in Europe/near East by around 45ka}. This also allowed us to
+include Neanderthal admixture into the OOA population at the
+previously-estimated $45$ka \citep{}. Fixing these parameters (i.e., east
+Africa-Eurasian split at 50ka, Neanderthal admixture at 45ka) resulted in an
+estimated value for the split between the ancestors of East and West African
+populations at around $60$ka.  \aprcomment{here: agrees with previous estimates
+    of West African-European splits from LD \citep{Ragsdale2019-nt}, which used
+    different populations from the Thousand Genomes Project than here. It is
+    also in close alignment with inferred divergence times from an SFS-based
+    demographic inference ($\sim65$ka), which used synonymous variation from a
+subset of populations used in this study -- Mende (MSL), British (GBR), and Han
+Chinese (CHB) (Section or Supp fig?).} We took this value (60ka) as fixed in
+most subsequent analyses. We discuss the sensitivity of our model to these
+assumptions in Section~\ref{sec:change-dates}.
 

### 2022-12-23 paper.tex
Update paper.tex

@@
 to the expansion (Figure~\ref{fig:proposed-models}A or Figure~\ref{fig:proposed-models}D). As expected
-\citep{Ragsdale2019-nt}, this first model preferred a recent expansion model but was also a poor fit to the data
+\citep{Ragsdale2019-nt}, this first model was a poor fit to the data
 qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
@@
 major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
-eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
+eastern (Stem 1E) and western/central Africa (Stem 2). Geographic association was informed by 
+the present population location with the greatest ancestry contribution from each stem. For example,
+Stem 1S contributes 71% to the ancestral formation of the Khoe-San. The extent of the isolation >400ka
+between Stem 1S and Stem 1E
+suggests that these two stems were not proximate to each other. While the length of
 isolation among the stems is variable across model fits, models with a period of 
@@
 Glacial Period, indicative that this gene flow likely occurred in
-western/central Africa (Table~\ref{tab:migration-rates}).
+western/central Africa (Table~\ref{tab:migration-rates}). Such an interpretation is reinforced
+by the differential migration rates between regions, i.e. the gene flow from Stem 2 to western Africans is 5x
+that of the rate to eastern Africans during this period.
 

### 2022-12-23 paper.tex
Geographic affiliation in Discussion
Update paper.tex

@@
 the present population location with the greatest ancestry contribution from each stem. For example,
-Stem 1S contributes 71% to the ancestral formation of the Khoe-San. The extent of the isolation >400ka
-between Stem 1S and Stem 1E
-suggests that these two stems were not proximate to each other. While the length of
+Stem 1S contributes 71\% to the ancestral formation of the Khoe-San. The extent of the isolation >400ka
+between Stem 1S, Stem 1E and Stem 2 suggests that these stems were not proximate to each other. While the length of
 isolation among the stems is variable across model fits, models with a period of 

### 2022-12-23 paper.tex
Revisions to address Reviewer 5

@@
 River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
-sapiens} anatomical features were intermittently present across the continent
+sapiens} anatomical features appeared across the continent
 300-100ka. Archaeological sites from the Middle Stone Age, of which some have
@@
 
-Multiple studies have shown a correspondence between phenotypic
-differentiation, usually assessed with measurements of the cranium, and genetic
-differentiation among human populations and between humans and Neanderthals
-\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}
-(see also SI Section~\ref{sec:morpho_diff}).
-This correspondence allows predictions of our model to be related to the fossil record.
-The fossil record of Africa is sparse during the time period of the stems, but of the
-available fossils, some are very similar in morphology to contemporary humans
-(e.g., from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}),
-others are similar in some morphological features but not others
-(e.g., from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}),
-and others are very different in morphology
-(e.g., from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}).
-If, as our model predicts, the
-genetic differences between the stems were comparable to those among
-contemporary human populations, the most morphologically divergent fossils are unlikely to
-represent branches that contributed appreciably to contemporary human
-ancestries.
+Multiple studies have shown a correspondence between phenotypic differentiation, 
+usually assessed with measurements of the cranium, and genetic differentiation 
+among human populations and between humans and Neanderthals 
+\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI 
+Section~\ref{sec:morpho_diff}). This correspondence potentially allows predictions 
+of our model to be related to the fossil record. The fossil record of Africa is 
+sparse during the time period of the stems, but of the fossils that date to this 
+time period, some are fairly similar in morphology to contemporary humans (e.g., 
+Omo 1 from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are 
+similar in some morphological features to contemporary humans (e.g., Irhoud 1 
+from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}), and others are 
+different enough in morphology to have been assigned to species other than 
+\emph{Homo sapiens} (e.g., DH1 from Dinaledi, South Africa 
+\citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
+differences between the stems were comparable to those among contemporary human 
+populations, the most morphologically divergent fossils are unlikely to represent 
+branches that contributed appreciably to contemporary human ancestries.
 

### 2022-12-23 supplement.tex
Update section on fixed OOA time

@@
 
+\subsubsection{Fixing the East African and OOA divergence times}
+
 In our preliminary models, the unconstrained inferred split times between East
-African populations and Europeans was inferred to have occurred by around
-$40$ka, which is more recent than we may have expected given the
-well-documented presence of modern humans outside of Africa at that time
-\citep[e.g.]{Hublin2020-nk,Hajdinjak2021-jo}, as well as previous
-genetics-based dating of splits between African and Eurasian populations
-\citep[e.g.]{Jouganous2017-pq,Kamm2020-vn}. The unconstrained inferred split
-between East and West African populations occurred about $10$ka earlier. These
-recent split dates seem to be at odds with archaeological evidence for earlier
-Out-of-Africa (OOA) expansion, and with the estimated date for Neanderthal
-admixture into OOA populations. The parameter uncertainty due to finite genomes
-does not account for this difference: in the high-likelihood models, the $95\%$
-confidence intervals for the East African-European split was roughly $38 \pm
-5$ka, and for the East/West African split was $45 \pm 10$ka, with the
-uncertainties reflecting 2 standard deviations according to bootstrap of
-genomic regions.  Thus we expect these discrepancies to result from model
-misspecification in the details of the divergence between the ancestors of OOA
-and east African populations. Our models suppose a single split prior to the
-OOA event followed by constant symmetric migrations between eastern and OOA
-populations, which is certainly an oversimplification.
-
-Since a detailed model for the Out of Africa and Back-to-Africa was not our
-main focus, we chose to keep this simplified model, but fix the east
-African-Eurasian split at $50$ka \aprcomment{references of archaeological
-evidence of humans in Europe/near East by around 45ka}. This also allowed us to
-include Neanderthal admixture into the OOA population at the
-previously-estimated $45$ka \citep{}. Fixing these parameters (i.e., east
-Africa-Eurasian split at 50ka, Neanderthal admixture at 45ka) resulted in an
-estimated value for the split between the ancestors of East and West African
-populations at around $60$ka.  \aprcomment{here: agrees with previous estimates
-    of West African-European splits from LD \citep{Ragsdale2019-nt}, which used
-    different populations from the Thousand Genomes Project than here. It is
-    also in close alignment with inferred divergence times from an SFS-based
-    demographic inference ($\sim65$ka), which used synonymous variation from a
-subset of populations used in this study -- Mende (MSL), British (GBR), and Han
-Chinese (CHB) (Section or Supp fig?).} We took this value (60ka) as fixed in
-most subsequent analyses. We discuss the sensitivity of our model to these
+African populations and Europeans was inferred to have occurred around $40$ka,
+and the inferred split between East and West African populations occurred about
+$10$ka earlier. These recent split dates are at odds with archaeological
+evidence for an earlier Out-of-Africa (OOA) expansion
+\citep[e.g.]{Hublin2020-nk,Hajdinjak2021-jo}, with the estimated date for
+Neanderthal admixture into Eurasian populations \citep[e.g.][]{Reilly2022-ym},
+and with previous genetics-based dating of splits between African and Eurasian
+populations \citep[e.g.]{Gutenkunst2009-ff,Jouganous2017-pq,Kamm2020-vn}.  The
+parameter uncertainty of our inferred divergence times (which accounts for the
+non-independence of sites) does not account for these differences: in the
+high-likelihood models, the $95\%$ confidence intervals for the East
+African-European split was roughly $38 \pm 5$ka, and for the East/West African
+split was $45 \pm 10$ka, with the uncertainties reflecting 2 standard
+deviations according to bootstrap of genomic regions. We therefore expect these
+discrepancies to result from model misspecification in the details of the
+divergence between the ancestors of OOA and East African populations.
+
+Our models suppose a single split leading to the OOA event followed by constant
+symmetric migrations between eastern African and OOA populations, which is an
+oversimplification. Previous studies that have focused on reconstructing the
+OOA divergence time and subsequent migration rates typically included more
+Eurasian populations and multiple epochs that allow for differing continuous
+migration rates and population sizes. Since a detailed model for the OOA (and
+Back-to-Africa) event was not our main focus, we chose to keep this simplified
+model, but fix the east African-Eurasian split at $50$ka \citep{Hublin2020-nk}.
+This also allowed us to include Neanderthal admixture into the OOA population
+at $45$ka.
+
+Fixing these parameters (i.e., East Africa-Eurasian split at 50ka, Neanderthal
+admixture at 45ka) resulted in an estimated value for the split between the
+ancestors of East and West African populations at around $60$ka. This timing
+agrees with previous LD-based estimates of West African-Eurasian divergence
+times \citep{Ragsdale2019-nt}, which used different populations than those used
+in this study. A date of $60$ka is also in close agreement with inferred
+divergence times from an SFS-based demographic inference (of $\sim65$ka), using
+synonymous variation from a subset of populations used in this study (Supp.
+Figure~\ref{fig:supp-OOA-SFS}) \aprcomment{TODO: add this figure}.
+We therefore took this value ($60$ka) as fixed
+in most subsequent analyses. We discuss the sensitivity of our model to these
 assumptions in Section~\ref{sec:change-dates}.
-
-%\sgcomment{I feel like we should improve the discussion of relationship
-%with previous estimates below, or remove.}
-%
-%In addition to archaeological evidence, previous genetic dating suggested 
-%earlier dates for a split between European and African populations. Most of 
-%these results focused on populations of recent West African ancestry. 
-%
-%Additionally, most previous
-%estimates of the time of the OOA expansion compared Eurasian and West African
-%populations. Our inferred time of the East/West African divergence aligns with
-%previous estimates of the OOA event at $\approx 60$ka (themselves inferred from
-%oversimplified demographic models).
 
@@
 
-\begin{figure}
-
+\begin{figure}[ht!]
+    \centering
     \includegraphics{figures/toy_models/toy_model_combo}
-
-    \caption{The impact of model specification on coalescence rate and inference.
-    We show five simplified demographic models that differ in demographic events, 
-    occurring between 200ka and 100ka, that mark the transition between a single 
-    ancestral population and two daughter populations. A) A clean split 100ka, 
-B) A clean split 200ka, C) population growth at 200ka followed by split 100ka, 
-D) split 200ka followed by migration until 100ka, and E) split 200ka followed by 
-merger 105ka and a new split 100ka. The left column shows model representation, 
-the middle column shows instantaneous coalescence rates for pairs of lineages from the
-same daughter population, the right column shows instantaneous coalescence rates for
-pairs of lineages coming from distinct populations.  
+    \caption{
+        \textbf{The impact of model specification on coalescence rate and
+        inference.} We show five simplified demographic models that differ in
+        demographic events, occurring between 200ka and 100ka, that mark the
+        transition between a single ancestral population and two daughter
+        populations. A: A clean split 100ka, B: A clean split 200ka, C:
+        population growth at 200ka followed by split 100ka, D: split 200ka
+        followed by migration until 100ka, and E: split 200ka followed by
+        merger 105ka and a new split 100ka. The left column shows model
+        representation, the middle column shows instantaneous coalescence rates
+        for pairs of lineages from the same daughter population, the right
+        column shows instantaneous coalescence rates for pairs of lineages
+        coming from distinct populations.  
+    }
     \label{fig:toymodels}
-    }
-
 \end{figure}
@@
 
-\subsection{Batch effects from merged data biases inference}\label{sec:merged-bias}
+\subsection{Batch effects from merged data biases inference}
+\label{sec:merged-bias}
 
@@
     \caption{
-        \textbf{Average error of statistics from low coverage and merged datasets.}
-        Low coverage data provides allelic diversity and linkage
-        disequilibrium measures consistent with those from high coverage data (left panel). On the other hand,
-        merging separately called datasets can lead to large discrepancies
-        in computed LD statistics (middle and right panels).
+        \textbf{Average error of statistics from low coverage and merged
+        datasets.} Low coverage data provides allelic diversity and linkage
+        disequilibrium measures consistent with those from high coverage data
+        (left panel). On the other hand, merging separately called datasets can
+        lead to large discrepancies in computed LD statistics (middle and right
+        panels).
     }

### 2022-12-23 paper.tex

@@
 River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
-sapiens} anatomical features appeared across the continent
+sapiens} anatomical features were found across the continent
 300-100ka. Archaeological sites from the Middle Stone Age, of which some have

### 2022-12-23 supplement.tex

@@
 
-\subsection{Optimization using \moments}
-\label{sec:optimization}
-
-\texttt{moments-LD} uses a composite likelihood approach to simultaneously fit relative
-pairwise diversity and LD statistics over a range of recombination distances.
-Likelihoods were computed independently for pairwise diversity and each
-recombination distance bin using a multivariate Gaussian likelihood function as
-described in Section \ref{sec:computing-stats}. These were multiplied across
-bins and the single set of heterozygosity statistics following the approach
-detailed in \citet{Ragsdale2019-nt}. For each model tested, we ran multiple
-rounds of optimization, alternating between the \emph{optimize\_log\_fmin},
-\emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
-parameter space and hone in on the best fit parameters. Initial guesses for
-parameters were chosen from demographically plausible starting points and then
-perturbed to explore space, using gradient descent (on the log of the
-parameters). The best fits from these initial rounds of optimization were then
-chosen as the starting points for optimization using the Powell and/or the
-L-BFGS-B methods (as implemented in the SciPy optimization package
-\citep{Virtanen2020-kr}). This process was repeated with alternating
-optimization methods until the best fit parameter set converged consistently.
-
-\subsection{Uncertainty and confidence intervals}
-
-To compute likelihoods, we needed an estimate of the joint
+\subsection{Likelihoods, uncertainty, and confidence intervals}
+
+\texttt{moments-LD} uses a composite multivariate Gaussian likelihood approach 
+to simultaneously fit relative pairwise diversity and LD statistics over a range of 
+recombination distances.
+This composite likelihood and Gaussian assumption are described in more detail 
+ and validated in \citet{Ragsdale2019-nt}, but we outline them here.  
+To compute likelihoods, we need an estimate of the joint
 distribution of summary statistics. 
-We estimated uncertainty due to the finite amount of genetic material
+We estimate uncertainty due to the finite amount of genetic material
 used in inference using bootstrap over 500 segments along the genome with
 roughly equal lengths of retained sequences within each segment. First, for
-each distance bin, we used these bootstrap samples to obtain a
+each distance bin, we use these bootstrap samples to obtain a
 variance-covariances matrix across all statistics. This variance-covariance
-matrix was used to obtain a model likelihood for each recombination distance
+matrix is used to obtain a model likelihood for each recombination distance
 bin and single-locus nucleotide diversity, as a multivariate Gaussian
-likelihood. The full model likelihood was taken as the product of likelihoods
-over each bin. In other words, we optimized a composite likelihood where
+likelihood. 
+The full model likelihood is taken as the product of likelihoods
+over each bin. In other words, we followed the approach of  \citet{Ragsdale2019-nt} 
+in optimizing a composite likelihood where
 observations in different bins were taken to be independent.

### 2022-12-23 paper.tex
OOA free vs fixed in Discussion

@@
 coalescence and fragmentation model \citep{Scerri2019-xg}. Additional African populations such as those 
-from Central Africa, or pre-Holocene ancient DNA samples, could further test our proposed models.
-Given that our samples cover major ancestry components across Africa, we predict that any new admixture signal is likely to be localized.
+from Central Africa, other Khoe-San groups or pre-Holocene ancient DNA samples, could further test our proposed models.
+Given that our samples cover major ancestry components across Africa, we predict that any new admixture signal is likely to be regionally localized.
 
@@
 by the differential migration rates between regions, i.e. the gene flow from Stem 2 to western Africans is 5x
-that of the rate to eastern Africans during this period.
+that of the rate to eastern Africans during this period. We performed a variety of validation tests
+to explore sensitivity of our assumptions, including relaxing fixed parameters (SI Section~\ref{sec:model_validation_strategy}).
+Most validation tests resulted in parameters similar to the models discussed above. However, one exception
+was the inferred Out of Africa and Eastern/Western divergence which were 10-15ka younger than our fixed parameters.
+Complexity of this event warrants additional Eurasian populations and greater detail; nonetheless, our model mis-specification
+of this event does not change other features of the merger model.
 

### 2022-12-23 paper.tex
archaic admixture more explicitly referenced

--- 2022-12-23 paper.tex
@@
 Neanderthal admixture with humans in Eurasia, several recent articles
-have shown that introducing an archaic hominin ghost population contributing to African
-populations in the period surrounding the Out-of-Africa migration event
+have shown that introducing an archaic hominin ghost population contributing to 
+African populations in the period surrounding the Out-of-Africa migration event
 substantially improves the description of genetic data relative to
-single-origin models
-\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}.
+single-origin models, mostly in Western Africa
+ \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}.
+%Plagnol: YRI
+
+%Hammer: Mandenka, Biaka, and San. Title:
+%"Genetic evidence for archaic admixture in Africa"
+%Abstract:"allow us to infer that contemporary African populations contain a small proportion of genetic material (≈2%) that introgressed ≈35 kya from an archaic population that split from the ancestors of anatomically modern humans ≈700 kya."
+
+%Hey: Strongest signal in Yoruba, Baka
+% Abstract: "consistent with human/archaic divergence."
+
+%Ragsdale: YRI
+%title: arcahic admixture from...
+
+%Lorente Galdos: Khoesan, Mbuti, W. AFR
+%title:"reveals archaic gene flow"
+
+%Durvasula: West African
+%title: "Recovering signals of ghost archaic introgression"
+%Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans."
+but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} and central Africa \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz}.
+populations.
 This has driven speculation about the geographic range of this ghost
@@
 Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
-archaic hominin admixture having also occurred in Africa. For this reason, previous 
-work has focused on archaic hominin admixture to explain patterns of polymorphism 
-inconsistent with a single-origin model. Here, we have shown that weakly-structured-stem 
-models better capture these patterns. Preferring weakly-structured-stem models over 
-archaic-admixture models potentially has ecological implications. With a weakly 
+archaic hominin admixture having also occurred in Africa. Previous 
+work that sought to explain patterns of polymorphism inconsistent with a single-origin model thus 
+focused on archaic hominin admixture as an alternative model. 
+The additional (or "ghost") branches required to fit the data
+are systematically referred to as "archaic" 
+\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}, 
+and assumed (or inferred) to be deeply diverged.
+We believe that this has strongly oriented the interpretation of these results (e.g., regarding selection \cite{Wall2019-ao} 
+or the fossil record \cite{Grun2020}).
+Here, we have shown that weakly-structured-stem 
+models better captures the apparently inconsistent patterns of polymorphisms. Preferring weakly-structured-stem models over 
+archaic-admixture models potentially has a range of implications.
+
+ First, with a weakly 
 structured stem, there is no need to posit that an archaic hominin population in Africa stayed 
@@
 
-There is evidence for both deleterious and adaptive archaic-hominin-derived alleles in
+Second, there is evidence for both deleterious and adaptive archaic-hominin-derived alleles in
 contemporary genomes in the form of a depletion of Neanderthal ancestry in regulatory
@@
 much higher than Neanderthal gene flow, and would have plausibly been fertile
-ground for dramatic selection for or against archaic-hominin-derived haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
+ground for dramatic selection for or against archaic-hominin-derived haplotypes\citep{Wall2019-ao}. 
+By contrast, adaptation under a weakly
 structured stem would have occurred continuously over much longer periods.
@@
 
-Multiple studies have shown a correspondence between phenotypic differentiation, 
+Third, multiple studies have shown a correspondence between phenotypic differentiation, 
 usually assessed with measurements of the cranium, and genetic differentiation 

### 2022-12-24 paper.tex
more lit searcg
Update paper.tex

@@
 single-origin models, mostly in Western Africa
- \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}.
+\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td},
 %Plagnol: YRI 
-
+%Here we present a new method for addressing whether archaic human groups contributed to the modern gene pool (called ancient admixture),
+%we find strong evidence for ancient admixture in both a European and a West African population
+%
 %Hammer: Mandenka, Biaka, and San. Title:
@@
 %Abstract:"allow us to infer that contemporary African populations contain a small proportion of genetic material (≈2%) that introgressed ≈35 kya from an archaic population that split from the ancestors of anatomically modern humans ≈700 kya."
-
+%
 %Hey: Strongest signal in Yoruba, Baka
 % Abstract: "consistent with human/archaic divergence."
-
+%
 %Ragsdale: YRI
-%title: arcahic admixture from...
-
+%abstract: We estimate that an unidentified, deeply diverged population admixed with modern humans within Africa both before and after the split of African and Eurasian populations, contributing 4 − 8% genetic ancestry to individuals in world-wide populations.
+%
 %Lorente Galdos: Khoesan, Mbuti, W. AFR
 %title:"reveals archaic gene flow"
-
+%
 %Durvasula: West African
@@
 %Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans."
-but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} and central Africa \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz}.
+but also in southern \citep{Hammer2011-bx,Lorente-Galdos2019-vz} and central African \citep{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz}.
 populations.
@@
 population, possible links to specific fossils, and the possibility of
-finding ancient DNA evidence (e.g.\citep{Hsieh2016-gk}). However, these prior
+finding ancient DNA evidence \citep[e.g.,][]{Hsieh2016-gk}. However, these prior
 articles share two weaknesses. First, they only contrast a single-origin model

### 2022-12-27 supplement.tex
refs for fixing dates

@@
 evidence for an earlier Out-of-Africa (OOA) expansion
-\citep[e.g.]{Hublin2020-nk,Hajdinjak2021-jo}, with the estimated date for
-Neanderthal admixture into Eurasian populations \citep[e.g.][]{Reilly2022-ym},
-and with previous genetics-based dating of splits between African and Eurasian
-populations \citep[e.g.]{Gutenkunst2009-ff,Jouganous2017-pq,Kamm2020-vn}.  The
-parameter uncertainty of our inferred divergence times (which accounts for the
+\citep[e.g.]{Fu2014-am,Hublin2020-nk,Hajdinjak2021-jo}, with the estimated date
+for Neanderthal admixture into Eurasian populations
+\citep[e.g.][]{Reilly2022-ym}, and with previous genetics-based dating of
+splits between African and Eurasian populations
+\citep[e.g.]{Gutenkunst2009-ff,Jouganous2017-pq,Kamm2020-vn}. The parameter
+uncertainty of our inferred divergence times (which accounts for the
 non-independence of sites) does not account for these differences: in the
@@
 Back-to-Africa) event was not our main focus, we chose to keep this simplified
-model, but fix the east African-Eurasian split at $50$ka \citep{Hublin2020-nk}.
-This also allowed us to include Neanderthal admixture into the OOA population
-at $45$ka.
+model, but fix the East African-Eurasian split at $50$ka
+\citep{Fewlass2020-se,Hublin2020-nk}. This also allowed us to include
+Neanderthal admixture into the OOA population at $45$ka.
 

### 2022-12-28 supplement.tex
edits to supplement

@@
 agrees with previous LD-based estimates of West African-Eurasian divergence
-times \citep{Ragsdale2019-nt}, which used different populations than those used
-in this study. A date of $60$ka is also in close agreement with inferred
-divergence times from an SFS-based demographic inference (of $\sim65$ka), using
-synonymous variation from a subset of populations used in this study (Supp.
-Figure~\ref{fig:supp-OOA-SFS}) \aprcomment{TODO: add this figure}.
-We therefore took this value ($60$ka) as fixed
-in most subsequent analyses. We discuss the sensitivity of our model to these
-assumptions in Section~\ref{sec:change-dates}.
+times, of $\sim61$ka ($50-64$ka, $95\%$ confidence interval)
+\citep{Ragsdale2019-nt}, which used a different set populations than those in
+this study. A date of $60$ka is also in close agreement with inferred
+divergence times of $\sim65$ka from both SFS-based demographic inference, using
+synonymous variation from a subset of populations used in this study (Mende,
+British, and Han Chinese), as well as that of LD-based inference using these
+same groups, of $53$ka ($43-63$ka, $95\%$ confidence interval). Taken
+altogether, we decided to fix this value ($60$ka) in most subsequent analyses.
+We discuss the sensitivity of our model to these assumptions in
+Section~\ref{sec:change-dates}.
 
@@
 
-All supplementary inference results are provided in Demes format
-\citep{Gower2022-yn} as a supplemental zipped archive. For each scenario, we
-refit the continuous migration and multiple merger models, with stem migration.
+All supplementary inference results discussed in this section are provided in
+Demes format \citep{Gower2022-yn} as a supplemental zipped archive. For each
+scenario, we refit the continuous migration and multiple merger models, with
+stem migration.
 
@@
 best-fit inferred E/W divergence time and the OOA expansion were more recent
-(35--40ka, 45--55ka, resp.). The remainder of the parameters were largely
+($45-55$ka and $35-40$ka, resp.). The remainder of the parameters were largely
 unaffected by these shifts to more recent divergence times, and our
@@
 
-\aprcomment{Here, addressing major comment 3 from reviewer 4, migration between
-Nama and Mene and Nama and GBR.}
+In preliminary analyses, we found that inferred migration rates between the
+Nama and non-East African groups were orders of magnitude smaller than
+migration rates between other populations included in our inferences.
+Therefore, to reduce the number of parameters needing to be fit we fixed those
+rates to zero in subsequent fits, and our reported best-fit models did not
+include migration between the Mende and Eurasians, nor between the Nama and
+Eurasian populations (aside from recent colonial-period admixture). To assess
+the impact of this choice, we re-inferred four of our reported models (single
+origin, continuous migration, and merger with and without stem migration),
+while allowing for additional parameters to include continuous, symmetric
+migration for the duration of the coexistence of the Nama and Mende of the Nama
+and British.
+
+Indeed, the inferred migration rates for both of these pairs of populations
+were small, with $2N_em \ll 1$ in all cases. In the single origin model, the
+Mende-British migration rate was 1.04E-6 ($2N_em = 0.02$) and the Nama-British
+rate was 2.67E-6 ($2N_em = 0.05$). In the continuous migration model, these
+rates were 2.69E-6 ($2N_em = 0.038$) and 3.38E-6 ($2N_em = 0.048$), resp. In
+the merger without stem migration model, these rates were 2.96E-6 ($2N_em =
+0.064$) and 5.46E-6 ($2N_em=0.119$), resp. And in the merger with stem
+migration model, these rates were 2.64E-6 ($2N_em = 0.055$) and 3.19E-6 ($2N_em
+= 0.066$), resp. In comparison with other migration rates, with $2N_em \gtrsim
+1$, these migrations have negligible impact on underlying statistics. The fits
+to other parameters were negligibly different from those reported with these
+migrations fixed to zero, and log-likelihoods were unimproved.
 
@@
 
-\aprcomment{Minor comment 3 from reviewer 4.}
+We used observed $\pi_2$ and pairwise diversity in the Mende to normalize two-
+and one-locus statistics (see Section~\ref{sec:supp-ld-stats}). To confirm that
+this choice of normalization did not impact inferred model parameters, we reran
+each of our reported models using data normalized by $\pi_2$ and pairwise
+diversity in the Gumuz. As expected, the best fit parameters from fits using
+the Gumuz-normalized data were very close to those reported in
+Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration},
+and they were all well within the reported confidence intervals. Our results
+are therefore unlikely to be sensitive to this choice of normalization.
+
+Reinferred demographic models from this and preceding sections are reported in
+Demes format \citep{Gower2022-yn} as a supplemental archive.
 
@@
 \end{table}
+
+%\begin{table}[ht]
+%    \caption{
+%        \label{tab:supp-OOA-SFS}
+%        \textbf{OOA parameters using LD and SFS statistics.}
+%        \aprcomment{To justify 60ka}
+%    }
+%    \centering
+%    \begin{tabular}[t]{rcc}
+%        \toprule
+%        Parameter & SFS from synonymous variation & LD from intergenic variation \\
+%        \midrule
+%        $N_e$ &  & \\
+%        $N_{AF}$ &  & \\
+%        $N_{B}$ &  & \\
+%        $N_{EU0}$ &  & \\
+%        $N_{EUF}$ &   & \\
+%        $N_{AS0}$ &  & \\
+%        $N_{ASF}$ &  & \\
+%        $m_{AF-B}$  & & \\
+%        $m_{AF-EU}$  & & \\
+%        $m_{AF-AS}$  & & \\
+%        $m_{EU-AS}$  & & \\
+%        $T_{EXP}$ &  & \\
+%        $T_{OOA}$ &  & \\
+%        $T_{EU-AS}$  & & \\
+%        \bottomrule
+%    \end{tabular}
+%\end{table}
 

### 2022-12-29 supplement.tex
section of bootstrap fits

@@
 \subsection{Likelihoods, uncertainty, and confidence intervals}
+\label{sec:likelihoods-uncertainty}
 
@@
 
-For the best fit parameters, we computed confidence intervals using the same
-set of bootstrapped samples and Godambe Information approach
+For the best fit parameters, we computed $95\%$ confidence intervals using the
+same set of bootstrapped samples and Godambe Information approach
 \citep{Coffman2016-yq}, which corrects composite likelihood estimates of
 confidence intervals to account for nonindependence in the data, including
-linkage between loci and nonindependence of recombination bins.
-
-\aprcomment{To ensure that our confidence intervals computed using the approach
-    from \citet{Coffman2016-yq} do not underestimate true CIs, we also
-    performed block bootstrap resampling of our data, 200 times, and refit each
-    model to the bootstrapped datasets. From these, we evaluated the
-distribution of inferred parameters, which should be similar to or narrower
-than the CIs estimated from the GIM approach. See
-Section~\ref{sec:bootstrap-CIs}.}
+linkage between loci and nonindependence of recombination bins.  To ensure that
+our confidence intervals computed using the approach from
+\citet{Coffman2016-yq} do not underestimate true CIs, we also performed block
+bootstrap resampling of our data, 200 times, and refit each model to the
+bootstrapped datasets. From these, we evaluated the distribution of inferred
+parameters, which should be similar to or narrower than the CIs estimated from
+the GIM approach (Section~\ref{sec:bootstrap-CIs}).
 
@@
 
-\aprcomment{Here, point to tables for CIs, or SEs from Godambe vs reinferred
-models from bootstrapped data. See Table~\ref{tab:supp-confidence-intervals}}
+We created 200 block-bootstrap replicate datasets and refit our four reported
+models to each replicate (Section~\ref{sec:likelihoods-uncertainty}). From
+these refit models, we estimated the empirical uncertainty by taking the
+standard deviation of the distribution of each refit parameter. We compared
+these to our estimated standard errors from the Godambe Information Matrix
+(GIM) approach to check that the confidence intervals from the GIM approach do
+not underestimate true uncertainty (Table~\ref{tab:supp-confidence-intervals}).
+
+The GIM-computed standard errors are generally close to those from the fits to
+the bootstrap data, and some are larger (roughly 2--5x) than those from the
+bootstrap fits. A possible explanation for this discrepancy is that some
+parameters of our models come close to the lower and upper bounds allowed for
+those parameters (e.g., small recent population size in the Nama, a bottleneck
+in one of the stems). In such cases, the distribution of those reinferred
+parameters and other correlated parameters in the model may be narrower than
+the true uncertainties in those parameters. Model constraints and parameter
+bounds in optimization may also lead to parameters fits that are away from
+their true optimal value. In this case, the GIM-estimated confidence intervals
+should be larger and contain what would be found from an unconstrained fit.
+
+To demonstrate this effect, we simulated data under a simple
+isolation-with-migration model, in which population sizes are allowed to vary,
+and one population has a severe bottleneck in the most recent 1000 years
+(Figure~\ref{fig:supp-uncerts-toy}). We simulated 200 1Mb regions with
+$u=r=10^{-8}$, sampling 10 diploid individuals from each present-day population
+and computed LD statistics. From these, we inferred the best fit parameters
+without imposing bounds on the parameters, and we also reinferred the model
+imposing a lower bound of 300 on population sizes (the bottleneck size). From
+our fits, we computed confidence intervals using the Fisher information matrix
+(FIM) and GIM approaches, in both the unbounded and bounded fits. We also
+reinferred parameters over 200 bootstrapped replicate datasets to compare
+empirical uncertainties to those estimated CIs.
+
+As expected the FIM CIs underestimate the true uncertainty and the GIM CIs
+closely match empirical uncertainty from the bootstrap fits. In the fit that
+imposed a lower bound on population sizes, some GIM-estimated standard errors
+were larger than those from the unbounded fits.  However, the fit parameters
+were slightly shifted from the unbounded fits, due to hitting the lower bound
+of the bottleneck size. In these cases, the estimated CIs were larger but
+contained the true CIs from the unbounded GIM and bootstrap fits. For parameters
+that were not strongly affected by hitting the bounds, the CIs matched closely
+(Figure~\ref{fig:supp-uncerts-toy}).
+
+\begin{figure}[htb]
+    \centering
+    \includegraphics{./figures/supp-uncerts-toy.pdf}
+    \caption{
+        \textbf{Evaluation of GIM confidence intervals using moments-LD.}
+        The GIM approach accurately estimates confidence intervals, accounting
+        for non-independence between regions of linked loci. GIM-estimated CIs
+        from the constrained fit are large enough to contain the true
+        uncertainty. Vertical dashed lines indicate simulated parameter values.
+    }
+    \label{fig:supp-uncerts-toy}
+\end{figure}
 
@@
 \end{table}
-
-%\begin{table}[ht]
-%    \caption{
-%        \label{tab:supp-OOA-SFS}
-%        \textbf{OOA parameters using LD and SFS statistics.}
-%        \aprcomment{To justify 60ka}
-%    }
-%    \centering
-%    \begin{tabular}[t]{rcc}
-%        \toprule
-%        Parameter & SFS from synonymous variation & LD from intergenic variation \\
-%        \midrule
-%        $N_e$ &  & \\
-%        $N_{AF}$ &  & \\
-%        $N_{B}$ &  & \\
-%        $N_{EU0}$ &  & \\
-%        $N_{EUF}$ &   & \\
-%        $N_{AS0}$ &  & \\
-%        $N_{ASF}$ &  & \\
-%        $m_{AF-B}$  & & \\
-%        $m_{AF-EU}$  & & \\
-%        $m_{AF-AS}$  & & \\
-%        $m_{EU-AS}$  & & \\
-%        $T_{EXP}$ &  & \\
-%        $T_{OOA}$ &  & \\
-%        $T_{EU-AS}$  & & \\
-%        \bottomrule
-%    \end{tabular}
-%\end{table}
 

### 2023-01-03 supplement.tex
minor style suggestion

@@
 those parameters (e.g., small recent population size in the Nama, a bottleneck
-in one of the stems). In such cases, the distribution of those reinferred
-parameters and other correlated parameters in the model may be narrower than
-the true uncertainties in those parameters. Model constraints and parameter
+in one of the stems). In such cases, the distribution of
+parameters under constraints, and that of other correlated parameters in the model
+may be narrower than
+the true (unconstrained) uncertainties in those parameters. Model constraints and parameter
 bounds in optimization may also lead to parameters fits that are away from

### 2023-01-03 supplement.tex
removed speculation
refit LLs

--- 2022-12-29 supplement.tex
@@
 
-\begin{figure}[ht!]
+\begin{figure}[htb!]
     \centering
@@
 when using the diversity and LD statistics that we have in our analyses.
+
+\subsection{Model choice from likelihood differences}
+\label{sec:refit-simulated-data}
+
+The set of models we tested with the full dataset have varying complexity,
+ranging from a single origin expansion to multiple stem populations connected
+by migration and gene flow. Thus, our models have varying numbers of parameters
+($21$ in the simplest model, and $30$ in the most heavily parameterized). Our
+analyses found that the more complex models (multiple stems with ongoing
+migration) provided the best fit to the data. To ensure that these more complex
+models are not simply fitting noise in our data, we simulated data under both
+simpler and more complex models, with same sample sizes and similar genome
+length to that used in this study. From these, we fit each of our models and
+checked that the correct model was chosen.
+
+While simulations of datasets of this size are rapid, computing LD statistics
+is computationally burdensome. We simulated 500 1Mb regions under our best fit
+single origin, continuous migration, and merger with stem migration models.
+Using these simulated regions, we constructed 500 bootstrap replicate datasets
+by sampling with replacement. Each of these 500 replicate sets were fit using
+the models reported in the main text, and we compared log-likelihoods across
+each fit (Figure~\ref{fig:refit-lls}). In each case, even in the simpler single
+origin simulations, the correct model was recovered.  We conclude that our
+inferences are not simply fitting statistical noise, and the observed
+differences in log-likelihoods indicate a better recapitulation of the true
+history.
+
+\begin{table}[ht]
+    \caption{
+        \label{tab:confusion-matrix}
+        \textbf{Something like a confusion matrix}
+        \aprcomment{fix/fill caption}
+    }
+    \centering
+    \begin{tabular}[t]{lcccc}
+        \toprule
+        Simulated model & Single origin & Cont. migration & Merger w/out stem mig. & Merger w/stem mig. \\
+        \midrule
+        Single origin & 494 & 3 & 1 & 2 \\
+        Cont. migration & 0 & 500 & 0 & 0 \\
+        Merger w/stem mig. & 0 & 0 & 1 & 499 \\
+        \bottomrule
+    \end{tabular}
+\end{table}
+
+The single origin model is nested within the more complex models (with
+migration from the additional stem population fixed to zero), those more
+complex models can always recover the log-likelihood found by the single origin
+model. We therefore explored the broader parameter space that included gene
+flow from the additional stem population. In these fits, the optimization
+converged either back to the nested single origin model, or it found a local
+optimum with lower log-likelihood than the simpler model (see first panel in
+Figure~\ref{fig:refit-lls}).
+
+\begin{figure}[htb]
+    \centering
+    \includegraphics{./figures/supp-refit-lls.pdf}
+    \caption{
+        \textbf{Fits of four models to simulated data.} In three scenarios, a
+        simple single origin model and our preferred continuous migration and
+        merger with stem migration models, we simulated 500 replicates of 500
+        Mb with the same populations and sample sizes from our reported best
+        fit models. We then fit the four models reported in the main text to
+        each of these simulated datasets and compared log-likelihoods. In each
+        case, the simulated model is correctly chosen, even under the single
+        origin scenario. We conclude our more complex models (with more
+        parameters) are not simply fitting statistical noise.
+    }
+    \label{fig:refit-lls}
+\end{figure}
 

### 2023-01-03 supplement.tex

@@
 those parameters (e.g., small recent population size in the Nama, a bottleneck
-in one of the stems). In such cases, the distribution of those reinferred
-parameters and other correlated parameters in the model may be narrower than
-the true uncertainties in those parameters. Model constraints and parameter
-bounds in optimization may also lead to parameters fits that are away from
-their true optimal value. In this case, the GIM-estimated confidence intervals
-should be larger and contain what would be found from an unconstrained fit.
+in one of the stems). In such cases, the distribution of
+parameters under constraints, and that of other correlated parameters in the model
+may be narrower than
+the true (unconstrained) uncertainties in those parameters. 
+%Model constraints and parameter
+%bounds in optimization may also lead to parameters fits that are away from
+%their true optimal value. In this case, the GIM-estimated confidence intervals
+%should be larger and contain what would be found from an unconstrained fit.
 

### 2023-01-06 supplement.tex
reinference from simulated data

@@
 the models reported in the main text, and we compared log-likelihoods across
-each fit (Figure~\ref{fig:refit-lls}). In each case, even in the simpler single
-origin simulations, the correct model was recovered.  We conclude that our
-inferences are not simply fitting statistical noise, and the observed
-differences in log-likelihoods indicate a better recapitulation of the true
-history.
+each fit (Figure~\ref{fig:refit-lls}). The correct model was recovered in the
+vast majority of cases based on log-likelihoods alone
+(Table~\ref{tab:confusion-matrix}). We conclude that our inferences are not
+simply fitting statistical noise, and the observed differences in
+log-likelihoods indicate a better recapitulation of the true history.
 
@@
         \label{tab:confusion-matrix}
-        \textbf{Something like a confusion matrix}
-        \aprcomment{fix/fill caption}
+        \textbf{Model choice from fits to simulated data.} We simulated three
+        models, based on the best fit parameters from fits to data. Using these
+        simulated datasets, we reinferred four models that we tested in our
+        analyses in the main text, including simple (single origin) and complex
+        models. Among 500 fits, we almost exclusively recover the simulated
+        model, even when fitting more complex (i.e., more heavily
+        parameterized) models to simpler simulated models (such as the single
+        origin model).
     }
@@
         \midrule
-        Single origin & 494 & 3 & 1 & 2 \\
+        Single origin & 500 & 0 & 0 & 0 \\
         Cont. migration & 0 & 500 & 0 & 0 \\
-        Merger w/stem mig. & 0 & 0 & 1 & 499 \\
+        Merger w/stem mig. & 0 & 0 & 10 & 490 \\
         \bottomrule
@@
 flow from the additional stem population. In these fits, the optimization
-converged either back to the nested single origin model, or it found a local
-optimum with lower log-likelihood than the simpler model (see first panel in
+converged either back to the nested single origin model (in which case the
+single origin model was chosen), or it found a local optimum with lower
+log-likelihood than the simpler model (see first panel in
 Figure~\ref{fig:refit-lls}).
@@
     \caption{
-        \textbf{Fits of four models to simulated data.} In three scenarios, a
-        simple single origin model and our preferred continuous migration and
-        merger with stem migration models, we simulated 500 replicates of 500
-        Mb with the same populations and sample sizes from our reported best
-        fit models. We then fit the four models reported in the main text to
-        each of these simulated datasets and compared log-likelihoods. In each
-        case, the simulated model is correctly chosen, even under the single
-        origin scenario. We conclude our more complex models (with more
-        parameters) are not simply fitting statistical noise.
+        \textbf{Log-likelihoods of fits of four models to simulated data.} In
+        three scenarios, a simple single origin model and our preferred
+        continuous migration and merger with stem migration models, we
+        simulated 500 replicates of 500 Mb with the same populations and sample
+        sizes from our reported best fit models. We then fit the four models
+        reported in the main text to each of these simulated datasets and
+        compared log-likelihoods. In each case, the simulated model is
+        correctly chosen, even under the single origin scenario. We conclude
+        our more complex models (with more parameters) are not simply fitting
+        statistical noise.
     }

### 2023-01-06 paper.tex
typos

@@
 %Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans."
-but also in southern \citep{Hammer2011-bx,Lorente-Galdos2019-vz} and central African \citep{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz}.
+but also in southern \citep{Hammer2011-bx,Lorente-Galdos2019-vz}
+and central African \citep{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz}
 populations.
@@
 the present population location with the greatest ancestry contribution from each stem. For example,
-Stem 1S contributes 71\% to the ancestral formation of the Khoe-San. The extent of the isolation >400ka
+Stem 1S contributes 71\% to the ancestral formation of the Khoe-San. The extent of the isolation \>400ka
 between Stem 1S, Stem 1E and Stem 2 suggests that these stems were not proximate to each other. While the length of
-isolation among the stems is variable across model fits, models with a period of 
+isolation among the stems is variable across fits, models with a period of 
 divergence, isolation and then a merger event (i.e. a ``reticulation'')
@@
 that of the rate to eastern Africans during this period. We performed a variety of validation tests
-to explore sensitivity of our assumptions, including relaxing fixed parameters (SI Section~\ref{sec:model_validation_strategy}).
+to explore sensitivity of our assumptions, including relaxing fixed parameters (SI Section~\ref{sec:validation_strategy}).
 Most validation tests resulted in parameters similar to the models discussed above. However, one exception

### 2023-01-06 supplement.tex
update CIs with fits from perturbed initial params

@@
 
+\aprcomment{These paragraphs are outdated.. we drop Godambe. need to fix}
+
 For the best fit parameters, we computed $95\%$ confidence intervals using the
@@
 
+\aprcomment{This section will be removed}
+
 We created 200 block-bootstrap replicate datasets and refit our four reported
@@
     \toprule
-    Parameter & Description & Value & Std. err.\\
+    Parameter & Description & Value & 95\% CI \\
     \midrule
-    $N_e$ & Ancestral effective population size & 10198 & 403 \\
-    $N_{MH}$ & Size of human lineage between Neanderthal and Nama splits & 21111 & 529 \\
-    $N_{Nama_0}$ & Initial Nama size & 10224 & 370 \\
-    $N_{Nama_F}$ & Final Nama size & 222 & 9 \\
-    $N_{MSL_0}$ & Initial Mende size & 17211 & 769 \\
-    $N_{MSL_F}$ & Final Mende size & 16822 & 606 \\
-    $N_{EA}$ & Size of East African branch & 7139 & 273 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3831 & 131 \\
-    $N_{EP}$ & East African agriculturist size & 13033 & 491 \\
-    $N_{GBR_0}$ & Initial British size & 846 & 33 \\
-    $N_{GBR_F}$ & Final British size & 12121 & 507 \\
-    $N_{Neand}$ & Neanderthal size & 1867 & 105 \\
-    $T_{Nama}$ & Nama split time (years) & 110400 & 2525 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $2.82\times10^{-5}$ & $0.158\times10^{-5}$ \\
-    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.94\times10^{-5}$ & $0.197\times10^{-5}$ \\
-    $m_{MSL-EA}$ & Mende--East Africa migration rate & $18.76\times10^{-5}$ & $0.764\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.42\times10^{-5}$ & $0.239\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $41.28\times10^{-5}$ & $1.33\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.658 & 0.0039 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.279 & 0.0039 \\
-    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.150 & 0.0019 \\
+    $N_e$ & Ancestral effective population size & 9910 & 9600--10090 \\
+    $N_{MH}$ & Size of human lineage between Neanderthal and Nama splits & 20890 & 20640--21220 \\
+    $N_{Nama_0}$ & Initial Nama size & 10170 & 9950--10250 \\
+    $N_{Nama_F}$ & Final Nama size & 223 & 217--226 \\
+    $N_{MSL_0}$ & Initial Mende size & 17250 & 16770--17490 \\
+    $N_{MSL_F}$ & Final Mende size & 16760 & 16370--17060 \\
+    $N_{EA}$ & Size of East African branch & 7130 & 6940--7130 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3830 & 3690--3900 \\
+    $N_{EP}$ & East African agriculturist size & 13010 & 12720--13140 \\
+    $N_{GBR_0}$ & Initial British size & 843 & 836--862 \\
+    $N_{GBR_F}$ & Final British size & 12150 & 11870--12300 \\
+    $N_{Neand}$ & Neanderthal size & 1850 & 1650--1990 \\
+    $T_{Nama}$ & Nama split time (years) & 110700 & 107800--112200 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $2.82\times10^{-5}$ & $2.61\times10^{-5}$--$3.02\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.98\times10^{-5}$ & $4.82\times10^{-5}$--$5.31\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $18.7\times10^{-5}$ & $18.7\times10^{-5}$--$19.2\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.42\times10^{-5}$ & $4.13\times10^{-5}$--$4.54\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $41.4\times10^{-5}$ & $41.0\times10^{-5}$--$44.2\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.659 & 0.657--0.671 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.280 & 0.270--0.286 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.150 & 0.147--0.154 \\
     \bottomrule
@@
     \toprule
-    Parameter & Description & Value & Std. err.\\
+    Parameter & Description & Value & 95\% CIs \\
     \midrule
-    $N_e$ & Ancestral effective population size & 10543 & 460 \\
-    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 15297 & 1227 \\
-    $N_{stem2}$ & Size of stem 2 lineage & 9156 & 573 \\
-    $N_{Nama_0}$ & Initial Nama size & 11329 & 423 \\
-    $N_{Nama_F}$ & Final Nama size & 219 & 11 \\
-    $N_{MSL_0}$ & Initial Mende size & 11187 & 644 \\
-    $N_{MSL_F}$ & Final Mende size & 24056 & 945 \\
-    $N_{EA}$ & Size of East African branch & 7185 & 293 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3746 & 184 \\
-    $N_{EP}$ & East African agriculturist size & 13187 & 645 \\
-    $N_{GBR_0}$ & Initial British size & 937 & 41 \\
-    $N_{GBR_F}$ & Final British size & 11478 & 420 \\
-    $N_{Neand}$ & Neanderthal size & 2285 & 113 \\
-    $T_{stems}$ & Stem split time (years) & 459720 & 21704 \\
-    $T_{Nama}$ & Nama split time (years) & 139725 & 3250 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $1.93\times10^{-5}$ & $0.360\times10^{-5}$ \\
-    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.43\times10^{-5}$ & $0.309\times10^{-5}$ \\
-    $m_{MSL-EA}$ & Mende--East Africa migration rate & $20.8\times10^{-5}$ & $0.97\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.46\times10^{-5}$ & $0.171\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.4\times10^{-5}$ & $1.85\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.649 & 0.0038 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.268 & 0.0048 \\
-    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.153 & 0.0022 \\
+    $N_e$ & Ancestral effective population size & 10543 & 10810--11260 \\
+    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 15297 & 13930--15330 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 9156 & 9780--11250 \\
+    $N_{Nama_0}$ & Initial Nama size & 11329 & 10760--11280 \\
+    $N_{Nama_F}$ & Final Nama size & 219 & 216--226 \\
+    $N_{MSL_0}$ & Initial Mende size & 11187 & 10510--11090 \\
+    $N_{MSL_F}$ & Final Mende size & 24056 & 24050--25780 \\
+    $N_{EA}$ & Size of East African branch & 7185 & 6970--7200 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3746 & 2650--3860 \\
+    $N_{EP}$ & East African agriculturist size & 13187 & 12830--13250 \\
+    $N_{GBR_0}$ & Initial British size & 937 & 931--964 \\
+    $N_{GBR_F}$ & Final British size & 11478 & 11320--11760 \\
+    $N_{Neand}$ & Neanderthal size & 2285 & 2140--2450 \\
+    $T_{stems}$ & Stem split time (years) & 459720 & 446600--463300 \\
+    $T_{Nama}$ & Nama split time (years) & 139725 & 137300--146000 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $1.93\times10^{-5}$ & $1.36\times10^{-5}$--$1.81\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.43\times10^{-5}$ & $4.08\times10^{-5}$--$4.60\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $20.8\times10^{-5}$ & $20.7\times10^{-5}$--$21.3\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.46\times10^{-5}$ & $4.19\times10^{-5}$--$4.58\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.4\times10^{-5}$ & $34.6\times10^{-5}$--$37.0\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.649 & 0.644--0.660 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.268 & 0.264--0.279 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.153 & 0.148--0.156 \\
     $m_{stems}$ & Stem 1--stem 2 migration rate & $0$ & fixed \\
-    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $4.33\times10^{-5}$ & $0.891\times10^{-5}$ \\
-    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $11.3\times10^{-5}$ & $1.83\times10^{-5}$ \\
-    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $2.77\times10^{-5}$ & $0.699\times10^{-5}$ \\
+    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $4.33\times10^{-5}$ & $4.91\times10^{-5}$--$5.92\times10^{-5}$ \\
+    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $11.3\times10^{-5}$ & $12.4\times10^{-5}$--$14.4\times10^{-5}$ \\
+    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $2.77\times10^{-5}$ & $3.04\times10^{-5}$--$3.95\times10^{-5}$ \\
     \bottomrule
@@
     \toprule
-    Parameter & Description & Value & Std. err.\\
+    Parameter & Description & Value & 95\% CIs\\
     \midrule
-    $N_e$ & Ancestral effective population size & 7270 & 1777 \\
-    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 8256 & 1612 \\
-    $N_{stem2}$ & Size of stem 2 lineage & 13547 & 2488 \\
-    $N_{Nama_0}$ & Initial Nama size & 11939 & 2989 \\
-    $N_{Nama_F}$ & Final Nama size & 221 & 54 \\
-    $N_{MSL_0}$ & Initial Mende size & 9738 & 2479 \\
-    $N_{MSL_F}$ & Final Mende size & 28150 & 6628 \\
-    $N_{EA}$ & Size of East African branch & 7489 & 1841 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3728 & 915 \\
-    $N_{EP}$ & East African agriculturist size & 13072 & 3246 \\
-    $N_{GBR_0}$ & Initial British size & 959 & 231 \\
-    $N_{GBR_F}$ & Final British size & 11822 & 2889 \\
-    $N_{Neand}$ & Neanderthal size & 2670 & 591 \\
-    $T_{stems}$ & Stem split time (years) & 1163072 & 390803 \\
-    $T_{Nama}$ & Nama split time (years) & 134745 & 17775 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.98\times10^{-5}$ & $0.366\times10^{-5}$ \\
-    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.08\times10^{-5}$ & $1.02\times10^{-5}$ \\
-    $m_{MSL-EA}$ & Mende--East Africa migration rate & $21.4\times10^{-5}$ & $5.32\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.17\times10^{-5}$ & $1.02\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $33.6\times10^{-5}$ & $8.35\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.642 & 0.0037 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.255 & 0.0043 \\
-    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.0021 \\
-    $m_{stems}$ & Stem 1--stem 2 migration rate & $6.43\times10^{-5}$ & $1.05\times10^{-5}$ \\
-    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $5.82\times10^{-5}$ & $1.60\times10^{-5}$ \\
-    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $16.4\times10^{-5}$ & $4.19\times10^{-5}$ \\
-    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $3.10\times10^{-5}$ & $0.901\times10^{-5}$ \\
+    $N_e$ & Ancestral effective population size & 7270 & 4180--6530 \\
+    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 8260 & 7130--8400 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 13550 & 11330--13580 \\
+    $N_{Nama_0}$ & Initial Nama size & 11940 & 11160--12390 \\
+    $N_{Nama_F}$ & Final Nama size & 221 & 215--226 \\
+    $N_{MSL_0}$ & Initial Mende size & 9740 & 9500--10210 \\
+    $N_{MSL_F}$ & Final Mende size & 28150 & 25930--28930 \\
+    $N_{EA}$ & Size of East African branch & 7490 & 7270--7630 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3730 & 3560--3810 \\
+    $N_{EP}$ & East African agriculturist size & 13070 & 12800--13290 \\
+    $N_{GBR_0}$ & Initial British size & 959 & 940--981 \\
+    $N_{GBR_F}$ & Final British size & 11820 & 11490--12090 \\
+    $N_{Neand}$ & Neanderthal size & 2670 & 2380--27950 \\
+    $T_{stems}$ & Stem split time (years) & 1163000 & 1120000--1382000 \\ %%% Outside CI!!!
+    $T_{Nama}$ & Nama split time (years) & 134700 & 127800--139900 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.98\times10^{-5}$ & $0.41\times10^{-5}$--$1.43\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.08\times10^{-5}$ & $3.92\times10^{-5}$--$4.56\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $21.4\times10^{-5}$ & $21.2\times10^{-5}$--$21.8\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.17\times10^{-5}$ & $3.77\times10^{-5}$--$4.33\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $33.6\times10^{-5}$ & $33.0\times10^{-5}$--$36.3\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.642 & 0.639--0.656 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.255 & 0.245--0.263 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.151--0.160 \\
+    $m_{stems}$ & Stem 1--stem 2 migration rate & $6.43\times10^{-5}$ & $6.14\times10^{-5}$--$7.83\times10^{-5}$ \\
+    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $5.82\times10^{-5}$ & $4.50\times10^{-5}$--$7.04\times10^{-5}$ \\
+    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $16.4\times10^{-5}$ & $13.8\times10^{-5}$--$18.9\times10^{-5}$ \\
+    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $3.10\times10^{-5}$ & $2.18\times10^{-5}$--$3.65\times10^{-5}$ \\
     \bottomrule
@@
     \toprule
-    Parameter & Description & Value & Std. err.\\
+    Parameter & Description & Value & 95\% CIs\\
     \midrule
-    $N_e$ & Ancestral effective population size & 11258 & 326 \\
-    $N_{stem1}$ & Size of stem 1 lineage between stem 1--stem 2 split and stem 1E--stem 1S split & 113 & 76 \\
-    $N_{stem2}$ & Size of stem 2 lineage & 23984 & 1149 \\
-    $N_{Nama_0}$ & Initial Nama and stem 1S size & 13134 & 384 \\
-    $N_{Nama_F}$ & Final Nama size & 225 & 7.3 \\
-    $N_{MSL_0}$ & Initial Mende size & 11856 & 322 \\
-    $N_{MSL_F}$ & Final Mende size & 25558 & 987 \\
-    $N_{EA}$ & Size of East African and stem 1E branch & 9136 & 246 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3385 & 102 \\
-    $N_{EP}$ & East African agriculturist size & 13650 & 408 \\
-    $N_{GBR_0}$ & Initial British size & 931 & 29 \\
-    $N_{GBR_F}$ & Final British size & 12064 & 334 \\
-    $N_{Neand}$ & Neanderthal size & 1935 & 91 \\
-    $T_{stems}$ & Stem split time (years) & 420881 & 27380 \\
-    $T_{stem1}$ & Stem 1 split time into stem 1E and stem 1S (years) & 367434 & 19952 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.361\times10^{-5}$ & $0.113\times10^{-5}$ \\
-    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.00\times10^{-5}$ & $0.130\times10^{-5}$ \\
-    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.5\times10^{-5}$ & $0.548\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.77\times10^{-5}$ & $0.152\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $37.1\times10^{-5}$ & $1.26\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.647 & 0.0037 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.257 & 0.0042 \\
-    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.0021 \\
+    $N_e$ & Ancestral effective population size & 11300 & 10100--11290 \\
+    $N_{stem1}$ & Size of stem 1 lineage between stem 1--stem 2 split and stem 1E--stem 1S split & 113 & 113--606 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 24780 & 26500--35060 \\
+    $N_{Nama_0}$ & Initial Nama and stem 1S size & 13080 & 11310--12910 \\
+    $N_{Nama_F}$ & Final Nama size & 224 & 223--238 \\
+    $N_{MSL_0}$ & Initial Mende size & 11870 & 10280--11630 \\
+    $N_{MSL_F}$ & Final Mende size & 25570 & 25720--30180 \\
+    $N_{EA}$ & Size of East African and stem 1E branch & 9110 & 8680--9140 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3370 & 3270--3470 \\
+    $N_{EP}$ & East African agriculturist size & 13550 & 13120--13690 \\
+    $N_{GBR_0}$ & Initial British size & 927 & 920--963 \\
+    $N_{GBR_F}$ & Final British size & 12050 & 11790--12430 \\
+    $N_{Neand}$ & Neanderthal size & 1940 & 1770--2110 \\
+    $T_{stems}$ & Stem split time (years) & 411700 & 381400--410700 \\
+    $T_{stem1}$ & Stem 1 split time into stem 1E and stem 1S (years) & 355400 & 154300--320400 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.361\times10^{-5}$ & $0.205\times10^{-5}$--$0.697\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.04\times10^{-5}$ & $3.90\times10^{-5}$--$4.42\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.3\times10^{-5}$ & $19.8\times10^{-5}$--$20.8\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.76\times10^{-5}$ & $3.34\times10^{-5}$--$3.95\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $37.2\times10^{-5}$ & $35.2\times10^{-5}$--$39.6\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.648 & 0.641--0.661 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.258 & 0.245--0.264 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.152--0.161 \\
     $m_{stems}$ & Stem 1--stem 2 migration rate & $0$ & fixed \\
-    $T_{Nama}$ & Time of Nama merger event & 117392 & 8253 \\
-    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.707 & 0.0086 \\
-    $T_{EA}$ & Time of East Africa merger event & 94892 & 3648 \\
-    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.481 & 0.0074 \\
-    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 23922 & 570 \\
-    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.168 & 0.0036 \\
+    $T_{Nama}$ & Time of Nama merger event & 115400 & 101400--116800 \\
+    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.693 & 0.706--0.769 \\
+    $T_{EA}$ & Time of East Africa merger event & 92940 & 94500--10140 \\ %%% outside CI!
+    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.502 & 0.510--0.564 \\
+    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 23490 & 22920--25510 \\
+    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.170 & 0.174--0.196 \\
     \bottomrule
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -102,633.
+    29 years. This model gave a log-likelihood of -102,633. \aprcomment{check these at end}
+    \aprcomment{if the slightly shift, also fix elsewhere in supp and main text}
 }
@@
     \midrule
-    $N_e$ & Ancestral effective population size & 11479 & 1369 \\
-    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal split and stem 1E--stem 1S split & 117 & 838 \\
-    $N_{stem2}$ & Size of stem 2 lineage & 24393 & 6668 \\
-    $N_{Nama_0}$ & Initial Nama size & 13211 & 1514 \\
-    $N_{Nama_F}$ & Final Nama size & 223 & 31 \\
-    $N_{MSL_0}$ & Initial Mende size & 11444 & 1165 \\
-    $N_{MSL_F}$ & Final Mende size & 27417 & 4332 \\
-    $N_{EA}$ & Size of East African Branch & 9077 & 1628 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3402 & 337 \\
-    $N_{EP}$ & East African agriculturist size & 13506 & 1684 \\
-    $N_{GBR_0}$ & Initial British size & 953 & 122 \\
-    $N_{GBR_F}$ & Final British size & 12406 & 1678 \\
-    $N_{Neand}$ & Neanderthal size & 2416 & 235 \\
-    $T_{stems}$ & Stem split time (years) & 1442022 & 426449 \\
-    $T_{stem1}$ & Stem 1S--stem 1E split time (years) & 479401 & 166339 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.712\times10^{-5}$ & $0.401\times10^{-5}$ \\
-    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.35\times10^{-5}$ & $0.912\times10^{-5}$ \\
-    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.8\times10^{-5}$ & $2.57\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.87\times10^{-5}$ & $0.550\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.9\times10^{-5}$ & $5.36\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.640 & 0.0075 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.257 & 0.0049 \\
-    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.157 & 0.0031 \\
-    $m_{stems}$ & Stem 1--stem 2 migration rate & $11.6\times10^{-5}$ & $8.74\times10^{-5}$ \\
-    $T_{Nama}$ & Time of Nama merger event & 118547 & 28170 \\
-    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.714 & 0.067 \\
-    $T_{EA}$ & Time of East Africa merger event & 98083 & 8865 \\
-    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.495 & 0.059 \\
-    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 25119 & 641 \\
-    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.181 & 0.0085 \\
+    $N_e$ & Ancestral effective population size & 10960 & 64500--11910 \\
+    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal split and stem 1E--stem 1S split & 100 & 100--851 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 24100 & 22130--26340 \\
+    $N_{Nama_0}$ & Initial Nama size & 13530 & 11500--13560 \\
+    $N_{Nama_F}$ & Final Nama size & 219 & 214--231 \\
+    $N_{MSL_0}$ & Initial Mende size & 11500 & 9690--11290 \\
+    $N_{MSL_F}$ & Final Mende size & 27390 & 26820--32280 \\
+    $N_{EA}$ & Size of East African Branch & 9150 & 7970--8860 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3390 & 3270--3570 \\
+    $N_{EP}$ & East African agriculturist size & 13500 & 12780--13500 \\
+    $N_{GBR_0}$ & Initial British size & 950 & 928--1002 \\
+    $N_{GBR_F}$ & Final British size & 12370 & 11850--12600 \\
+    $N_{Neand}$ & Neanderthal size & 2510 & 2100--2730 \\
+    $T_{stems}$ & Stem split time (years) & 1475000 & 695000--1785000 \\
+    $T_{stem1}$ & Stem 1S--stem 1E split time (years) & 479100 & 275600--478300 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $1.06\times10^{-5}$ & $0.66\times10^{-5}$--$1.57\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.49\times10^{-5}$ & $4.45\times10^{-5}$--$5.52\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.6\times10^{-5}$ & $20.0\times10^{-5}$--$21.7\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.85\times10^{-5}$ & $3.25\times10^{-5}$--$4.27\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $36.0\times10^{-5}$ & $33.7\times10^{-5}$--$40.9\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.639 & 0.636--0.667 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.259 & 0.237--0.267 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.159 & 0.154--0.167 \\
+    $m_{stems}$ & Stem 1--stem 2 migration rate & $11.8\times10^{-5}$ & $3.95\times10^{-5}$--$12.3\times10^{-5}$ \\
+    $T_{Nama}$ & Time of Nama merger event & 121600 & 101300--124600 \\
+    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.692 & 0.694--0.758 \\
+    $T_{EA}$ & Time of East Africa merger event & 96500 & 95400--111000 \\
+    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.511 & 0.514--0.589 \\
+    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 24890 & 22210--26300 \\
+    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.182 & 0.182--0.211 \\
     \bottomrule
@@
 
-\begin{table}[ht]
-    \caption{
-        \label{tab:supp-confidence-intervals}
-        \textbf{GIM-computed confidence intervals are not underestimated.}
-        CIs computed with the Godambe information matrix (GIM) are largely
-        similar (i.e., within an order of magnitude) to those estimated by
-        creating 200 block bootstrap-replicate datasets (BS) and refitting each
-        modelto these datasets (see Section~\ref{sec:bootstrap-CIs}). Standard
-        deviations for each inferred parameter were computed from these 200
-        model fits. Parameters are as labeled and described in the above
-        tables. Blank standard errors indicate that parameter was not relevant
-        to the given model.
-    }
-    \centering
-    \begin{tabular}[t]{l|cc|cc|cc|cc}
-        \toprule
-        & \multicolumn{2}{c|}{Single origin} & \multicolumn{2}{c|}{Cont. mig.}
-        & \multicolumn{2}{c|}{Merger w/o mig.} & \multicolumn{2}{c}{Merger w/mig.} \\
-        Parameter & GIM & BS & GIM & BS & GIM & BS & GIM & BS \\
-        \midrule
-        $N_e$ & 403 & 127 & 1777 & 951 & 326 & 406 & 1369 & 1408 \\
-        $N_{MH}$ & 529 & 158 & & & & & & \\
-        $N_{stem1}$ & & & 1612 & 406 & 76 & 122 & 838 & 164 \\
-        $N_{stem2}$ & & & 2488 & 1001 & 1149 & 2344 & 6668 & 1154 \\
-        $N_{Nama_0}$ & 370 & 79 & 2989 & 512 & 384 & 496 & 1514 & 538 \\
-        $N_{Nama_F}$ & 9 & 2 & 54 & 4 & 7 & 4 & 31 & 5 \\
-        $N_{MSL_0}$ & 769 & 190 & 2479 & 285 & 322 & 384 & 1165 & 421 \\
-        $N_{MSL_F}$ & 606 & 185 & 6628 & 1250 & 987 & 1031 & 4332 & 1404 \\
-        $N_{EA}$ & 273 & 49 & 1841 & 138 & 246 & 154 & 1628 & 252 \\
-        $N_{Gumuz_F}$ & 131 & 56 & 915 & 80 & 102 & 72 & 337 & 83 \\
-        $N_{EP}$ & 491 & 106 & 3246 & 168 & 408 & 177 & 1684 & 185 \\
-        $N_{GBR_0}$ & 33 & 7 & 231 & 17 & 29 & 17 & 122 & 21 \\
-        $N_{GBR_F}$ & 507 & 117 & 2889 & 295 & 334 & 203 & 1678 & 218 \\
-        $N_{Neand}$ & 105 & 88 & 591 & 97 & 91 & 94 & 235 & 158 \\
-        $T_{stems}$ & & & 391E3 & 67E3 & 27E3 & 7.6E3 & 426E3 & 284E3 \\
-        $T_{stem1}$ & & & & & 20E3 & 45E3 & 166E3 & 58E3 \\
-        $m_{Nama-MSL}$ & 1.6E-6 & 1.2E-6 & 3.7E-6 & 2.7E-6 & 1.1E-6 & 1.3E-6 & 4.0E-6 & 2.6E-6 \\
-        $m_{Nama-EA}$ & 2.0E-6 & 1.3E-6 & 1.0E-5 & 1.8E-6 & 1.3E-6 & 1.8E-6 & 9.1E-6 & 2.8E-6 \\
-        $m_{MSL-EA}$ & 7.6E-6 & 1.1E-6 & 5.3E-5 & 3.0E-6 & 5.5E-6 & 3.3E-6 & 2.6E-5 & 4.4E-6 \\
-        $m_{EA-GBR}$ & 2.4E-6 & 1.1E-6 & 1.0E-5 & 1.9E-6 & 1.5E-6 & 2.2E-6 & 5.5E-6 & 2.6E-6 \\
-        $m_{EA-EA}$ & 1.3E-5 & 7.5E-6 & 8.4E-5 & 1.4E-5 & 1.3E-5 & 1.4E-5 & 5.4E-5 & 1.9E-5 \\
-        $f_{GBR \rightarrow EP}$ & 0.0039 & 0.0036 & 0.0037 & 0.0071 & 0.0037 & 0.0066 & 0.0075 & 0.0084 \\
-        $f_{EP \rightarrow Nama}$ & 0.0039 & 0.0040 & 0.0043 & 0.0072 & 0.0042 & 0.0068 & 0.0049 & 0.0085 \\
-        $f_{GBR \rightarrow Nama}$ & 0.0019 & 0.0020 & 0.0021 & 0.0030 & 0.0021 & 0.0030 & 0.0031 & 0.0035 \\
-        $m_{stems}$ & & & 1.1E-5 & 7.7E-6 & & & 8.7E-5 & 2.5E-5 \\
-        $m_{stem2-Nama}$ & & & 1.6E-5 & 1.1E-5 & & & & \\
-        $m_{stem2-MSL}$ & & & 4.2E-5 & 1.7E-5 & & & & \\
-        $m_{stem2-EA}$ & & & 9.0E-6 & 5.4E-6 & & & & \\
-        $T_{Nama}$ & 2525 & 1188 & 18E3 & 6400 & 8253 & 4763 & 28E3 & 6283 \\
-        $f_{stem 2 \rightarrow Nama}$ & & & & & 0.0086 & 0.017 & 0.067 & 0.018 \\
-        $T_{EA}$ & & & & & 3648 & 2047 & 8865 & 4114 \\
-        $f_{stem 2 \rightarrow EA}$ & & & & & 0.0074 & 0.015 & 0.059 & 0.021 \\
-        $T_{MSL}$ & & & & & 570 & 1373 & 641 & 1613 \\
-        $f_{stem 2 \rightarrow MSL}$ & & & & & 0.0036 & 0.0066 & 0.0085 & 0.0076 \\
-        \bottomrule
-    \end{tabular}
-\end{table}
+%\begin{table}[ht]
+%    \caption{
+%        \label{tab:supp-confidence-intervals}
+%        \textbf{GIM-computed confidence intervals are not underestimated.}
+%        \aprcomment{This is out of date, needs to be removed}
+%        CIs computed with the Godambe information matrix (GIM) are largely
+%        similar (i.e., within an order of magnitude) to those estimated by
+%        creating 200 block bootstrap-replicate datasets (BS) and refitting each
+%        modelto these datasets (see Section~\ref{sec:bootstrap-CIs}). Standard
+%        deviations for each inferred parameter were computed from these 200
+%        model fits. Parameters are as labeled and described in the above
+%        tables. Blank standard errors indicate that parameter was not relevant
+%        to the given model.
+%    }
+%    \centering
+%    \begin{tabular}[t]{l|cc|cc|cc|cc}
+%        \toprule
+%        & \multicolumn{2}{c|}{Single origin} & \multicolumn{2}{c|}{Cont. mig.}
+%        & \multicolumn{2}{c|}{Merger w/o mig.} & \multicolumn{2}{c}{Merger w/mig.} \\
+%        Parameter & GIM & BS & GIM & BS & GIM & BS & GIM & BS \\
+%        \midrule
+%        $N_e$ & 403 & 127 & 1777 & 951 & 326 & 406 & 1369 & 1408 \\
+%        $N_{MH}$ & 529 & 158 & & & & & & \\
+%        $N_{stem1}$ & & & 1612 & 406 & 76 & 122 & 838 & 164 \\
+%        $N_{stem2}$ & & & 2488 & 1001 & 1149 & 2344 & 6668 & 1154 \\
+%        $N_{Nama_0}$ & 370 & 79 & 2989 & 512 & 384 & 496 & 1514 & 538 \\
+%        $N_{Nama_F}$ & 9 & 2 & 54 & 4 & 7 & 4 & 31 & 5 \\
+%        $N_{MSL_0}$ & 769 & 190 & 2479 & 285 & 322 & 384 & 1165 & 421 \\
+%        $N_{MSL_F}$ & 606 & 185 & 6628 & 1250 & 987 & 1031 & 4332 & 1404 \\
+%        $N_{EA}$ & 273 & 49 & 1841 & 138 & 246 & 154 & 1628 & 252 \\
+%        $N_{Gumuz_F}$ & 131 & 56 & 915 & 80 & 102 & 72 & 337 & 83 \\
+%        $N_{EP}$ & 491 & 106 & 3246 & 168 & 408 & 177 & 1684 & 185 \\
+%        $N_{GBR_0}$ & 33 & 7 & 231 & 17 & 29 & 17 & 122 & 21 \\
+%        $N_{GBR_F}$ & 507 & 117 & 2889 & 295 & 334 & 203 & 1678 & 218 \\
+%        $N_{Neand}$ & 105 & 88 & 591 & 97 & 91 & 94 & 235 & 158 \\
+%        $T_{stems}$ & & & 391E3 & 67E3 & 27E3 & 7.6E3 & 426E3 & 284E3 \\
+%        $T_{stem1}$ & & & & & 20E3 & 45E3 & 166E3 & 58E3 \\
+%        $m_{Nama-MSL}$ & 1.6E-6 & 1.2E-6 & 3.7E-6 & 2.7E-6 & 1.1E-6 & 1.3E-6 & 4.0E-6 & 2.6E-6 \\
+%        $m_{Nama-EA}$ & 2.0E-6 & 1.3E-6 & 1.0E-5 & 1.8E-6 & 1.3E-6 & 1.8E-6 & 9.1E-6 & 2.8E-6 \\
+%        $m_{MSL-EA}$ & 7.6E-6 & 1.1E-6 & 5.3E-5 & 3.0E-6 & 5.5E-6 & 3.3E-6 & 2.6E-5 & 4.4E-6 \\
+%        $m_{EA-GBR}$ & 2.4E-6 & 1.1E-6 & 1.0E-5 & 1.9E-6 & 1.5E-6 & 2.2E-6 & 5.5E-6 & 2.6E-6 \\
+%        $m_{EA-EA}$ & 1.3E-5 & 7.5E-6 & 8.4E-5 & 1.4E-5 & 1.3E-5 & 1.4E-5 & 5.4E-5 & 1.9E-5 \\
+%        $f_{GBR \rightarrow EP}$ & 0.0039 & 0.0036 & 0.0037 & 0.0071 & 0.0037 & 0.0066 & 0.0075 & 0.0084 \\
+%        $f_{EP \rightarrow Nama}$ & 0.0039 & 0.0040 & 0.0043 & 0.0072 & 0.0042 & 0.0068 & 0.0049 & 0.0085 \\
+%        $f_{GBR \rightarrow Nama}$ & 0.0019 & 0.0020 & 0.0021 & 0.0030 & 0.0021 & 0.0030 & 0.0031 & 0.0035 \\
+%        $m_{stems}$ & & & 1.1E-5 & 7.7E-6 & & & 8.7E-5 & 2.5E-5 \\
+%        $m_{stem2-Nama}$ & & & 1.6E-5 & 1.1E-5 & & & & \\
+%        $m_{stem2-MSL}$ & & & 4.2E-5 & 1.7E-5 & & & & \\
+%        $m_{stem2-EA}$ & & & 9.0E-6 & 5.4E-6 & & & & \\
+%        $T_{Nama}$ & 2525 & 1188 & 18E3 & 6400 & 8253 & 4763 & 28E3 & 6283 \\
+%        $f_{stem 2 \rightarrow Nama}$ & & & & & 0.0086 & 0.017 & 0.067 & 0.018 \\
+%        $T_{EA}$ & & & & & 3648 & 2047 & 8865 & 4114 \\
+%        $f_{stem 2 \rightarrow EA}$ & & & & & 0.0074 & 0.015 & 0.059 & 0.021 \\
+%        $T_{MSL}$ & & & & & 570 & 1373 & 641 & 1613 \\
+%        $f_{stem 2 \rightarrow MSL}$ & & & & & 0.0036 & 0.0066 & 0.0085 & 0.0076 \\
+%        \bottomrule
+%    \end{tabular}
+%\end{table}
 

### 2023-01-06 supplement.tex
final single origin best fit values

@@
 \texttt{moments-LD} uses a composite multivariate Gaussian likelihood approach
-to simultaneously fit relative pairwise diversity and LD statistics over a range of 
-recombination distances.
-This composite likelihood and Gaussian assumption are described in more detail 
- and validated in \citet{Ragsdale2019-nt}, but we outline them here.  
-To compute likelihoods, we need an estimate of the joint
-distribution of summary statistics. 
-We estimate uncertainty due to the finite amount of genetic material
-used in inference using bootstrap over 500 segments along the genome with
-roughly equal lengths of retained sequences within each segment. First, for
-each distance bin, we use these bootstrap samples to obtain a
-variance-covariances matrix across all statistics. This variance-covariance
-matrix is used to obtain a model likelihood for each recombination distance
-bin and single-locus nucleotide diversity, as a multivariate Gaussian
-likelihood. 
-The full model likelihood is taken as the product of likelihoods
-over each bin. In other words, we followed the approach of  \citet{Ragsdale2019-nt} 
-in optimizing a composite likelihood where
-observations in different bins were taken to be independent.
-
-\aprcomment{These paragraphs are outdated.. we drop Godambe. need to fix}
-
-For the best fit parameters, we computed $95\%$ confidence intervals using the
-same set of bootstrapped samples and Godambe Information approach
-\citep{Coffman2016-yq}, which corrects composite likelihood estimates of
-confidence intervals to account for nonindependence in the data, including
-linkage between loci and nonindependence of recombination bins.  To ensure that
-our confidence intervals computed using the approach from
-\citet{Coffman2016-yq} do not underestimate true CIs, we also performed block
-bootstrap resampling of our data, 200 times, and refit each model to the
-bootstrapped datasets. From these, we evaluated the distribution of inferred
-parameters, which should be similar to or narrower than the CIs estimated from
-the GIM approach (Section~\ref{sec:bootstrap-CIs}).
+to simultaneously fit relative pairwise diversity and LD statistics over a
+range of recombination distances. This composite likelihood and Gaussian
+assumption are described in more detail and validated in
+\citet{Ragsdale2019-nt}, but we outline them here. To compute likelihoods, we
+need an estimate of the joint distribution of summary statistics. We estimate
+uncertainty due to the finite amount of genetic material used in inference
+using bootstrap over 500 segments along the genome with roughly equal lengths
+of retained sequences within each segment. First, for each distance bin, we use
+these bootstrap samples to obtain a variance-covariances matrix across all
+statistics. This variance-covariance matrix is used to obtain a model
+likelihood for each recombination distance bin and single-locus nucleotide
+diversity, as a multivariate Gaussian likelihood. The full model likelihood is
+taken as the product of likelihoods over each bin. In other words, we followed
+the approach of \citet{Ragsdale2019-nt} in optimizing a composite likelihood
+where observations in different bins were taken to be independent. We computed
+$95\%$ confidence intervals for each model using inferences from 200
+block-bootstrapped datasets, which were fit using the same procedure as the
+original data. From the resulting parameters from these fits, we removed the
+bottom $2.5\%$ and top $97.5\%$ values and report the range of the remaining
+fit parameter values.
 
@@
 give the best fit values for all parameters in the four models considered here,
-along with standard errors using computed using the Godambe method.
+along with 95\% confidence intervals from the refit bootstrapped datasets.
 Figures~\ref{fig:supp-single-origin-fits}--\ref{fig:supp-merger-with-stem-migration-fits}
@@
 
-\begin{figure}[htb!]
+\begin{figure}[ht!]
     \centering
@@
 
-\begin{figure}[htb]
+\begin{figure}[ht!]
     \centering
@@
     \label{fig:refit-lls}
-\end{figure}
-
-\subsection{Evaluation of confidence intervals}
-\label{sec:bootstrap-CIs}
-
-\aprcomment{This section will be removed}
-
-We created 200 block-bootstrap replicate datasets and refit our four reported
-models to each replicate (Section~\ref{sec:likelihoods-uncertainty}). From
-these refit models, we estimated the empirical uncertainty by taking the
-standard deviation of the distribution of each refit parameter. We compared
-these to our estimated standard errors from the Godambe Information Matrix
-(GIM) approach to check that the confidence intervals from the GIM approach do
-not underestimate true uncertainty (Table~\ref{tab:supp-confidence-intervals}).
-
-The GIM-computed standard errors are generally close to those from the fits to
-the bootstrap data, and some are larger (roughly 2--5x) than those from the
-bootstrap fits. A possible explanation for this discrepancy is that some
-parameters of our models come close to the lower and upper bounds allowed for
-those parameters (e.g., small recent population size in the Nama, a bottleneck
-in one of the stems). In such cases, the distribution of
-parameters under constraints, and that of other correlated parameters in the model
-may be narrower than
-the true (unconstrained) uncertainties in those parameters. 
-%Model constraints and parameter
-%bounds in optimization may also lead to parameters fits that are away from
-%their true optimal value. In this case, the GIM-estimated confidence intervals
-%should be larger and contain what would be found from an unconstrained fit.
-
-To demonstrate this effect, we simulated data under a simple
-isolation-with-migration model, in which population sizes are allowed to vary,
-and one population has a severe bottleneck in the most recent 1000 years
-(Figure~\ref{fig:supp-uncerts-toy}). We simulated 200 1Mb regions with
-$u=r=10^{-8}$, sampling 10 diploid individuals from each present-day population
-and computed LD statistics. From these, we inferred the best fit parameters
-without imposing bounds on the parameters, and we also reinferred the model
-imposing a lower bound of 300 on population sizes (the bottleneck size). From
-our fits, we computed confidence intervals using the Fisher information matrix
-(FIM) and GIM approaches, in both the unbounded and bounded fits. We also
-reinferred parameters over 200 bootstrapped replicate datasets to compare
-empirical uncertainties to those estimated CIs.
-
-As expected the FIM CIs underestimate the true uncertainty and the GIM CIs
-closely match empirical uncertainty from the bootstrap fits. In the fit that
-imposed a lower bound on population sizes, some GIM-estimated standard errors
-were larger than those from the unbounded fits.  However, the fit parameters
-were slightly shifted from the unbounded fits, due to hitting the lower bound
-of the bottleneck size. In these cases, the estimated CIs were larger but
-contained the true CIs from the unbounded GIM and bootstrap fits. For parameters
-that were not strongly affected by hitting the bounds, the CIs matched closely
-(Figure~\ref{fig:supp-uncerts-toy}).
-
-\begin{figure}[htb]
-    \centering
-    \includegraphics{./figures/supp-uncerts-toy.pdf}
-    \caption{
-        \textbf{Evaluation of GIM confidence intervals using moments-LD.}
-        The GIM approach accurately estimates confidence intervals, accounting
-        for non-independence between regions of linked loci. GIM-estimated CIs
-        from the constrained fit are large enough to contain the true
-        uncertainty. Vertical dashed lines indicate simulated parameter values.
-    }
-    \label{fig:supp-uncerts-toy}
 \end{figure}
@@
     \midrule
-    $N_e$ & Ancestral effective population size & 9910 & 9600--10090 \\
-    $N_{MH}$ & Size of human lineage between Neanderthal and Nama splits & 20890 & 20640--21220 \\
+    $N_e$ & Ancestral effective population size & 9860 & 9600--10090 \\
+    $N_{MH}$ & Size of human lineage between Neanderthal and Nama splits & 20850 & 20640--21220 \\
     $N_{Nama_0}$ & Initial Nama size & 10170 & 9950--10250 \\
-    $N_{Nama_F}$ & Final Nama size & 223 & 217--226 \\
-    $N_{MSL_0}$ & Initial Mende size & 17250 & 16770--17490 \\
-    $N_{MSL_F}$ & Final Mende size & 16760 & 16370--17060 \\
+    $N_{Nama_F}$ & Final Nama size & 222 & 217--226 \\
+    $N_{MSL_0}$ & Initial Mende size & 17280 & 16770--17490 \\
+    $N_{MSL_F}$ & Final Mende size & 16730 & 16370--17060 \\
     $N_{EA}$ & Size of East African branch & 7130 & 6940--7130 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3830 & 3690--3900 \\
-    $N_{EP}$ & East African agriculturist size & 13010 & 12720--13140 \\
-    $N_{GBR_0}$ & Initial British size & 843 & 836--862 \\
-    $N_{GBR_F}$ & Final British size & 12150 & 11870--12300 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3800 & 3690--3900 \\
+    $N_{EP}$ & East African agriculturist size & 12990 & 12720--13140 \\
+    $N_{GBR_0}$ & Initial British size & 845 & 836--862 \\
+    $N_{GBR_F}$ & Final British size & 12140 & 11870--12300 \\
     $N_{Neand}$ & Neanderthal size & 1850 & 1650--1990 \\
-    $T_{Nama}$ & Nama split time (years) & 110700 & 107800--112200 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $2.82\times10^{-5}$ & $2.61\times10^{-5}$--$3.02\times10^{-5}$ \\
+    $T_{Nama}$ & Nama split time (years) & 110900 & 107800--112200 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $2.83\times10^{-5}$ & $2.61\times10^{-5}$--$3.02\times10^{-5}$ \\
     $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.98\times10^{-5}$ & $4.82\times10^{-5}$--$5.31\times10^{-5}$ \\
     $m_{MSL-EA}$ & Mende--East Africa migration rate & $18.7\times10^{-5}$ & $18.7\times10^{-5}$--$19.2\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.42\times10^{-5}$ & $4.13\times10^{-5}$--$4.54\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $41.4\times10^{-5}$ & $41.0\times10^{-5}$--$44.2\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.659 & 0.657--0.671 \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.38\times10^{-5}$ & $4.13\times10^{-5}$--$4.54\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $43.8\times10^{-5}$ & $41.0\times10^{-5}$--$44.2\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.661 & 0.657--0.671 \\
     $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.280 & 0.270--0.286 \\
@@
     $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.693 & 0.706--0.769 \\
-    $T_{EA}$ & Time of East Africa merger event & 92940 & 94500--10140 \\ %%% outside CI!
+    $T_{EA}$ & Time of East Africa merger event & 92940 & 94500--10140 \\
     $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.502 & 0.510--0.564 \\
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -102,633. \aprcomment{check these at end}
+    29 years. This model gave a log-likelihood of -102,633.
+    \aprcomment{check these at end for each model}
     \aprcomment{if the slightly shift, also fix elsewhere in supp and main text}
@@
 
-%\begin{table}[ht]
-%    \caption{
-%        \label{tab:supp-confidence-intervals}
-%        \textbf{GIM-computed confidence intervals are not underestimated.}
-%        \aprcomment{This is out of date, needs to be removed}
-%        CIs computed with the Godambe information matrix (GIM) are largely
-%        similar (i.e., within an order of magnitude) to those estimated by
-%        creating 200 block bootstrap-replicate datasets (BS) and refitting each
-%        modelto these datasets (see Section~\ref{sec:bootstrap-CIs}). Standard
-%        deviations for each inferred parameter were computed from these 200
-%        model fits. Parameters are as labeled and described in the above
-%        tables. Blank standard errors indicate that parameter was not relevant
-%        to the given model.
-%    }
-%    \centering
-%    \begin{tabular}[t]{l|cc|cc|cc|cc}
-%        \toprule
-%        & \multicolumn{2}{c|}{Single origin} & \multicolumn{2}{c|}{Cont. mig.}
-%        & \multicolumn{2}{c|}{Merger w/o mig.} & \multicolumn{2}{c}{Merger w/mig.} \\
-%        Parameter & GIM & BS & GIM & BS & GIM & BS & GIM & BS \\
-%        \midrule
-%        $N_e$ & 403 & 127 & 1777 & 951 & 326 & 406 & 1369 & 1408 \\
-%        $N_{MH}$ & 529 & 158 & & & & & & \\
-%        $N_{stem1}$ & & & 1612 & 406 & 76 & 122 & 838 & 164 \\
-%        $N_{stem2}$ & & & 2488 & 1001 & 1149 & 2344 & 6668 & 1154 \\
-%        $N_{Nama_0}$ & 370 & 79 & 2989 & 512 & 384 & 496 & 1514 & 538 \\
-%        $N_{Nama_F}$ & 9 & 2 & 54 & 4 & 7 & 4 & 31 & 5 \\
-%        $N_{MSL_0}$ & 769 & 190 & 2479 & 285 & 322 & 384 & 1165 & 421 \\
-%        $N_{MSL_F}$ & 606 & 185 & 6628 & 1250 & 987 & 1031 & 4332 & 1404 \\
-%        $N_{EA}$ & 273 & 49 & 1841 & 138 & 246 & 154 & 1628 & 252 \\
-%        $N_{Gumuz_F}$ & 131 & 56 & 915 & 80 & 102 & 72 & 337 & 83 \\
-%        $N_{EP}$ & 491 & 106 & 3246 & 168 & 408 & 177 & 1684 & 185 \\
-%        $N_{GBR_0}$ & 33 & 7 & 231 & 17 & 29 & 17 & 122 & 21 \\
-%        $N_{GBR_F}$ & 507 & 117 & 2889 & 295 & 334 & 203 & 1678 & 218 \\
-%        $N_{Neand}$ & 105 & 88 & 591 & 97 & 91 & 94 & 235 & 158 \\
-%        $T_{stems}$ & & & 391E3 & 67E3 & 27E3 & 7.6E3 & 426E3 & 284E3 \\
-%        $T_{stem1}$ & & & & & 20E3 & 45E3 & 166E3 & 58E3 \\
-%        $m_{Nama-MSL}$ & 1.6E-6 & 1.2E-6 & 3.7E-6 & 2.7E-6 & 1.1E-6 & 1.3E-6 & 4.0E-6 & 2.6E-6 \\
-%        $m_{Nama-EA}$ & 2.0E-6 & 1.3E-6 & 1.0E-5 & 1.8E-6 & 1.3E-6 & 1.8E-6 & 9.1E-6 & 2.8E-6 \\
-%        $m_{MSL-EA}$ & 7.6E-6 & 1.1E-6 & 5.3E-5 & 3.0E-6 & 5.5E-6 & 3.3E-6 & 2.6E-5 & 4.4E-6 \\
-%        $m_{EA-GBR}$ & 2.4E-6 & 1.1E-6 & 1.0E-5 & 1.9E-6 & 1.5E-6 & 2.2E-6 & 5.5E-6 & 2.6E-6 \\
-%        $m_{EA-EA}$ & 1.3E-5 & 7.5E-6 & 8.4E-5 & 1.4E-5 & 1.3E-5 & 1.4E-5 & 5.4E-5 & 1.9E-5 \\
-%        $f_{GBR \rightarrow EP}$ & 0.0039 & 0.0036 & 0.0037 & 0.0071 & 0.0037 & 0.0066 & 0.0075 & 0.0084 \\
-%        $f_{EP \rightarrow Nama}$ & 0.0039 & 0.0040 & 0.0043 & 0.0072 & 0.0042 & 0.0068 & 0.0049 & 0.0085 \\
-%        $f_{GBR \rightarrow Nama}$ & 0.0019 & 0.0020 & 0.0021 & 0.0030 & 0.0021 & 0.0030 & 0.0031 & 0.0035 \\
-%        $m_{stems}$ & & & 1.1E-5 & 7.7E-6 & & & 8.7E-5 & 2.5E-5 \\
-%        $m_{stem2-Nama}$ & & & 1.6E-5 & 1.1E-5 & & & & \\
-%        $m_{stem2-MSL}$ & & & 4.2E-5 & 1.7E-5 & & & & \\
-%        $m_{stem2-EA}$ & & & 9.0E-6 & 5.4E-6 & & & & \\
-%        $T_{Nama}$ & 2525 & 1188 & 18E3 & 6400 & 8253 & 4763 & 28E3 & 6283 \\
-%        $f_{stem 2 \rightarrow Nama}$ & & & & & 0.0086 & 0.017 & 0.067 & 0.018 \\
-%        $T_{EA}$ & & & & & 3648 & 2047 & 8865 & 4114 \\
-%        $f_{stem 2 \rightarrow EA}$ & & & & & 0.0074 & 0.015 & 0.059 & 0.021 \\
-%        $T_{MSL}$ & & & & & 570 & 1373 & 641 & 1613 \\
-%        $f_{stem 2 \rightarrow MSL}$ & & & & & 0.0036 & 0.0066 & 0.0085 & 0.0076 \\
-%        \bottomrule
-%    \end{tabular}
-%\end{table}
-
 \newcommand{\specialcell}[2][c]{%

### 2023-01-06 supplement.tex
final continuous wo stem mig best fit values

@@
     \midrule
-    $N_e$ & Ancestral effective population size & 10543 & 10810--11260 \\
-    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 15297 & 13930--15330 \\
-    $N_{stem2}$ & Size of stem 2 lineage & 9156 & 9780--11250 \\
-    $N_{Nama_0}$ & Initial Nama size & 11329 & 10760--11280 \\
-    $N_{Nama_F}$ & Final Nama size & 219 & 216--226 \\
-    $N_{MSL_0}$ & Initial Mende size & 11187 & 10510--11090 \\
-    $N_{MSL_F}$ & Final Mende size & 24056 & 24050--25780 \\
-    $N_{EA}$ & Size of East African branch & 7185 & 6970--7200 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3746 & 2650--3860 \\
-    $N_{EP}$ & East African agriculturist size & 13187 & 12830--13250 \\
-    $N_{GBR_0}$ & Initial British size & 937 & 931--964 \\
-    $N_{GBR_F}$ & Final British size & 11478 & 11320--11760 \\
-    $N_{Neand}$ & Neanderthal size & 2285 & 2140--2450 \\
-    $T_{stems}$ & Stem split time (years) & 459720 & 446600--463300 \\
-    $T_{Nama}$ & Nama split time (years) & 139725 & 137300--146000 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $1.93\times10^{-5}$ & $1.36\times10^{-5}$--$1.81\times10^{-5}$ \\
-    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.43\times10^{-5}$ & $4.08\times10^{-5}$--$4.60\times10^{-5}$ \\
+    $N_e$ & Ancestral effective population size & 10980 & 10810--11260 \\
+    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 13980 & 13930--15330 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 10390 & 9780--11250 \\
+    $N_{Nama_0}$ & Initial Nama size & 11240 & 10760--11280 \\
+    $N_{Nama_F}$ & Final Nama size & 221 & 216--226 \\
+    $N_{MSL_0}$ & Initial Mende size & 10810 & 10510--11090 \\
+    $N_{MSL_F}$ & Final Mende size & 25080 & 24050--25780 \\
+    $N_{EA}$ & Size of East African branch & 7130 & 6970--7200 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3770 & 2650--3860 \\
+    $N_{EP}$ & East African agriculturist size & 13070 & 12830--13250 \\
+    $N_{GBR_0}$ & Initial British size & 946 & 931--964 \\
+    $N_{GBR_F}$ & Final British size & 11560 & 11320--11760 \\
+    $N_{Neand}$ & Neanderthal size & 2290 & 2140--2450 \\
+    $T_{stems}$ & Stem split time (years) & 459700 & 446600--463300 \\
+    $T_{Nama}$ & Nama split time (years) & 145400 & 137300--146000 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $1.46\times10^{-5}$ & $1.36\times10^{-5}$--$1.81\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.22\times10^{-5}$ & $4.08\times10^{-5}$--$4.60\times10^{-5}$ \\
     $m_{MSL-EA}$ & Mende--East Africa migration rate & $20.8\times10^{-5}$ & $20.7\times10^{-5}$--$21.3\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.46\times10^{-5}$ & $4.19\times10^{-5}$--$4.58\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.4\times10^{-5}$ & $34.6\times10^{-5}$--$37.0\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.649 & 0.644--0.660 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.268 & 0.264--0.279 \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.36\times10^{-5}$ & $4.19\times10^{-5}$--$4.58\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.8\times10^{-5}$ & $34.6\times10^{-5}$--$37.0\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.651 & 0.644--0.660 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.271 & 0.264--0.279 \\
     $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.153 & 0.148--0.156 \\
     $m_{stems}$ & Stem 1--stem 2 migration rate & $0$ & fixed \\
-    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $4.33\times10^{-5}$ & $4.91\times10^{-5}$--$5.92\times10^{-5}$ \\
-    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $11.3\times10^{-5}$ & $12.4\times10^{-5}$--$14.4\times10^{-5}$ \\
-    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $2.77\times10^{-5}$ & $3.04\times10^{-5}$--$3.95\times10^{-5}$ \\
+    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $5.63\times10^{-5}$ & $4.91\times10^{-5}$--$5.92\times10^{-5}$ \\
+    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $14.2\times10^{-5}$ & $12.4\times10^{-5}$--$14.4\times10^{-5}$ \\
+    $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $3.82\times10^{-5}$ & $3.04\times10^{-5}$--$3.95\times10^{-5}$ \\
     \bottomrule
@@
     \midrule
-    $N_e$ & Ancestral effective population size & 11300 & 10100--11290 \\
-    $N_{stem1}$ & Size of stem 1 lineage between stem 1--stem 2 split and stem 1E--stem 1S split & 113 & 113--606 \\
-    $N_{stem2}$ & Size of stem 2 lineage & 24780 & 26500--35060 \\
-    $N_{Nama_0}$ & Initial Nama and stem 1S size & 13080 & 11310--12910 \\
+    $N_e$ & Ancestral effective population size & 11410 & 10100--11290 \\
+    $N_{stem1}$ & Size of stem 1 lineage between stem 1--stem 2 split and stem 1E--stem 1S split & 102 & 113--606 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 24720 & 26500--35060 \\
+    $N_{Nama_0}$ & Initial Nama and stem 1S size & 13180 & 11310--12910 \\
     $N_{Nama_F}$ & Final Nama size & 224 & 223--238 \\
-    $N_{MSL_0}$ & Initial Mende size & 11870 & 10280--11630 \\
-    $N_{MSL_F}$ & Final Mende size & 25570 & 25720--30180 \\
-    $N_{EA}$ & Size of East African and stem 1E branch & 9110 & 8680--9140 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3370 & 3270--3470 \\
-    $N_{EP}$ & East African agriculturist size & 13550 & 13120--13690 \\
-    $N_{GBR_0}$ & Initial British size & 927 & 920--963 \\
-    $N_{GBR_F}$ & Final British size & 12050 & 11790--12430 \\
-    $N_{Neand}$ & Neanderthal size & 1940 & 1770--2110 \\
-    $T_{stems}$ & Stem split time (years) & 411700 & 381400--410700 \\
-    $T_{stem1}$ & Stem 1 split time into stem 1E and stem 1S (years) & 355400 & 154300--320400 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.361\times10^{-5}$ & $0.205\times10^{-5}$--$0.697\times10^{-5}$ \\
-    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.04\times10^{-5}$ & $3.90\times10^{-5}$--$4.42\times10^{-5}$ \\
-    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.3\times10^{-5}$ & $19.8\times10^{-5}$--$20.8\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.76\times10^{-5}$ & $3.34\times10^{-5}$--$3.95\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $37.2\times10^{-5}$ & $35.2\times10^{-5}$--$39.6\times10^{-5}$ \\
+    $N_{MSL_0}$ & Initial Mende size & 11760 & 10280--11630 \\
+    $N_{MSL_F}$ & Final Mende size & 25770 & 25720--30180 \\
+    $N_{EA}$ & Size of East African and stem 1E branch & 9190 & 8680--9140 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3360 & 3270--3470 \\
+    $N_{EP}$ & East African agriculturist size & 13590 & 13120--13690 \\
+    $N_{GBR_0}$ & Initial British size & 936 & 920--963 \\
+    $N_{GBR_F}$ & Final British size & 12080 & 11790--12430 \\
+    $N_{Neand}$ & Neanderthal size & 1950 & 1770--2110 \\
+    $T_{stems}$ & Stem split time (years) & 415100 & 381400--410700 \\
+    $T_{stem1}$ & Stem 1 split time into stem 1E and stem 1S (years) & 360700 & 154300--320400 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.523\times10^{-5}$ & $0.205\times10^{-5}$--$0.697\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.09\times10^{-5}$ & $3.90\times10^{-5}$--$4.42\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.5\times10^{-5}$ & $19.8\times10^{-5}$--$20.8\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.74\times10^{-5}$ & $3.34\times10^{-5}$--$3.95\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $30.7\times10^{-5}$ & $35.2\times10^{-5}$--$39.6\times10^{-5}$ \\
     $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.648 & 0.641--0.661 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.258 & 0.245--0.264 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.259 & 0.245--0.264 \\
     $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.152--0.161 \\
     $m_{stems}$ & Stem 1--stem 2 migration rate & $0$ & fixed \\
-    $T_{Nama}$ & Time of Nama merger event & 115400 & 101400--116800 \\
+    $T_{Nama}$ & Time of Nama merger event & 118200 & 101400--116800 \\
     $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.693 & 0.706--0.769 \\
-    $T_{EA}$ & Time of East Africa merger event & 92940 & 94500--10140 \\
-    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.502 & 0.510--0.564 \\
-    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 23490 & 22920--25510 \\
-    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.170 & 0.174--0.196 \\
+    $T_{EA}$ & Time of East Africa merger event & 94300 & 94500--101400 \\
+    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.497 & 0.510--0.564 \\
+    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 23890 & 22920--25510 \\
+    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.171 & 0.174--0.196 \\
     \bottomrule

### 2023-01-07 supplement.tex
final fits and LLs

@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -189,434.
+    29 years. This model gave a log-likelihood of -189,271.
 }
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -126,644. This model allows for
+    29 years. This model gave a log-likelihood of -126,540. This model allows for
     continuous migration between Stem 2 and contemporary populations, but disallows
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -115,500.
+    29 years. This model gave a log-likelihood of -115,280.
 }
@@
     \midrule
-    $N_e$ & Ancestral effective population size & 7270 & 4180--6530 \\
-    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 8260 & 7130--8400 \\
-    $N_{stem2}$ & Size of stem 2 lineage & 13550 & 11330--13580 \\
-    $N_{Nama_0}$ & Initial Nama size & 11940 & 11160--12390 \\
+    $N_e$ & Ancestral effective population size & 6090 & 4180--6530 \\
+    $N_{stem1}$ & Size of stem 1 lineage between Neanderthal and Nama splits & 8080 & 7130--8400 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 12340 & 11330--13580 \\
+    $N_{Nama_0}$ & Initial Nama size & 11860 & 11160--12390 \\
     $N_{Nama_F}$ & Final Nama size & 221 & 215--226 \\
-    $N_{MSL_0}$ & Initial Mende size & 9740 & 9500--10210 \\
-    $N_{MSL_F}$ & Final Mende size & 28150 & 25930--28930 \\
-    $N_{EA}$ & Size of East African branch & 7490 & 7270--7630 \\
+    $N_{MSL_0}$ & Initial Mende size & 9880 & 9500--10210 \\
+    $N_{MSL_F}$ & Final Mende size & 27730 & 25930--28930 \\
+    $N_{EA}$ & Size of East African branch & 7460 & 7270--7630 \\
     $N_{Gumuz_F}$ & Final Gumuz size & 3730 & 3560--3810 \\
-    $N_{EP}$ & East African agriculturist size & 13070 & 12800--13290 \\
-    $N_{GBR_0}$ & Initial British size & 959 & 940--981 \\
-    $N_{GBR_F}$ & Final British size & 11820 & 11490--12090 \\
-    $N_{Neand}$ & Neanderthal size & 2670 & 2380--27950 \\
-    $T_{stems}$ & Stem split time (years) & 1163000 & 1120000--1382000 \\ %%% Outside CI!!!
-    $T_{Nama}$ & Nama split time (years) & 134700 & 127800--139900 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.98\times10^{-5}$ & $0.41\times10^{-5}$--$1.43\times10^{-5}$ \\
-    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.08\times10^{-5}$ & $3.92\times10^{-5}$--$4.56\times10^{-5}$ \\
-    $m_{MSL-EA}$ & Mende--East Africa migration rate & $21.4\times10^{-5}$ & $21.2\times10^{-5}$--$21.8\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.17\times10^{-5}$ & $3.77\times10^{-5}$--$4.33\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $33.6\times10^{-5}$ & $33.0\times10^{-5}$--$36.3\times10^{-5}$ \\
+    $N_{EP}$ & East African agriculturist size & 13110 & 12800--13290 \\
+    $N_{GBR_0}$ & Initial British size & 951 & 940--981 \\
+    $N_{GBR_F}$ & Final British size & 11874 & 11490--12090 \\
+    $N_{Neand}$ & Neanderthal size & 2620 & 2380--27950 \\
+    $T_{stems}$ & Stem split time (years) & 1223000 & 1199000--1382000 \\
+    $T_{Nama}$ & Nama split time (years) & 134200 & 127800--139900 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.92\times10^{-5}$ & $0.41\times10^{-5}$--$1.43\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.11\times10^{-5}$ & $3.92\times10^{-5}$--$4.56\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $21.3\times10^{-5}$ & $21.2\times10^{-5}$--$21.8\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.16\times10^{-5}$ & $3.77\times10^{-5}$--$4.33\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $33.7\times10^{-5}$ & $33.0\times10^{-5}$--$36.3\times10^{-5}$ \\
     $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.642 & 0.639--0.656 \\
     $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.255 & 0.245--0.263 \\
-    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.151--0.160 \\
-    $m_{stems}$ & Stem 1--stem 2 migration rate & $6.43\times10^{-5}$ & $6.14\times10^{-5}$--$7.83\times10^{-5}$ \\
-    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $5.82\times10^{-5}$ & $4.50\times10^{-5}$--$7.04\times10^{-5}$ \\
-    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $16.4\times10^{-5}$ & $13.8\times10^{-5}$--$18.9\times10^{-5}$ \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.155 & 0.151--0.160 \\
+    $m_{stems}$ & Stem 1--stem 2 migration rate & $6.26\times10^{-5}$ & $6.14\times10^{-5}$--$7.83\times10^{-5}$ \\
+    $m_{stem2-Nama}$ & Stem 2--Nama migration rate & $5.85\times10^{-5}$ & $4.50\times10^{-5}$--$7.04\times10^{-5}$ \\
+    $m_{stem2-MSL}$ & Stem 2--Mende migration rate & $16.2\times10^{-5}$ & $13.8\times10^{-5}$--$18.9\times10^{-5}$ \\
     $m_{stem2-EA}$ & Stem 2--East Africa migration rate & $3.10\times10^{-5}$ & $2.18\times10^{-5}$--$3.65\times10^{-5}$ \\
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -107,652.
+    29 years. This model gave a log-likelihood of -107,694.
 }
@@
     \midrule
-    $N_e$ & Ancestral effective population size & 11410 & 10100--11290 \\
-    $N_{stem1}$ & Size of stem 1 lineage between stem 1--stem 2 split and stem 1E--stem 1S split & 102 & 113--606 \\
-    $N_{stem2}$ & Size of stem 2 lineage & 24720 & 26500--35060 \\
-    $N_{Nama_0}$ & Initial Nama and stem 1S size & 13180 & 11310--12910 \\
-    $N_{Nama_F}$ & Final Nama size & 224 & 223--238 \\
-    $N_{MSL_0}$ & Initial Mende size & 11760 & 10280--11630 \\
-    $N_{MSL_F}$ & Final Mende size & 25770 & 25720--30180 \\
-    $N_{EA}$ & Size of East African and stem 1E branch & 9190 & 8680--9140 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3360 & 3270--3470 \\
-    $N_{EP}$ & East African agriculturist size & 13590 & 13120--13690 \\
-    $N_{GBR_0}$ & Initial British size & 936 & 920--963 \\
-    $N_{GBR_F}$ & Final British size & 12080 & 11790--12430 \\
-    $N_{Neand}$ & Neanderthal size & 1950 & 1770--2110 \\
-    $T_{stems}$ & Stem split time (years) & 415100 & 381400--410700 \\
-    $T_{stem1}$ & Stem 1 split time into stem 1E and stem 1S (years) & 360700 & 154300--320400 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.523\times10^{-5}$ & $0.205\times10^{-5}$--$0.697\times10^{-5}$ \\
-    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.09\times10^{-5}$ & $3.90\times10^{-5}$--$4.42\times10^{-5}$ \\
-    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.5\times10^{-5}$ & $19.8\times10^{-5}$--$20.8\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.74\times10^{-5}$ & $3.34\times10^{-5}$--$3.95\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $30.7\times10^{-5}$ & $35.2\times10^{-5}$--$39.6\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.648 & 0.641--0.661 \\
-    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.259 & 0.245--0.264 \\
+    $N_e$ & Ancestral effective population size & 10990 & 10100--11290 \\
+    $N_{stem1}$ & Size of stem 1 lineage between stem 1--stem 2 split and stem 1E--stem 1S split & 278 & 113--606 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 27850 & 26500--35060 \\
+    $N_{Nama_0}$ & Initial Nama and stem 1S size & 12640 & 11310--12910 \\
+    $N_{Nama_F}$ & Final Nama size & 227 & 223--238 \\
+    $N_{MSL_0}$ & Initial Mende size & 1190 & 10280--11630 \\
+    $N_{MSL_F}$ & Final Mende size & 27440 & 25720--30180 \\
+    $N_{EA}$ & Size of East African and stem 1E branch & 9016 & 8680--9140 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3440 & 3270--3470 \\
+    $N_{EP}$ & East African agriculturist size & 13680 & 13120--13690 \\
+    $N_{GBR_0}$ & Initial British size & 922 & 920--963 \\
+    $N_{GBR_F}$ & Final British size & 12020 & 11790--12430 \\
+    $N_{Neand}$ & Neanderthal size & 2100 & 1770--2110 \\
+    $T_{stems}$ & Stem split time (years) & 400300 & 381400--410700 \\
+    $T_{stem1}$ & Stem 1 split time into stem 1E and stem 1S (years) & 281600 & 154300--320400 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.229\times10^{-5}$ & $0.205\times10^{-5}$--$0.697\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.15\times10^{-5}$ & $3.90\times10^{-5}$--$4.42\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.9\times10^{-5}$ & $19.8\times10^{-5}$--$20.8\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.99\times10^{-5}$ & $3.34\times10^{-5}$--$3.95\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.5\times10^{-5}$ & $35.2\times10^{-5}$--$39.6\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.641 & 0.641--0.661 \\
+    $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.257 & 0.245--0.264 \\
     $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.156 & 0.152--0.161 \\
     $m_{stems}$ & Stem 1--stem 2 migration rate & $0$ & fixed \\
-    $T_{Nama}$ & Time of Nama merger event & 118200 & 101400--116800 \\
-    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.693 & 0.706--0.769 \\
-    $T_{EA}$ & Time of East Africa merger event & 94300 & 94500--101400 \\
-    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.497 & 0.510--0.564 \\
-    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 23890 & 22920--25510 \\
-    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.171 & 0.174--0.196 \\
+    $T_{Nama}$ & Time of Nama merger event & 112300 & 101400--116800 \\
+    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.731 & 0.706--0.769 \\
+    $T_{EA}$ & Time of East Africa merger event & 95000 & 94500--101400 \\
+    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.522 & 0.510--0.564 \\
+    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 23510 & 22920--25510 \\
+    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.183 & 0.174--0.196 \\
     \bottomrule
@@
     Inferred values are scaled to physical units assuming a generation time of
-    29 years. This model gave a log-likelihood of -102,633.
-    \aprcomment{check these at end for each model}
+    29 years. This model gave a log-likelihood of -101,618.
     \aprcomment{if the slightly shift, also fix elsewhere in supp and main text}
@@
     \midrule
-    $N_e$ & Ancestral effective population size & 10960 & 64500--11910 \\
+    $N_e$ & Ancestral effective population size & 9640 & 6410--11910 \\
     $N_{stem1}$ & Size of stem 1 lineage between Neanderthal split and stem 1E--stem 1S split & 100 & 100--851 \\
-    $N_{stem2}$ & Size of stem 2 lineage & 24100 & 22130--26340 \\
-    $N_{Nama_0}$ & Initial Nama size & 13530 & 11500--13560 \\
+    $N_{stem2}$ & Size of stem 2 lineage & 24210 & 22130--26340 \\
+    $N_{Nama_0}$ & Initial Nama size & 13440 & 11500--13560 \\
     $N_{Nama_F}$ & Final Nama size & 219 & 214--231 \\
-    $N_{MSL_0}$ & Initial Mende size & 11500 & 9690--11290 \\
-    $N_{MSL_F}$ & Final Mende size & 27390 & 26820--32280 \\
-    $N_{EA}$ & Size of East African Branch & 9150 & 7970--8860 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3390 & 3270--3570 \\
-    $N_{EP}$ & East African agriculturist size & 13500 & 12780--13500 \\
-    $N_{GBR_0}$ & Initial British size & 950 & 928--1002 \\
-    $N_{GBR_F}$ & Final British size & 12370 & 11850--12600 \\
-    $N_{Neand}$ & Neanderthal size & 2510 & 2100--2730 \\
-    $T_{stems}$ & Stem split time (years) & 1475000 & 695000--1785000 \\
-    $T_{stem1}$ & Stem 1S--stem 1E split time (years) & 479100 & 275600--478300 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $1.06\times10^{-5}$ & $0.66\times10^{-5}$--$1.57\times10^{-5}$ \\
-    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.49\times10^{-5}$ & $4.45\times10^{-5}$--$5.52\times10^{-5}$ \\
-    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.6\times10^{-5}$ & $20.0\times10^{-5}$--$21.7\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.85\times10^{-5}$ & $3.25\times10^{-5}$--$4.27\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $36.0\times10^{-5}$ & $33.7\times10^{-5}$--$40.9\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.639 & 0.636--0.667 \\
+    $N_{MSL_0}$ & Initial Mende size & 11380 & 9690--11290 \\
+    $N_{MSL_F}$ & Final Mende size & 27760 & 26820--32280 \\
+    $N_{EA}$ & Size of East African Branch & 9100 & 7970--8860 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3410 & 3270--3570 \\
+    $N_{EP}$ & East African agriculturist size & 13510 & 12780--13500 \\
+    $N_{GBR_0}$ & Initial British size & 958 & 928--1002 \\
+    $N_{GBR_F}$ & Final British size & 12300 & 11850--12600 \\
+    $N_{Neand}$ & Neanderthal size & 2520 & 2100--2730 \\
+    $T_{stems}$ & Stem split time (years) & 1692000 & 695000--1785000 \\
+    $T_{stem1}$ & Stem 1S--stem 1E split time (years) & 478900 & 275600--478300 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $0.98\times10^{-5}$ & $0.66\times10^{-5}$--$1.57\times10^{-5}$ \\
+    $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.48\times10^{-5}$ & $4.45\times10^{-5}$--$5.52\times10^{-5}$ \\
+    $m_{MSL-EA}$ & Mende--East Africa migration rate & $19.7\times10^{-5}$ & $20.0\times10^{-5}$--$21.7\times10^{-5}$ \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $3.82\times10^{-5}$ & $3.25\times10^{-5}$--$4.27\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $35.9\times10^{-5}$ & $33.7\times10^{-5}$--$40.9\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.640 & 0.636--0.667 \\
     $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.259 & 0.237--0.267 \\
-    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.159 & 0.154--0.167 \\
-    $m_{stems}$ & Stem 1--stem 2 migration rate & $11.8\times10^{-5}$ & $3.95\times10^{-5}$--$12.3\times10^{-5}$ \\
-    $T_{Nama}$ & Time of Nama merger event & 121600 & 101300--124600 \\
-    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.692 & 0.694--0.758 \\
-    $T_{EA}$ & Time of East Africa merger event & 96500 & 95400--111000 \\
-    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.511 & 0.514--0.589 \\
-    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 24890 & 22210--26300 \\
-    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.182 & 0.182--0.211 \\
+    $f_{GBR \rightarrow Nama}$ & Ancestry proportion from Europeans to Nama 10 generations ago & 0.158 & 0.154--0.167 \\
+    $m_{stems}$ & Stem 1--stem 2 migration rate & $12.6\times10^{-5}$ & $3.95\times10^{-5}$--$12.3\times10^{-5}$ \\
+    $T_{Nama}$ & Time of Nama merger event & 119500 & 101300--124600 \\
+    $f_{stem 2 \rightarrow Nama}$ & Proportion of stem 2 ancestry making up initial Nama lineage ($1-f$ from stem 1S) & 0.698 & 0.694--0.758 \\
+    $T_{EA}$ & Time of East Africa merger event & 96300 & 95400--111000 \\
+    $f_{stem 2 \rightarrow EA}$ & Proportion of stem 2 ancestry making up initial East Africa lineage ($1-f$ from stem 1E) & 0.515 & 0.514--0.589 \\
+    $T_{MSL}$ & Time of secondary admixture from stem 2 to Mende & 25120 & 22210--26300 \\
+    $f_{stem 2 \rightarrow MSL}$ & Proportion of ancestry from secondary stem 2 admixture to Mende & 0.187 & 0.182--0.211 \\
     \bottomrule

### 2023-01-08 paper.tex
fixed references

@@
 %Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans."
-but also in southern \citep{Hammer2011-bx,Lorente-Galdos2019-vz}
-and central African \citep{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz}
+but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} 
+and central African \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz,Hey2018-pw}.
 populations.

### 2023-01-08 supplement.tex

--- 2023-01-06 supplement.tex
@@
 
-\begin{figure}[ht!]
+\begin{figure}[htb]
     \centering
@@
     \midrule
-    $N_e$ & Ancestral effective population size & 9860 & 9600--10090 \\
-    $N_{MH}$ & Size of human lineage between Neanderthal and Nama splits & 20850 & 20640--21220 \\
+    $N_e$ & Ancestral effective population size & 9910 & 9600--10090 \\
+    $N_{MH}$ & Size of human lineage between Neanderthal and Nama splits & 20890 & 20640--21220 \\
     $N_{Nama_0}$ & Initial Nama size & 10170 & 9950--10250 \\
-    $N_{Nama_F}$ & Final Nama size & 222 & 217--226 \\
-    $N_{MSL_0}$ & Initial Mende size & 17280 & 16770--17490 \\
-    $N_{MSL_F}$ & Final Mende size & 16730 & 16370--17060 \\
+    $N_{Nama_F}$ & Final Nama size & 223 & 217--226 \\
+    $N_{MSL_0}$ & Initial Mende size & 17250 & 16770--17490 \\
+    $N_{MSL_F}$ & Final Mende size & 16760 & 16370--17060 \\
     $N_{EA}$ & Size of East African branch & 7130 & 6940--7130 \\
-    $N_{Gumuz_F}$ & Final Gumuz size & 3800 & 3690--3900 \\
-    $N_{EP}$ & East African agriculturist size & 12990 & 12720--13140 \\
-    $N_{GBR_0}$ & Initial British size & 845 & 836--862 \\
-    $N_{GBR_F}$ & Final British size & 12140 & 11870--12300 \\
+    $N_{Gumuz_F}$ & Final Gumuz size & 3830 & 3690--3900 \\
+    $N_{EP}$ & East African agriculturist size & 13010 & 12720--13140 \\
+    $N_{GBR_0}$ & Initial British size & 843 & 836--862 \\
+    $N_{GBR_F}$ & Final British size & 12150 & 11870--12300 \\
     $N_{Neand}$ & Neanderthal size & 1850 & 1650--1990 \\
-    $T_{Nama}$ & Nama split time (years) & 110900 & 107800--112200 \\
-    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $2.83\times10^{-5}$ & $2.61\times10^{-5}$--$3.02\times10^{-5}$ \\
+    $T_{Nama}$ & Nama split time (years) & 110700 & 107800--112200 \\
+    $m_{Nama-MSL}$ & Nama--Mende symmetric migration rate & $2.82\times10^{-5}$ & $2.61\times10^{-5}$--$3.02\times10^{-5}$ \\
     $m_{Nama-EA}$ & Nama--East Africa symmetric migration rate & $4.98\times10^{-5}$ & $4.82\times10^{-5}$--$5.31\times10^{-5}$ \\
     $m_{MSL-EA}$ & Mende--East Africa migration rate & $18.7\times10^{-5}$ & $18.7\times10^{-5}$--$19.2\times10^{-5}$ \\
-    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.38\times10^{-5}$ & $4.13\times10^{-5}$--$4.54\times10^{-5}$ \\
-    $m_{EA-EA}$ & Intra-East Africa migration rate & $43.8\times10^{-5}$ & $41.0\times10^{-5}$--$44.2\times10^{-5}$ \\
-    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.661 & 0.657--0.671 \\
+    $m_{EA-GBR}$ & East Africa--Europe migration rate & $4.42\times10^{-5}$ & $4.13\times10^{-5}$--$4.54\times10^{-5}$ \\
+    $m_{EA-EA}$ & Intra-East Africa migration rate & $41.4\times10^{-5}$ & $41.0\times10^{-5}$--$44.2\times10^{-5}$ \\
+    $f_{GBR \rightarrow EP}$ & Ancestry proportion of East African agriculturalists from GBR 12ka ($1-f$ from Gumuz) & 0.659 & 0.657--0.671 \\
     $f_{EP \rightarrow Nama}$ & Ancestry proportion from EA pastoralists to Nama 2ka & 0.280 & 0.270--0.286 \\

### 2023-01-08 supplement.tex
bootstrap section in supplement

--- 2023-01-07 supplement.tex
@@
 deeper population divergence than growth model C and reticulation model E.
+
+\subsection{Evaluation of estimated confidence intervals}
+\label{sec:confidence-intervals}
+
+We estimated 95\% confidence intervals (CIs) as reported in
+Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}
+by refitting each model to block-bootstrap replicated datasets. While our
+LD-based inference approach is expected to be unbiased
+\citep{Ragsdale2019-nt,Ragsdale2020-nz}, we tested whether our
+bootstrap-estimated CIs contain the true parameters using simulated data from
+our two highlighted models (the continuous migration model and the merger with
+stem migration model).
+
+To do this, we simulated 500 1Mb regions using the same sample sizes as our
+analyzed data, with per-base mutation and recombination rates $u=r=10^{-8}$,
+using \texttt{msprime} \citep{Baumdicker2022-mj}. From these simulated regions,
+we replicated our bootstrapping procedure by resampling (with replacement) 500
+regions. We repeated this 500 times (to obtain 500 bootstrap replicate
+datasets). We fit the full (non-bootstrapped) simulated dataset and each of the
+500 bootstrap replicates, and compared the inferred parameter values to the
+true simulated values (Figures~\ref{fig:supp-sim-BS-continuous-migration} and
+\ref{fig:supp-sim-BS-merger-with-stem-migration}).
+
+The distribution of refit values for vast majority of parameters overlapped
+with the true simulated value. This included every parameter from the
+continuous migration model, and all of the parameters in the merger with stem
+migration model except for the size, migration rate, and split time of Stem 1
+(which experienced a severe bottleneck in the deep past). Because the true
+simulated parameter for the bottleneck size in Stem 1 was at the lower bound
+($N=100$), this likely is due to the bottleneck re-inferences failing to
+fully converge. Overall, bootstrap-estimated CIs typically contain the true
+parameter, although CIs for parameters that lie at an optimization bound
+should be treated with caution.
+
+Finally, many features of recent human history were consistently parameterized
+across each of our tested models (such as recent migration rates and population
+sizes). We evaluated the consistency of these recent parameters across our
+different models (Figure~\ref{fig:supp-bootstrap-overlap}). In the case that
+the model provides a very poor fit to the data (the single origin model) the
+CIs of some parameters do not overlap with those of the other models (e.g.,
+Mende population sizes, and some migration rates). Many parameters are largely
+consistent across each of the models, showing that inferences of recent history
+are largely robust to the details of population structure in the distant past.
 
@@
     \centering
+    \includegraphics{figures/supp-simulated-continuous_migration-bootstrap-fits.pdf}
+    \caption{
+        \textbf{Evaluation of confidence intervals in the continuous migration
+        model.} From 500 replicate bootstrap datasets, the true (simulated)
+        parameter value lies within the distribution of inferred values from
+        the bootstrap datasets. Orange line: simulated parameter value. Green
+        line: inferred parameter value. Blue histogram: inferred parameter
+        values from 500 bootstrap replicate datasets.
+    }
+    \label{fig:supp-sim-BS-continuous-migration}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-simulated-merger_with_stem_migration-bootstrap-fits.pdf}
+    \caption{
+        \textbf{Evaluation of confidence intervals in the merger with stem
+        migration model.} From 500 replicate bootstrap datasets, the true
+        (simulated) parameter value typically lies within the distribution of
+        inferred values from the bootstrap datasets. Stem 1 experienced a
+        strong bottleneck, at the lower bound of allowed inferred values. In
+        this case, and other parameters directly related to Stem 1,
+        bootstrap-inferred values do not overlap with the true value, although
+        this may be due to incomplete convergence of the optimization at these
+        bounds of parameter space. Orange line: simulated parameter value.
+        Green line: inferred parameter value. Blue histogram: inferred
+        parameter values from 500 bootstrap replicate datasets.
+    }
+    \label{fig:supp-sim-BS-merger-with-stem-migration}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{figures/supp-bootstrap-fits-overlap.pdf}
+    \caption{
+        \textbf{Overlap of bootstrap distributions of inferred parameters
+        across multiple parameterized models.} While the details of deep
+        history differ substantially between the different inferred models, the
+        parameterization of recent history (population sizes, migration rates)
+        are similar. Despite the differences in the deep history, many
+        parameters related to recent history are consistent across all tested
+        models. The single origin model (blue) more often has inferred values
+        distinct from other models, although this model provides a far worse
+        fit to the data than the other models.
+    }
+    \label{fig:supp-bootstrap-overlap}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
     \includegraphics[width=5.75in]{figures/supp-admixture-panels}

### 2023-01-08 paper.tex
edits to main text, formatting

@@
 qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
-(log-likelihood ($LL$) $\approx -189,400$, Table~\ref{tab:supp-single-origin}).
+(log-likelihood ($LL$) $\approx -189,300$, Table~\ref{tab:supp-single-origin}).
 We next explored a suite of parameterized models in which population structure predates the
@@
 agricultural populations \citep{Molinaro2019-rf}, comprising over half of their
-genetic ancestry, estimated to be 64--65\%.  We observe significant gene flow
+genetic ancestry, estimated to be approximately 65\%. We observe significant gene flow
 from the Amhara and Oromo into the Nama, a signal which is likely a proxy for
@@
         migration rates over those intervals. The
-        merger-with-stem-migration model (B, with  $LL=-102,600$) outperformed the
-        continuous-migration model (A, with $LL=-115,500$ ).
+        merger-with-stem-migration model (B, with  $LL=-101,600$) outperformed the
+        continuous-migration model (A, with $LL=-115,300$).
         Colors are used to distinguish overlapping branches and link to
@@
         \textbf{Continuous Model} & & & & & \\
-        $LL= -115,500$ & a & Stem 1, Stem 2 & 1,163 & 6.43e-5 & 1,028 \\
-        (Table~\ref{tab:supp-continuous-migration}) & b & Stem 2, Nama & NA & 5.82e-5 & 130 \\
-        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.64e-4} & 130, 55 \\
-        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.8e-6 & 135, 60 \\
-        & g & Mende, East Africans & 60 & \textbf{2.14e-4} & 60 \\
-        & h & East Africans, British & 50 & 4.17e-5 & 50 \\
-        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.36e-4} & 12 \\
+        $LL= -115,300$ & a & Stem 1, Stem 2 & 1,223 & 6.26e-5 & 1,089 \\
+        (Table~\ref{tab:supp-continuous-migration}) & b & Stem 2, Nama & NA & 5.85e-5 & 129 \\
+        & c, d & Stem 2, Other Africans* & NA & 3.10e-5, \textbf{1.62e-4} & 129, 55 \\
+        & e, f & Nama, Other Africans* & 135 & 4.1e-5, 9.2e-6 & 134, 60 \\
+        & g & Mende, East Africans & 60 & \textbf{2.13e-4} & 60 \\
+        & h & East Africans, British & 50 & 4.16e-5 & 50 \\
+        & i & Gumuz, Amhara/Oromo & 12 &\textbf{3.37e-4} & 12 \\
         \textbf{Merger Model} & & & & & \\
-        $LL= -102,600$ & a & Stem 1, Stem 2 & 1,442 & \textbf{1.16e-4} & 963 \\
+        $LL= -101,600$ & a & Stem 1, Stem 2 & 1,692 & \textbf{1.26e-4} & 1,213 \\
         (Table~\ref{tab:supp-merger-with-stem-migration}) & -- & Stem 1S, Stem 1E & 479 & 0 (fixed) & -- \\
-        & b & Stem 2 to Nama & 119 & \textbf{0.71} & pulse \\
-        & c & Stem 2 to Stem 1E & 98 & \textbf{0.50} & pulse \\
-        & d & Stem 2 to Mende & 25 & \textbf{0.18} & pulse \\
-        & e, f & Nama, Other Africans* & 119 & 4.4e-5, 7.1e-6 & 119, 60 \\
-        & g & Mende, East Africans* & 60 & \textbf{1.98e-4} & 60 \\
-        & h & East Africans, British & 50 & 3.87e-5 & 50 \\
+        & b & Stem 2 to Nama & 119 & \textbf{0.70} & pulse \\
+        & c & Stem 2 to Stem 1E & 98 & \textbf{0.52} & pulse \\
+        & d & Stem 2 to Mende & 25 & \textbf{0.19} & pulse \\
+        & e, f & Nama, Other Africans* & 119 & 4.5e-5, 9.8e-6 & 120, 60 \\
+        & g & Mende, East Africans* & 60 & \textbf{1.97e-4} & 60 \\
+        & h & East Africans, British & 50 & 3.82e-5 & 50 \\
         & i & Gumuz, Amhara/Oromo &12 & \textbf{3.59e-4} & 12 \\
@@
 improved the fits relative to zero migration between stems
-($LL \approx -102,600$ vs. $-107,700$ in the merger model and
-$LL \approx -115,500$ vs. $-126,600$ in the continuous migration model).
+($LL \approx -101,600$ vs. $-107,700$ in the merger model and
+$LL \approx -115,300$ vs. $-126,500$ in the continuous migration model).
 With continuous migration between stems, population
@@
 migrant lineages each generation estimated as
-$m=6.4\times10^{-5}-1.2\times10^{-4}$. For comparison, this is similar
+$m=6.3\times10^{-5}-1.3\times10^{-4}$. For comparison, this is similar
 to inferred migration rates between connected contemporary populations over the
@@
 ($m=1.6\times10^{-4}$) compared to the Nama and East African populations
-($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
+($m=5.9\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
 allowed to occur until 5ka. A
@@
 varying proportions to form the different contemporary groups. 
-We observe a sharp bottleneck in Stem 1 down to $N_e=117$ after the split of
+We observe a sharp bottleneck in Stem 1 down to $N_e=100$ after the split of
 the Neanderthal branch. This represents the lower bound allowed in our
-optimization (i.e. 1\% of the ancestral $N_e$), although the size of this
-bottleneck is poorly constrained (standard error (SE) $=838$). After a long period of
+optimization (i.e. an $N_e$ of 100), although the size of this
+bottleneck is poorly constrained (95\% confidence interval 100--851). After a long period of
 exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
-479ka. The timing of this divergence was also poorly constrained (SE $=166$ka).
+478ka. The timing of this divergence was also poorly constrained (95\% CI 276–-478ka).
 These populations evolve independently until 119ka
-(SE $=28$ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
-proportions 29\%, 71\% respectively. Similarly, Stem 1E and Stem 2 combine in
+(101--125ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
+proportions 30\%, 70\% respectively. Similarly, Stem 1E and Stem 2 combine in
 equal proportions (50\% each) to form the ancestors of the western Africans and
@@
 Out of Africa event). Finally, the Mende receive a large additional pulse of
-gene flow from Stem 2, replacing 18\% of their population 25ka (SE $=0.6$ka).
+gene flow from Stem 2, replacing 19\% of their population 25ka (22--26ka).
 The later Stem 2 contribution to the western African Mende resulted in
@@
 here include reticulation and migration between early human populations rather
-than archaic hominin admixture from long-isolated branches (Figure~\ref{fig:proposed-models}C). 
-Elements of a recent single-origin or African multi-regionalism (Figure~\ref{fig:proposed-models}A,D)
-feature in our best fit models, as indicated in the recent time of contemporary population divergence
+than archaic hominin admixture from long-isolated branches
+(Figure~\ref{fig:proposed-models}C). 
+Elements of a recent single-origin or African multi-regionalism
+(Figure~\ref{fig:proposed-models}A,D)
+feature in our best fit models, as indicated in the recent time of
+contemporary population divergence
 or the gene flow between disparate stems, respectively.
-We caution that we cannot rule out that more complex models involving additional stems, continuous structure, or
+We caution that we cannot rule out that more complex models involving
+additional stems, continuous structure, or
 hybrid models including both weak structure and archaic hominin admixture may
@@
 interpret the weakly structured stem as consistent with a population
-coalescence and fragmentation model \citep{Scerri2019-xg}. Additional African populations such as those 
-from Central Africa, other Khoe-San groups or pre-Holocene ancient DNA samples, could further test our proposed models.
-Given that our samples cover major ancestry components across Africa, we predict that any new admixture signal is likely to be regionally localized.
+coalescence and fragmentation model \citep{Scerri2019-xg}.
+Additional African populations such as those 
+from Central Africa, other Khoe-San groups or pre-Holocene ancient DNA samples,
+could further test our proposed models.
+Given that our samples cover major ancestry components across Africa, we predict
+that any new admixture signal is likely to be regionally localized.
 
@@
 
-Our inferred models paint a consistent picture of the Middle to Late Pleistocene as a 
-critical period of change, assuming that estimates from the recombination clock accurately relate to geological chronologies
-(SI Section~\ref{sec:supp-mutation}). 
-During the late Middle Pleistocene, the multiple merger model indicates three
-major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
-eastern (Stem 1E) and western/central Africa (Stem 2). Geographic association was informed by 
-the present population location with the greatest ancestry contribution from each stem. For example,
-Stem 1S contributes 71\% to the ancestral formation of the Khoe-San. The extent of the isolation \>400ka
-between Stem 1S, Stem 1E and Stem 2 suggests that these stems were not proximate to each other. While the length of
-isolation among the stems is variable across fits, models with a period of 
+Our inferred models paint a consistent picture of the Middle to Late
+Pleistocene as a critical period of change, assuming that estimates from the
+recombination clock accurately relate to geological chronologies (SI
+Section~\ref{sec:supp-mutation}). During the late Middle Pleistocene, the
+multiple merger model indicates three major stem lineages in Africa,
+tentatively assigned to southern (Stem 1S), eastern (Stem 1E) and
+western/central Africa (Stem 2). Geographic association was informed by the
+present population location with the greatest ancestry contribution from each
+stem. For example, Stem 1S contributes 70\% to the ancestral formation of the
+Khoe-San. The extent of the isolation \>400ka between Stem 1S, Stem 1E and Stem
+2 suggests that these stems were not proximate to each other. While the length
+of isolation among the stems is variable across fits, models with a period of
 divergence, isolation and then a merger event (i.e. a ``reticulation'')
@@
 geographic expansion of one or both stems. For example, if during MIS 5, either
-Stem 1S (Figure~\ref{fig:best-fit-models}B) 
-from southern Africa moved northward thereby encountering Stem 2,
-or Stem 2 moved from central/western Africa southward into Stem 1S -- then we
-could observe disproportionate ancestry contributions from different stems in
-contemporary groups. We observed two merger events. The first, between Stem 1S and
-Stem 2, results in the formation of an ancestral Khoe-San population $\approx120$ka.
-The second event $\approx100$ka between Stem 1E and Stem 2,
-results in the formation of the
-ancestors of Eastern/Western Africans including the ancestors of ``Out of Africans''. The rapid
-rise in sea levels and increased precipitation during MIS 5e interglacial, following a
-glacial period of aridity across Africa \citep{Blome2012-lw}, might have
-triggered migration inland away from the coasts, as has been suggested, e.g., for the
-Paleo-Agulhas plain \citep{Marean2014-pg}.
+Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
+northward thereby encountering Stem 2, or Stem 2 moved from central/western
+Africa southward into Stem 1S -- then we could observe disproportionate
+ancestry contributions from different stems in contemporary groups. We observed
+two merger events. The first, between Stem 1S and Stem 2, results in the
+formation of an ancestral Khoe-San population $\approx120$ka. The second event
+$\approx100$ka between Stem 1E and Stem 2, results in the formation of the
+ancestors of Eastern/Western Africans including the ancestors of ``Out of
+Africans''. The rapid rise in sea levels and increased precipitation during MIS
+5e interglacial, following a glacial period of aridity across Africa
+\citep{Blome2012-lw}, might have triggered migration inland away from the
+coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
+\citep{Marean2014-pg}.
 
@@
 subpopulations which then persist over the past $\sim$120ka. These
-subpopulations can be linked to contemporary groups despite subsequent gene flow
-across the continent; for example, a genetic lineage sampled in the Gumuz has a
-$0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
-(Stem 1E) $150$ka versus $0.03$ probability of being inherited from the ‘southern’
-subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2
-(see Table~\ref{tab:supp-lineage-probabilitys} for additional comparisons). We also
-find that Stem 2 continued to contribute to western Africans during the Last
-Glacial Period, indicative that this gene flow likely occurred in
-western/central Africa (Table~\ref{tab:migration-rates}). Such an interpretation is reinforced
-by the differential migration rates between regions, i.e. the gene flow from Stem 2 to western Africans is 5x
-that of the rate to eastern Africans during this period. We performed a variety of validation tests
-to explore sensitivity of our assumptions, including relaxing fixed parameters (SI Section~\ref{sec:validation_strategy}).
-Most validation tests resulted in parameters similar to the models discussed above. However, one exception
-was the inferred Out of Africa and Eastern/Western divergence which were 10-15ka younger than our fixed parameters.
-Complexity of this event warrants additional Eurasian populations and greater detail; nonetheless, our model mis-specification
-of this event does not change other features of the merger model.
+subpopulations can be linked to contemporary groups despite subsequent gene
+flow across the continent; for example, a genetic lineage sampled in the Gumuz
+has a $0.44$ probability of being inherited from the ancestral ``eastern''
+subpopulation (Stem 1E) $150$ka versus $0.03$ probability of being inherited
+from the ``southern'' subpopulation (Stem 1S) and $0.53$ probability of being
+inherited from Stem 2 (see Table~\ref{tab:supp-lineage-probabilitys} for
+additional comparisons). We also find that Stem 2 continued to contribute to
+western Africans during the Last Glacial Period, indicative that this gene flow
+likely occurred in western/central Africa (Table~\ref{tab:migration-rates}).
+Such an interpretation is reinforced by the differential migration rates
+between regions, i.e. the gene flow from Stem 2 to western Africans is 5x that
+of the rate to eastern Africans during this period. We performed a variety of
+validation tests to explore sensitivity of our assumptions, including relaxing
+fixed parameters (SI Section~\ref{sec:validation_strategy}). Most validation
+tests resulted in parameters similar to the models discussed above. However,
+one exception was the inferred Out of Africa and eastern/western African
+divergences which were 10-15ka younger than our fixed parameters. Complexity of
+thes events warrant the inclusion of additional Eurasian populations and
+greater detail in modeling these population expansions; nonetheless, this model
+mis-specification does not change other features of our inferred model.
 
@@
 
-Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
-archaic hominin admixture having also occurred in Africa. Previous 
-work that sought to explain patterns of polymorphism inconsistent with a single-origin model thus 
-focused on archaic hominin admixture as an alternative model. 
-The additional (or "ghost") branches required to fit the data
-are systematically referred to as "archaic" 
-\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td}, 
-and assumed (or inferred) to be deeply diverged.
-We believe that this has strongly oriented the interpretation of these results (e.g., regarding selection \cite{Wall2019-ao} 
-or the fossil record \cite{Grun2020}).
-Here, we have shown that weakly-structured-stem 
-models better captures the apparently inconsistent patterns of polymorphisms. Preferring weakly-structured-stem models over 
-archaic-admixture models potentially has a range of implications.
+Evidence for archaic hominin admixture in Eurasia has bolstered the
+plausibility of archaic hominin admixture having also occurred in Africa.
+Previous work that sought to explain patterns of polymorphism inconsistent with
+a single-origin model thus focused on archaic hominin admixture as an
+alternative model. The additional (or ``ghost'') branches required to fit the
+data are systematically referred to as ``archaic''
+\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
+Lorente-Galdos2019-vz,Durvasula2020-td}, and assumed (or inferred) to be deeply
+diverged. We believe that this has strongly oriented the interpretation of
+these results (e.g., regarding selection \cite{Wall2019-ao} or the fossil
+record \cite{Grun2020-su}). Here, we have shown that a model of a weakly
+structured stem better captures the apparently inconsistent patterns of
+polymorphisms.
 
- First, with a weakly 
-structured stem, there is no need to posit that an archaic hominin population in Africa stayed 
-reproductively isolated from the ancestral human lineage for tens thousands of years before 
-the initiation of gene flow. Instead, there would simply have been continuous or recurrent 
-contact between two or more groups present in Africa.
+Preferring weakly-structured-stem models over archaic-admixture models
+potentially has a range of implications. First, with a weakly structured stem,
+there is no need to posit that an archaic hominin population in Africa stayed
+reproductively isolated from the ancestral human lineage for tens thousands of
+years before the initiation of gene flow. Instead, there would simply have
+been continuous or recurrent contact between two or more groups present in
+Africa.
 
-Second, there is evidence for both deleterious and adaptive archaic-hominin-derived alleles in
-contemporary genomes in the form of a depletion of Neanderthal ancestry in regulatory
-regions \citep{Petr2019-xo} or an increased frequency of archaic-hominin-related
-haplotypes such as at \emph{EPAS1} among Tibetans \citep{Zhang2021-xx}. Under previous
-African archaic hominin admixture models, the estimated 8--10\% introgression rate is
+Second, there is evidence for both deleterious and adaptive
+archaic-hominin-derived alleles in contemporary genomes in the form of a
+depletion of Neanderthal ancestry in regulatory regions \citep{Petr2019-xo} or
+an increased frequency of archaic-hominin-related haplotypes such as at
+\emph{EPAS1} among Tibetans \citep{Zhang2021-xx}. Under previous African
+archaic hominin admixture models, the estimated 8--10\% introgression rate is
 much higher than Neanderthal gene flow, and would have plausibly been fertile
-ground for dramatic selection for or against archaic-hominin-derived haplotypes\citep{Wall2019-ao}. 
-By contrast, adaptation under a weakly
+ground for dramatic selection for or against archaic-hominin-derived
+haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
 structured stem would have occurred continuously over much longer periods.
@@
 
-Third, multiple studies have shown a correspondence between phenotypic differentiation, 
-usually assessed with measurements of the cranium, and genetic differentiation 
-among human populations and between humans and Neanderthals 
+Third, multiple studies have shown a correspondence between phenotypic
+differentiation, usually assessed with measurements of the cranium, and genetic
+differentiation among human populations and between humans and Neanderthals
 \citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI
-Section~\ref{sec:morpho_diff}). This correspondence potentially allows predictions 
-of our model to be related to the fossil record. The fossil record of Africa is 
-sparse during the time period of the stems, but of the fossils that date to this 
-time period, some are fairly similar in morphology to contemporary humans (e.g., 
-Omo 1 from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are 
-similar in some morphological features to contemporary humans (e.g., Irhoud 1 
-from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}), and others are 
-different enough in morphology to have been assigned to species other than 
-\emph{Homo sapiens} (e.g., DH1 from Dinaledi, South Africa 
-\citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
-differences between the stems were comparable to those among contemporary human 
-populations, the most morphologically divergent fossils are unlikely to represent 
-branches that contributed appreciably to contemporary human ancestries.
+Section~\ref{sec:morpho_diff}). This correspondence potentially allows
+predictions of our model to be related to the fossil record. The fossil record
+of Africa is sparse during the time period of the stems, but of the fossils
+that date to this time period, some are fairly similar in morphology to
+contemporary humans (e.g., Omo 1 from Omo Kibish, Ethiopia
+\citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological
+features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco
+\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in
+morphology to have been assigned to species other than \emph{Homo sapiens}
+(e.g., DH1 from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}). If,
+as our model predicts, the genetic differences between the stems were
+comparable to those among contemporary human populations, the most
+morphologically divergent fossils are unlikely to represent branches that
+contributed appreciably to contemporary human ancestries.
 
@@
 generation, including Tommy Carensten, Deepti Gurdasani, and Manj Sandhu. We
-thank Luke Anderson-Trocm\'e and Georgette Femerling for assistance in creating the map in
-Figure~\ref{fig:diversity} and Figure \ref{fig:toymodels}, respectively. This research was supported by an NIH grant R35GM133531 (to
-BMH). The content is solely the responsibility of the authors and does
-not necessarily represent the official views of the National Institutes of
-Health.
+thank Luke Anderson-Trocm\'e and Georgette Femerling for assistance in creating
+the map in Figure~\ref{fig:diversity} and Figure \ref{fig:toymodels},
+respectively. This research was supported by an NIH grant R35GM133531 (to BMH).
+The content is solely the responsibility of the authors and does not
+necessarily represent the official views of the National Institutes of Health.
 

### 2023-01-08 paper.tex
edits to main text

@@
 \citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
-Neanderthal admixture with humans in Eurasia, several recent articles
+Neanderthal admixture with humans in Eurasia, several recent studies
 have shown that introducing an archaic hominin ghost population contributing to 
@@
 single-origin models, mostly in Western Africa
-\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td},
+\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
+Lorente-Galdos2019-vz,Durvasula2020-td},
 %Plagnol: YRI 
@@
 but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} 
-and central African \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz,Hey2018-pw}.
+and central African \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz,Hey2018-pw}
 populations.
@@
 population, possible links to specific fossils, and the possibility of
-finding ancient DNA evidence \citep[e.g.,][]{Hsieh2016-gk}. However, these prior
-articles share two weaknesses. First, they only contrast a single-origin model
+finding ancient DNA evidence\citep{Hsieh2016-gk}. However, these prior
+studies share two weaknesses. First, they only contrast a single-origin model
 with an archaic hominin admixture model, leaving out other plausible models
@@
 (Figure~\ref{fig:diversity}C). 
-While ancient DNA from Eurasia has helped us understand early human history
+While ancient DNA from Eurasia has helped to clarify early human history
 outside of Africa, there is no comparably ancient DNA to elucidate early history
@@
 multi-regional evolution, Figure~\ref{fig:proposed-models}), using 290
-genomes of individuals from southern, eastern, and western Africa as well as Eurasia. By
-including geographically and genetically diverse populations across Africa, we
+genomes of individuals from southern, eastern, and western Africa as well as Eurasia.
+By including geographically and genetically diverse populations across Africa, we
 infer demographic models that explain more features of genetic diversity
-in more
-populations than previously reported. These analyses confirm the inadequacy of
+in more populations than previously reported. These analyses confirm the inadequacy of
 tree-like models and provide an opportunity to directly evaluate a wide range
@@
         We used these conceptual models as starting points to build detailed
-        parameterized demographic models (SI Section~\ref{modelspec})
+        parameterized demographic models (SI Section~\ref{sec:modelspec})
         that were then fit to genetic data. 
@@
 unstructured source followed by migration between populations, without allowing
-for contribution from an African archaic hominin lineage or population structure prior
-to the expansion (Figure~\ref{fig:proposed-models}A or Figure~\ref{fig:proposed-models}D). As expected
-\citep{Ragsdale2019-nt}, this first model was a poor fit to the data
-qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
-(log-likelihood ($LL$) $\approx -189,300$, Table~\ref{tab:supp-single-origin}).
-We next explored a suite of parameterized models in which population structure predates the
-differentiation of contemporary groups (Supporting Information (SI) Section~\ref{modelspec}). 
-Depending on the parameters, these encompassed models allowing for ancestral
-reticulation (such as fragmentation-and-coalescence or meta-population models,
+for contribution from an African archaic hominin lineage or population
+structure prior to the expansion (Figure~\ref{fig:proposed-models}A or
+Figure~\ref{fig:proposed-models}D). As expected \citep{Ragsdale2019-nt}, this
+first model was a poor fit to the data qualitatively
+(Figure~\ref{fig:supp-single-origin-fits}) and quantitatively (log-likelihood
+($LL$) $\approx -189,300$, Table~\ref{tab:supp-single-origin}).  We next
+explored a suite of parameterized models in which population structure predates
+the differentiation of contemporary groups (Supporting Information (SI)
+Section~\ref{sec:modelspec}).  Depending on the parameters, these encompassed
+models allowing for ancestral reticulation (such as
+fragmentation-and-coalescence or meta-population models,
 Figure~\ref{fig:proposed-models}B), archaic hominin admixture
@@
 
-Regardless of the model choice for early epochs, maximum-likelihood inference of human demographic
-history for the last 150ka was remarkably robust.
-In a reticulated model, we use ``divergence'' between populations to mean
-the time of their most recent shared ancestry.
-The earliest divergence among
+Regardless of the model choice for early epochs, maximum-likelihood inference
+of human demographic history for the last 150ka was remarkably robust.  In a
+reticulated model, we use ``divergence'' between populations to mean the time
+of their most recent shared ancestry. The earliest divergence among
 contemporary human populations differentiates the southern African Nama from
@@
 high-likelihood models which we explored did the divergence between Nama and
-other populations exceed $\sim$140ka.
-We conclude that geographic patterns of contemporary \emph{Homo sapiens}
-population structure likely arise during MIS 5.
-Although we find evidence for earlier population structure in Africa (next section),
-contemporary populations cannot be easily mapped onto the more ancient
-`stem' groups as only a small proportion of drift between contemporary populations can
-be attributed to drift between stems (SI Section \ref{sec:f4},
+other populations exceed $\sim$140ka. We conclude that geographic patterns of
+contemporary \emph{Homo sapiens} population structure likely arise during MIS
+5. Although we find evidence for earlier population structure in Africa (next
+section), contemporary populations cannot be easily mapped onto the more
+ancient `stem' groups as only a small proportion of drift between contemporary
+populations can be attributed to drift between stems (SI Section \ref{sec:f4},
 Figures~\ref{fig:predictions} and
@@
 We quantify migration rates among populations after their divergence
-$\approx120ka$. Prior to agripastoralist expansion 5ka, migration between the ancestors
-of the Nama and other groups is an order of magnitude weaker than that observed
-between between western and eastern Africans (Table \ref{tab:migration-rates}).
-All models infer relatively high gene flow between Ethiopia and western Africa
-($m\approx2\times10^{-4}$, the constant proportion of immigrant lineages per
-generation over 60ka). We further find that Back-to-Africa gene flow at the
-beginning of the Holocene primarily affected the ancestors of the Ethiopian
-agricultural populations \citep{Molinaro2019-rf}, comprising over half of their
-genetic ancestry, estimated to be approximately 65\%. We observe significant gene flow
-from the Amhara and Oromo into the Nama, a signal which is likely a proxy for
-the movement of eastern African caprid and cattle pastoralists
+$\sim$120ka. Prior to agripastoralist expansion 5ka, migration between the
+ancestors of the Nama and other groups is an order of magnitude weaker than
+that observed between between western and eastern Africans (Table
+\ref{tab:migration-rates}). All models infer relatively high gene flow between
+Ethiopia and western Africa ($m\approx2\times10^{-4}$, the constant proportion
+of immigrant lineages per generation over 60ka). We further find that
+Back-to-Africa gene flow at the beginning of the Holocene primarily affected
+the ancestors of the Ethiopian agricultural populations
+\citep{Molinaro2019-rf}, comprising over half of their genetic ancestry,
+estimated to be approximately 65\%. We observe significant gene flow from the
+Amhara and Oromo into the Nama, a signal which is likely a proxy for the
+movement of eastern African caprid and cattle pastoralists
 \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
-contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE
-plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped into the
-Khoe-San component which has drifted appreciably from its ancestral eastern
-African source. Colonial period admixture from Europeans into the Nama was
-estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
+contribution 2,000 years ago. While this gene flow is not apparent from the
+ADMIXTURE plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped
+into the Khoe-San component which has drifted appreciably from its ancestral
+eastern African source. Colonial period admixture from Europeans into the Nama
+was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
 (Figure~\ref{fig:diversity}).
@@
 
-To account for population structure prior to 135ka, three of our four models allowed for two or more
-``stem'' populations which could diverge either before or after the Neanderthal
-split. We considered models both with and without migration between these stem
-populations, and in both cases we tested two different types of gene exchange
-during the expansion phase, as illustrated in Figure \ref{fig:mergers}:
-1) One of the stem population expands (splits into
+To account for population structure prior to 135ka, three of our four models
+allowed for two or more ``stem'' populations which could diverge either before
+or after the Neanderthal split. We considered models both with and without
+migration between these stem populations, and in both cases we tested two
+different types of gene exchange during the expansion phase, as illustrated in
+Figure \ref{fig:mergers}: 1) One of the stem population expands (splits into
 contemporary populations), and the other stem population(s) has continuous
@@
 populations expands, with instantaneous pulse (or ``merger'') events from the
-other stem population, so that recent populations are formed by mergers
-of multiple ancestral populations. Depending on parameter values,
-this scenario encompasses archaic hominin introgression and
-fragmentation-and-coalescence models (such as Figure~\ref{fig:proposed-models}B and C).
-For many parameters, confidence intervals based on bootstrapping are relatively narrow
+other stem population, so that recent populations are formed by mergers of
+multiple ancestral populations. Depending on parameter values, this scenario
+encompasses archaic hominin introgression and fragmentation-and-coalescence
+models (such as Figure~\ref{fig:proposed-models}B and C).  For many parameters,
+confidence intervals based on bootstrapping are relatively narrow
 (Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
-reflecting an informative statistical approach. Model assumptions
-have, however, a larger impact on parameter estimates (and thus, real uncertainty).
+reflecting an informative statistical approach. Model assumptions have,
+however, a larger impact on parameter estimates (and thus, real uncertainty).
 To convey model uncertainty, we highlight features of the two inferred models
-with high likelihoods. These are referred to as the
-``multiple-merger'' and the ``continuous-migration'' models. Both allow for
-migration between stem branches, but differ primarily in the timing of the
-early divergence of stem populations and their relative $N_e$
-(Figure~\ref{fig:best-fit-models}). The two models also differ in the mode of divergence, with the multiple-merger featuring a population reticulation during the Middle Pleistocene.
+with high likelihoods. These are referred to as the ``multiple-merger'' and the
+``continuous-migration'' models. Both allow for migration between stem
+branches, but differ primarily in the timing of the early divergence of stem
+populations and their relative $N_e$ (Figure~\ref{fig:best-fit-models}). The
+two models also differ in the mode of divergence, with the multiple-merger
+featuring a population reticulation during the Middle Pleistocene.
 
 Allowing for continuous migration between the stem populations substantially
-improved the fits relative to zero migration between stems
-($LL \approx -101,600$ vs. $-107,700$ in the merger model and
-$LL \approx -115,300$ vs. $-126,500$ in the continuous migration model).
-With continuous migration between stems, population
-structure extends back to 1.1--1.4Ma (Table~\ref{tab:migration-rates}).
-Migration between the stems in these models is moderate, with a fraction of
-migrant lineages each generation estimated as
-$m=6.3\times10^{-5}-1.3\times10^{-4}$. For comparison, this is similar
-to inferred migration rates between connected contemporary populations over the
+improved the fits relative to zero migration between stems ($LL \approx
+-101,600$ vs. $-107,700$ in the merger model and $LL \approx -115,300$ vs.
+$-126,500$ in the continuous migration model).  With continuous migration
+between stems, population structure extends back to up to 1.5Ma
+(Table~\ref{tab:migration-rates}).  Migration between the stems in these models
+is moderate, with a fraction of migrant lineages each generation estimated as
+$m=6.3\times10^{-5}-1.3\times10^{-4}$. For comparison, this is similar to
+inferred migration rates between connected contemporary populations over the
 past 50ka (Table~\ref{tab:migration-rates}). This ongoing (or at least,
 periodic) gene flow qualitatively distinguishes these models from previously
-proposed archaic hominin admixture models (Figure~\ref{fig:proposed-models}C) as
-the early branches remain closely related, and each branch contributes large
+proposed archaic hominin admixture models (Figure~\ref{fig:proposed-models}C)
+as the early branches remain closely related, and each branch contributes large
 amounts to all contemporary populations (Figure~\ref{fig:predictions}).
 Because of this relatedness, only $1\%$ to $4\%$ of genetic differentiation
-among contemporary populations can be traced back
-to this early population structure (SI Section~\ref{sec:f4})
+among contemporary populations can be traced back to this early population
+structure (SI Section~\ref{sec:f4})
 
@@
 ($m=5.9\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
-allowed to occur until 5ka. A
-sampled lineage from the Nama, Mende, and Gumuz have probabilities of being in
-Stem 2 at the time of Stem 1 expansion (135ka) of approximately $0.145$, $0.2$,
-and $0.13$, respectively, though these probabilities change over time,
-precluding the notion of a fixed admixture proportion.
+allowed to occur until 5ka. A sampled lineage from the Nama, Mende, and Gumuz
+have probabilities of being in Stem 2 at the time of Stem 1 expansion (135ka)
+of approximately $0.145$, $0.2$, and $0.13$, respectively, though these
+probabilities change over time, precluding the notion of a fixed admixture
+proportion.
 
 In contrast, under the multiple-merger model, stem populations merge with
-varying proportions to form the different contemporary groups. 
-We observe a sharp bottleneck in Stem 1 down to $N_e=100$ after the split of
-the Neanderthal branch. This represents the lower bound allowed in our
-optimization (i.e. an $N_e$ of 100), although the size of this
-bottleneck is poorly constrained (95\% confidence interval 100--851). After a long period of
-exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
-478ka. The timing of this divergence was also poorly constrained (95\% CI 276–-478ka).
-These populations evolve independently until 119ka
-(101--125ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
-proportions 30\%, 70\% respectively. Similarly, Stem 1E and Stem 2 combine in
-equal proportions (50\% each) to form the ancestors of the western Africans and
-eastern Africans (and thus also all individuals who later disperse during the
-Out of Africa event). Finally, the Mende receive a large additional pulse of
-gene flow from Stem 2, replacing 19\% of their population 25ka (22--26ka).
-The later Stem 2 contribution to the western African Mende resulted in
-better model fits ($\Delta LL \approx 60,000$).
-Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
-central Africa, broadly speaking. The differing proportions in the Nama and
-eastern Africans may also indicate geographic separation of Stem 1S in southern
-Africa and Stem 1E in eastern Africa.
+varying proportions to form the different contemporary groups.  We observe a
+sharp bottleneck in Stem 1 down to $N_e=100$ after the split of the Neanderthal
+branch. This represents the lower bound allowed in our optimization (i.e. an
+$N_e$ of 100), although the size of this bottleneck is poorly constrained (95\%
+confidence interval 100--851). After a long period of exchange with Stem 2,
+Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S'' 478ka. The timing of
+this divergence was also poorly constrained (95\% CI 276–-478ka).  These
+populations evolve independently until 119ka (101--125ka) when Stem 1S and Stem
+2 combine to form the ancestors of the Nama, with proportions 30\%, 70\%
+respectively. Similarly, Stem 1E and Stem 2 combine in equal proportions (50\%
+each) to form the ancestors of the western Africans and eastern Africans (and
+thus also all individuals who later disperse during the Out of Africa event).
+Finally, the Mende receive a large additional pulse of gene flow from Stem 2,
+replacing 19\% (18--21\%) of their population 25ka (22--26ka). The later Stem 2
+contribution to the western African Mende resulted in better model fits
+($\Delta LL \approx 60,000$).  Speculatively, this may indicate that an
+ancestral Stem 2 population occupied western or central Africa, broadly
+speaking. The differing proportions in the Nama and eastern Africans may also
+indicate geographic separation of Stem 1S in southern Africa and Stem 1E in
+eastern Africa.
 
@@
 \end{figure}
+
 To assess robustness of the inferred models to analysis and reference
@@
 
-Previous studies have found support for archaic hominin admixture in Africa using
-two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the conditional SFS
-(cSFS) \citep{Durvasula2020-td}, and reconstruction of gene genealogies
-\citep{Speidel2019-nj}. However, none of these studies considered a weakly
-structured stem. We validated our inferred models with additional independent
-approaches. We find that the observed cSFS (conditioned on the derived allele
-being carried in the Neanderthal sample) is very well-described by the merger
-model (Figures~\ref{fig:validation}A-C and
+Previous studies have found support for archaic hominin admixture in Africa
+using two-locus statistics \citep{Hsieh2016-gk,Ragsdale2019-nt}, the
+conditional SFS (cSFS) \citep{Durvasula2020-td}, and reconstruction of gene
+genealogies \citep{Speidel2019-nj}. However, none of these studies considered a
+weakly structured stem. We validated our inferred models with additional
+independent approaches. We find that the observed cSFS (conditioned on the
+derived allele being carried in the Neanderthal sample) is very well-described
+by the merger model (Figures~\ref{fig:validation}A-C and
 \ref{fig:supp-csfs-single-origin}--\ref{fig:supp-csfs-merger-with-stem-migration}),
-even though this statistic was not used in the fit. Our best-fit models outperform
-archaic hominin admixture models fit directly to the cSFS
-(for example, compare to Figure~1 in
-Durvasula and Sankararaman (2020)\citep{Durvasula2020-td}). 
-Specifically, it is the addition of migration between stems that results in a qualitative improvement 
-of the agreement 
-(compare Figures \ref{fig:supp-csfs-merger-without-stem-migration}  and \ref{fig:supp-csfs-merger-with-stem-migration}).
+even though this statistic was not used in the fit. Our best-fit models
+outperform archaic hominin admixture models fit directly to the cSFS (for
+example, compare to Figure~1 in Durvasula and Sankararaman
+(2020)\citep{Durvasula2020-td}).  Specifically, it is the addition of migration
+between stems that results in a qualitative improvement of the agreement
+(compare Figures \ref{fig:supp-csfs-merger-without-stem-migration}  and
+\ref{fig:supp-csfs-merger-with-stem-migration}).
 

### 2023-01-08 paper.tex
Acknowledgements / accessions #

@@
 
-All sequencing data are available from XXX,
-accession number: YYY.
+Nama sequencing data are available from the European Genome-Phenome Archive (EGA),
+accession number: EGAD00001006198. The African Diversity Reference Panel can be
+found at accession: EGAS00001000960.
 
@@
 the map in Figure~\ref{fig:diversity} and Figure \ref{fig:toymodels},
-respectively. This research was supported by an NIH grant R35GM133531 (to BMH).
+respectively. We thank Nuria Melisa Morales-Garcia for designing Figure \ref{fig:proposed-models} 
+and \ref{fig:best-fit-models}, and extensive data visualization discussion. 
+This research was supported by an NIH grant R35GM133531 (to BMH).
 The content is solely the responsibility of the authors and does not

### 2023-01-08 paper.tex
Divergence Discussion

@@
 ancestors of Eastern/Western Africans including the ancestors of ``Out of
-Africans''. The rapid rise in sea levels and increased precipitation during MIS
+Africans''. While $120$ka can be interpreted as the basal human population divergence,
+we suggest concpetualizing this event as the time of most recent shared ancestry among sampled populations.
+"Population divergence" in population genetics is a complex concept because of the co-estimation of divergence and migration;
+many methods make differing assumptions regarding the degree of a clean split between ancestral groups vs. isolation-with-migration 
+(Figure~\ref{fig:supp-IM-misspecification}). Therefore, wide variation exists in estimates of 
+divergence time in the literature (e.g. \citep{Pagani2015-pz }).
+
+The rapid rise in sea levels and increased precipitation during MIS
 5e interglacial, following a glacial period of aridity across Africa

### 2023-01-08 paper.tex
Update paper.tex

@@
 coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
-\citep{Marean2014-pg}.
-
-Following these merger events, the stems subsequently fracture into
+\citep{Marean2014-pg}. Following these merger events, the stems subsequently fracture into
 subpopulations which then persist over the past $\sim$120ka. These
@@
 flow across the continent; for example, a genetic lineage sampled in the Gumuz
-has a $0.44$ probability of being inherited from the ancestral ``eastern''
-subpopulation (Stem 1E) $150$ka versus $0.03$ probability of being inherited
-from the ``southern'' subpopulation (Stem 1S) and $0.53$ probability of being
-inherited from Stem 2 (see Table~\ref{tab:supp-lineage-probabilitys} for
+has a $0.70$ probability of being inherited from the ancestral ``eastern''
+subpopulation $55$ka versus $0.06$ probability of being inherited
+from the ``southern'' subpopulation (see Table~\ref{tab:supp-lineage-probabilitys} for
 additional comparisons). We also find that Stem 2 continued to contribute to

### 2023-01-08 paper.tex
Minor format and tense edits

@@
 northward thereby encountering Stem 2, or Stem 2 moved from central/western
-Africa southward into Stem 1S -- then we could observe disproportionate
+Africa southward into Stem 1S, then we could observe disproportionate
 ancestry contributions from different stems in contemporary groups. We observed
-two merger events. The first, between Stem 1S and Stem 2, results in the
+two merger events. The first, between Stem 1S and Stem 2, resulted in the
 formation of an ancestral Khoe-San population $\approx120$ka. The second event
-$\approx100$ka between Stem 1E and Stem 2, results in the formation of the
+$\approx100$ka between Stem 1E and Stem 2, resulted in the formation of the
 ancestors of Eastern/Western Africans including the ancestors of ``Out of
 Africans''. While $120$ka can be interpreted as the basal human population divergence,
-we suggest concpetualizing this event as the time of most recent shared ancestry among sampled populations.
-"Population divergence" in population genetics is a complex concept because of the co-estimation of divergence and migration;
-many methods make differing assumptions regarding the degree of a clean split between ancestral groups vs. isolation-with-migration 
-(Figure~\ref{fig:supp-IM-misspecification}). Therefore, wide variation exists in estimates of 
-divergence time in the literature (e.g. \citep{Pagani2015-pz }).
+we suggest concpetualizing this event as the time of most recent
+shared ancestry among sampled populations.
+``Population divergence'' in population genetics is a complex concept because 
+of the co-estimation of divergence time and subsequent migration;
+methods make differing assumptions regarding the degree of a ``clean'' split 
+between descendant groups vs. isolation-with-migration 
+(Figure~\ref{fig:supp-IM-misspecification}).
+Therefore, wide variation exists in estimates of 
+divergence time in the literature (e.g.\citep{Pagani2015-pz}).
 
@@
 coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
-\citep{Marean2014-pg}. Following these merger events, the stems subsequently fracture into
+\citep{Marean2014-pg}. Following these merger events, the stems subsequently fractured into
 subpopulations which then persist over the past $\sim$120ka. These
@@
 Such an interpretation is reinforced by the differential migration rates
-between regions, i.e. the gene flow from Stem 2 to western Africans is 5x that
+between regions, i.e. the gene flow from Stem 2 to western Africans is estimated to be 5x that
 of the rate to eastern Africans during this period. We performed a variety of

### 2023-01-08 paper.tex
Typos

@@
 one exception was the inferred Out of Africa and eastern/western African
-divergences which were 10-15ka younger than our fixed parameters. Complexity of
-thes events warrant the inclusion of additional Eurasian populations and
+divergences which were 10-15ka younger than our fixed parameters. The complexity of
+these events warrants the inclusion of additional Eurasian populations and
 greater detail in modeling these population expansions; nonetheless, this model

### 2023-01-08 supplement.tex
edits to main text, formatting
edits to main text
Small revisions to 3.1.1

@@
 \section{Model specification and fitting}
-\label{modelspec}
+\label{sec:modelspec}
 
@@
 and the inferred split between East and West African populations occurred about
-$10$ka earlier. These recent split dates are at odds with archaeological
+$10$ka earlier. These recent split dates are at odds with paleoanthropological 
 evidence for an earlier Out-of-Africa (OOA) expansion
-\citep[e.g.]{Fu2014-am,Hublin2020-nk,Hajdinjak2021-jo}, with the estimated date
+\citep[e.g.][]{Fu2014-am,Hublin2020-nk,Hajdinjak2021-jo}, with the estimated date
 for Neanderthal admixture into Eurasian populations
@@
 Back-to-Africa) event was not our main focus, we chose to keep this simplified
-model, but fix the East African-Eurasian split at $50$ka
+model, but fix the East African-Eurasian split at $50$ka to be older than 
+archaeological dates for when the OOA reached Europe 
 \citep{Fewlass2020-se,Hublin2020-nk}. This also allowed us to include 
@@
     29 years. This model gave a log-likelihood of -101,618.
-    \aprcomment{if the slightly shift, also fix elsewhere in supp and main text}
 }

### 2023-01-08 paper.tex
Supp fig cite

@@
 or the gene flow between disparate stems, respectively.
+
 We caution that we cannot rule out that more complex models involving
@@
 between descendant groups vs. isolation-with-migration 
-(Figure~\ref{fig:supp-IM-misspecification}).
+(Figure ~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
 Therefore, wide variation exists in estimates of 
@@
 of Africa is sparse during the time period of the stems, but of the fossils
-that date to this time period, some are fairly similar in morphology to
+that date to this time period, some are fairly similar overall in morphology to
 contemporary humans (e.g., Omo 1 from Omo Kibish, Ethiopia

### 2023-01-09 supplement.tex
update with converged merger with stem mig

--- 2023-01-08 supplement.tex
@@
 each fit (Figure~\ref{fig:refit-lls}). The correct model was recovered in the
-vast majority of cases based on log-likelihoods alone
+all of the scenarios based on log-likelihoods alone
 (Table~\ref{tab:confusion-matrix}). We conclude that our inferences are not
@@
         analyses in the main text, including simple (single origin) and complex
-        models. Among 500 fits, we almost exclusively recover the simulated
+        models. Among 500 fits, we exclusively recover the simulated
         model, even when fitting more complex (i.e., more heavily
@@
         Cont. migration & 0 & 500 & 0 & 0 \\
-        Merger w/stem mig. & 0 & 0 & 10 & 490 \\
+        Merger w/stem mig. & 0 & 0 & 0 & 500 \\
         \bottomrule

### 2023-01-09 paper.tex
Small edits from replying to reviewer 4

@@
 improved the fits relative to zero migration between stems ($LL \approx
--101,600$ vs. $-107,700$ in the merger model and $LL \approx -115,300$ vs.
-$-126,500$ in the continuous migration model).  With continuous migration
-between stems, population structure extends back to up to 1.5Ma
+-101,600$ vs.\ $-107,700$ in the merger model, Tables
+\ref{tab:supp-merger-without-stem-migration} and
+\ref{tab:supp-merger-with-stem-migration}, and $LL \approx -115,300$ vs.\
+$-126,500$ in the continuous migration model, Tables
+\ref{tab:supp-continuous-migration-no-mig} and
+\ref{tab:supp-continuous-migration-no-mig}).  With continuous migration between
+stems, population structure extends back to up to 1.5Ma
 (Table~\ref{tab:migration-rates}).  Migration between the stems in these models
@@
 midpoint estimates of rCCR are poor estimates for population divergence, often
-underestimating divergence time by 50\% or greater (e.g., Mende vs. Gumuz
+underestimating divergence time by 50\% or greater (e.g., Mende vs.\ Gumuz
 $\sim$15ka compared to a true divergence of 60ka), and recent migration can
@@
 methods make differing assumptions regarding the degree of a ``clean'' split 
-between descendant groups vs. isolation-with-migration 
+between descendant groups vs.\ isolation-with-migration 
 (Figure ~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).

### 2023-01-09 supplement.tex
Small edits from replying to reviewer 4

@@
 \emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
-parameter space and hone in on the best fit parameters. Initial guesses for
-parameters were chosen from demographically plausible starting points and then
-perturbed to explore space, using gradient descent (on the log of the
-parameters). The best fits from these initial rounds of optimization were then
-chosen as the starting points for optimization using the Powell and/or the
-L-BFGS-B methods (as implemented in the SciPy optimization package
-\citep{Virtanen2020-kr}). This process was repeated with alternating
+parameter space and hone in on the best fit parameters.  We imposed bounds on
+many of the parameters in the fits (most of which did not hit the imposed
+bounds). Population sizes had a lower bound of $N_e=100$, migration rates were
+bounded between 0 and $10^{-3}$, and admixture proportions between 0 and 1.
+Event times were only constrained to occur in the proper order (so that
+populations were required to have positive existence times, for example).
+Initial guesses for parameters were chosen from demographically plausible
+starting points and then perturbed to explore space, using gradient descent (on
+the log of the parameters). The best fits from these initial rounds of
+optimization were then chosen as the starting points for optimization using the
+Powell and/or the L-BFGS-B methods (as implemented in the SciPy optimization
+package \citep{Virtanen2020-kr}). This process was repeated with alternating
 optimization methods until the best fit parameter set converged consistently.
@@
     $N_{GBR_F}$ & Final British size & 11874 & 11490--12090 \\
-    $N_{Neand}$ & Neanderthal size & 2620 & 2380--27950 \\
+    $N_{Neand}$ & Neanderthal size & 2620 & 2380--2750 \\
     $T_{stems}$ & Stem split time (years) & 1223000 & 1199000--1382000 \\
@@
     $N_{Nama_F}$ & Final Nama size & 227 & 223--238 \\
-    $N_{MSL_0}$ & Initial Mende size & 1190 & 10280--11630 \\
+    $N_{MSL_0}$ & Initial Mende size & 11190 & 10280--11630 \\
     $N_{MSL_F}$ & Final Mende size & 27440 & 25720--30180 \\

### 2023-01-09 paper.tex
Interglacial Discussion

@@
 
-The rapid rise in sea levels and increased precipitation during MIS
-5e interglacial, following a glacial period of aridity across Africa
-\citep{Blome2012-lw}, might have triggered migration inland away from the
-coasts, as has been suggested, e.g., for the Paleo-Agulhas plain
-\citep{Marean2014-pg}. Following these merger events, the stems subsequently fractured into
+Shifts in wet/dry conditions across the African continent $140$ka to $100$ka may have
+promoted these merger events between divergent stems. 
+Precipitation does not neatly track interglacial cycles in Africa and heterogeneity across regions may mean 
+that beginning of an arid period in eastern Africa is conversely the start of a wet period in southern Africa\citep{Blome2012-lw}.
+The rapid rise in sea levels during the MIS 5e interglacial might have triggered migration inland away from the
+coasts, as has been suggested, e.g., for the Paleo-Agulhas plain\citep{Marean2014-pg}. 
+Following these merger events, the stems subsequently fractured into
 subpopulations which then persist over the past $\sim$120ka. These
@@
 from the ``southern'' subpopulation (see Table~\ref{tab:supp-lineage-probabilitys} for
-additional comparisons). We also find that Stem 2 continued to contribute to
+additional comparisons).
+
+We also find that Stem 2 continued to contribute to
 western Africans during the Last Glacial Period, indicative that this gene flow

### 2023-01-09 paper.tex
Update paper.tex

@@
 of optimizing a large number of parameters, we fixed several parameters related
-to recent population history so as to focus on more ancient events. These
+to recent population history so as to focus on more ancient events (SI Section {sec:models}). These
 parameters were ones supported by multiple genetic and archaeological studies.

### 2023-01-09 paper.tex
Update paper.tex

--- 2023-01-09 paper.tex
@@
 of optimizing a large number of parameters, we fixed several parameters related
-to recent population history so as to focus on more ancient events. These
+to recent population history so as to focus on more ancient events (SI Section \ref{sec:models}). These
 parameters were ones supported by multiple genetic and archaeological studies.

### 2023-01-09 supplement.tex
clarifications in supp

@@
 Using these simulated regions, we constructed 500 bootstrap replicate datasets
-by sampling with replacement. Each of these 500 replicate sets were fit using
+by sampling genomic regions with replacement. Each of these 500 replicate sets were fit using
 the models reported in the main text, and we compared log-likelihoods across
@@
 simply fitting statistical noise, and the observed differences in
-log-likelihoods indicate a better recapitulation of the true history.
+log-likelihoods indicate different abilities to capture signal present in the data.
 
@@
         \toprule
-        Simulated model & Single origin & Cont. migration & Merger w/out stem mig. & Merger w/stem mig. \\
+        Simulated model$\downarrow$ & Single origin & Cont. migration & Merger w/out stem mig. & Merger w/stem mig. \\
         \midrule

### 2023-01-09 supplement.tex
clarifications in supp

@@
 
+A simulation-based validation of the confidence interval calculation is provided in Section 
+\ref{sec:confidence-intervals}.
+
 \subsection{Optimization using \moments}
@@
 sizes). We evaluated the consistency of these recent parameters across our
-different models (Figure~\ref{fig:supp-bootstrap-overlap}). In the case that
-the model provides a very poor fit to the data (the single origin model) the
-CIs of some parameters do not overlap with those of the other models (e.g.,
-Mende population sizes, and some migration rates). Many parameters are largely
+different models (Figure~\ref{fig:supp-bootstrap-overlap}). Since all models use 
+the same set of bootstrapped simulations, systematic differences between models 
+reflect the effect model specification rather than statistical uncertainty. The effects 
+of statistical uncertainty of model specifications need not be of the same order, but 
+we find it useful to compare the two sources of uncertainties. Because model 
+specification differences are mostly in the distant past, we expect to find more 
+differences caused by model specifications in the distant past.
+Many recent parameters are largely
 consistent across each of the models, showing that inferences of recent history
 are largely robust to the details of population structure in the distant past.
+The model that provides a very poor fit to the data (the single origin model) shows 
+stronger model specification effects even for recent parameters (e.g.,
+Mende population sizes, and some migration rates).
 

### 2023-01-09 supplement.tex
clarifications in supp

@@
 the same set of bootstrapped simulations, systematic differences between models 
-reflect the effect model specification rather than statistical uncertainty. The effects 
-of statistical uncertainty of model specifications need not be of the same order, but 
+reflect the effect of model specification rather than statistical uncertainty. The effects 
+of statistical uncertainty and of model specifications need not be of the same order, but 
 we find it useful to compare the two sources of uncertainties. Because model 

### 2023-01-11 paper.tex
latex issue
formatting consistency

@@
 sapiens} anatomical features were found across the continent
-300-100ka. Archaeological sites from the Middle Stone Age, of which some have
+$300-100$ka. Archaeological sites from the Middle Stone Age, of which some have
 been associated with \emph{Homo sapiens}, are also widely distributed across
@@
 \emph{Homo sapiens} by which populations in many regions of the continent
-contributed to the formation of \emph{Homo sapiens} beginning at least 300ka
+contributed to the formation of \emph{Homo sapiens} beginning at least $300$ka
 \citep{Stringer2016-mj,Scerri2018-nl,Scerri2019-xg}.
@@
 for gene flow from Neanderthals into non-Africans and gauge the relative
-time depth of divergence, assuming Neanderthals diverged 550ka from a common
+time depth of divergence, assuming Neanderthals diverged $550$ka from a common
 stem. We computed one- and two-locus statistics whose expectation within and across 
@@
 Regardless of the model choice for early epochs, maximum-likelihood inference
-of human demographic history for the last 150ka was remarkably robust.  In a
+of human demographic history for the last $150$ka was remarkably robust.  In a
 reticulated model, we use ``divergence'' between populations to mean the time
@@
 contemporary human populations differentiates the southern African Nama from
-other African groups between 110--135ka, with low to moderate levels of
+other African groups between $110-135$ka, with low to moderate levels of
 subsequent gene flow (Table \ref{tab:migration-rates}). In none of the
 high-likelihood models which we explored did the divergence between Nama and
-other populations exceed $\sim$140ka. We conclude that geographic patterns of
+other populations exceed $\sim140$ka. We conclude that geographic patterns of
 contemporary \emph{Homo sapiens} population structure likely arise during MIS
@@
 Fixed parameters included the time of divergence between western and eastern
-African populations, set to 60ka, just prior to the split of Eurasians and East
-Africans set to 50ka. We also fixed the amount of admixture from Neanderthals
+African populations, set to $60$ka, just prior to the split of Eurasians and East
+Africans set to $50$ka. We also fixed the amount of admixture from Neanderthals
 to Europeans directly following the out-of-Africa migration which was set to
-1.5\% at 45ka (SI Section~\ref{sec:models}).
+$1.5\%$ at $45$ka (SI Section~\ref{sec:models}).
 
 We quantify migration rates among populations after their divergence
-$\sim$120ka. Prior to agripastoralist expansion 5ka, migration between the
+$\sim120$ka. Prior to agripastoralist expansion $5$ka, migration between the
 ancestors of the Nama and other groups is an order of magnitude weaker than
@@
 Ethiopia and western Africa ($m\approx2\times10^{-4}$, the constant proportion
-of immigrant lineages per generation over 60ka). We further find that
+of immigrant lineages per generation over $60$ka). We further find that
 Back-to-Africa gene flow at the beginning of the Holocene primarily affected
@@
 \citep{Molinaro2019-rf}, comprising over half of their genetic ancestry,
-estimated to be approximately 65\%. We observe significant gene flow from the
+estimated to be approximately $65\%$. We observe significant gene flow from the
 Amhara and Oromo into the Nama, a signal which is likely a proxy for the
 movement of eastern African caprid and cattle pastoralists
-\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
+\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a $25\%$ ancestry
 contribution 2,000 years ago. While this gene flow is not apparent from the
@@
 eastern African source. Colonial period admixture from Europeans into the Nama
-was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
+was estimated at $15\%$, similar to proportions suggested by \texttt{ADMIXTURE}
 (Figure~\ref{fig:diversity}).
@@
         In both models, the branch ancestral to the Nama shares a common ancestral population
-        with the other African groups $\sim$120--135ka. Following this divergence,
+        with the other African groups $\sim120-135$ka. Following this divergence,
         the population ancestral to other African groups branches into West and East African
-        groups 60ka.
+        groups $60$ka.
         *Migration rates and durations are shown between branches ancestral to 
@@
 
-To account for population structure prior to 135ka, three of our four models
+To account for population structure prior to $135$ka, three of our four models
 allowed for two or more ``stem'' populations which could diverge either before
@@
 inferred migration rates between connected contemporary populations over the
-past 50ka (Table~\ref{tab:migration-rates}). This ongoing (or at least,
+past $50$ka (Table~\ref{tab:migration-rates}). This ongoing (or at least,
 periodic) gene flow qualitatively distinguishes these models from previously
@@
 ($m=5.9\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
-allowed to occur until 5ka. A sampled lineage from the Nama, Mende, and Gumuz
+allowed to occur until $5$ka. A sampled lineage from the Nama, Mende, and Gumuz
 have probabilities of being in Stem 2 at the time of Stem 1 expansion (135ka)
@@
 Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S'' 478ka. The timing of
-this divergence was also poorly constrained (95\% CI 276–-478ka).  These
-populations evolve independently until 119ka (101--125ka) when Stem 1S and Stem
+this divergence was also poorly constrained ($95\%$ CI $276-478$ka).  These
+populations evolve independently until $119$ka ($101-125$ka) when Stem 1S and Stem
 2 combine to form the ancestors of the Nama, with proportions 30\%, 70\%
@@
 Finally, the Mende receive a large additional pulse of gene flow from Stem 2,
-replacing 19\% (18--21\%) of their population 25ka (22--26ka). The later Stem 2
+replacing 19\% (18--21\%) of their population $25$ka ($22-26$ka). The later Stem 2
 contribution to the western African Mende resulted in better model fits
@@
         contemporary genetic differentiation.
-        Most present differentiation dates back to the last 100ka.
+        Most present differentiation dates back to the last $100$ka.
     }
@@
         inference. Across the three populations, ancestral state
-        misidentification was consistently inferred to be 1.5--1.7\% for
+        misidentification was consistently inferred to be $1.5-1.7\%$ for
         intergenic loci (SI \ref{sec:misspecification}). (D, E) We used Relate
@@
 models, including the single-origin model, recapitulate an inferred ancestral
-increase in $N_e$ between 100ka-1Ma extending deeper in the past than the
+increase in $N_e$ between $100$ka--$1$Ma extending deeper in the past than the
 Relate estimate (Figure~\ref{fig:supp-iicr-sim} and SI
@@
 Published estimates of the earliest human divergences with rCCR, which range
-from 150ka-100ka \citep{Bergstrom2021-iw}, may be significantly biased when
+from $150$ka--$100$ka \citep{Bergstrom2021-iw}, may be significantly biased when
 compared to more complex models with gene flow as inferred here. We find that
@@
 underestimating divergence time by 50\% or greater (e.g., Mende vs.\ Gumuz
-$\sim$15ka compared to a true divergence of 60ka), and recent migration can
+$\sim15$ka compared to a true divergence of $60$ka), and recent migration can
 lead to the misordering of divergence events (Figure~\ref{fig:validation}E). We
@@
 stem. For example, Stem 1S contributes 70\% to the ancestral formation of the
-Khoe-San. The extent of the isolation \>400ka between Stem 1S, Stem 1E and Stem
+Khoe-San. The extent of the isolation $\>400$ka between Stem 1S, Stem 1E and Stem
 2 suggests that these stems were not proximate to each other. While the length
@@
 Following these merger events, the stems subsequently fractured into
-subpopulations which then persist over the past $\sim$120ka. These
+subpopulations which then persist over the past $\sim120$ka. These
 subpopulations can be linked to contemporary groups despite subsequent gene

### 2023-01-11 supplement.tex
clarify optimization

@@
 \emph{optimize\_log\_powell}, and \emph{optimize\_lbfgsb} methods to explore
-parameter space and hone in on the best fit parameters.  We imposed bounds on
-many of the parameters in the fits (most of which did not hit the imposed
-bounds). Population sizes had a lower bound of $N_e=100$, migration rates were
+parameter space and hone in on the best fit parameters.  
+We imposed bounds on many of the parameters in the fits.
+This is done primarily to avoid optimisation exploring parts of the parameter space 
+that are unrealistic (e.g., negative migration rates), where the model calculation can 
+be costly (e.g., population splits billions of years ago), and where model assumptions 
+can break down (e.g., very small populations).  Population sizes had a lower bound of 
+$N_e=100$, migration rates were
 bounded between 0 and $10^{-3}$, and admixture proportions between 0 and 1.
@@
 populations were required to have positive existence times, for example).
+Most best-fit parameters were away from these boundaries.
+
 Initial guesses for parameters were chosen from demographically plausible

### 2023-01-13 paper.tex
clarified comment on reticulation which was missing a verb

@@
 Africans''. While $120$ka can be interpreted as the basal human population divergence,
-we suggest concpetualizing this event as the time of most recent
+we suggest conceptualizing this event as the time of most recent
 shared ancestry among sampled populations.
@@
 of the co-estimation of divergence time and subsequent migration;
-methods make differing assumptions regarding the degree of a ``clean'' split 
-between descendant groups vs.\ isolation-with-migration 
+methods assuming clean 
+vs reticulated splits can make reach different conclusions 
 (Figure ~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).

### 2023-01-13 paper.tex
extra word

@@
 of the co-estimation of divergence time and subsequent migration;
-methods assuming clean 
-vs reticulated splits can make reach different conclusions 
+methods assuming clean vs reticulated splits can infer different split dates
 (Figure ~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
-Therefore, wide variation exists in estimates of 
-divergence time in the literature (e.g.\citep{Pagani2015-pz}).
+Therefore, wide variation exists in estimates of divergence time in the literature (e.g.\citep{Pagani2015-pz}))
 

### 2023-01-14 paper.tex
small edits

--- 2023-01-11 paper.tex
@@
 
-We also find that Stem 2 continued to contribute to
-western Africans during the Last Glacial Period, indicative that this gene flow
-likely occurred in western/central Africa (Table~\ref{tab:migration-rates}).
-Such an interpretation is reinforced by the differential migration rates
-between regions, i.e. the gene flow from Stem 2 to western Africans is estimated to be 5x that
-of the rate to eastern Africans during this period. We performed a variety of
-validation tests to explore sensitivity of our assumptions, including relaxing
-fixed parameters (SI Section~\ref{sec:validation_strategy}). Most validation
-tests resulted in parameters similar to the models discussed above. However,
-one exception was the inferred Out of Africa and eastern/western African
-divergences which were 10-15ka younger than our fixed parameters. The complexity of
-these events warrants the inclusion of additional Eurasian populations and
-greater detail in modeling these population expansions; nonetheless, this model
-mis-specification does not change other features of our inferred model.
+We also find that Stem 2 continued to contribute to western Africans during the
+Last Glacial Period, indicative that this gene flow likely occurred in
+western/central Africa (Table~\ref{tab:migration-rates}).  Such an
+interpretation is reinforced by the differential migration rates between
+regions, i.e. the gene flow from Stem 2 to western Africans is estimated to be
+five times that of the rate to eastern Africans during this period. We
+performed a variety of validation tests to explore sensitivity of our
+assumptions, including relaxing fixed parameters (SI
+Section~\ref{sec:validation_strategy}). Most validation tests resulted in
+parameters similar to the models discussed above. However, one exception was
+the inferred Out of Africa and eastern/western African divergences which were
+10-15ka younger than our fixed parameters, which seem at odds with the
+archaeological record \citep{Hublin2020-nk,Hajdinjak2021-jo} and some
+genetic estimates \citep{Schlebusch2017-nl,Kamm2020-vn}. The likely
+complexity of the demographic events that we model as population splits
+warrants the inclusion of additional Eurasian populations and greater modeling
+detail, particularly for the out-of-Africa expansion. However, other features
+of our inferred models are weakly affected by this analysis choice.
 

### 2023-01-14 paper.tex

@@
 of the co-estimation of divergence time and subsequent migration;
-methods make differing assumptions regarding the degree of a ``clean'' split 
-between descendant groups vs.\ isolation-with-migration 
+methods assuming clean vs reticulated splits can infer different split dates
 (Figure ~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
-Therefore, wide variation exists in estimates of 
-divergence time in the literature (e.g.\citep{Pagani2015-pz}).
+Therefore, wide variation exists in estimates of divergence time in the literature (e.g.\citep{Pagani2015-pz}))
 

### 2023-01-14 paper.tex
model clarifications

@@
 Section~\ref{sec:modelspec}).  Depending on the parameters, these encompassed
-models allowing for ancestral reticulation (such as
-fragmentation-and-coalescence or meta-population models,
-Figure~\ref{fig:proposed-models}B), archaic hominin admixture
+models allowing for ancestral reticulation, such as
+fragmentation-and-coalescence or meta-population models
+(Figure~\ref{fig:proposed-models}B), archaic hominin admixture
 (Figure~\ref{fig:proposed-models}C), and African multi-regionalism
-(Figure~\ref{fig:proposed-models}D).
+(Figure~\ref{fig:proposed-models}D). The recent expansion and the African
+multi-regional models (Figure~\ref{fig:proposed-models}A and D) have the same
+topology, so interpretation of the model depends on the specified or inferred
+divergence times.
 

### 2023-01-16 paper.tex
formatting edits throughout and comments

@@
 populations connected by gene flow over
-hundreds of thousands of years (i.e. a weakly structured stem). We find that
+hundreds of thousands of years (i.e., a weakly structured stem). We find that
 weakly structured stem models provide more likely explanations of polymorphism
@@
 Neanderthal admixture with humans in Eurasia, several recent studies
-have shown that introducing an archaic hominin ghost population contributing to 
+have shown that introducing an archaic hominin ``ghost'' population contributing to 
 African populations in the period surrounding the Out-of-Africa migration event
 substantially improves the description of genetic data relative to
-single-origin models, mostly in Western Africa
+single-origin models, mostly in western Africa
 \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
@@
 (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively (log-likelihood
-($LL$) $\approx -189,300$, Table~\ref{tab:supp-single-origin}). We next
+($LL$) $\approx-$189,300, Table~\ref{tab:supp-single-origin}). We next
 explored a suite of parameterized models in which population structure predates
@@
 of optimizing a large number of parameters, we fixed several parameters related
-to recent population history so as to focus on more ancient events (SI Section \ref{sec:models}). These
-parameters were ones supported by multiple genetic and archaeological studies.
-Fixed parameters included the time of divergence between western and eastern
-African populations, set to $60$ka, just prior to the split of Eurasians and East
-Africans set to $50$ka. We also fixed the amount of admixture from Neanderthals
-to Europeans directly following the out-of-Africa migration which was set to
-$1.5\%$ at $45$ka (SI Section~\ref{sec:models}).
+to recent population history so as to focus on more ancient events (SI Section
+\ref{sec:models}). These parameters were ones supported by multiple genetic and
+archaeological studies. Fixed parameters included the time of divergence
+between western and eastern African populations, set to $60$ka, just prior to
+the split of Eurasians and East Africans set to $50$ka. We also fixed the
+amount of admixture from Neanderthals to Europeans directly following the
+out-of-Africa migration to $1.5\%$ at $45$ka.
 
@@
 \ref{tab:migration-rates}). All models infer relatively high gene flow between
-Ethiopia and western Africa ($m\approx2\times10^{-4}$, the constant proportion
+eastern and western Africa ($m\approx2\times10^{-4}$, the constant proportion
 of immigrant lineages per generation over $60$ka). We further find that
@@
 contribution 2,000 years ago. While this gene flow is not apparent from the
-ADMIXTURE plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped
+\texttt{ADMIXTURE} plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped
 into the Khoe-San component which has drifted appreciably from its ancestral
@@
 Figure \ref{fig:mergers}: 1) One of the stem population expands (splits into
-contemporary populations), and the other stem population(s) has continuous
-symmetric migration with those populations; or 2) one or more of the stem
+contemporary populations), followed by continuous symmetric migration with
+the other stem population(s); or 2) one or more of the stem
 populations expands, with instantaneous pulse (or ``merger'') events from the
@@
 (Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
-reflecting an informative statistical approach. Model assumptions have,
-however, a larger impact on parameter estimates (and thus, real uncertainty).
+reflecting an informative statistical approach. However, model assumptions have
+a larger impact on parameter estimates (and thus real uncertainty).
 To convey model uncertainty, we highlight features of the two inferred models
@@
 improved the fits relative to zero migration between stems ($LL \approx
--101,600$ vs.\ $-107,700$ in the merger model, Tables
+-$101,600 vs.\ $-$107,700 in the merger model, Tables
 \ref{tab:supp-merger-without-stem-migration} and
-\ref{tab:supp-merger-with-stem-migration}, and $LL \approx -115,300$ vs.\
-$-126,500$ in the continuous migration model, Tables
+\ref{tab:supp-merger-with-stem-migration}, and $LL \approx -$115,300 vs.\
+$-$126,500 in the continuous migration model, Tables
 \ref{tab:supp-continuous-migration-no-mig} and
@@
 populations evolve independently until $119$ka ($101-125$ka) when Stem 1S and Stem
-2 combine to form the ancestors of the Nama, with proportions 30\%, 70\%
+2 combine to form the ancestors of the Nama, with proportions 30\% and 70\%,
 respectively. Similarly, Stem 1E and Stem 2 combine in equal proportions (50\%
 each) to form the ancestors of the western Africans and eastern Africans (and
-thus also all individuals who later disperse during the Out of Africa event).
+thus also all individuals who later disperse during the Out-of-Africa event).
 Finally, the Mende receive a large additional pulse of gene flow from Stem 2,
@@
 models, including the single-origin model, recapitulate an inferred ancestral
-increase in $N_e$ between $100$ka--$1$Ma extending deeper in the past than the
-Relate estimate (Figure~\ref{fig:supp-iicr-sim} and SI
+increase in $N_e$ between $100$ka--$1$Ma
+%extending deeper in the past than the Relate estimate
+% APR: I don't understand this phrase I commented out
+(Figure~\ref{fig:supp-iicr-sim} and SI
 Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
@@
 Relative cross-coalescence rates (rCCR) have recently been used to estimate
-divergence between a pair of populations, as measured by the rate of
+divergence between pairs of populations, as measured by the rate of
 coalescence between two groups divided by the mean within population
@@
 inconsistent divergence times, some of which are older than those we find here
-\cite{Henn2018-rf,Bergstrom2021-iw}. In the SI
-(Section~\ref{sec:IM-reinfer}), we show that this discrepancy can be explained
-by model misspecification: if divergence is estimated using an isolation with
-migration (IM) model with constant population sizes, but the correct model has
-ancient population growth or population structure, the split time in the
-inferred IM model is much deeper than in the correct model. Intuitively, growth
-or structure in the ancestral population will each increase coalescence times
-relative to a randomly mating population of constant size, so a model that
-assumes constant population sizes would require an older divergence time to fit
-the observed distribution of coalescence times and related statistics
+\cite{Henn2018-rf,Bergstrom2021-iw}. In the SI (Section~\ref{sec:IM-reinfer}),
+we show that this discrepancy can be explained by model misspecification: if
+divergence is estimated using an isolation with migration (IM) model with
+constant population sizes, but the correct model has ancient population growth
+or population structure, the divergence time in the inferred IM model is much
+deeper than in the correct model. Intuitively, growth or structure in the
+ancestral population will each increase coalescence times relative to a
+randomly mating population of constant size, so a model that assumes constant
+population sizes would require an older divergence time to fit the observed
+distribution of coalescence times and related statistics
 \citep{Momigliano2021-th,Shchur2022-wa}.
@@
 We caution that we cannot rule out that more complex models involving
-additional stems, continuous structure, or
+additional stems, more complex population structure, or
 hybrid models including both weak structure and archaic hominin admixture may
@@
 two merger events. The first, between Stem 1S and Stem 2, resulted in the
-formation of an ancestral Khoe-San population $\approx120$ka. The second event
-$\approx100$ka between Stem 1E and Stem 2, resulted in the formation of the
-ancestors of Eastern/Western Africans including the ancestors of ``Out of
-Africans''. While $120$ka can be interpreted as the basal human population divergence,
+formation of an ancestral Khoe-San population $\sim120$ka. The second event
+$\sim100$ka between Stem 1E and Stem 2, resulted in the formation of the
+ancestors of eastern/western Africans including the ancestors of
+``Out-of-Africans''. While $120$ka can be interpreted as the basal
+\aprcomment{maybe ``earliest'' instead of ``basal''?} human population divergence,
 we suggest conceptualizing this event as the time of most recent
-shared ancestry among sampled populations.
-``Population divergence'' in population genetics is a complex concept because 
-of the co-estimation of divergence time and subsequent migration;
+shared ancestry among sampled populations. \aprcomment{discounting very recent admixture..?}
+``Population divergence'' in population genetics is a complex concept due
+to the co-estimation of divergence time and subsequent migration;
 methods assuming clean vs reticulated splits can infer different split dates
-(Figure ~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
-Therefore, wide variation exists in estimates of divergence time in the literature (e.g.\citep{Pagani2015-pz}))
+(Figure~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
+Therefore, wide variation exists in estimates of divergence time in the
+literature (e.g.\citep{Pagani2015-pz}))
 
@@
 promoted these merger events between divergent stems. 
-Precipitation does not neatly track interglacial cycles in Africa and heterogeneity across regions may mean 
-that beginning of an arid period in eastern Africa is conversely the start of a wet period in southern Africa\citep{Blome2012-lw}.
-The rapid rise in sea levels during the MIS 5e interglacial might have triggered migration inland away from the
+Precipitation does not neatly track interglacial cycles in Africa,
+and heterogeneity across regions may mean that the beginning of an arid period
+in eastern Africa is conversely the start of a wet period in
+southern Africa\citep{Blome2012-lw}.
+The rapid rise in sea levels during the MIS 5e interglacial might have
+triggered migration inland away from the
 coasts, as has been suggested, e.g., for the Paleo-Agulhas plain\citep{Marean2014-pg}.
@@
 parameters similar to the models discussed above. However, one exception was
-the inferred Out of Africa and eastern/western African divergences which were
-10-15ka younger than our fixed parameters, which seem at odds with the
+the inferred Out-of-Africa and eastern/western African divergences which were
+10-15ka younger than our fixed parameters, seemingly at odds with the
 archaeological record \citep{Hublin2020-nk,Hajdinjak2021-jo} and some
@@
 warrants the inclusion of additional Eurasian populations and greater modeling
-detail, particularly for the out-of-Africa expansion. However, other features
+detail, particularly for the Out-of-Africa expansion. However, other features
 of our inferred models are weakly affected by this analysis choice.
@@
 there is no need to posit that an archaic hominin population in Africa stayed
-reproductively isolated from the ancestral human lineage for tens thousands of
+reproductively isolated from the ancestral human lineage for hundreds of thousands of
 years before the initiation of gene flow. Instead, there would simply have
@@
 an increased frequency of archaic-hominin-related haplotypes such as at
-\emph{EPAS1} among Tibetans \citep{Zhang2021-xx}. Under previous African
-archaic hominin admixture models, the estimated 8--10\% introgression rate is
+\emph{EPAS1} among Tibetans \citep{Zhang2021-xx}. Under previous
+African-archaic-hominin admixture models, the estimated 8--10\% introgression rate is
 much higher than Neanderthal gene flow, and would have plausibly been fertile
@@
 
-We are grateful for the DNA contribution from each participant which enabled
+We are grateful for the DNA contribution from each participant that enabled
 this study; in particular we wish to highlight the generous participation of
@@
 the map in Figure~\ref{fig:diversity} and Figure \ref{fig:toymodels},
-respectively. We thank Nuria Melisa Morales-Garcia for designing Figure \ref{fig:proposed-models} 
-and \ref{fig:best-fit-models}, and extensive data visualization discussion. 
-This research was supported by an NIH grant R35GM133531 (to BMH).
-The content is solely the responsibility of the authors and does not
-necessarily represent the official views of the National Institutes of Health.
+respectively. We thank Nuria Melisa Morales-Garcia for extensive data
+visualization discussion and designing Figures \ref{fig:proposed-models} and
+\ref{fig:best-fit-models}. This research was supported by an NIH grant
+R35GM133531 (to BMH). The content is solely the responsibility of the authors
+and does not necessarily represent the official views of the National
+Institutes of Health.
 

### 2023-01-16 paper.tex
Fixed validation sentence
Erroneous citations, awk wording, incorrect aspersions. 685-691

@@
 the inferred Out-of-Africa and eastern/western African divergences which were
-10-15ka younger than our fixed parameters, seemingly at odds with the
-archaeological record \citep{Hublin2020-nk,Hajdinjak2021-jo} and some
-genetic estimates \citep{Schlebusch2017-nl,Kamm2020-vn}. The likely
-complexity of the demographic events that we model as population splits
-warrants the inclusion of additional Eurasian populations and greater modeling
-detail, particularly for the Out-of-Africa expansion. However, other features
+10-15ka younger than our fixed parameters. The
+complexity of the OOA expansion warrants the inclusion of additional Eurasian populations 
+and more detailed modeling. However, other features
 of our inferred models are weakly affected by this analysis choice.

### 2023-01-16 paper.tex
minor edits

@@
 could further test our proposed models.
-Given that our samples cover major ancestry components across Africa, we predict
-that any new admixture signal is likely to be regionally localized.
 
@@
 Therefore, wide variation exists in estimates of divergence time in the
-literature (e.g.\citep{Pagani2015-pz}))
+literature (e.g.\citep{Pagani2015-pz}).
 
@@
 the inferred Out-of-Africa and eastern/western African divergences which were
-10-15ka younger than our fixed parameters. The
+10--15ka younger than our fixed parameters. The
 complexity of the OOA expansion warrants the inclusion of additional Eurasian populations 

### 2023-01-17 supplement.tex
update best fit yaml models
Update supplement.tex

@@
 \begin{itemize}
-    \item Neanderthal introgression to Eurasian branch: 45ka
-    \item Mende population expansion and Gumuz population decline: 5ka \citep{Gopalan2022-pw}
-    \item East African pastoralist to South African admixture: 2ka \citep{Uren2016-nn}
-    \item European to South African admixture: 10 generations ago
-    \item South African population decline following colonial admixture: 9 generations ago
+    \item Western African vs. Eastern African divergence: 60ka \citep{Pagani2015-pz}
+%Add Niang et al. "The Middle Stone Age occupations of Tiemassa, coastal West Africa, between 62 and 25 thousand years ago" Journal of Archaeological Science, 2020 as citation
+%Add 
+    \item Eastern Africa vs. British divergence [i.e. OOA]: 50ka
+    \item Neanderthal introgression to Eurasian branch: 45ka \citep{{Reilly2022-ym}
+    \item Mende population expansion: 5ka 
+    \item Gumuz population decline: 5ka
+    \item East African pastoralist to South African admixture: 2ka 
+    \item European to South African admixture: 10 generations ago \citep{Uren2016-nn}
+    \item South African population decline following colonial admixture: 9 generations ago \citep{Van_Eeden2022-od}
 \end{itemize}
-The total number of parameters that were
-ultimately inferred were 21-31, depending on the complexity
+The total number of parameters that were ultimately inferred were 21-31, depending on the complexity
 of the model.
@@
 
-In our preliminary models, the unconstrained inferred split times between East
+In our preliminary models, the unconstrained inferred split times between Eastern
 African populations and Europeans was inferred to have occurred around $40$ka,
-and the inferred split between East and West African populations occurred about
+and the inferred split between Eastern and Western African populations occurred about
 $10$ka earlier. These recent split dates are at odds with paleoanthropological 
@@
 non-independence of sites) does not account for these differences: in the
-high-likelihood models, the $95\%$ confidence intervals for the East
-African-European split was roughly $38 \pm 5$ka, and for the East/West African
+high-likelihood models, the $95\%$ confidence intervals for the Eastern
+African-European split was roughly $38 \pm 5$ka, and for the Eastern/Western African
 split was $45 \pm 10$ka, with the uncertainties reflecting 2 standard
@@
 discrepancies to result from model misspecification in the details of the
-divergence between the ancestors of OOA and East African populations.
+divergence between the ancestors of OOA and Eastern African populations.
 
@@
 \label{sec:validation_results}
-\subsection{Model fit using alternative datasets}
+\subsection{Model fits using alternative datasets}
 \label{sec:supp-alt-data}

### 2023-01-17 supplement.tex
Citation updates for fixed parameters

@@
 %Add Niang et al. "The Middle Stone Age occupations of Tiemassa, coastal West Africa, between 62 and 25 thousand years ago" Journal of Archaeological Science, 2020 as citation
-%Add 
-    \item Eastern Africa vs. British divergence [i.e. OOA]: 50ka
-    \item Neanderthal introgression to Eurasian branch: 45ka \citep{{Reilly2022-ym}
+%Add Chevrier et al. "Between continuity and discontinuity: An overview of the West African Paleolithic over the last 200,000 years". Quaternary International, 2018 as citation
+    \item Eastern Africa vs. British divergence [i.e. OOA]: 50ka \citep{Fu2014-am,Hublin2020-nk}
+%Add Mellars 2011 "The earliest modern humans in Europe", Nature.
+This is a nice commentary and Mellars was a wel-respected synthesizer of the European UP.
+Hajdinjak is not a paleo citation and not direct.
+    \item Neanderthal introgression to Eurasian branch: 45ka \citep{Reilly2022-ym}
     \item Mende population expansion: 5ka
+%Add Tennessen et al. 2012 "Evolution and Functional Impact of Rare Coding Variation from Deep Sequencing of Human Exomes"
     \item Gumuz population decline: 5ka
+    %We actually don't have a citation for this. Leaving blank. 
     \item East African pastoralist to South African admixture: 2ka 
@@
 evidence for an earlier Out-of-Africa (OOA) expansion
-\citep[e.g.][]{Fu2014-am,Hublin2020-nk,Hajdinjak2021-jo}, with the estimated date
+\citep[e.g.][]{Fu2014-am,Hublin2020-nk} and with the estimated date
 for Neanderthal admixture into Eurasian populations
-\citep[e.g.][]{Reilly2022-ym}, and with previous genetics-based dating of
-splits between African and Eurasian populations
-\citep[e.g.]{Gutenkunst2009-ff,Jouganous2017-pq,Kamm2020-vn}. The parameter
+\citep[e.g.][]{Reilly2022-ym}. The parameter
 uncertainty of our inferred divergence times (which accounts for the

### 2023-01-17 supplement.tex
Update supplement.tex
add fixed params citations, format consistency

@@
 \begin{itemize}
-    \item Western African vs. Eastern African divergence: 60ka \citep{Pagani2015-pz}
-%Add Niang et al. "The Middle Stone Age occupations of Tiemassa, coastal West Africa, between 62 and 25 thousand years ago" Journal of Archaeological Science, 2020 as citation
-%Add Chevrier et al. "Between continuity and discontinuity: An overview of the West African Paleolithic over the last 200,000 years". Quaternary International, 2018 as citation
-    \item Eastern Africa vs. British divergence [i.e. OOA]: 50ka \citep{Fu2014-am,Hublin2020-nk}
-%Add Mellars 2011 "The earliest modern humans in Europe", Nature.
-This is a nice commentary and Mellars was a wel-respected synthesizer of the European UP.
-Hajdinjak is not a paleo citation and not direct.
+    \item Western African vs. eastern African divergence: 60ka
+        \citep{Pagani2015-pz,Chevrier2018-vv,Niang2020-xt}
+    \item Eastern Africa vs. British divergence (i.e., OOA): 50ka
+        \citep{Mellars2011-sr,Fu2014-am,Hublin2020-nk}
     \item Neanderthal introgression to Eurasian branch: 45ka \citep{Reilly2022-ym}
-    \item Mende population expansion: 5ka
-%Add Tennessen et al. 2012 "Evolution and Functional Impact of Rare Coding Variation from Deep Sequencing of Human Exomes"
+    \item Mende population expansion: 5ka \citep{Tennessen2012-oy}
     \item Gumuz population decline: 5ka
     %We actually don't have a citation for this. Leaving blank. 
-    \item East African pastoralist to South African admixture: 2ka 
+    \item East African pastoralist to South African admixture: 2ka \citep{Henn2008-xo}
     \item European to South African admixture: 10 generations ago \citep{Uren2016-nn}
@@
 
-In our preliminary models, the unconstrained inferred split times between Eastern
+In our preliminary models, the unconstrained inferred split times between eastern
 African populations and Europeans was inferred to have occurred around $40$ka,
-and the inferred split between Eastern and Western African populations occurred about
+and the inferred split between eastern and western African populations occurred about
 $10$ka earlier. These recent split dates are at odds with paleoanthropological 
@@
 non-independence of sites) does not account for these differences: in the
-high-likelihood models, the $95\%$ confidence intervals for the Eastern
-African-European split was roughly $38 \pm 5$ka, and for the Eastern/Western African
+high-likelihood models, the $95\%$ confidence intervals for the eastern
+African-European split was roughly $38 \pm 5$ka, and for the eastern/western African
 split was $45 \pm 10$ka, with the uncertainties reflecting 2 standard
@@
 discrepancies to result from model misspecification in the details of the
-divergence between the ancestors of OOA and Eastern African populations.
+divergence between the ancestors of OOA and eastern African populations.
 
@@
 different West African population (the Yoruba from Ibidan, Nigeria, YRI), and
-two different European populations (Western Europeans from Utah, CEU, and
+two different European populations (western Europeans from Utah, CEU, and
 Toscani in Italy, TSI). The CEU are closely related to GBR and are thus likely
@@
 We replaced the Mende (MSL) and British (GBR) population sets with Yoruba (YRI)
-and Western and Northern Europeans (CEU), respectively, and refit our inferred
+and western and northern Europeans (CEU), respectively, and refit our inferred
 demographic models to diversity and LD statistics computed from those
@@
 constrained by historical records (Section~\ref{sec:models}), including the
-Eurasian out-of-Africa expansion at 50ka and the divergence of Western and
-Eastern African groups at 60ka. To evaluate the effects of these choices on the
+Eurasian out-of-Africa expansion at 50ka and the divergence of western and
+eastern African groups at 60ka. To evaluate the effects of these choices on the
 inference of other parameters in the model, we first fixed those modeled
@@
 
-When we allowed the out-of-Africa expansion and Eastern/Western African
+When we allowed the out-of-Africa expansion and eastern/western African
 divergence times to be jointly fit in model optimization, we found that the

### 2023-01-17 paper.tex
citation update in discussion
comment on sentence needing clarification

@@
 ancestors of eastern/western Africans including the ancestors of
-``Out-of-Africans''. While $120$ka can be interpreted as the basal
-\aprcomment{maybe ``earliest'' instead of ``basal''?} human population divergence,
+``Out-of-Africans''.
+% The following sentence may need clarification
+While $120$ka can be interpreted as the basal human population divergence,
 we suggest conceptualizing this event as the time of most recent
-shared ancestry among sampled populations. \aprcomment{discounting very recent admixture..?}
+shared ancestry among sampled populations.
 ``Population divergence'' in population genetics is a complex concept due
@@
 Therefore, wide variation exists in estimates of divergence time in the
-literature (e.g.\citep{Pagani2015-pz}).
+literature\citep{Bergstrom2021-iw,Henn2018-rf}.
 

### 2023-01-18 paper.tex
basal population split

@@
 % The following sentence may need clarification
-While $120$ka can be interpreted as the basal human population divergence,
-we suggest conceptualizing this event as the time of most recent
+Reticulated models do not have a unique and 
+well-defined "basal" human population divergence.
+We suggest conceptualizing the events $120$ka as the time of most recent
 shared ancestry among sampled populations.
-``Population divergence'' in population genetics is a complex concept due
+However, interpreting population divergence times in population genetics is 
+always challenging due
 to the co-estimation of divergence time and subsequent migration;

### 2023-02-15 paper.tex
Fix quote format
shaved words from abstract

@@
 \begin{abstract}
-While it is now broadly accepted that \emph{Homo sapiens}
+Despite broad agreement that \emph{Homo sapiens}
 originated within Africa, considerable
 uncertainty surrounds specific models of divergence and migration across the
-continent. Progress is hampered by a paucity of fossil and genomic data, as
+continent. Progress is hampered by limited fossil and genomic data, as
 well as variability in prior divergence time estimates.
-Here we use linkage disequilibrium and
+We use linkage disequilibrium and
 diversity-based statistics, optimized for rapid, complex demographic inference
@@
 populations across Africa, including representatives from eastern and western
-groups, as well as 44 newly whole-genome sequenced individuals from the Nama
-(Khoe-San). Despite the complexity of African population history, contemporary
+groups, as well as new whole-genome sequence data for 44 Nama (Khoe-San) individuals. 
+We infer a reticulated African population history in which contemporary
 population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
@@
 the Khoe-San and other groups. Prior to the divergence of contemporary African
-groups, we infer long-lasting structure between two or more weakly
+groups, we infer structure between two or more weakly
 differentiated ancestral \emph{Homo}
 populations connected by gene flow over
-hundreds of thousands of years (i.e., a weakly structured stem). We find that
-weakly structured stem models provide more likely explanations of polymorphism
+hundreds of thousands of years (i.e., a weakly structured stem). Weakly structured stem 
+models provide more likely explanations of patterns of polymorphism
 that had previously been attributed to contributions from archaic hominins in
@@
 Reticulated models do not have a unique and 
-well-defined "basal" human population divergence.
+well-defined ``basal'' human population divergence.
 We suggest conceptualizing the events $120$ka as the time of most recent

### 2023-02-20 paper.tex
start formatting changes for Nature

@@
 \usepackage{lineno}
-\linenumbers
+%\linenumbers
 \renewcommand{\Affilfont}{\normalsize}
@@
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.5\textwidth]{figures/proposed-models.png}
-    \caption{
-        \textbf{Proposed conceptual models of early human history in Africa.}
-        These models have been designed to translate models from the 
-        paleoanthropological literature into genetically testable demographic
-        models \citep{Henn2018-rf}.
-        We used these conceptual models as starting points to build detailed
-        parameterized demographic models (SI Section~\ref{sec:modelspec})
-        that were then fit to genetic data. 
-    }
-    \label{fig:proposed-models}
-\end{figure}
-
 \section*{Results}
@@
 \citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
-(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
+(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}).
+\aprcomment{See Supplementary Information (SI) section 1.2 for details of
+sample collection and consent.}
+The Amhara and Oromo
 populations, despite speaking distinct Afro-Asiatic languages, are highly
@@
 
-\begin{figure}[ht!]
-    \centering
-    \includegraphics{figures/samples-diversity.pdf}
-    \caption[width=\textwidth]{
-        \textbf{Genetic diversity across Africa.}
-        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
-        illustrate diversity from western, eastern and southern Africa.
-        We chose representatives from each region (bold labels)
-        to build parameterized models,
-        including the newly-sequenced Nama from South Africa, Mende
-        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
-        British and Vindija Neanderthal individual.
-        (B, C) PCA highlights the range of genetic divergence anchored 
-        by western Africans, Nama, Gumuz and the British.
-        Percentages show variance explained by each principal component.
-        (D) \texttt{ADMIXTURE} with $K=4$ illustrates signatures of recent gene flow in Africa
-        which reflect colonial-period migration into the Nama,
-        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
-    }
-    \label{fig:diversity}
-\end{figure}
-
 \subsection*{A Late Pleistocene common ancestry for contemporary humans}
@@
 explored a suite of parameterized models in which population structure predates
-the differentiation of contemporary groups (Supporting Information (SI)
+the differentiation of contemporary groups (SI
 Section~\ref{sec:modelspec}). Depending on the parameters, these encompassed
@@
 (Figure~\ref{fig:diversity}).
-
-\begin{figure}[t!]
-    \centering
-    \includegraphics[width=0.65\textwidth]{figures/best-fit-models}
-    \caption{
-        \textbf{A weakly structured stem best describes two-locus statistics.}
-        In the two best-fitting parameterizations of early population structure,
-        continuous migration (A) and multiple mergers (B), models
-        that include ongoing migration between stem populations outperform
-        those in which stem populations are isolated. Most recent populations are
-        connected by continuous, reciprocal migration that are indicated by 
-        double-headed arrows (labels matched to migration rates and divergence
-        times in Table~\ref{tab:migration-rates}). These migrations last for the
-        duration of co-existence of contemporaneous populations with constant
-        migration rates over those intervals. The
-        merger-with-stem-migration model (B, with  $LL=-101,600$) outperformed the
-        continuous-migration model (A, with $LL=-115,300$).
-        Colors are used to distinguish overlapping branches and link to
-        Figure~\ref{fig:diversity}.}
-    \label{fig:best-fit-models}
-\end{figure}
 
@@
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics{figures/predictions.pdf}
-    \caption{
-        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
-        From the best fit models of our two parameterizations (A and B:
-        continuous migration, C and D: merger with stem migration),
-        we predicted pairwise differences $H_{i,j}$ between individuals
-        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
-        (B and D) To understand how drift between stems explains contemporary structure, 
-        we also computed the proportion $\alpha^2$ of drift between pairs of
-        sampled contemporary
-        populations that aligns with drift between past populations
-        (here Nama and Mende, see SI Section~S\ref{sec:f4}
-        for details and additional comparisons in
-        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
-        Both models infer deep population structure with modest contributions to
-        contemporary genetic differentiation.
-        Most present differentiation dates back to the last $100$ka.
-    }
-    \label{fig:predictions}
-\end{figure}
-
 To assess robustness of the inferred models to analysis and reference
@@
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics{figures/validation}
-    \caption{
-        \textbf{Model validation using independent statistics.} (A--C) Using
-        our best fit models to linkage disequilibrium and pairwise diversity statistics,
-        we simulated expected conditional site-frequency-spectra (cSFS) and
-        compared to the observed cSFS from the data. Our inferred models
-        provide a good fit to the data, even though these were not used in our
-        inference. Across the three populations, ancestral state
-        misidentification was consistently inferred to be $1.5-1.7\%$ for
-        intergenic loci (SI \ref{sec:misspecification}). (D, E) We used Relate
-        \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
-        which we used to estimate coalescence rate trajectories
-        and cross-coalescence rates between pairs of populations. While
-        coalescence rate distributions are informative statistics about past
-        evolutionary processes, interpretation can be hindered by migration and
-        population structure, and translating relative cross-coalescence rate
-        curves (RCCR) into population divergence times is especially prone
-        to misinterpretation. For example, the Mende-Gumuz comparison shows
-        a more recent increased RCCR than either population with the British,
-        a pattern that is recapitulated under our best-fit model, even though
-        the Mende-Gumuz split occurs prior to the Gumuz-British split.
-    }
-    \label{fig:validation}
-\end{figure}
-
 We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
@@
 increase in $N_e$ between $100$ka--$1$Ma
-%extending deeper in the past than the Relate estimate
-% APR: I don't understand this phrase I commented out
 (Figure~\ref{fig:supp-iicr-sim} and SI
@@
 
-\subsection*{The Formation of Population Structure in Africa}
+\subsection*{The formation of population structure in Africa}
 
@@
 
-\section*{Data availability}
+\bibliographystyle{naturemag}
+\bibliography{paper}
 
-Nama sequencing data are available from the European Genome-Phenome Archive (EGA),
-accession number: EGAD00001006198. The African Diversity Reference Panel can be
-found at accession: EGAS00001000960.
-
-\section*{Code availability}
-
-Code for the software used in this paper is found at the following locations:\\
-moments-LD (\url{https://bitbucket.org/simongravel/moments}),\\
-Demes (\url{https://github.com/popsim-consortium/demes-python}),\\
-Relate (\url{https://myersgroup.github.io/relate/}),\\
-msprime (\url{https://github.com/tskit-dev/msprime}),\\
-tskit (\url{https://github.com/tskit-dev/tskit}),\\
-and scripts implementing analyses using each of these software are available at\\
-\url{https://github.com/apragsdale/african-structure-paper}.
+\section*{Supplementary information}
+Available in the online version of the paper.
 
@@
 
-\bibliographystyle{naturemag}
-\bibliography{paper}
+\section*{Competing interests}
+The authors declare no competing financial interests.
+
+\section*{Author contributions}
+\aprcomment{To do: fill in (e.g., designed the study, collected data,
+analyzed data, wrote and edited the manuscript)}
+\begin{itemize}
+    \item APR:
+    \item TDW:
+    \item EGA: 
+    \item EH:
+    \item MM:
+    \item BMH:
+    \item SG:
+\end{itemize}
+
+\section*{Methods}
+
+\aprcomment{This can be very short}
+
+\section*{Data availability}
+
+Nama sequencing data are available from the European Genome-Phenome Archive (EGA),
+accession number: EGAD00001006198. The African Diversity Reference Panel can be
+found at accession: EGAS00001000960.
+
+\section*{Code availability}
+
+Code for the software used in this paper is found at the following locations:\\
+moments-LD (\url{https://bitbucket.org/simongravel/moments}),\\
+Demes (\url{https://github.com/popsim-consortium/demes-python}),\\
+Relate (\url{https://myersgroup.github.io/relate/}),\\
+msprime (\url{https://github.com/tskit-dev/msprime}),\\
+tskit (\url{https://github.com/tskit-dev/tskit}),\\
+and scripts implementing analyses using each of these software are available at\\
+\url{https://github.com/apragsdale/african-structure-paper}.
+
+\break
+
+\section*{Figure captions}
+
+\begin{figure}[ht]
+    \centering
+    %\includegraphics[width=0.5\textwidth]{figures/proposed-models.png}
+    \caption{
+        \textbf{Proposed conceptual models of early human history in Africa.}
+        These models have been designed to translate models from the 
+        paleoanthropological literature into genetically testable demographic
+        models \citep{Henn2018-rf}.
+        We used these conceptual models as starting points to build detailed
+        parameterized demographic models (SI Section~\ref{sec:modelspec})
+        that were then fit to genetic data. 
+    }
+    \label{fig:proposed-models}
+\end{figure}
+
+\begin{figure}[ht!]
+    \centering
+    %\includegraphics{figures/samples-diversity.pdf}
+    \caption[width=\textwidth]{
+        \textbf{Genetic diversity across Africa.}
+        (A) Select populations from the 1000 Genomes and African Diversity Reference Panels
+        illustrate diversity from western, eastern and southern Africa.
+        We chose representatives from each region (bold labels)
+        to build parameterized models,
+        including the newly-sequenced Nama from South Africa, Mende
+        from Sierre Leone, Gumuz, Oromo and Amhara from Ethiopia, and the
+        British and Vindija Neanderthal individual.
+        (B, C) PCA highlights the range of genetic divergence anchored 
+        by western Africans, Nama, Gumuz and the British.
+        Percentages show variance explained by each principal component.
+        (D) \texttt{ADMIXTURE} with $K=4$ illustrates signatures of recent gene flow in Africa
+        which reflect colonial-period migration into the Nama,
+        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
+    }
+    \label{fig:diversity}
+\end{figure}
+
+\begin{figure}[t!]
+    \centering
+    %\includegraphics[width=0.65\textwidth]{figures/best-fit-models}
+    \caption{
+        \textbf{A weakly structured stem best describes two-locus statistics.}
+        In the two best-fitting parameterizations of early population structure,
+        continuous migration (A) and multiple mergers (B), models
+        that include ongoing migration between stem populations outperform
+        those in which stem populations are isolated. Most recent populations are
+        connected by continuous, reciprocal migration that are indicated by 
+        double-headed arrows (labels matched to migration rates and divergence
+        times in Table~\ref{tab:migration-rates}). These migrations last for the
+        duration of co-existence of contemporaneous populations with constant
+        migration rates over those intervals. The
+        merger-with-stem-migration model (B, with  $LL=-101,600$) outperformed the
+        continuous-migration model (A, with $LL=-115,300$).
+        Colors are used to distinguish overlapping branches and link to
+        Figure~\ref{fig:diversity}.}
+    \label{fig:best-fit-models}
+\end{figure}
+
+\begin{figure}[t!]
+    \centering
+    %\includegraphics{figures/predictions.pdf}
+    \caption{
+        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
+        From the best fit models of our two parameterizations (A and B:
+        continuous migration, C and D: merger with stem migration),
+        we predicted pairwise differences $H_{i,j}$ between individuals
+        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
+        (B and D) To understand how drift between stems explains contemporary structure, 
+        we also computed the proportion $\alpha^2$ of drift between pairs of
+        sampled contemporary
+        populations that aligns with drift between past populations
+        (here Nama and Mende, see SI Section~S\ref{sec:f4}
+        for details and additional comparisons in
+        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
+        Both models infer deep population structure with modest contributions to
+        contemporary genetic differentiation.
+        Most present differentiation dates back to the last $100$ka.
+    }
+    \label{fig:predictions}
+\end{figure}
+
+\begin{figure}[t!]
+    \centering
+    %\includegraphics{figures/validation}
+    \caption{
+        \textbf{Model validation using independent statistics.} (A--C) Using
+        our best fit models to linkage disequilibrium and pairwise diversity statistics,
+        we simulated expected conditional site-frequency-spectra (cSFS) and
+        compared to the observed cSFS from the data. Our inferred models
+        provide a good fit to the data, even though these were not used in our
+        inference. Across the three populations, ancestral state
+        misidentification was consistently inferred to be $1.5-1.7\%$ for
+        intergenic loci (SI \ref{sec:misspecification}). (D, E) We used Relate
+        \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
+        which we used to estimate coalescence rate trajectories
+        and cross-coalescence rates between pairs of populations. While
+        coalescence rate distributions are informative statistics about past
+        evolutionary processes, interpretation can be hindered by migration and
+        population structure, and translating relative cross-coalescence rate
+        curves (RCCR) into population divergence times is especially prone
+        to misinterpretation. For example, the Mende-Gumuz comparison shows
+        a more recent increased RCCR than either population with the British,
+        a pattern that is recapitulated under our best-fit model, even though
+        the Mende-Gumuz split occurs prior to the Gumuz-British split.
+    }
+    \label{fig:validation}
+\end{figure}
+
 \end{document}

### 2023-02-20 supplement.tex
start formatting changes for Nature

@@
 \begin{document}
-\title{Supporting Information for\\
+\title{Supplementary Information for\\
 ``A weakly structured stem for human origins in Africa''}
@@
 
-\addcontentsline{toc}{section}{Supporting tables and figures}
-\section*{Supporting tables and figures}
+\addcontentsline{toc}{section}{Supplementary tables and figures}
+\section*{Supplementary tables and figures}
 

### 2023-02-20 paper.tex
an attempt to condense the abstract

@@
 robust inferences about deep history.
+\end{abstract}
+
+\begin{abstract}
+    \aprcomment{An attempt at condensing a bit -- this is just under 230 words}
+
+    Despite broad agreement that \emph{Homo sapiens} originated within Africa,
+    considerable uncertainty surrounds specific models of divergence and
+    migration across the continent. Progress is hampered by limited fossil and
+    genomic data, as well as variability in prior divergence time estimates.
+    We use linkage disequilibrium and diversity-based statistics, optimized for
+    rapid, complex demographic inference to discriminate among such models.  We
+    infer detailed demographic models for populations across Africa, including
+    representatives from eastern and western groups and newly sequenced whole
+    genomes from 44 Nama (Khoe-San) individuals. We infer a reticulated African
+    population history in which contemporary population structure dates back to
+    Marine Isotope Stage (MIS) 5. The earliest population divergence among
+    contemporary populations occurs 120--135ka and was preceded by structure
+    between two or more weakly differentiated ancestral \emph{Homo} populations
+    connected by gene flow over hundreds of thousands of years. Such weakly
+    structured stem models provide more likely explanations of patterns of
+    polymorphism that were previously attributed to contributions from archaic
+    hominins in Africa. In contrast to models with archaic introgression, we
+    predict fossil remains from coexisting ancestral populations should be
+    genetically and morphologically similar, and an inferred 1--4\% of genetic
+    differentiation among contemporary human populations can be attributed to
+    genetic drift between stem populations. We show that model misspecification
+    explains variation in previous divergence time estimates and argue that
+    studying a suite of models is key to robust inferences about deep history.
+
 \end{abstract}
@@
 data for four diverse African populations, comprising the Nama (Khoe-San from
-South Africa, newly presented here), Mende (from Sierra Leone, MSL from the
+South Africa, newly presented here, see Supplementary Information (SI) section
+1.2 for details of
+sample collection and consent), Mende (from Sierra Leone, MSL from the
 Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
@@
 \citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
-(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}).
-\aprcomment{See Supplementary Information (SI) section 1.2 for details of
-sample collection and consent.}
-The Amhara and Oromo
+(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
 populations, despite speaking distinct Afro-Asiatic languages, are highly

### 2023-02-20 paper.tex
minor typo fixes

@@
 ancestors of the Nama and other groups is an order of magnitude weaker than
-that observed between between western and eastern Africans (Table
+that observed between western and eastern Africans (Table
 \ref{tab:migration-rates}). All models infer relatively high gene flow between
@@
 sharp bottleneck in Stem 1 down to $N_e=100$ after the split of the Neanderthal
-branch. This represents the lower bound allowed in our optimization (i.e. an
+branch. This represents the lower bound allowed in our optimization (i.e., an
 $N_e$ of 100), although the size of this bottleneck is poorly constrained (95\%
@@
 strategies and parameter optimization strategies. While we find some
-differences in the inferred parameters (see in particular Sections
-\ref{sec:replacing} and \ref{sec:change-dates}), the best fit model across all
+differences in the inferred parameters (see SI sections
+\ref{sec:replacing} and \ref{sec:change-dates}), the best fit models across all
 reanalyses are quantitatively consistent.
@@
 Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
-increase in $N_e$ during that period, the best-fit model recapitulate this
+increase in $N_e$ during that period, the best-fit models recapitulate this
 pattern without corresponding population size changes.
@@
 of isolation among the stems is variable across fits, models with a period of
-divergence, isolation and then a merger event (i.e. a ``reticulation'')
+divergence, isolation and then a merger event (i.e., a ``reticulation'')
 out-performed models with bifurcating divergence and continuous gene flow.
@@
 interpretation is reinforced by the differential migration rates between
-regions, i.e. the gene flow from Stem 2 to western Africans is estimated to be
+regions, i.e., the gene flow from Stem 2 to western Africans is estimated to be
 five times that of the rate to eastern Africans during this period. We

### 2023-02-21 paper.tex
missing period

@@
 among contemporary populations can be traced back to this early population
-structure (SI Section~\ref{sec:f4})
+structure (SI Section~\ref{sec:f4}).
 

### 2023-02-22 paper.tex
minor edits to abstract

@@
     between two or more weakly differentiated ancestral \emph{Homo} populations
-    connected by gene flow over hundreds of thousands of years. Such weakly
-    structured stem models provide more likely explanations of patterns of
+    connected by gene flow over hundreds of thousands of years. Such \emph{weakly
+    structured stem} models better explain patterns of
     polymorphism that were previously attributed to contributions from archaic
@@
     predict fossil remains from coexisting ancestral populations should be
-    genetically and morphologically similar, and an inferred 1--4\% of genetic
+    genetically and morphologically similar, and that an inferred 1--4\% of genetic
     differentiation among contemporary human populations can be attributed to

### 2023-02-24 paper.tex
added author contributions draft

@@
 analyzed data, wrote and edited the manuscript)}
-\begin{itemize}
-    \item APR:
-    \item TDW:
-    \item EGA: 
-    \item EH:
-    \item MM:
-    \item BMH:
-    \item SG:
-\end{itemize}
+\sgcomment{Very rough first pass:}
+APR, BMH, and SG designed the study. BMH, EGA, EH, and MM designed and conducted 
+participant recruitment.   APR, BMH, and SG designed statistical analyses.
+APR conducted statistical analyses. 
+APR, TDW, BMH, and SG interpreted the results and wrote the first draft of the article. 
+All authors read and edited the article.
 

### 2023-02-25 paper.tex
added funding
Shorten section headers

@@
 
-\subsection*{A Late Pleistocene common ancestry for contemporary humans}
+\subsection*{A Late Pleistocene common ancestry}
 
@@
 
-\subsection*{Deep but connected population structure within Africa}
+\subsection*{A weakly structured stem within Africa}
 
@@
 
-\subsection*{Reconciling multiple lines of genetic evidence}
+\subsection*{Reconciling lines of genetic evidence}
 
@@
 
-\subsection*{The formation of population structure in Africa}
+\subsection*{Population structure formation in Africa}
 
@@
 
-\subsection*{Contrasting archaic hominin admixture and a weakly structured stem}
+\subsection*{Contrasting ancestral structure models}
 
@@
 visualization discussion and designing Figures \ref{fig:proposed-models} and
-\ref{fig:best-fit-models}. This research was supported by an NIH grant
+\ref{fig:best-fit-models}. This research was supported by CIHR project grant 437576,
+Natural Sciences and Engineering Research Council of Canada (NSERC) grant RGPIN-2017-04816, 
+and the Canada Research Chair program. 
+This research was supported by an NIH grant
 R35GM133531 (to BMH). The content is solely the responsibility of the authors

### 2023-02-26 paper.tex
Save condensed abstract

@@
 \begin{abstract}
-Despite broad agreement that \emph{Homo sapiens}
-originated within Africa, considerable
-uncertainty surrounds specific models of divergence and migration across the
-continent. Progress is hampered by limited fossil and genomic data, as
-well as variability in prior divergence time estimates.
-We use linkage disequilibrium and
-diversity-based statistics, optimized for rapid, complex demographic inference
-to discriminate among such models.
-We infer detailed demographic models for
-populations across Africa, including representatives from eastern and western
-groups, as well as new whole-genome sequence data for 44 Nama (Khoe-San) individuals. 
-We infer a reticulated African population history in which contemporary
-population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
-population divergence among contemporary populations occurs 120-135ka, between
-the Khoe-San and other groups. Prior to the divergence of contemporary African
-groups, we infer structure between two or more weakly
-differentiated ancestral \emph{Homo}
-populations connected by gene flow over
-hundreds of thousands of years (i.e., a weakly structured stem). Weakly structured stem 
-models provide more likely explanations of patterns of polymorphism
-that had previously been attributed to contributions from archaic hominins in
-Africa.
-In contrast to models with archaic introgression, we predict that fossil
-remains from coexisting ancestral populations should be morphologically
-similar.
-Despite genetic similarity between these populations, an inferred 1--4\%
-of genetic differentiation among contemporary human populations can be
-attributed to genetic drift between stem populations.
-We show that model misspecification explains variation in previous
-divergence time estimates and argue that studying a suite of models is key to
-robust inferences about deep history.
-\end{abstract}
-
-\begin{abstract}
-    \aprcomment{An attempt at condensing a bit -- this is just under 230 words}
 

### 2023-02-26 paper.tex
Changed OOA wording
Discussion, last paragraph of Population structure formation in Africa

@@
 ancestors of eastern/western Africans including the ancestors of
-``Out-of-Africans''.
-% The following sentence may need clarification
-Reticulated models do not have a unique and 
+``Out-of-Africans''. Reticulated models do not have a unique and 
 well-defined ``basal'' human population divergence.
@@
 the inferred Out-of-Africa and eastern/western African divergences which were
-10--15ka younger than our fixed parameters. The
-complexity of the OOA expansion warrants the inclusion of additional Eurasian populations 
-and more detailed modeling. However, other features
-of our inferred models are weakly affected by this analysis choice.
+10--15ka younger than our fixed parameters. These younger dates are clearly at 
+odds with the accepted timing of Out-of-Africa expansion at approximately $\sim>50$ka, 
+as based on archaeological, climatic and paleo-morphological information [cite]. 
+We interpret the free estimate for eastern African vs. European divergence as reflecting 
+our inclusion of only a single OOA population in the model, the lack of a nearby 
+source for back-to-Africa gene flow, and other regionally complex parameters. 
+However, older pan-African features of our inferred models are weakly affected by 
+this analysis choice (SI Section~\ref{sec:models}).
 

### 2023-02-26 paper.tex
Update paper.tex
comma

--- 2023-02-25 paper.tex
@@
     between two or more weakly differentiated ancestral \emph{Homo} populations
-    connected by gene flow over hundreds of thousands of years. Such \emph{weakly
-    structured stem} models better explain patterns of
+    connected by gene flow over hundreds of thousands of years. Such weakly
+    structured stem models provide more likely explanations of patterns of
     polymorphism that were previously attributed to contributions from archaic
@@
     predict fossil remains from coexisting ancestral populations should be
-    genetically and morphologically similar, and that an inferred 1--4\% of genetic
+    genetically and morphologically similar, and an inferred 1--4\% of genetic
     differentiation among contemporary human populations can be attributed to
@@
 
-\subsection*{A Late Pleistocene common ancestry}
+\subsection*{A Late Pleistocene common ancestry for contemporary humans}
 
@@
         2) Nama and Mende, respectively.
-        ``Divergence times'' correspond to the most recent common ancestral population,
+        ``Divergence times'' correspond to the most recent common ancestral population
         and does not account for continuous migration or earlier reticulations.
@@
 
-\subsection*{A weakly structured stem within Africa}
+\subsection*{Deep but connected population structure within Africa}
 
@@
 
-\subsection*{Reconciling lines of genetic evidence}
+\subsection*{Reconciling multiple lines of genetic evidence}
 
@@
 
-\subsection*{Population structure formation in Africa}
+\subsection*{The formation of population structure in Africa}
 
@@
 
-\subsection*{Contrasting ancestral structure models}
+\subsection*{Contrasting archaic hominin admixture and a weakly structured stem}
 
@@
 visualization discussion and designing Figures \ref{fig:proposed-models} and
-\ref{fig:best-fit-models}. This research was supported by CIHR project grant 437576,
-Natural Sciences and Engineering Research Council of Canada (NSERC) grant RGPIN-2017-04816, 
-and the Canada Research Chair program. 
-This research was supported by an NIH grant
+\ref{fig:best-fit-models}. This research was supported by an NIH grant
 R35GM133531 (to BMH). The content is solely the responsibility of the authors
@@
 analyzed data, wrote and edited the manuscript)}
-\sgcomment{Very rough first pass:}
-APR, BMH, and SG designed the study. BMH, EGA, EH, and MM designed and conducted 
-participant recruitment.   APR, BMH, and SG designed statistical analyses.
-APR conducted statistical analyses. 
-APR, TDW, BMH, and SG interpreted the results and wrote the first draft of the article. 
-All authors read and edited the article.
+\begin{itemize}
+    \item APR:
+    \item TDW:
+    \item EGA: 
+    \item EH:
+    \item MM:
+    \item BMH:
+    \item SG:
+\end{itemize}
 

### 2023-02-26 paper.tex

--- 2023-02-26 paper.tex
@@
 
-\subsection*{Population structure formation in Africa}
+\subsection*{Formation of Population Structure in Africa}
 

### 2023-02-26 paper.tex
add references

@@
 odds with the accepted timing of Out-of-Africa expansion at approximately $\sim>50$ka, 
-as based on archaeological, climatic and paleo-morphological information [cite]. 
+as based on archaeological, climatic and paleo-morphological information
+\citep{Fu2014-am,Groucutt2015-do,Prufer2021-ao,Beyer2021-ym}. 
 We interpret the free estimate for eastern African vs. European divergence as reflecting 

### 2023-02-27 paper.tex
minor wording tweaks, removed extra comments

@@
 South Africa, newly presented here, see Supplementary Information (SI) section
-1.2 for details of
-sample collection and consent), Mende (from Sierra Leone, MSL from the
+1.2 for ethical and practical aspects of
+sample collection), Mende (from Sierra Leone, MSL from the
 Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
@@
 \section*{Author contributions}
-\aprcomment{To do: fill in (e.g., designed the study, collected data,
-analyzed data, wrote and edited the manuscript)}
-\sgcomment{Very rough first pass:}
+
+\sgcomment{Rough first pass for contributions:}
 APR, BMH, and SG designed the study. BMH, EGA, EH, and MM designed and conducted 

### 2023-02-27 paper.tex
minor wording tweaks, removed extra comments

@@
 South Africa, newly presented here, see Supplementary Information (SI) section
-1.2 for ethical and practical aspects of
-sample collection), Mende (from Sierra Leone, MSL from the
+1.2 for ethical and practical aspects of participant recruitment), Mende (from Sierra Leone, MSL from the
 Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),

### 2023-02-27 paper.tex
added refs to abstract

@@
     considerable uncertainty surrounds specific models of divergence and
-    migration across the continent. Progress is hampered by limited fossil and
-    genomic data, as well as variability in prior divergence time estimates.
-    We use linkage disequilibrium and diversity-based statistics, optimized for
-    rapid, complex demographic inference to discriminate among such models.  We
+    migration across the continent\citep{Henn2018-rf}. Progress is hampered by limited fossil and
+    genomic data, as well as variability in prior divergence time estimates\citep{Henn2018-rf}.
+    To discriminate among such models, we consider linkage disequilibrium and diversity-based 
+    statistics, optimized for rapid, complex demographic inference \citep{Ragsdale2019-nt}.  We
     infer detailed demographic models for populations across Africa, including
@@
     polymorphism that were previously attributed to contributions from archaic
-    hominins in Africa. In contrast to models with archaic introgression, we
+    hominins in Africa \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
+Lorente-Galdos2019-vz,Durvasula2020-td}. In contrast to models with archaic introgression, we
     predict fossil remains from coexisting ancestral populations should be

### 2023-02-27 paper.tex
start methods in main text

--- 2023-02-27 paper.tex
@@
 South Africa, newly presented here, see Supplementary Information (SI) section
-1.2 for ethical and practical aspects of
-sample collection), Mende (from Sierra Leone, MSL from the
+1.2 for details of
+sample collection and consent), Mende (from Sierra Leone, MSL from the
 Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
@@
 \section*{Author contributions}
-
-\sgcomment{Rough first pass for contributions:}
+\aprcomment{To do: fill in (e.g., designed the study, collected data,
+analyzed data, wrote and edited the manuscript)}
+\sgcomment{Very rough first pass:}
 APR, BMH, and SG designed the study. BMH, EGA, EH, and MM designed and conducted 
@@
 
-\aprcomment{This can be very short}
+\subsection*{Data and sequencing}
+
+We used merged sequencing data from the African Diversity Reference Panel
+
+\subsection*{LD and diversity statistics}
+
+\subsection*{Model specification and fitting}
+
+\subsection*{Gene genalogy reconstruction}
+
+\subsection*{Conditional site frequency spectrum}
 

### 2023-02-27 paper.tex
typos

--- 2023-02-27 paper.tex
@@
 
-\subsection*{Formation of Population Structure in Africa}
+\subsection*{Formation of population structure in Africa}
 
@@
 10--15ka younger than our fixed parameters. These younger dates are clearly at 
-odds with the accepted timing of Out-of-Africa expansion at approximately $\sim>50$ka, 
+odds with the accepted timing of Out-of-Africa expansion at approximately $\>50$ka, 
 as based on archaeological, climatic and paleo-morphological information

### 2023-02-27 paper.tex
add some methods

--- 2023-02-27 paper.tex
@@
 \ref{sec:models}). These parameters were ones supported by multiple genetic and
-archaeological studies. Fixed parameters included the time of divergence
+archaeological studies \citep{Bergstrom2021-iw}.
+Fixed parameters included the time of divergence
 between western and eastern African populations, set to $60$ka, just prior to
@@
 
-\aprcomment{This can be very short}
+\subsection*{Data and sequencing}
+
+We used a subset of populations from the merged sequencing data from the
+African Diversity Reference Panel (ADRP) \citep{Gurdasani2015-qy,Pagani2015-pz}
+and the Thousand Genomes Project (1KGP)
+\citep{1000_Genomes_Project_Consortium2015-zq}. These included the Amhara,
+Bakiga, Gumuz, Nama, Oromo, Somali, and Zulu from ADRP, and ESN, GWD, LWK, MSL,
+YRI, CEU, GBR, CHB, and PJL from 1KGP. After filtering for relatedness and
+retaining Nama individuals with $>70\%$ estimated Khoe-San ancestry, we used
+data from 289 individuals, including 44 Nama individuals. These were merged
+with the high-coverage Neanderthal genome from Vindija Cave
+\citep{Prufer2017-kk}. We kept variants from regions that fell within the 1KGP
+srict callability mask, overlapped with at least 100 continuously called base
+pairs in the Neanderthal genome, and were annotated as intergenic.
+\texttt{ADMIXTURE} and PCA analyses were conducted on a subset of variants
+filtered to remove variants in high LD ($r^2$ threshold of $0.1$). See SI
+section \ref{sec:data-sequencing} for additional details on data and
+sequencing.
+
+\subsection*{LD and diversity statistics}
+
+We used multi-population LD and pairwise diversity statistics to fit
+parameterized demographic models to data \citep{Ragsdale2019-nt}. Unbiased LD
+statistics were computed from all variants in retained intergenic regions
+\citep{Ragsdale2020-nz}, for pairs of variants separated by recombination
+distances $r = 5\times10^{-6}$ to $5\times10^{-3}$. These were assigned to 16
+recombination distance bins, and average statistics were computed within each
+bin (SI section \ref{sec:computing-stats}). Expected statistics were computed
+in \texttt{moments}, which also performed likelihood-based parameter
+optimization. Conditional site frequency spectra (cSFS) were computed
+conditioned on the Vindija Neanderthal carrying the derived allele relative to
+the ancestral allele determined by a six-primate alignment
+\citep{1000_Genomes_Project_Consortium2015-zq}.
+
+\subsection*{Model specification and fitting}
+
+Model parameters include population sizes and size changes, split times,
+continuous migration rates, and admixture times and proportions. The simplest
+model we tested was a bifurcating tree-like structure, allowing for subsequent
+migrations and recent known admixture events. To include ancestral population
+structure, we tested models that included multiple stem groups, which each were
+allowed their own population size and could be connected by continuous
+migration. We tested multiple scenarios of early population structure,
+including long-lasting continuous migration between stem populations and
+scenarios of periods of isolation with subsequent merger events (SI
+section~\ref{sec:modelspec}). To avoid overfitting, we to incrementally added
+copmlexity to our model optimization, and we fixed a number of parameters that
+are constrained by historical records or are consistently estimated across
+multiple models and previous studies (SI section~\ref{sec:models}). Likelihoods
+were computed using a composite multivariate Gaussian likelihood approach, and
+confidence intervals were estimated by refitting each model to 200
+block-bootstrap replicate datasets (SI
+section~\ref{sec:likelihoods-uncertainty}). We iteratively used gradient
+descent and L-BFGS-B optimization routines to fit each parameterized model (SI
+section~\ref{sec:optimization}).
+
+\subsection*{Gene genalogy reconstruction}
+
+\subsection*{Conditional site frequency spectrum}
 

### 2023-02-27 paper.tex
edits to methods

@@
 
-\sgcomment{Rough first pass for contributions:}
-APR, BMH, and SG designed the study. BMH, EGA, EH, and MM designed and conducted 
-participant recruitment.   APR, BMH, and SG designed statistical analyses.
+APR, BMH, and SG designed the study.
+BMH, EGA, EH, and MM designed and conducted participant recruitment.
+APR, BMH, and SG designed statistical analyses.
 APR conducted statistical analyses. 
@@
 
-\subsection*{Conditional site frequency spectrum}
+We used \texttt{Relate} \citep{Speidel2019-nj} to reconstruct genome-wide gene
+genealogies from the retained populations in the merged ADRP and 1KGP datasets
+(SI section~\ref{sec:relate}). From reconstructed genealogies, we computed
+coalescence rates within and between populations, which provide an estimate for
+effective population sizes over time and the relative cross-coalescence rates
+between pairs of populations. To compare reconstructed genealogies from data to
+model predictions, we used \texttt{msprime}
+\citep{Kelleher2016-lw,Baumdicker2022-mj} to simulate genomic data for equal
+numbers of samples for each population in our inferred models. We then applied
+\texttt{Relate} to these simulated datasets using the same mutation and
+recombination rates and generation time.
 
@@
 
-Nama sequencing data are available from the European Genome-Phenome Archive (EGA),
-accession number: EGAD00001006198. The African Diversity Reference Panel can be
-found at accession: EGAS00001000960.
+Nama sequencing data are available from the European Genome-Phenome Archive
+(EGA), accession number: EGAD00001006198. The African Diversity Reference Panel
+can be found at accession: EGAS00001000960.
 

### 2023-02-27 paper.tex
minor edits to methods section

@@
 and the Thousand Genomes Project (1KGP)
-\citep{1000_Genomes_Project_Consortium2015-zq}. These included the Amhara,
+\citep{1000_Genomes_Project_Consortium2015-zq} \sgcomment{This sounds like the Nama are from the 2015 paper. Perhaps: 
+We generated a sequencing dataset by combining existing and new populations recruited as part of the 
+African Diversity Reference Panel (ADRP) \citep{Gurdasani2015-qy,Pagani2015-pz},
+as well as Thousand Genomes Project populations (1KGP)
+\citep{1000_Genomes_Project_Consortium2015-zq} 
+}. These included the Amhara,
 Bakiga, Gumuz, Nama, Oromo, Somali, and Zulu from ADRP, and ESN, GWD, LWK, MSL,
 YRI, CEU, GBR, CHB, and PJL from 1KGP. After filtering for relatedness and
-retaining Nama individuals with $>70\%$ estimated Khoe-San ancestry, we used
+retaining Nama individuals with $>70\%$ estimated Khoe-San ancestry, we focused on
 data from 289 individuals, including 44 Nama individuals. These were merged
@@
 recombination distance bins, and average statistics were computed within each
-bin (SI section \ref{sec:computing-stats}). Expected statistics were computed
+bin (SI section \ref{sec:computing-stats}). Expected statistics under each model were computed
 in \texttt{moments}, which also performed likelihood-based parameter
@@
 migrations and recent known admixture events. To include ancestral population
-structure, we tested models that included multiple stem groups, which each were
+structure, we tested models that included multiple stem groups, each of which were
 allowed their own population size and could be connected by continuous
@@
 scenarios of periods of isolation with subsequent merger events (SI
-section~\ref{sec:modelspec}). To avoid overfitting, we to incrementally added
-copmlexity to our model optimization, and we fixed a number of parameters that
+section~\ref{sec:modelspec}). To avoid overfitting, we incrementally added
+complexity to our model optimization, and we fixed a number of parameters that
 are constrained by historical records or are consistently estimated across
@@
 We used \texttt{Relate} \citep{Speidel2019-nj} to reconstruct genome-wide gene
-genealogies from the retained populations in the merged ADRP and 1KGP datasets
+genealogies from the focal populations in the merged ADRP and 1KGP datasets
 (SI section~\ref{sec:relate}). From reconstructed genealogies, we computed

### 2023-02-27 paper.tex
tweak to previous suggestion

@@
 \citep{1000_Genomes_Project_Consortium2015-zq} \sgcomment{This sounds like the Nama are from the 2015 paper. Perhaps: 
-We generated a sequencing dataset by combining existing and new populations recruited as part of the 
+We generated a sequencing dataset by combining existing and newly recruited populations who are now part of the 
 African Diversity Reference Panel (ADRP) \citep{Gurdasani2015-qy,Pagani2015-pz},

### 2023-02-27 paper.tex
spell out OOA since undefined

@@
 We interpret the free estimate for eastern African vs. European divergence as reflecting 
-our inclusion of only a single OOA population in the model, the lack of a nearby 
+our inclusion of only a single Out-of-Africa population in the model, the lack of a nearby 
 source for back-to-Africa gene flow, and other regionally complex parameters. 

### 2023-02-27 paper.tex
accept suggestions

@@
     considerable uncertainty surrounds specific models of divergence and
-    migration across the continent\citep{Henn2018-rf}. Progress is hampered by limited fossil and
-    genomic data, as well as variability in prior divergence time estimates\citep{Henn2018-rf}.
-    To discriminate among such models, we consider linkage disequilibrium and diversity-based 
-    statistics, optimized for rapid, complex demographic inference \citep{Ragsdale2019-nt}.  We
-    infer detailed demographic models for populations across Africa, including
+    migration across the continent\citep{Henn2018-rf}. Progress is hampered by
+    limited fossil and genomic data, as well as variability in prior divergence
+    time estimates\citep{Henn2018-rf}.  To discriminate among such models, we
+    consider linkage disequilibrium and diversity-based statistics, optimized
+    for rapid, complex demographic inference \citep{Ragsdale2019-nt}.  We infer
+    detailed demographic models for populations across Africa, including
     representatives from eastern and western groups and newly sequenced whole
@@
     between two or more weakly differentiated ancestral \emph{Homo} populations
-    connected by gene flow over hundreds of thousands of years. Such \emph{weakly
-    structured stem} models better explain patterns of
+    connected by gene flow over hundreds of thousands of years. Such
+    \emph{weakly structured stem} models better explain patterns of
     polymorphism that were previously attributed to contributions from archaic
-    hominins in Africa \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
-Lorente-Galdos2019-vz,Durvasula2020-td}. In contrast to models with archaic introgression, we
-    predict fossil remains from coexisting ancestral populations should be
-    genetically and morphologically similar, and that an inferred 1--4\% of genetic
-    differentiation among contemporary human populations can be attributed to
-    genetic drift between stem populations. We show that model misspecification
-    explains variation in previous divergence time estimates and argue that
-    studying a suite of models is key to robust inferences about deep history.
+    hominins in Africa
+    \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
+    Lorente-Galdos2019-vz,Durvasula2020-td}. In contrast to models with archaic
+    introgression, we predict fossil remains from coexisting ancestral
+    populations should be genetically and morphologically similar, and that an
+    inferred 1--4\% of genetic differentiation among contemporary human
+    populations can be attributed to genetic drift between stem populations. We
+    show that model misspecification explains variation in previous divergence
+    time estimates and argue that studying a suite of models is key to robust
+    inferences about deep history.
 
@@
 
-We used a subset of populations from the merged  sequencing data from the
-African Diversity Reference Panel (ADRP) \citep{Gurdasani2015-qy,Pagani2015-pz}
-and the Thousand Genomes Project (1KGP)
-\citep{1000_Genomes_Project_Consortium2015-zq} \sgcomment{This sounds like the Nama are from the 2015 paper. Perhaps: 
-We generated a sequencing dataset by combining existing and newly recruited populations who are now part of the 
-African Diversity Reference Panel (ADRP) \citep{Gurdasani2015-qy,Pagani2015-pz},
-as well as Thousand Genomes Project populations (1KGP)
-\citep{1000_Genomes_Project_Consortium2015-zq} 
-}. These included the Amhara,
-Bakiga, Gumuz, Nama, Oromo, Somali, and Zulu from ADRP, and ESN, GWD, LWK, MSL,
-YRI, CEU, GBR, CHB, and PJL from 1KGP. After filtering for relatedness and
-retaining Nama individuals with $>70\%$ estimated Khoe-San ancestry, we focused on
-data from 289 individuals, including 44 Nama individuals. These were merged
-with the high-coverage Neanderthal genome from Vindija Cave
-\citep{Prufer2017-kk}. We kept variants from regions that fell within the 1KGP
-srict callability mask, overlapped with at least 100 continuously called base
-pairs in the Neanderthal genome, and were annotated as intergenic.
-\texttt{ADMIXTURE} and PCA analyses were conducted on a subset of variants
-filtered to remove variants in high LD ($r^2$ threshold of $0.1$). See SI
-section \ref{sec:data-sequencing} for additional details on data and
+We generated a sequencing dataset by combining existing and newly recruited
+populations who are now part of the African Diversity Reference Panel (ADRP)
+\citep{Gurdasani2015-qy,Pagani2015-pz}, as well as Thousand Genomes Project
+populations (1KGP) \citep{1000_Genomes_Project_Consortium2015-zq}. These
+included the Amhara, Bakiga, Gumuz, Nama (newly generated), Oromo, Somali, and
+Zulu from ADRP, and ESN, GWD, LWK, MSL, YRI, CEU, GBR, CHB, and PJL from 1KGP.
+After filtering for relatedness and retaining Nama individuals with $>70\%$
+estimated Khoe-San ancestry, we focused on data from 289 individuals, including
+44 Nama individuals. These were merged with the high-coverage Neanderthal
+genome from Vindija Cave \citep{Prufer2017-kk}. We kept variants from regions
+that fell within the 1KGP srict callability mask, overlapped with at least 100
+continuously called base pairs in the Neanderthal genome, and were annotated as
+intergenic. \texttt{ADMIXTURE} and PCA analyses were conducted on a subset of
+variants filtered to remove variants in high LD ($r^2$ threshold of $0.1$). See
+SI section \ref{sec:data-sequencing} for additional details on data and
 sequencing.

### 2023-03-01 paper.tex
updated author contributions
Acknowledgement and access

@@
 Natural Sciences and Engineering Research Council of Canada (NSERC) grant RGPIN-2017-04816, 
-and the Canada Research Chair program. 
+and the Canada Research Chair program to SG. 
 This research was supported by an NIH grant
-R35GM133531 (to BMH). The content is solely the responsibility of the authors
+R35GM133531 to BMH. The content is solely the responsibility of the authors
 and does not necessarily represent the official views of the National
-Institutes of Health.
+Institutes of Health. MM and EH acknowledge the support of the DSI-NRF Centre of Excellence for 
+Biomedical Tuberculosis Research, South African Medical Research Council Centre for Tuberculosis 
+Research, Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, 
+Stellenbosch University, Cape Town, South Africa.
 
@@
 APR, BMH, and SG designed the study.
-BMH, EGA, EH, and MM designed and conducted participant recruitment.
+BMH, EH, and MM designed and conducted participant recruitment.
+BMH and EGA performed quality control of data.
 APR, BMH, and SG designed statistical analyses.
@@
 Nama sequencing data are available from the European Genome-Phenome Archive
-(EGA), accession number: EGAD00001006198. The African Diversity Reference Panel
-can be found at accession: EGAS00001000960.
+(EGA), accession number: EGAD00001006198. Data access is permitted for non-commercial, 
+population origins or ancestry research
+upon application to the South African Data Access Committee with appropriate IRB approval. 
+The African Diversity Reference Panel can be found at accession: EGAS00001000960.
 

### 2023-03-02 paper.tex
more funding acknowledgements

@@
 \ref{fig:best-fit-models}. This research was supported by CIHR project grant 437576,
-Natural Sciences and Engineering Research Council of Canada (NSERC) grant RGPIN-2017-04816, 
-and the Canada Research Chair program to SG. 
+Natural Sciences and Engineering Research Council of Canada (NSERC) grant RGPIN-2017-04816, the Canada Research Chair program to SG, and the Canada Foundation for Innovation. 
 This research was supported by an NIH grant

### 2023-03-03 paper.tex
Addressing Editor’s comments

@@
 
-Evidence for archaic hominin admixture in Eurasia has bolstered the
-plausibility of archaic hominin admixture having also occurred in Africa.
-Previous work that sought to explain patterns of polymorphism inconsistent with
-a single-origin model thus focused on archaic hominin admixture as an
-alternative model. The additional (or ``ghost'') branches required to fit the
-data are systematically referred to as ``archaic''
-\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
-Lorente-Galdos2019-vz,Durvasula2020-td}, and assumed (or inferred) to be deeply
-diverged. We believe that this has strongly oriented the interpretation of
-these results (e.g., regarding selection \cite{Wall2019-ao} or the fossil
-record \cite{Grun2020-su}). Here, we have shown that a model of a weakly
-structured stem better captures the apparently inconsistent patterns of
-polymorphisms.
+Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
+archaic hominin admixture having also occurred in Africa. Previous work that sought to 
+explain patterns of polymorphism inconsistent with a single-origin model thus focused on 
+archaic hominin admixture as an alternative model, by referring to additional 
+(or ``ghost'') branches required to fit the data as ``archaic'' 
+\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td} 
+and assuming (or inferring) deep divergences. This perspective has oriented interpretations 
+of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil (e.g., evolution of 
+\emph{Homo sapiens} \cite{Grun2020-su} and late survival of archaic hominins and gene 
+flow from them \tdwcomment{add Harvati et al. 2011 ``The Later Stone Age Calvaria from 
+Iwo Eleru, Nigeria: Morphology and Chronology”, Crevecoeur 2012 ``The Upper Paleolithic 
+Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity”, Crevecoeur 
+et al. 2016 ``Late Stone Age human remains from Ishango (Democratic Republic of Congo): 
+New insights on Late Pleistocene modern human diversity in Africa”)}. Here, we have shown 
+that a model of a weakly structured stem better captures the apparently inconsistent 
+patterns of polymorphisms.
 
-Preferring weakly-structured-stem models over archaic-admixture models
-potentially has a range of implications. First, with a weakly structured stem,
-there is no need to posit that an archaic hominin population in Africa stayed
-reproductively isolated from the ancestral human lineage for hundreds of thousands of
-years before the initiation of gene flow. Instead, there would simply have
-been continuous or recurrent contact between two or more groups present in
-Africa.
+Preferring weakly-structured-stem models over archaic-admixture models potentially has 
+a range of implications. First, with a weakly structured stem, there is no need to posit 
+that an archaic hominin population in Africa stayed reproductively isolated from the 
+ancestral human lineage for hundreds of thousands of years before the initiation of 
+gene flow. Instead, there would simply have been continuous or recurrent contact between 
+two or more groups present in Africa. Additionally, because we would have expected to 
+detect admixture from late surviving archaic hominins, the distinctive morphologies of 
+certain fairly recent \emph{Homo sapiens} fossils (\tdwcomment{add Harvati et al. 2011 
+``The Later Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology”, 
+Crevecoeur 2012 ``The Upper Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and 
+Past Modern Human Diversity”, Crevecoeur et al. 2016 ``Late Stone Age human remains 
+from Ishango (Democratic Republic of Congo): New insights on Late Pleistocene modern 
+human diversity in Africa”)}) may have causes other than admixture \citep{Bergstrom2021-iw.
 

### 2023-03-03 paper.tex
Update paper.tex

@@
 flow from them \tdwcomment{add Harvati et al. 2011 ``The Later Stone Age Calvaria from 
-Iwo Eleru, Nigeria: Morphology and Chronology”, Crevecoeur 2012 ``The Upper Paleolithic 
-Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity”, Crevecoeur 
+Iwo Eleru, Nigeria: Morphology and Chronology'', Crevecoeur 2012 ``The Upper Paleolithic 
+Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity'', Crevecoeur 
 et al. 2016 ``Late Stone Age human remains from Ishango (Democratic Republic of Congo): 
-New insights on Late Pleistocene modern human diversity in Africa”)}. Here, we have shown 
+New insights on Late Pleistocene modern human diversity in Africa''}). Here, we have shown 
 that a model of a weakly structured stem better captures the apparently inconsistent 
@@
 certain fairly recent \emph{Homo sapiens} fossils (\tdwcomment{add Harvati et al. 2011 
-``The Later Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology”, 
+``The Later Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology'', 
 Crevecoeur 2012 ``The Upper Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and 
-Past Modern Human Diversity”, Crevecoeur et al. 2016 ``Late Stone Age human remains 
+Past Modern Human Diversity'', Crevecoeur et al. 2016 ``Late Stone Age human remains 
 from Ishango (Democratic Republic of Congo): New insights on Late Pleistocene modern 
-human diversity in Africa”)}) may have causes other than admixture \citep{Bergstrom2021-iw.
+human diversity in Africa''}) may have causes other than admixture \citep{Bergstrom2021-iw.
 

### 2023-03-03 paper.tex
Update paper.tex

@@
 \emph{Homo sapiens} \cite{Grun2020-su} and late survival of archaic hominins and gene 
-flow from them \tdwcomment{add Harvati et al. 2011 ``The Later Stone Age Calvaria from 
+flow from them ADD Harvati et al. 2011 ``The Later Stone Age Calvaria from 
 Iwo Eleru, Nigeria: Morphology and Chronology'', Crevecoeur 2012 ``The Upper Paleolithic 
 Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity'', Crevecoeur 
-et al. 2016 ``Late Stone Age human remains from Ishango (Democratic Republic of Congo): 
-New insights on Late Pleistocene modern human diversity in Africa''}). Here, we have shown 
+et al. 2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): 
+New Insights on Late Pleistocene Modern Human Diversity in Africa''}). Here, we have shown 
 that a model of a weakly structured stem better captures the apparently inconsistent 
@@
 detect admixture from late surviving archaic hominins, the distinctive morphologies of 
-certain fairly recent \emph{Homo sapiens} fossils (\tdwcomment{add Harvati et al. 2011 
+certain fairly recent \emph{Homo sapiens} fossils (ADD Harvati et al. 2011 
 ``The Later Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology'', 
 Crevecoeur 2012 ``The Upper Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and 
-Past Modern Human Diversity'', Crevecoeur et al. 2016 ``Late Stone Age human remains 
-from Ishango (Democratic Republic of Congo): New insights on Late Pleistocene modern 
-human diversity in Africa''}) may have causes other than admixture \citep{Bergstrom2021-iw.
+Past Modern Human Diversity'', Crevecoeur et al. 2016 ``Late Stone Age Human Remains 
+from Ishango (Democratic Republic of Congo): New Insights on Late Pleistocene Modern 
+Human Diversity in Africa''}) may have causes other than admixture \citep{Bergstrom2021-iw.
 

### 2023-03-03 paper.tex
Update paper.tex

@@
 from Ishango (Democratic Republic of Congo): New Insights on Late Pleistocene Modern 
-Human Diversity in Africa''}) may have causes other than admixture \citep{Bergstrom2021-iw.
+Human Diversity in Africa''}) may have causes other than admixture \citep{Bergstrom2021-iw}.
 

### 2023-03-03 paper.tex
Update paper.tex

--- 2023-03-03 paper.tex
@@
 et al. 2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): 
-New Insights on Late Pleistocene Modern Human Diversity in Africa''}). Here, we have shown 
+New Insights on Late Pleistocene Modern Human Diversity in Africa''). Here, we have shown 
 that a model of a weakly structured stem better captures the apparently inconsistent 
@@
 from Ishango (Democratic Republic of Congo): New Insights on Late Pleistocene Modern 
-Human Diversity in Africa''}) may have causes other than admixture \citep{Bergstrom2021-iw.
+Human Diversity in Africa'') may have causes other than admixture \citep{Bergstrom2021-iw}.
 

### 2023-03-03 paper.tex
Update paper.tex

@@
 \emph{Homo sapiens} \cite{Grun2020-su} and late survival of archaic hominins and gene 
-flow from them ADD Harvati et al. 2011 ``The Later Stone Age Calvaria from 
+flow from them \textcolor{red}{ADD} Harvati et al. 2011 ``The Later Stone Age Calvaria from 
 Iwo Eleru, Nigeria: Morphology and Chronology'', Crevecoeur 2012 ``The Upper Paleolithic 
@@
 detect admixture from late surviving archaic hominins, the distinctive morphologies of 
-certain fairly recent \emph{Homo sapiens} fossils (ADD Harvati et al. 2011 
+certain fairly recent \emph{Homo sapiens} fossils (\textcolor{red}{ADD} Harvati et al. 2011 
 ``The Later Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology'', 

### 2023-03-03 paper.tex
Addressing Editor's comments

--- 2023-03-03 paper.tex
@@
 \emph{Homo sapiens} \cite{Grun2020-su} and late survival of archaic hominins and gene 
-flow from them ADD Harvati et al. 2011 ``The Later Stone Age Calvaria from 
-Iwo Eleru, Nigeria: Morphology and Chronology'', Crevecoeur 2012 ``The Upper Paleolithic 
-Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity'', Crevecoeur 
-et al. 2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): 
-New Insights on Late Pleistocene Modern Human Diversity in Africa''). Here, we have shown 
-that a model of a weakly structured stem better captures the apparently inconsistent 
+flow from them \textcolor{red}{ADD REFERENCES} Harvati et al. 2011 ``The Later Stone Age 
+Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology'', Crevecoeur 2012 ``The Upper 
+Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity'', 
+Crevecoeur et al. 2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of 
+Congo): New Insights on Late Pleistocene Modern Human Diversity in Africa''). Here, we have 
+shown that a model of a weakly structured stem better captures the apparently inconsistent 
 patterns of polymorphisms.
@@
 detect admixture from late surviving archaic hominins, the distinctive morphologies of 
-certain fairly recent \emph{Homo sapiens} fossils (ADD Harvati et al. 2011 
-``The Later Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology'', 
-Crevecoeur 2012 ``The Upper Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and 
-Past Modern Human Diversity'', Crevecoeur et al. 2016 ``Late Stone Age Human Remains 
-from Ishango (Democratic Republic of Congo): New Insights on Late Pleistocene Modern 
-Human Diversity in Africa'') may have causes other than admixture \citep{Bergstrom2021-iw}.
+certain fairly recent \emph{Homo sapiens} fossils (\textcolor{red}{ADD REFERENCES} 
+Harvati et al. 2011 ``The Later Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology 
+and Chronology'', Crevecoeur 2012 ``The Upper Paleolithic Human Remains of Nazlet Khater 2 
+(Egypt) and Past Modern Human Diversity'', Crevecoeur et al. 2016 ``Late Stone Age Human 
+Remains from Ishango (Democratic Republic of Congo): New Insights on Late Pleistocene 
+Modern Human Diversity in Africa'') may have causes other than admixture \citep{Bergstrom2021-iw}.
 

### 2023-03-04 paper.tex
minor tweak in discussion
Update paper.tex

--- 2023-03-03 paper.tex
@@
 explain patterns of polymorphism inconsistent with a single-origin model thus focused on 
-archaic hominin admixture as an alternative model, by referring to additional 
-(or ``ghost'') branches required to fit the data as ``archaic'' 
-\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td} 
-and assuming (or inferring) deep divergences. This perspective has oriented interpretations 
-of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil (e.g., evolution of 
-\emph{Homo sapiens} \cite{Grun2020-su} and late survival of archaic hominins and gene 
-flow from them ADD Harvati et al. 2011 ``The Later Stone Age Calvaria from 
-Iwo Eleru, Nigeria: Morphology and Chronology'', Crevecoeur 2012 ``The Upper Paleolithic 
-Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity'', Crevecoeur 
-et al. 2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): 
-New Insights on Late Pleistocene Modern Human Diversity in Africa''). Here, we have shown 
-that a model of a weakly structured stem better captures the apparently inconsistent 
-patterns of polymorphisms.
+archaic hominin admixture as an alternative model, by referring to additional (or ``ghost'') 
+branches required to fit the data as ``archaic'' 
+\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,
+Durvasula2020-td} and assuming (or inferring) deep divergences. These perspectives have 
+oriented interpretations of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil 
+(e.g., evolution of \emph{Homo sapiens} \cite{Grun2020-su}) data. Here, we have shown that 
+a model of a weakly structured stem better captures the apparently inconsistent patterns 
+of polymorphisms.
 
-Preferring weakly-structured-stem models over archaic-admixture models potentially has 
-a range of implications. First, with a weakly structured stem, there is no need to posit 
-that an archaic hominin population in Africa stayed reproductively isolated from the 
-ancestral human lineage for hundreds of thousands of years before the initiation of 
-gene flow. Instead, there would simply have been continuous or recurrent contact between 
-two or more groups present in Africa. Additionally, because we would have expected to 
-detect admixture from late surviving archaic hominins, the distinctive morphologies of 
-certain fairly recent \emph{Homo sapiens} fossils (ADD Harvati et al. 2011 
-``The Later Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology'', 
-Crevecoeur 2012 ``The Upper Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and 
-Past Modern Human Diversity'', Crevecoeur et al. 2016 ``Late Stone Age Human Remains 
-from Ishango (Democratic Republic of Congo): New Insights on Late Pleistocene Modern 
-Human Diversity in Africa'') may have causes other than admixture \citep{Bergstrom2021-iw}.
+Preferring weakly-structured-stem models over archaic-admixture models potentially has a 
+range of implications. First, with a weakly structured stem, there is no need to posit that 
+an archaic hominin population in Africa stayed reproductively isolated from the ancestral 
+human lineage for hundreds of thousands of years before the initiation of gene flow. Instead, 
+there would simply have been continuous or recurrent contact between two or more groups 
+present in Africa.
 

### 2023-03-04 paper.tex
Update paper.tex

@@
 
-Third, multiple studies have shown a correspondence between phenotypic
-differentiation, usually assessed with measurements of the cranium, and genetic
-differentiation among human populations and between humans and Neanderthals
-\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI
-Section~\ref{sec:morpho_diff}). This correspondence potentially allows
-predictions of our model to be related to the fossil record. The fossil record
-of Africa is sparse during the time period of the stems, but of the fossils
-that date to this time period, some are fairly similar overall in morphology to
-contemporary humans (e.g., Omo 1 from Omo Kibish, Ethiopia
-\citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological
-features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco
-\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in
-morphology to have been assigned to species other than \emph{Homo sapiens}
-(e.g., DH1 from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}). If,
-as our model predicts, the genetic differences between the stems were
-comparable to those among contemporary human populations, the most
-morphologically divergent fossils are unlikely to represent branches that
-contributed appreciably to contemporary human ancestries.
+Third, multiple studies have shown a correspondence between phenotypic differentiation, 
+usually assessed with measurements of the cranium, and genetic differentiation among 
+human populations and between humans and Neanderthals \citep{Relethford1994-mh,
+Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI Section~\ref{sec:morpho_diff}). 
+This correspondence potentially allows predictions of our model to be related to the 
+fossil record. \emph{Homo sapiens} fossils from Iho Eleru, Nigeria (13ka) 
+\citep{Bergstrom2021-iw} \textcolor{red}{ADD REFERENCE} Harvati et al. 2011 ``The Later 
+Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology'', Ishango, 
+Democratic Republic of Congo (25-20ka) \textcolor{red}{ADD REFERENCE} Crevecoeur et al. 
+2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): New 
+Insights on Late Pleistocene Modern Human Diversity in Africa'', and Nazlet Khater, 
+Egypt (40-35ka) \textcolor{red}{ADD REFERENCE} Crevecoeur 2012 ``The Upper 
+Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity'' 
+have morphological features that may indicate gene flow from archaic hominins. If 
+this interpretation is correct, these fossils are unlikely to represent populations 
+ancestral to contemporary human populations, because we would have expected to detect 
+admixture from late surviving archaic hominins. In our multiple-merger model, we do 
+infer late gene flow from one of the stems into West African populations (25ka), but 
+this is not admixture from an archaic hominin lineage. The fossil record of Africa is 
+sparse during the time period of the stems, but of the fossils that date to this time 
+period, some are fairly similar overall in morphology to contemporary humans (e.g., 
+Omo 1 from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are similar 
+in some morphological features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, 
+Morocco \citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in 
+morphology to have been assigned to species other than \emph{Homo sapiens} (e.g., DH1 
+from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}). If, as our model 
+predicts, the genetic differences between the stems were comparable to those among 
+contemporary human populations, the most morphologically divergent fossils are unlikely 
+to represent branches that contributed appreciably to contemporary human ancestries.
 

### 2023-03-04 paper.tex
Update paper.tex

@@
 fossil record. \emph{Homo sapiens} fossils from Iho Eleru, Nigeria (13ka) 
-\citep{Bergstrom2021-iw} \textcolor{red}{ADD REFERENCE} Harvati et al. 2011 ``The Later 
-Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology'', Ishango, 
-Democratic Republic of Congo (25-20ka) \textcolor{red}{ADD REFERENCE} Crevecoeur et al. 
+\citep{Bergstrom2021-iw} \textcolor{red}{ADD REFERENCE Harvati et al. 2011 ``The Later 
+Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology''}, Ishango, 
+Democratic Republic of Congo (25-20ka) \textcolor{red}{ADD REFERENCE Crevecoeur et al. 
 2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): New 
-Insights on Late Pleistocene Modern Human Diversity in Africa'', and Nazlet Khater, 
-Egypt (40-35ka) \textcolor{red}{ADD REFERENCE} Crevecoeur 2012 ``The Upper 
-Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity'' 
+Insights on Late Pleistocene Modern Human Diversity in Africa''}, and Nazlet Khater, 
+Egypt (40-35ka) \textcolor{red}{ADD REFERENCE Crevecoeur 2012 ``The Upper 
+Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity''} 
 have morphological features that may indicate gene flow from archaic hominins. If 

### 2023-03-04 paper.tex
Revisions to address Editor's comments

--- 2023-03-04 paper.tex
@@
 oriented interpretations of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil 
-(e.g., evolution of \emph{Homo sapiens} \cite{Grun2020-su}) data. Here, we have shown that 
-a model of a weakly structured stem better captures the apparently inconsistent patterns 
+(e.g., evolution of early \emph{Homo sapiens} \cite{Grun2020-su}) data. Here, we have shown 
+that a model of a weakly structured stem better captures the apparently inconsistent patterns 
 of polymorphisms.
@@
 
-Third, multiple studies have shown a correspondence between phenotypic
-differentiation, usually assessed with measurements of the cranium, and genetic
-differentiation among human populations and between humans and Neanderthals
-\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI
-Section~\ref{sec:morpho_diff}). This correspondence potentially allows
-predictions of our model to be related to the fossil record. The fossil record
-of Africa is sparse during the time period of the stems, but of the fossils
-that date to this time period, some are fairly similar overall in morphology to
-contemporary humans (e.g., Omo 1 from Omo Kibish, Ethiopia
-\citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological
-features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco
-\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in
-morphology to have been assigned to species other than \emph{Homo sapiens}
-(e.g., DH1 from Dinaledi, South Africa \citep{Berger2015-bq,Dirks2017-uk}). If,
-as our model predicts, the genetic differences between the stems were
-comparable to those among contemporary human populations, the most
-morphologically divergent fossils are unlikely to represent branches that
+Third, multiple studies have shown a correspondence between phenotypic differentiation, 
+usually assessed with measurements of the cranium, and genetic differentiation among 
+human populations and between humans and Neanderthals \citep{Relethford1994-mh,
+Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI Section~\ref{sec:morpho_diff}). 
+This correspondence potentially allows predictions of our model to be related to the 
+fossil record. Some \emph{Homo sapiens} fossils, from Iho Eleru, Nigeria (13ka) 
+\citep{Bergstrom2021-iw} \textcolor{red}{ADD REFERENCE Harvati et al. 2011 ``The Later 
+Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology''}, Ishango, 
+Democratic Republic of Congo (25-20ka) \textcolor{red}{ADD REFERENCE Crevecoeur et al. 
+2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): New 
+Insights on Late Pleistocene Modern Human Diversity in Africa''}, and Nazlet Khater, 
+Egypt (40-35ka) \textcolor{red}{ADD REFERENCE Crevecoeur 2012 ``The Upper 
+Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity''}, 
+have morphological features that may reflect gene flow from archaic hominins. If 
+this interpretation is correct, these fossils are unlikely to represent populations 
+ancestral to contemporary human populations, because we would have expected to detect 
+this gene flow and we do not. In our multiple-merger model, we do infer late gene flow 
+from one of the stems into West African populations at 25ka, but this is not admixture 
+with an archaic hominin lineage. The fossil record of Africa is sparse during the time 
+period of the stems, but of the fossils that date to this time period, some are fairly 
+similar overall in morphology to contemporary humans (e.g., Omo 1 from Omo Kibish, Ethiopia 
+\citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological features to 
+contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco 
+\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in morphology to 
+have been assigned to species other than \emph{Homo sapiens} (e.g., DH1 from Dinaledi, 
+South Africa \citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
+differences between the stems were comparable to those among contemporary human populations, 
+the most morphologically divergent fossils are unlikely to represent branches that 
 contributed appreciably to contemporary human ancestries.

### 2023-03-04 paper.tex
Minor additional edit

@@
 Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology''}, Ishango, 
-Democratic Republic of Congo (25-20ka) \textcolor{red}{ADD REFERENCE Crevecoeur et al. 
+Democratic Republic of Congo (20-25ka) \textcolor{red}{ADD REFERENCE Crevecoeur et al. 
 2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): New 

### 2023-03-04 paper.tex
More minor changes

@@
 Insights on Late Pleistocene Modern Human Diversity in Africa''}, and Nazlet Khater, 
-Egypt (40-35ka) \textcolor{red}{ADD REFERENCE Crevecoeur 2012 ``The Upper 
+Egypt (35-40ka) \textcolor{red}{ADD REFERENCE Crevecoeur 2012 ``The Upper 
 Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity''}, 
@@
 ancestral to contemporary human populations, because we would have expected to detect 
-this gene flow and we do not. In our multiple-merger model, we do infer late gene flow 
-from one of the stems into West African populations at 25ka, but this is not admixture 
-with an archaic hominin lineage. The fossil record of Africa is sparse during the time 
-period of the stems, but of the fossils that date to this time period, some are fairly 
-similar overall in morphology to contemporary humans (e.g., Omo 1 from Omo Kibish, Ethiopia 
-\citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological features to 
-contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco 
+this gene flow and it not present in our models. In our multiple-merger model, we do 
+infer late gene flow from one of the stems into West African populations at 25ka, but 
+this is not admixture with an archaic hominin lineage. The fossil record of Africa is 
+sparse during the time period of the stems, but of the fossils that date to this time period, 
+some are fairly similar overall in morphology to contemporary humans (e.g., Omo 1 from 
+Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological 
+features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco 
 \citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in morphology to 

### 2023-03-04 paper.tex
Minor edits

@@
 ancestral to contemporary human populations, because we would have expected to detect 
-this gene flow and it not present in our models. In our multiple-merger model, we do 
+this gene flow, and it is not present in our models. In our multiple-merger model, we do 
 infer late gene flow from one of the stems into West African populations at 25ka, but 

### 2023-03-05 paper.tex
add refs, small edit suggestion

@@
 fossil record. Some \emph{Homo sapiens} fossils, from Iho Eleru, Nigeria (13ka) 
-\citep{Bergstrom2021-iw} \textcolor{red}{ADD REFERENCE Harvati et al. 2011 ``The Later 
-Stone Age Calvaria from Iwo Eleru, Nigeria: Morphology and Chronology''}, Ishango, 
-Democratic Republic of Congo (20-25ka) \textcolor{red}{ADD REFERENCE Crevecoeur et al. 
-2016 ``Late Stone Age Human Remains from Ishango (Democratic Republic of Congo): New 
-Insights on Late Pleistocene Modern Human Diversity in Africa''}, and Nazlet Khater, 
-Egypt (35-40ka) \textcolor{red}{ADD REFERENCE Crevecoeur 2012 ``The Upper 
-Paleolithic Human Remains of Nazlet Khater 2 (Egypt) and Past Modern Human Diversity''}, 
+\citep{Bergstrom2021-iw,Harvati2011-xs}, Ishango, 
+Democratic Republic of Congo (20-25ka) \citep{Crevecoeur2016-cl}, and Nazlet Khater, 
+Egypt (35-40ka) \citep{Crevecoeur2012-we}, 
 have morphological features that may reflect gene flow from archaic hominins. If 
@@
 infer late gene flow from one of the stems into West African populations at 25ka, but 
-this is not admixture with an archaic hominin lineage. The fossil record of Africa is 
+this should not be interpreted as admixture with an archaic hominin lineage as
+differentiation between stems remains low. The fossil record of Africa is 
 sparse during the time period of the stems, but of the fossils that date to this time period, 
@@
 visualization discussion and designing Figures \ref{fig:proposed-models} and
-\ref{fig:best-fit-models}. This research was supported by CIHR project grant 437576,
-Natural Sciences and Engineering Research Council of Canada (NSERC) grant RGPIN-2017-04816, the Canada Research Chair program to SG, and the Canada Foundation for Innovation. 
-This research was supported by an NIH grant
-R35GM133531 to BMH. The content is solely the responsibility of the authors
-and does not necessarily represent the official views of the National
-Institutes of Health. MM and EH acknowledge the support of the DSI-NRF Centre of Excellence for 
-Biomedical Tuberculosis Research, South African Medical Research Council Centre for Tuberculosis 
-Research, Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, 
-Stellenbosch University, Cape Town, South Africa.
+\ref{fig:best-fit-models}. This research was supported by CIHR project grant
+437576, Natural Sciences and Engineering Research Council of Canada (NSERC)
+grant RGPIN-2017-04816, the Canada Research Chair program to SG, and the Canada
+Foundation for Innovation. This research was supported by an NIH grant
+R35GM133531 to BMH. The content is solely the responsibility of the authors and
+does not necessarily represent the official views of the National Institutes of
+Health. MM and EH acknowledge the support of the DSI-NRF Centre of Excellence
+for Biomedical Tuberculosis Research, South African Medical Research Council
+Centre for Tuberculosis Research, Division of Molecular Biology and Human
+Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University,
+Cape Town, South Africa.
 

### 2023-03-06 paper.tex
clarification of stem period

@@
 \usepackage{booktabs}
-
+\usepackage{amssymb}
 \usepackage{authblk}
@@
 fossil record. Some \emph{Homo sapiens} fossils, from Iho Eleru, Nigeria (13ka) 
-\citep{Bergstrom2021-iw,Harvati2011-xs}, Ishango, 
+\citep{Harvati2011-xs}, Ishango, 
 Democratic Republic of Congo (20-25ka) \citep{Crevecoeur2016-cl}, and Nazlet Khater, 
@@
 differentiation between stems remains low. The fossil record of Africa is 
-sparse during the time period of the stems, but of the fossils that date to this time period, 
+sparse during the time period of the stems \aprcomment{($\gtrsim120$ka)},
+but of the fossils that date to this time period, 
 some are fairly similar overall in morphology to contemporary humans (e.g., Omo 1 from 

### 2023-03-07 paper.tex
Minor wording change

@@
 this gene flow, and it is not present in our models. In our multiple-merger model, we do 
-infer late gene flow from one of the stems into West African populations at 25ka, but 
+infer gene flow at 25ka from one of the stems into West African populations, but 
 this should not be interpreted as admixture with an archaic hominin lineage as
-differentiation between stems remains low. The fossil record of Africa is 
+differentiation between the stems remains low. The fossil record of Africa is 
 sparse during the time period of the stems \aprcomment{($\gtrsim120$ka)},

### 2023-03-10 paper.tex
tried to clarify the logical flow of conclusion

@@
 Egypt (35-40ka) \citep{Crevecoeur2012-we}, 
-have morphological features that may reflect gene flow from archaic hominins. If 
+have morphological features that may reflect relatively recent gene flow from archaic hominins \sgcomment{Could we have a citation for this?}. If 
 this interpretation is correct, these fossils are unlikely to represent populations 
@@
 differentiation between the stems remains low. The fossil record of Africa is 
-sparse during the time period of the stems \aprcomment{($\gtrsim120$ka)},
+sparse during the earlier time period of the stems \aprcomment{($\gtrsim120$ka)},
 but of the fossils that date to this time period, 
@@
 differences between the stems were comparable to those among contemporary human populations, 
-the most morphologically divergent fossils are unlikely to represent branches that 
+the most morphologically divergent fossils are also unlikely to represent branches that 
 contributed appreciably to contemporary human ancestries.

### 2023-03-11 supplement.tex
add some methods
add Brenna's supp section

@@
 \section{Data and sequencing}
+\label{sec:data-sequencing}
 
@@
 From the ADRP, these included the Amhara, Bakiga, Gumuz, Nama, Oromo, Somali,
-Nama, and Zulu. This larger set of populations were used in the PCA and
+and Zulu. This larger set of populations were used in the PCA and
 \texttt{Admixture} analyses. The \texttt{Relate} analysis used a slightly
@@
 
-\subsubsection{Fixing the East African and OOA divergence times}
-
-In our preliminary models, the unconstrained inferred split times between eastern
-African populations and Europeans was inferred to have occurred around $40$ka,
-and the inferred split between eastern and western African populations occurred about
-$10$ka earlier. These recent split dates are at odds with paleoanthropological 
-evidence for an earlier Out-of-Africa (OOA) expansion
-\citep[e.g.][]{Fu2014-am,Hublin2020-nk} and with the estimated date
-for Neanderthal admixture into Eurasian populations
-\citep[e.g.][]{Reilly2022-ym}. The parameter
-uncertainty of our inferred divergence times (which accounts for the
-non-independence of sites) does not account for these differences: in the
-high-likelihood models, the $95\%$ confidence intervals for the eastern
-African-European split was roughly $38 \pm 5$ka, and for the eastern/western African
-split was $45 \pm 10$ka, with the uncertainties reflecting 2 standard
-deviations according to bootstrap of genomic regions. We therefore expect these
-discrepancies to result from model misspecification in the details of the
-divergence between the ancestors of OOA and eastern African populations.
-
-Our models suppose a single split leading to the OOA event followed by constant
-symmetric migrations between eastern African and OOA populations, which is an
-oversimplification. Previous studies that have focused on reconstructing the
-OOA divergence time and subsequent migration rates typically included more
-Eurasian populations and multiple epochs that allow for differing continuous
-migration rates and population sizes. Since a detailed model for the OOA (and
-Back-to-Africa) event was not our main focus, we chose to keep this simplified
-model, but fix the East African-Eurasian split at $50$ka to be older than 
-archaeological dates for when the OOA reached Europe 
-\citep{Fewlass2020-se,Hublin2020-nk}. This also allowed us to include 
-Neanderthal admixture into the OOA population at $45$ka.
-
-Fixing these parameters (i.e., East Africa-Eurasian split at 50ka, Neanderthal
-admixture at 45ka) resulted in an estimated value for the split between the
-ancestors of East and West African populations at around $60$ka. This timing
-agrees with previous LD-based estimates of West African-Eurasian divergence
-times, of $\sim61$ka ($50-64$ka, $95\%$ confidence interval)
-\citep{Ragsdale2019-nt}, which used a different set populations than those in
-this study. A date of $60$ka is also in close agreement with inferred
-divergence times of $\sim65$ka from both SFS-based demographic inference, using
-synonymous variation from a subset of populations used in this study (Mende,
-British, and Han Chinese), as well as that of LD-based inference using these
-same groups, of $53$ka ($43-63$ka, $95\%$ confidence interval). Taken
-altogether, we decided to fix this value ($60$ka) in most subsequent analyses.
-We discuss the sensitivity of our model to these assumptions in
-Section~\ref{sec:change-dates}.
+\subsubsection{Evidence for the Out of Africa timing}
+
+Above (Section~\ref{sec:models}), we assume that the eastern Africa-Eurasian
+population divergence occurs at 50ka and Neanderthal admixture into this
+Eurasian ancestral population occurs at 45ka. The choice of these dates
+deserves additional discussion and consideration. Our focus is on the possible
+expansion Out of Africa (OOA) that led to a sustained, widespread population
+from whom present-day OOA populations are derived.
+
+In this context, a number of possible models have been outlined
+\citep{Groucutt2015-do}. Among these, \citeauthor{Groucutt2015-do} enumerate
+variants on an early onset dispersal model. There is persuasive
+paleo-morphological evidence for a Homo sapiens presence in Arabia between
+120--85ka, coincident with Marine Isotope Stage (MIS) 5. The crania identified
+at Skhul and Qafzeh in the Levant, as well a finger bone from southern Arabian
+Peninsula point to a dispersal OOA when conditions in North Africa facilitated
+a connection through the Sinai land bridge, and/or a possible connection via
+crossing the Red Sea. Ecological conditions would have been similar between
+North Africa and the Arabian Peninsula at this time, with greater abundance in
+game during the higher rainfall MIS 5. While Skhul and Qafzeh are taxonomically
+assigned to Homo sapiens, we note that their morphological diversity (i.e. a
+mix of derived and archaic traits) is unique among contemporary human
+populations. Under some formulations of the early onset dispersal model,
+archaeologists posit a further expansion of humans into South Asia or as far as
+southeastern Asia \citep{Groucutt2015-do}.
+
+However, the replacement of Homo sapiens by Neanderthals in the Levant (e.g.
+Kebara 2) complicates the persistence of modern humans in the Arabian Peninsula
+after 85ka. The end of MIS 5 coincides with increasing aridity in the region
+which would have rendered much the Peninsula uninhabitable, except for a
+secluded southern refugia, decreasing the probability of migrations through the
+Sinai land bridge \citep{Beyer2021-ym}. The northern route OOA remains a likely
+barrier to human migration (as modeled by precipitation amount) until the
+Holocene. A southern route across the Red Sea becomes possible during MIS 4
+(70ka) until 30ka \citep{Beyer2021-ym}. Even during MIS 4, there were few
+plausible routes out of the southern Arabian Peninsula, except along the Gulf
+of Oman or gaps in the Arabian Desert through the Zagros Mountains. Neanderthal
+remains dating to this period have been identified in the Zagros Mountains,
+adding a further potential barrier to expansion
+\citep{Yousefi2020-et,Heydari-Guran2021-xy}.
+
+Thus, either an incipient OOA population must have expanded beyond the Near
+East by 85ka and then given rise to today’s Eurasian populations or a
+successful expansion would likely post-date 65ka. Morphological and
+archaeological evidence for Homo sapiens in the Near East reappears by 50ka in
+the Levant with the Initial Upper Paleolithic stone tool industry
+\citep{Boaretto2021-oi} and a cranium from Manot Cave
+\citep{Hershkovitz2015-lq}. There is also strong consensus that a widespread
+human presence in Europe and Asia is dated to after 45ka. Archaeological
+\citep[e.g.,][]{Fewlass2020-se}, morphological \citep[e.g.,][]{Hublin2020-nk},
+and genetic data \citep[e.g.,][]{Prufer2021-ao} all indicate that about 45ka
+\emph{Homo sapiens} had expanded into eastern Europe and within 5,000 years
+rapidly colonized the entire span of southern Europe. This period further
+coincides with the dates for the last Neanderthal presence in Europe at
+approximately 40ka \citep{Deviese2017-de,Hublin2017-uy}. Similarly, by 45ka, an
+unequivocal \emph{H. sapiens} femur from southern Siberia contains derived
+alleles shared by later Eurasian populations \citep{Fu2014-am}. The Ust’-Ishim
+specimen contains evidence of admixture with Neanderthals, at a level similar
+to contemporary East Asian genomes. This supports estimates of Neanderthal
+admixture occurring earlier than 45ka, estimated to be by 47-65ka as inferred
+from Neanderthal tract lengths \citep{Sankararaman2012-uw}.
+
+In summary, a sustained, rapid human expansion across Eurasia dates to just
+prior to 45ka and emerges from a Near Eastern population. The time of
+population divergence between eastern Africans and this Near Eastern group is
+difficult to pinpoint. An exit OOA and persistence prior to 65ka appears
+unlikely to have contributed to the ancestors of most Eurasians. By 50ka, a
+Homo sapiens population occupied the Levant, subsequently admixed and replaced
+Neanderthals, and is a likely source for the Eurasian populations. Hence, we
+fixed the eastern African/Eurasian divergence at 50ka, though arguably dates
+even earlier up to 65ka would be appropriate (see SI Section 7.2).
+
+We tested our models for sensitivity to a subset of the fixed parameters,
+including those relevant to the OOA timing. When unconstrained, the inferred
+population split times between eastern African populations and Europeans
+occurred around 40ka, and the inferred split between eastern and western
+African populations occurred about 10ka earlier. The 95\% CI for the inferred
+divergence times (which accounts for the non-independence of sites) does not
+contain the values of the fixed estimates. Under the high-likelihood models,
+the 95\% confidence intervals for the eastern African-European divergence were
+roughly $38\pm5$ka, and for the eastern/western African divergence was
+$45\pm10$ka, with the uncertainties reflecting 2 standard deviations according
+to bootstrap of genomic regions.
+
+We hypothesize that these discrepancies result from model misspecification
+related to the divergence between the ancestors of OOA and eastern African
+populations. As is typical, our models suppose a single split leading to the
+OOA event followed by constant, symmetric migration between eastern African and
+OOA populations until 12ka. We consider the British population (GBR) as a proxy
+for all Eurasians populations, in part to avoid adding many additional
+parameters associated with including 2 or 3 Eurasian populations. Further, our
+study and most others lack representative genomes from the Arabian Peninsula
+and Levant.
 
@@
 mis-specification (as discussed in Section~\ref{sec:models}).
+
+In summary, fixing these parameters (i.e., East Africa-Eurasian split at 50ka,
+Neanderthal admixture at 45ka, see Section~\ref{sec:models}) resulted in an
+estimated value for the split between the ancestors of East and West African
+populations at around $60$ka. This timing agrees with previous LD-based
+estimates of West African-Eurasian divergence times, of $\sim61$ka ($50-64$ka,
+$95\%$ confidence interval) \citep{Ragsdale2019-nt}, which used a different set
+populations than those in this study. A date of $60$ka is also in close
+agreement with inferred divergence times of $\sim65$ka from both SFS-based
+demographic inference, using synonymous variation from a subset of populations
+used in this study (Mende, British, and Han Chinese), as well as that of
+LD-based inference using these same groups, of $53$ka ($43-63$ka, $95\%$
+confidence interval). Taken altogether, we decided to fix this value ($60$ka)
+in most subsequent analyses.
 

### 2023-03-11 paper.tex

@@
 Egypt (35-40ka) \citep{Crevecoeur2012-we}, 
-have morphological features that may reflect relatively recent gene flow from archaic hominins \sgcomment{Could we have a citation for this?}. If 
-this interpretation is correct, these fossils are unlikely to represent populations 
-ancestral to contemporary human populations, because we would have expected to detect 
-this gene flow, and it is not present in our models. In our multiple-merger model, we do 
-infer gene flow at 25ka from one of the stems into West African populations, but 
-this should not be interpreted as admixture with an archaic hominin lineage as
-differentiation between the stems remains low. The fossil record of Africa is 
+have morphological features that may reflect relatively recent gene flow from archaic hominins \citep{Harvati2011-xs, Crevecoeur2016-cl}, which has been used in support of previously inferred archaic admixture events in human ancestry \cite{Scerri2018-nl, Durvasula2020-td, Bergstrom2021-iw}.
+The weakly structured stem model is not incompatible with archaic admixture having occurred in the ancestry of these fossils, but would imply, by contrast, that such individuals are unlikely to have contributed much ancestry to  contemporary humans.
+The fossil record of Africa is 
 sparse during the earlier time period of the stems \aprcomment{($\gtrsim120$ka)},

### 2023-03-13 supplement.tex
update section 3.1.1

@@
 coincides with the dates for the last Neanderthal presence in Europe at
-approximately 40ka \citep{Deviese2017-de,Hublin2017-uy}. Similarly, by 45ka, an
-unequivocal \emph{H. sapiens} femur from southern Siberia contains derived
-alleles shared by later Eurasian populations \citep{Fu2014-am}. The Ust’-Ishim
+approximately 40ka \citep{Deviese2017-de,Hublin2017-uy}. Similarly in Asia,
+evidence of the Initial Upper Paleolithic occurs 39-45ka as far east as Lake
+Baikal. Characteristics of the IUP support a coherent, cultural package which
+may have been transmitted by demic expansion \citep{Kuhn2014-kj,Zwyns2021-zv}.
+By 45ka, there is additionally an unequivocal \emph{H. sapiens} femur from
+southern Siberia which contains derived alleles shared with later Eurasian
+populations \citep{Fu2014-am}. The Ust’-Ishim
 specimen contains evidence of admixture with Neanderthals, at a level similar

### 2023-03-13 paper.tex
final edits?

@@
 \affil[1]{Department of Integrative Biology, University of Wisconsin--Madison, WI, USA}
-\affil[2]{Department of Anthropology, University of California, Davis, Davis, CA, USA}
+\affil[2]{Department of Anthropology, University of California, Davis, CA, USA}
 \affil[3]{Department of Molecular and Human Genetics, Baylor College of Medicine, Houston, TX, USA}
 \affil[4]{DSI-NRF Centre of Excellence for Biomedical Tuberculosis Research; South African Medical Research Council Centre for Tuberculosis Research; Division of Molecular Biology and Human Genetics, Faculty of Medicine and Health Sciences, Stellenbosch University, Cape Town, South Africa}
-\affil[5]{UC Davis Genome Center, University of California, Davis, Davis, CA, USA}
+\affil[5]{UC Davis Genome Center, University of California, Davis, CA, USA}
 \affil[6]{Department of Human Genetics, McGill University, Montreal, QC, Canada}
@@
     polymorphism that were previously attributed to contributions from archaic
-    hominins in Africa
-    \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
-    Lorente-Galdos2019-vz,Durvasula2020-td}. In contrast to models with archaic
-    introgression, we predict fossil remains from coexisting ancestral
-    populations should be genetically and morphologically similar, and that an
-    inferred 1--4\% of genetic differentiation among contemporary human
-    populations can be attributed to genetic drift between stem populations. We
-    show that model misspecification explains variation in previous divergence
-    time estimates and argue that studying a suite of models is key to robust
-    inferences about deep history.
+    hominins in Africa \citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,
+    Ragsdale2019-nt, Lorente-Galdos2019-vz,Durvasula2020-td}. In contrast to
+    models with archaic introgression, we predict fossil remains from
+    coexisting ancestral populations should be genetically and morphologically
+    similar, and that an inferred 1--4\% of genetic differentiation among
+    contemporary human populations can be attributed to genetic drift between
+    stem populations. We show that model misspecification explains variation in
+    previous divergence time estimates and argue that studying a suite of
+    models is key to robust inferences about deep history.
 
@@
 River, South Africa \citep{Deacon1995-rx} demonstrate that derived \emph{Homo
-sapiens} anatomical features were found across the continent
-$300-100$ka. Archaeological sites from the Middle Stone Age, of which some have
-been associated with \emph{Homo sapiens}, are also widely distributed across
-Africa. It is unclear whether these fossils and archaeological sites represent
+sapiens} anatomical features were found across the continent $300-100$ka.
+Archaeological sites from the Middle Stone Age, of which some have been
+associated with \emph{Homo sapiens}, are also widely distributed across Africa.
+It is unclear whether these fossils and archaeological sites represent
 populations which contributed to contemporary \emph{Homo sapiens} as population
@@
 population coalescence and fragmentation \citep{Scerri2019-xg}, but these
-approaches are more challenging to interpret and fit to data.
-However, new population genetic tools now allow for inference involving tens to
-hundreds of genomes from multiple populations and greater complexity
+approaches are more challenging to interpret and fit to data.  However, new
+population genetic tools now allow for inference involving tens to hundreds of
+genomes from multiple populations and greater complexity
 \citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
-Neanderthal admixture with humans in Eurasia, several recent studies
-have shown that introducing an archaic hominin ``ghost'' population contributing to 
+Neanderthal admixture with humans in Eurasia, several recent studies have shown
+that introducing an archaic hominin ``ghost'' population contributing to
 African populations in the period surrounding the Out-of-Africa migration event
@@
 Lorente-Galdos2019-vz,Durvasula2020-td},
-%Plagnol: YRI 
-%Here we present a new method for addressing whether archaic human groups contributed to the modern gene pool (called ancient admixture),
-%we find strong evidence for ancient admixture in both a European and a West African population
+%Plagnol: YRI Here we present a new method for addressing whether archaic human
+%groups contributed to the modern gene pool (called ancient admixture), we find
+%strong evidence for ancient admixture in both a European and a West African
+%population
 %
-%Hammer: Mandenka, Biaka, and San. Title:
-%"Genetic evidence for archaic admixture in Africa"
-%Abstract:"allow us to infer that contemporary African populations contain a small proportion of genetic material (≈2%) that introgressed ≈35 kya from an archaic population that split from the ancestors of anatomically modern humans ≈700 kya."
+%Hammer: Mandenka, Biaka, and San. Title: "Genetic evidence for archaic
+%admixture in Africa" Abstract:"allow us to infer that contemporary African
+%populations contain a small proportion of genetic material (≈2%) that
+%introgressed ≈35 kya from an archaic population that split from the ancestors
+%of anatomically modern humans ≈700 kya."
 %
-%Hey: Strongest signal in Yoruba, Baka
-% Abstract: "consistent with human/archaic divergence."
+%Hey: Strongest signal in Yoruba, Baka Abstract: "consistent with human/archaic
+%divergence."
 %
-%Ragsdale: YRI
-%abstract: We estimate that an unidentified, deeply diverged population admixed with modern humans within Africa both before and after the split of African and Eurasian populations, contributing 4 − 8% genetic ancestry to individuals in world-wide populations.
+%Ragsdale: YRI abstract: We estimate that an unidentified, deeply diverged
+%population admixed with modern humans within Africa both before and after the
+%split of African and Eurasian populations, contributing 4 − 8% genetic
+%ancestry to individuals in world-wide populations.
 %
-%Lorente Galdos: Khoesan, Mbuti, W. AFR
-%title:"reveals archaic gene flow"
+%Lorente Galdos: Khoesan, Mbuti, W. AFR title:"reveals archaic gene flow"
 %
-%Durvasula: West African
-%title: "Recovering signals of ghost archaic introgression"
-%Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans."
-but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} 
-and central African \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz,Hey2018-pw}
-populations.
-This has driven speculation about the geographic range of this ghost
-population, possible links to specific fossils, and the possibility of
-finding ancient DNA evidence\citep{Hsieh2016-gk}. However, these prior
-studies share two weaknesses. First, they only contrast a single-origin model
-with an archaic hominin admixture model, leaving out other plausible models
-\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
-Second, they
-focus on a small subset of African diversity, either because of small sample
-sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
-recruited populations of recent West African or Bantu-speaking ancestry
-(Figure~\ref{fig:diversity}C). 
-While ancient DNA from Eurasia has helped to clarify early human history
-outside of Africa, there is no comparably ancient DNA to elucidate early history
-in Africa \citep{Lipson2022-xf}.
+%Durvasula: West African title: "Recovering signals of ghost archaic
+%introgression" Abstract:"Our analyses of site frequency spectra indicate that
+%these populations derive 2 to 19% of their genetic ancestry from an archaic
+%population that diverged before the split of Neanderthals and modern humans." 
+but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} and central
+African \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz,Hey2018-pw}
+populations.  This has driven speculation about the geographic range of this
+ghost population, possible links to specific fossils, and the possibility of
+finding ancient DNA evidence\citep{Hsieh2016-gk}. However, these prior studies
+share two weaknesses. First, they only contrast a single-origin model with an
+archaic hominin admixture model, leaving out other plausible models
+\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).  Second, they focus on
+a small subset of African diversity, either because of small sample sizes (2-5
+genomes) or because they rely on 1000 Genomes data which only recruited
+populations of recent West African or Bantu-speaking ancestry
+(Figure~\ref{fig:diversity}C).  While ancient DNA from Eurasia has helped to
+clarify early human history outside of Africa, there is no comparably ancient
+DNA to elucidate early history in Africa \citep{Lipson2022-xf}.
 
-We therefore aim to discriminate among a broader set of demographic
-models by studying the genomes of contemporary populations. We take as our
-starting point four classes of models (single population expansion, single
-population expansion with regional persistence, archaic hominin admixture, and
-multi-regional evolution, Figure~\ref{fig:proposed-models}), using 290
-genomes of individuals from southern, eastern, and western Africa as well as Eurasia.
-By including geographically and genetically diverse populations across Africa, we
-infer demographic models that explain more features of genetic diversity
-in more populations than previously reported. These analyses confirm the inadequacy of
-tree-like models and provide an opportunity to directly evaluate a wide range
-of alternative models.
+We therefore aim to discriminate among a broader set of demographic models by
+studying the genomes of contemporary populations. We take as our starting point
+four classes of models (single population expansion, single population
+expansion with regional persistence, archaic hominin admixture, and
+multi-regional evolution, Figure~\ref{fig:proposed-models}), using 290 genomes
+of individuals from southern, eastern, and western Africa as well as Eurasia.
+By including geographically and genetically diverse populations across Africa,
+we infer demographic models that explain more features of genetic diversity in
+more populations than previously reported. These analyses confirm the
+inadequacy of tree-like models and provide an opportunity to directly evaluate
+a wide range of alternative models.
 
@@
 South Africa, newly presented here, see Supplementary Information (SI) section
-1.2 for ethical and practical aspects of participant recruitment), Mende (from Sierra Leone, MSL from the
-Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
-Gumuz (recent descendants of a hunter-gatherer group from Ethiopia
-\citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
-(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
-populations, despite speaking distinct Afro-Asiatic languages, are highly
-genetically similar \citep{Pagani2015-pz,Gopalan2022-pw} and thus the two
-groups were combined for a larger sample size (Figure~\ref{fig:diversity}). We also
-included the British (GBR) from the 1000 Genomes Project in our demographic
-models as a representative source of back-to-Africa gene flow and recent
-colonial admixture in South Africa. Finally, we used a high-coverage ancient
-Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
-for gene flow from Neanderthals into non-Africans and gauge the relative
-time depth of divergence, assuming Neanderthals diverged $550$ka from a common
-stem. We computed one- and two-locus statistics whose expectation within and across 
-populations can be computed efficiently and that are well suited for both low- and
-high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
-maximum-likelihood inference framework, we then fit to these statistics a
-family of parameterized demographic models that involve population splits, size
-changes, continuous and variable migration rates, and punctuated admixture
-events, to learn about the nature of population structure over the past million
-years.
+1.2 for ethical and practical aspects of participant recruitment), Mende (from
+Sierra Leone, MSL from the Phase 3 1000 Genomes Project
+\citep{1000_Genomes_Project_Consortium2015-zq}), Gumuz (recent descendants of a
+hunter-gatherer group from Ethiopia \citep{Gurdasani2015-qy,Gopalan2022-pw}),
+and eastern African agriculturalists (Amhara and Oromo from Ethiopia
+\citep{Gurdasani2015-qy}). The Amhara and Oromo populations, despite speaking
+distinct Afro-Asiatic languages, are highly genetically similar
+\citep{Pagani2015-pz,Gopalan2022-pw} and thus the two groups were combined for
+a larger sample size (Figure~\ref{fig:diversity}). We also included the British
+(GBR) from the 1000 Genomes Project in our demographic models as a
+representative source of back-to-Africa gene flow and recent colonial admixture
+in South Africa. Finally, we used a high-coverage ancient Neanderthal genome
+from Vindija Cave, Croatia \citep{Prufer2017-kk} to account for gene flow from
+Neanderthals into non-Africans and gauge the relative time depth of divergence,
+assuming Neanderthals diverged $550$ka from a common stem. We computed one- and
+two-locus statistics whose expectation within and across populations can be
+computed efficiently and that are well suited for both low- and high-coverage
+genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a maximum-likelihood
+inference framework, we then fit to these statistics a family of parameterized
+demographic models that involve population splits, size changes, continuous and
+variable migration rates, and punctuated admixture events, to learn about the
+nature of population structure over the past million years.
 
@@
 (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively (log-likelihood
-($LL$) $\approx-$189,300, Table~\ref{tab:supp-single-origin}). We next
-explored a suite of parameterized models in which population structure predates
-the differentiation of contemporary groups (SI
-Section~\ref{sec:modelspec}). Depending on the parameters, these encompassed
-models allowing for ancestral reticulation, such as
-fragmentation-and-coalescence or meta-population models
+($LL$) $\approx-$189,300, Table~\ref{tab:supp-single-origin}). We next explored
+a suite of parameterized models in which population structure predates the
+differentiation of contemporary groups (SI Section~\ref{sec:modelspec}).
+Depending on the parameters, these encompassed models allowing for ancestral
+reticulation, such as fragmentation-and-coalescence or meta-population models
 (Figure~\ref{fig:proposed-models}B), archaic hominin admixture
@@
 \ref{sec:models}). These parameters were ones supported by multiple genetic and
-archaeological studies \citep{Bergstrom2021-iw}.
-Fixed parameters included the time of divergence
-between western and eastern African populations, set to $60$ka, just prior to
-the split of Eurasians and East Africans set to $50$ka. We also fixed the
-amount of admixture from Neanderthals to Europeans directly following the
-out-of-Africa migration to $1.5\%$ at $45$ka.
+archaeological studies \citep{Bergstrom2021-iw}.  Fixed parameters included the
+time of divergence between western and eastern African populations, set to
+$60$ka, just prior to the split of Eurasians and East Africans set to $50$ka.
+We also fixed the amount of admixture from Neanderthals to Europeans directly
+following the out-of-Africa migration to $1.5\%$ at $45$ka.
 
@@
 of immigrant lineages per generation over $60$ka). We further find that
-Back-to-Africa gene flow at the beginning of the Holocene primarily affected
+back-to-Africa gene flow at the beginning of the Holocene primarily affected
 the ancestors of the Ethiopian agricultural populations
-\citep{Molinaro2019-rf}, comprising over half of their genetic ancestry,
-estimated to be approximately $65\%$. We observe significant gene flow from the
-Amhara and Oromo into the Nama, a signal which is likely a proxy for the
-movement of eastern African caprid and cattle pastoralists
-\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a $25\%$ ancestry
-contribution 2,000 years ago. While this gene flow is not apparent from the
-\texttt{ADMIXTURE} plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped
-into the Khoe-San component which has drifted appreciably from its ancestral
-eastern African source. Colonial period admixture from Europeans into the Nama
-was estimated at $15\%$, similar to proportions suggested by \texttt{ADMIXTURE}
-(Figure~\ref{fig:diversity}).
+\citep{Molinaro2019-rf}, comprising close to 65\% of their genetic ancestry. We
+observe significant gene flow from the Amhara and Oromo into the Nama, a signal
+which is likely a proxy for the movement of eastern African caprid and cattle
+pastoralists \citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a
+$25\%$ ancestry contribution 2,000 years ago. While this gene flow is not
+apparent from the \texttt{ADMIXTURE} plot (Figure~\ref{fig:diversity}), the
+ancestry is likely grouped into the Khoe-San component which has drifted
+appreciably from its ancestral eastern African source. Colonial period
+admixture from Europeans into the Nama was estimated at $15\%$, similar to
+proportions suggested by \texttt{ADMIXTURE} (Figure~\ref{fig:diversity}).
 
@@
 Figure \ref{fig:mergers}: 1) One of the stem population expands (splits into
-contemporary populations), followed by continuous symmetric migration with
-the other stem population(s); or 2) one or more of the stem
-populations expands, with instantaneous pulse (or ``merger'') events from the
-other stem population, so that recent populations are formed by mergers of
-multiple ancestral populations. Depending on parameter values, this scenario
-encompasses archaic hominin introgression and fragmentation-and-coalescence
-models (such as Figure~\ref{fig:proposed-models}B and C). For many parameters,
-confidence intervals based on bootstrapping are relatively narrow
+contemporary populations), followed by continuous symmetric migration with the
+other stem population(s); or 2) one or more of the stem populations expands,
+with instantaneous pulse (or ``merger'') events from the other stem population,
+so that recent populations are formed by mergers of multiple ancestral
+populations. Depending on parameter values, this scenario encompasses archaic
+hominin introgression and fragmentation-and-coalescence models (such as
+Figure~\ref{fig:proposed-models}B and C). For many parameters, confidence
+intervals based on bootstrapping are relatively narrow
 (Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
 reflecting an informative statistical approach. However, model assumptions have
-a larger impact on parameter estimates (and thus real uncertainty).
-To convey model uncertainty, we highlight features of the two inferred models
-with high likelihoods. These are referred to as the ``multiple-merger'' and the
+a larger impact on parameter estimates (and thus real uncertainty).  To convey
+model uncertainty, we highlight features of the two inferred models with high
+likelihoods. These are referred to as the ``multiple-merger'' and the
 ``continuous-migration'' models. Both allow for migration between stem
@@
 this divergence was also poorly constrained ($95\%$ CI $276-478$ka).  These
-populations evolve independently until $119$ka ($101-125$ka) when Stem 1S and Stem
-2 combine to form the ancestors of the Nama, with proportions 30\% and 70\%,
-respectively. Similarly, Stem 1E and Stem 2 combine in equal proportions (50\%
-each) to form the ancestors of the western Africans and eastern Africans (and
-thus also all individuals who later disperse during the Out-of-Africa event).
-Finally, the Mende receive a large additional pulse of gene flow from Stem 2,
-replacing 19\% (18--21\%) of their population $25$ka ($22-26$ka). The later Stem 2
-contribution to the western African Mende resulted in better model fits
-($\Delta LL \approx 60,000$). Speculatively, this may indicate that an
+populations evolve independently until $119$ka ($101-125$ka) when Stem 1S and
+Stem 2 combine to form the ancestors of the Nama, with proportions 30\% and
+70\%, respectively. Similarly, Stem 1E and Stem 2 combine in equal proportions
+(50\% each) to form the ancestors of the western Africans and eastern Africans
+(and thus also all individuals who later disperse during the Out-of-Africa
+event).  Finally, the Mende receive a large additional pulse of gene flow from
+Stem 2, replacing 19\% (18--21\%) of their population $25$ka ($22-26$ka). The
+later Stem 2 contribution to the western African Mende resulted in better model
+fits ($\Delta LL \approx 60,000$). Speculatively, this may indicate that an
 ancestral Stem 2 population occupied western or central Africa, broadly
@@
 To assess robustness of the inferred models to analysis and reference
-population choices, SI Sections \ref{sec:validation_strategy}
-and \ref{sec:validation_results} present re-analyses changing the European and
-West African populations, as well as the recombination maps, filtering
-strategies and parameter optimization strategies. While we find some
-differences in the inferred parameters (see SI sections
-\ref{sec:replacing} and \ref{sec:change-dates}), the best fit models across all
-reanalyses are quantitatively consistent.
+population choices, SI Sections \ref{sec:validation_strategy} and
+\ref{sec:validation_results} present re-analyses changing the European and West
+African populations, as well as the recombination maps, filtering strategies
+and parameter optimization strategies. While we find some differences in the
+inferred parameters (see SI sections \ref{sec:replacing} and
+\ref{sec:change-dates}), the best fit models across all reanalyses are
+quantitatively consistent.
 
@@
 models, including the single-origin model, recapitulate an inferred ancestral
-increase in $N_e$ between $100$ka--$1$Ma
-(Figure~\ref{fig:supp-iicr-sim} and SI
-Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
-increase in $N_e$ during that period, the best-fit models recapitulate this
-pattern without corresponding population size changes.
+increase in $N_e$ between $100$ka--$1$Ma (Figure~\ref{fig:supp-iicr-sim} and SI
+Section~\ref{inferred_relate}). Whereas the single-origin model achieves this
+by an increase in $N_e$ during that period, the best-fit models recapitulate
+this pattern without corresponding population size changes.
 
 Relative cross-coalescence rates (rCCR) have recently been used to estimate
-divergence between pairs of populations, as measured by the rate of
-coalescence between two groups divided by the mean within population
-coalescence. Simulations of rCCR accuracy, however, focus on a ‘clean split’
-between populations whereby groups diverge without subsequent gene flow.
-Published estimates of the earliest human divergences with rCCR, which range
-from $150$ka--$100$ka \citep{Bergstrom2021-iw}, may be significantly biased when
+divergence between pairs of populations, as measured by the rate of coalescence
+between two groups divided by the mean within population coalescence.
+Simulations of rCCR accuracy, however, focus on a ‘clean split’ between
+populations whereby groups diverge without subsequent gene flow.  Published
+estimates of the earliest human divergences with rCCR, which range from
+$150$ka--$100$ka \citep{Bergstrom2021-iw}, may be significantly biased when
 compared to more complex models with gene flow as inferred here. We find that
@@
 population structure, we sought to capture model uncertainty by exploring
-multiple parameterizations of early history. The best fit models presented
-here include reticulation and migration between early human populations rather
-than archaic hominin admixture from long-isolated branches
-(Figure~\ref{fig:proposed-models}C). 
-Elements of a recent single-origin or African multi-regionalism
-(Figure~\ref{fig:proposed-models}A,D)
-feature in our best fit models, as indicated in the recent time of
-contemporary population divergence
-or the gene flow between disparate stems, respectively.
+multiple parameterizations of early history. The best fit models presented here
+include reticulation and migration between early human populations rather than
+archaic hominin admixture from long-isolated branches
+(Figure~\ref{fig:proposed-models}C). Elements of a recent single-origin or
+African multi-regionalism (Figure~\ref{fig:proposed-models}A,D) feature in our
+best fit models, as indicated in the recent time of contemporary population
+divergence or the gene flow between disparate stems, respectively.
 
 We caution that we cannot rule out that more complex models involving
-additional stems, more complex population structure, or
-hybrid models including both weak structure and archaic hominin admixture may
-better explain the data. Because parameters related to the split time,
-migration rates, and relative sizes of the early stems were variable across
-models, reflecting a degree of confounding among these parameters, we refrained
-from introducing additional branches associated with more parameters during
-that period. Rather than interpreting the two stems as representing
-well-defined and stable populations over hundreds of thousands of years, we
-interpret the weakly structured stem as consistent with a population
-coalescence and fragmentation model \citep{Scerri2019-xg}.
-Additional African populations such as those 
-from Central Africa, other Khoe-San groups or pre-Holocene ancient DNA samples,
+additional stems, more complex population structure, or hybrid models including
+both weak structure and archaic hominin admixture may better explain the data.
+Because parameters related to the split time, migration rates, and relative
+sizes of the early stems were variable across models, reflecting a degree of
+confounding among these parameters, we refrained from introducing additional
+branches associated with more parameters during that period. Rather than
+interpreting the two stems as representing well-defined and stable populations
+over hundreds of thousands of years, we interpret the weakly structured stem as
+consistent with a population coalescence and fragmentation model
+\citep{Scerri2019-xg}. Additional African populations such as those from
+Central Africa, other Khoe-San groups or pre-Holocene ancient DNA samples,
 could further test our proposed models.
@@
 stem. For example, Stem 1S contributes 70\% to the ancestral formation of the
-Khoe-San. The extent of the isolation $\>400$ka between Stem 1S, Stem 1E and Stem
-2 suggests that these stems were not proximate to each other. While the length
-of isolation among the stems is variable across fits, models with a period of
-divergence, isolation and then a merger event (i.e., a ``reticulation'')
-out-performed models with bifurcating divergence and continuous gene flow.
+Khoe-San. The extent of the isolation $\>400$ka between Stem 1S, Stem 1E and
+Stem 2 suggests that these stems were not proximate to each other. While the
+length of isolation among the stems is variable across fits, models with a
+period of divergence, isolation and then a merger event (i.e., a
+``reticulation'') out-performed models with bifurcating divergence and
+continuous gene flow.
 
@@
 northward thereby encountering Stem 2, or Stem 2 moved from central/western
-Africa southward into Stem 1S, then we could observe disproportionate
-ancestry contributions from different stems in contemporary groups. We observed
-two merger events. The first, between Stem 1S and Stem 2, resulted in the
-formation of an ancestral Khoe-San population $\sim120$ka. The second event
-$\sim100$ka between Stem 1E and Stem 2, resulted in the formation of the
-ancestors of eastern/western Africans including the ancestors of
-``Out-of-Africans''. Reticulated models do not have a unique and 
-well-defined ``basal'' human population divergence.
-We suggest conceptualizing the events $120$ka as the time of most recent
-shared ancestry among sampled populations.
-However, interpreting population divergence times in population genetics is 
-always challenging due
-to the co-estimation of divergence time and subsequent migration;
-methods assuming clean vs reticulated splits can infer different split dates
-(Figure~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
-Therefore, wide variation exists in estimates of divergence time in the
+Africa southward into Stem 1S, then we could observe disproportionate ancestry
+contributions from different stems in contemporary groups. We observed two
+merger events. The first, between Stem 1S and Stem 2, resulted in the formation
+of an ancestral Khoe-San population $\sim120$ka. The second event $\sim100$ka
+between Stem 1E and Stem 2, resulted in the formation of the ancestors of
+eastern/western Africans including the ancestors of ``Out-of-Africans''.
+Reticulated models do not have a unique and well-defined ``basal'' human
+population divergence.  We suggest conceptualizing the events at $120$ka as the
+time of most recent shared ancestry among sampled populations. However,
+interpreting population divergence times in population genetics is always
+challenging due to the co-estimation of divergence time and subsequent
+migration; methods assuming clean vs reticulated splits can infer different
+split dates (Figure~\ref{fig:supp-rccr-data} and
+Figure~\ref{fig:supp-IM-misspecification}). Therefore, wide variation exists in
+estimates of divergence time in the
 literature\citep{Bergstrom2021-iw,Henn2018-rf}.
 
-Shifts in wet/dry conditions across the African continent $140$ka to $100$ka may have
-promoted these merger events between divergent stems. 
-Precipitation does not neatly track interglacial cycles in Africa,
-and heterogeneity across regions may mean that the beginning of an arid period
-in eastern Africa is conversely the start of a wet period in
-southern Africa\citep{Blome2012-lw}.
+Shifts in wet/dry conditions across the African continent $140$ka to $100$ka
+may have promoted these merger events between divergent stems.  Precipitation
+does not neatly track interglacial cycles in Africa, and heterogeneity across
+regions may mean that the beginning of an arid period in eastern Africa is
+conversely the start of a wet period in southern Africa\citep{Blome2012-lw}.
 The rapid rise in sea levels during the MIS 5e interglacial might have
-triggered migration inland away from the
-coasts, as has been suggested, e.g., for the Paleo-Agulhas plain\citep{Marean2014-pg}.
-Following these merger events, the stems subsequently fractured into
-subpopulations which then persist over the past $\sim120$ka. These
-subpopulations can be linked to contemporary groups despite subsequent gene
-flow across the continent; for example, a genetic lineage sampled in the Gumuz
-has a $0.70$ probability of being inherited from the ancestral ``eastern''
-subpopulation $55$ka versus $0.06$ probability of being inherited
-from the ``southern'' subpopulation (see Table~\ref{tab:supp-lineage-probabilitys} for
-additional comparisons).
+triggered migration inland away from the coasts, as has been suggested, e.g.,
+for the Paleo-Agulhas plain\citep{Marean2014-pg}. Following these merger
+events, the stems subsequently fractured into subpopulations which then persist
+over the past $\sim120$ka. These subpopulations can be linked to contemporary
+groups despite subsequent gene flow across the continent; for example, a
+genetic lineage sampled in the Gumuz has a $0.70$ probability of being
+inherited from the ancestral ``eastern'' subpopulation $55$ka versus $0.06$
+probability of being inherited from the ``southern'' subpopulation (see
+Table~\ref{tab:supp-lineage-probabilitys} for additional comparisons).
 
@@
 the inferred Out-of-Africa and eastern/western African divergences which were
-10--15ka younger than our fixed parameters. These younger dates are clearly at 
-odds with the accepted timing of Out-of-Africa expansion at approximately $\>50$ka, 
-as based on archaeological, climatic and paleo-morphological information
-\citep{Fu2014-am,Groucutt2015-do,Prufer2021-ao,Beyer2021-ym}. 
-We interpret the free estimate for eastern African vs. European divergence as reflecting 
-our inclusion of only a single Out-of-Africa population in the model, the lack of a nearby 
-source for back-to-Africa gene flow, and other regionally complex parameters. 
-However, older pan-African features of our inferred models are weakly affected by 
+10--15ka younger than our fixed parameters. These younger dates are at odds
+with the accepted timing of the Out-of-Africa expansion that contributed to
+later human populations at approximately $\>50$ka, as based on archaeological,
+climatic and fossil information \citep{Fu2014-am,Groucutt2015-do,
+Prufer2021-ao,Beyer2021-ym}. We interpret the free estimate for eastern African
+vs. European divergence as reflecting our inclusion of only a single
+Out-of-Africa population in the model, the lack of a nearby source for
+back-to-Africa gene flow, and other regionally complex parameters. However,
+older pan-African features of our inferred models are minimally affected by
 this analysis choice (SI Section~\ref{sec:models}).
@@
 
-Evidence for archaic hominin admixture in Eurasia has bolstered the plausibility of 
-archaic hominin admixture having also occurred in Africa. Previous work that sought to 
-explain patterns of polymorphism inconsistent with a single-origin model thus focused on 
-archaic hominin admixture as an alternative model, by referring to additional (or ``ghost'') 
-branches required to fit the data as ``archaic'' 
-\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,
-Durvasula2020-td} and assuming (or inferring) deep divergences. These perspectives have 
-oriented interpretations of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil 
-(e.g., evolution of early \emph{Homo sapiens} \cite{Grun2020-su}) data. Here, we have shown 
-that a model of a weakly structured stem better captures the apparently inconsistent patterns 
-of polymorphisms.
+Evidence for archaic hominin admixture in Eurasia has bolstered the
+plausibility of archaic hominin admixture having also occurred in Africa.
+Previous work that sought to explain patterns of polymorphism inconsistent with
+a single-origin model thus focused on archaic hominin admixture as an
+alternative model, by referring to additional (or ``ghost'') branches required
+to fit the data as ``archaic'' \citep{Plagnol2006-lt,Hammer2011-bx,
+Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz, Durvasula2020-td} and
+assuming (or inferring) deep divergences. These perspectives have oriented
+interpretations of both genomic (e.g., selection \cite{Wall2019-ao}) and fossil
+(e.g., evolution of early \emph{Homo sapiens} \cite{Grun2020-su}) data. Here,
+we have shown that a model of a weakly structured stem better captures the
+apparently inconsistent patterns of polymorphisms.
 
-Preferring weakly-structured-stem models over archaic-admixture models potentially has a 
-range of implications. First, with a weakly structured stem, there is no need to posit that 
-an archaic hominin population in Africa stayed reproductively isolated from the ancestral 
-human lineage for hundreds of thousands of years before the initiation of gene flow. Instead, 
-there would simply have been continuous or recurrent contact between two or more groups 
-present in Africa.
+Preferring weakly-structured-stem models over archaic-admixture models has a
+range of implications. First, with a weakly structured stem, there is no need
+to posit that an archaic hominin population in Africa stayed reproductively
+isolated from the ancestral human lineage for hundreds of thousands of years
+before the initiation of gene flow. Instead, there would simply have been
+continuous or recurrent contact between two or more groups present in Africa.
 
@@
 \emph{EPAS1} among Tibetans \citep{Zhang2021-xx}. Under previous
-African-archaic-hominin admixture models, the estimated 8--10\% introgression rate is
-much higher than Neanderthal gene flow, and would have plausibly been fertile
-ground for dramatic selection for or against archaic-hominin-derived
+African-archaic-hominin admixture models, the estimated 8--10\% introgression
+rate is much higher than Neanderthal gene flow, and would have plausibly been
+fertile ground for dramatic selection for or against archaic-hominin-derived
 haplotypes\citep{Wall2019-ao}. By contrast, adaptation under a weakly
@@
 
-Third, multiple studies have shown a correspondence between phenotypic differentiation, 
-usually assessed with measurements of the cranium, and genetic differentiation among 
-human populations and between humans and Neanderthals \citep{Relethford1994-mh,
-Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also SI Section~\ref{sec:morpho_diff}). 
-This correspondence potentially allows predictions of our model to be related to the 
-fossil record. Some \emph{Homo sapiens} fossils, from Iho Eleru, Nigeria (13ka) 
-\citep{Harvati2011-xs}, Ishango, 
-Democratic Republic of Congo (20-25ka) \citep{Crevecoeur2016-cl}, and Nazlet Khater, 
-Egypt (35-40ka) \citep{Crevecoeur2012-we}, 
-have morphological features that may reflect relatively recent gene flow from archaic hominins \citep{Harvati2011-xs, Crevecoeur2016-cl}, which has been used in support of previously inferred archaic admixture events in human ancestry \cite{Scerri2018-nl, Durvasula2020-td, Bergstrom2021-iw}.
-The weakly structured stem model is not incompatible with archaic admixture having occurred in the ancestry of these fossils, but would imply, by contrast, that such individuals are unlikely to have contributed much ancestry to  contemporary humans.
-The fossil record of Africa is 
-sparse during the earlier time period of the stems \aprcomment{($\gtrsim120$ka)},
-but of the fossils that date to this time period, 
-some are fairly similar overall in morphology to contemporary humans (e.g., Omo 1 from 
-Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are similar in some morphological 
-features to contemporary humans (e.g., Irhoud 1 from Jebel Irhoud, Morocco 
-\citep{Hublin2017-cq,Richter2017-zu}), and others are different enough in morphology to 
-have been assigned to species other than \emph{Homo sapiens} (e.g., DH1 from Dinaledi, 
-South Africa \citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic 
-differences between the stems were comparable to those among contemporary human populations, 
-the most morphologically divergent fossils are also unlikely to represent branches that 
-contributed appreciably to contemporary human ancestries.
+Third, multiple studies have shown a correspondence between phenotypic
+differentiation, usually assessed with measurements of the cranium, and genetic
+differentiation among human populations and between humans and Neanderthals
+\citep{Relethford1994-mh, Weaver2008-ho,Von_Cramon-Taubadel2009-zb} (see also
+SI Section~\ref{sec:morpho_diff}).  This correspondence potentially allows
+predictions of our model to be related to the fossil record. Some \emph{Homo
+sapiens} fossils, from Iho Eleru, Nigeria (13ka) \citep{Harvati2011-xs},
+Ishango, Democratic Republic of Congo (20-25ka) \citep{Crevecoeur2016-cl}, and
+Nazlet Khater, Egypt (35-40ka) \citep{Crevecoeur2012-we}, have morphological
+features that may reflect relatively recent gene flow from archaic hominins
+\citep{Harvati2011-xs, Crevecoeur2016-cl}, which has been used in support of
+previously inferred archaic admixture events in human ancestry
+\cite{Scerri2018-nl, Durvasula2020-td, Bergstrom2021-iw}.  The weakly
+structured stem model is not incompatible with archaic admixture having
+occurred in the ancestry of these fossils, but would imply, by contrast, that
+such individuals are unlikely to have contributed much ancestry to contemporary
+humans. The fossil record of Africa is sparse during the earlier time period of
+the stems ($\gtrsim200$ka), but of the fossils that date to this time period,
+some are fairly similar overall in morphology to contemporary humans (e.g., Omo
+1 from Omo Kibish, Ethiopia \citep{Day1969-rh,Vidal2022-qe}), others are
+similar in some morphological features to contemporary humans (e.g., Irhoud 1
+from Jebel Irhoud, Morocco \citep{Hublin2017-cq,Richter2017-zu}), and others
+are different enough in morphology to have been assigned to species other than
+\emph{Homo sapiens} (e.g., DH1 from Dinaledi, South Africa
+\citep{Berger2015-bq,Dirks2017-uk}). If, as our model predicts, the genetic
+differences between the stems were comparable to those among contemporary human
+populations, the most morphologically divergent fossils are also unlikely to
+represent branches that contributed appreciably to contemporary human
+ancestries.
 
@@
 thank Luke Anderson-Trocm\'e and Georgette Femerling for assistance in creating
-the map in Figure~\ref{fig:diversity} and Figure \ref{fig:toymodels},
+the map in Figure~\ref{fig:diversity} and Figure~\ref{fig:toymodels},
 respectively. We thank Nuria Melisa Morales-Garcia for extensive data
-visualization discussion and designing Figures \ref{fig:proposed-models} and
+visualization discussion and designing Figures~\ref{fig:proposed-models} and
 \ref{fig:best-fit-models}. This research was supported by CIHR project grant
@@
 
-\subsection*{Gene genalogy reconstruction}
+\subsection*{Gene genealogy reconstruction}
 
@@
         which reflect colonial-period migration into the Nama,
-        Back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
+        back-to-Africa gene flow among some Ethiopians, and Khoe-San admixture in the Zulu.
     }
@@
     \caption{
-        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
-        From the best fit models of our two parameterizations (A and B:
-        continuous migration, C and D: merger with stem migration),
-        we predicted pairwise differences $H_{i,j}$ between individuals
-        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
-        (B and D) To understand how drift between stems explains contemporary structure, 
-        we also computed the proportion $\alpha^2$ of drift between pairs of
-        sampled contemporary
-        populations that aligns with drift between past populations
-        (here Nama and Mende, see SI Section~S\ref{sec:f4}
-        for details and additional comparisons in
+        \textbf{Structure among stems is weak and present-day structure is
+        mostly recent.} From the best fit models of our two parameterizations
+        (A and B: continuous migration, C and D: merger with stem migration),
+        we predicted differentiation and shared drift between populations at
+        past times. (A, C) We computed expected pairwise differences $H_{i,j}$
+        between individuals sampled from populations $i$ and $j$ existing at
+        time $t$. (B, D) To understand how drift between stems explains
+        contemporary structure, we computed the proportion $\alpha^2$ of drift
+        between pairs of sampled contemporary populations that aligns with
+        drift between past populations (here Nama and Mende, see SI
+        Section~S\ref{sec:f4} for details and additional comparisons in
         Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
         Both models infer deep population structure with modest contributions to
-        contemporary genetic differentiation.
-        Most present differentiation dates back to the last $100$ka.
+        contemporary genetic differentiation.  Most present-day differentiation
+        dates back to the last $100$ka.
     }
@@
         curves (RCCR) into population divergence times is especially prone
-        to misinterpretation. For example, the Mende-Gumuz comparison shows
-        a more recent increased RCCR than either population with the British,
-        a pattern that is recapitulated under our best-fit model, even though
-        the Mende-Gumuz split occurs prior to the Gumuz-British split.
+        to misinterpretation.
+        For example, in our model, the Mende-Gumuz split occurs prior to the
+        Gumuz-British split. However, the model also predicts recent elevated
+        Mende-Gumuz RCCR. This pattern, also observed in the data, does not
+        indicate that the Mende and Gumuz split more recently.
     }

### 2023-03-14 supplement.tex
simon comments in supp

@@
 proportions, and population sizes and size changes. Thus, parameter space for a
-given model topology is large. In addition, the space of possible model
-topologies is itself large – as the number of populations increases, the number
-of possible topologies also increases, as there are more possibilities for the
-order of divergence and admixture events.
+given model topology is large. In addition, as the number of populations
+increases, the number of possible topologies also increases, as there are more
+possibilities for the order of divergence and admixture events.
 
@@
     \item Mende population expansion: 5ka \citep{Tennessen2012-oy}
-    \item Gumuz population decline: 5ka
-    %We actually don't have a citation for this. Leaving blank. 
+    \item Gumuz population decline: 5ka \citep{Gopalan2022-pw}
     \item East African pastoralist to South African admixture: 2ka \citep{Henn2008-xo}
@@
 \end{itemize}
-The total number of parameters that were ultimately inferred were 21-31, depending on the complexity
-of the model.
+The total number of parameters that were ultimately inferred were twenty-one to
+thirty-one, depending on the complexity of the model.
 

### 2023-03-15 supplement.tex
minor edits
question

@@
 evidence of the Initial Upper Paleolithic occurs 39-45ka as far east as Lake
-Baikal. Characteristics of the IUP support a coherent, cultural package which
+Baikal. Characteristics of the IUP \sgcomment{define?} support a coherent, cultural package which
 may have been transmitted by demic expansion \citep{Kuhn2014-kj,Zwyns2021-zv}.

### 2023-03-15 paper.tex

@@
 the inferred Out-of-Africa and eastern/western African divergences which were
-10--15ka younger than our fixed parameters. These younger dates are at odds
-with the accepted timing of the Out-of-Africa expansion that contributed to
-later human populations at approximately $\>50$ka, as based on archaeological,
-climatic and fossil information \citep{Fu2014-am,Groucutt2015-do,
-Prufer2021-ao,Beyer2021-ym}. We interpret the free estimate for eastern African
-vs. European divergence as reflecting our inclusion of only a single
-Out-of-Africa population in the model, the lack of a nearby source for
-back-to-Africa gene flow, and other regionally complex parameters. However,
-older pan-African features of our inferred models are minimally affected by
-this analysis choice (SI Section~\ref{sec:models}).
+10–15ka younger than our fixed parameters. These younger dates are at odds with
+the accepted timing of the Out-of-Africa expansion that contributed to later
+human populations at approximately 50ka, as based on archaeological, climatic
+and fossil information \citep{Fu2014-am,Groucutt2015-do,
+Prufer2021-ao,Beyer2021-ym}. Because the inference approach is unbiased in
+simulations, we interpret the free estimate for eastern African vs. European
+divergence as reflecting our inclusion of only a single Out-of-Africa
+population in the model, the lack of a nearby source for back-to-Africa gene
+flow, and other regionally complex parameters, rather than to a systematic bias
+that may affect all parameters in the model. Older pan-African features of our
+inferred models are minimally affected by the choice of these fixed parameters
+(SI Section~\ref{sec:change-dates}).
 

### 2023-03-15 supplement.tex
abbreviation

--- 2023-03-14 supplement.tex
@@
 secluded southern refugia, decreasing the probability of migrations through the
-Sinai land bridge \citep{Beyer2021-ym}. The northern route OOA remains a likely
-barrier to human migration (as modeled by precipitation amount) until the
-Holocene. A southern route across the Red Sea becomes possible during MIS 4
-(70ka) until 30ka \citep{Beyer2021-ym}. Even during MIS 4, there were few
-plausible routes out of the southern Arabian Peninsula, except along the Gulf
-of Oman or gaps in the Arabian Desert through the Zagros Mountains. Neanderthal
-remains dating to this period have been identified in the Zagros Mountains,
-adding a further potential barrier to expansion
+Sinai land bridge \citep{Beyer2021-ym}. This barrier to human migration OOA
+via the northern route likely remained until the Holocene (as modeled by
+precipitation amount).  A southern route across the Red Sea becomes possible
+during MIS 4 (70ka) until 30ka \citep{Beyer2021-ym}. Even during MIS 4, there
+were few plausible routes out of the southern Arabian Peninsula, except along
+the Gulf of Oman or gaps in the Arabian Desert through the Zagros Mountains.
+Neanderthal remains dating to this period have been identified in the Zagros
+Mountains, adding a further potential barrier to expansion
 \citep{Yousefi2020-et,Heydari-Guran2021-xy}.
@@
 approximately 40ka \citep{Deviese2017-de,Hublin2017-uy}. Similarly in Asia,
-evidence of the Initial Upper Paleolithic occurs 39-45ka as far east as Lake
+evidence of the Initial Upper Paleolithic (IUP) occurs 39-45ka as far east as Lake
 Baikal. Characteristics of the IUP support a coherent, cultural package which

### 2023-03-16 supplement.tex
Address Tim's comments in supp

@@
 variants on an early onset dispersal model. There is persuasive
-paleo-morphological evidence for a Homo sapiens presence in Arabia between
-120--85ka, coincident with Marine Isotope Stage (MIS) 5. The crania identified
-at Skhul and Qafzeh in the Levant, as well a finger bone from southern Arabian
-Peninsula point to a dispersal OOA when conditions in North Africa facilitated
-a connection through the Sinai land bridge, and/or a possible connection via
-crossing the Red Sea. Ecological conditions would have been similar between
-North Africa and the Arabian Peninsula at this time, with greater abundance in
-game during the higher rainfall MIS 5. While Skhul and Qafzeh are taxonomically
-assigned to Homo sapiens, we note that their morphological diversity (i.e. a
-mix of derived and archaic traits) is unique among contemporary human
-populations. Under some formulations of the early onset dispersal model,
-archaeologists posit a further expansion of humans into South Asia or as far as
-southeastern Asia \citep{Groucutt2015-do}.
-
-However, the replacement of Homo sapiens by Neanderthals in the Levant (e.g.
-Kebara 2) complicates the persistence of modern humans in the Arabian Peninsula
-after 85ka. The end of MIS 5 coincides with increasing aridity in the region
-which would have rendered much the Peninsula uninhabitable, except for a
-secluded southern refugia, decreasing the probability of migrations through the
-Sinai land bridge \citep{Beyer2021-ym}. This barrier to human migration OOA
-via the northern route likely remained until the Holocene (as modeled by
-precipitation amount).  A southern route across the Red Sea becomes possible
-during MIS 4 (70ka) until 30ka \citep{Beyer2021-ym}. Even during MIS 4, there
-were few plausible routes out of the southern Arabian Peninsula, except along
-the Gulf of Oman or gaps in the Arabian Desert through the Zagros Mountains.
-Neanderthal remains dating to this period have been identified in the Zagros
-Mountains, adding a further potential barrier to expansion
-\citep{Yousefi2020-et,Heydari-Guran2021-xy}.
+paleoanthropological evidence for a \emph{Homo sapiens} presence in Arabia
+between 85--120ka, coincident with Marine Isotope Stage (MIS) 5. The crania
+identified at Skhul and Qafzeh in the Levant, as well a finger bone from
+southern Arabian Peninsula point to a dispersal OOA when conditions in North
+Africa facilitated a connection through the Sinai land bridge, and/or a
+possible connection via crossing the Red Sea. Ecological conditions would have
+been similar between North Africa and the Arabian Peninsula at this time, with
+greater abundance in game during the higher rainfall MIS 5. While Skhul and
+Qafzeh are taxonomically assigned to \emph{Homo sapiens}, we note that they are
+morphologically quite different from contemporary populations
+\citep{Pearson2008-ok}. Under some formulations of the early onset dispersal
+model, archaeologists posit a further expansion of humans into South Asia or as
+far as southeastern Asia \citep{Groucutt2015-do}.
+
+However, the replacement of \emph{Homo sapiens} by Neanderthals in the Levant
+(e.g.  Kebara 2) complicates the persistence of modern humans in the Arabian
+Peninsula after 85ka \citep{Valladas1987-bk}. The end of MIS 5 coincides with
+increasing aridity in the region which would have rendered much the Peninsula
+uninhabitable, except for a secluded southern refugia, decreasing the
+probability of migrations through the Sinai land bridge \citep{Beyer2021-ym}.
+This barrier to human migration OOA via the northern route likely remained
+until the Holocene (as modeled by precipitation amount). A southern route
+across the Red Sea becomes possible during MIS 4 (70ka) until 30ka
+\citep{Beyer2021-ym}. Even during MIS 4, there were few plausible routes out of
+the southern Arabian Peninsula, except along the Gulf of Oman or gaps in the
+Arabian Desert through the Zagros Mountains. Neanderthal remains dating to this
+period have been identified in the Zagros Mountains, adding a further potential
+barrier to expansion \citep{Yousefi2020-et,Heydari-Guran2021-xy}.
 
@@
 successful expansion would likely post-date 65ka. Morphological and
-archaeological evidence for Homo sapiens in the Near East reappears by 50ka in
-the Levant with the Initial Upper Paleolithic stone tool industry
+archaeological evidence for \emph{Homo sapiens} in the Near East reappears by
+50ka in the Levant with the Initial Upper Paleolithic stone tool industry
 \citep{Boaretto2021-oi} and a cranium from Manot Cave
@@
 \citep[e.g.,][]{Fewlass2020-se}, morphological \citep[e.g.,][]{Hublin2020-nk},
-and genetic data \citep[e.g.,][]{Prufer2021-ao} all indicate that about 45ka
+and genetic \citep[e.g.,][]{Prufer2021-ao} data all indicate that about 45ka
 \emph{Homo sapiens} had expanded into eastern Europe and within 5,000 years
@@
 approximately 40ka \citep{Deviese2017-de,Hublin2017-uy}. Similarly in Asia,
-evidence of the Initial Upper Paleolithic (IUP) occurs 39-45ka as far east as Lake
-Baikal. Characteristics of the IUP support a coherent, cultural package which
-may have been transmitted by demic expansion \citep{Kuhn2014-kj,Zwyns2021-zv}.
-By 45ka, there is additionally an unequivocal \emph{H. sapiens} femur from
-southern Siberia which contains derived alleles shared with later Eurasian
-populations \citep{Fu2014-am}. The Ust’-Ishim
-specimen contains evidence of admixture with Neanderthals, at a level similar
-to contemporary East Asian genomes. This supports estimates of Neanderthal
-admixture occurring earlier than 45ka, estimated to be by 47-65ka as inferred
-from Neanderthal tract lengths \citep{Sankararaman2012-uw}.
+evidence of the Initial Upper Paleolithic (IUP) occurs 39-45ka as far east as
+Lake Baikal. Characteristics of the IUP support a coherent, cultural package
+which may have been transmitted by demic expansion
+\citep{Kuhn2014-kj,Zwyns2021-zv}.  By 45ka, there is additionally an
+unequivocal \emph{Homo sapiens} femur from southern Siberia which contains
+derived alleles shared with later Eurasian populations \citep{Fu2014-am}. The
+Ust’-Ishim specimen contains evidence of admixture with Neanderthals, at a
+level similar to contemporary East Asian genomes. This supports estimates of
+Neanderthal admixture occurring earlier than 45ka, estimated to be by 47-65ka
+as inferred from Neanderthal tract lengths \citep{Sankararaman2012-uw}.
 
@@
 unlikely to have contributed to the ancestors of most Eurasians. By 50ka, a
-Homo sapiens population occupied the Levant, subsequently admixed and replaced
-Neanderthals, and is a likely source for the Eurasian populations. Hence, we
-fixed the eastern African/Eurasian divergence at 50ka, though arguably dates
-even earlier up to 65ka would be appropriate (see SI Section 7.2).
+\emph{Homo sapiens} population occupied the Levant, subsequently admixed and
+replaced Neanderthals, and is a likely source for the Eurasian populations.
+Hence, we fixed the eastern African/Eurasian divergence at 50ka, though
+arguably dates even earlier up to 65ka would be appropriate (see SI Section
+7.2).
 
